---
title: "Lab Assignment 7: Conjoint Analysis"
subtitle: "Week 15"
author: "Your Name Here"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    theme:
      - lecture.scss
      - cosmo
    highlight-style: github
    mainfont: "Clear Sans, -apple-system, system-ui, sans-serif"
    fontsize: 1.1em
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Assignment Overview

-   **Total Points**: 10 points
-   **Due Date**: April 26, 2026
-   **Submission**: Upload your completed .qmd (no need to submit the rendered HTML file). I'll render your .qmd to HTML to check your work and assign points, so please make sure your file renders without errors before submitting.

## Learning Objectives

This assignment assesses your ability to:

-   Fit and interpret a rating-based conjoint model using `lm()`
-   Calculate attribute importance scores from part-worth utilities
-   Compute willingness-to-pay (WTP) estimates
-   Predict market shares using proportional allocation
-   Fit and interpret a choice-based conjoint model using `mlogit()`

## Instructions

1.  Complete all 3 tasks below
2.  Fill in the blanks (replace `___` with appropriate code or values)
3.  After completing each code chunk, change `#| eval: false` to `#| eval: true` to run it
4.  Answer all interpretation questions (1-3 sentences each)
5.  Render the document and review the output to confirm all code runs correctly and all parts (code, plots, interpretations) are complete and accurate before submitting

------------------------------------------------------------------------

# Grading Rubric

| Task | Points | Breakdown |
|----|----|----|
| Task 1: Rating-Based Conjoint | 4 | 0.5 model + 0.5 importance + 0.5 importance plot + 0.5 WTP + 2 interpretation |
| Task 2: Market Share Prediction | 3 | 0.5 prediction + 0.5 market share + 2 interpretation |
| Task 3: Choice-Based Conjoint | 3 | 0.5 data prep + 0.5 model + 2 interpretation |
| **Total** | **10** |  |

**Interpretation Guidance**: Your interpretations should be brief but specific. For example:

-   **Good**: "The WTP for extended hours is \$45.00, meaning patients would pay \$45 more per visit for a clinic open 7am-11pm instead of 9am-5pm. This suggests that convenience of access has meaningful monetary value to patients, which could inform staffing and scheduling investments."
-   **Too vague**: "Patients prefer extended hours."

------------------------------------------------------------------------

# Required Packages

-   Load the required packages using the code below:

```{r}
#| label: required-packages

library(tidyverse)
library(mlogit)
library(dfidx)
```

# Assignment Tasks

## Task 1: Rating-Based Conjoint: Urgent Care Center Design (4 points)

A healthcare system is planning to open a new **urgent care center** and wants to understand patient preferences. They conducted a conjoint survey where 120 respondents each rated 16 hypothetical urgent care profiles (full factorial design: 2×2×2×2) on a 1-10 scale.

The study examines four attributes:

| Attribute | Level 1 | Level 2 | Description |
|----|----|----|----|
| Wait Time | 15 min | 45 min | Expected wait before being seen |
| Hours | Extended (7am-11pm) | Standard (9am-5pm) | Operating hours |
| Provider Type | Physician (MD) | Nurse Practitioner (NP) | Who provides care |
| Visit Cost | \$75 | \$150 | Out-of-pocket cost per visit |

Load the data using the code below and review the key variables.

```{r}
#| label: task1-data

urgentcare_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/lab7_urgentcare_ratings.csv")

head(urgentcare_data, 5)
```

**Key variables**:

| Variable | Description | Type |
|----|----|----|
| `wait_15min` | Short wait indicator (1 = 15 min, 0 = 45 min) | Binary |
| `hours_extended` | Extended hours indicator (1 = 7am-11pm, 0 = 9am-5pm) | Binary |
| `provider_md` | Physician indicator (1 = MD, 0 = NP) | Binary |
| `visit_cost` | Out-of-pocket cost per visit (\$75 or \$150) | Continuous |
| `rating` | Respondent's rating on 1-10 scale — **outcome variable** | Continuous |

**Step 1 (Model Fitting: 0.5 points)**: Fit a linear regression model to estimate part-worth utilities for each attribute.

```{r}
#| label: task1-model
#| eval: false

# Fit rating-based conjoint model
model_uc <- lm(___ ~ ___ + ___ + ___ + ___,
               data = ___)

summary(___)
```

**Step 2 (Attribute Importance: 0.5 points)**: Calculate importance scores for each attribute.

```{r}
#| label: task1-importance
#| eval: false

# Extract coefficients
coef_uc <- coef(model_uc)

# Calculate utility ranges
range_wait <- abs(coef_uc["___"] * 1)          # Binary: 15 min vs 45 min
range_hours <- abs(coef_uc["___"] * 1)         # Binary: Extended vs Standard
range_provider <- abs(coef_uc["___"] * 1)      # Binary: MD vs NP
range_cost <- abs(coef_uc["___"] * (150 - 75)) # $75 cost difference

# Calculate total range and importance percentages
total_range <- range_wait + range_hours + range_provider + range_cost

importance_uc <- data.frame(
  attribute = c("Wait Time", "Operating Hours", "Provider Type", "Visit Cost"),
  utility_range = c(range_wait, range_hours, range_provider, range_cost),
  importance_pct = c(range_wait, range_hours, range_provider, range_cost) / total_range * 100
)

importance_uc
```

**Step 3 (Importance Visualization: 0.5 points)**: Create a horizontal bar chart of attribute importance.

```{r}
#| label: task1-importance-plot
#| eval: false

ggplot(importance_uc, aes(x = reorder(___, importance_pct),
                           y = ___)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(aes(label = paste0(round(importance_pct, 1), "%")),
            hjust = -0.1, size = 4) +
  coord_flip() +
  labs(title = "Attribute Importance: Urgent Care Center Design",
       x = "",
       y = "Relative Importance (%)") +
  theme_minimal() +
  ylim(0, max(importance_uc$importance_pct) * 1.2)
```

**Step 4 (Willingness to Pay: 0.5 points)**: Calculate how much patients would pay **per visit** for each attribute improvement using the WTP formula: $\text{WTP} = -\frac{\beta_{\text{attribute}}}{\beta_{\text{price}}}$. Since `visit_cost` is measured per visit, the resulting WTP values represent the additional dollar amount a patient would pay on a single visit to obtain that attribute improvement.

```{r}
#| label: task1-wtp
#| eval: false

# Calculate WTP for each attribute
wtp_wait <- -coef_uc["___"] / coef_uc["___"]
wtp_hours <- -coef_uc["___"] / coef_uc["___"]
wtp_provider <- -coef_uc["___"] / coef_uc["___"]

wtp_table <- data.frame(
  Attribute = c("Short Wait (15 min vs 45 min)",
                "Extended Hours (7am-11pm vs 9am-5pm)",
                "MD Provider (vs NP)"),
  WTP = paste0("$", round(c(wtp_wait, wtp_hours, wtp_provider), 2))
)

wtp_table
```

**Interpretation (2 points)**:

1.  **(0.5 points)** Report the sign (positive or negative) of the `wait_15min` coefficient. Does this direction make sense? Why?
2.  **(0.5 points)** Which attribute has the highest importance percentage? Which has the lowest?
3.  **(1.0 point)** Report the WTP for short wait time. If maintaining 15-minute wait times costs the hospital \$60 per visit through additional staffing, is the investment justified? (Compare the WTP to the \$60 cost.)

::: {.callout-note icon="false" title="Your Answer (1-3 sentences each)"}
*\[Write your answers here\]*

1.  Question 1:

2.  Question 2:

3.  Question 3:
:::

------------------------------------------------------------------------

## Task 2: Market Share Prediction (3 points)

Using the model from Task 1, predict market shares for two competing urgent care center designs.

-   **Center A (Premium)**: 15-min wait, Extended hours, MD provider, \$150 per visit
-   **Center B (Value)**: 45-min wait, Standard hours, NP provider, \$75 per visit

**Step 1 (Prediction: 0.5 points)**: Predict ratings for both centers using the conjoint model.

```{r}
#| label: task2-predict
#| eval: false

# Define competing centers
center_A <- data.frame(wait_15min = 1, hours_extended = 1,
                       provider_md = 1, visit_cost = 150)
center_B <- data.frame(wait_15min = 0, hours_extended = 0,
                       provider_md = 0, visit_cost = 75)

# Predict ratings
rating_A <- predict(___, newdata = center_A)
rating_B <- predict(___, newdata = center_B)

cat("Center A (Premium) predicted rating:", round(rating_A, 2), "\n")
cat("Center B (Value) predicted rating:", round(rating_B, 2), "\n")
```

**Step 2 (Market Share: 0.5 points)**: Calculate proportional market shares.

```{r}
#| label: task2-marketshare
#| eval: false

# Calculate proportional market share
total_rating <- rating_A + rating_B
share_A <- ___ / total_rating
share_B <- ___ / total_rating

market_share <- data.frame(
  Center = c("Center A (Premium)", "Center B (Value)"),
  Configuration = c("15-min wait, Extended, MD, $150",
                    "45-min wait, Standard, NP, $75"),
  Predicted_Rating = round(c(rating_A, rating_B), 2),
  Market_Share = paste0(round(c(share_A, share_B) * 100, 1), "%")
)

market_share
```

**Interpretation (2 points)**:

1.  Report the predicted rating and market share for each center. Which center captures the larger share?
2.  Center A charges twice as much (\$150 vs. \$75) but offers shorter waits, extended hours, and an MD. Based on the market share split, would you recommend the healthcare system open Center A, Center B, or both? Why?

::: {.callout-note icon="false" title="Your Answer (1-3 sentences each)"}
*\[Write your answers here\]*

1.  Question 1:

2.  Question 2:
:::

------------------------------------------------------------------------

## Task 3: Choice-Based Conjoint: Outpatient Clinic Selection (3 points)

A hospital network wants to understand how patients choose between outpatient clinics. They conducted a discrete choice experiment where 100 respondents each completed 10 choice tasks, selecting their preferred clinic from pairs of options.

The study examines four attributes:

| Attribute | Level 1 | Level 2 | Description |
|----|----|----|----|
| Appointment Availability | Within 1 week | 3+ weeks | How soon patients can be seen |
| Travel Time | 10 minutes | 30 minutes | Door-to-door travel time |
| Online Portal | Full portal | No portal | Access to results, messaging, scheduling |
| Copay | \$30 | \$60 | Out-of-pocket cost per visit |

The choice data has already been prepared in **long format** for you. Load it and prepare it for `mlogit`.

```{r}
#| label: task3-data

clinic_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/lab7_clinic_choices_long.csv")

head(clinic_data, 5)
```

**Key variables**:

| Variable | Description | Type |
|----|----|----|
| `choice_set` | Unique ID for each choice task | ID |
| `alternative` | Which option (A or B) | Categorical |
| `appt_1week` | Appointment within 1 week (1 = yes, 0 = no) | Binary |
| `travel_10min` | Short travel time (1 = 10 min, 0 = 30 min) | Binary |
| `portal_full` | Full online portal (1 = yes, 0 = no) | Binary |
| `copay` | Copay amount (\$30 or \$60) | Continuous |
| `choice` | Whether this alternative was chosen (1 = yes, 0 = no) — **outcome** | Binary |

**Step 1 (Data Preparation: 0.5 points)**: Create the indexed data format required by `mlogit`.

```{r}
#| label: task3-dfidx
#| eval: false

# Create indexed data format for mlogit
clinic_mlogit <- dfidx(clinic_data,
                       shape = "___",
                       choice = "___",
                       idx = c("___", "___"))
```

**Step 2 (Model Fitting: 0.5 points)**: Fit a multinomial logit model.

```{r}
#| label: task3-model
#| eval: false

# Fit choice-based conjoint model
model_clinic <- mlogit(choice ~ ___ + ___ + ___ + ___ | 0,
                       data = ___)

summary(___)
```

**Interpretation (2 points)**:

1.  Report the sign of the `appt_1week` and `copay` coefficients. Do these directions make intuitive sense? Why?
2.  Among the three binary attributes (`appt_1week`, `travel_10min`, `portal_full`), which coefficient has the largest absolute value? What does this suggest about what matters most to patients when choosing an outpatient clinic? (Note: `copay` is excluded from this comparison because it is continuous and on a different scale, so its raw coefficient is not directly comparable to the binary coefficients.)

::: {.callout-note icon="false" title="Your Answer (1-3 sentences each)"}
*\[Write your answers here\]*

1.  Question 1:

2.  Question 2:
:::

------------------------------------------------------------------------

# Submission Checklist

Before submitting, verify:

-   [ ] All 3 tasks are completed
-   [ ] All blanks (`___`) are filled in with appropriate values
-   [ ] All `#| eval: false` changed to `#| eval: true`
-   [ ] All code chunks run without errors
-   [ ] All interpretation questions are answered (1-3 sentences each)
-   [ ] Visualizations have appropriate titles and axis labels
-   [ ] File renders to HTML successfully

------------------------------------------------------------------------

# Points Summary

| Task      | Points   | Topic                    | Method                         |
|-----------|----------|--------------------------|--------------------------------|
| Task 1    | 4.0      | Urgent care preferences  | Rating-Based Conjoint (lm)     |
| Task 2    | 3.0      | Market share prediction  | WTP & Proportional Allocation  |
| Task 3    | 3.0      | Outpatient clinic choice | Choice-Based Conjoint (mlogit) |
| **Total** | **10.0** |                          |                                |

------------------------------------------------------------------------

**Good luck!**
