---
title: "Conjoint Analysis in Healthcare Management"
subtitle: "MBA642: Business Analytics in Healthcare Management"
Week: "Week 14"
date: 04/13/2026
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: false
    theme: [lecture.scss, cosmo]
    highlight-style: github
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Load required packages
library(tidyverse)
library(broom)
library(mlogit)
library(dfidx)

# Set seed for reproducibility
set.seed(2026)
```

# Introduction to Conjoint Analysis

## What is Conjoint Analysis?

Conjoint analysis is a survey-based statistical technique used to understand how people make decisions when facing **trade-offs** between multiple attributes of a product or service.

The core idea is simple: in real life, we rarely get everything we want. When choosing a health insurance plan, for example, we might prefer both low premiums AND low deductibles, but plans with low premiums often come with high deductibles. Conjoint analysis helps us understand how people navigate these trade-offs.

The term "conjoint" comes from "considered jointly" (we're studying how attributes are evaluated together rather than in isolation). This is crucial because asking someone "Do you prefer low cost?" will always get a "yes," but asking them to choose between a low-cost option with long wait times vs. a high-cost option with short wait times reveals their true priorities.


## Why Use Conjoint Analysis in Healthcare Management?

Healthcare managers face constant decisions about service design:

- **Insurance plan design**: What combination of premium, deductible, and coverage will attract people?
- **Clinic operations**: Should we prioritize shorter wait times or extended hours?
- **Telehealth services**: Do patients value video consultations enough to pay more for them?
- **Hospital services**: What amenities matter most to patients choosing where to receive care?

Conjoint analysis answers these questions by quantifying preferences. Instead of relying on intuition or simple surveys (e.g., "Would you like shorter wait times?"), conjoint analysis reveals how much each feature actually matters when people must make realistic trade-offs.


## Key Outputs from Conjoint Analysis

By using conjoint analysis, we can get several valuable insights:

### 1. Part-Worth Utilities

**Part-worth utilities** (also called "part-worths" or "utility scores") are the building blocks of conjoint analysis. They represent the value or satisfaction that a person derives from a specific level of an attribute.

**What is "utility"?** In this context, "utility" is an economics term meaning satisfaction or value. Think of it as a numeric score representing how much someone values something (Higher utility = more preferred, lower utility = less preferred).

Think of it this way: every product or service is a bundle of attributes. A health insurance plan is a bundle of (premium, deductible, network size, etc.). The total utility (satisfaction) a person gets from a plan is the sum of the part-worth utilities for each attribute level in that plan.

For example, if the part-worth for "Large Network" is +3 and for "Small Network" is 0, then having a large network adds 3 utility units to the overall plan value. These numbers are relative; they represent **relative values**, not absolute values on their own.


### 2. Attribute Importance

Once we have part-worth utilities, we can calculate how important each attribute is relative to others. This is done by looking at the **range** of utilities within each attribute (highest level minus lowest level) and comparing these ranges across attributes.

For example, if premium ranges from 0 to 5 utility units and network size ranges from 0 to 2 units, then premium has a larger impact on preferences and is therefore more "**important**."


### 3. Willingness-to-Pay (WTP)

**When price or cost is one of the attributes**, we can convert utility differences into dollar values. This tells us how much extra a person would be willing to pay for an improvement in another attribute. (Without a price attribute in the study design, WTP cannot be calculated because there is no price coefficient to serve as the conversion rate between utility and dollars)

For example, if people would pay $50 more per month for a large network versus a small network, this helps managers decide whether investing in network expansion is worthwhile.


### 4. Market Share Prediction

By knowing the utilities for each attribute level, we can predict what percentage of people would choose one option over another. This is invaluable for **forecasting demand and competitive positioning**.


## Two Main Approaches

In this class, we will cover the two most common conjoint methods:

### 1. Rating-Based Conjoint

- **What respondents do**: Rate each profile on a scale (e.g., 1-10)
- **Analysis approach**: Linear regression (OLS)
- **Advantages**: Simple to understand and implement; builds on familiar regression
- **Limitations**: Respondents may rate all options similarly, not revealing true trade-offs

### 2. Choice-Based Conjoint (Discrete Choice Experiment)

- **What respondents do**: Choose their preferred option from a set
- **Analysis approach**: Multinomial logistic regression
- **Advantages**: Mimics real-world decisions; forces explicit trade-offs
- **Limitations**: Requires more data preparation and more complicated statistical method

Both methods ultimately produce the same types of outputs (part-worths, importance, WTP, market share), but choice-based conjoint is generally preferred in academic research and practice because it better reflects actual decision-making behavior (Louviere et al., 2000).

Now that we understand what conjoint analysis is and why it's valuable in healthcare management, let's work through a practical example. We'll use one comprehensive case study that demonstrates both rating-based and choice-based approaches, allowing us to see how the methods compare and when you might use each one.

# Scenario: Health Insurance Plan Selection

## Context and Background

One of the most studied applications of conjoint analysis in healthcare management is understanding **how employees choose health insurance plans** during open enrollment periods. This is a high-stakes decision for both employees (who will live with the plan for a year) and employers (who subsidize the plans).

Research has shown that employees often struggle with this decision due to the complexity of plan attributes. Studies by Sinaiko and Hirth (2011) in the *Journal of Health Economics* found that many employees make "dominated choices" (selecting plans that are objectively worse than alternatives on all dimensions). Understanding which attributes drive choices can help employers design better communication strategies and plan offerings.

Our example uses three key attributes that consistently appear in health plan conjoint studies:

| Attribute | Level 1 | Level 2 | Why It Matters |
|-----------|---------|---------|----------------|
| Monthly Premium | $200 | $300 | Direct, visible cost paid every month |
| Annual Deductible | $500 | $1,500 | Amount paid before insurance kicks in |
| Network Size | Large | Small | Access to preferred doctors/hospitals |

With 3 attributes and 2 levels each, there are $2^3 = 8$ possible profile combinations (a **full factorial design**). This is small enough to show every combination to respondents. In practice, when the number of attributes and levels is larger (e.g., 5 attributes × 3 levels = 243 profiles), researchers use a **fractional factorial design** that presents only a carefully selected subset of profiles.


## 1. Rating-Based Conjoint Analysis

### 1.1 Conceptual Overview

In rating-based conjoint analysis, we show respondents hypothetical profiles one at a time and ask them to rate how attractive each option is. For example:

> **Profile 1**: Monthly premium $200, Deductible $1,500, Small network
> *How likely would you be to choose this plan? (Rate 1-10)*

> **Profile 2**: Monthly premium $300, Deductible $500, Large network
> *How likely would you be to choose this plan? (Rate 1-10)*

A respondent might rate Profile 1 as 4/10 and Profile 2 as 7/10.

By collecting ratings for multiple profiles across many respondents, we can use regression analysis to decompose these ratings into the contribution of each attribute. This works because we have systematically varied the attributes across profiles.


### 1.2 The Statistical Model

Rating-based conjoint uses **ordinary least squares (OLS) regression**, the same technique you've learned in previous courses. The model assumes that a person's rating is a linear combination of the utilities they assign to each attribute level:

$$\text{Rating} = \beta_0 + \beta_1(\text{Premium}) + \beta_2(\text{Deductible}) + \beta_3(\text{Network}) + \varepsilon$$

Where:

- $\beta_0$ is the intercept (baseline rating)
- $\beta_1, \beta_2, \beta_3$ are the **part-worth utilities** for each attribute 
  - For continuous attributes like premium, this is the marginal utility per dollar 
  - For categorical attributes like network size, this is the utility difference between levels
- $\varepsilon$ is random error (individual variation, measurement error, etc.)

**The coefficients we estimate ARE the part-worth utilities**. They tell us how much each unit change in an attribute affects the rating.

**Important note about coding variables**

In conjoint analysis, we can code attributes in different ways depending on their nature:

- **Continuous variables** (e.g., premium in dollars, deductible in dollars): Treated as numeric values. The coefficient tells us the utility change per unit (e.g., per dollar).
- **Categorical variables** (e.g., network size): Coded as dummy variables (0 or 1). The coefficient represents the utility difference between categories (e.g., Large network vs. Small network).

In our insurance example, Premium and Deductible are continuous (dollar amounts), while Network Size is categorical (Large vs. Small).

### 1.3 Generating the Experimental Design

First, we create all possible combinations of our attribute levels. With 3 attributes × 2 levels each, we have $2^3 = 8$ possible profiles. This is called a **full factorial design**.

```{r}
#| label: rating-insurance-data
#| echo: false

# Create full factorial design for insurance plans
insurance_profiles <- expand.grid(
  premium = c(200, 300),
  deductible = c(500, 1500),
  network = c("Small", "Large")
)
insurance_profiles$profile_id <- 1:nrow(insurance_profiles)

# Display all 8 profiles
knitr::kable(insurance_profiles,
             col.names = c("Premium (\\$)", "Deductible (\\$)", "Network", "Profile ID"),
             align = c("c", "c", "c", "c"),
             caption = "Health Insurance Plan Profiles (Full Factorial Design)")
```

### 1.4 Loading Survey Data

Let's load survey data collected using these profiles. In practice, you would design and administer the survey yourself, but here we use simulated data so that we can verify our analysis against known true values:

```{r}
#| label: rating-insurance-simulate

rating_insurance <- read.csv("week14_rating_insurance_data.csv")

# Show first few rows
head(rating_insurance, 10)
```

**Understanding the dataset:**

The `rating_insurance` dataset has 800 observations (100 respondents × 8 profiles each):

| Variable | Description |
|----------|-------------|
| `respondent_id` | Unique identifier for each survey respondent (1 to 100) |
| `profile_id` | Identifier for each insurance plan profile (1 to 8) |
| `premium` | Monthly premium cost in dollars (200 or 300) |
| `deductible` | Annual deductible amount in dollars (500 or 1500) |
| `network` | Network size (Large or Small) |
| `rating` | Respondent's rating of the plan on a 1-10 scale (outcome variable) |

### 1.5 Estimating Part-Worth Utilities

Now we use linear regression to estimate the part-worth utilities. We convert Network to a factor with "Small" as the reference level, so that the coefficient represents the utility gain from having a large network (vs. small):

```{r}
#| label: rating-insurance-estimation

# Convert network to a factor (Small = reference level)
rating_insurance <- rating_insurance %>%
  mutate(network = factor(network, levels = c("Small", "Large")))

# Fit linear regression model
model_rating_ins <- lm(rating ~ premium + deductible + network,
                       data = rating_insurance)

# Display results
summary(model_rating_ins)
```


```{r}
#| label: rating-insurance-coef-table
#| echo: false

# Extract and display coefficients
coef_rating_ins <- tidy(model_rating_ins) %>%
  mutate(p.value = ifelse(p.value < 0.001, "< .001", as.character(round(p.value, 4))))

knitr::kable(coef_rating_ins, digits = 4,
             align = "c",
             caption = "Estimated Part-Worth Utilities (Rating-Based Conjoint)")
```

```{r}
#| label: rating-insurance-coef-extract
#| echo: false

# Extract coefficients for inline text
coef_intercept <- round(coef(model_rating_ins)["(Intercept)"], 2)
coef_premium <- round(coef(model_rating_ins)["premium"], 4)
coef_deductible <- format(round(coef(model_rating_ins)["deductible"], 4), nsmall=4, scientific = FALSE)
coef_network <- round(coef(model_rating_ins)["networkLarge"], 2)

# Calculate meaningful effects
effect_premium_100 <- abs(round(coef(model_rating_ins)["premium"] * 100, 2))
effect_deductible_1000 <- format(abs(round(coef(model_rating_ins)["deductible"], 4) * 1000), nsmall=2, scientific = FALSE)
```


**Interpreting the coefficients (Part-Worth Utilities):**

- **Intercept (`r coef_intercept`)**: The predicted rating when premium = $0, deductible = $0, and network = Small (the reference level). Since no real plan has a $0 premium and $0 deductible, this value is an extrapolation far outside our data range and should not be interpreted substantively. It simply serves as the mathematical starting point from which the attribute effects shift the predicted rating.
- **Premium (`r coef_premium`)**: Each $100 increase in monthly premium decreases the rating by about `r effect_premium_100` points.
- **Deductible (`r coef_deductible`)**: Each $1,000 increase in annual deductible decreases rating by about `r effect_deductible_1000` points.
- **networkLarge (`r coef_network`)**: Having a large network (vs. small) increases rating by about `r coef_network` points.

The regression coefficients (excluding the intercept) ARE the part-worth utilities. They directly tell us how much each attribute contributes to the overall rating.

### 1.6 Calculating Attribute Importance

Attribute importance tells us which features have the biggest impact on preferences. The calculation:

1. For each attribute, calculate the **utility range**: difference between highest and lowest part-worth values
2. Sum all the ranges
3. Express each range as a percentage of the total

**Why does a larger range mean more important?**

An attribute's range tells us how much it can change someone's overall evaluation. If *premium* has a utility range of 5 points and *network* has a range of 2 points, switching from worst to best premium gives you 5 utility points, while switching network levels only gives 2 points. The attribute with the larger range has more power to influence preferences.

**CRITICAL CAVEAT**: These importance scores are completely dependent on the range of levels you test. Testing premium of $200 vs. $201 would make premium appear unimportant, while $200 vs. $500 would make it appear very important. Importance scores do NOT tell us which attributes are "objectively most important." They tell us which attributes have the most impact **GIVEN THE SPECIFIC LEVELS YOU TESTED**. Always choose attribute levels that reflect realistic market variation.


Let's get regression coefficients and calculate importance:

```{r}
#| label: rating-insurance-importance
#| code-fold: true

# 1. Extract coefficients
coef_ins <- coef(model_rating_ins)

# 2. Calculate utility at each level and the range
importance_rating_ins <- data.frame(
  attribute = c("Monthly Premium", "Annual Deductible", "Network Size"),
  highest_utility = c(
    coef_ins["premium"] * 200,       # Lowest premium ($200) = highest utility
    coef_ins["deductible"] * 500,     # Lowest deductible ($500) = highest utility
    coef_ins["networkLarge"] * 1      # Large network = highest utility
  ),
  lowest_utility = c(
    coef_ins["premium"] * 300,        # Highest premium ($300) = lowest utility
    coef_ins["deductible"] * 1500,    # Highest deductible ($1500) = lowest utility
    0                                  # Small network (reference) = 0
  )
)

importance_rating_ins <- importance_rating_ins %>%
  mutate(
    utility_range = abs(highest_utility - lowest_utility),
    importance_pct = utility_range / sum(utility_range) * 100
  )

knitr::kable(importance_rating_ins, digits = 2,
             col.names = c("Attribute", "Highest Utility", "Lowest Utility",
                           "Utility Range", "Importance (%)"),
             align = "c",
             caption = "Attribute Importance Scores")
```

**Visualization:**

```{r}
#| label: rating-insurance-importance-plot
#| code-fold: true
#| fig-width: 6
#| fig-height: 4

ggplot(importance_rating_ins, aes(x = reorder(attribute, importance_pct),
                                   y = importance_pct)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(aes(label = paste0(round(importance_pct, 2), "%")),
            hjust = -0.1, size = 4) +
  coord_flip() +
  labs(title = "Attribute Importance: Health Insurance Plan Selection",
       subtitle = "Rating-Based Conjoint Analysis",
       x = "",
       y = "Relative Importance (%)",
       caption = "Importance = utility range as % of total range across all attributes") +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold")) +
  ylim(0, max(importance_rating_ins$importance_pct) * 1.2)
```

```{r}
#| label: rating-insurance-importance-interpretation
#| echo: false

# Determine the order for interpretation
importance_order <- importance_rating_ins %>%
  arrange(desc(importance_pct))

most_important <- importance_order$attribute[1]
second_important <- importance_order$attribute[2]
third_important <- importance_order$attribute[3]
```

**Interpretation**: `r most_important` is the most important attribute (`r round(importance_order$importance_pct[1], 1)`%), followed by `r second_important` (`r round(importance_order$importance_pct[2], 1)`%), and then `r third_important` (`r round(importance_order$importance_pct[3], 1)`%). This suggests that when choosing health plans, employees are most sensitive to differences in `r tolower(most_important)` (within the ranges tested).


### 1.7 Calculating Willingness-to-Pay (WTP)

Willingness-to-pay converts abstract utility values into concrete dollar amounts. This is possible because we included **a price attribute** (monthly premium) in our study.

**The Logic**:

If we know (1) how much utility a person gets from each attribute and (2) how much utility they lose per dollar of cost, we can calculate the dollar amount that would exactly offset any utility gain.

For example, if the utility of a large network is +0.6 points, and the utility loss per dollar of premium is -0.012 points, then: 0.6 / 0.012 = $50. This means respondents value a large network as much as they would value a $50 reduction in monthly premium.

**The Formula**:
$$\text{WTP} = -\frac{\beta_{\text{attribute}}}{\beta_{\text{price}}}$$

```{r}
#| label: rating-insurance-wtp
#| echo: false

# Extract coefficients
beta_network <- coef_ins["networkLarge"]
beta_premium <- coef_ins["premium"]
beta_deductible <- coef_ins["deductible"]

# Calculate WTP for large network (vs small)
wtp_network <- -beta_network / beta_premium

# Calculate WTP for $1000 lower deductible
wtp_deductible_reduction <- -(beta_deductible * -1000) / beta_premium
```

**WTP Calculations:**

From our regression model:

- Premium coefficient: `r round(beta_premium, 4)` utility points per dollar
- Network coefficient: `r round(beta_network, 4)` utility points for large network
- Deductible coefficient: `r format(round(beta_deductible, 4), nsmall = 4, scientific = FALSE)` utility points per dollar

**WTP for Large Network (vs. Small Network)**:

$$\text{WTP}_{\text{network}} = -\frac{`r round(beta_network, 4)`}{`r round(beta_premium, 4)`} = \$`r round(wtp_network, 2)`$$

**Interpretation**: Respondents would be willing to pay an additional **$`r round(wtp_network, 2)` per month** to have access to a large provider network instead of a small network.

**WTP for $1,000 Lower Deductible**:

Applying the same logic, but scaling the deductible coefficient to a $1,000 change:

$$\text{WTP}_{\text{deductible}} = -\frac{\beta_{\text{deductible}} \times (-1000)}{\beta_{\text{premium}}} = -\frac{`r format(round(beta_deductible * -1000, 4), nsmall = 4, scientific = FALSE)`}{`r round(beta_premium, 4)`} = \$`r round(wtp_deductible_reduction, 2)`$$

**Interpretation**: Respondents would be willing to pay an additional **$`r round(wtp_deductible_reduction, 2)` per month** to reduce their annual deductible by $1,000.

**Summary of WTP Estimates:**

| Attribute | WTP per Month |
|:---------:|:-------------:|
| Large Network (vs. Small) | $`r round(wtp_network, 2)` |
| $1,000 Lower Deductible | $`r round(wtp_deductible_reduction, 2)` |

**Managerial Application**:

These WTP values help managers make cost-benefit decisions:

- **Network expansion decision**: If expanding the provider network costs the insurer $30 per member per month, and members are willing to pay $`r round(wtp_network, 2)` for it, the expansion is financially viable: the insurer can charge an additional $`r round(wtp_network, 2)` in premium while only incurring $30 in costs, creating a $`r round(wtp_network - 30, 2)` per member margin.

- **Premium pricing for deductible tiers**: Members are willing to pay $`r round(wtp_deductible_reduction, 2)` per month (or $`r round(wtp_deductible_reduction * 12, 0)` annually) to reduce their deductible by $1,000. Offering a $1,000 lower deductible means insurance kicks in sooner, costing the insurer approximately $400 per year in additional claims. Since members would pay ~$`r round(wtp_deductible_reduction * 12, 0)`/year in extra premium for this benefit, the insurer captures a margin of approximately $`r round(wtp_deductible_reduction * 12 - 400, 0)` per member per year.

### 1.8 Market Share Simulation

One of the most powerful applications of conjoint analysis is predicting how the market would respond to different product configurations.

**The Process for Rating-Based Conjoint**:

1. Define two (or more) competing options
2. Calculate the predicted rating for each option using the regression model
3. Convert ratings to market shares using proportional allocation

**Proportional Allocation**: If Plan A receives a predicted rating of 8 and Plan B receives 6, we allocate market share proportionally:

$$\text{Share}_A = \frac{8}{8+6} = 57\% \qquad \text{Share}_B = \frac{6}{8+6} = 43\%$$

**Example Scenario:**

Let's compare two common plan designs:

- **Plan A (Premium Plan)**: $250/month premium, $500 annual deductible, Large network
- **Plan B (Budget Plan)**: $200/month premium, $1,500 annual deductible, Small network

Neither plan dominates the other: Plan A is better on deductible and network, while Plan B is better on premium.

```{r}
#| label: rating-insurance-marketshare
#| echo: false

# Define two competing health plans
plan_A <- data.frame(
  name = "Premium Plan",
  premium = 250,
  deductible = 500,
  network = "Large",
  description = "Mid-price, Low deductible, Large network"
)

plan_B <- data.frame(
  name = "Budget Plan",
  premium = 200,
  deductible = 1500,
  network = "Small",
  description = "Low-price, High deductible, Small network"
)

# Predict ratings for each plan
rating_A <- predict(model_rating_ins, plan_A)
rating_B <- predict(model_rating_ins, plan_B)

# Calculate market share using proportional allocation
total_rating <- rating_A + rating_B
share_A <- rating_A / total_rating
share_B <- rating_B / total_rating
```

**Step 1: Calculate predicted ratings**

Using the regression equation $\text{Rating} = \beta_0 + \beta_1(\text{Premium}) + \beta_2(\text{Deductible}) + \beta_3(\text{Network})$:

**Plan A** (Premium Plan: $250, $500 deductible, Large network):

$$\text{Rating}_A = `r coef_intercept` + (`r coef_premium` \times 250) + (`r coef_deductible` \times 500) + (`r coef_network` \times 1) = `r round(rating_A, 2)`$$

**Plan B** (Budget Plan: $200, $1,500 deductible, Small network):

$$\text{Rating}_B = `r coef_intercept` + (`r coef_premium` \times 200) + (`r coef_deductible` \times 1500) + (`r coef_network` \times 0) = `r round(rating_B, 2)`$$

**Step 2: Convert to market shares using proportional allocation**

$$\text{Share}_A = \frac{`r round(rating_A, 2)`}{`r round(rating_A, 2)` + `r round(rating_B, 2)`} = `r round(share_A * 100, 1)`\% \qquad \text{Share}_B = \frac{`r round(rating_B, 2)`}{`r round(rating_A, 2)` + `r round(rating_B, 2)`} = `r round(share_B * 100, 1)`\%$$


```{r}
#| label: rating-insurance-marketshare-display
#| echo: false

# Display results
market_share_rating <- data.frame(
  Plan = c(plan_A$name, plan_B$name),
  Description = c(plan_A$description, plan_B$description),
  Predicted_Rating = round(c(rating_A, rating_B), 2),
  Market_Share = paste0(round(c(share_A, share_B) * 100, 1), "%")
)

knitr::kable(market_share_rating,
             align = "c",
             col.names = c("Plan", "Configuration", "Predicted Rating", "Predicted Share"),
             caption = "Market Share Simulation: Two Competing Plans (Rating-Based)")
```


**Interpretation**: Based on the predicted ratings, approximately `r round(share_A * 100, 0)`% of respondents would prefer the Premium Plan despite its higher cost, because the low deductible and large network compensate for the price difference. This helps managers:

- Forecast enrollment in different plan designs
- Understand competitive positioning
- Optimize plan configurations

**Note on Method**: Proportional allocation is a simple and intuitive approach, but it lacks theoretical grounding. It assumes that if Plan A gets a rating of 8 and Plan B gets a rating of 6, then exactly 57% of people will choose Plan A (8 ÷ 14 = 0.57). However, in reality, people make discrete yes/no choices, not proportional splits—someone either picks Plan A or they don't. The method assumes market shares directly mirror rating differences, which oversimplifies how people actually choose. For important strategic decisions, choice-based methods with multinomial logit formulas (shown later) are theoretically more sound and preferred in practice because they model the actual choice process.

## 2. Practice: Rating-Based Conjoint Analysis {.unnumbered}

::: {.callout-note}
## Your Turn!

Practice estimating and interpreting a rating-based conjoint model using a different healthcare scenario.

:::

### Scenario

The COVID-19 pandemic accelerated the adoption of telehealth services, and healthcare organizations now face decisions about how to structure "hybrid" care models. Understanding patient preferences for telehealth attributes is crucial for designing services that people will actually use.

Research by Savira et al. (2023) in *PLOS ONE* examined consumer preferences for telehealth in Australia, finding that attributes like wait time, consultation mode, and cost significantly influenced preferences.

Your healthcare organization is designing a new telehealth program and needs to understand which attributes matter most to patients. You will analyze three key attributes:

| Attribute | Level 1 | Level 2 | Why It Matters |
|-----------|---------|---------|----------------|
| Wait Time | Same-day | 3 days | How quickly patients can be seen |
| Mode | Video | Phone | Communication richness |
| Co-pay | $20 | $40 | Out-of-pocket cost per visit |

With 3 attributes × 2 levels each, there are $2^3 = 8$ possible profiles:

```{r}
#| label: activity1-profile
#| echo: false

# Create full factorial design
telehealth_profiles <- expand.grid(
  wait_sameday = c(1, 0),  # 1 = Same-day, 0 = 3 days
  mode_video = c(1, 0),    # 1 = Video, 0 = Phone
  copay = c(20, 40)
)
telehealth_profiles$profile_id <- 1:nrow(telehealth_profiles)

# Create display version
telehealth_display <- telehealth_profiles %>%
  mutate(
    wait_time = ifelse(wait_sameday == 1, "Same-day", "3 days"),
    mode = ifelse(mode_video == 1, "Video", "Phone")
  ) %>%
  select(wait_time, mode, copay, profile_id)

knitr::kable(telehealth_display,
             col.names = c("Wait Time", "Mode", "Co-pay ($)", "Profile ID"),
             align = "c",
             caption = "Telehealth Program Profiles")
```

### Data Loading

Run this code to load the dataset:

```{r}
#| label: activity1-data

rating_telehealth <- read.csv("week14_practice_data1.csv")
head(rating_telehealth, 10)
```

**Dataset Variables:**

The `rating_telehealth` dataset contains 800 observations (100 respondents × 8 profiles each).

| Variable | Description | Type |
|----------|-------------|------|
| `respondent_id` | Unique identifier (1-100) | Numeric |
| `profile_id` | Profile identifier (1-8) | Numeric |
| `wait_sameday` | Binary indicator (1 = Same-day, 0 = 3 days) | Numeric (binary) |
| `mode_video` | Binary indicator (1 = Video, 0 = Phone) | Numeric (binary) |
| `copay` | Co-payment amount (20 or 40) | Numeric (continuous) |
| `rating` | Satisfaction rating on 1-10 scale (outcome variable) | Numeric (continuous) |

### Task 1.1: Estimate Part-Worth Utilities

Fit a linear regression model to estimate the part-worth utilities (coefficients) for each attribute.

```{r}
#| label: activity1-task1
#| eval: false

# Fill in the blanks
model_rating_th <- lm(rating ~ ___ + ___ + ___,
                      data = rating_telehealth)

summary(model_rating_th)
```

**Hint:** The formula should include all three attribute variables from the dataset.

**Questions:**

1. What is the sign and magnitude of each coefficient? Do they match your expectations?
2. Is each coefficient statistically significant (p < 0.05)?


### Task 1.2: Calculate Attribute Importance

Calculate which attributes have the most impact on preferences.

```{r}
#| label: activity1-task2
#| eval: false

# Extract coefficients
coef_th <- coef(model_rating_th)

# Calculate utility ranges for each attribute (fill in the blanks)
range_wait <- abs(coef_th["___"] * 1)  # Wait Time: Binary 0 vs 1
range_mode <- abs(coef_th["___"] * 1)  # Mode: Binary 0 vs 1
range_copay <- abs(coef_th["___"] * (40 - 20))  # Copay: $20 difference

# Calculate total range
total_range_th <- range_wait + range_mode + range_copay

# Calculate importance percentages
importance_rating_th <- data.frame(
  attribute = c("Wait Time", "Consultation Mode", "Co-pay"),
  importance_pct = c(range_wait, range_mode, range_copay) / total_range_th * 100
)

# Display results
importance_rating_th
```

**Hint:** Fill in the coefficient names from your regression model. These should match the variable names you used in Task 1.1.

**Question:** Which attribute is most important? Does this match your intuition about telehealth preferences?


### Task 1.3: Calculate Willingness-to-Pay (WTP)

Calculate how much patients would pay for same-day access and video consultation.

```{r}
#| label: activity1-task3
#| eval: false

# Calculate WTP for same-day access (vs 3-day wait) - fill in the blanks
wtp_sameday <- -coef_th["___"] / coef_th["___"]

# Calculate WTP for video consultation (vs phone)
wtp_video <- -coef_th["___"] / coef_th["___"]

# Display results
cat("WTP for Same-day access: $", round(wtp_sameday, 2), " per visit\n")
cat("WTP for Video consultation: $", round(wtp_video, 2), " per visit\n")
```

**Hint:** WTP = -β_attribute / β_price. The price attribute here is `copay`.

**Question:** If adding same-day scheduling costs the organization $30 per visit, is it financially viable based on WTP?


### Task 1.4: Predict Market Share

Compare two program designs and predict which would be more popular:

- **Program A (Premium)**: Same-day access, Video consultation, $40 co-pay
- **Program B (Basic)**: 3-day wait, Phone consultation, $20 co-pay

```{r}
#| label: activity1-task4
#| eval: false

# Define two competing programs
program_A <- data.frame(wait_sameday = 1, mode_video = 1, copay = 40)  # Premium: Same-day, Video, $40
program_B <- data.frame(wait_sameday = 0, mode_video = 0, copay = 20)  # Basic: 3-day, Phone, $20

# Predict ratings (fill in the blanks)
rating_A_th <- predict(model_rating_th, ____)
rating_B_th <- predict(model_rating_th, ____)

# Calculate proportional market share
total_rating_th <- rating_A_th + rating_B_th
share_A_th <- rating_A_th / total_rating_th
share_B_th <- rating_B_th / total_rating_th

# Display results
cat("Premium Program (Same-day Video, $40): ", round(share_A_th * 100, 1), "%\n")
cat("Basic Program (3-day Phone, $20): ", round(share_B_th * 100, 1), "%\n")
```

**Hint:** Use the `predict()` function with your model and the program data frames.

**Question:** Which program captures the larger market share? Given that Program A costs $20 more per visit, does the higher share justify offering it as the primary program?

------------------------------------------------------------------------

## 3. Choice-Based Conjoint Analysis

### 3.1 Conceptual Overview

Choice-based conjoint (CBC), also known as a **Discrete Choice Experiment (DCE)**, is considered the gold standard in conjoint analysis because it closely mimics real-world decision-making (Louviere, Hensher, & Swait, 2000).

Instead of rating profiles individually, respondents are shown **choice sets** containing two or more options and asked to pick their preferred one:

> **Choice Set 1:**
>
> | | Plan A | Plan B |
> |---|--------|--------|
> | Premium | $200 | $300 |
> | Deductible | $1,500 | $500 |
> | Network | Small | Large |
>
> **Which plan would you choose?** ☐ Plan A  ☐ Plan B

By forcing a choice, we ensure that respondents make explicit trade-offs. Someone who wants both low premium AND low deductible must decide which matters more.

### 3.2 Multinomial Logit Model

Choice-based conjoint uses **multinomial logistic regression** to analyze choice data. This method can handle choice sets with any number of alternatives (2, 3, 4, or more). We'll use the `mlogit` package in R, which is the standard tool for discrete choice analysis.

**Key Concept**: 

The model assumes people choose the alternative with the highest total utility. Each alternative's total utility has two components:

$$U_i = V_i + \varepsilon_i$$

- $V_i = \beta_1 X_{i1} + \beta_2 X_{i2} + ... + \beta_k X_{ik}$ is the **systematic** (predictable) component, based on observed attributes
- $\varepsilon_i$ is the **random** component, representing individual taste variation and unmeasured factors

The logit formula converts the systematic utilities into choice probabilities (the random component is integrated out through the derivation):

$$P(\text{Choose alternative } i) = \frac{e^{V_i}}{\sum_{j=1}^{J} e^{V_j}}$$

where:

- $J$ is the total number of alternatives in the choice set
- The coefficients ($\beta_1, \beta_2, ..., \beta_k$) are the **part-worth utilities**

**Intuition**: Alternatives with higher utility have exponentially higher choice probabilities. The model estimates the part-worth utilities ($\beta$ coefficients) that best explain the observed choices.

### 3.3 Choice Data Format

The `mlogit` package requires data in **long format** — each row represents one alternative within a choice set, not one choice set per row. Note that the `choice_set` identifier must be **unique across all respondents**. That's why we use IDs like `1_1`, `1_2`, ... (meaning "respondent 1, choice set 1", "respondent 1, choice set 2", etc.) rather than simple numbers like 1, 2, 3. If two respondents both had a choice set labeled "1", `mlogit` would incorrectly treat them as alternatives within the same choice set.

In other words, we need to clarify **what each respondent saw and chose in each choice set**. 

The dataset must have this structure:

| Column | Description |
|--------|-------------|
| `choice_set` | Identifier linking alternatives in the same set (e.g., "1_1" = respondent 1, choice set 1) |
| `alternative` | Label for each option (e.g., A, B) |
| attribute columns | One column per attribute (e.g., `premium`, `deductible`, `network_large`) |
| `choice` | Binary: 1 = chosen, 0 = not chosen (sums to 1 within each choice set) |

For example, one choice set with two alternatives looks like this:

| choice_set | alternative | premium | deductible | network_large | choice |
|:----------:|:-----------:|:-------:|:----------:|:-------------:|:------:|
| 1_1        | A           | 200     | 1500       | 0             | 1      |
| 1_1        | B           | 300     | 500        | 1             | 0      |

```{r}
#| label: choice-insurance-data

# Load choice data (already in long format)
choice_long <- read.csv("week14_choice_insurance_long.csv")

# Display the data structure
head(choice_long, 8)
```

**Understanding the dataset**: Each pair of rows represents one choice set. The `choice` variable indicates which alternative was selected (1 = chosen, 0 = not chosen). Within each choice set, the `choice` values sum to 1.

### 3.4 Indexing and Model Estimation

The `dfidx()` function prepares the data for mlogit by creating an indexed structure:

```{r}
#| label: choice-insurance-dfidx

# Create indexed data format for mlogit
choice_mlogit <- dfidx(choice_long,
                       shape = "long",           # data is already in long format (one row per alternative)
                       choice = "choice",         # column indicating which alternative was chosen (1/0)
                       idx = c("choice_set",      # column that groups alternatives into the same choice set
                               "alternative"))    # column that identifies each alternative (A, B, etc.)
```

**What dfidx() does**: It adds special index attributes telling mlogit which rows belong to the same choice set. This is essential for correctly estimating choice probabilities.

Now we can estimate the model using the `mlogit` function in R:

```{r}
#| label: choice-insurance-mlogit-model

# Fit multinomial logit model
model_mlogit_ins <- mlogit(choice ~ premium + deductible + network_large | 0,
                           data = choice_mlogit)

summary(model_mlogit_ins)
```

**Understanding the formula**:

- `choice ~ premium + deductible + network_large` specifies the attributes
- `| 0` means "no alternative-specific constants"

We exclude alternative-specific constants because "Plan A" and "Plan B" are just generic labels (i.e., there's no inherent preference for "A" vs. "B" independent of their attributes).

```{r}
#| label: choice-insurance-coef-table
#| echo: false

coef_choice_ins <- tidy(model_mlogit_ins) %>%
  mutate(p.value = ifelse(p.value < 0.001, "< .001", as.character(round(p.value, 4))))

knitr::kable(coef_choice_ins, digits = 4,
             align = "c",
             caption = "Part-Worth Utilities (Choice-Based Conjoint)")
```

**Interpretation**: The estimated coefficients are the part-worth utilities, representing how much each attribute contributes to the **systematic utility** ($V_i$) of an alternative.

- **Premium** (`r round(coef(model_mlogit_ins)["premium"], 4)`): Each additional $1 increase in monthly premium **decreases the utility** of that plan option by `r abs(round(coef(model_mlogit_ins)["premium"], 4))` points
- **Deductible** (`r format(round(coef(model_mlogit_ins)["deductible"], 4), nsmall = 4, scientific = FALSE)`): Each additional $1 increase in annual deductible **decreases the utility** of that plan option by `r format(abs(round(coef(model_mlogit_ins)["deductible"], 4)), nsmall = 4, scientific = FALSE)` points
- **Network_large** (`r round(coef(model_mlogit_ins)["network_large"], 4)`): Having a large network (vs. small) **increases the utility** of that plan option by `r round(coef(model_mlogit_ins)["network_large"], 4)` points

**How do utilities translate to choice probabilities?**

The multinomial logit formula converts utilities into choice probabilities:

$$P(\text{Choose alternative } i) = \frac{e^{V_i}}{\sum_{j} e^{V_j}}$$

where $V_i = \beta_{\text{premium}} \times \text{premium}_i + \beta_{\text{deductible}} \times \text{deductible}_i + \beta_{\text{network}} \times \text{network}_i$ = systematic utility of alternative $i$

**Key insight**: Higher utility → Higher probability of being chosen, but the relationship is **nonlinear** (exponential). For example, with two alternatives having utilities of 2.0 and 1.0, the first would be chosen about 73% of the time (not 67% as a simple proportional model would suggest), because the logit formula exponentiates the utilities before computing shares.

### 3.5 Attribute Importance (Choice-Based)

The calculation is identical to rating-based conjoint:

1. First, get the range of utilities for each attribute (highest - lowest)
2. Then, calculate total range (sum all the ranges)
3. Finally, get each range as a percentage of the total 

```{r}
#| label: choice-insurance-importance
#| code-fold: true

# 1. Extract coefficients
coef_choice <- coef(model_mlogit_ins)

# 2. Calculate utility at each level and the range
importance_choice_ins <- data.frame(
  attribute = c("Monthly Premium", "Annual Deductible", "Network Size"),
  highest_utility = c(
    coef_choice["premium"] * 200,        # Lowest premium ($200) = highest utility
    coef_choice["deductible"] * 500,      # Lowest deductible ($500) = highest utility
    coef_choice["network_large"] * 1      # Large network = highest utility
  ),
  lowest_utility = c(
    coef_choice["premium"] * 300,         # Highest premium ($300) = lowest utility
    coef_choice["deductible"] * 1500,     # Highest deductible ($1500) = lowest utility
    0                                      # Small network (reference) = 0
  )
)

importance_choice_ins <- importance_choice_ins %>%
  mutate(
    utility_range = abs(highest_utility - lowest_utility),
    importance_pct = utility_range / sum(utility_range) * 100
  )

knitr::kable(importance_choice_ins, digits = 2,
             col.names = c("Attribute", "Highest Utility", "Lowest Utility",
                           "Utility Range", "Importance (%)"),
             align = "c",
             caption = "Attribute Importance (Choice-Based Conjoint)")
```

```{r}
#| label: choice-insurance-importance-plot
#| fig-width: 6
#| fig-height: 4
#| echo: false

ggplot(importance_choice_ins, aes(x = reorder(attribute, importance_pct),
                                   y = importance_pct)) +
  geom_col(fill = "darkorange", width = 0.6) +
  geom_text(aes(label = paste0(round(importance_pct, 1), "%")),
            hjust = -0.1, size = 4) +
  coord_flip() +
  labs(title = "Attribute Importance: Health Insurance Plan Selection",
       subtitle = "Choice-Based Conjoint Analysis",
       x = "",
       y = "Relative Importance (%)",
       caption = "Importance = utility range as % of total range") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold")) +
  ylim(0, max(importance_choice_ins$importance_pct) * 1.2)
```

### 3.6 Willingness-to-Pay (Choice-Based)

We apply the same WTP formula as in the rating-based analysis. Because both methods estimate part-worth utilities and include a price attribute, the conversion logic is identical:

$$\text{WTP} = -\frac{\beta_{\text{attribute}}}{\beta_{\text{price}}}$$

```{r}
#| label: choice-insurance-wtp
#| echo: false

# Extract coefficients
beta_network_c <- coef_choice["network_large"]
beta_premium_c <- coef_choice["premium"]
beta_deductible_c <- coef_choice["deductible"]

# Calculate WTP
wtp_network_c <- -beta_network_c / beta_premium_c
wtp_deductible_c <- -(beta_deductible_c * -1000) / beta_premium_c
```

**WTP Calculations:**

**WTP for Large Network (vs. Small Network)**:

$$\text{WTP}_{\text{network}} = -\frac{`r round(beta_network_c, 4)`}{`r round(beta_premium_c, 4)`} = \$`r round(wtp_network_c, 2)`$$

**WTP for $1,000 Lower Deductible** (scaling the per-dollar coefficient to a $1,000 change):

$$\text{WTP}_{\text{deductible}} = -\frac{\beta_{\text{deductible}} \times (-1000)}{\beta_{\text{premium}}} = -\frac{`r format(round(beta_deductible_c * -1000, 4), nsmall = 4, scientific = FALSE)`}{`r round(beta_premium_c, 4)`} = \$`r round(wtp_deductible_c, 2)`$$

**Summary of WTP Estimates:**

| Attribute | WTP per Month |
|:---------:|:-------------:|
| Large Network (vs. Small) | $`r round(wtp_network_c, 2)` |
| $1,000 Lower Deductible | $`r round(wtp_deductible_c, 2)` |

**Interpretation**: Respondents are willing to pay **$`r round(wtp_network_c, 2)` more per month** for a large network and **$`r round(wtp_deductible_c, 2)` more per month** for a plan with a $1,000 lower deductible.

**Managerial Application**: These WTP values from the choice model can be directly compared to the rating-based estimates. Consistent WTP findings across methods strengthen managerial confidence in the results. When the two methods produce similar WTP estimates, managers can trust that the preference data is robust — the underlying preferences are stable regardless of how respondents express them.

### 3.7 Market Share Simulation (Choice-Based)

For choice-based conjoint, we use the **multinomial logit formula** to predict market shares:

$$P(\text{Choose alternative } j) = \frac{e^{U_j}}{\sum_{k=1}^{J} e^{U_k}}$$

Let's compare the same two plans:

```{r}
#| label: choice-insurance-plan-comparison
#| echo: false

plan_comparison <- data.frame(
  Plan = c("Plan A (Premium Plan)", "Plan B (Budget Plan)"),
  Premium = c("$250/month", "$200/month"),
  Deductible = c("$500", "$1,500"),
  Network = c("Large", "Small")
)

knitr::kable(plan_comparison,
             col.names = c("Plan", "Monthly Premium", "Annual Deductible", "Network Size"),
             align = "c",
             caption = "Two Competing Health Insurance Plans")
```

```{r}
#| label: choice-insurance-marketshare
#| echo: false

# Calculate utilities for each plan
coef_mlogit <- coef(model_mlogit_ins)

utility_A_c <- coef_mlogit["premium"] * 250 +
               coef_mlogit["deductible"] * 500 +
               coef_mlogit["network_large"] * 1

utility_B_c <- coef_mlogit["premium"] * 200 +
               coef_mlogit["deductible"] * 1500 +
               coef_mlogit["network_large"] * 0

# Apply multinomial logit formula
share_A_c <- exp(utility_A_c) / (exp(utility_A_c) + exp(utility_B_c))
share_B_c <- exp(utility_B_c) / (exp(utility_A_c) + exp(utility_B_c))
```

**Step 1: Calculate systematic utilities**

Using $V = \beta_{\text{premium}} \times \text{premium} + \beta_{\text{deductible}} \times \text{deductible} + \beta_{\text{network}} \times \text{network}$:

**Plan A** (Premium Plan: $250, $500 deductible, Large network):

$$V_A = (`r round(coef_mlogit["premium"], 4)` \times 250) + (`r format(round(coef_mlogit["deductible"], 4), nsmall = 4, scientific = FALSE)` \times 500) + (`r round(coef_mlogit["network_large"], 4)` \times 1) = `r round(utility_A_c, 2)`$$

**Plan B** (Budget Plan: $200, $1,500 deductible, Small network):

$$V_B = (`r round(coef_mlogit["premium"], 4)` \times 200) + (`r format(round(coef_mlogit["deductible"], 4), nsmall = 4, scientific = FALSE)` \times 1500) + (`r round(coef_mlogit["network_large"], 4)` \times 0) = `r round(utility_B_c, 2)`$$

**Step 2: Convert to market shares using the multinomial logit formula**

$$\text{Share}_A = \frac{e^{`r round(utility_A_c, 2)`}}{e^{`r round(utility_A_c, 2)`} + e^{`r round(utility_B_c, 2)`}} = `r round(share_A_c * 100, 1)`\% \qquad \text{Share}_B = \frac{e^{`r round(utility_B_c, 2)`}}{e^{`r round(utility_A_c, 2)`} + e^{`r round(utility_B_c, 2)`}} = `r round(share_B_c * 100, 1)`\%$$

**Key Difference from Rating-Based**: Instead of dividing ratings proportionally, the logit formula exponentiates the utilities before computing shares. This is theoretically grounded in how people make discrete choices, making it more appropriate for predicting actual market behavior.

**Interpretation**: Based on the choice model, the Premium Plan is predicted to capture approximately `r round(share_A_c * 100, 0)`% of the market. This is consistent with the rating-based results, reinforcing that the low deductible and large network more than compensate for the higher premium. From a managerial perspective:

- The Premium Plan captures approximately `r round(share_A_c * 100, 0)`% of the market despite costing $50/month more, indicating that the low deductible and large network are valued enough to justify the higher price
- The Budget Plan still captures approximately `r round(share_B_c * 100, 0)`%, suggesting a price-sensitive segment exists that prioritizes lower premiums

## 4. Practice: Choice-Based Conjoint Analysis {.unnumbered}

::: {.callout-note}
## Your Turn!

Practice estimating and interpreting a choice-based conjoint model using a different healthcare scenario.

:::

### Scenario

As healthcare systems become more competitive, understanding what drives patients' choice of primary care clinic is crucial. Patients must trade off multiple factors: convenience (location, wait times), provider credentials, and cost.

You'll analyze patient preferences for primary care clinics using **choice-based conjoint analysis** with four key attributes:

| Attribute | Level 1 | Level 2 | Why It Matters |
|-----------|---------|---------|----------------|
| Distance from Home | 5 minutes | 25 minutes | Travel convenience |
| Appointment Wait | Same-week | 3 weeks | Access to care |
| Provider Type | MD (Physician) | NP (Nurse Practitioner) | Professional credentials |
| Copay | $25 | $50 | Out-of-pocket cost |

### Data Loading

The data is provided in long format (one row per alternative per choice set). Run this code to load and index it:

```{r}
#| label: activity2-data

# Load clinic choice data (already in long format)
clinic_long <- read.csv("week14_practice_data2_long.csv")
head(clinic_long, 8)
```

**Dataset Variables:**

The `clinic_long` dataset contains 1,600 rows (100 respondents × 8 choice sets × 2 alternatives per set).

| Variable | Description | Type |
|----------|-------------|------|
| `choice_set` | Unique choice set identifier (e.g., "1_1") | Character |
| `alternative` | Alternative label (A or B) | Character |
| `distance_5min` | Binary (1 = 5 min away, 0 = 25 min away) | Numeric (binary) |
| `wait_sameweek` | Binary (1 = Same-week, 0 = 3 weeks) | Numeric (binary) |
| `provider_md` | Binary (1 = MD, 0 = NP) | Numeric (binary) |
| `copay` | Copay amount (25 or 50) | Numeric (continuous) |
| `choice` | Binary (1 = chosen, 0 = not chosen; sums to 1 per choice set) | Numeric (binary) |

```{r}
#| label: activity2-index

# Index for mlogit
clinic_mlogit <- dfidx(clinic_long,
                       shape = "long",
                       choice = "choice",
                       idx = c("choice_set", "alternative"))
```

### Task 2.1: Estimate Multinomial Logit Model

Fit the choice model to estimate part-worth utilities.

```{r}
#| label: activity2-task1
#| eval: false

# Fill in the blanks
model_clinic <- mlogit(choice ~ ___ + ___ + ___ + ___ | 0,
                       data = clinic_mlogit)

summary(model_clinic)
```

**Hint:** Include all four attributes from your dataset. The `| 0` means no alternative-specific constants.

**Questions:**

1. What do the signs of the coefficients tell you about patient preferences?
2. Which coefficients are statistically significant?


### Task 2.2: Calculate Attribute Importance

Determine which attributes matter most in clinic selection.

```{r}
#| label: activity2-task2
#| eval: false

# Extract coefficients
coef_clinic <- coef(model_clinic)

# Calculate utility ranges (fill in the blanks)
range_distance <- abs(coef_clinic["___"] * 1)   # Distance: Binary 5 min vs 25 min
range_wait <- abs(coef_clinic["___"] * 1)       # Wait Time: Binary Same-week vs 3 weeks
range_provider <- abs(coef_clinic["___"] * 1)   # Provider Type: Binary MD vs NP
range_copay <- abs(coef_clinic["___"] * (50 - 25))  # Copay: $25 difference

# Calculate total range
total_range <- range_distance + range_wait + range_provider + range_copay

# Calculate importance percentages
importance_clinic <- data.frame(
  attribute = c("Distance", "Appointment Wait", "Provider Type", "Copay"),
  importance_pct = c(range_distance, range_wait, range_provider, range_copay) / total_range * 100
)

# Display results
importance_clinic
```

**Hint:** Fill in the coefficient names matching the model variables.


### Task 2.3: Calculate Willingness-to-Pay

Calculate how much patients would pay for each clinic feature.

```{r}
#| label: activity2-task3
#| eval: false

# Calculate WTP for each attribute (fill in the blanks)
wtp_nearby <- -coef_clinic["___"] / coef_clinic["___"]      # 5 min vs 25 min distance
wtp_sameweek <- -coef_clinic["___"] / coef_clinic["___"]    # Same-week vs 3-week wait
wtp_md <- -coef_clinic["___"] / coef_clinic["___"]          # MD vs NP

# Display results
cat("WTP for Nearby Clinic (5 min vs 25 min): $", round(wtp_nearby, 2), " per visit\n")
cat("WTP for Same-week Appointment: $", round(wtp_sameweek, 2), " per visit\n")
cat("WTP for MD vs NP: $", round(wtp_md, 2), " per visit\n")
```

**Hint:** WTP = -β_attribute / β_copay. The copay variable is the "price" in this analysis.

**Question:** Which feature has the highest WTP? What does this suggest for clinic management?


### Task 2.4: Predict Market Share

Compare two competing clinics using the multinomial logit formula:

- **Clinic A (Convenient)**: 5-minute drive, Same-week appointment, NP provider, $50 copay
- **Clinic B (Credentialed)**: 25-minute drive, 3-week wait, MD provider, $25 copay

```{r}
#| label: activity2-task4
#| eval: false

# Calculate systematic utilities
utility_A_clinic <- coef_clinic["distance_5min"] * 1 +
                    coef_clinic["wait_sameweek"] * 1 +
                    coef_clinic["provider_md"] * 0 +
                    coef_clinic["copay"] * 50

utility_B_clinic <- coef_clinic["distance_5min"] * 0 +
                    coef_clinic["wait_sameweek"] * 0 +
                    coef_clinic["provider_md"] * 1 +
                    coef_clinic["copay"] * 25

# Apply multinomial logit formula (fill in the blanks)
share_A_clinic <- _____ / (exp(utility_A_clinic) + exp(utility_B_clinic))
share_B_clinic <- _____ / (exp(utility_A_clinic) + exp(utility_B_clinic))

# Display results
cat("Clinic A (Convenient, NP, Higher copay): ", round(share_A_clinic * 100, 1), "%\n")
cat("Clinic B (Far, MD, Lower copay): ", round(share_B_clinic * 100, 1), "%\n")
```

**Hint:** The probability of choosing alternative *i* is `exp(U_i)` divided by the sum of `exp(U_j)` for all alternatives.

**Question:** Which clinic captures the larger market share? Should clinics prioritize convenience (nearby location, shorter wait) or staffing MDs over NPs?

------------------------------------------------------------------------

# Summary and Key Takeaways

## Two Approaches, Same Goal

| Aspect | Rating-Based | Choice-Based |
|--------|:------------:|:------------:|
| Data Collection | Rate each profile (e.g., 1-10 scale) | Choose preferred option from a set |
| Analysis Method | OLS Linear Regression (`lm`) | Multinomial Logit (`mlogit`) |
| Data Format | One row per profile per respondent | Long format (one row per alternative) |
| Market Share Prediction | Proportional allocation | Logit formula ($e^V$ based) |
| Realism | Lower: no forced trade-offs | Higher: mimics actual decisions |

Both methods produce the same types of outputs: **part-worth utilities**, **attribute importance**, **WTP**, and **market share predictions**. Choice-based conjoint is generally preferred in practice because it forces respondents to make explicit trade-offs, mimicking real-world decision-making.

## Key Takeaways

1. **Part-worth utilities** (regression coefficients) tell us how much each attribute level contributes to overall preference

2. **Attribute importance** is calculated from the range of utilities within each attribute — the wider the range, the more influence that attribute has on preferences

3. **WTP** converts utility into dollar values using the price coefficient as the exchange rate: $\text{WTP} = -\beta_{\text{attribute}} / \beta_{\text{price}}$. This requires a price attribute in the study design

4. **Market share simulation** predicts how respondents would split across competing options — using proportional allocation (rating-based) or the logit formula (choice-based)

5. **Data structure matters**: Choice-based conjoint requires long-format data with proper indexing so that `mlogit` knows which alternatives belong to each choice set

6. **The managerial insight**: Conjoint analysis transforms subjective preferences into quantitative data that supports concrete decisions about pricing, feature prioritization, and competitive strategy


---

# References and Further Reading

## Statistical Methods

- Green, P. E., & Srinivasan, V. (1990). Conjoint analysis in marketing: New developments with implications for research and practice. *Journal of Marketing*, 54(4), 3-19.
- Louviere, J. J., Hensher, D. A., & Swait, J. D. (2000). *Stated Choice Methods: Analysis and Applications*. Cambridge University Press.
- McFadden, D. (1974). Conditional logit analysis of qualitative choice behavior. In P. Zarembka (Ed.), *Frontiers in Econometrics* (pp. 105-142). Academic Press.
- Orme, B. (2010). *Getting Started with Conjoint Analysis: Strategies for Product Design and Pricing Research* (2nd ed.). Research Publishers LLC.

## Healthcare Applications

Examples are motivated by the following:

- Sinaiko, A. D., & Hirth, R. A. (2011). Consumers, health insurance and dominated choices. *Journal of Health Economics*, 30(2), 450-457. https://doi.org/10.1016/j.jhealeco.2010.12.007

- Savira, F., Robinson, S., Toll, K., Spark, L., Thomas, E., Nesbitt, J., Frean, I., & Norman, R. (2023). Consumer preferences for telehealth in Australia: A discrete choice experiment. *PLOS ONE*, 18(3), e0283821. https://doi.org/10.1371/journal.pone.0283821
