---
title: "Conjoint Analysis in Healthcare Management (Practice Workbook)"
subtitle: "Rating-Based and Choice-Based Methods with R"
author: ""
Week: "Week 14"
date: 01/01/2026
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: false
    code-copy: true
    theme: custom-theme.scss
    highlight-style: github
    mainfont: "Clear Sans, -apple-system, system-ui, sans-serif"
    fontsize: 1.1em
    grid:
      sidebar-width: 200px
      body-width: 900px
      margin-width: 300px
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Install required packages if needed
required_packages <- c("ggplot2", "dplyr", "tidyr", "broom", "mlogit", "dfidx")
new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

library(ggplot2)
library(dplyr)
library(tidyr)
library(broom)
library(mlogit)
set.seed(2026)
```

# Introduction

## Recap: What is Conjoint Analysis?

Conjoint analysis is a powerful survey-based technique that reveals how people make decisions when facing trade-offs between multiple attributes of a product or service. Instead of simply asking "Do you prefer low prices?" (which always gets a "yes"), conjoint analysis forces respondents to make realistic choices that reveal their true priorities.

The key insight is that people evaluate products and services as bundles of attributes. A health insurance plan isn't just about the premium—it's a combination of premium, deductible, network size, and coverage. Conjoint analysis decomposes overall preferences into the contribution of each individual attribute, giving us **part-worth utilities** that quantify how much each feature matters.

## Two Methods You've Learned

In the lecture, you saw two approaches to conjoint analysis:

1. **Rating-Based Conjoint**: Respondents rate each profile on a scale (e.g., 1-10). We analyze this using ordinary least squares (OLS) regression to estimate part-worth utilities.

2. **Choice-Based Conjoint**: Respondents choose their preferred option from a set of alternatives. We analyze this using multinomial logistic regression (mlogit) to estimate part-worth utilities. This method better mimics real-world decision-making.

Both methods produce the same key outputs: part-worth utilities, attribute importance scores, willingness-to-pay estimates, and market share predictions.

## What You'll Do in This Workbook

Now it's your turn to apply these techniques. You'll work through two practice activities:

- **Part 1**: Rating-based conjoint analysis of telehealth program preferences
- **Part 2**: Choice-based conjoint analysis of primary care clinic selection 

Each activity includes starter code with blanks (indicated by `___`) for you to fill in. Take your time, refer back to the lecture notes, and focus on understanding what each step accomplishes.

---

# Practice Activity

## Case: Telehealth Program Preferences

### Background and Context

The COVID-19 pandemic accelerated the adoption of telehealth services, and healthcare organizations now face decisions about how to structure "hybrid" care models. Understanding patient preferences for telehealth attributes is crucial for designing services that people will actually use.

Research by Savira et al. (2023) in *PLOS ONE* examined consumer preferences for telehealth in Australia, finding that attributes like wait time, consultation mode, and cost significantly influenced preferences.

Your healthcare organization is designing a new telehealth program and needs to understand which attributes matter most to patients. You will analyze three key attributes:

| Attribute | Level 1 | Level 2 | Why It Matters |
|-----------|---------|---------|----------------|
| Wait Time | Same-day | 3 days | How quickly patients can be seen |
| Mode | Video | Phone | Communication richness |
| Co-pay | $20 | $40 | Out-of-pocket cost per visit |

### Your Tasks (30 minutes)

**Part 1: Rating-Based Conjoint Analysis (15 minutes)**

Using the `rating_telehealth` dataset provided below:

1. Estimate part-worth utilities using linear regression
2. Calculate attribute importance scores
3. Calculate willingness-to-pay for same-day appointments and video mode
4. Predict market shares for two program designs:
   - **Premium Program**: Same-day, Video, $40 co-pay
   - **Basic Program**: 3-day wait, Phone, $20 co-pay

**Part 2: Reflection and Discussion (5 minutes)**

Consider these questions:

- Which attribute matters most to patients? Does this match your intuition?
- If you were managing this telehealth program, how would you use these WTP estimates?
- What would you recommend based on the market share predictions?

**Instructions for Filling in Code Blanks:**

Throughout this workbook, you'll see blanks marked with `___`. Here's how to fill them:

- Most blanks should be filled with **variable names** from the dataset (e.g., `wait_sameday`, `copay`)
- Some blanks need **numeric values** (these are indicated in nearby comments)
- If you get stuck, refer back to the lecture notes, especially the Insurance Example section
- The variable names you need are listed in the "Dataset Information" section above each task

### Generate the Data

**STEP 1: Generate the Data**

Run the code chunk below first. This will create the `rating_telehealth` dataset that you'll use in your analysis.

```{r}
#| label: telehealth-data

# Create full factorial design
telehealth_profiles <- expand.grid(
  wait_sameday = c(1, 0),  # 1 = Same-day, 0 = 3 days
  mode_video = c(1, 0),    # 1 = Video, 0 = Phone
  copay = c(20, 40)
)
telehealth_profiles$profile_id <- 1:nrow(telehealth_profiles)

# Create display version
telehealth_display <- telehealth_profiles %>%
  mutate(
    wait_time = ifelse(wait_sameday == 1, "Same-day", "3 days"),
    mode = ifelse(mode_video == 1, "Video", "Phone")
  ) %>%
  select(wait_time, mode, copay, profile_id)

knitr::kable(telehealth_display,
             col.names = c("Wait Time", "Mode", "Co-pay ($)", "Profile ID"),
             align = "c",
             caption = "Telehealth Program Profiles")

# Simulate ratings from 100 respondents
n_respondents <- 100
rating_telehealth <- data.frame()

for (resp in 1:n_respondents) {
  for (i in 1:nrow(telehealth_profiles)) {
    utility <- 50 +
      4 * telehealth_profiles$wait_sameday[i] +
      2 * telehealth_profiles$mode_video[i] +
      -0.1 * telehealth_profiles$copay[i] +
      rnorm(1, 0, 2)

    rating <- round(pmin(pmax(utility / 5, 1), 10), 1)

    rating_telehealth <- rbind(rating_telehealth, data.frame(
      respondent_id = resp,
      profile_id = i,
      wait_sameday = telehealth_profiles$wait_sameday[i],
      mode_video = telehealth_profiles$mode_video[i],
      copay = telehealth_profiles$copay[i],
      rating = rating
    ))
  }
}

# Display first few rows
head(rating_telehealth, 10)
```

**Dataset Information:**

The `rating_telehealth` dataset contains 800 observations (100 respondents × 8 profiles each).

**Variables:**

- `respondent_id`: Unique identifier (1-100)
- `profile_id`: Profile identifier (1-8)
- `wait_sameday`: Binary indicator (1 = Same-day, 0 = 3 days)
- `mode_video`: Binary indicator (1 = Video, 0 = Phone)
- `copay`: Co-payment amount (20 or 40; continuous scale)
- `rating`: Satisfaction rating on 1-10 scale (outcome variable)

### Task 1: Estimate Part-Worth Utilities

Fit a linear regression model to estimate the part-worth utilities (coefficients) for each attribute.

```{r}
#| label: practice-task1-model
#| eval: false

# Estimate part-worth utilities using linear regression
model_rating_th <- lm(rating ~ ___ + ___ + ___,
                      data = rating_telehealth)

summary(model_rating_th)
```

**Hint:** The formula should include all three attribute variables from the dataset.

---

### Task 2: Calculate Attribute Importance

Calculate which attributes have the most impact on preferences.

```{r}
#| label: practice-task2-importance
#| eval: false

# Extract coefficients
coef_th <- coef(model_rating_th)

# Calculate utility ranges for each attribute
range_wait <- abs(coef_th["___"] * 1)  # Wait Time: Binary 0 vs 1
range_mode <- abs(coef_th["___"] * 1)  # Mode: Binary 0 vs 1
range_copay <- abs(coef_th["___"] * (40 - 20))  # Copay: $20 difference

# Calculate total range
total_range_th <- range_wait + range_mode + range_copay

# Calculate importance percentages
importance_rating_th <- data.frame(
  attribute = c("Wait Time", "Consultation Mode", "Co-pay"),
  importance_pct = c(range_wait, range_mode, range_copay) / total_range_th * 100
)

# Display results
knitr::kable(importance_rating_th, digits = 2,
             col.names = c("Attribute", "Importance (%)"),
             align = "c")
```

**Hint:** Fill in the coefficient names from your regression model. These should match the variable names you used in Task 1.

---

### Task 3: Calculate Willingness-to-Pay (WTP)

Calculate how much patients would pay for same-day access and video consultation.

```{r}
#| label: practice-task3-wtp
#| eval: false

# Calculate WTP for same-day access (vs 3-day wait)
wtp_sameday <- -coef_th["___"] / coef_th["___"]

# Calculate WTP for video consultation (vs phone)
wtp_video <- -coef_th["___"] / coef_th["___"]

# Display results
cat("WTP for Same-day access: $", round(wtp_sameday, 2), " per visit\n")
cat("WTP for Video consultation: $", round(wtp_video, 2), " per visit\n")
```

**Hint:** WTP = -β_attribute / β_price. The price attribute here is `copay`.

---

### Task 4: Predict Market Share

Compare two program designs and predict which would be more popular.

```{r}
#| label: practice-task4-marketshare
#| eval: false

# Define two competing programs
program_A <- data.frame(wait_sameday = 1, mode_video = 1, copay = 40)  # Premium: Same-day, Video, $40
program_B <- data.frame(wait_sameday = 0, mode_video = 0, copay = 20)  # Basic: 3-day, Phone, $20

# Predict ratings
rating_A_th <- predict(model_rating_th, ____)
rating_B_th <- predict(model_rating_th, ____)

# Calculate proportional market share
total_rating_th <- rating_A_th + rating_B_th
share_A_th <- rating_A_th / total_rating_th
share_B_th <- rating_B_th / total_rating_th

# Display results
cat("Premium Program (Same-day Video, $40): ", round(share_A_th * 100, 1), "%\n")
cat("Basic Program (3-day Phone, $20): ", round(share_B_th * 100, 1), "%\n")
```

**Hint:** Use the `predict()` function with your model and the program data frames.

---

## Part 2: Choice-Based Conjoint Practice (Optional/Advanced)

### Case: Primary Care Clinic Selection

**Background and Context**

As healthcare systems become more competitive, understanding what drives patients' choice of primary care clinic is crucial. Patients must trade off multiple factors: convenience (location, wait times), provider credentials, and cost.

You'll analyze patient preferences for primary care clinics using **choice-based conjoint analysis** with four key attributes:

| Attribute | Level 1 | Level 2 | Why It Matters |
|-----------|---------|---------|----------------|
| Distance from Home | 5 minutes | 25 minutes | Travel convenience |
| Appointment Wait | Same-week | 3 weeks | Access to care |
| Provider Type | MD (Physician) | NP (Nurse Practitioner) | Professional credentials |
| Copay | $25 | $50 | Out-of-pocket cost |

### Your Tasks (20-25 minutes)

**Using choice-based conjoint (mlogit):**

1. Generate the choice data (code provided below)
2. Reshape data to long format
3. Index data with dfidx()
4. Estimate multinomial logit model
5. Calculate attribute importance
6. Calculate WTP for key attributes
7. Predict market shares for two clinic designs

### Generate the Choice Data

```{r}
#| label: clinic-choice-data

# Create full factorial design for clinic profiles
clinic_profiles <- expand.grid(
  distance_5min = c(1, 0),     # 1 = 5 min, 0 = 25 min
  wait_sameweek = c(1, 0),     # 1 = Same-week, 0 = 3 weeks
  provider_md = c(1, 0),        # 1 = MD, 0 = NP
  copay = c(25, 50)
)
clinic_profiles$profile_id <- 1:nrow(clinic_profiles)

# Simulate choice data from 100 respondents
n_respondents <- 100
n_choice_sets <- 8

choice_clinic <- data.frame()

for (resp in 1:n_respondents) {
  for (cs in 1:n_choice_sets) {
    # Randomly select two different profiles
    profiles <- sample(1:16, 2)

    alt_A <- clinic_profiles[profiles[1], ]
    alt_B <- clinic_profiles[profiles[2], ]

    # Calculate utilities (true preferences for simulation)
    # Distance: +2.5 for nearby clinic
    # Wait: +3.0 for same-week appointments
    # Provider: +1.0 for MD
    # Copay: -0.08 per dollar
    utility_A <- 2.5 * alt_A$distance_5min +
                 3.0 * alt_A$wait_sameweek +
                 1.0 * alt_A$provider_md +
                 -0.08 * alt_A$copay +
                 rnorm(1, 0, 0.5)

    utility_B <- 2.5 * alt_B$distance_5min +
                 3.0 * alt_B$wait_sameweek +
                 1.0 * alt_B$provider_md +
                 -0.08 * alt_B$copay +
                 rnorm(1, 0, 0.5)

    choice_A <- as.integer(utility_A > utility_B)

    choice_clinic <- rbind(choice_clinic, data.frame(
      respondent_id = resp,
      choice_set = cs,
      distance_5min_A = alt_A$distance_5min,
      wait_sameweek_A = alt_A$wait_sameweek,
      provider_md_A = alt_A$provider_md,
      copay_A = alt_A$copay,
      distance_5min_B = alt_B$distance_5min,
      wait_sameweek_B = alt_B$wait_sameweek,
      provider_md_B = alt_B$provider_md,
      copay_B = alt_B$copay,
      choice_A = choice_A
    ))
  }
}

# Display first few rows
head(choice_clinic, 10)
```

**Dataset Information:**

The `choice_clinic` dataset contains 800 observations (100 respondents × 8 choice sets each).

**Variables:**

- `respondent_id`: Unique identifier (1-100)
- `choice_set`: Choice set identifier (1-8)
- `distance_5min_A/B`: Binary (1 = 5 min away, 0 = 25 min away)
- `wait_sameweek_A/B`: Binary (1 = Same-week, 0 = 3 weeks)
- `provider_md_A/B`: Binary (1 = MD, 0 = NP)
- `copay_A/B`: Copay amount (25 or 50)
- `choice_A`: Binary outcome (1 = chose A, 0 = chose B)

### Task 1: Reshape Data to Long Format

Choice-based conjoint requires data in "long format" where each alternative gets its own row.

```{r}
#| label: practice-clinic-task1-reshape
#| eval: false

library(tidyr)

# Create unique choice set IDs
choice_clinic <- choice_clinic %>%
  mutate(choice_set_id = paste0(respondent_id, "_", choice_set))

# Reshape from wide to long format
clinic_long <- choice_clinic %>%
  pivot_longer(
    cols = c(distance_5min_A, distance_5min_B,
             wait_sameweek_A, wait_sameweek_B,
             provider_md_A, provider_md_B,
             copay_A, copay_B),
    names_to = c(".value", "alternative"),
    names_pattern = "(.*)_([AB])$"  # Match everything before the final _A or _B
  ) %>%
  mutate(
    # Create the choice variable (1 = chosen, 0 = not chosen)
    # Logic: If alternative A was chosen (choice_A = 1):
    #   - Row for alternative A gets choice = 1
    #   - Row for alternative B gets choice = 0
    # If alternative B was chosen (choice_A = 0):
    #   - Row for alternative A gets choice = 0
    #   - Row for alternative B gets choice = 1
    choice = case_when(
      alternative == "A" & choice_A == 1 ~ ___,
      alternative == "A" & choice_A == 0 ~ ___,
      alternative == "B" & choice_A == 1 ~ ___,
      alternative == "B" & choice_A == 0 ~ ___
    )
  ) %>%
  select(choice_set = choice_set_id, alternative,
         distance_5min, wait_sameweek, provider_md, copay, choice) %>%
  arrange(choice_set, alternative) %>%
  as.data.frame()

# Check the result
head(clinic_long, 10)
```

**Hint:** Fill in 1 or 0 for each case_when condition based on whether that alternative/choice combination means "chosen."

---

### Task 2: Index Data for mlogit

Prepare the data structure that mlogit requires.

```{r}
#| label: practice-clinic-task2-index
#| eval: false

library(dfidx)

# Create indexed data format
clinic_mlogit <- dfidx(clinic_long,
                       shape = "___",      # Data format type
                       choice = "___",     # Column containing choice outcome
                       idx = c("___", "___"))  # c(choice set identifier, alternative identifier)
```

**Hint:** The shape should indicate the data format. The choice parameter is the name of your binary choice variable. The idx parameter needs the names of your choice set identifier and alternative identifier columns.

---

### Task 3: Estimate Multinomial Logit Model

Fit the choice model to estimate part-worth utilities.

```{r}
#| label: practice-clinic-task3-model
#| eval: false

library(mlogit)

# Estimate multinomial logit model
model_clinic <- mlogit(choice ~ ___ + ___ + ___ + ___ | 0,
                       data = clinic_mlogit)

summary(model_clinic)
```

**Hint:** Include all four attributes from your dataset. Refer to the variable names listed in the "Dataset Information" section above. The `| 0` means no alternative-specific constants.

---

### Task 4: Calculate Attribute Importance

Determine which attributes matter most in clinic selection.

```{r}
#| label: practice-clinic-task4-importance
#| eval: false

# Extract coefficients
coef_clinic <- coef(model_clinic)

# Calculate utility ranges
range_distance <- abs(coef_clinic["___"] * 1)   # Distance: Binary 5 min vs 25 min
range_wait <- abs(coef_clinic["___"] * 1)       # Wait Time: Binary Same-week vs 3 weeks
range_provider <- abs(coef_clinic["___"] * 1)   # Provider Type: Binary MD vs NP
range_copay <- abs(coef_clinic["___"] * (50 - 25))  # Copay: $25 difference

# Calculate total range
total_range <- range_distance + range_wait + range_provider + range_copay

# Calculate importance percentages
importance_clinic <- data.frame(
  attribute = c("Distance", "Appointment Wait", "Provider Type", "Copay"),
  importance_pct = c(range_distance, range_wait, range_provider, range_copay) / total_range * 100
)

# Display results
knitr::kable(importance_clinic, digits = 2,
             col.names = c("Attribute", "Importance (%)"),
             align = "c")
```

**Hint:** Fill in the coefficient names matching the model variables.

---

### Task 5: Calculate Willingness-to-Pay

Calculate how much patients would pay for each clinic feature.

```{r}
#| label: practice-clinic-task5-wtp
#| eval: false

# Calculate WTP for each attribute
wtp_nearby <- -coef_clinic["___"] / coef_clinic["___"]      # For Distance: 5 min vs 25 min distance
wtp_sameweek <- -coef_clinic["___"] / coef_clinic["___"]    # For Wait Time: Same-week vs 3-week wait
wtp_md <- -coef_clinic["___"] / coef_clinic["___"]          # For Provider Type: MD vs NP

# Display results
cat("WTP for Nearby Clinic (5 min vs 25 min): $", round(wtp_nearby, 2), " per visit\n")
cat("WTP for Same-week Appointment: $", round(wtp_sameweek, 2), " per visit\n")
cat("WTP for MD vs NP: $", round(wtp_md, 2), " per visit\n")
```

**Hint:** WTP = -β_attribute / β_copay. The copay variable is the "price" in this analysis.

---

### Task 6: Predict Market Share

Compare two competing clinics using the multinomial logit formula.

```{r}
#| label: practice-clinic-task6-marketshare
#| eval: false

# Define two competing clinics
# Clinic A: Near (5 min), Same-week, NP, $50 copay
# Clinic B: Far (25 min), 3-week wait, MD, $25 copay

# Calculate systematic utilities
utility_A_clinic <- coef_clinic["distance_5min"] * 1 +
                    coef_clinic["wait_sameweek"] * 1 +
                    coef_clinic["provider_md"] * 0 +
                    coef_clinic["copay"] * 50

utility_B_clinic <- coef_clinic["distance_5min"] * 0 +
                    coef_clinic["wait_sameweek"] * 0 +
                    coef_clinic["provider_md"] * 1 +
                    coef_clinic["copay"] * 25

# Apply multinomial logit formula
share_A_clinic <- _____ / (exp(utility_A_clinic) + exp(utility_B_clinic))
share_B_clinic <- _____ / (exp(utility_A_clinic) + exp(utility_B_clinic))

# Display results
cat("Clinic A (Convenient, NP, Higher copay): ", round(share_A_clinic * 100, 1), "%\n")
cat("Clinic B (Far, MD, Lower copay): ", round(share_B_clinic * 100, 1), "%\n")
```

**Hint:** Fill in the numerators using the multinomial logit formula. The probability of choosing alternative i is exp(U_i) divided by the sum of exp(U_j) for all alternatives.

### Key Questions to Consider

After completing your analysis, reflect on:

1. **Which attribute is most important for clinic choice?**
   - Does distance or appointment wait matter more?
   - How important is provider type (MD vs NP)?

2. **What do the WTP estimates tell you?**
   - How much would patients pay to reduce travel time from 25 to 5 minutes?
   - Is the WTP for same-week appointments larger than for MD vs NP?

3. **Based on market share predictions:**
   - Should clinics prioritize convenience or provider credentials?
   - What's the optimal clinic configuration?

4. **How do these results compare to the telehealth example?**
   - Are similar patterns emerging (e.g., access time most important)?
   - What's different about in-person vs virtual care preferences?

**Hints:**

- The choice variable should be 1 for chosen alternative, 0 otherwise
- For dfidx: shape = "long", choice = "choice", idx = c("choice_set", "alternative")
- Model formula: Include all four attributes (distance_5min, wait_sameweek, provider_md, copay)
- For WTP: WTP = -β_attribute / β_copay

---

# References

## Healthcare Applications

Examples are motivated by the following:

- Sinaiko, A. D., & Hirth, R. A. (2011). Consumers, health insurance and dominated choices. *Journal of Health Economics*, 30(2), 450-457. https://doi.org/10.1016/j.jhealeco.2010.12.007

- Savira, F., Robinson, S., Toll, K., Spark, L., Thomas, E., Nesbitt, J., Frean, I., & Norman, R. (2023). Consumer preferences for telehealth in Australia: A discrete choice experiment. *PLOS ONE*, 18(3), e0283821. https://doi.org/10.1371/journal.pone.0283821
