---
title: "Linear Regression in Healthcare Analytics"
subtitle: "Part 1: Simple and Multiple Linear Regression"
author: "MBA642: Business Analytics in Healthcare Management"
Week: "Week 10"
date: 01/01/2026
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: false
    code-copy: true
    theme: custom-theme.scss
    highlight-style: github
    mainfont: "Clear Sans, -apple-system, system-ui, sans-serif"
    fontsize: 1.1em
    grid:
      sidebar-width: 200px
      body-width: 900px
      margin-width: 300px
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Load required packages
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
library(car)
library(GGally)
library(patchwork)

# Set seed for reproducibility
set.seed(42)
```

# Introduction to Linear Regression

## Why Linear Regression Matters

Linear regression is one of the most fundamental and widely used statistical techniques in data analytics. Healthcare managers will frequently encounter situations where understanding relationships between variables is crucial for decision-making:

- **Resource allocation**: How does nurse staffing level affect patient outcomes?
- **Cost management**: What factors drive hospital readmission costs?
- **Quality improvement**: How does patient wait time impact satisfaction scores?
- **Performance prediction**: Can we predict length of stay based on patient characteristics?

Linear regression provides a powerful framework to:

1. **Quantify relationships** between variables (e.g., "For every additional nurse per patient, satisfaction increases by X points")
2. **Test hypotheses** about whether relationships are statistically significant
3. **Make predictions** about future outcomes
4. **Control for confounding variables** when examining effects of interest

## Overview of Linear Regression Types

| Type | Number of Predictors | Example |
|------|---------------------|---------|
| Simple Linear Regression | 1 | Patient age → Length of stay |
| Multiple Linear Regression | 2+ | Patient age + Severity + Insurance → Length of stay |

## Key Concepts and Terminology

Before we dive in, let's establish key terminology:

- **Dependent Variable (DV)**: The outcome we want to predict or explain (also called response variable or Y)
- **Independent Variable (IV)**: The predictor(s) we use to explain the outcome (also called predictor variable or X)
- **Regression Coefficient (β)**: The slope that quantifies the relationship between an IV and the DV
- **Intercept (α or β₀)**: The predicted value of Y when all X variables equal zero
- **Residual (ε)**: The difference between observed and predicted values

------------------------------------------------------------------------

# Simple Linear Regression

## What is it?

Simple linear regression models the relationship between **one independent variable (X)** and **one dependent variable (Y)** using a linear model (straight line). It is the foundation for understanding more complex regression models and allows us to quantify how changes in one variable are associated with changes in another.

## When to Use It

Use simple linear regression when you want to:

- Examine the relationship between two continuous variables
- Predict an outcome based on a single predictor
- Quantify the strength and direction of a linear relationship
- Test whether a relationship is statistically significant

## The Concept

Simple linear regression finds the "best-fitting" line through the data points using a method called Ordinary Least Squares (OLS).

### The Model Equation

The simple linear regression model is expressed as:

$$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$$

Where:

- $Y_i$ = Observed value of the dependent variable for observation $i$
- $\beta_0$ = Intercept (the value of Y when X = 0)
- $\beta_1$ = Slope (the change in Y for a one-unit increase in X)
- $X_i$ = Value of the independent variable for observation $i$
- $\epsilon_i$ = Residual for observation $i$

### Understanding the Components

**The Intercept (β₀)**:
The intercept represents the predicted value of Y when X equals zero. In healthcare contexts, this may or may not have a meaningful interpretation. For example, if X is "years of nursing experience," an intercept would represent the predicted outcome for a nurse with zero years of experience.

**The Slope (β₁)**:
The slope is the key parameter of interest. It tells us:

- **Direction**: Positive slope = positive relationship; Negative slope = negative relationship
- **Magnitude**: How much Y changes for each one-unit increase in X
- **Statistical significance**: Whether the relationship is likely to exist in the population (not just due to chance)

## The Ordinary Least Squares (OLS) Method

### How OLS Works

Linear regression finds the "best" line by minimizing the **sum of squared residuals (SSR)**:

$$SSR = \sum_{i=1}^{n} (Y_i - \hat{Y}_i)^2 = \sum_{i=1}^{n} \epsilon_i^2$$

Where $\hat{Y}_i$ is the predicted value for observation $i$.

Why squared residuals?

1. Squaring ensures positive and negative errors don't cancel out
2. Squaring penalizes larger errors more heavily
3. Leads to closed-form mathematical solutions

### Visual Intuition

```{r}
#| label: fig-ols-concept
#| fig-cap: "OLS finds the line that minimizes the sum of squared residuals (vertical distances)"
#| fig-width: 6
#| fig-height: 4

# Create example data
example_data <- tibble(
  x = c(1, 2, 3, 4, 5, 6, 7),
  y = c(1.8, 2.5, 3.8, 3.7, 5.1, 5.8, 6.2)
)

# Fit model
fit <- lm(y ~ x, data = example_data)

# Create plot showing residuals
example_data <- example_data %>%
  mutate(predicted = predict(fit),
         residual = y - predicted)

ggplot(example_data, aes(x = x, y = y)) +
  geom_segment(aes(xend = x, yend = predicted),
               color = "red", linetype = "dashed", size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "blue", size = 1) +
  geom_point(size = 4, color = "black") +
  annotate("text", x = 4.5, y = 3.5,
           label = paste("Residuals: red dashed lines\nOLS minimizes:",
                        expression(sum(epsilon^2))),
           hjust = 0, size = 3.5) +
  labs(x = "Independent Variable (X)",
       y = "Dependent Variable (Y)",
       title = "OLS Regression: Minimizing Squared Residuals") +
  theme_minimal(base_size = 12)
```

## Using the Fitted Model for Predictions

Once OLS finds the best-fitting line by estimating $\beta_0$ (intercept) and $\beta_1$ (slope), we can use the fitted model to make predictions for new observations.

**The prediction equation** is:

$$\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 X$$

Where:

- $\hat{Y}$ = Predicted value of the dependent variable
- $\hat{\beta}_0$ = Estimated intercept from the fitted model
- $\hat{\beta}_1$ = Estimated slope from the fitted model
- $X$ = Value of the independent variable for which we want to make a prediction

**Key insight**: The "hat" notation (^) indicates these are **estimated** values from our sample data, not the true population parameters. We use the estimated coefficients to predict outcomes for:

1. **Observations within our dataset** - to assess model fit
2. **New observations** - to forecast outcomes based on predictor values
3. **Hypothetical scenarios** - to evaluate "what if" questions (e.g., "What if we reduce wait time to 30 minutes?")

Let's see this in action with a real healthcare example.

------------------------------------------------------------------------

## Example Case: Patient Satisfaction and Wait Time

### Background and Context

Research has consistently shown that emergency department wait times are negatively associated with patient satisfaction. A 2021 study found that being informed about waiting time showed a moderately strong positive correlation with patients' overall rating, highlighting the importance of wait time management for patient experience.

Let's examine whether emergency department wait time affects patient satisfaction scores using synthetic data that mirrors these real-world relationships.

### Creating Synthetic Healthcare Data

```{r}
#| label: create-simple-data

# Generate synthetic ED data
n <- 150

ed_data <- tibble(
  patient_id = 1:n,
  wait_time_minutes = round(rnorm(n, mean = 45, sd = 20), 0),
  satisfaction_score = round(85 - 0.4 * wait_time_minutes + rnorm(n, 0, 8), 1)
) %>%
  # Ensure realistic bounds
  mutate(
    wait_time_minutes = pmax(5, pmin(wait_time_minutes, 120)),
    satisfaction_score = pmax(0, pmin(satisfaction_score, 100))
  )

# Preview the data
head(ed_data, 10) %>%
  kable(caption = "Sample of Emergency Department Patient Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Exploratory Data Analysis

Before running regression, always visualize your data. Let's take a look at the pattern using a scatterplot.

```{r}
#| label: fig-eda-simple
#| fig-cap: "Relationship between ED wait time and patient satisfaction"
#| fig-height: 4
#| fig-width: 6

# Scatter plot with trend line
ggplot(ed_data, aes(x = wait_time_minutes, y = satisfaction_score)) +
  geom_point(alpha = 0.6, size = 2, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", fill = "pink") +
  labs(
    x = "Wait Time (minutes)",
    y = "Patient Satisfaction Score (0-100)",
    title = "ED Wait Time vs. Patient Satisfaction",
    subtitle = "Each point represents one patient"
  ) +
  theme_minimal(base_size = 12)
```

### Fitting the Simple Linear Regression Model

Let's fit the simple linear regression model to quantify the relationship. We can fit a linear model using `lm()` function in R.

```{r}
#| label: fit-simple-model

# Fit the model
model_simple <- lm(satisfaction_score ~ wait_time_minutes, data = ed_data)

# Display model summary
summary(model_simple)
```

### The Fitted Regression Equation

Based on the model output above, we can write our fitted regression equation with the estimated coefficients:

```{r}
#| echo: false
# Extract coefficients for the equation
b0 <- round(coef(model_simple)[1], 2)
b1 <- round(coef(model_simple)[2], 3)
```

$$\widehat{\text{Satisfaction}} = `r b0` + (`r b1`) \times \text{Wait Time}$$

Or more generally:

$$\hat{Y}_i = `r b0` + (`r b1`) \times X_i$$

Where:

- $\hat{Y}_i$ = Predicted patient satisfaction score for patient $i$
- $X_i$ = Wait time in minutes for patient $i$
- `r b0` = Estimated intercept ($\hat{\beta}_0$)
- `r b1` = Estimated slope ($\hat{\beta}_1$)

This equation allows us to predict patient satisfaction for any given wait time. For example:

- If wait time = 30 minutes: $\hat{Y} = `r b0` + (`r b1`) \times 30 = `r round(b0 + b1 * 30, 1)`$
- If wait time = 60 minutes: $\hat{Y} = `r b0` + (`r b1`) \times 60 = `r round(b0 + b1 * 60, 1)`$

### Interpreting the Results

Let's break down the output systematically:

#### 1. The Intercept

```{r}
#| echo: false
intercept <- round(coef(model_simple)[1], 2)
```

**Value**: β₀ = `r intercept`

**Interpretation**: When wait time is 0 minutes, the predicted patient satisfaction score is `r intercept` points. This represents the baseline satisfaction when there is no wait.

#### 2. The Slope (Regression Coefficient)

```{r}
#| echo: false
slope <- round(coef(model_simple)[2], 3)
p_value <- summary(model_simple)$coefficients[2, 4]
```

**Value**: β₁ = `r slope`

**Interpretation** (The Three Key Questions):

1. **Direction**: The coefficient is **negative** (`r slope`), indicating that longer wait times are associated with **lower** satisfaction scores.

2. **Magnitude**: For every **one-minute increase** in wait time, patient satisfaction decreases by **`r abs(slope)` points** on the 0-100 scale.

3. **Statistical Significance**: The p-value is `r format(p_value, scientific = TRUE)`, which is less than 0.05. We **reject the null hypothesis** that there is no relationship between wait time and satisfaction. The relationship is statistically significant.

#### 3. Model Fit (R-squared)

```{r}
#| echo: false
r_squared <- round(summary(model_simple)$r.squared, 3)
```

**R² = `r r_squared`**

**Interpretation**: Wait time explains **`r r_squared * 100`%** of the variance in patient satisfaction scores. The remaining `r (1 - r_squared) * 100`% is explained by other factors not included in this model (e.g., staff friendliness, treatment effectiveness, facility cleanliness).

#### Visualizing R-squared: Explained vs. Unexplained Variance

To understand R² visually, let's see how the regression model reduces prediction error compared to simply using the mean:

```{r}
#| label: fig-rsquared-concept
#| fig-cap: "R² measures how much variance is explained by the regression model"
#| fig-height: 6
#| fig-width: 10

# Calculate variance components
ed_data_variance <- ed_data %>%
  mutate(
    predicted = predict(model_simple),
    mean_y = mean(satisfaction_score),
    # Total variance: deviation from mean
    total_dev = satisfaction_score - mean_y,
    # Explained variance: deviation of prediction from mean
    explained_dev = predicted - mean_y,
    # Residual variance: deviation from prediction
    residual_dev = satisfaction_score - predicted
  )

# Calculate sums of squares
TSS <- sum(ed_data_variance$total_dev^2)
ESS <- sum(ed_data_variance$explained_dev^2)
RSS <- sum(ed_data_variance$residual_dev^2)
R2 <- ESS / TSS

# Create two comparison plots
p1 <- ggplot(ed_data_variance, aes(x = wait_time_minutes, y = satisfaction_score)) +
  geom_hline(yintercept = mean(ed_data_variance$satisfaction_score),
             color = "darkgreen", size = 1.2, linetype = "dashed") +
  geom_segment(aes(xend = wait_time_minutes, yend = mean_y),
               color = "purple", alpha = 0.4, size = 0.6) +
  geom_point(size = 2, color = "black", alpha = 0.6) +
  annotate("text", x = 100, y = mean(ed_data_variance$satisfaction_score) + 3,
           label = "Mean (baseline model)", color = "darkgreen", size = 3, hjust = 1) +
  annotate("text", x = 20, y = 95,
           label = paste0("Total Variance (TSS)\nAll deviations from mean\nTSS = ",
                         round(TSS, 0)),
           color = "purple", size = 3, hjust = 0) +
  labs(
    x = "Wait Time (minutes)",
    y = "Patient Satisfaction Score",
    title = "Baseline Model: Using Only the Mean",
    subtitle = "Purple lines show total variance (what we're trying to explain)"
  ) +
  theme_minimal(base_size = 10)

p2 <- ggplot(ed_data_variance, aes(x = wait_time_minutes, y = satisfaction_score)) +
  geom_hline(yintercept = mean(ed_data_variance$satisfaction_score),
             color = "darkgreen", size = 0.8, linetype = "dotted", alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "blue", size = 1.2) +
  geom_segment(aes(xend = wait_time_minutes, yend = predicted),
               color = "red", alpha = 0.5, size = 0.6, linetype = "dashed") +
  geom_point(size = 2, color = "black", alpha = 0.6) +
  annotate("text", x = 100, y = 45,
           label = "Regression line", color = "blue", size = 3, hjust = 1) +
  annotate("text", x = 20, y = 95,
           label = paste0("Residual Variance (RSS)\nUnexplained by model\nRSS = ",
                         round(RSS, 0)),
           color = "red", size = 3, hjust = 0) +
  annotate("text", x = 20, y = 82,
           label = paste0("R² = ", round(R2, 3), " = ", round(R2*100, 1), "%\n",
                         "of variance explained"),
           color = "blue", size = 3, hjust = 0, fontface = "bold") +
  labs(
    x = "Wait Time (minutes)",
    y = "Patient Satisfaction Score",
    title = "Regression Model: Using Wait Time to Predict Satisfaction",
    subtitle = "Red lines show remaining (unexplained) variance after using the model"
  ) +
  theme_minimal(base_size = 10)

# Combine plots
p1 + p2
```

**Key Insight**:

- **Left plot**: Without any predictors, we'd predict every patient's satisfaction as the mean (green line). The purple lines show how far each observation deviates from this mean. This is the **total variance (TSS)** we want to explain.

- **Right plot**: With wait time as a predictor, the regression line (blue) does a better job. The red dashed lines show the remaining errors (residuals). This is the **residual variance (RSS)** left unexplained.

- **R² = 1 - (RSS/TSS)** = The proportion of total variance that our model explains. An R² of `r round(R2, 3)` means the regression model reduces prediction error by `r round(R2*100, 1)`% compared to just using the mean.

### Making Predictions

One powerful application of regression is predicting outcomes for new observations. Let's assume that we have new patients and their waiting times.

```{r}
#| label: simple-new-data-points

# Predict satisfaction for specific wait times
new_patients <- tibble(
  wait_time_minutes = c(15, 30, 45, 60, 90)
)
head(new_patients)

```


By using the `predict` function we can get predicted satisfaction scores for these new patients based on their wait times.

```{r}
#| label: predictions

predictions <- new_patients %>%
  mutate(
    predicted_satisfaction = predict(model_simple, newdata = new_patients),
    prediction_interval = predict(model_simple, newdata = new_patients,
                                  interval = "prediction") %>%
      as_tibble()
  ) %>%
  unnest(prediction_interval)

predictions %>%
  mutate(across(where(is.numeric), ~round(., 2))) %>%
  rename(
    "Wait Time (min)" = wait_time_minutes,
    "Predicted Score" = predicted_satisfaction,
    "Fit" = fit,
    "Lower 95%" = lwr,
    "Upper 95%" = upr
  ) %>%
  select(-Fit) %>%
  kable(caption = "Predicted Satisfaction Scores by Wait Time (with 95% Prediction Intervals)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

#### Manual Calculation Example

Let's manually calculate one prediction to understand what the `predict()` function is doing. We'll predict satisfaction for a patient with a **45-minute wait time**.

**Step 1: Recall the fitted equation**

$$\widehat{\text{Satisfaction}} = `r b0` + (`r b1`) \times \text{Wait Time}$$

**Step 2: Substitute the wait time value**

For Wait Time = 45 minutes:

$$\widehat{\text{Satisfaction}} = `r b0` + (`r b1`) \times 45$$

**Step 3: Calculate**

```{r}
#| label: manual-prediction-calculation

# Extract coefficients
intercept_val <- coef(model_simple)[1]
slope_val <- coef(model_simple)[2]

# Manual calculation
wait_time_example <- 45
predicted_satisfaction_manual <- intercept_val + slope_val * wait_time_example

# Display the calculation
cat("Intercept (β₀):", round(intercept_val, 2), "\n")
cat("Slope (β₁):", round(slope_val, 3), "\n")
cat("Wait Time:", wait_time_example, "minutes\n\n")
cat("Calculation:", round(intercept_val, 2), "+",
    paste0("(", round(slope_val, 3), ")"), "×", wait_time_example, "\n")
cat("          =", round(intercept_val, 2), "+",
    round(slope_val * wait_time_example, 2), "\n")
cat("          =", round(predicted_satisfaction_manual, 2), "\n")
```

**Result**: The predicted satisfaction score for a patient with a 45-minute wait is **`r round(predicted_satisfaction_manual, 2)`** points.

You can verify this matches the value in the prediction table above for the 45-minute wait time!

#### Practical Application

If your ED currently has an average wait time of 60 minutes and you implement process improvements to reduce it to 30 minutes, you could expect satisfaction scores to increase by approximately `r abs(round(30 * slope, 1))` points (for every one-minute reduction in waiting time, satisfaction scores improve by `r abs(round(slope, 3))` points).

### Visualizing the Fitted Model

```{r}
#| label: fig-fitted-model
#| fig-cap: "Fitted regression line with prediction intervals"
#| fig-height: 4
#| fig-width: 6

# Create prediction data
pred_data <- tibble(
  wait_time_minutes = seq(5, 120, by = 1)
)

# Get prediction intervals and bind to data
pred_intervals <- predict(model_simple, newdata = pred_data, interval = "prediction")
pred_data <- bind_cols(pred_data, as_tibble(pred_intervals))

ggplot() +
  geom_ribbon(data = pred_data,
              aes(x = wait_time_minutes, ymin = lwr, ymax = upr),
              fill = "lightblue", alpha = 0.5) +
  geom_point(data = ed_data,
             aes(x = wait_time_minutes, y = satisfaction_score),
             alpha = 0.5, color = "steelblue") +
  geom_line(data = pred_data,
            aes(x = wait_time_minutes, y = fit),
            color = "darkred", size = 1.2) +
  labs(
    x = "Wait Time (minutes)",
    y = "Patient Satisfaction Score",
    title = "Simple Linear Regression: ED Wait Time → Patient Satisfaction",
    subtitle = paste0("Satisfaction = ", round(intercept, 1), " + (",
                     round(slope, 2), ") × Wait Time")
  ) +
  annotate("text", x = 90, y = 90,
           label = paste0("R² = ", round(r_squared, 3)),
           size = 4, fontface = "bold") +
  theme_minimal(base_size = 12)
```

### Practice Checkpoint 1 {.unnumbered}

::: {.callout-important}
## Try It Yourself! (10 minutes)

**Pause here!** Open `week11_linear_regression_practice.qmd` and complete:

- **Activity 1: Simple Linear Regression Practice**
  - Task 1.1: Fit a simple regression model
  - Task 1.2: Interpret the coefficients
  - Task 1.3: Visualize the relationship
  - Task 1.4: Make predictions

This will help you practice fitting and interpreting simple linear regression models on your own.

**After 10 minutes**, we'll continue with the lecture.
:::

------------------------------------------------------------------------

# Multiple Linear Regression

## What is it?

Multiple linear regression extends simple regression to model the relationship between **multiple independent variables** and **one dependent variable**. It allows us to examine how several predictors simultaneously influence an outcome.

## When to Use It

Use multiple linear regression when you need to:

- Include multiple predictors to better explain variance in an outcome
- Control for confounding variables when examining specific relationships
- Compare the relative importance of different factors
- Make predictions based on multiple characteristics simultaneously

## Extending to Multiple Predictors

In reality, outcomes of interest are influenced by many factors simultaneously. Multiple linear regression allows us to account for this complexity.

### The Model Equation

$$Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_k X_{ki} + \epsilon_i$$

Where:

- $\beta_0$ = Intercept
- $\beta_1, \beta_2, \ldots, \beta_k$ = Regression coefficients for each predictor
- $X_{1i}, X_{2i}, \ldots, X_{ki}$ = Values of predictors for observation $i$
- $\epsilon_i$ = Residual

In multiple regression, each coefficient represents the effect of that variable **holding all other variables constant** (also called the "partial" or "conditional" effect).

## An Example Case: Hospital Length of Stay

### Background and Context

A 2023 systematic review identified age, comorbidities, and severity as the most frequently cited predictors of hospital length of stay across 15 studies. Understanding these factors is crucial for healthcare operations and resource planning.

Let's predict hospital length of stay (LOS) based on three key patient characteristics using synthetic data that reflects these evidence-based relationships:

1. **Age** - Patient age in years
2. **Severity Score** - Illness severity on a 1-10 scale
3. **Number of Comorbidities** - Count of concurrent chronic conditions

### Creating Comprehensive Healthcare Data

Let's create a dataset using the code below.

```{r}
#| label: create-multiple-data

# Generate synthetic hospital data
n <- 300

hospital_data <- tibble(
  patient_id = 1:n,
  age = round(rnorm(n, mean = 55, sd = 18)),
  severity_score = round(runif(n, 1, 10), 1),  # 1-10 scale
  num_comorbidities = rpois(n, lambda = 2)
) %>%
  mutate(
    # Ensure realistic age bounds
    age = pmax(18, pmin(age, 95)),
    # Generate LOS based on predictors
    length_of_stay = round(
      1.5 +                          # Base stay
      0.04 * age +                   # Age effect
      0.6 * severity_score +         # Severity effect
      0.8 * num_comorbidities +      # Comorbidity effect
      rnorm(n, 0, 1.5),              # Random error
      1
    ),
    # Ensure positive LOS
    length_of_stay = pmax(1, length_of_stay)
  )

# Summary statistics
hospital_data %>%
  select(age, severity_score, num_comorbidities, length_of_stay) %>%
  summary()
```

### Exploratory Data Analysis

Before fitting the model, let's visually explore the distributions of our variables and their pairwise relationships. This helps us understand the data structure and identify potential patterns.

```{r}
#| label: fig-eda-multiple
#| fig-cap: "Pairwise relationships between predictors and length of stay"
#| fig-height: 7
#| fig-width: 9

# Pairs plot for continuous variables
hospital_data %>%
  select(age, severity_score, num_comorbidities, length_of_stay) %>%
  ggpairs(
    lower = list(continuous = wrap("smooth", alpha = 0.3, size = 0.5)),
    upper = list(continuous = wrap("cor", size = 3)),
    diag = list(continuous = wrap("densityDiag", alpha = 0.5))
  ) +
  theme_minimal(base_size = 10)
```

### Fitting the Multiple Linear Regression Model

Let's fit the linear model. Same as simple linear regression, we can use the `lm()` function in R.

```{r}
#| label: fit-multiple-model

# Fit multiple regression model
model_multiple <- lm(length_of_stay ~ age + severity_score + num_comorbidities,
                     data = hospital_data)

# Display summary
summary(model_multiple)
```

### Interpreting Multiple Regression Results

Now let's extract and examine the regression coefficients to understand how each predictor affects length of stay.

```{r}
#| label: interpret-multiple

# Extract coefficients
multiple_coefs <- tidy(model_multiple, conf.int = TRUE) %>%
  mutate(
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ ""
    )
  )

# Display formatted table
multiple_coefs %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "Multiple Regression Coefficients for Length of Stay") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

#### Interpreting Each Coefficient

Each coefficient represents the effect of that variable **while holding all other variables constant**.

```{r}
#| echo: false
coefs <- coef(model_multiple)
```

1. **Intercept (β₀ = `r round(coefs[1], 2)`)**: The predicted length of stay for a patient who is 0 years old, has a severity score of 0, and has 0 comorbidities. This value is not practically meaningful but is mathematically necessary.

2. **Age (β₁ = `r round(coefs[2], 3)`)**: Holding severity and comorbidities constant, each additional year of age is associated with a **`r round(coefs[2], 3)`-day increase** in length of stay.

3. **Severity Score (β₂ = `r round(coefs[3], 3)`)**: Holding age and comorbidities constant, each one-point increase in severity score is associated with a **`r round(coefs[3], 2)`-day increase** in length of stay.

4. **Number of Comorbidities (β₃ = `r round(coefs[4], 3)`)**: Holding age and severity constant, each additional comorbidity is associated with a **`r round(coefs[4], 2)`-day increase** in length of stay.

#### Model Fit Statistics

Let's examine overall model fit using R² and adjusted R² to see how well our predictors explain variation in length of stay.

```{r}
#| label: model-fit-multiple

# Model fit statistics
fit_stats <- glance(model_multiple)

fit_stats %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, df, nobs) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value") %>%
  kable(caption = "Model Fit Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


**R² vs. Adjusted R²**:

- **R² = `r round(fit_stats$r.squared, 3)`**: `r round(fit_stats$r.squared * 100, 1)`% of variance in LOS is explained by the model
- **Adjusted R² = `r round(fit_stats$adj.r.squared, 3)`**: Adjusted for the number of predictors; use this when comparing models with different numbers of variables

### Practical Application: Predicting LOS

Now let's use our fitted model to predict length of stay for different patient profiles. This demonstrates how the model can support clinical decision-making and resource planning. Let's assume that we have several new patients and their characteristics as below.

```{r}
#| label: multiple-new-patients
#| echo: false

# Create example patients
example_patients <- tibble(
  patient_profile = c("Young, Low Severity", "Elderly, Moderate",
                      "Middle-aged, High Severity", "Elderly, Complex"),
  age = c(35, 75, 50, 80),
  severity_score = c(2, 5, 8, 7),
  num_comorbidities = c(0, 2, 3, 5)
)

head(example_patients)

```


Let's get the predicted LOS. We can do this using the `predict()` function in R.

```{r}
#| label: multiple-predictions

# Make predictions
example_patients <- example_patients %>%
  mutate(
    predicted_los = round(predict(model_multiple, newdata = .), 2)
  )

example_patients %>%
  kable(caption = "Predicted Length of Stay by Patient Profile",
        col.names = c("Patient Profile", "Age", "Severity",
                     "Comorbidities", "Predicted LOS (days)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

#### Manual Calculation Example

Let's manually calculate one prediction to understand how multiple regression works. We'll predict LOS for the **"Elderly, Moderate"** patient (Age = 75, Severity = 5, Comorbidities = 2; the second new patient in the table above).

**Step 1: Write the fitted regression equation**

First, recall our fitted model equation with estimated coefficients:

```{r}
#| echo: false
# Extract coefficients for the equation
b0_multi <- round(coef(model_multiple)[1], 2)
b1_multi <- round(coef(model_multiple)[2], 3)
b2_multi <- round(coef(model_multiple)[3], 3)
b3_multi <- round(coef(model_multiple)[4], 3)
```

$$\widehat{\text{LOS}} = `r b0_multi` + `r b1_multi` \times \text{Age} + `r b2_multi` \times \text{Severity} + `r b3_multi` \times \text{Comorbidities}$$

Or in general notation:

$$\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 X_1 + \hat{\beta}_2 X_2 + \hat{\beta}_3 X_3$$

Where:

- $\hat{Y}$ = Predicted length of stay
- $X_1$ = Age (years)
- $X_2$ = Severity score (1-10)
- $X_3$ = Number of comorbidities
- $\hat{\beta}_0$ = `r b0_multi` (Intercept)
- $\hat{\beta}_1$ = `r b1_multi` (Age coefficient)
- $\hat{\beta}_2$ = `r b2_multi` (Severity coefficient)
- $\hat{\beta}_3$ = `r b3_multi` (Comorbidity coefficient)

**Step 2: Substitute the patient's values**

For the "Elderly, Moderate" patient:

- Age = 75 years
- Severity = 5
- Comorbidities = 2

$$\widehat{\text{LOS}} = `r b0_multi` + `r b1_multi` \times 75 + `r b2_multi` \times 5 + `r b3_multi` \times 2$$

**Step 3: Calculate step-by-step**

```{r}
#| label: manual-multiple-prediction
#| echo: false

# Extract coefficients
b0_val <- coef(model_multiple)[1]
b1_val <- coef(model_multiple)[2]
b2_val <- coef(model_multiple)[3]
b3_val <- coef(model_multiple)[4]

# Patient characteristics
age_val <- 75
severity_val <- 5
comorbid_val <- 2

# Manual calculation
pred_los_manual <- b0_val + b1_val * age_val + b2_val * severity_val + b3_val * comorbid_val

# Display step-by-step calculation
cat("Step-by-step calculation:\n\n")
cat("Intercept:              ", round(b0_val, 2), "\n")
cat("Age effect:             ", round(b1_val, 3), "×", age_val, "=", round(b1_val * age_val, 2), "\n")
cat("Severity effect:        ", round(b2_val, 3), "×", severity_val, "=", round(b2_val * severity_val, 2), "\n")
cat("Comorbidity effect:     ", round(b3_val, 3), "×", comorbid_val, "=", round(b3_val * comorbid_val, 2), "\n")
cat("                        ", strrep("-", 40), "\n")
cat("Predicted LOS:          ", round(pred_los_manual, 2), "days\n")
```

**Result**: The predicted length of stay for a 75-year-old patient with severity score of 5 and 2 comorbidities is **`r round(pred_los_manual, 2)` days**.

You can verify this matches the value in the prediction table above for the "Elderly, Moderate" patient!

In multiple regression, each predictor contributes to the final prediction. We simply add up all the effects (intercept + age effect + severity effect + comorbidity effect) to get the total predicted outcome.


### Visualizing Multiple Regression Effects

Let's visualize how each predictor independently affects length of stay while holding other variables at their mean values. This helps us understand the individual contribution of each factor.

```{r}
#| label: fig-multiple-effects
#| fig-cap: "Effect of each predictor on length of stay (holding others at mean)"
#| fig-height: 4
#| fig-width: 10

# Create effect plots
# Age effect
age_effect <- tibble(
  age = seq(20, 90, by = 5),
  severity_score = mean(hospital_data$severity_score),
  num_comorbidities = mean(hospital_data$num_comorbidities)
) %>%
  mutate(predicted = predict(model_multiple, newdata = .))

p1 <- ggplot(age_effect, aes(x = age, y = predicted)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point() +
  labs(x = "Patient Age (years)", y = "Predicted LOS (days)",
       title = "Effect of Age") +
  theme_minimal()

# Severity effect
severity_effect <- tibble(
  age = mean(hospital_data$age),
  severity_score = seq(1, 10, by = 0.5),
  num_comorbidities = mean(hospital_data$num_comorbidities)
) %>%
  mutate(predicted = predict(model_multiple, newdata = .))

p2 <- ggplot(severity_effect, aes(x = severity_score, y = predicted)) +
  geom_line(color = "darkred", size = 1.2) +
  geom_point() +
  labs(x = "Severity Score (1-10)", y = "Predicted LOS (days)",
       title = "Effect of Severity") +
  theme_minimal()

# Comorbidities effect
comorbid_effect <- tibble(
  age = mean(hospital_data$age),
  severity_score = mean(hospital_data$severity_score),
  num_comorbidities = 0:8
) %>%
  mutate(predicted = predict(model_multiple, newdata = .))

p3 <- ggplot(comorbid_effect, aes(x = num_comorbidities, y = predicted)) +
  geom_line(color = "darkgreen", size = 1.2) +
  geom_point(size = 2) +
  labs(x = "Number of Comorbidities", y = "Predicted LOS (days)",
       title = "Effect of Comorbidities") +
  theme_minimal()

# Combine plots
p1 + p2 + p3
```

### Practice Checkpoint 2 {.unnumbered}

::: {.callout-important}
## Try It Yourself! (10 minutes)

Open `week11_linear_regression_practice.qmd` and complete:

- **Activity 2: Multiple Linear Regression Practice**
  - Task 2.1: Fit a multiple regression model
  - Task 2.2: Interpret the coefficients
  - Task 2.3: Compare model fit (R²) with simple regression
  - Task 2.4: Make predictions for new patients

This will help you practice building and interpreting multiple regression models.

**After 10 minutes**, we'll continue with model diagnostics.
:::

------------------------------------------------------------------------

# Model Assumptions and Diagnostics

## Key Assumptions of Linear Regression

Linear regression relies on several assumptions. Understanding these assumptions helps ensure valid statistical inference and reliable predictions.

### 1. Linearity:

- The relationship between predictor(s) and outcome is linear.
- If the true relationship is curved but we fit a straight line, predictions will be systematically biased.
- We can visually check by drawing a scatterplot
- **If violated**: Consider transforming variables (log, square root) or using polynomial terms.

### 2. Independence of Observations

- Each observation's residual is unrelated to others' residuals.
- Correlated errors (e.g., repeated measures, clustered data) underestimate standard errors, making p-values too small.
- Know your data structure (longitudinal? clustered?)
- **If violated**: Use mixed models, clustered standard errors, or time-series methods.

### 3. Homoscedasticity (Equal Variance)

- Variance of residuals is constant across all fitted values.
- Unequal variance (heteroscedasticity) makes standard errors unreliable and confidence intervals inaccurate.
- We can visually inspect by drawing a Residuals vs Fitted plot (vertical spread should be roughly equal across x-axis)
- **If violated**: Use weighted least squares or robust standard errors; consider transforming Y.

### 4. Normality of Residuals

- Residuals follow a normal distribution.
- Critical for valid p-values and confidence intervals
- We can visually inspect using Q-Q plot (points should follow diagonal line)
- **If violated**: Consider transforming Y or using robust methods.

### 5. No (Perfect) Multicollinearity (Multiple Regression Only)

- Predictors are not highly correlated with each other.
- High correlation between predictors inflates standard errors, making coefficients unstable and hard to interpret.
- We can check this visually (with correlation matrix) and check Variance Inflation Factor (VIF; VIF > 10 indicates problematic multicollinearity)
- **If violated**: Remove redundant predictors, combine correlated predictors, or use penalized regression (we will cover this later in the semester).

---

### Summary: Which Plot Checks Which Assumption?

| Assumption | Primary Diagnostic |
|------------|-------------------|
| **Linearity** | Residuals vs Fitted (look for no pattern) |
| **Independence** | Know your study design |
| **Homoscedasticity** | Residuals vs Fitted (equal vertical spread) |
| **Normality** | Q-Q plot |
| **Multicollinearity** | VIF values (not a plot) |


## Diagnostic Plots

We can draw diagnostic plots by using the `plot()` function in R.

```{r}
#| label: fig-diagnostics
#| fig-cap: "Regression diagnostic plots"
#| fig-height: 4
#| fig-width: 8

# Create diagnostic plots
par(mfrow = c(1, 2))  # 1 row, 2 columns
plot(model_multiple, which = 1) # Residual vs. fitted plot
plot(model_multiple, which = 2) # Q-Q plot
par(mfrow = c(1, 1))  # Reset to default
```

### Interpreting the Two Standard Diagnostic Plots

**Plot 1: Residuals vs Fitted**

- **What to look for**: Random scatter around horizontal line at 0
- **What it checks**: Linearity and homoscedasticity
- **Good sign**: No pattern, equal spread across x-axis
- **Red flags**:
  - Curved pattern = non-linearity
  - Funnel shape = heteroscedasticity

**Plot 2: Normal Q-Q (Quantile-Quantile)**

- A Q-Q plot compares the distribution of your residuals to a theoretical normal distribution.
- **What to look for**: Points following the diagonal reference line
- **What it checks**: Normality of residuals
- Points close around the diagonal indicate residuals are normally distributed


## Checking for Multicollinearity

Multicollinearity occurs when predictors are highly correlated, which can:

- Inflate standard errors
- Make individual coefficients unstable
- Lead to counterintuitive results

### Variance Inflation Factor (VIF)

VIF quantifies how much multicollinearity inflates the variance (and standard errors) of regression coefficients.

**How VIF works**:

For each predictor $X_1$:

1. Regress $X_1$ on all other predictors
2. Get $R^2$ from that regression (how well other predictors explain $X_1$)
3. Calculate: $\text{VIF} = \frac{1}{1 - R^2}$

- **Rule of Thumb**: VIF > 10 (or Tolerance < 0.1) indicates problematic multicollinearity.

We can get VIF values by using the `vif()` function in R.

```{r}
#| label: vif-check

# Calculate VIF
vif_values <- car::vif(model_multiple)

tibble(
  Variable = names(vif_values),
  VIF = round(vif_values, 2),
  Tolerance = round(1/vif_values, 3),
  Interpretation = case_when(
    vif_values < 5 ~ "Acceptable",
    vif_values < 10 ~ "Moderate concern",
    TRUE ~ "High - consider removing"
  )
) %>%
  kable(caption = "Multicollinearity Diagnostics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Demonstrating the Impact of Multicollinearity

Let's see what happens when we include two highly correlated predictors in the same model.

**Scenario**: We want to understand what predicts longer hospital length of stay. We have two predictors:

- **Number of comorbidities** (chronic conditions like diabetes, COPD, etc.)
- **Number of medications** (prescribed at admission)

These variables are highly correlated (r ≈ 0.90) because patients with more conditions naturally take more medications.

```{r}
#| label: multicollinearity-demo
#| code-fold: false

# Generate demonstration data
set.seed(2026)
n <- 200

demo_data <- tibble(
  num_comorbidities = sample(0:5, n, replace = TRUE, prob = c(0.15, 0.25, 0.25, 0.20, 0.10, 0.05)),
  num_medications = round(2.5 * num_comorbidities + rnorm(n, mean = 1, sd = 1.2)),
  los = round(3 + 2.0 * num_comorbidities - 0.4 * num_medications + rnorm(n, 0, 1.5), 1)
) %>%
  mutate(
    num_medications = pmax(0, num_medications),  # No negative medications
    los = pmax(1, los)  # Minimum 1 day stay
  )

# Check correlation
correlation <- cor(demo_data$num_comorbidities, demo_data$num_medications)
cat("Correlation between comorbidities and medications:", round(correlation, 3), "\n")
```

**Now let's fit three models and compare**:

```{r}
#| label: multicollinearity-comparison

# Model 1: Comorbidities alone
model_1 <- lm(los ~ num_comorbidities, data = demo_data)

# Model 2: Medications alone
model_2 <- lm(los ~ num_medications, data = demo_data)

# Model 3: Both together (multicollinearity!)
model_3 <- lm(los ~ num_comorbidities + num_medications, data = demo_data)

# Compare coefficients and significance
comparison <- tibble(
  Model = c("Model 1: Comorbidities only",
            "Model 2: Medications only",
            "Model 3: Both: Comorbidities",
            "Model 3: Both: Medications"),
  Coefficient = c(
    coef(model_1)[2],
    coef(model_2)[2],
    coef(model_3)[2],
    coef(model_3)[3]
  ),
  SE = c(
    summary(model_1)$coefficients[2, 2],
    summary(model_2)$coefficients[2, 2],
    summary(model_3)$coefficients[2, 2],
    summary(model_3)$coefficients[3, 2]
  ),
  `p-value` = c(
    summary(model_1)$coefficients[2, 4],
    summary(model_2)$coefficients[2, 4],
    summary(model_3)$coefficients[2, 4],
    summary(model_3)$coefficients[3, 4]
  ),
  VIF = c(
    NA,  # Not applicable for single predictor
    NA,  # Not applicable for single predictor
    car::vif(model_3)[1],
    car::vif(model_3)[2]
  )
) %>%
  mutate(
    Coefficient = round(Coefficient, 3),
    SE = round(SE, 3),
    `p-value` = ifelse(`p-value` < 0.001, "<.001", round(`p-value`, 3)),
    VIF = round(VIF, 2),
    Significance = case_when(
      `p-value` == "<.001" ~ "***",
      `p-value` < 0.01 ~ "**",
      `p-value` < 0.05 ~ "*",
      TRUE ~ ""
    )
  )

comparison %>%
  kable(caption = "Impact of Multicollinearity on Coefficient Estimates") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  footnote(general = "Note: *** p < .001, ** p < .01, * p < .05")
```

**What happened?**

1. **Model 1** (comorbidities alone):
   - Significant positive effect (each additional comorbidity increases LOS)

2. **Model 2** (medications alone):
   - Significant positive effect (each additional medication increases LOS)

3. **Model 3** (both together): The truth is revealed!
   - **Comorbidities**: Remains positive and significant
   - **Medications**: Coefficient dramatically changes (from +`r round(coef(model_2)[2], 2)` to `r round(coef(model_3)[3], 2)`)!
   - Standard errors inflate dramatically (VIF ≈ `r round(car::vif(model_3)[1], 1)`)

**Why the dramatic change?**

- When medications is alone in the model, it picks up the effect of comorbidities (since they're correlated). It looks important!

- But when both are included, the model separates their effects. Now we see that **after controlling for** comorbidities, medications shows its true relationship (which may be weak, non-significant, or even negative)

- Multicollinearity doesn't just inflate standard errors. It can completely mislead you about which predictors matter and even the direction of their effects!

**What to do**:

- Choose the most theoretically important predictor (usually **comorbidities**)
- Or create a composite measure (e.g., "patient complexity index")
- Check correlation matrix before modeling to identify problems early

### Practice Checkpoint 3 {.unnumbered}

::: {.callout-important}
## Try It Yourself! (10 minutes)

Open `week11_linear_regression_practice.qmd` and complete:

- **Activity 3: Model Diagnostics Practice**
  - Task 3.1: Create and interpret diagnostic plots
  - Task 3.2: Check for multicollinearity using VIF
  - Task 3.3: Assess model assumptions

This will help you practice evaluating regression model assumptions.

**After 10 minutes**, we'll wrap up with a summary.
:::

------------------------------------------------------------------------

# Summary

## Key Takeaways

### Simple vs. Multiple Linear Regression

- **Simple regression**: Examines how one factor relates to an outcome ("If X increases by 1, how much does Y change?")
- **Multiple regression**: Examines how several factors together relate to an outcome, showing each factor's unique contribution while holding others constant

### How OLS Finds the Best Line

- Ordinary Least Squares (OLS) finds the line that minimizes the sum of squared residuals
- "Residuals" = vertical distance between actual data points and the fitted line

### Model Fit: R²

- R² tells you what percentage of variation your model explains
- Adjusted R² is better for multiple regression (accounts for number of predictors)
- Higher is better, but context matters

### Making Predictions

- Use fitted models to predict outcomes for new situations
- Useful for "what-if" scenarios and planning

### Interpreting Coefficients

Ask three questions for each coefficient:

1. **Direction**: Positive or negative relationship?
2. **Magnitude**: How large is the effect practically?
3. **Significance**: Is p < 0.05? (But practical significance matters too!)

**Key concept**: In multiple regression, coefficients show effects **while controlling for other variables**.

### Multicollinearity

- **Problem**: When predictors are highly correlated, coefficients become unstable and can even flip signs
- **Check**: VIF > 10 indicates problems
- **Solution**: Remove or combine correlated predictors

## Next Steps: Practice and Comprehensive Exercises

::: {.callout-note icon=false}
## Practice Activity (Completion-Based)

Complete the hands-on practice in **`week11_linear_regression_practice.qmd`**:

### Activities 1-3: Guided Practice (with fill-in-the-blank)

- Activity 1: Simple Linear Regression Practice
- Activity 2: Multiple Linear Regression Practice
- Activity 3: Model Diagnostics Practice

**Format**: Fill-in-the-blank code templates with hints and solutions

**Skills practiced**: All regression techniques from lecture

### Exercises 1-3: Comprehensive Analysis (open-ended)

- Exercise 1: Simple regression with hospital readmission data
- Exercise 2: Multiple regression controlling for confounders
- Exercise 3: Making predictions and scenario analysis

**Format**: Independent work with provided solutions

**Expected time**: 2-3 hours (in-class practice + homework)

**Purpose**: Build confidence and fluency with linear regression analysis
:::

------------------------------------------------------------------------

# References and Further Reading

## Statistical Methods

- James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning with Applications in R* (2nd ed.). Springer.
- Fox, J. (2015). *Applied Regression Analysis and Generalized Linear Models* (3rd ed.). SAGE Publications.
- Gelman, A., & Hill, J. (2006). *Data Analysis Using Regression and Multilevel/Hierarchical Models*. Cambridge University Press.

## Healthcare Applications

Examples are motivated by the following:

- Alreshidi, N. M., Alaseeri, R. M., Alzoraigi, U. S., et al. (2021). Patient satisfaction with the emergency department services at an academic teaching hospital. *BMC Health Services Research*, 21, Article 423. https://pmc.ncbi.nlm.nih.gov/articles/PMC8144764/

- Stone, K., Zwiggelaar, R., Jones, P., & Mac Parthaláin, N. (2023). A systematic review of the prediction of hospital length of stay: Towards a unified framework. *PLOS Digital Health*, 2(2), e0000017. https://pmc.ncbi.nlm.nih.gov/articles/PMC9931263/

- McHugh, M. D., Tubbs-Cooley, H. L., Ghazarian, S. R., Hegarty, S. E., et al. (2021). Effects of nurse-to-patient ratio legislation on nurse staffing and patient mortality, readmissions, and length of stay: a prospective study in a panel of hospitals. *The Lancet*, 397(10288), 1905-1913. https://pmc.ncbi.nlm.nih.gov/articles/PMC8408834/
