---
title: "Linear Regression 1: Simple and Multiple Linear Regression"
subtitle: "MBA642: Business Analytics in Healthcare Management"
Week: "Week 11"
date: 03/23/2026
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: false
    theme: [lecture.scss, cosmo]
    highlight-style: github
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Load required packages
library(tidyverse)
library(car)
library(GGally)
library(patchwork)

# Set seed for reproducibility
set.seed(2026)
```

# Introduction

## From Correlation to Regression

In Week 8, we used correlation to answer whether there is a relationship between nurse staffing hours and patient satisfaction (e.g., r = 0.85).

**Linear regression extends this** by answering more detailed questions:

-   *How many points does satisfaction increase for each additional nursing hour?* (quantifying the relationship)
-   *If we increase staffing to 8 hours per patient day, what satisfaction score can we expect?* (making predictions)
-   *Among wait time, staffing, and facility age, which factors significantly explain patient satisfaction?* (identifying key drivers)
-   *Does staffing still matter after controlling for facility age and patient severity?* (controlling for confounding variables)

## Why Linear Regression Matters

Healthcare managers frequently encounter situations where understanding relationships between variables is crucial for decision-making:

-   **Resource allocation**: How does nurse staffing level affect patient outcomes?
-   **Cost management**: What factors drive hospital readmission costs?
-   **Quality improvement**: How does patient wait time impact satisfaction scores?
-   **Performance prediction**: Can we predict length of stay based on patient characteristics?

Linear regression provides a powerful framework to:

1.  **Quantify relationships** between variables (e.g., "For every additional nurse per patient, satisfaction increases by X points")
2.  **Test hypotheses** about whether relationships are statistically significant
3.  **Make predictions** about future outcomes
4.  **Identify key drivers** (among multiple factors, determine which significantly explain the outcome)
5.  **Control for confounding variables** when examining effects of interest

## Overview of Linear Regression Types

| Type | Predictors | Example |
|:---|:--:|:---|
| Simple Linear Regression | 1 | Patient age → Length of stay |
| Multiple Linear Regression | 2+ | Patient age + Severity + Insurance → Length of stay |

------------------------------------------------------------------------

# Simple Linear Regression

## What is it?

Simple linear regression models the relationship between **one independent variable (X)** and **one dependent variable (Y)** using a linear model (straight line). It is the foundation for understanding more complex regression models, and allows us to quantify how changes in one variable are associated with changes in another.

## When to Use It

You can use simple linear regression when you want to:

-   Examine the (linear) relationship between two continuous variables
-   Predict an outcome based on a single predictor
-   Quantify the strength and direction of a linear relationship
-   Test whether a relationship is statistically significant

## The Concept

Simple linear regression finds the "best-fitting" line (linear model) through the data points using a method called Ordinary Least Squares (OLS).

### The Model Equation

The simple linear regression model is expressed as:

$$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$$

Where:

-   $Y_i$ (Dependent Variable; DV): The outcome we want to predict or explain. *(e.g., patient satisfaction score)*
-   $X_i$ (Independent Variable; IV): The predictor we use to explain the outcome. *(e.g., wait time in minutes)*
-   $\beta_0$ (Intercept): The predicted value of Y when X = 0. *(Think of it as the starting point of the regression line)*
-   $\beta_1$ (Regression Coefficient; Slope): The change in Y for a one-unit increase in X. *(i.e., For each one-unit increase in X, how much does Y change)*
-   $\epsilon_i$ (Residual): The difference between an observed value and the model's prediction *(i.e., the "leftover" the model does not explain)*

The following diagram shows how all components of the regression equation relate to each other using a simple example with three data points:

```{r}
#| label: fig-regression-components
#| fig-cap: "All components of the regression equation shown together"
#| fig-width: 7
#| fig-height: 5
#| echo: false

# Manually set a clean line for conceptual illustration
demo_b0 <- 3      # intercept
demo_b1 <- 0.5    # slope: Y increases by 0.5 for each 1-unit increase in X

# 3 data points positioned around the line
demo_pts <- tibble(
  x = c(2, 5, 8),
  y = c(3.6, 6.5, 6.0),                          # observed
  predicted = demo_b0 + demo_b1 * c(2, 5, 8)     # on the line: 4, 5.5, 7
)

# Build labels
intercept_label <- paste0("bold('Intercept')~(beta[0])==", demo_b0)
slope_label <- paste0("bold('Slope')~(beta[1])==", demo_b1)

ggplot(demo_pts, aes(x = x, y = y)) +
  # Explicit axis lines through origin
  geom_hline(yintercept = 0, linewidth = 0.6, color = "gray30") +
  geom_vline(xintercept = 0, linewidth = 0.6, color = "gray30") +
  # Regression line
  geom_abline(intercept = demo_b0, slope = demo_b1,
              color = "blue", linewidth = 1.2) +
  # Residual segments
  geom_segment(aes(xend = x, yend = predicted),
               color = "red", linewidth = 0.9, linetype = "dashed") +
  # Observed points
  geom_point(size = 4, color = "black") +
  # Predicted points on the line
  geom_point(aes(y = predicted), size = 3, color = "blue", shape = 17) +
  # Intercept point
  geom_point(aes(x = 0, y = demo_b0), size = 4, color = "darkgreen", shape = 15) +
  # --- Intercept label (above intercept, to the right) ---
  annotate("text", x = 0.3, y = demo_b0,
           label = intercept_label, parse = TRUE,
           color = "darkgreen", size = 3.5, hjust = 0) +
  # --- Legend area (top-right corner, away from all data) ---
  annotate("point", x = 2.0, y = 9, size = 3, color = "black") +
  annotate("text", x = 2.3, y = 9, parse = TRUE,
           label = "'Observed'~(Y[i])",
           color = "black", size = 3.5, hjust = 0) +
  annotate("point", x = 2.0, y = 8.3, size = 2.5, color = "blue", shape = 17) +
  annotate("text", x = 2.3, y = 8.3, parse = TRUE,
           label = "'Predicted'~(hat(Y)[i])",
           color = "blue", size = 3.5, hjust = 0) +
  # --- Residual label (right of 3rd point's residual) ---
  annotate("text", x = 8.5,
           y = mean(c(demo_pts$y[3], demo_pts$predicted[3])),
           label = "'Residual'~(epsilon[i])", parse = TRUE,
           color = "red", size = 3.5, hjust = 0) +
  # --- Slope bracket (between x=3 and x=4) ---
  annotate("segment", x = 3, xend = 4,
           y = demo_b0 + demo_b1 * 3, yend = demo_b0 + demo_b1 * 3,
           color = "purple", linewidth = 0.7) +
  annotate("segment", x = 4, xend = 4,
           y = demo_b0 + demo_b1 * 3, yend = demo_b0 + demo_b1 * 4,
           color = "purple", linewidth = 0.7) +
  annotate("text", x = 4.1, y = demo_b0 + demo_b1 * 3.5,
           label = slope_label, parse = TRUE,
           color = "purple", size = 3.5, hjust = 0) +
  annotate("text", x = 3.5, y = demo_b0 + demo_b1 * 3 - 0.4,
           label = "1 unit", color = "purple", size = 3, hjust = 0.5) +
  # X label near first point
  scale_x_continuous(limits = c(-0.5, 10.5), breaks = 0:10, expand = c(0, 0)) +
  scale_y_continuous(limits = c(-0.5, 9.8), breaks = 0:9, expand = c(0, 0)) +
  labs(x = "Independent Variable (X)",
       y = "Dependent Variable (Y)",
       title = expression("Anatomy of a Regression Equation: " ~
                           Y == beta[0] + beta[1] * X + epsilon)) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

## The Ordinary Least Squares (OLS) Method

### How OLS Works

The shape of a linear model (with one predictor) is determined by two components: the intercept and the slope.

The question we have now is: how do we define the "best" linear model given data? If we find a linear model that has the smallest *squared* differences between our predictions and the actual outcome values, we can say that the model is the best linear model.

Linear regression finds this "best" line by minimizing the **residual sum of squares (RSS)**, also called the sum of squared residuals:

$$RSS = \sum_{i=1}^{n} (Y_i - \hat{Y}_i)^2 = \sum_{i=1}^{n} \epsilon_i^2$$

Where $Y_i$ is the observed value and $\hat{Y}_i$ is the predicted value for observation $i$.

Why squared residuals?

1.  Squaring ensures positive and negative errors don't cancel out
2.  Squaring penalizes larger errors more heavily
3.  Leads to closed-form mathematical solutions

The figure below shows three different lines fitted to the same data. Each line has a different intercept and slope, producing different residuals (red dashed lines) and therefore different RSS values.

Notice that the OLS line (right panel) has the **smallest RSS**. No other intercept-slope pair can produce a smaller sum of squared residuals.

```{r}
#| label: fig-ols-comparison
#| echo: false
#| fig-width: 9
#| fig-height: 4

# Same data for all three panels
ols_demo <- tibble(
  x = c(1, 2, 3, 4, 5, 6, 7),
  y = c(1.8, 2.5, 3.8, 3.7, 5.1, 5.8, 6.2)
)

# OLS fit (the best line)
ols_fit <- lm(y ~ x, data = ols_demo)
ols_b0 <- coef(ols_fit)[1]
ols_b1 <- coef(ols_fit)[2]
ols_rss <- sum(residuals(ols_fit)^2)

# Three candidate lines: too flat, too steep, OLS best
candidates <- list(
  list(b0 = 2.5, b1 = 0.3, label = "Too Flat"),
  list(b0 = -0.5, b1 = 1.1, label = "Too Steep"),
  list(b0 = ols_b0, b1 = ols_b1, label = "OLS Best Fit")
)

plots <- map(candidates, function(cand) {
  pred_y <- cand$b0 + cand$b1 * ols_demo$x
  rss_val <- sum((ols_demo$y - pred_y)^2)

  ols_demo_aug <- ols_demo |>
    mutate(predicted = pred_y)

  ggplot(ols_demo_aug, aes(x = x, y = y)) +
    # Residual segments
    geom_segment(aes(xend = x, yend = predicted),
                 color = "darkred", linewidth = 0.7, linetype = "dashed") +
    # Candidate line
    geom_abline(intercept = cand$b0, slope = cand$b1,
                color = "steelblue", linewidth = 1.2) +
    # Observed points
    geom_point(size = 3, color = "black") +
    # RSS annotation
    annotate("label", x = 1, y = 7, size = 3.5, hjust = 0, fontface = "bold",
             label = paste0("RSS = ", round(rss_val, 2)),
             fill = if (cand$label == "OLS Best Fit") "#e8f5e9" else "#ffebee",
             color = if (cand$label == "OLS Best Fit") "#2e7d32" else "#c62828",
             label.size = 0) +
    # Equation annotation
    annotate("text", x = 1, y = 6.3, size = 3, hjust = 0, color = "steelblue",
             label = paste0("Y = ", round(cand$b0, 1), " + ", round(cand$b1, 2), " X")) +
    labs(title = cand$label, x = "X", y = "Y") +
    coord_cartesian(xlim = c(0, 8), ylim = c(0, 8)) +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12))
})

plots[[1]] + plots[[2]] + plots[[3]] +
  plot_layout(ncol = 3)
```

## Using the Fitted Model for Predictions

Once OLS finds the best-fitting line by estimating $\beta_0$ (intercept) and $\beta_1$ (slope), we can use the fitted model to make predictions for new observations.

**The prediction equation** is:

$$\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 X$$

Where:

-   $\hat{Y}$ = Predicted value of the dependent variable
-   $\hat{\beta}_0$ = Estimated intercept from the fitted model
-   $\hat{\beta}_1$ = Estimated slope from the fitted model
-   $X$ = Value of the independent variable for which we want to make a prediction

The "hat" notation (\^) indicates these are **estimated** values from our sample data, not the true population parameters. We use the estimated coefficients to predict outcomes for:

1.  **Observations within our dataset** to assess model fit
2.  **New observations** to forecast outcomes based on predictor values
3.  **Hypothetical scenarios** to evaluate "what if" questions (e.g., "What if we reduce wait time to 30 minutes?")

Let's see this in action with a real healthcare example.

------------------------------------------------------------------------

## Example Case: Patient Satisfaction and Wait Time

### Background and Context

Research has consistently shown that emergency department wait times are negatively associated with patient satisfaction. A 2021 study (Abass et al.) found that being informed about waiting time showed a moderately strong positive correlation with patients' overall rating, highlighting the importance of wait time management for patient experience.

Let's examine whether emergency department wait time affects patient satisfaction scores using synthetic data that mirrors these real-world relationships.

### Loading the Dataset

Run the code below to load the dataset and ruquired packages:

```{r}
#| label: create-simple-data

# Load required packages
library(tidyverse)
library(car)
library(GGally)
set.seed(2026)

# Load synthetic ED data
ed_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week11_ed_data.csv")

# Preview the data
head(ed_data, 10)
```

### Exploratory Data Analysis

Before running regression, always visualize your data. Let's take a look at the pattern using a scatterplot.

```{r}
#| label: fig-eda-simple
#| fig-cap: "Relationship between ED wait time and patient satisfaction"
#| echo: false
#| fig-height: 4
#| fig-width: 6

# Scatter plot with trend line
ggplot(ed_data, aes(x = wait_time_minutes, y = satisfaction_score)) +
  geom_point(alpha = 0.6, size = 2, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", fill = "pink") +
  labs(
    x = "Wait Time (minutes)",
    y = "Patient Satisfaction Score (0-100)",
    title = "ED Wait Time vs. Patient Satisfaction"
  ) +
  theme_minimal(base_size = 12)
```

### Fitting the Simple Linear Regression Model

Let's fit the simple linear regression model to quantify the relationship. We can fit a linear model using the `lm()` function in R.

```{r}
#| label: fit-simple-model

# Fit the model
model_simple <- lm(satisfaction_score ~ wait_time_minutes, data = ed_data)

# Display model summary
summary(model_simple)
```

### The Fitted Regression Equation

Based on the model output above, we can write our fitted regression equation with the estimated coefficients:

```{r}
#| echo: false
# Extract coefficients for the equation
b0 <- round(coef(model_simple)[1], 2)
b1 <- round(coef(model_simple)[2], 3)
```

$$\widehat{\text{Satisfaction}} = `r b0` + (`r b1`) \times \text{Wait Time}$$

Or more generally:

$$\hat{Y}_i = `r b0` + (`r b1`) \times X_i$$

Where:

-   $\hat{Y}_i$ = Predicted patient satisfaction score for patient $i$
-   $X_i$ = Wait time in minutes for patient $i$
-   `r b0` = Estimated intercept ($\hat{\beta}_0$)
-   `r b1` = Estimated slope ($\hat{\beta}_1$)

### Interpreting the Results

Let's break down the output systematically:

#### 1. The Intercept

```{r}
#| echo: false
intercept <- round(coef(model_simple)[1], 2)
```

**Value**: β₀ = `r intercept`

**Interpretation**: When wait time is 0 minutes, the predicted patient satisfaction score is `r intercept` points. This represents the baseline satisfaction when there is no wait.

*Note*: Intercepts should be interpreted cautiously. If zero falls outside the range of observed X values (e.g., if no patient had a 0-minute wait), the intercept is an extrapolation and may not represent a realistic scenario. Here, a 0-minute wait is plausible, so the interpretation is reasonable.

#### 2. The Slope (Regression Coefficient)

```{r}
#| echo: false
slope <- round(coef(model_simple)[2], 3)
p_value <- summary(model_simple)$coefficients[2, 4]
```

**Value**: β₁ = `r slope`

**Interpretation**:

1.  **Direction**: The coefficient is **negative** (`r slope`), indicating that longer wait times are associated with **lower** satisfaction scores.

2.  **Magnitude**: For every **one-minute increase** in wait time, patient satisfaction decreases by **`r abs(slope)` points** (on the 0-100 scale).

3.  **Statistical Significance**: In regression, the null and alternative hypotheses for each coefficient are:

    -   $H_0$: $\beta_1 = 0$ (no linear relationship between X and Y)
    -   $H_1$: $\beta_1 \neq 0$ (a linear relationship between X and Y)

    The p-value is `r ifelse(p_value < 2.2e-16, "< 2.2e-16", format(p_value, scientific = TRUE))`, which is less than 0.05. We **reject the null hypothesis** (there is a linear relationship between X and Y).

#### 3. Model Fit (R-squared)

```{r}
#| echo: false
r_squared <- round(summary(model_simple)$r.squared, 3)
```

**R² = `r r_squared`**

R² indicates the amount of the variance in the outcome variable explained by the model. In this case, wait time explains **`r r_squared * 100`%** of the variance in patient satisfaction scores. The remaining `r (1 - r_squared) * 100`% is attributable to other factors not included in this model (e.g., staff friendliness, treatment effectiveness, facility cleanliness).

To understand R² visually, let's see how the regression model explains more variance compared to simply using the mean:

```{r}
#| label: fig-rsquared-concept
#| echo: false
#| fig-cap: "R² measures how much variance is explained by the regression model"
#| fig-height: 6
#| fig-width: 10

# Calculate variance components
ed_data_variance <- ed_data %>%
  mutate(
    predicted = predict(model_simple),
    mean_y = mean(satisfaction_score),
    # Total variance: deviation from mean
    total_dev = satisfaction_score - mean_y,
    # Explained variance: deviation of prediction from mean
    explained_dev = predicted - mean_y,
    # Residual variance: deviation from prediction
    residual_dev = satisfaction_score - predicted
  )

# Calculate sums of squares
TSS <- sum(ed_data_variance$total_dev^2)
ESS <- sum(ed_data_variance$explained_dev^2)
RSS <- sum(ed_data_variance$residual_dev^2)
R2 <- ESS / TSS

# Create two comparison plots
p1 <- ggplot(ed_data_variance, aes(x = wait_time_minutes, y = satisfaction_score)) +
  geom_hline(yintercept = mean(ed_data_variance$satisfaction_score),
             color = "darkgreen", linewidth = 1.2, linetype = "dashed") +
  geom_segment(aes(xend = wait_time_minutes, yend = mean_y),
               color = "purple", alpha = 0.4, linewidth = 0.6) +
  geom_point(size = 2, color = "black", alpha = 0.6) +
  annotate("text", x = 100, y = mean(ed_data_variance$satisfaction_score) + 3,
           label = "Mean (baseline model)", color = "darkgreen", size = 3, hjust = 1) +
  annotate("text", x = 25, y = 95,
           label = paste0("Total Variance (TSS)\nAll deviations from mean\nTSS = ",
                         round(TSS, 0)),
           color = "purple", size = 3, hjust = 0) +
  labs(
    x = "Wait Time (minutes)",
    y = "Patient Satisfaction Score",
    title = "Baseline Model: Using Only the Mean",
    subtitle = "Purple lines show total variance (what we're trying to explain)"
  ) +
  theme_minimal(base_size = 10)

p2 <- ggplot(ed_data_variance, aes(x = wait_time_minutes, y = satisfaction_score)) +
  geom_hline(yintercept = mean(ed_data_variance$satisfaction_score),
             color = "darkgreen", linewidth = 0.8, linetype = "dotted", alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 1.2) +
  geom_segment(aes(xend = wait_time_minutes, yend = predicted),
               color = "red", alpha = 0.5, linewidth = 0.6, linetype = "dashed") +
  geom_point(size = 2, color = "black", alpha = 0.6) +
  annotate("text", x = 100, y = 45,
           label = "Regression line", color = "blue", size = 3, hjust = 1) +
  annotate("text", x = 25, y = 95,
           label = paste0("Residual Variance (RSS)\nUnexplained by model\nRSS = ",
                         round(RSS, 0)),
           color = "red", size = 3, hjust = 0) +
  annotate("text", x = 20, y = 82,
           label = paste0("R² = ", round(R2, 3), " = ", round(R2*100, 1), "%\n",
                         "of variance explained"),
           color = "blue", size = 3, hjust = 0, fontface = "bold") +
  labs(
    x = "Wait Time (minutes)",
    y = "Patient Satisfaction Score",
    title = "Regression Model: Using Wait Time to Predict Satisfaction",
    subtitle = "Red lines show remaining (unexplained) variance after using the model"
  ) +
  theme_minimal(base_size = 10)

# Combine plots
p1 + p2
```

-   **Left plot**: Without any predictors, we'd predict every patient's satisfaction as the mean (dark green line). The purple lines show how far each observation deviates from this mean. This is the **total variance (TSS)** we want to explain.

-   **Right plot**: With wait time as a predictor, the regression line (blue) does a better job. The red dashed lines show the remaining errors (residuals). This is the **residual variance (RSS)** left unexplained.

-   **R² = 1 - (RSS/TSS)** = The proportion of total variance that our model explains. An R² of `r round(R2, 3)` means the regression model explains `r round(R2*100, 1)`% of the variance in patient satisfaction compared to just using the mean.

### Making Predictions

One powerful application of regression is predicting outcomes for new observations. Let's assume that we have new patients and their waiting times.

```{r}
#| label: simple-new-data-points

# Predict satisfaction for specific wait times 
# New data need to be data frame or a list
new_patients <- data.frame(
  wait_time_minutes = c(15, 30, 45, 60, 90)
)
head(new_patients)

```

By using the `predict()` function we can get predicted satisfaction scores for these new patients based on their wait times.

```{r}
#| label: predictions

# Get predicted satisfaction scores
new_patients$predicted_score <- predict(model_simple, newdata = new_patients)
new_patients
```

#### Manual Calculation Example

Let's manually calculate one prediction to understand what the `predict()` function is doing. We'll predict satisfaction for a patient with a **45-minute wait time**.

**Step 1: Recall the fitted equation**

$$\widehat{\text{Satisfaction}} = `r b0` + (`r b1`) \times \text{Wait Time}$$

**Step 2: Substitute and calculate**

```{r}
#| label: manual-prediction-calculation
#| echo: false

# Extract coefficients
intercept_val <- coef(model_simple)[1]
slope_val <- coef(model_simple)[2]

# Manual calculation
wait_time_example <- 45
predicted_satisfaction_manual <- intercept_val + slope_val * wait_time_example
```

For Wait Time = 45 minutes:

$$\widehat{\text{Satisfaction}} = `r b0` + (`r b1`) \times 45 = `r round(predicted_satisfaction_manual, 2)`$$

The predicted satisfaction score for a patient with a 45-minute wait is **`r round(predicted_satisfaction_manual, 2)`** points. You can verify this matches the value in the prediction table above!

#### Practical Application

If your ED currently has an average wait time of 60 minutes and you implement process improvements to reduce it to 30 minutes, the model predicts satisfaction would increase by approximately `r abs(round(30 * slope, 2))` points. Since each additional minute of wait time decreases satisfaction by `r abs(round(slope, 3))` points, a 30-minute reduction corresponds to a meaningful improvement in patient experience.

### Practice 1: Simple Linear Regression {.unnumbered}

::: callout-note
## Your Turn!

Practice fitting and interpreting simple linear regression models on your own using a different healthcare scenario.
:::

### Scenario

You are analyzing the relationship between **nurse staffing levels** and **patient mortality rates** across 100 hospitals. Hospital administrators want to know **whether having more nurses per patient reduces mortality rates**.

This question has important implications for staffing policies and patient safety. If the data show that higher nurse-to-patient ratios are associated with lower mortality, hospital administrators could use the regression results to justify hiring additional nursing staff, estimate how much mortality might decrease with a specific staffing increase, and make evidence-based budget decisions that balance costs against patient outcomes.

### Data Loading

Run this code to load the dataset:

```{r}
#| label: generate-simple-practice-data

# Loading dataset

simple_practice_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week11_practice_data1.csv")

# Preview data
head(simple_practice_data, 5) 
```

**Dataset Variables:**

| Variable | Description | Type |
|----|----|----|
| `hospital_id` | Unique hospital identifier | Numeric |
| `nurses_per_patient` | Ratio of nurses to patients (higher = better staffing) | Numeric (continuous) |
| `mortality_rate` | Hospital mortality rate (deaths per 100 patients) | Numeric (continuous) |

### Task 1.1: Visualize the Relationship

Before fitting a model, create a scatterplot to visualize the relationship between nurse staffing and mortality rates.

```{r}
#| label: task-1-1
#| eval: false

# Fill in the blanks
ggplot(data = ___, aes(x = ___, y = ___)) +
  geom_point(alpha = 0.6, color = "steelblue", size = 2) +
  geom_smooth(method = "___", se = TRUE, color = "darkred") +
  labs(
    title = "Nurse Staffing vs. Mortality Rate",
    x = "Nurses per Patient",
    y = "Mortality Rate (%)"
  ) +
  theme_minimal()
```

**Question**: Based on the scatterplot, do you expect a positive or negative relationship? Why?

### Task 1.2: Fit the Simple Linear Regression Model

Now fit a simple linear regression model predicting `mortality_rate` from `nurses_per_patient`.

```{r}
#| label: task-1-2
#| eval: false

# Fill in the blanks
model_simple_practice <- lm(___ ~ ___, data = ___)

# Display the summary
summary(___)
```

**Questions:**

1.  What is the **intercept**? What does it represent in this context?
2.  What is the **slope** coefficient for `nurses_per_patient`?
    -   **Direction**: Is it positive or negative? What does this mean?
    -   **Magnitude**: If a hospital increases staffing by 1 nurse per patient, how much does mortality rate change?
    -   **Significance**: Is the p-value less than 0.05?
3.  What is the **R²** value? What percentage of variance in mortality rates is explained by nurse staffing?

### Task 1.3: Make Predictions

Use your fitted model to predict mortality rates for hospitals with different staffing levels.

```{r}
#| label: task-1-3
#| eval: false

# Create new data for prediction (fill in the blank)
new_hospitals <- data.frame(
  nurses_per_patient = c(1.0, 1.5, 2.0, 2.5, 3.0)
)

# Make predictions (fill in the blank)
new_hospitals$predicted_mortality = predict(___, newdata = ___)
new_hospitals

```

**Question**: A hospital currently has 1.5 nurses per patient and wants to reduce mortality. If they increase staffing to 2.5 nurses per patient, how much would you expect mortality to decrease?

------------------------------------------------------------------------

# Multiple Linear Regression

Now that we've seen how to model the relationship between a single predictor and an outcome, the next question we would have is *What happens when multiple factors influence the outcome simultaneously?* In healthcare, outcomes are rarely driven by a single variable. Patient satisfaction depends on more than just wait time, and length of stay depends on more than just age. Multiple linear regression addresses this reality.

## What is it?

Multiple linear regression extends simple regression to model the relationship between **multiple independent variables** and **one dependent variable**. It allows us to examine how several predictors ***simultaneously*** influence an outcome.

## When to Use It

Use multiple linear regression when you need to:

-   Include **multiple predictors** to better explain variance in an outcome
-   **Control for confounding variables** when examining specific relationships
-   Compare **the relative importance** of different factors
-   Make predictions based on multiple characteristics simultaneously

## Extending to Multiple Predictors

### The Model Equation

$$Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_k X_{ki} + \epsilon_i$$

Where:

-   $\beta_0$ = Intercept
-   $\beta_1, \beta_2, \ldots, \beta_k$ = Regression coefficients for each predictor
-   $X_{1i}, X_{2i}, \ldots, X_{ki}$ = Values of predictors for observation $i$
-   $\epsilon_i$ = Residual

In multiple regression, each coefficient represents the effect of that variable **holding all other variables constant** (also called the "partial" or "conditional" effect).

## An Example Case: Hospital Length of Stay

### Background and Context

A 2022 systematic review identified age, comorbidities, and severity as the most frequently cited predictors of hospital length of stay across 15 studies. Understanding these factors is crucial for healthcare operations and resource planning.

Let's predict hospital length of stay (LOS) based on three key patient characteristics using synthetic data that reflects these evidence-based relationships:

1.  **Age** - Patient age in years
2.  **Severity Score** - Illness severity on a 1-10 scale
3.  **Number of Comorbidities** - Count of concurrent chronic conditions

### Loading the Dataset

Let's load the dataset using the code below:

```{r}
#| label: create-multiple-data

# Loading the dataset

hospital_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week11_hospital_data.csv")

# Summary statistics
head(hospital_data, 5)
```

### Exploratory Data Analysis

Before fitting the model, let's visually explore the distributions of our variables and their pairwise relationships.

```{r}
#| label: fig-eda-multiple
#| code-fold: true
#| fig-cap: "Pairwise relationships between predictors and length of stay"
#| fig-height: 7
#| fig-width: 9

# Pairs plot for continuous variables
# To use ggpairs(), you need to install the GGally package.
# wrap() customizes what each section of the matrix displays:
#   "smooth"     = scatterplot with a trend line
#   "cor"        = correlation coefficient (numeric summary)
#   "densityDiag" = density curve (distribution shape)
hospital_data %>%
  select(age, severity_score, num_comorbidities, length_of_stay) %>%
  ggpairs(
    lower = list(continuous = wrap("smooth", alpha = 0.3, linewidth = 0.5)),
    upper = list(continuous = wrap("cor", size = 3)),
    diag = list(continuous = wrap("densityDiag", alpha = 0.5))
  ) +
  theme_minimal(base_size = 10)
```

### Fitting the Multiple Linear Regression Model

Let's fit the linear model. Same as simple linear regression, we can use the `lm()` function in R.

```{r}
#| label: fit-multiple-model

# Fit multiple regression model
model_multiple <- lm(length_of_stay ~ age + severity_score + num_comorbidities,
                     data = hospital_data)

# Display summary
summary(model_multiple)
```

### Interpreting the Results

Each coefficient represents the effect of that variable **while holding all other variables constant**.

```{r}
#| echo: false
coefs <- coef(model_multiple)
```

1.  **Intercept (β₀ = `r round(coefs[1], 2)`)**: The predicted length of stay for a patient who is 0 years old, has a severity score of 0, and has 0 comorbidities. This value is not practically meaningful but is mathematically necessary.

2.  **Age (β₁ = `r round(coefs[2], 3)`)**: Holding severity and comorbidities constant, each additional year of age is associated with a **`r round(coefs[2], 3)`-day increase** in length of stay.

3.  **Severity Score (β₂ = `r round(coefs[3], 3)`)**: Holding age and comorbidities constant, each one-point increase in severity score is associated with a **`r round(coefs[3], 2)`-day increase** in length of stay.

4.  **Number of Comorbidities (β₃ = `r round(coefs[4], 3)`)**: Holding age and severity constant, each additional comorbidity is associated with a **`r round(coefs[4], 2)`-day increase** in length of stay.

#### Model Fit (R² and Adjusted R²)

```{r}
#| echo: false
r_squared_multi <- round(summary(model_multiple)$r.squared, 3)
adj_r_squared_multi <- round(summary(model_multiple)$adj.r.squared, 3)
```

From the `summary()` output above:

-   **R² = `r r_squared_multi`**: `r round(r_squared_multi * 100, 1)`% of variance in LOS is explained by the model
-   **Adjusted R² = `r adj_r_squared_multi`**: Adjusted R² penalizes for the number of predictors and is preferred when comparing models with different numbers of variables

**Why Adjusted R²?** Regular R² can only increase (or stay the same) when you add more predictors, even if those predictors have no real relationship with the outcome. Adjusted R² penalizes for the number of predictors, so it only increases if a new predictor significantly improves the model. This makes it a fairer metric for comparing models with different numbers of variables.

### Practical Application: Predicting LOS

Now let's use our fitted model to predict length of stay for different patient profiles. This demonstrates how the model can support clinical decision-making and resource planning. Let's assume that we have several new patients and their characteristics as below.

```{r}
#| label: multiple-new-patients

# Create example patients with different characteristics
example_patients <- data.frame(
  age = c(35, 75, 50, 80),
  severity_score = c(2, 5, 8, 7),
  num_comorbidities = c(0, 2, 3, 5)
)
example_patients
```

Let's get the predicted LOS. We can do this using the `predict()` function in R.

```{r}
#| label: multiple-predictions

# Make predictions
example_patients$predicted_los <- predict(model_multiple, newdata = example_patients)
example_patients 
```

#### Manual Calculation Example

Let's manually calculate one prediction to understand how multiple regression works. We'll predict LOS for the **second patient** in the table above (Age = 75, Severity = 5, Comorbidities = 2).

**Step 1: Write the fitted regression equation**

First, recall our fitted model equation with estimated coefficients:

```{r}
#| echo: false
# Extract coefficients for the equation
b0_multi <- round(coef(model_multiple)[1], 2)
b1_multi <- round(coef(model_multiple)[2], 3)
b2_multi <- round(coef(model_multiple)[3], 3)
b3_multi <- round(coef(model_multiple)[4], 3)
```

$$\widehat{\text{LOS}} = `r b0_multi` + `r b1_multi` \times \text{Age} + `r b2_multi` \times \text{Severity} + `r b3_multi` \times \text{Comorbidities}$$

Or in general notation:

$$\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 X_1 + \hat{\beta}_2 X_2 + \hat{\beta}_3 X_3$$

Where:

-   $\hat{Y}$ = Predicted length of stay
-   $X_1$ = Age (years)
-   $X_2$ = Severity score (1-10)
-   $X_3$ = Number of comorbidities
-   $\hat{\beta}_0$ = `r b0_multi` (Intercept)
-   $\hat{\beta}_1$ = `r b1_multi` (Age coefficient)
-   $\hat{\beta}_2$ = `r b2_multi` (Severity coefficient)
-   $\hat{\beta}_3$ = `r b3_multi` (Comorbidity coefficient)

**Step 2: Substitute and calculate**

```{r}
#| label: manual-multiple-prediction
#| echo: false

# Manual calculation for Elderly, Moderate patient
pred_los_manual <- coef(model_multiple)[1] + coef(model_multiple)[2] * 75 +
                   coef(model_multiple)[3] * 5 + coef(model_multiple)[4] * 2
```

For the "Elderly, Moderate" patient (Age = 75, Severity = 5, Comorbidities = 2):

$$\widehat{\text{LOS}} = `r b0_multi` + `r b1_multi` \times 75 + `r b2_multi` \times 5 + `r b3_multi` \times 2 = `r round(pred_los_manual, 2)`$$

The predicted length of stay is **`r round(pred_los_manual, 2)` days**. You can verify this matches the value in the prediction result above!

### Practice 2: Multiple Linear Regression {.unnumbered}

::: callout-note
## Your Turn!

Practice building and interpreting multiple regression models with a new scenario.
:::

### Scenario

You are now analyzing **patient satisfaction scores** across hospitals. You hypothesize that satisfaction is influenced by multiple factors:

-   **Wait time** (minutes): Longer waits → lower satisfaction
-   **Nurse staffing** (nurses per patient): Better staffing → higher satisfaction
-   **Facility age** (years): Older facilities → lower satisfaction

Your task is to build a multiple regression model to understand how these factors simultaneously influence patient satisfaction.

### Loading the Dataset

Run this code to load the practice dataset:

```{r}
#| label: generate-multiple-practice-data

# Loading the dataset

multiple_practice_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week11_practice_data2.csv")

# Preview data
head(multiple_practice_data, 5)
```

**Dataset Variables:**

| Variable | Description | Type |
|----|----|----|
| `hospital_id` | Unique hospital identifier | Numeric |
| `wait_time_min` | Average patient wait time (minutes) | Numeric (continuous) |
| `nurse_ratio` | Nurses per patient ratio | Numeric (continuous) |
| `facility_age_years` | Age of hospital building (years) | Numeric (continuous) |
| `satisfaction_score` | Average patient satisfaction (0-100 scale) | Numeric (continuous) |

### Task 2.1: Fit the Multiple Regression Model

Fit a multiple regression model predicting `satisfaction_score` from all three predictors: `wait_time_min`, `nurse_ratio`, and `facility_age_years`.

```{r}
#| label: task-2-1
#| eval: false

# Fill in the blanks
model_multiple_practice <- lm(___ ~ ___ + ___ + ___, data = ___)

# Display the summary
summary(___)
```

**Questions:**

1.  What is the coefficient for **wait_time_min**? Holding nurse ratio and facility age constant, what happens to satisfaction when wait time increases by 1 minute? Is this effect statistically significant (p \< 0.05)?

2.  What is the coefficient for **nurse_ratio**? Holding wait time and facility age constant, what happens to satisfaction when nurse staffing increases by 1 nurse per patient? Is this effect statistically significant?

3.  What is the coefficient for **facility_age_years**? Holding wait time and nurse ratio constant, what happens to satisfaction for each additional year of facility age? Is this effect statistically significant?

### Task 2.2: Compare Model Fit

Compare the multiple regression model to a simple regression model that only includes `wait_time_min`.

```{r}
#| label: task-2-2
#| eval: false

# Fit simple model with only wait time (fill in the blank)
model_simple_wait <- lm(___ ~ ___, data = multiple_practice_data)

# Compare R-squared values
cat("Simple model R²:", round(summary(___)$r.squared, 3), "\n")
cat("Multiple model R²:", round(summary(___)$r.squared, 3), "\n")
```

**Question**: How much additional variance is explained by adding nurse ratio and facility age to the model? Is the multiple regression model better?

### Task 2.3: Make Predictions

Use your multiple regression model to predict satisfaction for two different hospital scenarios.

```{r}
#| label: task-2-3
#| eval: false

# Create new hospital new_hospitals_multi (fill in the blanks)
new_hospitals_multi <- data.frame(
  wait_time_min = c(20, 70),         # Short vs. long wait
  nurse_ratio = c(3.5, 1.8),         # High vs. low staffing
  facility_age_years = c(5, 35)      # New vs. old facility
)

# Make predictions (fill in the blank)
new_hospitals_multi$predicted_satisfaction <- predict(___, newdata = ___)
new_hospitals_multi
```

**Question**: What is the difference in predicted satisfaction between the best-case and worst-case scenarios? What does this tell you about the combined impact of these factors?

------------------------------------------------------------------------

# Model Assumptions and Diagnostics

## Key Assumptions of Linear Regression

Linear regression relies on several assumptions. Understanding these assumptions helps ensure valid statistical inference and reliable predictions.

### 1. Linearity

-   The relationship between predictor(s) and outcome is linear.
-   If the true relationship is curved but we fit a straight line, predictions will be systematically biased.
-   We can visually check by drawing a scatterplot.
-   **If violated**, consider transforming variables (log, square root) or using polynomial terms.

### 2. Independence of Observations

-   Each observation's residual is unrelated to others' residuals.
-   Correlated errors (e.g., repeated measures, clustered data) underestimate standard errors, making p-values too small.
-   Know your data structure and study design (longitudinal? clustered?)
-   **If violated**, use mixed models, clustered standard errors, or time-series methods.

### 3. Homoscedasticity (Equal Variance)

-   Variance of residuals is constant across all fitted values.
-   Unequal variance (heteroscedasticity) makes standard errors unreliable and confidence intervals inaccurate.
-   We can visually inspect by drawing a Residuals vs Fitted plot (vertical spread should be roughly equal across x-axis)
-   **If violated**, use robust standard errors or transform Y.

### 4. Normality of Residuals

-   Residuals follow a normal distribution.
-   Critical for valid p-values and confidence intervals
-   We can visually inspect using a Q-Q plot (points should follow the diagonal line)
-   **If violated**, consider transforming Y or using robust methods.

### 5. No (Perfect) Multicollinearity (Multiple Regression Only)

-   Predictors are not highly correlated with each other.
-   High correlation between predictors inflates standard errors, making coefficients unstable and hard to interpret.
-   We can check the Variance Inflation Factor (VIF; VIF \> 10 indicates problematic multicollinearity)
-   **If violated**, remove redundant predictors, combine correlated predictors, or use penalized regression (we will cover this later in the semester).

------------------------------------------------------------------------

### Summary: Which Plot Checks Which Assumption?

| Assumption            | Primary Diagnostic                          |
|-----------------------|---------------------------------------------|
| **Linearity**         | Residuals vs Fitted (look for no pattern)   |
| **Independence**      | Know your study design                      |
| **Homoscedasticity**  | Residuals vs Fitted (equal vertical spread) |
| **Normality**         | Q-Q plot                                    |
| **Multicollinearity** | VIF values (not a plot)                     |

## Diagnostic Plots

We can draw diagnostic plots by using the `plot()` function in R.

-   `which = 1`: Residuals vs Fitted
-   `which = 2`: Q-Q plot

```{r}
#| label: fig-diagnostics
#| fig-cap: "Regression diagnostic plots"
#| fig-height: 4
#| fig-width: 8

# Create diagnostic plots
par(mfrow = c(1, 2))  # 1 row, 2 columns
plot(model_multiple, which = 1) # Residual vs. fitted plot
plot(model_multiple, which = 2) # Q-Q plot
par(mfrow = c(1, 1))  # Reset to default
```

### Interpreting the Two Standard Diagnostic Plots

**Plot 1: Residuals vs Fitted (left)**

-   **What to look for**: Random scatter around horizontal line at 0
-   **What it checks**: Linearity and homoscedasticity
-   **Good sign**: No pattern, equal spread across x-axis
-   **Red flags**:
    -   Curved pattern = non-linearity
    -   Funnel shape = heteroscedasticity

**Plot 2: Normal Q-Q (Quantile-Quantile; right)**

-   A Q-Q plot compares the distribution of your residuals to a theoretical normal distribution.
-   **What to look for**: Points following the diagonal reference line
-   **What it checks**: Normality of residuals
-   Points close around the diagonal indicate residuals are normally distributed

## Checking for Multicollinearity

Multicollinearity occurs when predictors are highly correlated, which can:

-   Inflate standard errors
-   Make individual coefficients unstable
-   Lead to counterintuitive results

### Variance Inflation Factor (VIF)

VIF quantifies how much multicollinearity inflates the variance of regression coefficients.

**How VIF works**:

For each predictor $X_j$:

1.  Regress $X_j$ on all other predictors
2.  Get $R^2_j$ from that regression (how well other predictors explain $X_j$)
3.  Calculate: $\text{VIF}_j = \frac{1}{1 - R^2_j}$

-   **Rule of Thumb**: VIF \> 10 (or Tolerance \< 0.1) indicates problematic multicollinearity.

We can get VIF values by using the `vif()` function in R (to use it, we first need to load the package `car`).

```{r}
#| label: vif-check

# Calculate VIF
vif_values <- car::vif(model_multiple)
vif_values
```

As all of them are less than 10, they are not highly correlated with each other.

------------------------------------------------------------------------

# Summary

## Key Takeaways

### Simple vs. Multiple Linear Regression

-   **Simple regression**: Examines how one factor relates to an outcome ("If X increases by 1, how much does Y change?")
-   **Multiple regression**: Examines how several factors together relate to an outcome, showing each factor's unique contribution while holding others constant

### How OLS Finds the Best Line

-   Ordinary Least Squares (OLS) finds the line that minimizes the sum of squared residuals
-   "Residuals" = vertical distance between actual data points and the fitted line

### Model Fit: R²

-   R² tells you what percentage of variation your model explains
-   Adjusted R² is better for multiple regression (accounts for number of predictors)
-   Higher is better, but context matters

### Making Predictions

-   Use fitted models to predict outcomes for new situations
-   Useful for "what-if" scenarios and planning

### Interpreting Coefficients

Ask three questions for each coefficient:

1.  **Direction**: Positive or negative relationship?
2.  **Magnitude**: How large is the effect practically?
3.  **Significance**: Is p \< 0.05? (But practical significance matters too!)

In multiple regression, coefficients show effects **while controlling for other variables**.

### Multicollinearity

-   When predictors are highly correlated, coefficients become unstable and can even flip signs
-   **Check**: VIF \> 10 indicates problems
-   **Solution**: Remove or combine correlated predictors

------------------------------------------------------------------------

# References and Further Reading

## Statistical Methods

-   James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning with Applications in R* (2nd ed.). Springer.
-   Fox, J. (2015). *Applied Regression Analysis and Generalized Linear Models* (3rd ed.). SAGE Publications.
-   Gelman, A., & Hill, J. (2006). *Data Analysis Using Regression and Multilevel/Hierarchical Models*. Cambridge University Press.

## Healthcare Applications

Examples are motivated by the following:

-   Abass, G., Asery, A., Al Badr, A., AlMaghlouth, A., AlOtaiby, S., & Heena, H. (2021). Patient satisfaction with the emergency department services at an academic teaching hospital. *Journal of Family Medicine and Primary Care*, 10(4), 1718-1725.

-   Stone, K., Zwiggelaar, R., Jones, P., & Mac Parthaláin, N. (2022). A systematic review of the prediction of hospital length of stay: Towards a unified framework. *PLOS Digital Health*, 1(4), e0000017.

-   McHugh, M. D., Aiken, L. H., Sloane, D. M., Windsor, C., Douglas, C., & Yates, P. (2021). Effects of nurse-to-patient ratio legislation on nurse staffing and patient mortality, readmissions, and length of stay: a prospective study in a panel of hospitals. *The Lancet*, 397(10288), 1905-1913.
