---
title: "Conjoint Analysis in Healthcare Management"
subtitle: "Rating-Based and Choice-Based Methods with R"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: false
    theme: cosmo
    highlight-style: github
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(tidyr)
library(broom)
library(mlogit)
set.seed(2026)
```

# Introduction to Conjoint Analysis

## What is Conjoint Analysis?

Conjoint analysis is a survey-based statistical technique used to understand how people make decisions when facing trade-offs between multiple attributes of a product or service. The core idea is simple: in real life, we rarely get everything we want. When choosing a health insurance plan, for example, we might prefer both low premiums AND low deductibles, but plans with low premiums often come with high deductibles. Conjoint analysis helps us understand how people navigate these trade-offs.

The term "conjoint" comes from "considered jointly"—we're studying how attributes are evaluated together rather than in isolation. This is crucial because asking someone "Do you prefer low cost?" will always get a "yes," but asking them to choose between a low-cost option with long wait times versus a high-cost option with short wait times reveals their true priorities.

## Why Use Conjoint Analysis in Healthcare Management?

Healthcare managers face constant decisions about service design:

-   **Insurance plan design**: What combination of premium, deductible, and coverage will attract enrollees?
-   **Clinic operations**: Should we prioritize shorter wait times or extended hours?
-   **Telehealth services**: Do patients value video consultations enough to pay more for them?
-   **Hospital services**: What amenities matter most to patients choosing where to receive care?

Conjoint analysis answers these questions by quantifying preferences. Instead of relying on intuition or simple surveys ("Would you like shorter wait times?"), conjoint analysis reveals how much each feature actually matters when people must make realistic trade-offs.

## Key Outputs from Conjoint Analysis

Conjoint analysis produces several valuable insights:

### 1. Part-Worth Utilities

**Part-worth utilities** (also called "part-worths" or "utility scores") are the building blocks of conjoint analysis. They represent the value or satisfaction that a person derives from a specific level of an attribute.

Think of it this way: every product or service is a bundle of attributes. A health insurance plan is a bundle of (premium, deductible, network size, etc.). The total utility (satisfaction) a person gets from a plan is the sum of the part-worth utilities for each attribute level in that plan.

**Example**: If the part-worth for "Large Network" is +3 and for "Small Network" is 0, then having a large network adds 3 utility units to the overall plan value. These numbers are relative—they only make sense in comparison to other levels of the same attribute.

### 2. Attribute Importance

Once we have part-worths, we can calculate how important each attribute is relative to others. This is done by looking at the **range** of utilities within each attribute (highest level minus lowest level) and comparing these ranges.

**Example**: If premium ranges from 0 to 5 utility units and network size ranges from 0 to 2 units, then premium has a larger impact on preferences and is therefore more "important."

### 3. Willingness-to-Pay (WTP)

When price or cost is one of the attributes, we can convert utility differences into dollar values. This tells us how much extra a person would be willing to pay for an improvement in another attribute.

**Example**: If people would pay \$50 more per month for a large network versus a small network, this helps managers decide whether investing in network expansion is worthwhile.

### 4. Market Share Prediction

By knowing the utilities for each attribute level, we can predict what percentage of people would choose one option over another. This is invaluable for forecasting demand and competitive positioning.

## Two Main Approaches

This tutorial covers the two most common conjoint methods:

### Rating-Based Conjoint

-   **What respondents do**: Rate each profile on a scale (e.g., 1-10)
-   **Analysis approach**: Linear regression (OLS)
-   **Advantages**: Simple to understand and implement
-   **Limitations**: Respondents may rate all options similarly, not revealing true trade-offs

### Choice-Based Conjoint (Discrete Choice Experiment)

-   **What respondents do**: Choose their preferred option from a set
-   **Analysis approach**: Logistic regression / Multinomial logit
-   **Advantages**: Mimics real-world decisions; forces explicit trade-offs
-   **Limitations**: Requires more sophisticated experimental design

Both methods ultimately produce the same types of outputs (part-worths, importance, WTP, market share), but choice-based conjoint is generally preferred in academic research and practice because it better reflects actual decision-making behavior (Louviere et al., 2000).

------------------------------------------------------------------------

# Scenario 1: Health Insurance Plan Selection

## Context and Background

One of the most studied applications of conjoint analysis in healthcare management is understanding how employees choose health insurance plans during open enrollment periods. This is a high-stakes decision for both employees (who will live with the plan for a year) and employers (who subsidize the plans).

Research has shown that employees often struggle with this decision due to the complexity of plan attributes. Studies by Sinaiko and Hirth (2011) in the *Journal of Health Economics* found that many employees make "dominated choices"—selecting plans that are objectively worse than alternatives on all dimensions. Understanding which attributes drive choices can help employers design better choice architectures and communication strategies.

Our example uses three key attributes that consistently appear in health plan conjoint studies:

| Attribute         | Level 1 | Level 2 | Why It Matters                        |
|-------------------|---------|---------|---------------------------------------|
| Monthly Premium   | \$200   | \$300   | Direct, visible cost paid every month |
| Annual Deductible | \$500   | \$1,500 | Amount paid before insurance kicks in |
| Network Size      | Large   | Small   | Access to preferred doctors/hospitals |

------------------------------------------------------------------------

## Rating-Based Conjoint Analysis

### Conceptual Overview

In rating-based conjoint analysis, we show respondents one hypothetical profile at a time and ask them to rate how attractive it is. For example:

> **Plan A**: Monthly premium \$200, Deductible \$1,500, Small network
>
> "On a scale from 1 to 10, how likely would you be to choose this plan?"

The respondent might give this a 4 out of 10. Then we show them another profile:

> **Plan B**: Monthly premium \$300, Deductible \$500, Large network
>
> Rating: 7 out of 10

By collecting ratings for multiple profiles across many respondents, we can use regression analysis to decompose these ratings into the contribution of each attribute. This works because we've systematically varied the attributes across profiles.

### The Statistical Model

Rating-based conjoint uses **ordinary least squares (OLS) regression**, the same technique used in standard linear regression. The model assumes that a person's rating is a linear combination of the utilities they assign to each attribute level:

$$\text{Rating} = \beta_0 + \beta_1(\text{Premium}) + \beta_2(\text{Deductible}) + \beta_3(\text{Network}) + \varepsilon$$

Where: - $\beta_0$ is the intercept (baseline rating) - $\beta_1, \beta_2, \beta_3$ are the **part-worth utilities** for each attribute - $\varepsilon$ is random error (individual variation, measurement error, etc.)

The coefficients we estimate ARE the part-worths. They tell us how much each unit change in an attribute affects the rating.

### Generating the Experimental Design

First, we create all possible combinations of our attribute levels. With 3 attributes × 2 levels each, we have $2^3 = 8$ possible profiles. This is called a **full factorial design**.

```{r}
#| label: rating-insurance-data

# Create full factorial design for insurance plans
# This generates all possible combinations of attribute levels
insurance_profiles <- expand.grid(
  premium = c(200, 300),
  deductible = c(500, 1500),
  network = c("Large", "Small")
)
insurance_profiles$profile_id <- 1:nrow(insurance_profiles)

# Display all 8 profiles
knitr::kable(insurance_profiles,
             col.names = c("Premium ($)", "Deductible ($)", "Network", "Profile ID"),
             caption = "Health Insurance Plan Profiles (Full Factorial Design)")
```

### Simulating Respondent Data

In a real study, you would collect actual ratings from respondents. For this tutorial, we simulate data based on assumed "true" preferences. This allows us to verify that our analysis recovers the correct values.

**Our assumed true preferences:** - Premium: -0.02 utility per dollar (people prefer lower premiums) - Deductible: -0.003 utility per dollar (people prefer lower deductibles) - Network: +3 utility for Large network (people prefer larger networks)

These values reflect that a \$100 difference in monthly premium (-0.02 × 100 = -2 utility) has about the same impact as a large vs. small network (+3 utility), suggesting people would pay roughly \$150/month more for a large network.

```{r}
#| label: rating-insurance-simulate

# Simulate ratings from 100 respondents
# Each respondent rates all 8 profiles

n_respondents <- 100

rating_insurance <- data.frame()

for (resp in 1:n_respondents) {
  for (i in 1:nrow(insurance_profiles)) {
    # Calculate utility based on true preferences
    # The baseline of 50 centers our utility scale
    utility <- 50 +
      -0.02 * insurance_profiles$premium[i] +
      -0.003 * insurance_profiles$deductible[i] +
      3 * (insurance_profiles$network[i] == "Large") +
      rnorm(1, 0, 3)  # Add random noise to simulate individual variation

    # Convert utility to a 1-10 rating scale
    # We divide by 5 to scale appropriately and bound between 1 and 10
    rating <- round(pmin(pmax(utility / 5, 1), 10), 1)

    rating_insurance <- rbind(rating_insurance, data.frame(
      respondent_id = resp,
      profile_id = i,
      premium = insurance_profiles$premium[i],
      deductible = insurance_profiles$deductible[i],
      network = insurance_profiles$network[i],
      rating = rating
    ))
  }
}

# Show first few rows
cat("Dataset structure: ", nrow(rating_insurance), "rows (100 respondents × 8 profiles)\n\n")
head(rating_insurance, 10)
```

### Estimating Part-Worth Utilities

Now we use linear regression to estimate the part-worths. We code our variables numerically so the coefficients are directly interpretable:

-   Premium and deductible are kept as continuous variables (in dollars)
-   Network is coded as a binary indicator (1 = Large, 0 = Small)

```{r}
#| label: rating-insurance-estimation

# Create binary indicator for network size
rating_insurance <- rating_insurance %>%
  mutate(
    network_large = ifelse(network == "Large", 1, 0)
  )

# Fit linear regression model
model_rating_ins <- lm(rating ~ premium + deductible + network_large,
                       data = rating_insurance)

# Display full results
summary(model_rating_ins)
```

**Interpreting the Output:**

```{r}
#| label: rating-insurance-coef-table

# Extract and display coefficients nicely
coef_rating_ins <- tidy(model_rating_ins)
knitr::kable(coef_rating_ins, digits = 4,
             caption = "Estimated Part-Worth Utilities (Rating-Based Conjoint)")
```

**What do these coefficients mean?**

-   **Intercept (≈10)**: The baseline rating when all attributes are at their reference levels (but this is somewhat artificial given our coding)
-   **Premium (≈-0.004)**: Each additional dollar in monthly premium decreases the rating by about 0.004 points. So a \$100 increase reduces rating by 0.4 points.
-   **Deductible (≈-0.0006)**: Each additional dollar in deductible decreases rating by about 0.0006 points. A \$1,000 increase reduces rating by 0.6 points.
-   **Network_large (≈0.6)**: Having a large network (vs. small) increases rating by about 0.6 points.

Note: Our recovered coefficients are scaled versions of the true utilities because we transformed utility to a 1-10 rating scale.

### Calculating Attribute Importance

Attribute importance tells us which features have the biggest impact on preferences. The calculation is straightforward:

1.  For each attribute, calculate the **utility range**: the difference between the highest and lowest part-worth values
2.  Sum all the ranges
3.  Express each range as a percentage of the total

**Why use ranges?** Because importance depends on how much the attribute levels differ. If we tested premiums of \$200 vs. \$201, premium would appear unimportant even though cost matters greatly to people. The range captures the actual impact given the levels we tested.

```{r}
#| label: rating-insurance-importance

# Extract coefficients
coef_ins <- coef(model_rating_ins)

# Calculate utility ranges for each attribute
# For continuous variables: coefficient × (max level - min level)
# For binary variables: coefficient × 1

range_premium <- abs(coef_ins["premium"] * (300 - 200))  # $100 difference
range_deductible <- abs(coef_ins["deductible"] * (1500 - 500))  # $1000 difference
range_network <- abs(coef_ins["network_large"] * 1)  # Large vs Small

# Total range (sum of all ranges)
total_range <- range_premium + range_deductible + range_network

# Calculate importance as percentage
importance_rating_ins <- data.frame(
  attribute = c("Monthly Premium", "Annual Deductible", "Network Size"),
  utility_range = c(range_premium, range_deductible, range_network),
  importance_pct = c(range_premium, range_deductible, range_network) / total_range * 100
)

knitr::kable(importance_rating_ins, digits = 2,
             col.names = c("Attribute", "Utility Range", "Importance (%)"),
             caption = "Attribute Importance Scores")
```

**Visualization:**

```{r}
#| label: rating-insurance-importance-plot
#| fig-width: 8
#| fig-height: 5

ggplot(importance_rating_ins, aes(x = reorder(attribute, importance_pct),
                                   y = importance_pct)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(aes(label = paste0(round(importance_pct, 1), "%")),
            hjust = -0.1, size = 4) +
  coord_flip() +
  labs(title = "Attribute Importance: Health Insurance Plan Selection",
       subtitle = "Rating-Based Conjoint Analysis",
       x = "",
       y = "Relative Importance (%)",
       caption = "Importance = utility range as % of total range across all attributes") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold")) +
  ylim(0, max(importance_rating_ins$importance_pct) * 1.2)
```

**Interpretation**: The results show which attributes have the greatest impact on plan ratings given the levels we tested. This helps managers prioritize which plan features to emphasize in communications or modify in plan design.

### Calculating Willingness-to-Pay (WTP)

Willingness-to-pay converts the abstract utility values into concrete dollar amounts. This is possible because we included a price attribute (monthly premium) in our study.

**The Logic**: If we know how much utility a person gets from each attribute and how much utility they lose per dollar of cost, we can calculate the dollar amount that would exactly offset any utility gain.

**The Formula**: $$\text{WTP} = -\frac{\beta_{\text{attribute}}}{\beta_{\text{price}}}$$

The negative sign is needed because price coefficients are negative (people dislike higher prices) while beneficial attribute coefficients are positive.

**Example Calculation**: - Network coefficient = +0.6 (utility gain from large network) - Premium coefficient = -0.004 (utility loss per dollar) - WTP = -0.6 / (-0.004) = \$150

This means respondents value a large network as much as they would value a \$150 reduction in monthly premium.

```{r}
#| label: rating-insurance-wtp

# Extract relevant coefficients
beta_network <- coef_ins["network_large"]
beta_premium <- coef_ins["premium"]
beta_deductible <- coef_ins["deductible"]

# Calculate WTP for large network (vs small)
wtp_network <- -beta_network / beta_premium

# Calculate WTP for $1000 lower deductible
# (How much extra premium would someone pay to reduce deductible by $1000?)
wtp_deductible_reduction <- -(beta_deductible * -1000) / beta_premium

cat("Willingness-to-Pay Estimates:\n")
cat("━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━\n")
cat("Large Network (vs. Small): $", round(wtp_network, 2), " more per month\n", sep = "")
cat("$1,000 Lower Deductible:   $", round(wtp_deductible_reduction, 2), " more per month\n", sep = "")
cat("━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━\n")
```

**Practical Application**: These WTP values help managers make cost-benefit decisions. If expanding the provider network costs the insurer \$100 per member per month, and members are willing to pay \$150 for it, the expansion could be profitable while also improving member satisfaction.

### Market Share Simulation

One of the most powerful applications of conjoint analysis is predicting how the market would respond to different product configurations. Given the part-worth utilities, we can calculate the overall utility of any hypothetical product and predict choice shares.

**The Process**: 1. Define two (or more) competing options 2. Calculate total utility for each option by summing the part-worths 3. Convert utilities to choice probabilities using the **logit formula**

**Why the Logit Formula?**

The logit model comes from random utility theory (McFadden, 1974, who won the Nobel Prize for this work). It assumes that people choose the option with the highest utility, but there's some randomness in choices. The formula is:

$$P(\text{Choose A}) = \frac{e^{U_A}}{e^{U_A} + e^{U_B}}$$

Where $U_A$ and $U_B$ are the total utilities of options A and B. This formula has nice properties: - If $U_A > U_B$, then $P(A) > 0.5$ (higher utility = higher probability) - If $U_A = U_B$, then $P(A) = 0.5$ (equal utilities = equal chances) - Probabilities always sum to 1

```{r}
#| label: rating-insurance-marketshare

# Define two competing health plans
plan_A <- data.frame(
  name = "Premium Plan",
  premium = 250,
  deductible = 500,
  network_large = 1,
  description = "Mid-price, Low deductible, Large network"
)

plan_B <- data.frame(
  name = "Budget Plan",
  premium = 200,
  deductible = 1500,
  network_large = 0,
  description = "Low-price, High deductible, Small network"
)

# Calculate total utility for each plan using estimated coefficients
utility_A <- predict(model_rating_ins, plan_A)
utility_B <- predict(model_rating_ins, plan_B)

# Apply logit formula for market share
share_A <- exp(utility_A) / (exp(utility_A) + exp(utility_B))
share_B <- exp(utility_B) / (exp(utility_A) + exp(utility_B))

# Display results
market_share_rating <- data.frame(
  Plan = c(plan_A$name, plan_B$name),
  Description = c(plan_A$description, plan_B$description),
  Total_Utility = round(c(utility_A, utility_B), 3),
  Market_Share = paste0(round(c(share_A, share_B) * 100, 1), "%")
)

knitr::kable(market_share_rating,
             col.names = c("Plan", "Configuration", "Total Utility", "Predicted Share"),
             caption = "Market Share Simulation: Two Competing Plans")
```

**Interpretation**: The model predicts that approximately `r round(share_A * 100, 0)`% of enrollees would choose the Premium Plan despite its higher cost, because the low deductible and large network compensate for the price difference. This type of simulation helps managers:

-   Forecast enrollment in different plan designs
-   Understand competitive positioning
-   Optimize plan configurations

------------------------------------------------------------------------

## Choice-Based Conjoint Analysis

### Conceptual Overview

Choice-based conjoint (CBC), also known as a **Discrete Choice Experiment (DCE)**, is considered the gold standard in conjoint analysis because it closely mimics real-world decision-making (Louviere, Hensher, & Swait, 2000).

Instead of rating profiles individually, respondents are shown **choice sets** containing two or more options and asked to pick their preferred one:

> **Choice Set 1:**
>
> |            | Plan A  | Plan B |
> |------------|---------|--------|
> | Premium    | \$200   | \$300  |
> | Deductible | \$1,500 | \$500  |
> | Network    | Small   | Large  |
>
> **Which plan would you choose?** ☐ Plan A ☐ Plan B

By forcing a choice, we ensure that respondents make explicit trade-offs. Someone who wants both low premium AND low deductible must decide which matters more.

### The Statistical Model

When choice sets have exactly **two options** (binary choice), we can use **standard logistic regression**. The key insight is that the choice depends on the **difference** in utilities between the options.

**Mathematical Foundation**: - Utility of Plan A: $U_A = \beta_1 X_{A1} + \beta_2 X_{A2} + ...$ - Utility of Plan B: $U_B = \beta_1 X_{B1} + \beta_2 X_{B2} + ...$ - Person chooses A if $U_A > U_B$, i.e., if $U_A - U_B > 0$

This difference can be written as: $$U_A - U_B = \beta_1 (X_{A1} - X_{B1}) + \beta_2 (X_{A2} - X_{B2}) + ... = \beta_1 \Delta X_1 + \beta_2 \Delta X_2 + ...$$

So we model the probability of choosing A as a function of the **attribute differences**: $$P(\text{Choose A}) = \frac{1}{1 + e^{-(β_1 \Delta X_1 + β_2 \Delta X_2 + ...)}}$$

This is exactly the logistic regression model! The coefficients we estimate are the part-worth utilities.

### Generating Choice Data

We create choice sets by randomly pairing profiles, then simulate choices based on our assumed true utilities.

```{r}
#| label: choice-insurance-data

# Parameters for simulation
n_respondents <- 100
n_choice_sets <- 8  # Each respondent completes 8 choice tasks

choice_insurance <- data.frame()

for (resp in 1:n_respondents) {
  for (cs in 1:n_choice_sets) {
    # Randomly select two different profiles for this choice set
    profiles <- sample(1:8, 2)

    alt_A <- insurance_profiles[profiles[1], ]
    alt_B <- insurance_profiles[profiles[2], ]

    # Calculate utilities using same true values as before
    utility_A <- -0.02 * alt_A$premium -
                 0.003 * alt_A$deductible +
                 3 * (alt_A$network == "Large") +
                 rnorm(1, 0, 1)  # Random component

    utility_B <- -0.02 * alt_B$premium -
                 0.003 * alt_B$deductible +
                 3 * (alt_B$network == "Large") +
                 rnorm(1, 0, 1)

    # Choice: 1 if A chosen, 0 if B chosen
    choice_A <- as.integer(utility_A > utility_B)

    choice_insurance <- rbind(choice_insurance, data.frame(
      respondent_id = resp,
      choice_set = cs,
      # Alternative A attributes
      premium_A = alt_A$premium,
      deductible_A = alt_A$deductible,
      network_A = alt_A$network,
      # Alternative B attributes
      premium_B = alt_B$premium,
      deductible_B = alt_B$deductible,
      network_B = alt_B$network,
      # Choice outcome
      choice_A = choice_A
    ))
  }
}

# Display example choice sets
cat("Example choice data (first 6 choice sets):\n\n")
head(choice_insurance, 6)
```

### Method 1: Binary Logistic Regression

For binary choices, we create **difference variables** and use standard logistic regression.

```{r}
#| label: choice-insurance-logistic

# Create difference variables (A minus B)
choice_insurance <- choice_insurance %>%
  mutate(
    diff_premium = premium_A - premium_B,
    diff_deductible = deductible_A - deductible_B,
    diff_network = (network_A == "Large") - (network_B == "Large")
  )

# Fit binary logistic regression
# Note: We remove the intercept (-1) because there's no inherent preference
# for "Alternative A" vs "Alternative B" - only the attributes matter
model_logit_ins <- glm(choice_A ~ diff_premium + diff_deductible + diff_network - 1,
                       data = choice_insurance,
                       family = binomial)

summary(model_logit_ins)
```

**Important**: We exclude the intercept (`-1` in the formula) because the labels "A" and "B" are arbitrary. There's no reason someone would prefer "Plan A" over "Plan B" independent of their attributes.

```{r}
#| label: choice-insurance-logistic-table

coef_logit_ins <- tidy(model_logit_ins)
knitr::kable(coef_logit_ins, digits = 4,
             caption = "Part-Worth Utilities (Choice-Based: Binary Logistic Regression)")
```

**Interpreting Logistic Regression Coefficients**:

The coefficients are on the **log-odds scale**. A coefficient of -0.08 for premium means that each additional dollar in premium for Plan A (compared to Plan B) decreases the log-odds of choosing A by 0.08.

More intuitively: - **Negative premium coefficient**: People prefer lower premiums (choose the cheaper option) - **Negative deductible coefficient**: People prefer lower deductibles - **Positive network coefficient**: People prefer larger networks

### Method 2: Multinomial Logit (mlogit)

The `mlogit` package is designed for choice models and can handle: - More than 2 alternatives per choice set - Alternative-specific and individual-specific variables - More complex model specifications

For our binary choice data, `mlogit` should give similar results to standard logistic regression, confirming our approach.

**Data Restructuring**: `mlogit` requires data in "long" format where each row represents one alternative:

```{r}
#| label: choice-insurance-mlogit

# Reshape data to long format
# Each choice set becomes two rows (one per alternative)
choice_long <- data.frame()

for (i in 1:nrow(choice_insurance)) {
  row <- choice_insurance[i, ]

  # Add Alternative A
  choice_long <- rbind(choice_long, data.frame(
    respondent_id = row$respondent_id,
    choice_set = paste0(row$respondent_id, "_", row$choice_set),
    alternative = "A",
    premium = row$premium_A,
    deductible = row$deductible_A,
    network_large = as.integer(row$network_A == "Large"),
    choice = row$choice_A
  ))

  # Add Alternative B
  choice_long <- rbind(choice_long, data.frame(
    respondent_id = row$respondent_id,
    choice_set = paste0(row$respondent_id, "_", row$choice_set),
    alternative = "B",
    premium = row$premium_B,
    deductible = row$deductible_B,
    network_large = as.integer(row$network_B == "Large"),
    choice = 1 - row$choice_A
  ))
}

# Convert to mlogit data format
# idx specifies: choice_set ID, and alternative ID
choice_mlogit <- dfidx(choice_long,
                       choice = "choice",
                       idx = c("choice_set", "alternative"))

# Fit conditional logit model
# The "| 0" means no alternative-specific constants
model_mlogit_ins <- mlogit(choice ~ premium + deductible + network_large | 0,
                           data = choice_mlogit)

summary(model_mlogit_ins)
```

**Comparing the Two Approaches**:

The coefficients from `mlogit` should be very close to those from binary logistic regression, which validates that both methods are equivalent for binary choice data. The `mlogit` approach is more flexible and is the standard for choice-based conjoint with 3+ alternatives.

### Attribute Importance (Choice-Based)

The calculation is identical to rating-based conjoint: compute utility ranges and normalize.

```{r}
#| label: choice-insurance-importance

# Use binary logistic regression coefficients
coef_choice <- coef(model_logit_ins)

# Calculate utility ranges
range_premium_c <- abs(coef_choice["diff_premium"] * 100)  # $100 difference
range_deductible_c <- abs(coef_choice["diff_deductible"] * 1000)  # $1000 difference
range_network_c <- abs(coef_choice["diff_network"] * 1)  # Binary

total_range_c <- range_premium_c + range_deductible_c + range_network_c

importance_choice_ins <- data.frame(
  attribute = c("Monthly Premium", "Annual Deductible", "Network Size"),
  utility_range = c(range_premium_c, range_deductible_c, range_network_c),
  importance_pct = c(range_premium_c, range_deductible_c, range_network_c) / total_range_c * 100
)

knitr::kable(importance_choice_ins, digits = 2,
             col.names = c("Attribute", "Utility Range", "Importance (%)"),
             caption = "Attribute Importance (Choice-Based Conjoint)")
```

```{r}
#| label: choice-insurance-importance-plot
#| fig-width: 8
#| fig-height: 5

ggplot(importance_choice_ins, aes(x = reorder(attribute, importance_pct),
                                   y = importance_pct)) +
  geom_col(fill = "darkorange", width = 0.6) +
  geom_text(aes(label = paste0(round(importance_pct, 1), "%")),
            hjust = -0.1, size = 4) +
  coord_flip() +
  labs(title = "Attribute Importance: Health Insurance Plan Selection",
       subtitle = "Choice-Based Conjoint Analysis",
       x = "",
       y = "Relative Importance (%)",
       caption = "Importance = utility range as % of total range across all attributes") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold")) +
  ylim(0, max(importance_choice_ins$importance_pct) * 1.2)
```

### Willingness-to-Pay (Choice-Based)

The WTP calculation uses the same formula as before:

```{r}
#| label: choice-insurance-wtp

beta_network_c <- coef_choice["diff_network"]
beta_premium_c <- coef_choice["diff_premium"]
beta_deductible_c <- coef_choice["diff_deductible"]

wtp_network_c <- -beta_network_c / beta_premium_c
wtp_deductible_c <- -(beta_deductible_c * -1000) / beta_premium_c

cat("Willingness-to-Pay Estimates (Choice-Based):\n")
cat("━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━\n")
cat("Large Network (vs. Small): $", round(wtp_network_c, 2), " more per month\n", sep = "")
cat("$1,000 Lower Deductible:   $", round(wtp_deductible_c, 2), " more per month\n", sep = "")
cat("━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━\n")
```

### Market Share Simulation (Choice-Based)

Using the `mlogit` coefficients for market share prediction:

```{r}
#| label: choice-insurance-marketshare

# Get mlogit coefficients
coef_ml <- coef(model_mlogit_ins)

# Calculate utilities for the same two plans
utility_A_c <- coef_ml["premium"] * 250 +
               coef_ml["deductible"] * 500 +
               coef_ml["network_large"] * 1

utility_B_c <- coef_ml["premium"] * 200 +
               coef_ml["deductible"] * 1500 +
               coef_ml["network_large"] * 0

# Logit market share
share_A_c <- exp(utility_A_c) / (exp(utility_A_c) + exp(utility_B_c))
share_B_c <- exp(utility_B_c) / (exp(utility_A_c) + exp(utility_B_c))

market_share_choice <- data.frame(
  Plan = c("Premium Plan", "Budget Plan"),
  Description = c("$250, $500 ded., Large network",
                  "$200, $1500 ded., Small network"),
  Total_Utility = round(c(utility_A_c, utility_B_c), 3),
  Market_Share = paste0(round(c(share_A_c, share_B_c) * 100, 1), "%")
)

knitr::kable(market_share_choice,
             col.names = c("Plan", "Configuration", "Total Utility", "Predicted Share"),
             caption = "Market Share Simulation (Choice-Based Conjoint)")
```

------------------------------------------------------------------------

# Scenario 2: Telehealth Program Preferences

## Context and Background

The COVID-19 pandemic accelerated the adoption of telehealth services, and healthcare organizations now face decisions about how to structure "hybrid" care models that combine virtual and in-person visits. Understanding patient preferences for telehealth attributes is crucial for designing services that people will actually use.

Research by Bürkin et al. (2023) in *BMC Health Services Research* used discrete choice experiments to examine patient preferences for primary care consultations, finding that attributes like wait time and consultation mode significantly influenced preferences. Our example is inspired by this and similar studies.

## Attributes and Levels

| Attribute | Level 1  | Level 2 | Why It Matters                   |
|-----------|----------|---------|----------------------------------|
| Wait Time | Same-day | 3 days  | How quickly patients can be seen |
| Mode      | Video    | Phone   | Communication richness           |
| Co-pay    | \$20     | \$40    | Out-of-pocket cost per visit     |

------------------------------------------------------------------------

## Rating-Based Conjoint Analysis

### Generate Data

```{r}
#| label: rating-telehealth-data

# Create full factorial design
telehealth_profiles <- expand.grid(
  wait_time = c("Same-day", "3 days"),
  mode = c("Video", "Phone"),
  copay = c(20, 40)
)
telehealth_profiles$profile_id <- 1:nrow(telehealth_profiles)

knitr::kable(telehealth_profiles,
             col.names = c("Wait Time", "Mode", "Co-pay ($)", "Profile ID"),
             caption = "Telehealth Program Profiles")

# Simulate ratings
# True utilities:
# Wait: +4 for same-day (patients strongly prefer immediate access)
# Mode: +2 for video (visual communication preferred over audio-only)
# Copay: -0.1 per dollar

n_respondents <- 100

rating_telehealth <- data.frame()

for (resp in 1:n_respondents) {
  for (i in 1:nrow(telehealth_profiles)) {
    utility <- 50 +
      4 * (telehealth_profiles$wait_time[i] == "Same-day") +
      2 * (telehealth_profiles$mode[i] == "Video") +
      -0.1 * telehealth_profiles$copay[i] +
      rnorm(1, 0, 2)

    rating <- round(pmin(pmax(utility / 5, 1), 10), 1)

    rating_telehealth <- rbind(rating_telehealth, data.frame(
      respondent_id = resp,
      profile_id = i,
      wait_time = telehealth_profiles$wait_time[i],
      mode = telehealth_profiles$mode[i],
      copay = telehealth_profiles$copay[i],
      rating = rating
    ))
  }
}

cat("Dataset: ", nrow(rating_telehealth), " observations\n")
```

### Parameter Estimation

```{r}
#| label: rating-telehealth-estimation

# Create indicator variables
rating_telehealth <- rating_telehealth %>%
  mutate(
    wait_sameday = ifelse(wait_time == "Same-day", 1, 0),
    mode_video = ifelse(mode == "Video", 1, 0)
  )

# Fit model
model_rating_th <- lm(rating ~ wait_sameday + mode_video + copay,
                      data = rating_telehealth)

summary(model_rating_th)

knitr::kable(tidy(model_rating_th), digits = 4,
             caption = "Part-Worth Utilities: Telehealth (Rating-Based)")
```

### Attribute Importance

```{r}
#| label: rating-telehealth-importance

coef_th <- coef(model_rating_th)

range_wait <- abs(coef_th["wait_sameday"])
range_mode <- abs(coef_th["mode_video"])
range_copay <- abs(coef_th["copay"] * 20)  # $20 difference

total_range_th <- range_wait + range_mode + range_copay

importance_rating_th <- data.frame(
  attribute = c("Wait Time", "Consultation Mode", "Co-pay"),
  importance = c(range_wait, range_mode, range_copay) / total_range_th * 100
)

knitr::kable(importance_rating_th, digits = 2,
             col.names = c("Attribute", "Importance (%)"),
             caption = "Attribute Importance: Telehealth (Rating-Based)")

ggplot(importance_rating_th, aes(x = reorder(attribute, importance), y = importance)) +
  geom_col(fill = "forestgreen", width = 0.6) +
  geom_text(aes(label = paste0(round(importance, 1), "%")), hjust = -0.1, size = 4) +
  coord_flip() +
  labs(title = "Attribute Importance: Telehealth Program",
       subtitle = "Rating-Based Conjoint Analysis",
       x = "", y = "Relative Importance (%)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold")) +
  ylim(0, 60)
```

### Willingness-to-Pay

```{r}
#| label: rating-telehealth-wtp

beta_wait_th <- coef_th["wait_sameday"]
beta_mode_th <- coef_th["mode_video"]
beta_copay_th <- coef_th["copay"]

wtp_sameday <- -beta_wait_th / beta_copay_th
wtp_video <- -beta_mode_th / beta_copay_th

cat("Willingness-to-Pay Estimates:\n")
cat("━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━\n")
cat("Same-day Appointment (vs. 3 days): $", round(wtp_sameday, 2), "\n", sep = "")
cat("Video Mode (vs. Phone):            $", round(wtp_video, 2), "\n", sep = "")
cat("━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━\n")
```

**Interpretation**: Patients would be willing to pay approximately \$`r round(wtp_sameday, 0)` more for a same-day appointment compared to waiting 3 days. This suggests that reducing wait times could support a premium pricing strategy.

### Market Share Simulation

```{r}
#| label: rating-telehealth-marketshare

# Compare two program designs
program_A <- data.frame(
  wait_sameday = 1, mode_video = 1, copay = 40,
  desc = "Premium: Same-day, Video, $40"
)
program_B <- data.frame(
  wait_sameday = 0, mode_video = 0, copay = 20,
  desc = "Basic: 3-day wait, Phone, $20"
)

utility_A_th <- predict(model_rating_th, program_A)
utility_B_th <- predict(model_rating_th, program_B)

share_A_th <- exp(utility_A_th) / (exp(utility_A_th) + exp(utility_B_th))
share_B_th <- 1 - share_A_th

market_share_th <- data.frame(
  Program = c("Premium Program", "Basic Program"),
  Configuration = c(program_A$desc, program_B$desc),
  Market_Share = paste0(round(c(share_A_th, share_B_th) * 100, 1), "%")
)

knitr::kable(market_share_th,
             caption = "Predicted Market Share: Telehealth Programs (Rating-Based)")
```

------------------------------------------------------------------------

## Choice-Based Conjoint Analysis

### Generate Choice Data

```{r}
#| label: choice-telehealth-data

n_respondents <- 100
n_choice_sets <- 8

choice_telehealth <- data.frame()

for (resp in 1:n_respondents) {
  for (cs in 1:n_choice_sets) {
    profiles <- sample(1:8, 2)

    alt_A <- telehealth_profiles[profiles[1], ]
    alt_B <- telehealth_profiles[profiles[2], ]

    utility_A <- 4 * (alt_A$wait_time == "Same-day") +
                 2 * (alt_A$mode == "Video") -
                 0.1 * alt_A$copay + rnorm(1, 0, 0.5)
    utility_B <- 4 * (alt_B$wait_time == "Same-day") +
                 2 * (alt_B$mode == "Video") -
                 0.1 * alt_B$copay + rnorm(1, 0, 0.5)

    choice_A <- as.integer(utility_A > utility_B)

    choice_telehealth <- rbind(choice_telehealth, data.frame(
      respondent_id = resp,
      choice_set = cs,
      wait_A = alt_A$wait_time, mode_A = alt_A$mode, copay_A = alt_A$copay,
      wait_B = alt_B$wait_time, mode_B = alt_B$mode, copay_B = alt_B$copay,
      choice_A = choice_A
    ))
  }
}
```

### Parameter Estimation

```{r}
#| label: choice-telehealth-logistic

# Create difference variables
choice_telehealth <- choice_telehealth %>%
  mutate(
    diff_wait = (wait_A == "Same-day") - (wait_B == "Same-day"),
    diff_mode = (mode_A == "Video") - (mode_B == "Video"),
    diff_copay = copay_A - copay_B
  )

# Binary logistic regression
model_logit_th <- glm(choice_A ~ diff_wait + diff_mode + diff_copay - 1,
                      data = choice_telehealth, family = binomial)

summary(model_logit_th)

knitr::kable(tidy(model_logit_th), digits = 4,
             caption = "Part-Worth Utilities: Telehealth (Choice-Based)")
```

### Expanding to mlogit

```{r}
#| label: choice-telehealth-mlogit

# Reshape to long format
choice_long_th <- data.frame()

for (i in 1:nrow(choice_telehealth)) {
  row <- choice_telehealth[i, ]

  choice_long_th <- rbind(choice_long_th,
    data.frame(
      choice_set = paste0(row$respondent_id, "_", row$choice_set),
      alternative = "A",
      wait_sameday = as.integer(row$wait_A == "Same-day"),
      mode_video = as.integer(row$mode_A == "Video"),
      copay = row$copay_A,
      choice = row$choice_A
    ),
    data.frame(
      choice_set = paste0(row$respondent_id, "_", row$choice_set),
      alternative = "B",
      wait_sameday = as.integer(row$wait_B == "Same-day"),
      mode_video = as.integer(row$mode_B == "Video"),
      copay = row$copay_B,
      choice = 1 - row$choice_A
    )
  )
}

choice_mlogit_th <- dfidx(choice_long_th,
                          choice = "choice",
                          idx = c("choice_set", "alternative"))

model_mlogit_th <- mlogit(choice ~ wait_sameday + mode_video + copay | 0,
                          data = choice_mlogit_th)

summary(model_mlogit_th)
```

### Attribute Importance

```{r}
#| label: choice-telehealth-importance

coef_choice_th <- coef(model_logit_th)

range_wait_c <- abs(coef_choice_th["diff_wait"])
range_mode_c <- abs(coef_choice_th["diff_mode"])
range_copay_c <- abs(coef_choice_th["diff_copay"] * 20)

total_range_th_c <- range_wait_c + range_mode_c + range_copay_c

importance_choice_th <- data.frame(
  attribute = c("Wait Time", "Consultation Mode", "Co-pay"),
  importance = c(range_wait_c, range_mode_c, range_copay_c) / total_range_th_c * 100
)

knitr::kable(importance_choice_th, digits = 2,
             col.names = c("Attribute", "Importance (%)"),
             caption = "Attribute Importance: Telehealth (Choice-Based)")

ggplot(importance_choice_th, aes(x = reorder(attribute, importance), y = importance)) +
  geom_col(fill = "purple4", width = 0.6) +
  geom_text(aes(label = paste0(round(importance, 1), "%")), hjust = -0.1, size = 4) +
  coord_flip() +
  labs(title = "Attribute Importance: Telehealth Program",
       subtitle = "Choice-Based Conjoint Analysis",
       x = "", y = "Relative Importance (%)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold")) +
  ylim(0, 60)
```

### Willingness-to-Pay

```{r}
#| label: choice-telehealth-wtp

wtp_sameday_c <- -coef_choice_th["diff_wait"] / coef_choice_th["diff_copay"]
wtp_video_c <- -coef_choice_th["diff_mode"] / coef_choice_th["diff_copay"]

cat("Willingness-to-Pay Estimates (Choice-Based):\n")
cat("━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━\n")
cat("Same-day Appointment: $", round(wtp_sameday_c, 2), "\n", sep = "")
cat("Video Mode:           $", round(wtp_video_c, 2), "\n", sep = "")
cat("━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━\n")
```

### Market Share Simulation

```{r}
#| label: choice-telehealth-marketshare

coef_ml_th <- coef(model_mlogit_th)

utility_A_th_c <- coef_ml_th["wait_sameday"] * 1 +
                  coef_ml_th["mode_video"] * 1 +
                  coef_ml_th["copay"] * 40

utility_B_th_c <- coef_ml_th["wait_sameday"] * 0 +
                  coef_ml_th["mode_video"] * 0 +
                  coef_ml_th["copay"] * 20

share_A_th_c <- exp(utility_A_th_c) / (exp(utility_A_th_c) + exp(utility_B_th_c))
share_B_th_c <- 1 - share_A_th_c

market_share_th_c <- data.frame(
  Program = c("Premium Program", "Basic Program"),
  Configuration = c("Same-day, Video, $40", "3-day, Phone, $20"),
  Market_Share = paste0(round(c(share_A_th_c, share_B_th_c) * 100, 1), "%")
)

knitr::kable(market_share_th_c,
             caption = "Predicted Market Share: Telehealth Programs (Choice-Based)")
```

------------------------------------------------------------------------

# Summary and Comparison

## Method Comparison

```{r}
#| label: summary-table

comparison <- data.frame(
  Aspect = c(
    "Data Collection",
    "Analysis Method",
    "Key Assumption",
    "Realism",
    "Implementation Complexity",
    "When to Use"
  ),
  Rating_Based = c(
    "Rate each profile (e.g., 1-10 scale)",
    "OLS Linear Regression",
    "Utility → Rating is linear",
    "Lower: no forced trade-offs",
    "Simpler to design and analyze",
    "Exploratory research, simple products"
  ),
  Choice_Based = c(
    "Choose preferred option from set",
    "Logistic / Multinomial Logit",
    "Random utility maximization",
    "Higher: mimics actual decisions",
    "Requires careful experimental design",
    "When predicting actual choices matters"
  )
)

knitr::kable(comparison,
             col.names = c("Aspect", "Rating-Based", "Choice-Based"),
             caption = "Comparison of Conjoint Analysis Methods")
```

## Key Takeaways

1.  **Both methods yield similar part-worth utilities** when properly designed, because they're measuring the same underlying preferences

2.  **Choice-based is generally preferred** in academic research and industry practice because it better mimics real decision-making and forces explicit trade-offs

3.  **Binary logistic regression works well** for choice sets with exactly 2 alternatives—it's simpler and equivalent to multinomial logit

4.  **The `mlogit` package** extends to 3+ alternatives and offers more modeling flexibility

5.  **All outputs transfer across methods**: Attribute importance, WTP, and market share calculations use the same formulas regardless of how part-worths were estimated

6.  **The key insight**: Conjoint analysis transforms subjective preferences into quantitative data that managers can use for concrete decisions about pricing, feature prioritization, and competitive strategy

------------------------------------------------------------------------

# References and Sources

## Academic References

-   **Green, P. E., & Srinivasan, V. (1990)**. Conjoint analysis in marketing: New developments with implications for research and practice. *Journal of Marketing*, 54(4), 3-19. *(Classic overview of conjoint methods)*

-   **Louviere, J. J., Hensher, D. A., & Swait, J. D. (2000)**. *Stated Choice Methods: Analysis and Applications*. Cambridge University Press. *(Comprehensive textbook on choice-based conjoint and discrete choice experiments)*

-   **McFadden, D. (1974)**. Conditional logit analysis of qualitative choice behavior. In P. Zarembka (Ed.), *Frontiers in Econometrics* (pp. 105-142). Academic Press. *(Nobel Prize-winning work on discrete choice models)*

-   **Sinaiko, A. D., & Hirth, R. A. (2011)**. Consumers, health insurance and dominated choices. *Journal of Health Economics*, 30(2), 450-457. *(Health insurance plan choice study)*

-   **Bürkin, F., et al. (2023)**. Patient preferences for primary care consultations: A discrete choice experiment. *BMC Health Services Research*, 23, 123. *(Telehealth preferences study)*

## R Packages Used

-   **`mlogit`**: Croissant, Y. (2020). mlogit: Multinomial Logit Models. R package. https://CRAN.R-project.org/package=mlogit

-   **`broom`**: Robinson, D., Hayes, A., & Couch, S. (2023). broom: Convert Statistical Objects into Tidy Tibbles. https://CRAN.R-project.org/package=broom

## Code and Examples

The examples and R code in this tutorial were developed specifically for this document. The scenarios (health insurance plans and telehealth programs) are inspired by the academic literature cited above but use simulated data generated with assumed "true" utility values to demonstrate the methods.

The data simulation approach (generating responses based on assumed utilities plus random error) is a standard technique for teaching conjoint analysis, as it allows students to verify that the estimated parameters recover the known true values.

## Additional Resources

-   **Orme, B. (2010)**. *Getting Started with Conjoint Analysis: Strategies for Product Design and Pricing Research* (2nd ed.). Research Publishers LLC. *(Practical guide with intuitive explanations)*

-   **Sawtooth Software**: https://sawtoothsoftware.com/ *(Commercial software with excellent technical papers on conjoint methods)*
