---
title: "Mean Comparison 2: ANOVA (Practice Activities)"
subtitle: "MBA642: Business Analytics in Healthcare Management"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
    theme: cosmo
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(emmeans)
library(parameters)

# Set seed for reproducibility
set.seed(2026)
```

# Overview

In this practice session, you will apply the ANOVA techniques learned in the lecture. This is a learning activity designed to help you gain hands-on experience with one-way ANOVA and repeated-measures ANOVA.

**What you'll practice:**

- Conducting one-way ANOVA for comparing independent groups
- Performing Tukey's HSD post-hoc tests
- Interpreting ANOVA tables and post-hoc results
- Conducting repeated-measures ANOVA for longitudinal data
- Visualizing group comparisons
- Making evidence-based conclusions from ANOVA results

**Format**: Fill-in-the-blank code sections (marked with `___`)

**Instructions**:

1. Complete the code by replacing `___` with appropriate values or function names
2. Run each chunk to verify your analysis
3. Answer the interpretation questions
4. Compare your results with the solutions at the end

**Note**: This practice is completion-based. The goal is to learn and experiment—mistakes are part of the process!

---

# Recap: Key Concepts

Before starting the activities, let's review essential concepts from the lecture:

## One-Way ANOVA

**Purpose**: Compare means across **3 or more independent groups**

**When to use**: Different departments, treatment groups, facilities, or any categorical grouping

**Key steps**:

1. Run ANOVA using `aov()` function
2. Check if F-test is significant (p < 0.05)
3. If significant, run post-hoc tests to identify which groups differ

**Basic R syntax**:

```{r}
#| label: recap-oneway
#| eval: false

# Step 1: Run one-way ANOVA
anova_result <- aov(outcome_variable ~ grouping_variable, data = your_data)
summary(anova_result)

# Step 2: Post-hoc test using emmeans
emm <- emmeans(anova_result, ~ grouping_variable)
pairs(emm, adjust = "tukey")
```

## Repeated-Measures ANOVA

**Purpose**: Compare means across **3 or more time points or conditions** for the **same subjects**

**When to use**: Longitudinal studies, tracking patients over time, within-subject experiments

**Key difference**: Each subject is measured multiple times, so we account for individual differences

**Basic R syntax**:

```{r}
#| label: recap-repeated
#| eval: false

# Step 1: Run repeated-measures ANOVA
# Error(subject_id/time) specifies within-subjects design
rm_anova <- aov(outcome ~ time + Error(subject_id/time), data = your_data)
summary(rm_anova)

# Step 2: Post-hoc tests
emm_time <- emmeans(rm_anova, ~ time)
pairs(emm_time, adjust = "tukey")
```

## Understanding ANOVA Output

**ANOVA Table Components**:

- **Df**: Degrees of freedom
- **Sum Sq**: Sum of squares (variance)
- **Mean Sq**: Mean square (Sum Sq / Df)
- **F value**: Test statistic (ratio of between-group to within-group variance)
- **Pr(>F)**: p-value (if < 0.05, groups differ significantly)

**Post-Hoc Test Output**:

- **contrast**: Which two groups are being compared
- **estimate**: Difference between group means
- **p.value**: Adjusted p-value (if < 0.05, those two groups differ significantly)

---

# Common Errors and How to Fix Them

You **WILL** encounter errors while practicing—that's completely normal! Here are common ANOVA-specific errors:

## Error 1: "contrasts can be applied only to factors"

```r
Error in contrasts<-: contrasts can be applied only to factors with 2 or more levels
```

**Cause**: Your grouping variable is not a factor, or it has only 1 level

**Fix**:

```r
# Convert grouping variable to factor
your_data$group <- as.factor(your_data$group)

# Check number of levels
levels(your_data$group)  # Should show multiple levels
```

## Error 2: "variables must have same length"

**Cause**: Your data frame is incorrectly structured (often in repeated measures)

**Fix**: Make sure your data is in **long format** where:
- Each row = one measurement
- Subject ID column identifies which subject
- Time/condition column identifies when/which measurement

```r
# Check your data structure
head(your_data)
str(your_data)
```

## Error 3: emmeans can't find the model

```r
Error: No emmGrid object was supplied
```

**Cause**: You forgot to save the ANOVA result, or there's a typo in the model name

**Fix**:

```r
# Make sure you saved the ANOVA result
anova_result <- aov(outcome ~ group, data = your_data)

# Then use that saved result in emmeans
emm <- emmeans(anova_result, ~ group)  # Use the same name you saved
```

## Error 4: "Error() model is singular"

**Cause**: In repeated measures, subject ID is not unique or not properly specified

**Fix**:

```r
# Make sure subject_id is a factor
your_data$subject_id <- as.factor(your_data$subject_id)

# Check that each subject appears in all conditions
table(your_data$subject_id, your_data$time)
```

## Error 5: Forgetting to load packages

```r
Error: could not find function "emmeans"
```

**Cause**: The `emmeans` package isn't loaded

**Fix**: Run the setup chunk at the top of this file:

```r
library(emmeans)
```

## Still Stuck?

1. **Check your variable names**: Are they spelled correctly? Are they in the right data frame?
2. **Run previous chunks**: Make sure all data generation and setup code has been executed
3. **Read error messages carefully**: They often tell you exactly what's wrong
4. **Compare with lecture examples**: Look at the syntax used in similar examples
5. **Check data structure**: Use `str()` and `head()` to verify your data looks correct

---

# Practice Activity 1: Staff Training Programs (One-Way ANOVA)

## Scenario

A hospital compares the effectiveness of **three different training programs** for improving nurse medication safety scores (out of 100).

**Context**: Research on simulation-based and blended learning approaches has shown significant improvements in medication safety competency for nurses.

**Programs**:
- **Traditional**: Classroom lectures and written materials
- **Simulation**: Hands-on practice with realistic patient scenarios
- **Blended**: Combination of online modules, in-person sessions, and simulation

**Research Question**: Do the three training programs result in different medication safety competency scores?

## Dataset

```{r}
#| label: practice-oneway1-data

# Training program scores
program_traditional <- c(78, 82, 75, 80, 77, 83, 79, 81, 76, 80,
                        82, 78, 84, 77, 79, 81, 75, 83, 78, 80)

program_simulation <- c(85, 88, 82, 87, 89, 84, 86, 90, 83, 88,
                        87, 85, 89, 84, 86, 88, 82, 87, 85, 86)

program_blended <- c(88, 92, 86, 90, 93, 87, 89, 91, 85, 90,
                     92, 88, 91, 86, 89, 93, 87, 90, 88, 91)

# Create data frame
training_data <- data.frame(
  score = c(program_traditional, program_simulation, program_blended),
  program = factor(rep(c("Traditional", "Simulation", "Blended"), each = 20))
)

# View first few rows
head(training_data)
```

## Task 1.1: Explore the Data

Before running ANOVA, let's examine summary statistics:

```{r}
#| label: task-1-1
#| eval: false

# Calculate mean and SD by program
summary_stats <- training_data %>%
  group_by(___) %>%
  summarise(
    n = n(),
    mean = round(mean(___), 2),
    sd = round(sd(___), 2)
  )

# Display the table
knitr::kable(summary_stats)
```

**Question**: Which program appears to have the highest mean score? Do the means look different enough to suggest a real difference?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.2: Visualize the Comparison

Create a boxplot comparing scores across the three programs:

```{r}
#| label: task-1-2
#| eval: false

# Create boxplot
ggplot(data = training_data, mapping = aes(x = ___, y = ___, fill = ___)) +
  geom_boxplot(alpha = 0.7, width = 0.4) +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(
    title = "Medication Safety Scores by Training Program",
    x = "___",
    y = "Safety Score (out of 100)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Question**: Based on the boxplot, do you see differences in the median scores across programs? Do any programs show more variability (wider boxes)?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.3: Conduct One-Way ANOVA

Now run the ANOVA to test if the differences are statistically significant:

```{r}
#| label: task-1-3
#| eval: false

# Perform one-way ANOVA
anova_training <- aov(___ ~ ___, data = training_data)

# Display ANOVA table
summary(___)
```

**Questions**:

1. What is the F-statistic?
2. What is the p-value?
3. Is the result statistically significant at α = 0.05?
4. What is your conclusion?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answers**:

1. F-statistic =
2. p-value =
3. Significant? (Yes/No)
4. Conclusion:

</div>

## Task 1.4: Post-Hoc Tests

Since ANOVA is significant, run Tukey's HSD to identify which programs differ:

```{r}
#| label: task-1-4
#| eval: false

# Calculate estimated marginal means
emm_training <- emmeans(___, ~ ___)
emm_training

# Pairwise comparisons with Tukey adjustment
tukey_training <- pairs(___, adjust = "tukey")
tukey_training
```

**Questions**:

1. Which pairwise comparisons are statistically significant (p < 0.05)?
2. Which training program appears most effective?
3. What is the practical significance of these differences?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answers**:

1. Significant comparisons:
2. Most effective program:
3. Practical significance:

</div>

---

# Practice Activity 2: Hospital Operating Costs (One-Way ANOVA)

## Scenario

Compare average cost per patient discharge (\$) across **four different hospital types**.

**Context**: Hospital costs vary significantly by ownership type and facility characteristics, with meaningful differences observed between academic, community, and specialty hospitals.

**Hospital Types**:
- **Academic Medical**: Large teaching hospitals affiliated with medical schools
- **Community**: Non-teaching hospitals serving local populations
- **Critical Access**: Small rural hospitals (≤25 beds) serving remote areas
- **Specialty**: Hospitals focused on specific conditions (cardiac, orthopedic, etc.)

**Research Question**: Do operating costs per discharge differ significantly across hospital types?

## Dataset

```{r}
#| label: practice-oneway2-data

# Cost per discharge by hospital type
academic_medical <- c(15200, 16500, 14800, 15900, 16200, 15100, 16800, 15500,
                      16100, 15700, 14900, 16400, 15300, 16000, 15800)

community <- c(11500, 12200, 10800, 11900, 12500, 11200, 10500, 12100,
               11700, 10900, 12300, 11400, 10700, 12000, 11600)

critical_access <- c(13800, 14500, 13200, 14200, 14800, 13500, 13000, 14600,
                     14100, 13700, 14400, 13300, 14700, 13900, 14300)

specialty <- c(18500, 19200, 17800, 18900, 19500, 18200, 17500, 19100,
               18700, 17900, 19300, 18400, 17700, 19000, 18600)

# Create data frame
cost_data <- data.frame(
  cost = c(academic_medical, community, critical_access, specialty),
  hospital_type = factor(rep(c("Academic Medical", "Community",
                               "Critical Access", "Specialty"), each = 15))
)

head(cost_data)
```

## Task 2.1: Summary Statistics and Visualization

```{r}
#| label: task-2-1
#| eval: false

# Summary statistics
cost_summary <- cost_data %>%
  group_by(___) %>%
  summarise(
    n = n(),
    mean_cost = round(mean(___), 0),
    sd_cost = round(sd(___), 0)
  )

knitr::kable(cost_summary)

# Boxplot
ggplot(cost_data, aes(x = ___, y = ___, fill = ___)) +
  geom_boxplot(alpha = 0.7, width = 0.5) +
  labs(
    title = "___",
    x = "Hospital Type",
    y = "Cost per Discharge ($)"
  ) +
  scale_y_continuous(labels = scales::dollar) +  # Format y-axis as dollars
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

## Task 2.2: One-Way ANOVA

```{r}
#| label: task-2-2
#| eval: false

# Run ANOVA
anova_cost <- aov(___ ~ ___, data = ___)
summary(anova_cost)
```

**Question**: Are the cost differences statistically significant?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 2.3: Post-Hoc Comparisons

```{r}
#| label: task-2-3
#| eval: false

# Estimated marginal means
emm_cost <- emmeans(___, ~ ___)

# Pairwise comparisons
pairs(___, adjust = "tukey")
```

**Questions**:

1. Which hospital type has the highest costs? Which has the lowest?
2. Are all pairwise comparisons significant, or only some?
3. What might explain these cost differences?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answers**:

1. Highest: _________, Lowest: _________
2. All significant? _________
3. Explanation:

</div>

---

# Practice Activity 3: Pain Management Follow-up (Repeated-Measures ANOVA)

## Scenario

A pain clinic measures pain intensity (0-10 VAS scale) for 20 chronic pain patients at **five time points** during a 6-month treatment program.

**Context**: The Visual Analogue Scale (VAS) is a validated, reliable tool for assessing chronic pain intensity over time.

**Time Points**:
- **Baseline**: Before treatment starts
- **Month 1**: Early treatment phase
- **Month 2**: Continued intervention
- **Month 4**: Mid-term follow-up
- **Month 6**: End of program

**Research Question**: Does pain intensity decrease significantly over the course of the treatment program?

## Dataset

```{r}
#| label: practice-repeated1-data

# Pain scores at 5 time points for 20 patients
n_pain <- 20

pain_baseline <- c(7.5, 8.2, 6.8, 7.9, 8.5, 7.1, 8.0, 7.3, 8.4, 7.6,
                   7.8, 8.1, 6.9, 7.7, 8.3, 7.2, 8.6, 7.4, 7.9, 8.0)
pain_month1 <- pain_baseline - runif(n_pain, 0.3, 0.8)
pain_month2 <- pain_month1 - runif(n_pain, 0.4, 0.9)
pain_month4 <- pain_month2 - runif(n_pain, 0.3, 0.7)
pain_month6 <- pain_month4 - runif(n_pain, 0.2, 0.6)

# Create data frame in long format
pain_data <- data.frame(
  patient_id = rep(1:n_pain, 5),
  time = factor(rep(c("Baseline", "Month 1", "Month 2", "Month 4", "Month 6"),
                    each = n_pain),
                levels = c("Baseline", "Month 1", "Month 2", "Month 4", "Month 6")),
  pain = c(pain_baseline, pain_month1, pain_month2, pain_month4, pain_month6)
)

# IMPORTANT: Convert patient_id to factor
pain_data$patient_id <- as.factor(pain_data$patient_id)

head(pain_data, 10)
```

## Task 3.1: Summary Statistics by Time

```{r}
#| label: task-3-1
#| eval: false

# Calculate mean pain at each time point
pain_summary <- pain_data %>%
  group_by(___) %>%
  summarise(
    n = n(),
    mean_pain = round(mean(___), 2),
    sd_pain = round(sd(___), 2)
  )

knitr::kable(pain_summary)
```

**Question**: Looking at the means, does pain appear to decrease over time?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 3.2: Visualize Trajectories

Create a "spaghetti plot" showing individual patient pain trajectories:

```{r}
#| label: task-3-2
#| eval: false

# Spaghetti plot with individual lines and overall mean
ggplot(pain_data, aes(x = ___, y = ___, group = ___)) +
  geom_line(alpha = 0.3, color = "gray50") +  # Individual patient lines
  geom_point(alpha = 0.3) +
  stat_summary(aes(group = 1), fun = mean, geom = "line",  # Overall mean line
               linewidth = 2, color = "red") +
  stat_summary(aes(group = 1), fun = mean, geom = "point",
               size = 4, color = "red") +
  labs(
    title = "___",
    subtitle = "Red line shows mean across all patients",
    x = "Time Point",
    y = "Pain Intensity (VAS 0-10)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Question**: Do most patients show a declining trend? Are there any patients who don't improve?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 3.3: Repeated-Measures ANOVA

Now test if the changes are statistically significant:

```{r}
#| label: task-3-3
#| eval: false

# Repeated-measures ANOVA
# Error(patient_id/time) specifies within-subjects design
rm_anova_pain <- aov(___ ~ ___ + Error(___/___), data = pain_data)

# Display results
summary(rm_anova_pain)
```

**Questions**:

1. What is the F-value for the time effect?
2. What is the p-value?
3. Is there a significant change in pain over time?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answers**:

1. F =
2. p =
3. Significant? (Yes/No)

</div>

## Task 3.4: Post-Hoc Tests for Time Points

Identify which time points differ significantly:

```{r}
#| label: task-3-4
#| eval: false

# Estimated marginal means by time
emm_pain <- emmeans(___, ~ ___)
emm_pain

# Pairwise comparisons
pairs(___, adjust = "tukey")
```

**Questions**:

1. At which time point is pain significantly lower than baseline?
2. Is there continued improvement from Month 4 to Month 6?
3. What do these results suggest about the treatment timeline?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answers**:

1. Significantly lower than baseline:
2. Month 4 to Month 6 improvement:
3. Clinical implications:

</div>

---

# Practice Activity 4: Employee Engagement Survey (Repeated-Measures ANOVA)

## Scenario

An HR department surveys nurse engagement quarterly over one year (4 time points). Engagement is measured on a 1-100 scale.

**Time Points**:
- **Q1**: First quarter (winter)
- **Q2**: Second quarter (spring)
- **Q3**: Third quarter (summer)
- **Q4**: Fourth quarter (fall)

**Research Question**: Does nurse engagement change significantly across the four quarters?

## Dataset

```{r}
#| label: practice-repeated2-data

# Engagement scores (1-100) for 25 nurses over 4 quarters
n_engage <- 25

q1 <- c(68, 72, 65, 70, 74, 67, 71, 66, 73, 69,
        70, 64, 75, 68, 72, 66, 74, 67, 71, 69,
        73, 65, 70, 68, 72)
q2 <- q1 + rnorm(n_engage, mean = 3, sd = 2)
q3 <- q2 + rnorm(n_engage, mean = 2, sd = 2)
q4 <- q3 + rnorm(n_engage, mean = 2, sd = 2)

# Create data frame
engagement_data <- data.frame(
  nurse_id = rep(1:n_engage, 4),
  quarter = factor(rep(c("Q1", "Q2", "Q3", "Q4"), each = n_engage),
                   levels = c("Q1", "Q2", "Q3", "Q4")),
  engagement = c(q1, q2, q3, q4)
)

# Convert nurse_id to factor
engagement_data$nurse_id <- as.factor(engagement_data$nurse_id)

head(engagement_data)
```

## Task 4.1: Your Turn - Complete Analysis

This is your chance to conduct the entire analysis independently! Follow these steps:

1. Calculate summary statistics by quarter
2. Create a visualization (spaghetti plot or boxplot)
3. Run repeated-measures ANOVA
4. Interpret the F-test
5. If significant, run post-hoc tests
6. Draw conclusions

**Hint**: Follow the same structure as Practice Activity 3.

```{r}
#| label: task-4-1
#| eval: false

# Step 1: Summary statistics
engagement_summary <- ___

# Step 2: Visualization
# (Choose: spaghetti plot or boxplot by quarter)

# Step 3: Repeated-measures ANOVA
rm_anova_engage <- ___

# Step 4: View results
summary(___)

# Step 5: Post-hoc tests (if significant)
emm_engage <- ___
pairs(___)
```

**Questions**:

1. Does engagement change significantly over the year?
2. If yes, which quarters differ from each other?
3. What recommendations would you make to HR based on these results?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answers**:

1. Significant change?
2. Which quarters differ:
3. Recommendations:

</div>

---

# Summary and Reflection

## Key Takeaways

After completing this practice activity, you should be able to:

- ✓ Conduct one-way ANOVA to compare 3+ independent groups
- ✓ Perform and interpret Tukey's HSD post-hoc tests
- ✓ Conduct repeated-measures ANOVA for longitudinal data
- ✓ Visualize group comparisons and trajectories over time
- ✓ Interpret ANOVA tables and make evidence-based conclusions
- ✓ Distinguish when to use one-way vs. repeated-measures ANOVA

## When to Use Which Test?

| Situation | Test to Use |
|-----------|-------------|
| Comparing 3+ **different groups** (e.g., three hospitals) | **One-Way ANOVA** |
| Comparing **same subjects** at 3+ time points (e.g., baseline, 3 months, 6 months) | **Repeated-Measures ANOVA** |
| Comparing only 2 groups | **t-test** (independent or paired) |
| After significant ANOVA | **Post-hoc tests** (Tukey's HSD) |

## Common Pitfalls to Avoid

1. **Running multiple t-tests** instead of ANOVA → inflates Type I error
2. **Forgetting post-hoc tests** → knowing groups differ but not which ones
3. **Using one-way ANOVA** for repeated measures → ignores within-subject correlation
4. **Not checking data structure** → especially critical for repeated measures
5. **Confusing F-statistic with p-value** → F tells you strength, p tells you significance

---

# Activity Solutions

Solutions for these activities are available below. Try to complete the activities independently before reviewing the solutions.

## Practice Activity 1 Solutions

### Task 1.1 Solution

```{r}
#| label: solution-1-1
#| code-fold: true

summary_stats <- training_data %>%
  group_by(program) %>%
  summarise(
    n = n(),
    mean = round(mean(score), 2),
    sd = round(sd(score), 2)
  )

knitr::kable(summary_stats)
```

The Blended program appears to have the highest mean (around 89), followed by Simulation (around 86), and Traditional (around 79).

### Task 1.2 Solution

```{r}
#| label: solution-1-2
#| code-fold: true

ggplot(data = training_data, mapping = aes(x = program, y = score, fill = program)) +
  geom_boxplot(alpha = 0.7, width = 0.4) +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(
    title = "Medication Safety Scores by Training Program",
    x = "Training Program",
    y = "Safety Score (out of 100)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The boxplot shows clear differences in median scores, with Blended highest and Traditional lowest.

### Task 1.3 Solution

```{r}
#| label: solution-1-3
#| code-fold: true

anova_training <- aov(score ~ program, data = training_data)
summary(anova_training)
```

The ANOVA is highly significant (p < 0.001), indicating that the three programs produce different mean scores.

### Task 1.4 Solution

```{r}
#| label: solution-1-4
#| code-fold: true

emm_training <- emmeans(anova_training, ~ program)
emm_training

tukey_training <- pairs(emm_training, adjust = "tukey")
tukey_training
```

All three pairwise comparisons are significant. Blended performs best, followed by Simulation, then Traditional.

---

## Practice Activity 2 Solutions

### Task 2.1 Solution

```{r}
#| label: solution-2-1
#| code-fold: true

cost_summary <- cost_data %>%
  group_by(hospital_type) %>%
  summarise(
    n = n(),
    mean_cost = round(mean(cost), 0),
    sd_cost = round(sd(cost), 0)
  )

knitr::kable(cost_summary)

ggplot(cost_data, aes(x = hospital_type, y = cost, fill = hospital_type)) +
  geom_boxplot(alpha = 0.7, width = 0.5) +
  labs(
    title = "Cost per Discharge by Hospital Type",
    x = "Hospital Type",
    y = "Cost per Discharge ($)"
  ) +
  scale_y_continuous(labels = scales::dollar) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

### Task 2.2 Solution

```{r}
#| label: solution-2-2
#| code-fold: true

anova_cost <- aov(cost ~ hospital_type, data = cost_data)
summary(anova_cost)
```

The F-test is highly significant, indicating substantial cost differences across hospital types.

### Task 2.3 Solution

```{r}
#| label: solution-2-3
#| code-fold: true

emm_cost <- emmeans(anova_cost, ~ hospital_type)
pairs(emm_cost, adjust = "tukey")
```

Specialty hospitals have the highest costs, Community hospitals the lowest. All pairwise differences are significant except possibly Critical Access vs Academic Medical (depending on exact p-value).

---

## Practice Activity 3 Solutions

### Task 3.1 Solution

```{r}
#| label: solution-3-1
#| code-fold: true

pain_summary <- pain_data %>%
  group_by(time) %>%
  summarise(
    n = n(),
    mean_pain = round(mean(pain), 2),
    sd_pain = round(sd(pain), 2)
  )

knitr::kable(pain_summary)
```

Mean pain decreases from baseline (~7.7) to Month 6 (~5.1), suggesting improvement.

### Task 3.2 Solution

```{r}
#| label: solution-3-2
#| code-fold: true

ggplot(pain_data, aes(x = time, y = pain, group = patient_id)) +
  geom_line(alpha = 0.3, color = "gray50") +
  geom_point(alpha = 0.3) +
  stat_summary(aes(group = 1), fun = mean, geom = "line",
               linewidth = 2, color = "red") +
  stat_summary(aes(group = 1), fun = mean, geom = "point",
               size = 4, color = "red") +
  labs(
    title = "Individual Patient Pain Trajectories",
    subtitle = "Red line shows mean across all patients",
    x = "Time Point",
    y = "Pain Intensity (VAS 0-10)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Most patients show declining trends, though some variability exists.

### Task 3.3 Solution

```{r}
#| label: solution-3-3
#| code-fold: true

rm_anova_pain <- aov(pain ~ time + Error(patient_id/time), data = pain_data)
summary(rm_anova_pain)
```

The time effect is highly significant, confirming that pain changes significantly over the treatment period.

### Task 3.4 Solution

```{r}
#| label: solution-3-4
#| code-fold: true

emm_pain <- emmeans(rm_anova_pain, ~ time)
emm_pain

pairs(emm_pain, adjust = "tukey")
```

Each successive time point shows significant improvement from baseline, with most improvement occurring in early months.

---

## Practice Activity 4 Solutions

### Task 4.1 Solution

```{r}
#| label: solution-4-1
#| code-fold: true

# Step 1: Summary statistics
engagement_summary <- engagement_data %>%
  group_by(quarter) %>%
  summarise(
    n = n(),
    mean_engagement = round(mean(engagement), 2),
    sd_engagement = round(sd(engagement), 2)
  )

knitr::kable(engagement_summary)

# Step 2: Visualization
ggplot(engagement_data, aes(x = quarter, y = engagement, group = nurse_id)) +
  geom_line(alpha = 0.3, color = "gray50") +
  geom_point(alpha = 0.3) +
  stat_summary(aes(group = 1), fun = mean, geom = "line",
               linewidth = 2, color = "blue") +
  stat_summary(aes(group = 1), fun = mean, geom = "point",
               size = 4, color = "blue") +
  labs(
    title = "Nurse Engagement Over Four Quarters",
    subtitle = "Blue line shows mean across all nurses",
    x = "Quarter",
    y = "Engagement Score (1-100)"
  ) +
  theme_minimal()

# Step 3: Repeated-measures ANOVA
rm_anova_engage <- aov(engagement ~ quarter + Error(nurse_id/quarter),
                       data = engagement_data)

# Step 4: Results
summary(rm_anova_engage)

# Step 5: Post-hoc tests
emm_engage <- emmeans(rm_anova_engage, ~ quarter)
pairs(emm_engage, adjust = "tukey")
```

**Interpretation**: Engagement increases significantly across the year, with each quarter higher than the previous. This suggests successful HR interventions or improving work conditions.

---
