---
title: "Lab Assignment 4: Crosstabulation and Correlation (Week 08)"
subtitle: "MBA642: Business Analytics in Healthcare Management"
author: "Your Name"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: false
    code-copy: true
    theme:
      - cosmo
      - lecture.scss
    highlight-style: github
    mainfont: "Clear Sans, -apple-system, system-ui, sans-serif"
    fontsize: 1.1em
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Assignment Overview

- **Total Points**: 10 points
- **Due Date**: March 8, 2026
- **Submission**: Upload your completed .qmd (no need to submit the rendered HTML file)

## Learning Objectives

This assignment assesses your ability to:

- Construct and interpret contingency tables (crosstabulation)
- Conduct a chi-square test and interpret results
- Calculate and interpret Pearson and Spearman correlation coefficients
- Create appropriate visualizations for bivariate relationships
- Choose the correct statistical method based on variable types

## Instructions

1. Complete all 3 tasks below
2. Fill in the blanks (replace `___` with appropriate code or values)
3. After completing each code chunk, change `#| eval: false` to `#| eval: true` to run it
4. Answer all interpretation questions (1-2 sentences each)
5. Run all code chunks to verify they work before submitting

---

# Grading Rubric

| Task | Points | Breakdown |
|------|--------|-----------|
| Task 1: Crosstabulation | 4 | 1 coding + 1 percentages + 1 visualization + 1 interpretation |
| Task 2: Pearson Correlation | 3 | 1 visualization + 1 coding + 1 interpretation |
| Task 3: Spearman Correlation | 3 | 1 visualization + 1 coding + 1 interpretation |
| **Total** | **10** | |

**Interpretation Guidance**: Your interpretations should be brief but specific. For example:

- **Good**: "The chi-square test shows a significant association between insurance type and readmission (χ² = 8.52, p = 0.014). Medicaid patients have the highest readmission rate (30%), suggesting they may need enhanced discharge planning."
- **Too vague**: "The result is significant."

---

# Required Packages

- Load the required packages using the code below:

```{r}
#| label: required-packages

library(ggplot2)
library(dplyr)
```




# Assignment Tasks

## Task 1: Crosstabulation — Discharge Disposition and Readmission (4 points) {.unnumbered}

A hospital quality team wants to know if there is an association between **discharge disposition** (Home, Skilled Nursing Facility, Rehabilitation Center) and **30-day readmission** (Yes, No). Understanding this relationship helps target post-discharge interventions.

```{r}
#| label: task1-data

# Data Loading
discharge_data <- read.csv("week08_assignment_data1.csv")
head(discharge_data)
```

**Step 1 (Coding — 1 point)**: Create a contingency table and run a chi-square test.

```{r}
#| label: task1-chisq
#| eval: false

# Create contingency table
discharge_table <- table(discharge_data$___, discharge_data$___)
discharge_table

# Run chi-square test
chi_discharge <- chisq.test(___)

# Check expected frequencies
chi_discharge$___

# Display test results
chi_discharge
```

**Step 2 (Percentages — 1 point)**: Calculate row percentages to see readmission rates by disposition.

```{r}
#| label: task1-pct
#| eval: false

row_pct <- prop.table(___, margin = ___) * ___
row_pct
```

**Step 3 (Visualization — 1 point)**: Create a grouped bar chart of readmission rates by disposition. (Hint: Use the `row_pct` result from Step 2)

```{r}
#| label: task1-plot
#| eval: false

plot_data <- as.data.frame(___)
colnames(plot_data) <- c("disposition", "readmission", "percentage")

ggplot(plot_data, aes(x = ___, y = ___, fill = ___)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8, width = 0.4) +
  geom_text(aes(label = sprintf("%.1f%%", percentage)),
            position = position_dodge(width = 0.4),
            vjust = -0.5, size = 3.5) +
  labs(title = "Readmission Rates by Discharge Disposition",
       x = "Discharge Disposition",
       y = "Percentage of Patients") +
  theme_minimal() +
  theme(legend.position = "top")
```

**Step 4 (Interpretation — 1 point)**:

1. Are all expected frequencies ≥ 5? Is the chi-square test appropriate for this data?
2. Is the association between discharge disposition and readmission statistically significant? Report the χ² value, df, and p-value.
3. Which disposition type has the highest and lowest readmission rates? What does this suggest for care transition planning?

::: {.callout-note icon=false title="Your Answer (1-2 sentences each)"}
*[Write your answers here]*

1. Question 1:


2. Question 2:


3. Question 3:


:::

---

## Task 2: Pearson Correlation — Length of Stay and Total Charges (3 points) {.unnumbered}

A hospital finance team wants to understand the relationship between **length of stay** (LOS, in days) and **total charges** (in thousands of dollars). Both variables are continuous, so Pearson correlation is appropriate. This analysis helps with budgeting and resource planning.

```{r}
#| label: task2-data

# Data Loading
finance_data <- read.csv("week08_assignment_data2.csv")
head(finance_data)
```

**Step 1 (Visualization — 1 point)**: Create a scatterplot with a trend line.

```{r}
#| label: task2-plot
#| eval: false

ggplot(finance_data, aes(x = ___, y = ___)) +
  geom_point(alpha = 0.7, size = 3, color = "steelblue") +
  geom_smooth(method = "___", se = TRUE, color = "red") +
  labs(title = "Length of Stay vs. Total Charges",
       x = "Length of Stay (days)",
       y = "Total Charges ($1,000s)") +
  theme_minimal()
```

**Step 2 (Coding — 1 point)**: Run a Pearson correlation test.

```{r}
#| label: task2-pearson
#| eval: false

with(finance_data, cor.test(___, ___, method = "___"))
```

**Step 3 (Interpretation — 1 point)**:

1. What is the Pearson correlation coefficient (r)? What is the direction of the association, and how would you classify its strength (very weak, weak, moderate, strong, very strong)?
2. Is the correlation statistically significant? Report the t-statistic and p-value.
3. What does this relationship suggest for hospital financial planning?

::: {.callout-note icon=false title="Your Answer (1-2 sentences each)"}
*[Write your answers here]*

1. Question 1:


2. Question 2:


3. Question 3:


:::

---

## Task 3: Spearman Correlation — Triage Acuity and Time to Treatment (3 points) {.unnumbered}

An emergency department manager wants to examine the relationship between **triage acuity level** (severity of illness) and **time to treatment** (the number of minutes from arrival to first clinical intervention). Triage acuity is measured using the Emergency Severity Index (ESI), a 5-level system used in emergency departments to prioritize patients: Level 1 (most acute, life-threatening) through Level 5 (non-urgent, least acute). Because ESI is an ordinal scale (the levels indicate rank order but the intervals between them are not necessarily equal), Spearman correlation is appropriate.

**ESI Scale**: 1 = Resuscitation, 2 = Emergent, 3 = Urgent, 4 = Less Urgent, 5 = Non-urgent

```{r}
#| label: task3-data

# Data Loading
ed_triage_data <- read.csv("week08_assignment_data3.csv")
head(ed_triage_data)
```

**Step 1 (Visualization — 1 point)**: Create a scatterplot (use jitter since acuity is discrete).

```{r}
#| label: task3-plot
#| eval: false

ggplot(ed_triage_data, aes(x = ___, y = ___)) +
  geom_point(alpha = 0.6, size = 3,
             position = position_jitter(width = 0.15, height = 0)) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  scale_x_continuous(breaks = 1:5) +
  labs(title = "Time to Treatment by Triage Acuity Level",
       x = "Triage Acuity Level (1 = Most Acute, 5 = Non-urgent)",
       y = "Time to Treatment (minutes)") +
  theme_minimal()
```

**Step 2 (Coding — 1 point)**: Run a Spearman correlation test.

```{r}
#| label: task3-spearman
#| eval: false

with(ed_triage_data, cor.test(___, ___, method = "___"))
```

**Step 3 (Interpretation — 1 point)**:

1. What is the Spearman correlation coefficient (ρ)? Is it positive or negative? Explain why the direction makes clinical sense.
2. Is the correlation statistically significant? Report the p-value.
3. Why is Spearman more appropriate than Pearson for this data? (Hint: think about the measurement scale of acuity)

::: {.callout-note icon=false title="Your Answer (1-2 sentences each)"}
*[Write your answers here]*

1. Question 1:


2. Question 2:


3. Question 3:


:::

---

# Submission Checklist

Before submitting, verify:

- [ ] All 3 tasks are completed
- [ ] All blanks (`___`) are filled in with appropriate values
- [ ] All `#| eval: false` changed to `#| eval: true`
- [ ] All code chunks run without errors
- [ ] All interpretation questions are answered (1-2 sentences each)
- [ ] Visualizations have appropriate titles and axis labels
- [ ] File renders to HTML successfully

---

# Points Summary

| Task | Points | Topic | Method |
|------|--------|-------|--------|
| Task 1 | 4.0 | Discharge disposition & readmission | Crosstabulation (Chi-Square) |
| Task 2 | 3.0 | Length of stay & total charges | Pearson Correlation |
| Task 3 | 3.0 | Triage acuity & time to treatment | Spearman Correlation |
| **Total** | **10.0** | | |

---

**Good luck!**
