---
title: "Crosstabulation and Correlation (Practice Activities)"
subtitle: "Business Analytics in Healthcare Management"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
    theme: cosmo
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrplot)

# Set seed for reproducibility
set.seed(2026)
```

# Overview

In this practice session, you will apply the crosstabulation and correlation techniques learned in the lecture. This is a learning activity designed to help you gain hands-on experience with bivariate analysis methods.

**What you'll practice:**

- Creating contingency tables and calculating percentages
- Conducting chi-square tests and Fisher's exact tests
- Interpreting chi-square test results
- Creating scatterplots to visualize relationships
- Calculating and interpreting Pearson correlation
- Calculating and interpreting Spearman correlation
- Understanding when to use each method

**Format**: Fill-in-the-blank code sections (marked with `___`)

**Instructions**:

1. Complete the code by replacing `___` with appropriate values
2. Run each chunk to verify your analysis
3. Answer the interpretation questions
4. Compare your results with the solutions provided

**Note**: This practice is completion-based. The goal is to learn and experiment—mistakes are part of the process!

---

# Recap: Key Concepts

Before starting the activities, let's review essential concepts from the lecture:

## Crosstabulation (Chi-Square Test)

**Purpose**: Test whether there is a significant association between two categorical variables

**Key Functions**:
```{r}
#| label: recap-crosstab
#| eval: false

# Create contingency table
table(data$variable1, data$variable2)

# Calculate row percentages
prop.table(table(data$variable1, data$variable2), margin = 1) * 100

# Perform chi-square test
chi_result <- chisq.test(table(data$variable1, data$variable2))

# View expected frequencies
chi_result$expected

# Fisher's exact test (for small samples)
fisher.test(table(data$variable1, data$variable2))
```

**Interpretation**:
- **p < 0.05**: Variables are significantly associated (reject independence)
- **p ≥ 0.05**: No significant association (fail to reject independence)
- **Expected frequencies**: All should be ≥ 5 for chi-square; use Fisher's exact if < 5

## Correlation Analysis

**Purpose**: Test whether there is a significant linear (Pearson) or monotonic (Spearman) relationship between two continuous or ordinal variables

**Key Functions**:
```{r}
#| label: recap-correlation
#| eval: false

# Pearson correlation (for continuous variables)
cor.test(data$variable1, data$variable2, method = "pearson")

# Spearman correlation (for ordinal variables or when outliers present)
cor.test(data$variable1, data$variable2, method = "spearman", exact = FALSE)

# Correlation matrix
cor(data_subset, method = "pearson")
```

**Interpretation**:
- **Direction**: Positive (+) or negative (-)
- **Strength**: 0.0-0.19 (very weak), 0.20-0.39 (weak), 0.40-0.59 (moderate), 0.60-0.79 (strong), 0.80-1.0 (very strong)
- **Significance**: p < 0.05 indicates significant correlation

---

# Common Errors and How to Fix Them

You **WILL** encounter errors while practicing—that's completely normal! Here are common issues:

## Error 1: "object not found"

```r
Error: object 'surgery_data' not found
```

**Cause**: You didn't run the chunk that creates the dataset

**Fix**: Scroll up to find the dataset generation chunk and run it first

## Error 2: Incorrect table() syntax

```r
# ❌ Wrong:
table(surgery_data)  # This shows all variables

# ✅ Right:
table(surgery_data$approach, surgery_data$complications)  # Specify TWO variables
```

## Error 3: Forgot prop.table() margin parameter

```r
# ❌ Wrong:
prop.table(table(data$var1, data$var2))  # Shows overall proportions

# ✅ Right:
prop.table(table(data$var1, data$var2), margin = 1) * 100  # Row percentages
# margin = 1 means row percentages
# margin = 2 means column percentages
```

## Error 4: Using wrong correlation method

```r
# ❌ Wrong for ordinal data:
cor.test(ordinal_var1, ordinal_var2, method = "pearson")

# ✅ Right for ordinal data:
cor.test(ordinal_var1, ordinal_var2, method = "spearman", exact = FALSE)
```

## Error 5: Plotting categorical data incorrectly

For crosstabulation visualization, you typically need to create summary data first:

```r
# Create summary for plotting
plot_data <- data %>%
  group_by(var1, var2) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(var1) %>%
  mutate(percentage = count / sum(count) * 100)
```

---

# Practice Activity 1: Treatment Protocol and Complications

## Scenario

A surgical department wants to know if there's an association between the **surgical approach** (Minimally Invasive vs. Open) and whether patients experienced **post-operative complications** (Yes vs. No).

**Research Question**: Is there a statistically significant association between surgical approach and post-operative complications?

## Dataset Generation

Run this code to generate the surgical outcomes dataset:

```{r}
#| label: practice-crosstab-data

# Practice dataset
set.seed(2026)
n_surgery <- 200

surgical_approach <- sample(c("Minimally Invasive", "Open"), n_surgery,
                            replace = TRUE, prob = c(0.6, 0.4))

# Complications with different rates by approach
complications <- character(n_surgery)
for(i in 1:n_surgery) {
  if(surgical_approach[i] == "Minimally Invasive") {
    complications[i] <- sample(c("Yes", "No"), 1, prob = c(0.05, 0.95))
  } else {
    complications[i] <- sample(c("Yes", "No"), 1, prob = c(0.20, 0.80))
  }
}

surgery_data <- data.frame(
  approach = factor(surgical_approach),
  complications = factor(complications, levels = c("Yes", "No"))
)

# View the first few rows
head(surgery_data, 10)
```

## Task 1.1: Create a Contingency Table

Create a contingency table showing the counts of complications by surgical approach.

**Hint**: Use `table()` with two variables: `surgery_data$approach` and `surgery_data$complications`

```{r}
#| label: task-1-1
#| eval: false

# Fill in the blanks
___(surgery_data$___, surgery_data$___)
```

**Question**: Looking at the raw counts, which surgical approach appears to have more complications?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.2: Calculate Row Percentages

Calculate the percentage of patients with complications for each surgical approach (row percentages).

**Hint**: Use `prop.table()` with `margin = 1` for row percentages, then multiply by 100

```{r}
#| label: task-1-2
#| eval: false

# Fill in the blanks
prop.table(table(surgery_data$approach, surgery_data$complications),
           margin = ___) * ___
```

**Question**: What percentage of patients had complications in each surgical approach group?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.3: Check Expected Frequencies

Before conducting a chi-square test, check that all expected frequencies are ≥ 5.

**Hint**: First run `chisq.test()` and store it, then access `$expected`

```{r}
#| label: task-1-3
#| eval: false

# Fill in the blanks
chi_surgery <- chisq.test(table(surgery_data$___, surgery_data$___))

# View expected frequencies
chi_surgery$___
```

**Question**: Are all expected frequencies ≥ 5? Should we use chi-square test or Fisher's exact test?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.4: Conduct the Chi-Square Test

Perform the chi-square test and interpret the results.

**Hint**: You already created `chi_surgery` in Task 1.3, just display it

```{r}
#| label: task-1-4
#| eval: false

# Display the chi-square test results
___
```

**Question**: Based on the p-value, is there a statistically significant association between surgical approach and complications? What do you conclude?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.5: Visualize the Association (Optional)

Create a grouped bar chart showing the percentage of complications by surgical approach.

**Hint**: First create summary data with `group_by()` and `summarise()`, then use `ggplot()` with `geom_bar()`

```{r}
#| label: task-1-5
#| eval: false

# Create summary data
plot_data <- surgery_data %>%
  group_by(___, ___) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(approach) %>%
  mutate(percentage = count / sum(count) * ___)

# Create visualization
ggplot(plot_data, aes(x = ___, y = ___, fill = ___)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  labs(title = "___",
       x = "Surgical Approach",
       y = "Percentage of Patients") +
  theme_minimal()
```

---

# Practice Activity 2: ED Wait Time and Satisfaction

## Scenario

An emergency department manager wants to examine the relationship between **wait time** (in minutes) and **patient satisfaction score** (0-100).

**Research Question**: Is there a statistically significant correlation between ED wait time and patient satisfaction scores?

## Dataset Generation

```{r}
#| label: practice-correlation-data

# Practice dataset
set.seed(2026)
n_ed <- 60

wait_time <- round(runif(n_ed, 10, 120))
# Satisfaction decreases with wait time
satisfaction_ed <- round(90 - 0.4 * wait_time + rnorm(n_ed, 0, 10))
satisfaction_ed <- pmin(pmax(satisfaction_ed, 0), 100)

ed_data <- data.frame(
  wait_time_minutes = wait_time,
  satisfaction_score = satisfaction_ed
)

head(ed_data, 10)
```

## Task 2.1: Examine the Data

Get a quick summary of both variables.

**Hint**: Use `summary()` function

```{r}
#| label: task-2-1
#| eval: false

# Fill in the blanks
___(ed_data)
```

**Question**: What is the range of wait times? What is the mean satisfaction score?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 2.2: Create a Scatterplot

Create a scatterplot showing the relationship between wait time (x-axis) and satisfaction (y-axis).

**Hint**: Use `ggplot()` with `geom_point()` and `geom_smooth(method = "lm")`

```{r}
#| label: task-2-2
#| eval: false

# Fill in the blanks
ggplot(ed_data, aes(x = ___, y = ___)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "___", se = TRUE, color = "red") +
  labs(title = "___",
       x = "___",
       y = "___") +
  theme_minimal()
```

**Question**: Based on the scatterplot, does there appear to be a relationship? Is it positive or negative?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 2.3: Calculate Pearson Correlation

Calculate the Pearson correlation coefficient and test its significance.

**Hint**: Use `cor.test()` with `method = "pearson"`

```{r}
#| label: task-2-3
#| eval: false

# Fill in the blanks
cor.test(ed_data$___, ed_data$___, method = "___")
```

**Question**: What is the correlation coefficient? Is it statistically significant? Interpret the direction and strength.

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

Correlation coefficient (r):
Direction (positive/negative):
Strength (weak/moderate/strong):
p-value:
Significant? (yes/no):

</div>

## Task 2.4: Practical Implications

**Question**: What are the practical implications for ED management? If the hospital wants to improve patient satisfaction, what does this analysis suggest?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

# Practice Activity 3: Nursing Workload and Job Satisfaction

## Scenario

A nursing director wants to examine the relationship between **perceived workload** (measured on a 1-5 ordinal scale) and **job satisfaction** (also measured on a 1-5 ordinal scale) among nurses.

**Workload Scale**: 1 = Very light, 2 = Manageable, 3 = Moderate, 4 = Heavy, 5 = Overwhelming

**Job Satisfaction Scale**: 1 = Very dissatisfied, 2 = Dissatisfied, 3 = Neutral, 4 = Satisfied, 5 = Very satisfied

**Research Question**: Is there a significant monotonic relationship between workload and job satisfaction?

## Dataset Generation

```{r}
#| label: practice-spearman-data

# Practice dataset (n = 50 nurses)
set.seed(2027)
n_nurses_practice <- 50

# Generate ordinal workload ratings (1-5)
workload_rating <- sample(1:5, n_nurses_practice, replace = TRUE,
                          prob = c(0.05, 0.15, 0.35, 0.30, 0.15))

# Job satisfaction tends to decrease with workload (negative relationship)
satisfaction_base <- 6 - workload_rating + sample(-1:1, n_nurses_practice, replace = TRUE)
job_satisfaction <- pmin(pmax(satisfaction_base, 1), 5)

workload_data <- data.frame(
  nurse_id = 1:n_nurses_practice,
  workload = workload_rating,
  satisfaction = job_satisfaction
)

head(workload_data, 10)
```

## Task 3.1: Create a Scatterplot with Jitter

Since both variables are ordinal (discrete values 1-5), we need to add jitter to see overlapping points.

**Hint**: Use `position = position_jitter()` inside `geom_point()`

```{r}
#| label: task-3-1
#| eval: false

# Fill in the blanks
ggplot(workload_data, aes(x = ___, y = ___)) +
  geom_point(alpha = 0.6, size = 3,
             position = position_jitter(width = 0.15, height = 0.15)) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  scale_x_continuous(breaks = 1:5) +
  scale_y_continuous(breaks = 1:5) +
  labs(title = "___",
       x = "Perceived Workload (1 = Light, 5 = Overwhelming)",
       y = "___") +
  theme_minimal()
```

**Question**: Based on the scatterplot, what relationship do you observe?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 3.2: Why Use Spearman Instead of Pearson?

**Question**: Why is Spearman correlation more appropriate than Pearson correlation for this data?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer** (Hint: Think about the measurement scales):

</div>

## Task 3.3: Calculate Spearman Correlation

Calculate the Spearman rank correlation coefficient.

**Hint**: Use `cor.test()` with `method = "spearman"` and `exact = FALSE`

```{r}
#| label: task-3-3
#| eval: false

# Fill in the blanks
cor.test(workload_data$___, workload_data$___,
         method = "___", exact = ___)
```

**Question**: What is the Spearman correlation coefficient (rho)? Is it statistically significant? Interpret the results.

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

Spearman's rho (ρ):
Direction (positive/negative):
Strength (weak/moderate/strong):
p-value:
Significant? (yes/no):

</div>

## Task 3.4: Compare Pearson and Spearman (Optional)

Run Pearson correlation on the same data and compare results.

```{r}
#| label: task-3-4
#| eval: false

# Pearson correlation
cor.test(workload_data$workload, workload_data$satisfaction,
         method = "pearson")
```

**Question**: How do the Pearson and Spearman coefficients compare? Which one should you report for this analysis and why?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

# Practice Activity 4: Discharge Destination and Readmission

## Scenario

A hospital discharge planning team wants to examine whether there is an association between **discharge destination** (Home, Skilled Nursing Facility, Rehabilitation) and **30-day readmission** (Yes vs. No).

**Research Question**: Is discharge destination associated with 30-day readmission rates?

## Dataset Generation

```{r}
#| label: practice-discharge-data

# Discharge destination and readmission data
set.seed(2026)
n_discharge <- 250

discharge_dest <- sample(c("Home", "SNF", "Rehab"), n_discharge,
                         replace = TRUE, prob = c(0.5, 0.3, 0.2))

readmit <- character(n_discharge)
for(i in 1:n_discharge) {
  prob_readmit <- switch(discharge_dest[i],
                         "Home" = 0.12,
                         "SNF" = 0.22,
                         "Rehab" = 0.08)
  readmit[i] <- sample(c("Yes", "No"), 1, prob = c(prob_readmit, 1 - prob_readmit))
}

discharge_data <- data.frame(
  discharge_destination = factor(discharge_dest, levels = c("Home", "SNF", "Rehab")),
  readmitted_30_days = factor(readmit, levels = c("Yes", "No"))
)

head(discharge_data, 10)
```

## Task 4.1: Create Contingency Table and Calculate Percentages

```{r}
#| label: task-4-1
#| eval: false

# Contingency table
table(___, ___)

# Row percentages
prop.table(table(discharge_data$discharge_destination,
                 discharge_data$readmitted_30_days),
           margin = ___) * ___
```

**Question**: Which discharge destination has the highest readmission rate? Which has the lowest?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 4.2: Conduct Chi-Square Test

```{r}
#| label: task-4-2
#| eval: false

# Chi-square test
chi_discharge <- chisq.test(table(discharge_data$___,
                                   discharge_data$___))

# Check expected frequencies
chi_discharge$___

# Display results
chi_discharge
```

**Question**: Are all expected frequencies ≥ 5? Is there a significant association between discharge destination and readmission?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

# Practice Activity 5: Hospital Quality Metrics Correlation Matrix

## Scenario

A health system analyst wants to explore the relationships among multiple hospital quality metrics across 100 hospitals.

**Variables**:
- Bed size
- Average length of stay (days)
- Mortality rate (%)
- Patient satisfaction score (0-100)
- Nurse retention rate (%)

**Research Question**: How are these quality metrics related to each other?

## Dataset Generation

```{r}
#| label: practice-metrics-data

# Hospital quality metrics
set.seed(2026)
n_hosp <- 100

quality_data <- data.frame(
  hospital_id = 1:n_hosp,
  bed_size = sample(c(50, 100, 200, 300, 500), n_hosp, replace = TRUE),
  avg_los = round(runif(n_hosp, 3, 7), 1),
  mortality_rate = round(runif(n_hosp, 1, 5), 2),
  pt_satisfaction = round(runif(n_hosp, 60, 95), 1),
  nurse_retention = round(runif(n_hosp, 70, 98), 1)
)

# Add realistic correlation: satisfaction related to retention
quality_data$pt_satisfaction <- round(quality_data$pt_satisfaction +
                                       0.1 * quality_data$nurse_retention +
                                       rnorm(n_hosp, 0, 5), 1)
quality_data$pt_satisfaction <- pmin(pmax(quality_data$pt_satisfaction, 0), 100)

head(quality_data)
```

## Task 5.1: Calculate Correlation Matrix

Calculate the Pearson correlation matrix for all numeric variables except hospital_id.

**Hint**: Select columns 2 through 6, then use `cor()`

```{r}
#| label: task-5-1
#| eval: false

# Select only the numeric quality metrics (columns 2-6)
metrics_only <- quality_data[, ___:___]

# Calculate correlation matrix
cor_matrix <- cor(metrics_only, method = "___")

# Display rounded to 2 decimal places
round(cor_matrix, ___)
```

**Question**: Which pair of variables has the strongest positive correlation? The strongest negative correlation?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 5.2: Visualize Correlation Matrix

Create a correlation matrix heatmap using the `corrplot` package.

**Hint**: Use `corrplot()` with `method = "color"`

```{r}
#| label: task-5-2
#| eval: false

# Load corrplot if not already loaded
library(corrplot)

# Visualize correlation matrix
corrplot(___,
         method = "___",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45,
         title = "Correlation Matrix: Hospital Quality Metrics",
         mar = c(0, 0, 2, 0))
```

## Task 5.3: Test Specific Correlation

Test whether the correlation between patient satisfaction and nurse retention is statistically significant.

```{r}
#| label: task-5-3
#| eval: false

# Fill in the blanks
cor.test(quality_data$___, quality_data$___,
         method = "pearson")
```

**Question**: Is the correlation between patient satisfaction and nurse retention statistically significant? What might this suggest for hospital management?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

# Summary and Reflection

## Key Takeaways

After completing this practice activity, you should be able to:

- ✓ Create contingency tables and calculate row/column percentages
- ✓ Conduct chi-square tests and interpret results
- ✓ Determine when to use Fisher's exact test vs. chi-square test
- ✓ Create scatterplots to visualize relationships
- ✓ Calculate and interpret Pearson correlation coefficients
- ✓ Calculate and interpret Spearman correlation coefficients
- ✓ Understand when to use Pearson vs. Spearman
- ✓ Create and interpret correlation matrices

## Decision Guide

| **Research Question** | **Variable Types** | **Method** |
|----------------------|-------------------|-----------|
| Are two categorical variables associated? | Both categorical | Chi-square test (or Fisher's exact if expected < 5) |
| Is there a linear relationship between two continuous variables? | Both continuous | Pearson correlation |
| Is there a monotonic relationship with ordinal data? | Ordinal or continuous with outliers | Spearman correlation |
| How are multiple variables related? | Multiple continuous | Correlation matrix |

---

# Activity Solutions

Solutions for these activities are provided below. Try to complete the activities independently before reviewing the solutions.

## Practice Activity 1 Solutions

### Task 1.1 Solution

```{r}
#| label: solution-1-1
#| code-fold: true

table(surgery_data$approach, surgery_data$complications)
```

### Task 1.2 Solution

```{r}
#| label: solution-1-2
#| code-fold: true

prop.table(table(surgery_data$approach, surgery_data$complications),
           margin = 1) * 100
```

Minimally invasive surgery shows lower complication rates (~5-6%) compared to open surgery (~18-22%).

### Task 1.3 Solution

```{r}
#| label: solution-1-3
#| code-fold: true

chi_surgery <- chisq.test(table(surgery_data$approach, surgery_data$complications))
chi_surgery$expected
```

All expected frequencies are greater than 5, so chi-square test is appropriate.

### Task 1.4 Solution

```{r}
#| label: solution-1-4
#| code-fold: true

chi_surgery
```

With p < 0.05, there is a statistically significant association between surgical approach and complications. Minimally invasive surgery is associated with lower complication rates.

### Task 1.5 Solution

```{r}
#| label: solution-1-5
#| code-fold: true
#| fig-width: 8
#| fig-height: 5

plot_data <- surgery_data %>%
  group_by(approach, complications) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(approach) %>%
  mutate(percentage = count / sum(count) * 100)

ggplot(plot_data, aes(x = approach, y = percentage, fill = complications)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  geom_text(aes(label = sprintf("%.1f%%", percentage)),
            position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Post-Operative Complications by Surgical Approach",
       x = "Surgical Approach",
       y = "Percentage of Patients",
       fill = "Complications") +
  theme_minimal() +
  ylim(0, 110)
```

---

## Practice Activity 2 Solutions

### Task 2.1 Solution

```{r}
#| label: solution-2-1
#| code-fold: true

summary(ed_data)
```

### Task 2.2 Solution

```{r}
#| label: solution-2-2
#| code-fold: true
#| fig-width: 7
#| fig-height: 5

ggplot(ed_data, aes(x = wait_time_minutes, y = satisfaction_score)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Relationship Between ED Wait Time and Patient Satisfaction",
       x = "Wait Time (minutes)",
       y = "Satisfaction Score (0-100)") +
  theme_minimal()
```

There appears to be a negative relationship: as wait time increases, satisfaction decreases.

### Task 2.3 Solution

```{r}
#| label: solution-2-3
#| code-fold: true

cor.test(ed_data$wait_time_minutes, ed_data$satisfaction_score, method = "pearson")
```

The correlation is negative and statistically significant (p < 0.05), indicating that longer wait times are associated with lower satisfaction scores. The strength is typically moderate to strong.

### Task 2.4 Solution

**Practical implications**: The significant negative correlation suggests that reducing ED wait times could improve patient satisfaction scores. Hospital management should consider:
- Increasing staffing during peak hours
- Implementing fast-track systems for less acute cases
- Improving patient flow processes
- Setting target wait times and monitoring performance

---

## Practice Activity 3 Solutions

### Task 3.1 Solution

```{r}
#| label: solution-3-1
#| code-fold: true
#| fig-width: 7
#| fig-height: 5

ggplot(workload_data, aes(x = workload, y = satisfaction)) +
  geom_point(alpha = 0.6, size = 3,
             position = position_jitter(width = 0.15, height = 0.15)) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  scale_x_continuous(breaks = 1:5) +
  scale_y_continuous(breaks = 1:5) +
  labs(title = "Nursing Workload vs. Job Satisfaction",
       x = "Perceived Workload (1 = Light, 5 = Overwhelming)",
       y = "Job Satisfaction (1 = Very Dissatisfied, 5 = Very Satisfied)") +
  theme_minimal()
```

### Task 3.2 Solution

Spearman correlation is more appropriate because:
1. Both variables are measured on ordinal scales (1-5 ratings)
2. The intervals between scale points may not be equal (the difference between 1 and 2 may not be the same as between 4 and 5)
3. Spearman uses ranks, making it appropriate for ordinal data

### Task 3.3 Solution

```{r}
#| label: solution-3-3
#| code-fold: true

cor.test(workload_data$workload, workload_data$satisfaction,
         method = "spearman", exact = FALSE)
```

The Spearman correlation is negative and statistically significant, indicating that higher workload is associated with lower job satisfaction among nurses.

### Task 3.4 Solution

```{r}
#| label: solution-3-4
#| code-fold: true

# Pearson correlation for comparison
cor.test(workload_data$workload, workload_data$satisfaction,
         method = "pearson")
```

Both Pearson and Spearman show similar negative correlations, but we should report **Spearman** because our data are ordinal. Spearman is the appropriate method when variables are measured on ordinal scales.

---

## Practice Activity 4 Solutions

### Task 4.1 Solution

```{r}
#| label: solution-4-1
#| code-fold: true

# Contingency table
table(discharge_data$discharge_destination, discharge_data$readmitted_30_days)

# Row percentages
prop.table(table(discharge_data$discharge_destination,
                 discharge_data$readmitted_30_days),
           margin = 1) * 100
```

SNF (Skilled Nursing Facility) patients have the highest readmission rate (~22%), while Rehab patients have the lowest (~8%).

### Task 4.2 Solution

```{r}
#| label: solution-4-2
#| code-fold: true

chi_discharge <- chisq.test(table(discharge_data$discharge_destination,
                                   discharge_data$readmitted_30_days))

# Check expected frequencies
chi_discharge$expected

# Display results
chi_discharge
```

All expected frequencies are ≥ 5, so chi-square is appropriate. The significant p-value (p < 0.05) indicates that discharge destination is significantly associated with 30-day readmission rates.

---

## Practice Activity 5 Solutions

### Task 5.1 Solution

```{r}
#| label: solution-5-1
#| code-fold: true

# Select metrics (columns 2-6)
metrics_only <- quality_data[, 2:6]

# Calculate correlation matrix
cor_matrix <- cor(metrics_only, method = "pearson")

# Display rounded
round(cor_matrix, 2)
```

### Task 5.2 Solution

```{r}
#| label: solution-5-2
#| code-fold: true
#| fig-width: 8
#| fig-height: 7

corrplot(cor_matrix,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45,
         title = "Correlation Matrix: Hospital Quality Metrics",
         mar = c(0, 0, 2, 0))
```

### Task 5.3 Solution

```{r}
#| label: solution-5-3
#| code-fold: true

cor.test(quality_data$pt_satisfaction, quality_data$nurse_retention,
         method = "pearson")
```

The positive correlation between patient satisfaction and nurse retention is statistically significant (p < 0.05). This suggests that hospitals with better nurse retention tend to have higher patient satisfaction scores, likely because:
- Experienced, stable nursing staff provide better care
- Lower turnover means better team coordination
- Satisfied nurses may provide better patient experiences

---

# Final Reflection

::: {.callout-note}
## Congratulations!

You've completed all five practice activities on crosstabulation and correlation analysis. You should now feel confident:

- Choosing the appropriate bivariate analysis method based on variable types
- Conducting and interpreting chi-square tests and Fisher's exact tests
- Calculating and interpreting Pearson and Spearman correlations
- Visualizing relationships between variables
- Understanding the assumptions and limitations of each method

Remember: **Correlation does not imply causation!** Always consider alternative explanations and potential confounding variables when interpreting your results.
:::
