---
title: "Basic Statistical Concepts (Lecture Notes)"
subtitle: ""
author: ""
Week: "Week 3"
date: 01/01/2026
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: false
    code-copy: true
    theme: cosmo
    highlight-style: github
    mainfont: "Clear Sans, -apple-system, system-ui, sans-serif"
    fontsize: 1.1em
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(tidyverse)
library(knitr)
library(kableExtra)

# Set seed for reproducibility
set.seed(2026)

# Create healthcare datasets for the course
# Patient satisfaction data
n_patient1 <- 1000
patient_satisfaction <- tibble(
  patient_id = 1:n_patient1,
  hospital = sample(c("General Hospital", "University Medical", "Community Health"), n_patient1, replace = TRUE),
  department = sample(c("Emergency", "Surgery", "Internal Medicine", "Pediatrics"), n_patient1, replace = TRUE),
  age = round(rnorm(n_patient1, mean = 55, sd = 18)),
  gender = sample(c("Male", "Female"), n_patient1, replace = TRUE),
  insurance_type = sample(c("Private", "Medicare", "Medicaid", "Uninsured"), n_patient1,
                          prob = c(0.4, 0.35, 0.15, 0.1), replace = TRUE),
  wait_time_minutes = round(rgamma(n_patient1, shape = 2, rate = 0.05)),
  satisfaction_score = round(pmin(7, pmax(1, rnorm(n_patient1, mean = 5.2, sd = 1.3)))),
  would_recommend = sample(1:7, n_patient1, replace = TRUE,
                           prob = c(0.05, 0.05, 0.1, 0.15, 0.25, 0.25, 0.15)),
  length_of_stay = round(rgamma(n_patient1, shape = 1.5, rate = 0.3), 1),
  total_charges = round(rlnorm(n_patient1, meanlog = 9, sdlog = 1))
)

# Readmission data
readmission_data <- tibble(
  patient_id = 1:500,
  age_group = sample(c("18-44", "45-64", "65-74", "75+"), 500,
                     prob = c(0.15, 0.30, 0.30, 0.25), replace = TRUE),
  diagnosis = sample(c("Heart Failure", "Pneumonia", "COPD", "AMI", "Hip Replacement"),
                     500, replace = TRUE),
  readmitted_30day = sample(c("Yes", "No"), 500, prob = c(0.15, 0.85), replace = TRUE),
  pain_score = sample(0:10, 500, replace = TRUE),
  comorbidity_count = rpois(500, lambda = 2)
)

# Additional data for distribution shape examples
# Normal distribution: Systolic blood pressure (larger sample for smoother normal curve)
bp_data <- tibble(
  patient_id = 1:800,
  systolic_bp = round(rnorm(800, mean = 120, sd = 12))
)

# Left-skewed distribution: Patient satisfaction (skewed toward high scores)
high_satisfaction <- tibble(
  patient_id = 1:250,
  satisfaction_rating = round(100 - rgamma(250, shape = 3, rate = 0.2))
)

# Bimodal distribution: ED wait times by shift (more prominent bimodal pattern)
ed_wait_times <- tibble(
  patient_id = 1:500,
  shift = sample(c("Day", "Night"), 500, replace = TRUE, prob = c(0.5, 0.5)),
  wait_minutes = ifelse(shift == "Day",
                       rnorm(500, mean = 22, sd = 5),  # Day shift: faster, tighter
                       rnorm(500, mean = 55, sd = 6))  # Night shift: slower, tighter
) %>%
  mutate(wait_minutes = round(pmax(5, wait_minutes)))  # Minimum 5 minutes
```

# Introduction

## Why Statistical Concepts Matter in Healthcare Management

As a healthcare manager, you make decisions daily that affect:

- **Patient outcomes**: Quality of care, safety, satisfaction
- **Operational efficiency**: Staffing, resource allocation, workflow
- **Financial performance**: Cost management, budget planning
- **Strategic planning**: Service line development, capacity planning

Understanding basic statistical concepts helps you:

1. **Describe** your data accurately (What's typical? How much variation exists?)
2. **Compare** groups meaningfully (Are these differences real or random?)
3. **Make evidence-based decisions** (What does the data actually say?)
4. **Communicate** findings clearly to diverse stakeholders

---

# Learning Objectives

By the end of this session, you will be able to:

1. **Understand** commonly used measures of central tendency (mean, median, mode) and dispersion (range, variance, standard deviation)
2. **Identify** the four scales of measurement (or data type) and select appropriate statistics for each
3. **Recognize** how data distribution shapes affect the choice of central tendency measures
4. **Understand** confidence intervals and their role in quantifying uncertainty
5. **Understand** the basic logic of hypothesis testing and what p-values mean
6. **Apply** these concepts to healthcare management decision-making

---

# Central Tendency and Dispersion Measures

When working with healthcare data, we face a common challenge: **how do we make sense of hundreds or thousands of data points?**

Imagine describing patient satisfaction by reading each score individually: "Patient 1 scored 2, Patient 2 scored 4, Patient 3 scored 6..." With 200 patients, this provides too much detail to identify meaningful patterns.

A good solution is summarizing data using two key types of measures:

- A representative value that indicates where the "center" of the data lies (Central tendency)
- Measures that describe how spread out the data is around that center (dispersion)

Together, these summaries help us understand the overall pattern efficiently and communicate findings clearly.

In this section, we'll learn the most commonly used central tendency and dispersion measures, what they tell us, and how to calculate them.

---

## Central Tendency: The "Typical" Value

Central tendency measures answer questions like:

- What is the typical length of stay for hip replacement patients?
- What satisfaction score do most patients give?
- What is the average wait time in the ED?

### The Three Measures

#### Mean (Arithmetic Average)

$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n} = \frac{x_1 + x_2 + ... + x_n}{n}$$

- The **balancing point** of the data
- Sum all values and divide by the number of observations
- Sensitive to outliers (extreme values)
- Most common measure used in general

#### Median

- The **middle value** when data is arranged in order
- If n is even, it's the average of the two middle values
- Less affected by outliers
- Better representation when data is skewed

#### Mode

- The **most frequent** value in the dataset
- Can have multiple modes (bimodal, multimodal)
- Useful for categorical data and finding the most common value

---

### Example: Length of Stay for Nine Patients

Let's calculate central tendency measures for these hospital length of stay values (in days):

**Patient LOS data**: 2, 3, 3, 4, 5, 5, 6, 8, 15 days

```{r}
#| echo: false

# Create small dataset for demonstration
los_example <- c(2, 3, 3, 4, 5, 5, 6, 8, 15)

# Calculate measures
mean_los <- mean(los_example)
median_los <- median(los_example)
mode_los <- as.numeric(names(sort(table(los_example), decreasing = TRUE)[1:2]))

# Create summary table
los_summary <- tibble(
  Measure = c("Mean", "Median", "Mode", "Min", "Max", "N"),
  Value = c(
    round(mean_los, 2),
    median_los,
    paste(mode_los, collapse = ", "),
    min(los_example),
    max(los_example),
    length(los_example)
  )
)

kable(los_summary) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Calculations**:

- **Mean**: (2 + 3 + 3 + 4 + 5 + 5 + 6 + 8 + 15) / 9 = 51 / 9 = **`r round(mean_los, 2)` days**
- **Median**: Arrange in order and find middle value (5th value) = **`r median_los` days**
- **Mode**: Most frequent values = **3 days and 5 days** (both appear twice)

**Visualization**:

```{r}
#| label: fig-los-central
#| fig-cap: "Length of Stay with Central Tendency Measures"
#| echo: false

# Create visualization
los_df <- tibble(LOS = los_example)

ggplot(los_df, aes(x = LOS)) +
  geom_dotplot(binwidth = 1, fill = "steelblue", color = "white", dotsize = 0.8) +
  geom_vline(xintercept = mean_los, color = "#E41A1C", linewidth = 1.2) +
  geom_vline(xintercept = median_los, color = "#377EB8", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = mean_los + 0.5, y = 0.95, label = paste("Mean =", round(mean_los, 2)),
           color = "#E41A1C", hjust = 0, size = 4) +
  annotate("text", x = median_los - 0.5, y = 0.85, label = paste("Median =", median_los),
           color = "#377EB8", hjust = 1, size = 4) +
  scale_x_continuous(breaks = seq(0, 16, 2)) +
  labs(title = "Length of Stay: Mean vs. Median",
       x = "Length of Stay (Days)",
       y = "") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```


---

## Dispersion: How Spread Out is the Data?

Now that we understand central tendency (where the center is), let's learn about dispersion (how spread out the data is around that center).

**Dispersion measures** answer questions like:

- How consistent are our wait times? (Predictable vs. unpredictable)
- How much do patient outcomes vary across providers?
- Is there high variability that needs investigation?

**Why Dispersion Matters**

Knowing the center isn't enough. We also need to know how much variability exists around that center.

**Example**: Two hospitals both have an average wait time of 45 minutes, but:

- **Hospital A**: Wait times consistently 40-50 minutes → **Low variability** → Predictable
- **Hospital B**: Wait times range from 10-120 minutes → **High variability** → Unpredictable

Which provides a better patient experience? Which is easier to staff?

---

### Measures of Dispersion

#### Range

$$\text{Range} = \text{Maximum} - \text{Minimum}$$

- Simplest measure
- **Very sensitive to outliers**
- Useful for quick assessment

#### Variance

$$s^2 = \frac{\sum_{i=1}^{n} (x_i - \bar{x})^2}{n}$$

- Average of **squared** deviations from the mean
- Units are squared (e.g., days²) - hard to interpret directly
- Foundation for many statistical tests
- We divide by the number of data points

#### Standard Deviation (SD)

$$s = \sqrt{s^2} = \sqrt{\frac{\sum_{i=1}^{n} (x_i - \bar{x})^2}{n}}$$

- Square root of variance
- **Same units as original data** (e.g., days)
- Most commonly reported measure
- Interpretable: "Typical distance a value deviates from the mean"

---

### Calculating Dispersion: Step-by-Step Example

Let's use the **same length of stay data** from our central tendency example:

**Patient LOS data**: 2, 3, 3, 4, 5, 5, 6, 8, 15 days

```{r}
#| echo: false

# Use same dataset as central tendency example
los_example <- c(2, 3, 3, 4, 5, 5, 6, 8, 15)

# Step-by-step calculation
mean_los_disp <- mean(los_example)
deviations <- los_example - mean_los_disp
squared_dev <- deviations^2
variance_los <- sum(squared_dev) / (length(los_example) - 1)
sd_los <- sqrt(variance_los)

calculation_table <- tibble(
  `LOS (days)` = los_example,
  `Deviation from Mean` = round(deviations, 2),
  `Squared Deviation` = round(squared_dev, 2)
)

kable(calculation_table) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Calculation Steps**:

1. **Calculate mean**: (2 + 3 + 3 + 4 + 5 + 5 + 6 + 8 + 15) / 9 = 51 / 9 = **`r round(mean_los_disp, 2)` days**
2. **Calculate deviations**: Each value minus the mean (see table above)
3. **Square the deviations**: Eliminates negative values, emphasizes larger deviations
4. **Sum and average**: Sum of squared deviations / (n-1) = **Variance** = `r round(variance_los, 2)` days²
5. **Take square root**: **Standard Deviation** = **`r round(sd_los, 2)` days**


---

# Scales of Measurement

Now that you understand tools to describe data (mean, median, mode, SD, variance, range), let's talk about **when** each tool is appropriate to use.

Not all statistics work with all types of data. Using the wrong statistic for your data type can lead to meaningless or misleading results.

---

## What are Measurement Scales

In healthcare management, we may collect many types of data: patient demographics, satisfaction scores, diagnosis codes, length of stay, costs, and more.

The same concept can be measured in different ways. For example, to measure patient satisfaction, we could:

- Ask patients to rate satisfaction on a **7-point scale** (1 = very dissatisfied, 7 = very satisfied)
- Ask patients to select one of **three categories** (low, medium, high)
- Ask a simple **yes/no question** (satisfied or not satisfied)

These are all measuring "patient satisfaction," but each approach creates a different type of data that requires different statistical methods.

**Measurement scales** (also called levels of measurement or data types) are classification systems that describe the nature of information contained in data values and the mathematical properties they possess.

---

## The Four Scales

There are four different types of measurement scale (nominal, ordinal, interval, ratio). Let's take a look at them. 

```{r}
#| label: fig-scales-hierarchy
#| fig-cap: "Hierarchy of Measurement Scales"
#| echo: false

scales_data <- tibble(
  Scale = factor(c("Nominal", "Ordinal", "Interval", "Ratio"),
                 levels = c("Nominal", "Ordinal", "Interval", "Ratio")),
  Information = c(1, 2, 3, 4),
  Label = c("Categories\nonly", "Categories +\nOrder", "Categories +\nOrder +\nEqual intervals",
            "Categories +\nOrder +\nEqual intervals +\nAbsolute zero")
)

ggplot(scales_data, aes(x = Scale, y = Information, fill = Scale)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = Label), vjust = -0.3, size = 3) +
  scale_fill_manual(values = c("#E8E8E8", "#B8D4E8", "#6BAED6", "#2171B5")) +
  labs(title = "Measurement Scales: From Less to More Information",
       y = "Amount of Information") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.y = element_blank()) +
  ylim(0, 5.5)
```

---

## Nominal Scale

**Definition**: 

Numbers or labels indicate different categories or groups with NO inherent order.

**Examples**

- **Diagnosis codes**: [ICD-10 codes](https://www.cms.gov/medicare/coordination-benefits-recovery/overview/icd-code-lists) (e.g., I5031 for Acute Heart Failure)
- **Insurance type**: Private, Medicare, Medicaid, Uninsured
- **Hospital department**: Emergency, Surgery, Internal Medicine, etc.
- **Blood type**: A, B, AB, O
- **Gender**: Male, Female, Other


**Appropriate Statistics for Nominal Data**

Among the central tendency and dispersion measures you learned earlier, which are appropriate for nominal scale data? 

- **Central tendency**: Mode (most frequent response category)
- **Dispersion**: Frequency distribution, percentages across response categories

You CANNOT calculate a meaningful mean or median for nominal data. Even if categories are coded as numbers (1, 2, 3, 4), the average is meaningless!

**Example**: If we code Private=1, Medicare=2, Medicaid=3, Uninsured=4, an "average" of 2.3 is meaningless—it doesn't represent any real insurance category!

```{r}
#| label: tbl-insurance-freq
#| tbl-cap: "Distribution of Insurance Types in Patient Sample"
#| echo: false

patient_satisfaction %>%
  count(insurance_type) %>%
  mutate(Percentage = round(n/sum(n)*100, 1)) %>%
  rename(`Insurance Type` = insurance_type, Count = n) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

## Ordinal Scale

**Definition**:

Numbers or categories indicate relative order or rank, but **intervals between values are not necessarily equal**.

**Key characteristic**: 

We know that Category A > Category B > Category C, but we DON'T know if the difference between A and B equals the difference between B and C.

**Examples**

- **Pain scale categories**: None, Mild, Moderate, Severe
- **Age groups**: 18-44, 45-64, 65-74, 75+
- **NYHA Heart Failure Class**: I (mild) → II → III → IV (severe)
- **Triage categories**: Immediate, Emergent, Urgent, Less Urgent, Non-Urgent
- **Cancer staging**: Stage I → Stage II → Stage III → Stage IV
- **Functional status**: Independent, Needs minimal assistance, Needs moderate assistance, Fully dependent


**Appropriate Statistics for Ordinal Data**

Among the central tendency and dispersion measures you learned earlier, which are appropriate for ordinal scale data?

- **Central tendency**: Mode (most common category), Median (middle rank)
- **Dispersion**: Frequency distribution, Percentages, Range

```{r}
#| label: fig-age-groups
#| fig-cap: "Distribution of Age Groups in Readmission Data"
#| echo: false

readmission_data %>%
  mutate(age_group = factor(age_group, levels = c("18-44", "45-64", "65-74", "75+"))) %>%
  count(age_group) %>%
  ggplot(aes(x = age_group, y = n, fill = age_group)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.5) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Patient Age Group Distribution",
       x = "Age Group", y = "Number of Patients") +
  theme_minimal() +
  theme(legend.position = "none")
```

**Why You Cannot Average Ordinal Data**

A mean for ordinal data is not meaningful. Even if categories are coded as numbers (1, 2, 3, 4), arithmetic operations are meaningless because intervals are not equal!

This is a critical point that's often misunderstood. Let's see why averaging ordinal data produces meaningless results.

**Example**: Suppose we ask 100 patients to rate their post-operative pain using these categories:

- None (coded as 1) — 20 patients
- Mild (coded as 2) — 30 patients
- Moderate (coded as 3) — 35 patients
- Severe (coded as 4) — 15 patients

If we calculate the mean: (20×1 + 30×2 + 35×3 + 15×4) / 100 = **2.45**

**What does 2.45 mean?** Is it "somewhere between Mild and Moderate"? The answer is **we don't know!** because:

- The difference between None→Mild might be very small (minor discomfort)
- The difference between Moderate→Severe might be huge (unbearable pain)
- We cannot assume equal intervals, so averaging is meaningless!

**Better approach**: Report that the **median** is Moderate, and 50% of patients reported Moderate or Severe pain. This is interpretable and meaningful.


---

## Interval Scale

**Definition**:

Numbers have **equal intervals** between scale points, but no true zero point. Differences are meaningful, but ratios are not.

**Key characteristic**: 

The difference between 1 and 2 is the same as the difference between 6 and 7. This makes addition, subtraction, and averaging meaningful.

**Examples**

**Classic interval scales:**

- **Temperature** (Fahrenheit or Celsius) — 0° doesn't mean "no temperature"; it's an arbitrary reference point
- **Calendar year** — Year 0 is arbitrary; 2020 isn't "twice as recent" as 1010

**Healthcare applications** (less common):

- **Numeric satisfaction scales** (1-7 or 1-10 with equal intervals assumed)
- **Numeric pain scales** (0-10 where each point represents an equal increase in pain)
- **Quality of life scores** from validated instruments (0-100 with equal intervals assumed)


**Appropriate Statistics for Interval Data**

Among the central tendency and dispersion measures you learned earlier, which are appropriate for interval scale data?

- **Central tendency**: Mode, Median, **Mean** (now meaningful!)
- **Dispersion**: Range, **Variance, Standard Deviation**
- **All arithmetic operations** are meaningful: Addition, subtraction, averaging

```{r}
#| label: fig-satisfaction-dist
#| fig-cap: "Distribution of Patient Satisfaction Scores (1-7 Scale)"
#| echo: false

# Calculate mean and SD for satisfaction scores
mean_sat <- mean(patient_satisfaction$satisfaction_score)
sd_sat <- sd(patient_satisfaction$satisfaction_score)

ggplot(patient_satisfaction, aes(x = factor(satisfaction_score))) +
  geom_bar(fill = "#2171B5") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  annotate("text", x = 5.5, y = max(table(patient_satisfaction$satisfaction_score)) * 0.9,
           label = paste0("Mean = ", round(mean_sat, 2), "\nSD = ", round(sd_sat, 2)),
           hjust = 0, size = 4.5, fontface = "bold", color = "#E41A1C") +
  labs(title = "Patient Satisfaction Score Distribution",
       subtitle = "1 = Very Dissatisfied, 7 = Very Satisfied | Mean and SD are appropriate for interval scales",
       x = "Satisfaction Score", y = "Number of Patients") +
  theme_minimal()
```

**What you CAN do** with interval data:

- Calculate meaningful averages (mean)
- Compare differences: "Department A's satisfaction (5.2) is 1.3 points higher than Department B's (3.9)"

**What you CANNOT do** with interval data:

- Make ratio statements: You cannot say "40°C is twice as hot as 20°C" because there's no true zero

::: {.callout-note}
## Special Case: Numeric Satisfaction/Pain Scales (Quasi-Interval)

**The complexity**: Numeric rating scales (1-7, 1-10) for satisfaction or pain are technically ordinal, but in healthcare practice they're often treated as interval scales.

**Quick guide**:

- **Category labels** (Poor/Fair/Good/Excellent) → **Ordinal** → Use median only
- **Numeric scales** (1-7, 1-10) → **Quasi-Interval** → Can use mean, but report median too
:::

---

## Ratio Scale

**Definition**:

Has all properties of interval scale plus a **true zero (absolute zero) point**, where zero means "absence of the attribute." Ratios between values are meaningful.

**Key characteristic**: 

Zero means "none" or "absence," making statements like "twice as much" meaningful.

**Examples**

- **Length of stay** (days) — 0 days = no stay
- **Wait time** (minutes) — 0 minutes = no wait
- **Total charges** ($) — $0 = no charges
- **Number of comorbidities** — 0 = no comorbidities
- **Age** (years) — 0 = birth
- **Blood pressure** (mmHg), **Heart rate** (bpm)
- **Weight, height**
- **Readmission rate** (%)

**Appropriate Statistics for Ratio Data**

Among the central tendency and dispersion measures you learned earlier, which are appropriate for ratio scale data?

- **Central tendency**: Mode, Median, **Mean**
- **Dispersion**: Range, **Variance, Standard Deviation**
- **All mathematical operations** are meaningful: Addition, subtraction, multiplication, division

```{r}
#| label: fig-los-hist
#| fig-cap: "Distribution of Length of Stay (Days)"
#| echo: false

# Calculate mean and SD for length of stay
mean_los <- mean(patient_satisfaction$length_of_stay)
sd_los <- sd(patient_satisfaction$length_of_stay)

ggplot(patient_satisfaction, aes(x = length_of_stay)) +
  geom_histogram(binwidth = 1, fill = "#2171B5", color = "white") +
  geom_vline(xintercept = mean_los, color = "#E41A1C", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = mean_los + 1, y = max(table(patient_satisfaction$length_of_stay)) * 3,
           label = paste0("Mean = ", round(mean_los, 2), " days\nSD = ", round(sd_los, 2), " days"),
           hjust = 0, size = 4.5, fontface = "bold", color = "#E41A1C") +
  labs(title = "Length of Stay Distribution",
       x = "Length of Stay (Days)", y = "Number of Patients") +
  theme_minimal()
```

**What you CAN do** with ratio data:

- Calculate meaningful averages (mean)
- Compare differences: "Department A's wait time (45 min) is 15 minutes longer than Department B's (30 min)"
- **Make ratio statements**: "Hospital A's average LOS (6 days) is **twice** that of Hospital B (3 days)"

**Key difference from interval scale**:

The presence of a true zero allows meaningful ratio comparisons. You can say "twice as long" or "three times as many" with ratio data, but NOT with interval data.

---

# Distribution Shapes and Choosing Statistics

## Why Distribution Shape Matters

The shape of your data distribution affects which measure of central tendency best represents the "typical" value. This is crucial for accurate healthcare reporting and decision-making.

For example, if a hospital administrator asks, "What's our realistic expectation of patient length of stay?" The answer may not be that simple: it depends on the distribution!

---

## Four Common Distribution Shapes

### 1. Normal (Symmetric) Distribution

**Characteristics**:

- Bell-shaped and symmetric around the mean
- Mean ≈ Median ≈ Mode (all at the center)
- Most values cluster near the mean, with fewer values in the tails
- Use **Mean** as central tendency

**Healthcare Example: Systolic Blood Pressure**

Blood pressure in a healthy population typically follows a normal distribution, with most people having values near 120 mmHg.

```{r}
#| label: fig-normal-bp
#| fig-cap: "Normal Distribution: Systolic Blood Pressure"
#| echo: false

mean_bp <- mean(bp_data$systolic_bp)
median_bp <- median(bp_data$systolic_bp)
sd_bp <- sd(bp_data$systolic_bp)

ggplot(bp_data, aes(x = systolic_bp)) +
  geom_histogram(binwidth = 2, fill = "#6BAED6", color = "white") +
  geom_vline(xintercept = mean_bp, color = "#E41A1C", linewidth = 1.2) +
  annotate("text", x = mean_bp + 5, y = max(table(cut(bp_data$systolic_bp, breaks = seq(min(bp_data$systolic_bp), max(bp_data$systolic_bp) + 2, by = 2)))) * 0.85,
           label = paste0("Mean ≈ Median = ", round(mean_bp, 1), " mmHg\nSD = ", round(sd_bp, 1), " mmHg"),
           color = "#E41A1C", hjust = 0, fontface = "bold") +
  labs(title = "Systolic Blood Pressure: Normal Distribution",
       subtitle = "Symmetric distribution - Mean is the best measure of central tendency",
       x = "Systolic Blood Pressure (mmHg)", y = "Number of Patients") +
  theme_minimal()
```

**The 68-95-99.7 Rule (Empirical Rule)**

For **normally distributed** data, we can use standard deviation to describe where most values fall:

- **68%** of values fall within **±1 SD** of the mean
- **95%** of values fall within **±2 SD** of the mean
- **99.7%** of values fall within **±3 SD** of the mean

```{r}
#| label: fig-empirical-rule
#| fig-cap: "The Empirical Rule Applied to Blood Pressure"
#| echo: false

# Create theoretical normal curve
x_norm <- seq(mean_bp - 4*sd_bp, mean_bp + 4*sd_bp, length = 1000)
y_norm <- dnorm(x_norm, mean = mean_bp, sd = sd_bp)
normal_curve <- tibble(x = x_norm, y = y_norm * 300)  # Scale for visibility

ggplot(normal_curve, aes(x = x, y = y)) +
  # 3 SD region
  geom_area(data = filter(normal_curve, x >= mean_bp - 3*sd_bp & x <= mean_bp + 3*sd_bp),
            alpha = 0.2, fill = "#FED976") +
  # 2 SD region
  geom_area(data = filter(normal_curve, x >= mean_bp - 2*sd_bp & x <= mean_bp + 2*sd_bp),
            alpha = 0.3, fill = "#FEB24C") +
  # 1 SD region
  geom_area(data = filter(normal_curve, x >= mean_bp - sd_bp & x <= mean_bp + sd_bp),
            alpha = 0.4, fill = "#F03B20") +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = mean_bp, linetype = "dashed") +
  annotate("text", x = mean_bp, y = max(normal_curve$y) * 0.5, label = "68%", fontface = "bold", size = 5) +
  annotate("text", x = mean_bp + 1.5*sd_bp, y = max(normal_curve$y) * 0.2, label = "95%", fontface = "bold", size = 4) +
  annotate("text", x = mean_bp + 2.5*sd_bp, y = max(normal_curve$y) * 0.05, label = "99.7%", fontface = "bold", size = 3) +
  scale_x_continuous(breaks = c(mean_bp - 3*sd_bp, mean_bp - 2*sd_bp, mean_bp - sd_bp,
                                mean_bp, mean_bp + sd_bp, mean_bp + 2*sd_bp, mean_bp + 3*sd_bp),
                     labels = c("-3σ", "-2σ", "-1σ", "μ", "+1σ", "+2σ", "+3σ")) +
  labs(title = "The Empirical Rule (68-95-99.7)",
       subtitle = paste0("For BP: 68% fall within ", round(mean_bp - sd_bp, 0), "-", round(mean_bp + sd_bp, 0), " mmHg"),
       x = "Systolic Blood Pressure (mmHg)", y = "Density") +
  theme_minimal()
```

::: {.callout-tip}
## Practical Healthcare Application

For our blood pressure data with Mean = `r round(mean_bp, 1)` mmHg and SD = `r round(sd_bp, 1)` mmHg:

- **68%** of patients have BP between `r round(mean_bp - sd_bp, 0)`-`r round(mean_bp + sd_bp, 0)` mmHg
- **95%** of patients have BP between `r round(mean_bp - 2*sd_bp, 0)`-`r round(mean_bp + 2*sd_bp, 0)` mmHg
- **99.7%** of patients have BP between `r round(mean_bp - 3*sd_bp, 0)`-`r round(mean_bp + 3*sd_bp, 0)` mmHg

If a patient's BP is above `r round(mean_bp + 2*sd_bp, 0)` mmHg (>2 SD from mean), they're in the top 2.5% - a potential clinical concern!
:::

---

### 2. Right-Skewed (Positively Skewed) Distribution

**Characteristics**:

- Long tail to the right (toward higher values): there are (several) extremely high values
- Most values cluster on the left (lower end)
- Mean > Median > Mode
- Use **Median** as central tendency

**Healthcare Example: Length of Stay**

Hospital length of stay is typically **right-skewed** because most patients have shorter stays, but a few have very long stays.

```{r}
#| label: fig-skewed-los
#| fig-cap: "Right-Skewed Distribution: Length of Stay"
#| echo: false

mean_los_dist <- mean(patient_satisfaction$length_of_stay)
median_los_dist <- median(patient_satisfaction$length_of_stay)

ggplot(patient_satisfaction, aes(x = length_of_stay)) +
  geom_histogram(binwidth = 1, fill = "#6BAED6", color = "white") +
  geom_vline(xintercept = mean_los_dist, color = "blue", linewidth = 1.2) +
  geom_vline(xintercept = median_los_dist, color = "#4DAF4A", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = mean_los_dist + 1, y = 40, label = paste("Mean =", round(mean_los_dist, 1)),
           color = "blue", hjust = 0, fontface = "bold") +
  annotate("text", x = median_los_dist - 1, y = 35, label = paste("Median =", median_los_dist),
           color = "#4DAF4A", hjust = 1, fontface = "bold") +
  labs(title = "Length of Stay: Right-Skewed Distribution",
       subtitle = "Mean is pulled toward long stays; Median better represents 'typical' patient",
       x = "Length of Stay (Days)", y = "Number of Patients") +
  theme_minimal()
```

::: {.callout-important}
## Healthcare Management Insight

When reporting average length of stay or costs to stakeholders:

- **Median** often tells a fairer story for skewed data
- **Mean** is inflated by a few very long stays or expensive cases
- **Best practice**: Report both and explain the difference!

**Example statement**: "The median length of stay is `r median_los_dist` days, meaning half of patients stay `r median_los_dist` days or less. However, the mean is `r round(mean_los_dist, 1)` days due to a small number of patients with extended stays."
:::

---

### 3. Left-Skewed (Negatively Skewed) Distribution

**Characteristics**:

- Long tail to the left (toward lower values):  there are (several) extremely low values
- Most values cluster on the right (higher end)
- Mean < Median < Mode
- Use **Median** as central tendency

**Healthcare Example: Patient Satisfaction Ratings**

When patients rate satisfaction on a 0-100 scale, scores often cluster toward the high end (most patients are satisfied), with a tail toward lower scores (dissatisfied patients).

```{r}
#| label: fig-left-skewed-satisfaction
#| fig-cap: "Left-Skewed Distribution: Patient Satisfaction"
#| echo: false

mean_sat_ls <- mean(high_satisfaction$satisfaction_rating)
median_sat_ls <- median(high_satisfaction$satisfaction_rating)

ggplot(high_satisfaction, aes(x = satisfaction_rating)) +
  geom_histogram(binwidth = 2, fill = "#6BAED6", color = "white") +
  geom_vline(xintercept = mean_sat_ls, color = "#E41A1C", linewidth = 1.2) +
  geom_vline(xintercept = median_sat_ls, color = "#4DAF4A", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = mean_sat_ls - 2, y = 30, label = paste("Mean =", round(mean_sat_ls, 1)),
           color = "#E41A1C", hjust = 1, fontface = "bold") +
  annotate("text", x = median_sat_ls + 2, y = 25, label = paste("Median =", round(median_sat_ls, 1)),
           color = "#4DAF4A", hjust = 0, fontface = "bold") +
  labs(title = "Patient Satisfaction: Left-Skewed Distribution",
       subtitle = "Mean is pulled down by low scores; Median better represents typical experience",
       x = "Satisfaction Rating (0-100)", y = "Number of Patients") +
  theme_minimal()
```

**Key Insight**: Most patients are highly satisfied (scores cluster near 100), but a few very dissatisfied patients pull the mean down. The median better represents the typical patient experience.

---

### 4. Bimodal Distribution

**Characteristics**:

- Two distinct peaks (modes)
- Indicates two different subgroups in your data
- Mean and median fall between the peaks and don't represent either group well
- **Analyze subgroups separately** rather than combining them

**Healthcare Example: ED Wait Times by Shift**

Emergency department wait times show two peaks - one for day shift (faster) and one for night shift (slower, due to reduced staffing).

```{r}
#| label: fig-bimodal-ed
#| fig-cap: "Bimodal Distribution: ED Wait Times by Shift"
#| echo: false

overall_mean <- mean(ed_wait_times$wait_minutes)
overall_median <- median(ed_wait_times$wait_minutes)

ggplot(ed_wait_times, aes(x = wait_minutes)) +
  geom_histogram(binwidth = 3, fill = "#6BAED6", color = "white") +
  geom_vline(xintercept = overall_mean, color = "#E41A1C", linewidth = 1.2) +
  annotate("text", x = overall_mean + 2, y = max(table(cut(ed_wait_times$wait_minutes, breaks = seq(0, 80, by = 3)))) * 1,
           label = paste("Overall Mean =", round(overall_mean, 1), "min"),
           color = "#E41A1C", hjust = 0, fontface = "bold") +
  annotate("text", x = 21, y = max(table(cut(ed_wait_times$wait_minutes, breaks = seq(0, 80, by = 3)))) * 0.75,
           label = "Day Shift\nPeak", color = "#377EB8", fontface = "bold", size = 3.5) +
  annotate("text", x = 51, y = max(table(cut(ed_wait_times$wait_minutes, breaks = seq(0, 80, by = 3)))) * 0.75,
           label = "Night Shift\nPeak", color = "#377EB8", fontface = "bold", size = 3.5) +
  labs(title = "ED Wait Times: Bimodal Distribution",
       subtitle = "Two distinct peaks indicate different subgroups - analyze separately!",
       x = "Wait Time (minutes)", y = "Number of Patients") +
  theme_minimal()
```

**Better approach**: Analyze by shift separately:

```{r}
#| label: tbl-bimodal-subgroups
#| tbl-cap: "ED Wait Times by Shift (Subgroup Analysis)"
#| echo: false

ed_wait_times %>%
  group_by(shift) %>%
  summarise(
    `Mean Wait (min)` = round(mean(wait_minutes), 1),
    `Median Wait (min)` = round(median(wait_minutes), 1),
    `SD (min)` = round(sd(wait_minutes), 1),
    `N Patients` = n()
  ) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

::: {.callout-warning}
## Management Implication

**Overall average (`r round(overall_mean, 1)` min) is misleading!** It doesn't represent either day shift (~22 min) or night shift (~55 min) patients.

**Action**: Different shifts need different strategies. Don't make decisions based on the overall average when you have bimodal data.
:::

---

## Decision Guide: Choosing Central Tendency Based on Shape

| Distribution Shape | Best Measure | Why |
|--------------------|-------------|-----|
| **Normal (Symmetric)** | Mean | Represents balance point well; not distorted by outliers |
| **Right-skewed** | Median | Not pulled by high outliers; better represents "typical" value |
| **Left-skewed** | Median | Not pulled by low outliers; better represents "typical" value |
| **Bimodal** | Analyze subgroups separately | Mean/median fall between peaks, not representative of either group |

---

# Synthesis: Choosing the Right Statistics

## Overview

Now that you've learned about measurement scales, central tendency, dispersion, and distribution shapes, let's synthesize this knowledge into a practical guide for choosing appropriate statistics in healthcare analysis.

**Two key factors determine which statistics to use:**

1. **Measurement scale** → What statistics are *mathematically appropriate*
2. **Distribution shape** → Which statistics *best represent* your data

---

## Summary Table: Statistics by Measurement Scale

```{r}
#| label: tbl-scale-stats
#| tbl-cap: "Appropriate Statistics by Measurement Scale"
#| echo: false

scale_stats <- tibble(
  Scale = c("Nominal", "Ordinal", "Interval/Quasi-Interval", "Ratio"),
  `Key Property` = c("Categories only", "Order, but unequal intervals", "Equal intervals, no true zero", "Equal intervals + true zero"),
  `Central Tendency` = c("Mode", "Mode, Median", "Mode, Median, Mean", "Mode, Median, Mean"),
  Dispersion = c("Frequency, %", "Frequency, %, Range", "All + Variance, SD", "All + Variance, SD"),
  `Healthcare Examples` = c("Diagnosis, Insurance, Blood type", "Pain categories, Severity stages", "Satisfaction scores (numeric)", "LOS, Wait time, Cost, Age")
)

kable(scale_stats) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 11) %>%
  column_spec(1, bold = TRUE, width = "2cm") %>%
  column_spec(2, width = "3.5cm") %>%
  column_spec(3, width = "2.5cm") %>%
  column_spec(4, width = "3cm") %>%
  column_spec(5, width = "4cm")
```

---

# Introduction to Statistical Inference

## From Description to Inference

So far, we've focused on **descriptive statistics** (i.e., summarizing data we have). But healthcare managers need **inferential statistics**: drawing conclusions or making inferences about populations from samples.

We typically work with **samples** (subset of patients; data we have) but want to make conclusions about **populations** (all patients, future patients).

**Example**:

- **Sample**: 200 patients we surveyed about satisfaction
- **Population**: All patients who visit our hospital (past, present, future)
- Is the (observed) satisfaction difference between Department A and Department B in our sample statistically significant, or just random chance?

---

## Why We Need Hypothesis Testing

Let's think about this case: 

- Hospital A has a mean satisfaction score of 5.5 (n=100 patients) 
- Hospital B has 5.2 (n=100 patients)

**Questions a manager asks**:

1. Is this difference meaningful?
2. Could this difference just be due to random sampling?
3. If we surveyed 100 different patients, would we see the same pattern?

**This is where hypothesis testing helps!**

---

## The Logic of Null Hypothesis Testing

Broadly, there are two ways to make statistical inferences about populations from samples:

1. Bayesian approach
2. Frequentist approach (null hypothesis testing)

Both approaches have pros and cons, but we'll focus on the null hypothesis testing approach in this class for computational simplicity. 


Let's assume that we want to understand whether there's a significant difference in patient satisfaction between Hospital A and Hospital B. 

Broadly, null hypothesis testing uses the following approach:

1. **Assume there's NO difference** between the two hospitals (Null Hypothesis, H₀)
2. **Calculate** how likely our observed data would be if this assumption were true
3. **If very unlikely** (p-value < threshold) → Reject the assumption → Conclude there IS a difference (or the difference is statistically significant)

**Think of it like a criminal trial**:

- **Null hypothesis (H₀)**: Defendant is innocent (no effect, no difference)
- **Alternative hypothesis (H₁)**: Defendant is guilty (there is an effect/difference)
- **Evidence**: Our data
- **Verdict**: If evidence strongly contradicts innocence → Reject H₀

---

## Key Terminology

### Null Hypothesis (H₀)

The "no effect" or "no difference" assumption. This is what we're testing against.

**Healthcare examples**:

- H₀: Hospital A's satisfaction = Hospital B's satisfaction
- H₀: New discharge process has no effect on readmission rate
- H₀: Mean LOS for diabetic patients = Mean LOS for non-diabetic patients

### Alternative Hypothesis (H₁ or Hₐ)

What we're trying to find evidence for.

**Healthcare examples**:

- H₁: Hospital A's satisfaction ≠ Hospital B's satisfaction (two-tailed)
- H₁: New discharge process reduces readmission rate (one-tailed)
- H₁: Mean LOS differs between diabetic and non-diabetic patients (two-tailed)

### P-value

The probability of observing data as extreme or more extreme than what we observed, If the null hypothesis were true.

**Important**: The p-value is NOT the probability that H₀ or H₁ is true!

---

## Visualizing the Null Hypothesis

We can understand expected results and their probabilities when H₀ is true. 

```{r}
#| label: fig-null-model
#| fig-cap: "The Null Model: Expected Results When the Null is True"
#| echo: false

# Create smooth normal distribution curve for null hypothesis
x_null <- seq(-0.6, 0.6, length = 1000)
y_null <- dnorm(x_null, mean = 0, sd = 0.15)
null_curve <- tibble(x = x_null, y = y_null)

# Critical values for rejection regions (α = 0.05, two-tailed ≈ ±2 SD)
critical_value <- 0.3  # Approximately 2 * 0.15

ggplot(null_curve, aes(x = x, y = y)) +
  # Rejection regions (tails)
  geom_area(data = filter(null_curve, x <= -critical_value),
            fill = "#E41A1C", alpha = 0.3) +
  geom_area(data = filter(null_curve, x >= critical_value),
            fill = "#E41A1C", alpha = 0.3) +
  # Main curve
  geom_line(linewidth = 1.5, color = "#2171B5") +
  # Vertical line at null hypothesis (mean = 0)
  geom_vline(xintercept = 0, linetype = "dashed", color = "#2171B5", linewidth = 1) +
  # Critical value lines
  geom_vline(xintercept = c(-critical_value, critical_value),
             color = "#E41A1C", linewidth = 1, linetype = "dashed") +
  # Annotations
  annotate("text", x = 0, y = max(null_curve$y) * 1.05,
           label = "H0: No difference\n(mean = 0)",
           fontface = "bold", color = "#2171B5") +
  annotate("text", x = -0.45, y = max(null_curve$y) * 0.5,
           label = "Reject H0", color = "#E41A1C", hjust = 0.5, fontface = "bold") +
  annotate("text", x = 0.45, y = max(null_curve$y) * 0.5,
           label = "Reject H0", color = "#E41A1C", hjust = 0.5, fontface = "bold") +
  annotate("text", x = 0, y = max(null_curve$y) * 0.3,
           label = "Do NOT reject H0\n(consistent with null)",
           color = "#2171B5", hjust = 0.5, size = 3.5) +
  labs(title = "Null Model: Distribution of Differences When H₀ is True",
       subtitle = "If observed difference falls in red zone (tails), it's unlikely under H₀ → Reject H₀",
       x = "Difference Between Groups", y = "Probability Density") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

**Interpretation**:

- **Blue curve**: Expected distribution of differences if H₀ (no difference) is true
- **Middle region**: Observed differences here are likely outcomes with H₀ → Do NOT reject
- **Red tails**: Observed differences here are unlikely outcomes under H₀ → Reject H₀ (statistically significant; significant difference)

---

## Understanding P-values

**What is a P-value?**

A p-value is the probability of getting results at least as extreme as what we observed, **assuming H₀ is true**

**Small p-value (typically < 0.05)**: Our data would be surprising if H₀ were true → Evidence against H₀

**Large p-value (≥ 0.05)**: Our data wouldn't be surprising if H₀ were true → Not enough evidence against H₀

**What is not a P-value?**

Common Misconceptions are: 

1. A p-value means the probability of H₀ is true (X)
- p-value assumes H₀ is true, doesn't tell us probability H₀ is true

2. p ≥ 0.05 means there's no difference (X)
- It indicates that we don't have sufficient evidence to reject H₀ (absence of evidence ≠ evidence of absence)

3. p < 0.05 means the result is clinically and practically important (X)
- It means statistically significant, not necessarily clinically or practically meaningful

---

# Confidence Intervals

**What is a Confidence Interval?**

A **confidence interval (CI)** provides a range of plausible values for a **population** parameter (like a mean) based on sample data.

Instead of just reporting a single estimate, like "the average wait time is 45 minutes," a confidence interval communicates the uncertainty in this estimate due to sampling variability. 

**Understanding the 95% Confidence Level**

A 95% confidence interval means that if we were to repeatedly draw samples of the same size from the population and construct confidence intervals using the same method, approximately 95% of those intervals would contain the true population parameter. ([see this](https://seeing-theory.brown.edu/frequentist-inference/index.html#section2))

It does not mean that there's a 95% probability that the true mean lies within the specific interval calculated from the observed data. The population parameter is fixed; what varies from sample to sample is the interval itself. 

---

## Calculating a Confidence Interval

For a mean, the 95% CI is approximately:

$$\text{CI} = \bar{x} \pm 2 \times \frac{s}{\sqrt{n}}$$

Where:

- $\bar{x}$ = sample mean
- $s$ = sample standard deviation
- $n$ = sample size
- The value 2 is approximately the critical value for 95% confidence (more precisely, 1.96)

**What this tells us**:

- Larger samples lead to **Narrower** CI, reflecting more precise estimates
- More variability (larger SD) leads to **Wider** CI, reflecting greater uncertainty


**An Example: Wait Times**

**Sample data**: 100 patients, mean wait time = 45 minutes, SD = 15 minutes

**95% CI calculation**:
$$45 \pm 2 \times \frac{15}{\sqrt{100}} = 45 \pm 2 \times 1.5 = 45 \pm 3 = [42, 48]$$

**Interpretation**: 

Using this estimation procedure, we construct an interval from 42 to 48 minutes. If the population mean wait time were between 42 and 48, observing a sample mean like the one we obtained would not be surprising. 

---

## Using Confidence Intervals for Decision-Making

**Comparing Two Groups**

Let's assume that we have two hospitals' average satisfaction scores and their confidence intervals

Hospital A mean satisfaction = 5.5 [CI: 5.2, 5.8]
Hospital B mean satisfaction = 4.8 [CI: 4.5, 5.1]

**Question**: Are these hospitals' satisfaction scores significantly different?

**Answer**: Yes! The confidence intervals **don't overlap**, suggesting a significant difference exists.

If the CIs overlapped substantially (e.g., A: [5.0, 5.8], B: [4.8, 5.4]), we'd be less confident there's a true difference.

---


## Type I and Type II Errors

When we make decisions based on hypothesis testing, two types of errors are possible:

### Type I Error (False Positive)

- **Definition**: Concluding there's a difference/effect when there actually isn't
- **Probability**: α = 0.05 (the significance level we set)
- **Example**: Concluding a new treatment works when it actually doesn't → Waste resources, potential harm to patients

### Type II Error (False Negative)

- **Definition**: Failing to detect a real difference/effect that actually exists
- **Example**: Concluding a new treatment doesn't work when it actually does → Miss opportunity to improve care

**Key insight**: We can't eliminate both types of errors simultaneously. Setting a stricter threshold (e.g., 0.01 instead of 0.05) reduces Type I errors but increases Type II errors.

---

## The 0.05 Threshold

**Why p < 0.05?**

- It's a **convention**, not a magical number
- Represents a 5% chance of Type I error (false positive)
- Sometimes we use stricter thresholds (0.01, 0.001) for more important decisions
- Some fields use 0.01 or even lower

**For this course**: We'll use 0.05 as the conventional threshold unless there's any further instruction

---

# Key Takeaways

## Main Concepts to Remember

1. **Measurement scales** determine which statistics are appropriate
   - Nominal → Mode, Frequencies only (no ranking possible)
   - Ordinal → Mode, Median, Range (ranking exists, but intervals are NOT equal—cannot average)
   - Interval/Ratio → Mean, SD (plus all others; equal intervals make arithmetic meaningful)
   - **Quasi-Interval**: Numeric subjective scales (satisfaction, pain) treated as interval by convention

2. **Central tendency** describes the typical value; **dispersion** describes spread
   - Always report both!
   - Consider distribution shape when choosing measures

3. **Distribution shape matters**
   - Symmetric → Use mean
   - Skewed → Use median (usually better for healthcare cost/LOS data)
   - Bimodal → Analyze groups separately

4. **Normal distribution** is foundational for statistical inference
   - Empirical rule (68-95-99.7) helps understand variability

5. **Confidence intervals** quantify uncertainty around estimates
   - Provide a range of plausible values, not just a point estimate
   - Larger samples → narrower (more precise) intervals

6. **Hypothesis testing logic**
   - Assume no difference (H₀)
   - Calculate probability of observed data under this assumption (p-value)
   - Small p-value → Reject H₀
   - **Statistical significance ≠ Practical/clinical importance**

---

# Further Reading

- Lakens, D. (2014). [98% Capture Percentages and 99.9% Confidence Intervals](https://daniellakens.blogspot.com/2014/12/98-capture-percentages-and-999.html) - Understanding confidence intervals beyond the conventional 95%

- Lakens, D. (2017). [Understanding Common Misconceptions About P-values](https://daniellakens.blogspot.com/2017/12/understanding-common-misconceptions.html) - Clarifying frequent misinterpretations of statistical significance

- Navarro, D. [Learning Statistics with R](https://learningstatisticswithr.com/lsr-0.6.pdf) - Free comprehensive textbook covering statistical concepts with R applications
