---
title: "Basic Statistical Concepts (Practice Workbook)"
subtitle: ""
author: ""
Week: "Week 3"
date: 01/01/2026
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: false
    code-copy: true
    theme: cosmo
    highlight-style: github
execute:
  warning: false
  message: false
---

# Introduction

## What You Learned in Lecture

In today's lecture, you learned the basic statistical concepts needed for data analysis:

1. **Central Tendency & Dispersion** - The fundamental tools for summarizing data
   - Mean, median, mode (where the central, typical value is)
   - Standard deviation, variance, range (how spread out the data is)

2. **Measurement Scales** - When each tool is appropriate to use
   - Nominal → frequencies and mode only
   - Ordinal → add median (but NOT mean!)
   - Interval/Ratio → can use mean and SD

3. **Distribution Shapes** - Four common shapes and what they mean
   - Normal/Symmetric → use mean; Empirical rule (68-95-99.7)
   - Right-skewed → use median (typical for healthcare LOS, costs)
   - Left-skewed → use median (e.g., high satisfaction scores)
   - Bimodal → use the modes, analyze subgroups separately

4. **Confidence Intervals** - Quantifying uncertainty around estimates

## What You'll Do in This Workbook

This practice session gives you hands-on experience applying these concepts to a dataset. You'll work with patient satisfaction survey data to:

- Identify measurement scales and choose appropriate statistics
- Calculate descriptive statistics (mean, median, SD, Mode)
- Visualize distribution shapes
- Apply the empirical rule
- Calculate and interpret confidence intervals

**General Format**: Each code chunk has blanks (marked with `___`) for you to fill in. Try to complete them on your own first, then check the solutions at the end.

---

# Dataset: Patient Satisfaction Survey

## Background and Context

### The Scenario

You are a **Data Analyst** at a regional healthcare system that operates three hospitals: General Hospital, University Medical, and Community Health. The system recently conducted a comprehensive patient satisfaction survey across all facilities to:

- Assess overall patient experience
- Identify areas for quality improvement
- Compare performance across departments
- Meet regulatory reporting requirements (HCAHPS; Hospital Customer Assessment of Healthcare Providers and Systems)

The **Chief Quality Officer** has asked you to analyze the survey data and provide insights that will inform strategic decisions about:

- Resource allocation
- Quality improvement initiatives
- Staff training priorities
- Performance benchmarking

### About the Data

This hypothetical dataset contains **200 patient survey responses** collected over a 3-month period. Each row represents one patient's complete survey response and visit information.

**Key features of this dataset:**

- Includes different data types (nominal, ordinal, interval, ratio)
- All variables are commonly tracked in hospital quality programs
- Designed to support practical decision-making

**Data collection period**: Q1 2024 (January - March)
**Survey response rate**: 68% (typical for patient satisfaction surveys)
**Survey method**: Post-discharge email survey 

### Why This Dataset Matters

Patient satisfaction data is critical for healthcare organizations because:

1. **Regulatory compliance**: Centers for Medicare & Medicaid Services (CMS) ties hospital reimbursement to HCAHPS scores
2. **Quality indicator**: Satisfaction correlates with clinical outcomes, safety, and adherence
3. **Competitive advantage**: Public reporting influences patient choice
4. **Operational improvement**: Identifies specific pain points in care delivery

Understanding how to properly analyze this type of data is an essential skill for healthcare managers.

---

## Setup

First, let's run the following code to load required R packages. 

```{r}
#| label: setup

# Load required packages
library(tidyverse)
library(knitr)
library(kableExtra)
library(DescTools)

# Set seed for reproducibility
set.seed(2026)
```

## Create the Dataset

Run this code to generate the patient satisfaction dataset:

```{r}
#| label: create-data

# Create patient satisfaction data
patient_satisfaction <- tibble(
  patient_id = 1:200,
  hospital = sample(c("General Hospital", "University Medical", "Community Health"),
                    200, replace = TRUE),
  department = sample(c("Emergency", "Surgery", "Internal Medicine", "Pediatrics"),
                      200, replace = TRUE),
  age = round(rnorm(200, mean = 55, sd = 18)),
  gender = sample(c("Male", "Female"), 200, replace = TRUE),
  insurance_type = sample(c("Private", "Medicare", "Medicaid", "Uninsured"), 200,
                          prob = c(0.4, 0.35, 0.15, 0.1), replace = TRUE),
  wait_time_minutes = round(rgamma(200, shape = 2, rate = 0.05)),
  satisfaction_score = round(pmin(7, pmax(1, rnorm(200, mean = 5.2, sd = 1.3)))),
  would_recommend = sample(1:7, 200, replace = TRUE,
                           prob = c(0.05, 0.05, 0.1, 0.15, 0.25, 0.25, 0.15)),
  length_of_stay = round(rgamma(200, shape = 1.5, rate = 0.3), 1),
  total_charges = round(rlnorm(200, meanlog = 9, sdlog = 1))
) %>%
  mutate(
    age = pmax(18, pmin(age, 95)),  # Keep age realistic
    age_group = case_when(
      age < 45 ~ "18-44",
      age < 65 ~ "45-64",
      age < 75 ~ "65-74",
      TRUE ~ "75+"
    ),
    age_group = factor(age_group, levels = c("18-44", "45-64", "65-74", "75+")),
    # Create pain category variable (ordinal)
    pain_category = sample(c("None", "Mild", "Moderate", "Severe"), 200,
                          prob = c(0.25, 0.35, 0.30, 0.10), replace = TRUE),
    pain_category = factor(pain_category, levels = c("None", "Mild", "Moderate", "Severe"))
  )

# View first rows
head(patient_satisfaction, 10)
```

**What you should see**: A table showing the first 10 patients' data with all 13 variables. Each row represents one patient's complete information - demographics, visit details, and survey responses.

---

## Variable Descriptions

The dataset includes the following variables:

### **Demographic & Administrative Variables**

| Variable | Type | Description | Values/Range |
|----------|------|-------------|--------------|
| `patient_id` | Identifier | Unique patient identifier | 1-200 |
| `hospital` | Nominal | Facility where patient received care | General Hospital, University Medical, Community Health |
| `department` | Nominal | Department of primary service | Emergency, Surgery, Internal Medicine, Pediatrics |
| `age` | Ratio | Patient age in years | 18-95 |
| `age_group` | Ordinal | Age category | 18-44, 45-64, 65-74, 75+ |
| `gender` | Nominal | Patient gender | Male, Female |
| `insurance_type` | Nominal | Primary insurance payer | Private, Medicare, Medicaid, Uninsured |

### **Experience & Outcome Variables**

| Variable | Type | Description | Values/Range |
|----------|------|-------------|--------------|
| `satisfaction_score` | Quasi-Interval | Overall satisfaction rating | 1-7 (1=Very Dissatisfied, 7=Very Satisfied) |
| `would_recommend` | Quasi-Interval | Likelihood to recommend hospital | 1-7 (1=Definitely No, 7=Definitely Yes) |
| `pain_category` | Ordinal | Self-reported pain level | None, Mild, Moderate, Severe |
| `wait_time_minutes` | Ratio | Time from arrival to first provider contact | Continuous (minutes) |
| `length_of_stay` | Ratio | Total days in hospital | Continuous (days) |
| `total_charges` | Ratio | Total billing charges | Continuous (dollars) |

### **Important Notes on Measurement Scales**

- **Nominal variables** (hospital, department, insurance_type, gender): Categories with no inherent order. Use **mode**, **frequencies**, **shares** only.

- **Ordinal variables** (age_group, pain_category): Categories with meaningful order, but intervals between categories are NOT equal. Use **mode**, **median**, **frequencies**, and **shares**. Do NOT calculate means!

- **Quasi-Interval variables** (satisfaction_score, would_recommend): Numeric rating scales that are technically ordinal but treated as interval by convention in healthcare. Can use **mean and SD**.

- **Ratio variables** (age, wait_time_minutes, length_of_stay, total_charges): Have equal intervals AND a true zero point. Can use **all statistics** (mean, median, SD, etc.). Consider distribution shape when choosing which to report.

---

## What You'll Analyze

Using this patient satisfaction dataset, you will:

1. **Identify the correct measurement scale** for each variable and choose appropriate statistics
2. **Calculate descriptive statistics** to understand central tendency and variation
3. **Visualize distributions** to see patterns in the data (symmetric, skewed, bimodal)
4. **Apply the empirical rule** to identify typical ranges and outliers
5. **Calculate confidence intervals** to quantify uncertainty in your estimates

These fundamental skills form the foundation for data analysis in healthcare management.

**Let's get started!** 📊

---

# Activity 1: Understanding Measurement Scales

In the Variable Descriptions section above, each variable has been classified into a measurement scale (nominal, ordinal, (quasi) interval, or ratio). But why were they classified this way? Let's take some time and think about these classifications.

---

# Activity 2: Descriptive Statistics

Now let's apply what we learned about measurement scales by calculating appropriate descriptive statistics for different types of variables. We'll work through examples of each scale type.

## R Functions for Descriptive Statistics

Before we start calculating statistics, let's review the key R functions you'll need. These functions make it easy to calculate central tendency and dispersion measures.

### Central Tendency Functions

```{r}
#| label: demo-central-tendency
#| eval: false

# Create example data
example_data <- c(2, 3, 3, 4, 5, 5, 6, 8, 15)

# Mean (average)
mean(example_data)          

# Median (middle value)
median(example_data)        

# Mode (most frequent) 
Mode(example_data)  # The Mode function from the DescTools package
```

### Dispersion Functions

```{r}
#| label: demo-dispersion
#| eval: false

# Standard Deviation
sd(example_data)            # 3.64

# Variance (SD squared)
var(example_data)           # 13.25

# Range (max - min)
range(example_data)         # range function returns c(2, 15) - both min and max

# Minimum and Maximum
min(example_data)           # 2
max(example_data)           # 15

# By using these numbers we can get Range manually
max(example_data) - min(example_data)  # 13 
```

### Summary Function - Quick Overview

The `summary()` function gives you multiple statistics at once:

```{r}
#| label: demo-summary
#| eval: false

summary(example_data)
# Output:
#   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
#   2.00    3.50    5.00    5.67    6.50   15.00
```

### Working with Data Frames

When your data is in a tibble/data frame, use `$` to access specific columns/variables:

```{r}
#| label: demo-dataframe
#| eval: false

# Calculate mean age from patient_satisfaction dataset
mean(patient_satisfaction$age)

# Calculate SD for length of stay
sd(patient_satisfaction$length_of_stay)

# Summary of all numeric variables
summary(patient_satisfaction)
```

---

## Task 2.1: Nominal Data - Insurance Type

For nominal data, we can only use **mode** and **frequencies/percentages**. Let's analyze the distribution of insurance types.

```{r}
#| label: task2-insurance
#| eval: false

# Method 1: Quick answer - just the mode (most common insurance type)
Mode(____)

# Method 2: Complete picture - all categories with counts and percentages
insurance_summary <- patient_satisfaction |>
  count(____) |>
  mutate(percentage = round(n / sum(n) * 100, 1))

insurance_summary
```

**Questions**:

1. What is the mode (most common insurance type)?
2. What percentage of patients are uninsured?
3. Why can't we calculate a mean for insurance type?

---

## Task 2.2: Ordinal Data - Pain Categories

For ordinal data, we can use **mode**, **median**, and **frequencies**, but NOT the mean. Let's analyze pain categories.

```{r}
#| label: task2-pain
#| eval: false

# FILL IN: Count patients in each pain category
pain_summary <- patient_satisfaction |>
  count(____) |>
  mutate(percentage = round(n / sum(n) * 100, 1))

pain_summary

# FILL IN: Calculate median (appropriate for ordinal)
median_pain <- median(as.numeric(____))
cat("\nMedian pain category:", levels(patient_satisfaction$pain_category)[median_pain])

# WRONG approach (for demonstration - DO NOT do this in practice!):
mean_pain <- mean(as.numeric(____))
cat("\nMean pain category (MEANINGLESS!):", round(mean_pain, 2))
```

**Questions**:

1. What is the mode pain category?
2. What is the median pain category?
2. How can we interpret the mean pain category? Is it interpretable? 

---

## Task 2.3: Ratio Data - Patient Age

For ratio data, we can use ALL statistics. Calculate central tendency and dispersion for patient age.

```{r}
#| label: task2-age-stats
#| eval: false

# FILL IN: Calculate mean, median, and SD for age
age_stats <- patient_satisfaction |>
  summarise(
    N = n(),
    Mean = round(mean(___), 1),
    Median = median(___),
    SD = round(sd(___), 1),
    Min = min(___),
    Max = max(___)
  )

age_stats |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Questions**:

1. Are mean and median close? What does this suggest about the distribution shape?
2. What does the SD tell you about the variability in patient ages?

---

## Task 2.4: Visualize Distribution Shape - Length of Stay

Create a histogram to see the distribution shape of length of stay (ratio data).

```{r}
#| label: task2-los-histogram
#| eval: false

# FILL IN: Create histogram for length_of_stay
ggplot(patient_satisfaction, aes(x = ____)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Length of Stay",
    x = "Length of Stay (Days)",
    y = "Number of Patients"
  ) +
  theme_minimal()
```

**Questions**:

1. Is this distribution symmetric or skewed?
2. Based on the shape, should you report mean or median as the "typical" value?
3. Why is length of stay typically right-skewed in healthcare data?

---

## Task 2.5: Compare Mean vs. Median on Skewed Data

Visualize where mean and median length of stay fall on the distribution to see how skewness affects these measures.

```{r}
#| label: task2-los-measures
#| eval: false

# FILL IN: Calculate mean and median
mean_los <- mean(patient_satisfaction$___)
median_los <- median(patient_satisfaction$___)

# Create histogram with reference lines
ggplot(patient_satisfaction, aes(x = length_of_stay)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  geom_vline(xintercept = ___, color = "red", linewidth = 1.2) +
  geom_vline(xintercept = ___, color = "darkgreen", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = mean_los + 0.5, y = 35, label = "Mean", color = "red", hjust = 0) +
  annotate("text", x = median_los - 0.5, y = 30, label = "Median", color = "darkgreen", hjust = 1) +
  labs(
    title = "Length of Stay: Mean vs. Median",
    x = "Length of Stay (Days)",
    y = "Number of Patients"
  ) +
  theme_minimal()
```

**Interpretation**: Notice how the mean is pulled toward the long tail (higher values). This confirms that **median better represents the "typical" patient** for right-skewed data.

**Management insight**: If you reported only the mean LOS, it would be inflated by a few very long stays. The median gives a more accurate picture of what most patients experience.

---

# Activity 3: The Empirical Rule (68-95-99.7)

## Task 3.1: Apply the Empirical Rule to Age

If age is approximately normally distributed, use the empirical rule to find where most patients fall.

```{r}
#| label: task3-empirical-rule
#| eval: false

# FILL IN: Calculate mean and SD
age_mean <- mean(patient_satisfaction$___)
age_sd <- sd(patient_satisfaction$___)

# Calculate ranges
range_1sd <- c(age_mean - ___, age_mean + ___)  # ±1 SD
range_2sd <- c(age_mean - 2*___, age_mean + 2*___)  # ±2 SD

cat("Mean age:", round(age_mean, 1), "years\n")
cat("SD:", round(age_sd, 1), "years\n\n")
cat("Using the Empirical Rule:\n")
cat("  68% of patients are between", round(range_1sd[1], 1), "and", round(range_1sd[2], 1), "years\n")
cat("  95% of patients are between", round(range_2sd[1], 1), "and", round(range_2sd[2], 1), "years\n")
```

**Question**: If a patient is 80 years old, are they more than 2 standard deviations from the mean? Would you consider them an outlier?

---

# Activity 4: Confidence Intervals

## Task 4.1: Calculate a 95% Confidence Interval for Mean Wait Time

Calculate a confidence interval to quantify uncertainty around the mean wait time.

**Formula reminder**: CI = mean ± 2 × (SD / √n)

```{r}
#| label: task4-ci-waittime
#| eval: false

# FILL IN: Calculate mean and SD for wait time
wait_mean <- mean(patient_satisfaction$___)
wait_sd <- sd(patient_satisfaction$___)
n <- nrow(patient_satisfaction)

# Calculate standard error
se <- wait_sd / sqrt(___)

# Calculate 95% CI (using 2 as approximation for 1.96)
ci_lower <- wait_mean - 2 * ___
ci_upper <- wait_mean + 2 * ___

cat("Mean wait time:", round(wait_mean, 1), "minutes\n")
cat("95% Confidence Interval: [", round(ci_lower, 1), ",", round(ci_upper, 1), "]\n\n")
cat("Interpretation: If we repeatedly sampled patients and calculated CIs using\n")
cat("this method, approximately 95% of those intervals would contain the true\n")
cat("population mean wait time.\n")
```

**Question**: What happens to the confidence interval width if you increase the sample size?

---

# Solutions

::: {.callout-tip collapse="true"}
## Click for Complete Solutions

```{r}
#| label: solutions-all
#| eval: false

# =============================================================================
# ACTIVITY 1: UNDERSTANDING MEASUREMENT SCALES
# =============================================================================

# Activity 1 is discussion-based. There is no code to run.

# =============================================================================
# ACTIVITY 2: DESCRIPTIVE STATISTICS
# =============================================================================

# Task 2.1: Insurance type (nominal data - mode and frequencies only)
# Method 1: Quick answer - just the mode
Mode(patient_satisfaction$insurance_type)

# Method 2: Complete picture - all categories with counts and percentages
insurance_summary <- patient_satisfaction |>
  count(insurance_type) |>
  mutate(percentage = round(n / sum(n) * 100, 1))

insurance_summary

# Task 2.2: Pain categories (ordinal data - median and frequencies, NOT mean)
pain_summary <- patient_satisfaction |>
  count(pain_category) |>
  mutate(percentage = round(n / sum(n) * 100, 1))

pain_summary

median_pain <- median(as.numeric(patient_satisfaction$pain_category))
cat("\nMedian pain category:", levels(patient_satisfaction$pain_category)[median_pain])

mean_pain <- mean(as.numeric(patient_satisfaction$pain_category))
cat("\nMean pain category (MEANINGLESS!):", round(mean_pain, 2))

# Task 2.3: Age statistics (ratio data - all statistics appropriate)
age_stats <- patient_satisfaction |>
  summarise(
    N = n(),
    Mean = round(mean(age), 1),
    Median = median(age),
    SD = round(sd(age), 1),
    Min = min(age),
    Max = max(age)
  )

age_stats |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"))

# Task 2.4: Length of stay histogram (visualize distribution shape)
ggplot(patient_satisfaction, aes(x = length_of_stay)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Length of Stay",
    x = "Length of Stay (Days)",
    y = "Number of Patients"
  ) +
  theme_minimal()

# Task 2.5: Compare mean vs. median on skewed data
mean_los <- mean(patient_satisfaction$length_of_stay)
median_los <- median(patient_satisfaction$length_of_stay)

ggplot(patient_satisfaction, aes(x = length_of_stay)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  geom_vline(xintercept = mean_los, color = "red", linewidth = 1.2) +
  geom_vline(xintercept = median_los, color = "darkgreen", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = mean_los + 0.5, y = 35, label = "Mean", color = "red", hjust = 0) +
  annotate("text", x = median_los - 0.5, y = 30, label = "Median", color = "darkgreen", hjust = 1) +
  labs(
    title = "Length of Stay: Mean vs. Median",
    x = "Length of Stay (Days)",
    y = "Number of Patients"
  ) +
  theme_minimal()

# =============================================================================
# ACTIVITY 3: EMPIRICAL RULE
# =============================================================================

age_mean <- mean(patient_satisfaction$age)
age_sd <- sd(patient_satisfaction$age)

range_1sd <- c(age_mean - age_sd, age_mean + age_sd)
range_2sd <- c(age_mean - 2*age_sd, age_mean + 2*age_sd)

cat("Mean age:", round(age_mean, 1), "years\n")
cat("SD:", round(age_sd, 1), "years\n\n")
cat("Using the Empirical Rule:\n")
cat("  68% of patients are between", round(range_1sd[1], 1), "and", round(range_1sd[2], 1), "years\n")
cat("  95% of patients are between", round(range_2sd[1], 1), "and", round(range_2sd[2], 1), "years\n")

# =============================================================================
# ACTIVITY 4: CONFIDENCE INTERVALS
# =============================================================================

# Task 4.1: CI for wait time
wait_mean <- mean(patient_satisfaction$wait_time_minutes)
wait_sd <- sd(patient_satisfaction$wait_time_minutes)
n <- nrow(patient_satisfaction)

se <- wait_sd / sqrt(n)
ci_lower <- wait_mean - 2 * se
ci_upper <- wait_mean + 2 * se

cat("Mean wait time:", round(wait_mean, 1), "minutes\n")
cat("95% Confidence Interval: [", round(ci_lower, 1), ",", round(ci_upper, 1), "]\n\n")
cat("Interpretation: If we repeatedly sampled patients and calculated CIs using\n")
cat("this method, approximately 95% of those intervals would contain the true\n")
cat("population mean wait time.\n")

```
:::

---

# Key Takeaways

After completing this practice workbook, you should be able to:

✓ **Identify measurement scales** and choose appropriate statistics
✓ **Calculate descriptive statistics** (mean, median, Mode, sd)
✓ **Interpret distribution shapes** and choose mean vs. median
✓ **Apply the empirical rule** to understand data spread
✓ **Calculate and interpret confidence intervals**

**Remember**:

- **Ordinal data** → median, NOT mean!
- **Skewed data** → median better represents "typical" value
- **Always report** both mean and median for transparency when appropriate
- **Confidence intervals** quantify uncertainty (not probability the parameter is in THIS interval)
- **Statistical significance ≠ Clinical/practical importance**


