---
title: "Lab Assignment 3: ANOVA (Weeks 06-07)"
subtitle: "MBA642: Business Analytics in Healthcare Management"
author: "Your Name" 
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: false
    code-copy: true
    theme:
      - cosmo
      - lecture.scss
    highlight-style: github
    mainfont: "Clear Sans, -apple-system, system-ui, sans-serif"
    fontsize: 1.1em
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(tidyverse)
library(emmeans)
set.seed(2026)
```

# Assignment Overview

- **Total Points**: 10 points
- **Due Date**: March 1, 2026
- **Submission**: Upload your completed .qmd (no need to submit the rendered html file)

## Learning Objectives

This assignment assesses your ability to:

- Choose the appropriate ANOVA design for different research scenarios
- Conduct one-way, repeated-measures, and two-way ANOVA analyses in R
- Interpret ANOVA tables, interaction effects, and post-hoc comparisons
- Draw data-driven conclusions in a healthcare management context

## Instructions

1. Complete all 5 tasks below
2. Fill in the blanks (replace `___` with appropriate code or values)
3. After completing each code chunk, change `#| eval: false` to `#| eval: true` to run it
4. Answer all interpretation questions (1-2 sentences each)
5. Run all code chunks to verify they work before submitting

---

# Grading Rubric

Each task is worth 2 points. Points are allocated as follows:

| Component | Weight | Description |
|-----------|--------|-------------|
| Code correctness | 50% | Code runs without errors and produces correct output |
| Completeness | 30% | All blanks filled in, correct ANOVA type and arguments used |
| Interpretation | 20% | Thoughtful 1-2 sentence response addressing the question |

**Interpretation Guidance**: Your interpretations should be brief but specific. For example:

- **Good**: "The ANOVA shows a significant effect of department (F = 4.52, p = 0.012). Post-hoc tests reveal that the Emergency department has significantly lower satisfaction than Cardiology and Orthopedics."
- **Too vague**: "The result is significant."

---

# Required Packages

- Load the required packages using the code below: 

```{r}
#| label: required-packages

library(ggplot2)
library(tidyverse)
library(emmeans)
```




# Assignment Tasks

## Task 1: One-Way ANOVA — Hospital Department Efficiency (2 points) {.unnumbered}

A hospital operations manager wants to compare **average length of stay** (LOS, in days) across **four departments** to identify departments that may need process improvements.

**Departments**: Emergency, Cardiology, Orthopedics, Neurology

```{r}
#| label: task1-data

# Generate department LOS data
los_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week07_assignment_data1.csv")
```

**Step 1**: Compute summary statistics.

```{r}
#| label: task1-summary
#| eval: false

los_data %>%
  group_by(___) %>%
  summarise(
    n = n(),
    mean = round(mean(___), 2),
    sd = round(sd(___), 2),
    .groups = "drop"
  )
```

**Step 2**: Run a one-way ANOVA.

```{r}
#| label: task1-anova
#| eval: false

# One-way ANOVA
los_anova <- aov(___ ~ ___, data = los_data)
summary(___)
```

**Step 3**: Run post-hoc comparisons using `emmeans`.

```{r}
#| label: task1-posthoc
#| eval: false

# Post-hoc pairwise comparisons
emm_los <- emmeans(___, ~ ___)
pairs(___, adjust = "tukey")
```

**Step 4**: Create a box plot comparing LOS across departments.

```{r}
#| label: task1-plot
#| eval: false

ggplot(los_data, aes(x = ___, y = ___, fill = ___)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Length of Stay by Department",
       x = "Department",
       y = "Length of Stay (days)") +
  theme_minimal() +
  theme(legend.position = "none")
```

**Interpretation Questions**:

1. Is the overall ANOVA significant? What does this tell you?
2. Which departments differ significantly from each other according to the post-hoc test? What operational recommendations would you make?

::: {.callout-note icon=false title="Your Answer (1-2 sentences each)"}
*[Write your answers here]*

1. Question 1: 


2. Question 2:


:::

---

## Task 2: Repeated-Measures ANOVA — Patient Recovery Over Time (2 points) {.unnumbered}

A rehabilitation center tracks **functional mobility scores** for 25 patients at **four time points** after knee replacement surgery to evaluate the recovery trajectory.

**Time Points**: Week 1, Week 4, Week 8, Week 12

```{r}
#| label: task2-data

recovery_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week07_assignment_data2.csv")
recovery_data <- recovery_data |> 
  mutate(
    patient_id = factor(patient_id), 
    time = factor(time, levels = c("Week 1", "Week 4", "Week 8", "Week 12"))
  )
```

**Step 1**: Compute summary statistics by time point.

```{r}
#| label: task2-summary
#| eval: false

recovery_data %>%
  group_by(___) %>%
  summarise(
    mean = round(mean(___), 1),
    sd = round(sd(___), 1),
    .groups = "drop"
  )
```

**Step 2**: Run a repeated-measures ANOVA. Remember: the same patients are measured at all time points, so use `Error()`.

```{r}
#| label: task2-anova
#| eval: false

# Repeated-measures ANOVA
recovery_anova <- aov(___ ~ ___ + Error(___/___), data = recovery_data)
summary(___)
```

**Step 3**: Run post-hoc comparisons.

```{r}
#| label: task2-posthoc
#| eval: false

emm_recovery <- emmeans(___, ~ ___)
pairs(___, adjust = "tukey")
```

**Interpretation Questions**:

1. Is the effect of time significant? What does this mean for patient recovery?
2. Based on the post-hoc results, between which time points do we see the most significant improvement? Does recovery appear to plateau at any point?

::: {.callout-note icon=false title="Your Answer (1-2 sentences each)"}
*[Write your answers here]*

1. Question 1: 


2. Question 2: 


:::

---

## Task 3: Two-Way Between-Subjects ANOVA — Training Method x Experience Level (2 points) {.unnumbered}

A hospital evaluates **two training methods** for medication safety competency across **three experience levels**. Different nurses are in each combination of training method and experience level.

**Training Methods**: Traditional Lecture, High-Fidelity Simulation

**Experience Levels**: Novice (<1 year), Intermediate (1-5 years), Expert (>5 years)

```{r}
#| label: task3-data

training_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week07_assignment_data3.csv")
training_data <- training_data |> 
  mutate(
    experience = factor(experience, levels = c("Novice", "Intermediate", "Expert"))
  )
```

**Step 1**: Create an interaction plot.

```{r}
#| label: task3-plot
#| eval: false

training_summary <- training_data %>%
    group_by(___, ___) %>%
    summarise(mean = mean(___), .groups = "drop")

ggplot(training_summary, aes(x = ___, y = ___, color = ___, group = ___)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Competency by Training Method and Experience",
       x = "Experience Level",
       y = "Mean Competency Score",
       color = "Method") +
  theme_minimal()
```

**Step 2**: Run two-way between-subjects ANOVA.

```{r}
#| label: task3-anova
#| eval: false

training_anova <- aov(___ ~ ___ * ___, data = training_data)
summary(___)
```

**Step 3**: Run simple effects analysis.

```{r}
#| label: task3-simple
#| eval: false

emm_train <- emmeans(___, ~ ___ | ___)
pairs(___, adjust = "tukey")
```

**Interpretation Questions**:

1. Is the interaction significant? What does the interaction plot suggest about the pattern (parallel, ordinal, or disordinal)?
2. At which experience levels does the training method make a significant difference? What training strategy would you recommend?

::: {.callout-note icon=false title="Your Answer (1-2 sentences each)"}
*[Write your answers here]*

1. Question 1: 


2. Question 2: 



:::

---

## Task 4: Two-Way Within-Subjects ANOVA — Communication Mode x Instruction Complexity (2 points) {.unnumbered}

A clinic tests whether **patient comprehension** of discharge instructions varies by **communication mode** and **instruction complexity**. The same 18 patients are tested under all combinations (within-subjects design).

**Communication Modes**: Verbal Only, Written + Verbal

**Instruction Complexity**: Simple, Complex

```{r}
#| label: task4-data

comprehension_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week07_assignment_data4.csv")
comprehension_data <- comprehension_data |> 
  mutate(
    complexity = factor(complexity, levels = c("Simple", "Complex")),
    mode = factor(mode, levels = c("Verbal Only", "Written + Verbal")),
    patient_id = factor(patient_id)
  )

```

**Step 1**: Create an interaction plot.

```{r}
#| label: task4-plot
#| eval: false

comprehension_summary <- comprehension_data %>%
  group_by(___, ___) %>%
  summarise(mean = mean(___), .groups = "drop")

ggplot(comprehension_summary, aes(x = ___, y = ___, color = ___, group = ___)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Comprehension by Mode and Complexity",
       x = "Instruction Complexity",
       y = "Mean Comprehension Score",
       color = "Mode") +
  theme_minimal()
```

**Step 2**: Run two-way within-subjects ANOVA. Remember: both factors are within-subjects, so both go in `Error()`.

```{r}
#| label: task4-anova
#| eval: false

comp_anova <- aov(___ ~ ___ * ___ + Error(___/(___ * ___)),
                  data = comprehension_data)
summary(___)
```

**Step 3**: Run simple effects analysis.

```{r}
#| label: task4-simple
#| eval: false

emm_comp <- emmeans(___, ~ ___ | ___)
pairs(___, adjust = "tukey")
```

**Interpretation Questions**:

1. Is the interaction significant? What does this mean about how communication mode affects comprehension for simple vs. complex instructions?
2. For which instruction type does the communication mode matter more? What practical recommendation would you make for discharge planning?

::: {.callout-note icon=false title="Your Answer (1-2 sentences each)"}
*[Write your answers here]*

1. Question 1: 


2. Question 2: 



:::

---

## Task 5: Mixed Design ANOVA — Intervention Group x Measurement Time (2 points) {.unnumbered}

A diabetes clinic evaluates an **intensive education program** compared to **usual care**. Patients are assigned to one group (between-subjects) and measured at **three time points** (within-subjects).

**Groups**: Usual Care, Intensive Education

**Outcome**: HbA1c level (%), lower = better blood sugar control

```{r}
#| label: task5-data

diabetes_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week07_assignment_data5.csv")
diabetes_data <- diabetes_data |> 
  mutate(
    patient_id = factor(patient_id),
    time = factor(time, levels = c("Baseline", "3 Months", "6 Months")),
    group = factor(group, levels = c("Usual Care", "Intensive Education"))
  )
```

**Step 1**: Create an interaction plot.

```{r}
#| label: task5-plot
#| eval: false

diabetes_summary <- diabetes_data %>%
  group_by(___, ___) %>%
  summarise(mean = mean(___), .groups = "drop") 

ggplot(diabetes_summary, aes(x = ___, y = ___, color = ___, group = ___)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "HbA1c Trajectories by Education Program",
       x = "Time Point",
       y = "Mean HbA1c (%)",
       color = "Group") +
  theme_minimal()
```

**Step 2**: Run mixed design ANOVA. Remember: only the within-subjects factor (time) goes in `Error()`.

```{r}
#| label: task5-anova
#| eval: false

diabetes_anova <- aov(___ ~ ___ * ___ + Error(___/___),
                      data = diabetes_data)
summary(___)
```

**Step 3**: Run simple effects — compare groups at each time point.

```{r}
#| label: task5-simple
#| eval: false

emm_diabetes <- emmeans(___, ~ ___ | ___)
pairs(___, adjust = "tukey")
```

**Interpretation Questions**:

1. Is the interaction significant? What does this tell you about the trajectories of the two groups?
2. At which time points do the groups differ significantly? If you were a clinic administrator, would you fund this program based on the 3-month results alone? Why or why not?

::: {.callout-note icon=false title="Your Answer (1-2 sentences each)"}
*[Write your answers here]*

1. Question 1: 


2. Question 2: 



:::

---

# Submission Checklist

Before submitting, verify:

- [ ] All 5 tasks are completed
- [ ] All blanks (`___`) are filled in with appropriate values
- [ ] All `#| eval: false` changed to `#| eval: true`
- [ ] All code chunks run without errors
- [ ] All interpretation questions are answered (1-2 sentences each)
- [ ] Visualizations have appropriate titles and axis labels
- [ ] File renders to HTML successfully

---

# Points Summary

| Task | Points | Topic | Design |
|------|--------|-------|--------|
| Task 1 | 2.0 | Hospital department efficiency | One-Way ANOVA |
| Task 2 | 2.0 | Patient recovery over time | Repeated-Measures ANOVA |
| Task 3 | 2.0 | Training method × experience | Two-Way Between-Subjects |
| Task 4 | 2.0 | Communication mode × complexity | Two-Way Within-Subjects |
| Task 5 | 2.0 | Education program × time | Mixed Design ANOVA |
| **Total** | **10.0** | | |

---

**Good luck!**
