---
title: "ANOVA 2: Two-Way ANOVA (Practice Activities)"
subtitle: "MBA642: Business Analytics in Healthcare Management"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
    theme: cosmo
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(emmeans)
set.seed(2026)
```

# Overview

In this practice session, you will apply two-way ANOVA techniques learned in the lecture. This is a learning activity designed to help you gain hands-on experience with analyzing factorial designs in healthcare contexts.

**What you'll practice:**

- Two-way between-subjects ANOVA
- Two-way within-subjects ANOVA
- Mixed design ANOVA
- Interaction plots
- Simple effects analysis using `emmeans`

**Format**: Fill-in-the-blank code sections (marked with `___`)

**Instructions**:

1. Complete the code by replacing `___` with appropriate values
2. Run each chunk to verify your analyses
3. Answer the interpretation questions
4. Compare your results with the solutions at the end

**Note**: This practice includes both **lecture examples** (Activities 1-5) that you code along with during class, and **practice activities** (Activities 6-10) for independent work.

---

# Recap: Key Concepts

## Two-Way ANOVA Designs

| Design | Factor A | Factor B | Example | R Formula |
|--------|----------|----------|---------|-----------|
| **Between-Subjects** | Different subjects | Different subjects | Treatment × Severity | `aov(y ~ A * B)` |
| **Within-Subjects** | Same subjects | Same subjects | Medication × Time | `aov(y ~ A * B + Error(id/(A*B)))` |
| **Mixed** | Different subjects | Same subjects | Program × Time | `aov(y ~ A * B + Error(id/B))` |

## Three Effects Tested

1. **Main effect of Factor A**: Does Factor A affect the outcome (averaging across Factor B)?
2. **Main effect of Factor B**: Does Factor B affect the outcome (averaging across Factor A)?
3. **Interaction A × B**: Does the effect of A depend on the level of B?

## Interaction Patterns

- **No interaction**: Parallel lines (effect of A is constant across levels of B)
- **Ordinal interaction**: Converging/diverging lines (effect of A varies in magnitude)
- **Disordinal interaction**: Crossing lines (effect of A reverses direction)

## Simple Effects Analysis

When interaction is significant, use `emmeans` to examine effects at specific levels:

```{r}
#| eval: false

# Compare Factor A at each level of Factor B
emm <- emmeans(model, ~ A | B)
pairs(emm, adjust = "tukey")

# Compare Factor B at each level of Factor A
emm <- emmeans(model, ~ B | A)
pairs(emm, adjust = "tukey")
```

---

# Common Errors and How to Fix Them

## Error 1: "object not found"

**Cause**: You didn't run the data generation chunk

**Fix**: Scroll up and run the chunk that creates the dataset for that activity

## Error 2: Incorrect Error term in repeated measures

```r
# ❌ Wrong (missing Error term):
aov(score ~ medication * time, data = within_data)

# ✅ Right (within-subjects design):
aov(score ~ medication * time + Error(patient_id/(medication * time)), data = within_data)

# ✅ Right (mixed design):
aov(score ~ program * time + Error(nurse_id/time), data = mixed_data)
```

## Error 3: Factor not properly defined

**Cause**: Variable is numeric instead of factor

**Fix**: Use `factor()` to convert:

```r
data <- data %>%
  mutate(severity = factor(severity, levels = c("Mild", "Moderate", "Severe")))
```

## Error 4: emmeans formula syntax

```r
# ❌ Wrong:
emmeans(model, treatment | severity)

# ✅ Right (use tilde ~):
emmeans(model, ~ treatment | severity)
```

---

# Part 1: Between-Subjects ANOVA

## Activity 1: Treatment × Disease Severity

**Scenario**: A clinical trial compares two treatments (Standard vs. Novel) for chronic pain across three severity levels (Mild, Moderate, Severe). Does treatment effectiveness depend on disease severity?

### Generate Data

```{r}
#| label: activity1-data

n_per_cell <- 15

pain_data <- expand.grid(
  treatment = c("Standard", "Novel"),
  severity = c("Mild", "Moderate", "Severe"),
  subject = 1:n_per_cell
) %>%
  mutate(
    base = case_when(
      severity == "Mild" ~ 70,
      severity == "Moderate" ~ 55,
      severity == "Severe" ~ 40
    ),
    treatment_effect = case_when(
      treatment == "Standard" ~ 0,
      treatment == "Novel" & severity == "Mild" ~ 5,
      treatment == "Novel" & severity == "Moderate" ~ 12,
      treatment == "Novel" & severity == "Severe" ~ 22
    ),
    pain_reduction = base + treatment_effect + rnorm(n(), 0, 6)
  ) %>%
  mutate(
    severity = factor(severity, levels = c("Mild", "Moderate", "Severe")),
    treatment = factor(treatment, levels = c("Standard", "Novel"))
  )
```

### Task 1.1: Create Interaction Plot

```{r}
#| label: activity1-plot
#| eval: false

# Calculate means for each treatment-severity combination
plot_data <- pain_data %>%
  group_by(___, ___) %>%
  summarise(mean = mean(___), .groups = "drop")

# Create interaction plot
ggplot(plot_data, aes(x = ___, y = ___,
                      color = ___, group = ___)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Interaction Plot: Treatment × Severity",
       x = "___",
       y = "___",
       color = "___") +
  theme_minimal()
```

**Question**: Do the lines appear parallel, converging, or crossing? What does this suggest about interaction?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

### Task 1.2: Run Two-Way ANOVA

```{r}
#| label: activity1-anova
#| eval: false

# Two-way between-subjects ANOVA
anova_pain <- aov(___ ~ ___ * ___, data = pain_data)
summary(___)
```

**Question**: Which effects are significant (p < 0.05)? Is the interaction significant?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

### Task 1.3: Simple Effects Analysis

```{r}
#| label: activity1-simple
#| eval: false

# Get estimated marginal means
emm <- emmeans(___, ~ ___ | ___)

# Test simple effects
pairs(emm, adjust = "___")
```

**Question**: At which severity levels does the Novel treatment significantly outperform Standard treatment?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

## Activity 2: Training Method × Experience Level

**Scenario**: A hospital evaluates two training methods (Lecture vs. Simulation) for medication safety across three experience levels (Novice, Intermediate, Expert).

### Generate Data

```{r}
#| label: activity2-data

n_cell <- 12

training_data <- expand.grid(
  method = c("Lecture", "Simulation"),
  experience = c("Novice", "Intermediate", "Expert"),
  subject = 1:n_cell
) %>%
  mutate(
    score = case_when(
      method == "Lecture" & experience == "Novice" ~ 68,
      method == "Lecture" & experience == "Intermediate" ~ 78,
      method == "Lecture" & experience == "Expert" ~ 88,
      method == "Simulation" & experience == "Novice" ~ 82,
      method == "Simulation" & experience == "Intermediate" ~ 80,
      method == "Simulation" & experience == "Expert" ~ 79
    ) + rnorm(n(), 0, 4),
    experience = factor(experience, levels = c("Novice", "Intermediate", "Expert"))
  )
```

### Task 2.1: Create Interaction Plot

```{r}
#| label: activity2-plot
#| eval: false

training_data %>%
  group_by(___, ___) %>%
  summarise(mean = mean(___), .groups = "drop") %>%
  ggplot(aes(x = ___, y = ___, color = ___, group = ___)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Training Method × Experience Interaction",
       x = "___",
       y = "___") +
  theme_minimal()
```

**Question**: Describe the interaction pattern. Does one method work better for all experience levels?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

### Task 2.2: Run ANOVA and Simple Effects

```{r}
#| label: activity2-anova
#| eval: false

# ANOVA
anova_training <- aov(___ ~ ___ * ___, data = training_data)
summary(___)

# Simple effects
emm_train <- emmeans(___, ~ ___ | ___)
pairs(___, adjust = "tukey")
```

**Question**: Which training method is better for novice nurses? For expert nurses?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

## Activity 3: Intervention × Facility Type (Practice)

**Scenario**: A health system evaluates care transition programs (Standard vs. Enhanced) on readmission rates across facility types (Urban Academic, Suburban Community, Rural Critical Access).

### Generate Data

```{r}
#| label: activity3-data

n_cell <- 10

intervention_data <- expand.grid(
  program = c("Standard Discharge", "Enhanced Transition"),
  facility = c("Urban Academic", "Suburban Community", "Rural Critical Access"),
  id = 1:n_cell
) %>%
  mutate(
    readmit_rate = case_when(
      program == "Standard Discharge" & facility == "Urban Academic" ~ 14,
      program == "Standard Discharge" & facility == "Suburban Community" ~ 12,
      program == "Standard Discharge" & facility == "Rural Critical Access" ~ 16,
      program == "Enhanced Transition" & facility == "Urban Academic" ~ 13,
      program == "Enhanced Transition" & facility == "Suburban Community" ~ 9,
      program == "Enhanced Transition" & facility == "Rural Critical Access" ~ 8
    ) + rnorm(n(), 0, 1.5)
  )
```

### Your Code Here

```{r}
#| label: activity3-analysis
#| eval: false

# 1. Create interaction plot


# 2. Run two-way ANOVA


# 3. Simple effects analysis if interaction is significant

```

**Question**: Which facility type benefits most from the Enhanced Transition program?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

## Activity 4: Medication × Age Group (Practice)

**Scenario**: Compare two pain medications (Drug A vs. Drug B) across three age groups (Young 18-40, Middle 41-65, Older 65+) on pain relief scores.

### Generate Data

```{r}
#| label: activity4-data

n_cell <- 15

medication_data <- expand.grid(
  medication = c("Drug A", "Drug B"),
  age_group = c("Young (18-40)", "Middle (41-65)", "Older (65+)"),
  patient = 1:n_cell
) %>%
  mutate(
    pain_relief = case_when(
      medication == "Drug A" ~ 55 + rnorm(n(), 0, 6),
      medication == "Drug B" & age_group == "Young (18-40)" ~ 58 + rnorm(n(), 0, 6),
      medication == "Drug B" & age_group == "Middle (41-65)" ~ 68 + rnorm(n(), 0, 6),
      medication == "Drug B" & age_group == "Older (65+)" ~ 78 + rnorm(n(), 0, 6)
    )
  )
```

### Your Code Here

```{r}
#| label: activity4-analysis
#| eval: false

# 1. Create interaction plot


# 2. Run ANOVA


# 3. Simple effects analysis

```

**Question**: Does Drug B work equally well for all age groups?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

# Part 2: Within-Subjects ANOVA

## Activity 5: Medication Type × Time of Day

**Scenario**: The same 20 patients receive three medications (Med A, Med B, Med C) at two times (Morning, Evening). Does medication effectiveness depend on time of administration?

### Generate Data

```{r}
#| label: activity5-data

n_patients <- 20

within_data <- expand.grid(
  patient_id = 1:n_patients,
  medication = c("Med A", "Med B", "Med C"),
  time = c("Morning", "Evening")
) %>%
  mutate(
    patient_effect = rep(rnorm(n_patients, 0, 5), 6),
    fixed = case_when(
      medication == "Med A" & time == "Morning" ~ 63,
      medication == "Med A" & time == "Evening" ~ 72,
      medication == "Med B" & time == "Morning" ~ 70,
      medication == "Med B" & time == "Evening" ~ 71,
      medication == "Med C" & time == "Morning" ~ 77,
      medication == "Med C" & time == "Evening" ~ 65
    ),
    effectiveness = fixed + patient_effect + rnorm(n(), 0, 3)
  )
```

### Task 5.1: Create Interaction Plot

```{r}
#| label: activity5-plot
#| eval: false

within_data %>%
  group_by(___, ___) %>%
  summarise(mean = mean(___), .groups = "drop") %>%
  ggplot(aes(x = ___, y = ___, color = ___, group = ___)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Medication Effectiveness by Time of Day",
       x = "___",
       y = "___") +
  theme_minimal()
```

### Task 5.2: Run Within-Subjects ANOVA

**Important**: For within-subjects design, include Error term!

```{r}
#| label: activity5-anova
#| eval: false

# Two-way within-subjects ANOVA
# Note the Error term for repeated measures
within_anova <- aov(___ ~ ___ * ___ + Error(___/(___*___)),
                    data = within_data)
summary(___)
```

### Task 5.3: Simple Effects

```{r}
#| label: activity5-simple
#| eval: false

# Compare time at each medication level
emm_within <- emmeans(___, ~ ___ | ___)
pairs(___, adjust = "tukey")
```

**Question**: For which medications does timing matter?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

## Activity 6: Task Type × Shift Period (Practice)

**Scenario**: The same 15 nurses perform two task types (Documentation, Patient Care) at three shift periods (Beginning, Middle, End). Does performance change differently for different tasks?

### Generate Data

```{r}
#| label: activity6-data

n_nurses <- 15

task_data <- expand.grid(
  nurse_id = 1:n_nurses,
  task = c("Documentation", "Patient Care"),
  period = c("Beginning", "Middle", "End")
) %>%
  mutate(
    nurse_effect = rep(rnorm(n_nurses, 0, 4), 6),
    fixed = case_when(
      task == "Documentation" & period == "Beginning" ~ 90,
      task == "Documentation" & period == "Middle" ~ 82,
      task == "Documentation" & period == "End" ~ 70,
      task == "Patient Care" & period == "Beginning" ~ 88,
      task == "Patient Care" & period == "Middle" ~ 85,
      task == "Patient Care" & period == "End" ~ 80
    ),
    performance = fixed + nurse_effect + rnorm(n(), 0, 3),
    period = factor(period, levels = c("Beginning", "Middle", "End"))
  )
```

### Your Code Here

```{r}
#| label: activity6-analysis
#| eval: false

# 1. Interaction plot


# 2. Within-subjects ANOVA (remember Error term!)


# 3. Simple effects analysis

```

**Question**: Does documentation performance decline more than patient care performance over the shift?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

# Part 3: Mixed Design ANOVA

## Activity 7: Training Program × Assessment Time

**Scenario**: Nurses are randomly assigned to one of three training programs (Traditional, Simulation, Blended) and tested at three times (Pre, Post, 3 Months). Between-subjects: program; Within-subjects: time.

### Generate Data

```{r}
#| label: activity7-data

n_per_group <- 20

mixed_data <- expand.grid(
  nurse_id = 1:(n_per_group * 3),
  time = c("Pre", "Post", "3 Months")
) %>%
  mutate(
    program = rep(rep(c("Traditional", "Simulation", "Blended"),
                      each = n_per_group), 3),
    nurse_effect = rep(rnorm(n_per_group * 3, 0, 4), 3),
    fixed = case_when(
      program == "Traditional" & time == "Pre" ~ 65,
      program == "Traditional" & time == "Post" ~ 75,
      program == "Traditional" & time == "3 Months" ~ 66,
      program == "Simulation" & time == "Pre" ~ 65,
      program == "Simulation" & time == "Post" ~ 84,
      program == "Simulation" & time == "3 Months" ~ 79,
      program == "Blended" & time == "Pre" ~ 65,
      program == "Blended" & time == "Post" ~ 88,
      program == "Blended" & time == "3 Months" ~ 85
    ),
    score = fixed + nurse_effect + rnorm(n(), 0, 3),
    time = factor(time, levels = c("Pre", "Post", "3 Months"))
  )
```

### Task 7.1: Interaction Plot

```{r}
#| label: activity7-plot
#| eval: false

mixed_data %>%
  group_by(___, ___) %>%
  summarise(mean = mean(___), .groups = "drop") %>%
  ggplot(aes(x = ___, y = ___, color = ___, group = ___)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Competency Trajectories by Training Program",
       x = "___",
       y = "___") +
  theme_minimal()
```

### Task 7.2: Mixed Design ANOVA

**Important**: Error term only includes within-subjects factor (time)!

```{r}
#| label: activity7-anova
#| eval: false

# Mixed design ANOVA
# Between: program, Within: time
mixed_anova <- aov(___ ~ ___ * ___ + Error(___/___),
                   data = mixed_data)
summary(___)
```

### Task 7.3: Simple Effects - Programs at Each Time

```{r}
#| label: activity7-simple1
#| eval: false

# Compare programs at each time point
emm_mixed <- emmeans(___, ~ ___ | ___)
pairs(___, adjust = "tukey")
```

### Task 7.4: Simple Effects - Time Within Each Program

```{r}
#| label: activity7-simple2
#| eval: false

# Compare time points within each program
emm_time <- emmeans(___, ~ ___ | ___)
pairs(___, adjust = "tukey")
```

**Question**: Which programs maintain learning gains at 3 months? Which show decay?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

## Activity 8: Diabetes Intervention × Measurement Occasions

**Scenario**: Patients are assigned to Usual Care or Intensive Education (between-subjects) and HbA1c is measured at Baseline, 3 Months, 6 Months (within-subjects).

### Generate Data

```{r}
#| label: activity8-data

n_group <- 25

diabetes_mixed <- expand.grid(
  patient_id = 1:(n_group * 2),
  time = c("Baseline", "3 Months", "6 Months")
) %>%
  mutate(
    group = rep(rep(c("Usual Care", "Intensive Education"), each = n_group), 3),
    patient_effect = rep(rnorm(n_group * 2, 0, 0.5), 3),
    fixed = case_when(
      group == "Usual Care" & time == "Baseline" ~ 8.5,
      group == "Usual Care" & time == "3 Months" ~ 8.3,
      group == "Usual Care" & time == "6 Months" ~ 8.2,
      group == "Intensive Education" & time == "Baseline" ~ 8.5,
      group == "Intensive Education" & time == "3 Months" ~ 8.0,
      group == "Intensive Education" & time == "6 Months" ~ 7.1
    ),
    hba1c = fixed + patient_effect + rnorm(n(), 0, 0.3),
    time = factor(time, levels = c("Baseline", "3 Months", "6 Months"))
  )
```

### Your Code Here

```{r}
#| label: activity8-analysis
#| eval: false

# 1. Interaction plot


# 2. Mixed ANOVA


# 3. Simple effects: groups at each time point

```

**Question**: At which time point does the Intensive Education group show significant improvement?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

## Activity 9: Therapy Type × Session Number (Practice)

**Scenario**: Patients are assigned to CBT or Mindfulness therapy (between-subjects) and anxiety is measured at Session 1, Session 3, Session 5 (within-subjects).

### Generate Data

```{r}
#| label: activity9-data

n_group <- 18

therapy_data <- expand.grid(
  patient_id = 1:(n_group * 2),
  session = c("Session 1", "Session 3", "Session 5")
) %>%
  mutate(
    therapy = rep(rep(c("CBT", "Mindfulness"), each = n_group), 3),
    patient_effect = rep(rnorm(n_group * 2, 0, 4), 3),
    fixed = case_when(
      therapy == "CBT" & session == "Session 1" ~ 65,
      therapy == "CBT" & session == "Session 3" ~ 48,
      therapy == "CBT" & session == "Session 5" ~ 42,
      therapy == "Mindfulness" & session == "Session 1" ~ 65,
      therapy == "Mindfulness" & session == "Session 3" ~ 55,
      therapy == "Mindfulness" & session == "Session 5" ~ 40
    ),
    anxiety = fixed + patient_effect + rnorm(n(), 0, 3),
    session = factor(session, levels = c("Session 1", "Session 3", "Session 5"))
  )
```

### Your Code Here

```{r}
#| label: activity9-analysis
#| eval: false

# 1. Interaction plot


# 2. Mixed ANOVA


# 3. Simple effects analysis

```

**Question**: Which therapy shows faster initial improvement? Do they converge by Session 5?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

## Activity 10: Facility Type × Time (Practice)

**Scenario**: Three facility types (Urban, Suburban, Rural) implement a safety initiative (between-subjects) and safety scores are measured at Baseline and 6 Months (within-subjects).

### Generate Data

```{r}
#| label: activity10-data

n_facilities <- 15

safety_data <- expand.grid(
  facility_id = 1:(n_facilities * 3),
  time = c("Baseline", "6 Months")
) %>%
  mutate(
    type = rep(rep(c("Urban", "Suburban", "Rural"), each = n_facilities), 2),
    facility_effect = rep(rnorm(n_facilities * 3, 0, 3), 2),
    fixed = case_when(
      type == "Urban" & time == "Baseline" ~ 80,
      type == "Urban" & time == "6 Months" ~ 85,
      type == "Suburban" & time == "Baseline" ~ 78,
      type == "Suburban" & time == "6 Months" ~ 86,
      type == "Rural" & time == "Baseline" ~ 68,
      type == "Rural" & time == "6 Months" ~ 84
    ),
    score = fixed + facility_effect + rnorm(n(), 0, 2),
    time = factor(time, levels = c("Baseline", "6 Months"))
  )
```

### Your Code Here

```{r}
#| label: activity10-analysis
#| eval: false

# 1. Interaction plot


# 2. Mixed ANOVA


# 3. Simple effects

```

**Question**: Which facility type shows the greatest improvement? Why might this be important for resource allocation?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

# Summary

## What You've Practiced

- ✓ Between-subjects two-way ANOVA (different participants in each cell)
- ✓ Within-subjects two-way ANOVA (same participants in all conditions)
- ✓ Mixed design ANOVA (one between, one within factor)
- ✓ Creating interaction plots to visualize factorial designs
- ✓ Interpreting main effects and interactions
- ✓ Conducting simple effects analysis with `emmeans`

## Key Formulas Reference

### Between-Subjects ANOVA
```r
aov(outcome ~ factorA * factorB, data = data)
```

### Within-Subjects ANOVA
```r
aov(outcome ~ factorA * factorB + Error(subject_id/(factorA * factorB)), data = data)
```

### Mixed Design ANOVA
```r
# Between: factorA, Within: factorB
aov(outcome ~ factorA * factorB + Error(subject_id/factorB), data = data)
```

### Simple Effects
```r
emm <- emmeans(model, ~ factorA | factorB)  # A at each level of B
pairs(emm, adjust = "tukey")
```

---

# Solutions

Solutions are provided below. Try to complete each activity independently before checking the solutions!

## Activity 1 Solutions

### Task 1.1: Interaction Plot

```{r}
#| label: solution1-plot
#| code-fold: true

plot_data <- pain_data %>%
  group_by(treatment, severity) %>%
  summarise(mean = mean(pain_reduction), .groups = "drop")

ggplot(plot_data, aes(x = severity, y = mean,
                      color = treatment, group = treatment)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Interaction Plot: Treatment × Severity",
       x = "Disease Severity",
       y = "Mean Pain Reduction",
       color = "Treatment") +
  theme_minimal()
```

The lines converge, suggesting an ordinal interaction where the Novel treatment's benefit increases with severity.

### Task 1.2: ANOVA

```{r}
#| label: solution1-anova
#| code-fold: true

anova_pain <- aov(pain_reduction ~ treatment * severity, data = pain_data)
summary(anova_pain)
```

All three effects are significant: treatment, severity, and their interaction.

### Task 1.3: Simple Effects

```{r}
#| label: solution1-simple
#| code-fold: true

emm <- emmeans(anova_pain, ~ treatment | severity)
pairs(emm, adjust = "tukey")
```

The Novel treatment significantly outperforms Standard at all severity levels, with the largest effect for Severe patients.

---

## Activity 2 Solutions

### Task 2.1: Interaction Plot

```{r}
#| label: solution2-plot
#| code-fold: true

training_data %>%
  group_by(method, experience) %>%
  summarise(mean = mean(score), .groups = "drop") %>%
  ggplot(aes(x = experience, y = mean, color = method, group = method)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Training Method × Experience Interaction",
       x = "Experience Level",
       y = "Mean Competency Score") +
  theme_minimal()
```

The lines cross, indicating a disordinal interaction. Simulation is better for novices, while Lecture is better for experts.

### Task 2.2: ANOVA and Simple Effects

```{r}
#| label: solution2-anova
#| code-fold: true

anova_training <- aov(score ~ method * experience, data = training_data)
summary(anova_training)

emm_train <- emmeans(anova_training, ~ method | experience)
pairs(emm_train, adjust = "tukey")
```

Simulation is significantly better for Novice and Intermediate nurses, while Lecture is better for Expert nurses.

---

## Activity 3 Solutions

```{r}
#| label: solution3
#| code-fold: true

# Interaction plot
intervention_data %>%
  group_by(program, facility) %>%
  summarise(mean = mean(readmit_rate), .groups = "drop") %>%
  ggplot(aes(x = facility, y = mean, color = program, group = program)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Readmission Rates by Program and Facility Type",
       x = "Facility Type",
       y = "Mean Readmission Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# ANOVA
anova_int <- aov(readmit_rate ~ program * facility, data = intervention_data)
summary(anova_int)

# Simple effects
emmeans(anova_int, ~ program | facility) %>% pairs(adjust = "tukey")
```

Rural Critical Access hospitals benefit most from the Enhanced Transition program, showing the largest reduction in readmission rates.

---

## Activity 4 Solutions

```{r}
#| label: solution4
#| code-fold: true

# Interaction plot
medication_data %>%
  group_by(medication, age_group) %>%
  summarise(mean = mean(pain_relief), .groups = "drop") %>%
  ggplot(aes(x = age_group, y = mean, color = medication, group = medication)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Pain Relief by Medication and Age Group",
       x = "Age Group",
       y = "Mean Pain Relief Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# ANOVA
anova_med <- aov(pain_relief ~ medication * age_group, data = medication_data)
summary(anova_med)

# Simple effects
emmeans(anova_med, ~ medication | age_group) %>% pairs(adjust = "tukey")
```

Drug B shows increasing effectiveness with age, working best for older adults.

---

## Activity 5 Solutions

### Task 5.1: Interaction Plot

```{r}
#| label: solution5-plot
#| code-fold: true

within_data %>%
  group_by(medication, time) %>%
  summarise(mean = mean(effectiveness), .groups = "drop") %>%
  ggplot(aes(x = medication, y = mean, color = time, group = time)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Medication Effectiveness by Time of Day",
       x = "Medication",
       y = "Mean Effectiveness Score") +
  theme_minimal()
```

### Task 5.2: Within-Subjects ANOVA

```{r}
#| label: solution5-anova
#| code-fold: true

within_anova <- aov(effectiveness ~ medication * time +
                     Error(patient_id/(medication * time)),
                    data = within_data)
summary(within_anova)
```

### Task 5.3: Simple Effects

```{r}
#| label: solution5-simple
#| code-fold: true

emm_within <- emmeans(within_anova, ~ time | medication)
pairs(emm_within, adjust = "tukey")
```

Timing matters for Med A (better in evening) and Med C (better in morning), but not for Med B.

---

## Activity 6 Solutions

```{r}
#| label: solution6
#| code-fold: true

# Interaction plot
task_data %>%
  group_by(task, period) %>%
  summarise(mean = mean(performance), .groups = "drop") %>%
  ggplot(aes(x = period, y = mean, color = task, group = task)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Performance by Task and Shift Period",
       x = "Shift Period",
       y = "Mean Performance Score") +
  theme_minimal()

# Within-subjects ANOVA
task_anova <- aov(performance ~ task * period + Error(nurse_id/(task * period)),
                  data = task_data)
summary(task_anova)

# Simple effects
emmeans(task_anova, ~ task | period) %>% pairs(adjust = "tukey")
```

Documentation performance declines significantly more than patient care performance as the shift progresses.

---

## Activity 7 Solutions

### Tasks 7.1-7.4: Complete Analysis

```{r}
#| label: solution7
#| code-fold: true

# Interaction plot
mixed_data %>%
  group_by(program, time) %>%
  summarise(mean = mean(score), .groups = "drop") %>%
  ggplot(aes(x = time, y = mean, color = program, group = program)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Competency Trajectories by Training Program",
       x = "Assessment Time",
       y = "Mean Competency Score") +
  theme_minimal()

# Mixed ANOVA
mixed_anova <- aov(score ~ program * time + Error(nurse_id/time),
                   data = mixed_data)
summary(mixed_anova)

# Simple effects: programs at each time
emm_mixed <- emmeans(mixed_anova, ~ program | time)
pairs(emm_mixed, adjust = "tukey")

# Simple effects: time within each program
emm_time <- emmeans(mixed_anova, ~ time | program)
pairs(emm_time, adjust = "tukey")
```

Simulation and Blended programs maintain gains at 3 months, while Traditional shows significant decay.

---

## Activity 8 Solutions

```{r}
#| label: solution8
#| code-fold: true

# Interaction plot
diabetes_mixed %>%
  group_by(group, time) %>%
  summarise(mean = mean(hba1c), .groups = "drop") %>%
  ggplot(aes(x = time, y = mean, color = group, group = group)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "HbA1c Trajectories by Education Program",
       x = "Time Point",
       y = "Mean HbA1c (%)") +
  theme_minimal()

# Mixed ANOVA
diabetes_anova <- aov(hba1c ~ group * time + Error(patient_id/time),
                      data = diabetes_mixed)
summary(diabetes_anova)

# Simple effects
emmeans(diabetes_anova, ~ group | time) %>% pairs(adjust = "tukey")
```

The Intensive Education group shows significant improvement by 6 months, reaching the target HbA1c of 7.0%.

---

## Activity 9 Solutions

```{r}
#| label: solution9
#| code-fold: true

# Interaction plot
therapy_data %>%
  group_by(therapy, session) %>%
  summarise(mean = mean(anxiety), .groups = "drop") %>%
  ggplot(aes(x = session, y = mean, color = therapy, group = therapy)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Anxiety Trajectories by Therapy Type",
       x = "Session",
       y = "Mean Anxiety Score") +
  theme_minimal()

# Mixed ANOVA
therapy_anova <- aov(anxiety ~ therapy * session + Error(patient_id/session),
                     data = therapy_data)
summary(therapy_anova)

# Simple effects
emmeans(therapy_anova, ~ therapy | session) %>% pairs(adjust = "tukey")
```

CBT shows faster initial improvement, but both therapies converge to similar levels by Session 5.

---

## Activity 10 Solutions

```{r}
#| label: solution10
#| code-fold: true

# Interaction plot
safety_data %>%
  group_by(type, time) %>%
  summarise(mean = mean(score), .groups = "drop") %>%
  ggplot(aes(x = time, y = mean, color = type, group = type)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Safety Score Changes by Facility Type",
       x = "Time Point",
       y = "Mean Safety Score") +
  theme_minimal()

# Mixed ANOVA
safety_anova <- aov(score ~ type * time + Error(facility_id/time),
                    data = safety_data)
summary(safety_anova)

# Simple effects
emmeans(safety_anova, ~ type | time) %>% pairs(adjust = "tukey")
```

Rural facilities show the greatest improvement (+16 points), suggesting the initiative is most beneficial where baseline safety infrastructure was weakest.
