---
title: "Mean Comparison 3:"
subtitle: "Two-Way ANOVA"
author: ""
date: "2026-02-25"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    logo: ""
    footer: "Week 7 | Two-Way ANOVA"
    scrollable: true
    incremental: false
    code-line-numbers: false
    highlight-style: github
    navigation-mode: linear
execute:
  warning: false
  message: false
  echo: true
---

```{r setup}
#| include: false
library(ggplot2)
library(tidyverse)
library(emmeans)
set.seed(2026)
```

# Introduction {background-color="#40666e"}

## Where We Left Off

Last week, we learned two ANOVA designs for comparing means across 3+ groups:

::: {.incremental}
- **One-Way ANOVA**: tests whether group means differ across 3+ independent groups
- **Repeated-Measures ANOVA**: compares means when the **same subjects** are measured under multiple conditions
:::

. . .

<br>
Both designs examined the effect of **one factor** at a time.

. . .

But in healthcare, outcomes are rarely influenced by a single factor.

## The Limitation of One Factor

Let's think about this case: 

> You invest $500,000 in a new nurse training program, and the overall competency score improves by 5 points. Sounds like a success, until you discover it boosted novice nurses' scores by 15 points but *decreased* experienced nurses' scores by 5 points.

. . .

<br>

- A one-way ANOVA would only show the modest average improvement (training program = 5 points).

. . .


- **Two-way ANOVA** would have revealed this pattern before hospital-wide rollout (training program x experience levels).


## This Week's Question

This week, we extend ANOVA to examine **two factors simultaneously**:

. . .

<br>

> **Does the effect of one factor depend on the level of another factor?**

. . .

<br>

- This is the concept of an **interaction effect**, the most important idea in this lecture.

. . . 

- In the previous example, the effect of training program depends on the level of nurse experience. 


## Today's Learning Goals

By the end of this lecture, you will be able to:

::: {.incremental}
1. **Distinguish** between main effects and interaction effects
2. **Identify** three types of interaction patterns from interaction plots
3. **Choose** the correct two-way ANOVA design (between, within, mixed)
4. **Conduct** two-way ANOVA in R using `aov()` and interpret the results
5. **Perform** simple effects analysis using `emmeans()` when interactions are significant
:::


# What is Two-Way ANOVA? {background-color="#40666e"}

## Understanding "One-Way" vs. "Two-Way"

In ANOVA, the term "way" refers to the number of **factors** we examine:

. . .

<br> **One-Way ANOVA**: Effect of **one factor** on an outcome

::: {.smaller style="margin-left: 2em;"}
- e.g., Does satisfaction differ across 4 departments?
- *Factor*: Department (4 levels) → *Outcome*: Satisfaction score
:::

. . .

**Two-Way ANOVA**: Effects of **two factors** simultaneously

::: {.smaller style="margin-left: 2em;"}
- e.g., Does pain reduction depend on treatment type AND severity?
- *Factor A*: Treatment, *Factor B*: Severity → *Outcome*: Pain reduction
:::


## Why Study Two Factors Simultaneously?

**Two-way ANOVA** allows us to test three things simultaneously:

::: {.incremental}
1. Test the effect of **Factor A** (main effect of A)
2. Test the effect of **Factor B** (main effect of B)
3. Test whether the effect of A **depends on** B (interaction effect A × B)
:::



## Main Effects vs. Interaction Effects

Suppose we want to understand the effects of treatment and patient group on recovery score:

. . . 

- **Main Effect**: The effect of one factor on the outcome, **averaged across** all levels of the other factor.
  - *Main effect of Treatment*: Does the treatment affect recovery, regardless of patient group?
  - *Main effect of Patient Group*: Do patient groups differ in recovery, regardless of treatment?

. . .

- **Interaction Effect**: The effect of one factor **depends on** the level of another factor.
  - The treatment improves recovery for severe patients but not for mild patients

. . .

**When there's a significant interaction, main effects can be misleading!**

## Overview of Two-Way ANOVA Designs

| Design | Factor A | Factor B | Example |
|:------:|:--------:|:--------:|:-------:|
| **Between-Subjects** | Different subjects | Different subjects | Treatment type × Patient severity |
| **Within-Subjects** | Same subjects | Same subjects | Drug type × Time of day |
| **Mixed Design** | Different subjects | Same subjects | Training program × Assessment time |


# Understanding Interaction Patterns {background-color="#40666e"}

## Pattern 1: Parallel Lines (No Interaction)

:::: {.columns}
::: {.column width="60%"}
```{r}
#| label: no-interaction
#| echo: false
#| fig-width: 6
#| fig-height: 4.5

no_int <- data.frame(
  Treatment = rep(c("Standard", "New"), each = 2),
  Severity = rep(c("Mild", "Severe"), 2),
  Outcome = c(70, 50, 85, 65)
)

ggplot(no_int, aes(x = Severity, y = Outcome, color = Treatment, group = Treatment)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  scale_color_manual(values = c("#E63946", "#457B9D")) +
  labs(title = "No Interaction: Parallel Lines",
       subtitle = "New treatment is equally better (+15 points)",
       x = "Disease Severity", y = "Recovery Score") +
  coord_cartesian(ylim = c(40, 100)) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```
:::
::: {.column width="40%"}
<br>

- The new treatment improves outcomes by **15 points** for *both* mild and severe patients

- The treatment effect is **consistent** regardless of severity

- Lines are **parallel** → No interaction
:::
::::


## Pattern 2: Ordinal Interaction 

:::: {.columns}
::: {.column width="60%"}
```{r}
#| label: ordinal-interaction
#| echo: false
#| fig-width: 6
#| fig-height: 4.5

ord_int <- data.frame(
  Treatment = rep(c("Standard", "New"), each = 2),
  Severity = rep(c("Mild", "Severe"), 2),
  Outcome = c(70, 50, 80, 75)
)

ggplot(ord_int, aes(x = Severity, y = Outcome, color = Treatment, group = Treatment)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  scale_color_manual(values = c("#E63946", "#457B9D")) +
  labs(title = "Ordinal Interaction: Converging Lines",
       subtitle = "New treatment helps severe patients more (+25 vs +10)",
       x = "Disease Severity", y = "Recovery Score") +
  coord_cartesian(ylim = c(40, 100)) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```
:::
::: {.column width="40%"}
<br>

- New treatment helps *both* groups, but helps severe patients **much more**
  - Mild: +10 points
  - Severe: +25 points
- Lines **converge** but don't cross → Ordinal interaction
:::
::::


## Pattern 3: Disordinal (Crossover) Interaction

:::: {.columns}
::: {.column width="60%"}
```{r}
#| label: crossover-interaction
#| echo: false
#| fig-width: 6
#| fig-height: 4.5

cross_int <- data.frame(
  Treatment = rep(c("Standard", "New"), each = 2),
  Severity = rep(c("Mild", "Severe"), 2),
  Outcome = c(75, 50, 65, 80)
)

ggplot(cross_int, aes(x = Severity, y = Outcome, color = Treatment, group = Treatment)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  scale_color_manual(values = c("#E63946", "#457B9D")) +
  labs(title = "Disordinal (Crossover) Interaction",
       subtitle = "New treatment helps severe but harms mild patients",
       x = "Disease Severity", y = "Recovery Score") +
  coord_cartesian(ylim = c(40, 100)) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```
:::
::: {.column width="40%"}
<br>

- New treatment *helps* severe patients (+30) but *harms* mild patients (-10)
- The main effect might appear **small or non-significant** because effects cancel out!
- Lines **cross** → Disordinal interaction
:::
::::


## Why Interactions Matter in Healthcare

::: {.incremental}
1. **Personalized medicine**: Treatment effectiveness often varies by patient characteristics
2. **Resource allocation**: Interventions may work better in some settings than others
3. **Safety**: A treatment safe for one population may be harmful for another
4. **Policy decisions**: Program effectiveness may depend on implementation context
:::


# 1. Two-Way Between-Subjects ANOVA {background-color="#40666e"}

## What is it?

A **two-way between-subjects ANOVA** tests the effects of two factors where **different participants** are in each combination of factor levels.

. . .

<br>
**Key Characteristics**:

::: {.incremental}
- Two factors (each with 2+ levels)
- Each participant is in **only one cell** of the design
- Tests three effects: Main effect A, Main effect B, A × B interaction
:::

## When to Use It

Use when you have two categorical factors and each individual is assigned to **only one** combination:

::: {.incremental}
- **Treatment types** × **patient demographics** → pain scores, recovery time
  - e.g., Comparing opioid vs. non-opioid pain protocols across age groups (18–39, 40–64, 65+)
- **Interventions** × **facility types** → length of stay, readmission rates
  - e.g., Evaluating a discharge planning program across urban, suburban, and rural hospitals
- **Training programs** × **experience levels** → competency scores, error rates
  - e.g., Simulation-based vs. lecture-based training for novice vs. experienced nurses
:::


## Scenario: Treatment × Disease Severity

A clinical trial compares **two treatments** for chronic pain across **three severity levels**.

. . .

- **Treatments:**
  - **Standard Care**: Physical therapy, OTC pain relievers, patient education (~$800/patient)
  - **Novel Multimodal**: CBT for pain, targeted exercise, non-opioid prescription (~$3,200/patient)

. . .

- **Severity Levels:**
  - **Mild** (pain 1-3), **Moderate** (pain 4-6), **Severe** (pain 7-10)

. . .

**Outcome**: Pain reduction score (0-100, higher = better)

## Research Questions

1. Does the novel treatment reduce pain more than standard care? (**main effect of treatment**)

. . .

2. Does pain reduction vary by baseline severity? (**main effect of severity**)

. . .

3. Does the treatment effect depend on severity level? (**interaction**)

. . .

<br>
If the novel treatment only benefits severe patients, prescribing it for everyone would waste the extra $2,400 per patient.

## The Formula

Two-way ANOVA partitions variance into four components:

$$SS_{total} = SS_A + SS_B + SS_{A \times B} + SS_{error}$$

. . .

Three F-statistics are calculated:

$$F_A = \frac{MS_A}{MS_{error}}, \quad F_B = \frac{MS_B}{MS_{error}}, \quad F_{A \times B} = \frac{MS_{A \times B}}{MS_{error}}$$

Degrees of freedom: 

- $df_A = a - 1$ 
- $df_B = b - 1$ 
- $df_{A \times B} = (a-1)(b-1)$
- $df_{error} = N - ab$


## Demonstration: Loading the Package and Data

Let's load the packages and data. 

```{r}
#| label: loading-packages

library(tidyverse)
library(ggplot2)
library(emmeans)
```

```{r}
#| label: between-example1

# Pain reduction scores for 2 treatments × 3 severity levels
pain_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week07_pain_data.csv")

pain_data <- pain_data %>%
  mutate(
    severity = factor(severity, levels = c("Mild", "Moderate", "Severe")),
    treatment = factor(treatment, levels = c("Standard", "Novel"))
  )
```

## Demonstration: Summary Statistics

```{r}
#| echo: false
pain_summary <- pain_data %>%
  group_by(treatment, severity) %>%
  summarise(mean = round(mean(pain_reduction), 1),
            sd = round(sd(pain_reduction), 1),
            .groups = "drop")

knitr::kable(pain_summary,
             col.names = c("Treatment", "Severity", "Mean", "SD"),
             caption = "**Pain Reduction Scores by Treatment and Severity**",
             align = "c")
```


## Visualizing the Data

```{r}
#| label: between-viz1
#| code-fold: true
#| fig-width: 8
#| fig-height: 4.5

pain_means <- pain_data %>%
  group_by(treatment, severity) %>%
  summarise(mean = mean(pain_reduction), .groups = "drop")

ggplot(pain_means, aes(x = severity, y = mean, color = treatment, group = treatment)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Interaction Plot: Treatment × Severity",
       x = "Disease Severity", y = "Mean Pain Reduction",
       color = "Treatment") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```


## The R Formula

The R formula for two-way between-subjects ANOVA uses `*` for main effects plus interaction:

```r
aov(outcome ~ factorA * factorB, data = dataset)
```

. . .

<br>
The `*` is shorthand for `treatment + severity + treatment:severity`

. . .

<br>
**Note**: Unlike repeated-measures ANOVA from Week 6, there is **no `Error()` term** here because each participant appears in only one condition.


## Hypotheses

1. **Main effect of treatment**:
   - $H_0$: $\mu_{Standard} = \mu_{Novel}$
   - $H_A$: $\mu_{Standard} \neq \mu_{Novel}$

. . .

2. **Main effect of severity**:
   - $H_0$: $\mu_{Mild} = \mu_{Moderate} = \mu_{Severe}$
   - $H_A$: At least one severity mean differs

. . .

3. **Interaction**:
   - $H_0$: The treatment effect is the same at every severity level
   - $H_A$: The treatment effect depends on severity level

## Demonstration: Running Two-Way ANOVA

```{r}
#| label: between-test1

# Two-way ANOVA
anova_pain <- aov(pain_reduction ~ treatment * severity, data = pain_data)
summary(anova_pain)
```

## Interpreting the ANOVA Table

The output shows three F-tests:

::: {.incremental}
- **treatment**: Does pain reduction differ between treatments, averaging across severity?
- **severity**: Does pain reduction differ across severity, averaging across treatments?
- **treatment:severity**: Does the treatment difference vary across severity levels?
:::

. . .

<br>
Recall the ANOVA table columns: **Df** (degrees of freedom), **Sum Sq** (variation), **Mean Sq** ($SS/df$), **F value**, **Pr(>F)** (p-value)


## Results Interpretation

::: {.incremental}
- **Main effect of treatment**: Significant (p < 0.05). Pain reduction differs between treatments.
- **Main effect of severity**: Significant (p < 0.05). Pain reduction differs across severity levels.
- **Interaction (treatment:severity)**: Significant (p < 0.05). The treatment effect is not uniform (it differs depending on severity).
:::



## When Interaction is Significant

The main effect tells us the *average* treatment difference collapsed across severity. But the interaction tells us this average is **misleading**.

. . .

<br>
The Novel treatment's advantage:

::: {.smaller style="margin-left: 2em;"}
- Mild patients: ~2 points improvement
- Severe patients: ~12 points improvement
:::

. . .

<br>
An administrator who only looked at the main effect might conclude "the Novel treatment is better for everyone (7 vs. 13)." The interaction reveals the $2,400 extra cost may not be justified for mild cases.

. . .

<br>
**When the interaction is significant, examine simple effects to understand the pattern.**


## Simple Effects Analysis

Simple effects examine the effect of one factor **at each level** of the other factor:

```{r}
#| label: simple-effects1

# Get estimated marginal means
emm <- emmeans(anova_pain, ~ treatment | severity)

# Simple effects: treatment effect at each severity level
pairs(emm, adjust = "tukey")
```

## Interpreting Simple Effects

The output shows the treatment difference (Standard minus Novel) at each severity level:

::: {.incremental}
- **Mild**: Modest improvement; not statistically significant
- **Moderate**: Larger improvement; statistically significant
- **Severe**: Largest improvement; statistically significant
:::

. . .

<br>
This suggests a **targeted prescribing strategy**: prioritize the expensive Novel treatment for moderate-to-severe patients; standard treatment may be sufficient for mild cases.


## Practice 1: Intervention × Facility Type

- A health system evaluates a **care transition program** on 30-day readmission rates (%) across **facility types**.

. . .

**Care Transition Programs:**

::: {.smaller style="margin-left: 2em;"}
- **Standard Discharge**: Basic discharge instructions, prescription reconciliation, follow-up appointment
- **Enhanced Transition**: Standard + transition coordinator, 48-hour call, home medication review, community resource connections
:::

. . .

**Facility Types:**

::: {.smaller style="margin-left: 2em;"}
- **Urban Academic**: Large teaching hospitals with extensive care coordination
- **Suburban Community**: Mid-sized hospitals with standard services
- **Rural Critical Access**: Small hospitals (<25 beds) with limited specialists
:::

## Practice 1: Load the Data

Load the dataset using the code below. 

```{r}
#| label: practice-between1-data

intervention_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week07_intervention_data.csv")
```

**Task 1.0**: Explore the data

```{r}
#| label: practice-between1-explore
#| eval: false

intervention_data %>%
  group_by(___, ___) %>%
  summarise(
    n = n(),
    mean = round(mean(___), 1),
    sd = round(sd(___), 1),
    .groups = "drop"
  )
```

**Question**: Which program-facility combination has the lowest mean readmission rate? Which has the highest?

## Practice 1: Interaction Plot

**Task 1.1**: Create an interaction plot

```{r}
#| label: practice-between1-plot
#| eval: false

plot_data <- intervention_data %>%
  group_by(___, ___) %>%
  summarise(mean = mean(___), .groups = "drop")

ggplot(plot_data, aes(x = ___, y = ___,
                      color = ___, group = ___)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Readmission Rates by Program and Facility Type",
       x = "Care Transition Program",
       y = "Mean Readmission Rate (%)",
       color = "Facility Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Question**: Do the lines appear parallel, converging, or crossing?

## Practice 1: ANOVA and Simple Effects

**Task 1.2**: Run Two-Way ANOVA

```{r}
#| label: practice-between1-anova
#| eval: false

anova_int <- aov(___ ~ ___ * ___, data = intervention_data)
summary(___)
```

. . .

**Task 1.3**: Simple Effects Analysis

```{r}
#| label: practice-between1-simple
#| eval: false

emm_int <- emmeans(___, ~ ___ | ___)
pairs(emm_int, adjust = "tukey")
```

**Questions**:

::: {.smaller}
1. At which facility types does the Enhanced Transition program significantly reduce readmission rates?
2. Which facility type benefits the most? What are the implications for healthcare administrators?
:::


# 2. Two-Way Within-Subjects ANOVA {background-color="#40666e"}

## What is it?

A **two-way within-subjects ANOVA** tests the effects of two factors where the **same participants** experience all combinations of both factors.

. . .

<br>
**Key Characteristics**:

::: {.incremental}
- Same participants measured under **all** conditions
- Both factors are within-subjects (repeated measures)
- More powerful than between-subjects designs
- Requires fewer participants
:::


## When to Use It

Use when the same participants are measured under **all** combinations of two factors:

::: {.incremental}
- Same patients tested with **different medications** at **different times of day** → pain control scores
  - e.g., Each patient tries ibuprofen and acetaminophen in the morning and evening
- Same staff evaluated on **multiple tasks** across **multiple assessment periods** → performance scores
  - e.g., Each nurse is assessed on medication administration and charting at months 1 and 6
- Same clinics measured on **different metrics** in **different seasons** → satisfaction scores
  - e.g., Each clinic's wait time and communication scores are tracked in summer vs. winter
:::


## Scenario: Medication Type × Time of Day

A clinical study examines whether **medication effectiveness** varies by **time of administration**. The same 20 patients receive each of three medications at two times.

<br>

**Medications:**

::: {.smaller style="margin-left: 2em;"}
- **Med A**: Standard pain reliever
- **Med B**: Extended-release pain medication
- **Med C**: Fast-acting pain medication
:::

**Time:** Morning vs. Evening

**Outcome**: Effectiveness score (0-100, higher = better pain control)

. . .

**Study Design**: Each patient experiences all 6 conditions (3 medications × 2 times).

## Demonstration: Loading the Data

Let's load the data using the code below:

```{r}
#| label: within-example1

within_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week07_within_data.csv")

within_data <- within_data %>%
  mutate(medication = factor(medication, levels = c("Med A", "Med B", "Med C")),
         time = factor(time, levels = c("Morning", "Evening")),
         patient_id = factor(patient_id))
```

## Demonstration: Summary Statistics

```{r}
#| echo: false
within_summary <- within_data %>%
  group_by(medication, time) %>%
  summarise(mean = round(mean(effectiveness), 1),
            sd = round(sd(effectiveness), 1),
            .groups = "drop")

knitr::kable(within_summary,
             col.names = c("Medication", "Time", "Mean", "SD"),
             caption = "**Medication Effectiveness by Type and Time of Day**",
             align = "c")
```


## Visualizing the Data

```{r}
#| label: within-viz1
#| code-fold: true
#| fig-width: 8
#| fig-height: 4.5

within_means <- within_data %>%
  group_by(medication, time) %>%
  summarise(mean = mean(effectiveness), .groups = "drop")

ggplot(within_means, aes(x = medication, y = mean, color = time, group = time)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Medication Effectiveness by Time of Administration",
       subtitle = "Medications show different time-of-day patterns",
       x = "Medication", y = "Mean Effectiveness Score",
       color = "Time") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```


## Understanding the Error Term

Since the **same patients** experience all conditions, we account for this using `Error()`:

. . .

- **If the same person experiences all levels of a factor, that factor goes inside `Error(id/...)`**

. . .

<br>
Here, each patient receives all 3 medications AND both time points, so **both** factors go inside `Error()`:

```r
aov(outcome ~ factorA * factorB + Error(subject_id / (factorA * factorB)),
    data = dataset)
```


## Comparing All Formula Types

| Design | Formula | Why? |
|--------|---------|------|
| **Between-subjects** | `aov(y ~ A * B)` | Different people in each cell |
| **One-way repeated** (Week 6) | `aov(y ~ A + Error(id/A))` | Same people across levels of A |
| **Two-way within** (current) | `aov(y ~ A * B + Error(id/(A*B)))` | Same people across ALL combinations |


## Hypotheses

1. **Main effect of medication**:
   - $H_0$: $\mu_{Med A} = \mu_{Med B} = \mu_{Med C}$
   - $H_A$: At least one medication mean differs

. . .

2. **Main effect of time**:
   - $H_0$: $\mu_{Morning} = \mu_{Evening}$
   - $H_A$: $\mu_{Morning} \neq \mu_{Evening}$

. . .

3. **Interaction**:
   - $H_0$: The time effect is the same for all medications
   - $H_A$: The time effect depends on medication


## Demonstration: Running Within-Subjects ANOVA

```{r}
#| label: within-test1

# Two-way within-subjects ANOVA
within_anova <- aov(effectiveness ~ medication * time +
                     Error(patient_id/(medication * time)),
                   data = within_data)
summary(within_anova)
```

## Interpreting the Results

Unlike between-subjects output (one table), you'll see **multiple sections** labeled with different "Error" strata. Focus on **F value** and **Pr(>F)**:

::: {.incremental}
- **medication**: Significant. The three medications produce different effectiveness levels.
- **time**: Significant. Overall effectiveness differs between morning and evening.
- **medication:time**: Significant! The time-of-day effect is **different** for different medications.
:::

. . .

<br>
Although the main effect of time is significant, the significant interaction tells us the time effect is **not uniform**: some medications work better in the morning, others in the evening.


## Simple Effects Analysis

```{r}
#| label: within-posthoc1

# Simple effects: time effect at each medication level
emm_within <- emmeans(within_anova, ~ time | medication)
pairs(emm_within, adjust = "tukey")
```


## Interpreting Simple Effects

The simple effects reveal the time-of-day effect for each medication:

::: {.incremental}
- **Med A** (Standard): Significantly more effective in the **evening**
- **Med B** (Extended-release): **No significant difference** (designed for consistent levels)
- **Med C** (Fast-acting): Significantly more effective in the **morning**
:::

. . .

<br>
These findings support **chronotherapy** (timing medication to optimize effectiveness without changing the medication or increasing costs).


## Practice 2: Task Type × Shift Period

- A nursing director evaluates whether **nurse performance** on different **task types** changes throughout a 12-hour shift.

. . .

**Task Types:**

::: {.smaller style="margin-left: 2em;"}
- **Documentation**: Charting, EHR data entry, admission/discharge paperwork
- **Patient Care**: Vital signs, medication administration, patient education
:::

. . .

**Shift Periods:**

::: {.smaller style="margin-left: 2em;"}
- **Beginning** (hours 1-3), **Middle** (hours 4-9), **End** (hours 10-12)
:::

. . .

The same 15 nurses are evaluated on both task types across all three periods.


## Practice 2: Load the Data

Let's load the dataset and explore it. 

```{r}
#| label: practice-within1-data

task_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week07_task_data.csv")

task_data <- task_data %>%
  mutate(
    task = factor(task, levels = c("Documentation", "Patient Care")),
    period = factor(period, levels = c("Beginning", "Middle", "End")),
    nurse_id = factor(nurse_id)
  )
```

**Task 2.0**: Explore the data

```{r}
#| label: practice-within1-explore
#| eval: false

task_data %>%
  group_by(___, ___) %>%
  summarise(
    mean = round(mean(___), 1),
    sd = round(sd(___), 1),
    .groups = "drop"
  )
```


## Practice 2: Interaction Plot

**Task 2.1**: Create an interaction plot

```{r}
#| label: practice-within1-plot
#| eval: false

task_means <- task_data %>%
  group_by(___, ___) %>%
  summarise(mean = mean(___), .groups = "drop")

ggplot(task_means, aes(x = ___, y = ___, color = ___, group = ___)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Performance by Task Type and Shift Period",
       x = "Shift Period",
       y = "Mean Performance Score",
       color = "Task Type") +
  theme_minimal()
```

**Questions**:

1. Does one task type decline more than the other over the shift?
2. What type of interaction does this represent (no interaction, ordinal, or disordinal)?


## Practice 2: ANOVA and Simple Effects

**Task 2.2**: Run Within-Subjects ANOVA

```{r}
#| label: practice-within1-anova
#| eval: false

task_anova <- aov(___ ~ ___ * ___ + Error(___/(___ * ___)),
                  data = task_data)
summary(___)
```

. . .

**Task 2.3**: Simple Effects Analysis

```{r}
#| label: practice-within1-simple
#| eval: false

emm_task <- emmeans(___, ~ ___ | ___)
pairs(emm_task, adjust = "tukey")
```

::: {.smaller}
**Questions**:

1. At which shift periods is there a significant difference between documentation and patient care performance?
2. Does documentation decline more than patient care? What would you recommend to the nursing director?
:::


# 3. Mixed Design ANOVA {background-color="#40666e"}

## What is it?

A **mixed design ANOVA** combines both between-subjects and within-subjects factors. This is the **most common design in healthcare research**.

. . .

<br>
**Key Characteristics**:

::: {.incremental}
- **Between-subjects factor**: Different participants in each group (e.g., treatment groups)
- **Within-subjects factor**: Same participants measured repeatedly (e.g., time points)
- The interaction is usually the most important test: do the groups show *different trajectories*?
:::


## When to Use It

Use when different groups of participants are measured **repeatedly**:

::: {.incremental}
- **Clinical trials**: Treatment group × Time points → symptom severity
  - e.g., Drug vs. placebo groups measured at baseline, 4 weeks, and 12 weeks
- **Patient experience**: Hospital unit × Survey domains → satisfaction scores
  - e.g., ICU vs. med-surg patients surveyed on communication, responsiveness, and discharge 
- **Quality improvement**: Intervention group × Measurement periods → readmission rates
  - e.g., Hospitals with vs. without a care coordinator tracked quarterly for one year
:::


## Scenario: Training Program × Assessment Time

A hospital compares **three training programs** for nursing competency. 60 nurses are randomly assigned to one program (between) and assessed at **three time points** (within).

. . .

**Training Programs:**

::: {.smaller style="margin-left: 2em;"}
- **Traditional Lecture**: Classroom instruction (~$200/nurse)
- **Simulation-Based**: Hands-on practice with mannequins (~$800/nurse)
- **Blended Learning**: Online theory + in-person practice (~$600/nurse)
:::

. . .

**Time Points**: Pre, Post, 3 Months

. . .

**Key Management Question**: *Do more expensive methods maintain their advantage over time?*


## Demonstration: Loading the Data

Let's load the dataset using the code below. 

```{r}
#| label: mixed-example1

mixed_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week07_mixed_data.csv")

mixed_data <- mixed_data %>%
  mutate(
    program = factor(program, levels = c("Traditional", "Simulation", "Blended")),
    time = factor(time, levels = c("Pre", "Post", "3 Months")),
    nurse_id = factor(nurse_id)
  )
```


## Visualizing the Data

```{r}
#| label: mixed-viz1
#| code-fold: true
#| fig-width: 8
#| fig-height: 4.5

mixed_means <- mixed_data %>%
  group_by(program, time) %>%
  summarise(mean = mean(score), .groups = "drop")

ggplot(mixed_means, aes(x = time, y = mean, color = program, group = program)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Competency Trajectories by Training Program",
       x = "Assessment Time", y = "Mean Competency Test Score",
       color = "Program") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```


## Understanding the Error Term

Apply the same rule: **Does the same person experience all levels of this factor?**

::: {.incremental}
- **Program** (between-subjects): Each nurse is in *only one* program → **NOT** in `Error()`
- **Time** (within-subjects): Each nurse is tested at *all three* time points → **in** `Error()`
:::

. . .

<br>

```r
aov(outcome ~ between_factor * within_factor + Error(subject_id / within_factor),
    data = dataset)
```

## Complete Formula Comparison

| Design | Formula | Error Term Contains |
|--------|---------|-------------------|
| **Between-subjects** | `aov(y ~ A * B)` | Nothing (no repeated measures) |
| **Within-subjects** | `aov(y ~ A * B + Error(id/(A*B)))` | Both A and B |
| **Mixed** | `aov(y ~ A * B + Error(id/B))` | Only within-subjects factor B |


## Hypotheses

1. **Main effect of program**:
   - $H_0$: $\mu_{Traditional} = \mu_{Simulation} = \mu_{Blended}$
   - $H_A$: At least one program mean differs

. . .

2. **Main effect of time**:
   - $H_0$: $\mu_{Pre} = \mu_{Post} = \mu_{3 Months}$
   - $H_A$: At least one time point mean differs

. . .

3. **Interaction**:
   - $H_0$: The time trajectory is the same for all programs
   - $H_A$: The time trajectory differs across programs


## Demonstration: Running Mixed Design ANOVA

```{r}
#| label: mixed-test1

# Mixed design ANOVA
mixed_anova <- aov(score ~ program * time + Error(nurse_id/time),
                   data = mixed_data)
summary(mixed_anova)
```


## Interpreting the Results

The output has two error strata:

::: {.incremental}
- **program**: Main effect of training program (Significant)
- **time**: Main effect of time (Significant): competency scores change over time
- **program:time**: The interaction (Significant): programs show *different trajectories*!
:::

. . .

<br>

- Traditional starts equal but falls behind
- Simulation and Blended maintain their gains

## When Interaction is Significant

Without the interaction, an administrator might conclude all programs need refresher courses equally.

. . .

The interaction reveals:

::: {.smaller style="margin-left: 2em;"}
- Traditional: significant decay at 3 months
- Simulation & Blended: retain most of their gains
:::

. . .

<br>
**When the interaction is significant, examine simple effects to understand the pattern.**


## Simple Effects: Programs at Each Time

```{r}
#| label: mixed-simple1

# Compare programs at each time point
emm_mixed <- emmeans(mixed_anova, ~ program | time)
pairs(emm_mixed, adjust = "tukey")
```

## Interpreting: Programs at Each Time

At each time point:

::: {.incremental}
- **Pre**: No significant differences — all three groups started equal
- **Post**: Significant differences (Blended > Simulation > Traditional)
- **3 Months**: Significant differences persist (Blended > Simulation > Traditional)
:::


## Practice 3: Therapy Type × Session Number

- A mental health clinic compares **two therapy approaches** for anxiety. 36 patients are randomly assigned to one therapy (between) and measured at **three time points** (within).

. . .

**Therapies:**

::: {.smaller style="margin-left: 2em;"}
- **CBT**: Identifies and changes negative thought patterns; includes homework
- **Mindfulness-Based**: Focuses on present-moment awareness; includes meditation
:::

. . .

**Time Points**: Session 1, Session 3, Session 5

. . .

**Outcome**: Anxiety score (0-100, higher = more severe)


## Practice 3: Load the Data

Load the data using the code below and explore the data.

```{r}
#| label: practice-mixed1-data

therapy_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week07_therapy_data.csv")

therapy_data <- therapy_data %>%
  mutate(
    session = factor(session, levels = c("Session 1", "Session 3", "Session 5")),
    therapy = factor(therapy, levels = c("CBT", "Mindfulness")),
    patient_id = factor(patient_id)
  )
```

**Task 3.0**: Explore the data

```{r}
#| label: practice-mixed1-explore
#| eval: false

therapy_data %>%
  group_by(___, ___) %>%
  summarise(
    mean = round(mean(___), 1),
    sd = round(sd(___), 1),
    .groups = "drop"
  )
```


## Practice 3: Interaction Plot

**Task 3.1**: Create an interaction plot

```{r}
#| label: practice-mixed1-plot
#| eval: false

therapy_means <- therapy_data %>%
  group_by(___, ___) %>%
  summarise(mean = mean(___), .groups = "drop")

ggplot(therapy_means, aes(x = ___, y = ___, color = ___, group = ___)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  labs(title = "Anxiety Trajectories by Therapy Type",
       x = "Session",
       y = "Mean Anxiety Score",
       color = "Therapy") +
  theme_minimal()
```

**Questions**:

1. Do both therapies show the same rate of improvement?
2. Which therapy shows faster initial improvement?


## Practice 3: ANOVA and Simple Effects

**Task 3.2**: Run Mixed Design ANOVA

```{r}
#| label: practice-mixed1-anova
#| eval: false

# Between: therapy, Within: session
therapy_anova <- aov(___ ~ ___ * ___ + Error(___/___),
                     data = therapy_data)
summary(___)
```

. . .

**Task 3.3**: Simple Effects Analysis

```{r}
#| label: practice-mixed1-simple
#| eval: false

emm_therapy <- emmeans(___, ~ ___ | ___)
pairs(emm_therapy, adjust = "tukey")
```

::: {.smaller}
**Questions**:

1. At which sessions do the two therapy types differ significantly?
2. Do the therapies converge by Session 5? What would you tell patients about choosing between therapies?
:::


# Summary {background-color="#40666e"}

## Key Takeaways: Design Formulas

| Design | Between Factor | Within Factor | R Formula |
|--------|:-:|:-:|-----------|
| **Between** | A, B | — | `aov(y ~ A * B)` |
| **Within** | — | A, B | `aov(y ~ A*B + Error(id/(A*B)))` |
| **Mixed** | A | B | `aov(y ~ A*B + Error(id/B))` |


## Key Takeaways: Interaction Patterns

| Pattern | Lines | Interpretation |
|---------|-------|----------------|
| **No interaction** | Parallel | Effect of A is same at all levels of B |
| **Ordinal** | Converge/diverge | Effect of A varies in magnitude but not direction |
| **Disordinal** | Cross | Effect of A reverses direction across levels of B |


## Choosing the Right Design

Ask yourself two questions about each factor:

::: {.incremental}
1. **"Do different people experience different levels?"** → Between-subjects factor
2. **"Does each person experience all levels?"** → Within-subjects factor
:::

. . .

Then match:

::: {.incremental}
- Both between → **Two-Way Between-Subjects**
- Both within → **Two-Way Within-Subjects**
- One of each → **Mixed Design**
:::


## Interpreting Results: Step-by-Step

::: {.incremental}
1. **Check the interaction first**
   - Significant → Focus on simple effects. **Do not** interpret main effects in isolation.
   - Not significant → Interpret main effects directly.

2. **For significant interactions, run simple effects**
   - `emmeans()` with `pairs()` to compare at each level

3. **Connect to practical implications**
   - What does this mean for clinical practice or resource allocation?
   - Which subgroups benefit most?
:::


## Assignments

1. **Reflection**: Please share your thoughts on the lecture
   - 3 interesting things
   - 3 confusing things

2. **Lab Assignment 3**


# References {background-color="#40666e"}

## Statistical Methods

- Field, A. (2018). *Discovering Statistics Using IBM SPSS Statistics* (5th ed.). SAGE Publications.
- Maxwell, S. E., & Delaney, H. D. (2004). *Designing Experiments and Analyzing Data* (2nd ed.). Psychology Press.
- [Navarro, D. (2019). *Learning Statistics with R*.](https://learningstatisticswithr.com/lsr-0.6.pdf){target="_blank"}
- Wickham, H., & Grolemund, G. (2017). *R for Data Science*. O'Reilly Media.

## Healthcare Applications

Examples are motivated by the following:

- Interpretation of Chronic Pain Clinical Trial Outcomes: IMMPACT Recommended Considerations. (2020). *Pain*. https://pmc.ncbi.nlm.nih.gov/articles/PMC7572524/

- Effectiveness of a Simulation-Based Education Program to Improve Novice Nurses' Clinical Judgment Skills. (2024). *Cureus*. https://pmc.ncbi.nlm.nih.gov/articles/PMC11223944/

- Circadian Variation in Efficacy of Medications. (2020). *Clinical Pharmacology & Therapeutics*. https://pmc.ncbi.nlm.nih.gov/articles/PMC8268638/

- Mindfulness-Based Interventions for Anxiety and Depression. (2017). *The Psychiatric Clinics of North America*. https://pmc.ncbi.nlm.nih.gov/articles/PMC5679245/
