---
title: "Linear Regression in Healthcare Analytics"
subtitle: "Part 2: Dummy Variables and Interaction Effects"
author: "MBA642: Business Analytics in Healthcare Management"
Week: "Week 11"
date: 01/01/2026
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: false
    code-copy: true
    theme: custom-theme.scss
    highlight-style: github
    mainfont: "Clear Sans, -apple-system, system-ui, sans-serif"
    fontsize: 1.1em
    grid:
      sidebar-width: 200px
      body-width: 900px
      margin-width: 300px
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Load required packages
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
library(car)
library(emmeans)
library(interactions)
library(patchwork)

# Set seed for reproducibility
set.seed(2026)
```

# Introduction {.section-time data-time="10 min"}

## Moving Beyond Continuous Predictors

In our last lecture, we focused primarily on continuous predictors (e.g., age, severity scores) in linear regression. However, many important variables in healthcare research are **categorical**:

- **Hospital type**: Teaching vs. Community vs. Critical Access
- **Insurance type**: Medicare, Medicaid, Private, Uninsured
- **Treatment protocol**: Standard Care vs. Enhanced Recovery
- **Region**: Northeast, South, Midwest, West

This section covers two essential extensions to linear regression:

1. **Dummy Variables**: How to handle and include categorical predictors in regression
2. **Interaction Effects**: Whether the effect of one variable depends on another

## Why This Matters

Understanding these techniques allows you to:

- Compare outcomes across hospital types or treatment groups
- Determine if an intervention works differently for different patient populations
- Make evidence-based decisions about resource allocation across categories
- Design targeted interventions for specific subgroups

------------------------------------------------------------------------

# Dummy Variables (Indicator Variables) {.section-time data-time="30 min"}

## What is it?

Dummy variables (also called indicator variables) are binary variables that represent categorical group membership using 0s and 1s. They allow us to include categorical predictors in regression models, which mathematically require numerical inputs.

For a categorical variable with **k categories**, we create **k-1 dummy variables**, with one category serving as the reference group against which all others are compared.

For example, if we have a variable "Hospital Type" with three categories (Teaching, Community, Critical Access), we would create two dummy variables:

- Dummy 1: Teaching (1 if Teaching, 0 otherwise)
- Dummy 2: Critical Access (1 if Critical Access, 0 otherwise)

Community hospitals would be the reference category (both dummies = 0).

Dummy 1 captures the difference between Teaching and Community hospitals (the reference category), while Dummy 2 captures the difference between Critical Access and Community hospitals.

## When to Use It

Use dummy variables when you need to:

- Include categorical predictors in regression (hospital type, insurance status, treatment group)
- Compare outcomes across different categories while controlling for other variables
- Test whether group differences are statistically significant
- Quantify the magnitude of differences between categories

## The Challenge with Categorical Variables

Regression requires numerical inputs, but categorical variables are not inherently numerical. Simply assigning numbers (1, 2, 3) to categories implies an ordering and equal spacing that usually doesn't exist.

**Solution**: Create **dummy variables** that represent category membership using 0s and 1s.

## Binary Categorical Variables

For a variable with **two categories**, we need **one dummy variable**:

| Patient | Insurance Status | Dummy: Insured |
|---------|-----------------|----------------|
| 1 | Insured | 1 |
| 2 | Uninsured | 0 |
| 3 | Insured | 1 |
| 4 | Uninsured | 0 |


The coefficient on the dummy variable (Insured) represents the difference in the outcome between the two groups (Insured vs. Uninsured).


### Dummy Variable Example Case: Treatment Protocol Comparison

#### Background and Context

**The problem**: Traditional surgical recovery involves extended hospital stays, which are costly and can lead to complications.

**A better approach**: Enhanced Recovery After Surgery (ERAS) protocols use evidence-based practices to help patients recover faster. These include:

- Better pain management (using less opioid medication)
- Getting patients moving sooner after surgery
- Starting normal eating earlier

**Research evidence**: A 2025 study found that ERAS protocols significantly reduce hospital length of stay compared to traditional care, with patients recovering and being discharged substantially earlier.

Let's compare patient outcomes between standard care and ERAS protocols to quantify the treatment effect using a dummy variable.

#### Creating Synthetic Healthcare Data

We'll create a dataset with 200 surgical patients that includes:

**Variables in the dataset**:

- `patient_id`: Unique identifier for each patient (1 to 200)
- `protocol`: Treatment protocol assigned (Standard or ERAS)
- `age`: Patient age in years
- `bmi`: Body Mass Index (measure of body fat based on height and weight)
- `surgery_complexity`: Complexity level of the surgery (1 = Low, 2 = Medium, 3 = High)
- `length_of_stay`: Hospital length of stay in days (our outcome variable)

Let's create the dataset using the code below:

```{r}
#| label: binary-dummy-data

# Generate surgical patient data
n <- 200

surgical_data <- tibble(
  patient_id = 1:n,
  protocol = sample(c("Standard", "ERAS"), n, replace = TRUE),
  age = round(rnorm(n, 62, 12)),
  bmi = round(rnorm(n, 28, 5), 1),
  surgery_complexity = sample(1:3, n, replace = TRUE)  # 1=Low, 2=Medium, 3=High
) %>%
  mutate(
    age = pmax(30, pmin(age, 85)),
    bmi = pmax(18, pmin(bmi, 45)),
    # ERAS reduces LOS
    length_of_stay = round(
      3 +
      0.03 * age +
      0.05 * bmi +
      1.5 * surgery_complexity +
      -1.8 * (protocol == "ERAS") +  # ERAS effect
      rnorm(n, 0, 1),
      1
    ),
    length_of_stay = pmax(1, length_of_stay)
  )

# Summary by protocol
surgical_data %>%
  group_by(protocol) %>%
  summarise(
    n = n(),
    mean_los = round(mean(length_of_stay), 2),
    sd_los = round(sd(length_of_stay), 2),
    min_los = min(length_of_stay),
    max_los = max(length_of_stay)
  ) %>%
  kable(caption = "Length of Stay by Treatment Protocol") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The variable `protocol` is our binary categorical variable (Standard vs. ERAS). We'll create a dummy variable for ERAS (1 = ERAS, 0 = Standard) to include in our regression model.

We can do this using the `factor()` function in R, which automatically creates the necessary dummy variable.

```{r}
#| label: make-binary-dummy

# R automatically creates dummy variables for factors
surgical_data <- surgical_data %>%
  mutate(protocol = factor(protocol, levels = c("Standard", "ERAS")))
```


### Fitting the Model

We can fit a linear regression model using the `lm()` function in R. When we include the variable `protocol` in the model, which is a factor, R automatically creates the dummy variable for us.

```{r}
#| label: binary-dummy-model

# Fit model
model_protocol <- lm(length_of_stay ~ protocol + age + bmi + surgery_complexity,
                     data = surgical_data)

# Extract coefficients
coeffs <- coef(model_protocol)

summary(model_protocol)
```

### Understanding the Dummy Variable Output

```{r}
#| label: interpret-binary
#| echo: false

# Extract coefficients
coefs <- tidy(model_protocol, conf.int = TRUE)

coefs %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = "Regression Results: Protocol Comparison") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


**Key Interpretation**:

The coefficient for `protocolERAS` is approximately `r round(coef(model_protocol)["protocolERAS"], 2)`.

This means: **Patients in the ERAS protocol have a length of stay that is `r abs(round(coef(model_protocol)["protocolERAS"], 2))` days shorter than patients in the Standard protocol, holding age, BMI, and surgery complexity constant.**

The Standard protocol is the **reference category** (omitted from the output). All comparisons are made relative to this reference group.

### Manual Calculation with Dummy Variables

Let's manually calculate predicted length of stay to understand how dummy variables work in the regression equation.

**The fitted equation is**:

$\widehat{\text{LOS}} = \beta_0 + \beta_1(\text{ERAS}) + \beta_2(\text{Age}) + \beta_3(\text{BMI}) + \beta_4(\text{Complexity})$

where ERAS is a dummy variable:

- ERAS = 0 for Standard protocol
- ERAS = 1 for ERAS protocol


**With the fitted coefficients**, this becomes:

$\widehat{\text{LOS}} = `r round(coeffs[1], 3)` + `r round(coeffs[2], 3)`(\text{ERAS}) + `r round(coeffs[3], 3)`(\text{Age}) + `r round(coeffs[4], 3)`(\text{BMI}) + `r round(coeffs[5], 3)`(\text{Complexity})$

Let's get predicted LOS for both protocols for a patient who is 65 years old, has a BMI of 30, and underwent a medium complexity surgery (complexity = 2).


```{r}
#| label: manual-calc-dummy

# get coefficients
coefs <- tidy(model_protocol, conf.int = TRUE)
coeffs <- coef(model_protocol)

# Example patient: 65 years old, BMI 30, complexity=2
age_val <- 65
bmi_val <- 30
complexity_val <- 2

# Scenario 1: Standard protocol (dummy = 0)
pred_standard <- coeffs[1] +                    # Intercept
                 0 * coeffs[2] +                # protocolERAS = 0
                 age_val * coeffs[3] +          # Age effect
                 bmi_val * coeffs[4] +          # BMI effect
                 complexity_val * coeffs[5]     # Complexity effect

cat("Predicted LOS (Standard protocol):", round(pred_standard, 2), "days\n")

# Scenario 2: ERAS protocol (dummy = 1)
pred_eras <- coeffs[1] +                    # Intercept
             1 * coeffs[2] +                # protocolERAS = 1
             age_val * coeffs[3] +          # Age effect
             bmi_val * coeffs[4] +          # BMI effect
             complexity_val * coeffs[5]     # Complexity effect

cat("Predicted LOS (ERAS protocol):    ", round(pred_eras, 2), "days\n")
```


```{r}
#| label: dummy-diff
#| echo: false

# The difference is exactly the dummy variable coefficient
cat("\nDifference:", round(pred_eras - pred_standard, 2), "days")
cat("\nCoefficient for ERAS:", round(coeffs[2], 2), "days")
cat("\nThese should match!")
```


The dummy variable coefficient ($\beta_1$) represents the difference in predicted outcomes between the two groups, holding all other variables constant.

```{r}
#| label: fig-protocol-effect
#| fig-cap: "Comparison of length of stay by treatment protocol"
#| fig-height: 4
#| fig-width: 6
#| code-fold: true

# Visualize the effect
ggplot(surgical_data, aes(x = protocol, y = length_of_stay, fill = protocol)) +
  geom_boxplot(alpha = 0.7, width = 0.4) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  scale_fill_manual(values = c("Standard" = "coral", "ERAS" = "steelblue")) +
  labs(
    x = "Treatment Protocol",
    y = "Length of Stay (days)",
    title = "ERAS Protocol Reduces Hospital Length of Stay"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

### Practice Checkpoint 1 {.unnumbered}

::: {.callout-important}
## Try It Yourself! (5 minutes)

**Pause here!** Open `week12_linear-regression2_practice.qmd` and complete:

- **Task 1.1**: Fit a regression model with binary dummy variables (ERAS example)
- **Task 1.2**: Make predictions and interpret dummy coefficients

This will verify you understand how to create and interpret binary dummy variables.

**After 5 minutes**, we'll continue with the lecture.
:::

*→ We'll practice multi-level dummy variables in Tasks 1.3 and 1.4*

## Multi-Level Categorical Variables

When a categorical variable has **more than two levels**, we need **k-1 dummy variables** (where k is the number of categories).

### Why k-1 and Not k?

Including a dummy for every category would create **perfect multicollinearity** (the dummy variables would sum to 1 for every observation, which is perfectly correlated with the intercept).


### Multi-Level Categorical Variables Example: Hospital Type Comparison

#### Background and Context

**The context**: Hospitals come in different types:

- **Teaching hospitals**: Large academic medical centers that train medical residents and handle complex cases
- **Community hospitals**: General hospitals serving local populations
- **Critical Access hospitals**: Small rural hospitals in remote areas with limited resources

**Question**: Do 30-day readmission rates (the percentage of patients who return to the hospital within 30 days) differ across these hospital types?

- A 2017 study examining nearly 4,500 U.S. hospitals found that readmission rates vary by hospital type, reflecting differences in patient complexity, resources, and care patterns.

Let's compare 30-day readmission rates across three hospital types while controlling for patient characteristics and staffing levels using dummy variables.

#### Creating Synthetic Healthcare Data

Let's create a dataset using the code below:

```{r}
#| label: multi-dummy-data

# Generate hospital data
n_hospitals <- 250

hospital_type_data <- tibble(
  hospital_id = 1:n_hospitals,
  hospital_type = sample(c("Teaching", "Community", "Critical Access"),
                        n_hospitals, replace = TRUE,
                        prob = c(0.25, 0.50, 0.25)),
  bed_count = case_when(
    hospital_type == "Teaching" ~ round(rnorm(n_hospitals, 400, 100)),
    hospital_type == "Community" ~ round(rnorm(n_hospitals, 200, 80)),
    hospital_type == "Critical Access" ~ round(rnorm(n_hospitals, 40, 15))
  ),
  avg_patient_severity = round(runif(n_hospitals, 1, 3), 2),
  nurse_per_patient = round(runif(n_hospitals, 2, 5), 2)
) %>%
  mutate(
    bed_count = pmax(10, bed_count),
    # Generate readmission rate
    readmission_rate = round(
      15 +
      2.5 * avg_patient_severity +
      -0.8 * nurse_per_patient +
      case_when(
        hospital_type == "Teaching" ~ -2.5,       # Reference in effect
        hospital_type == "Community" ~ 0,
        hospital_type == "Critical Access" ~ 3.0
      ) +
      rnorm(n_hospitals, 0, 2),
      1
    ),
    readmission_rate = pmax(5, pmin(readmission_rate, 30))
  )

# Summary by hospital type
hospital_type_data %>%
  group_by(hospital_type) %>%
  summarise(
    n = n(),
    mean_readmission = round(mean(readmission_rate), 2),
    sd_readmission = round(sd(readmission_rate), 2)
  ) %>%
  kable(caption = "Readmission Rates by Hospital Type") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


### Fitting the Model with Multiple Categories

Before fitting the model, let's make the variable `hospital_type` as a factor with "Community" as the reference category using the `factor()` function.

```{r}
#| label: creating-multilevel-dummy
hospital_type_data <- hospital_type_data |>
  mutate(
    hospital_type = factor(hospital_type, levels = c("Community", "Teaching", "Critical Access"))
    )
```

We can fit a linear regression model using the `lm()` function in R. When we include the variable `hospital_type` in the model, which is a factor, R automatically creates the dummy variables for us.

```{r}
#| label: multi-dummy-model

# Fit model
model_hospital <- lm(readmission_rate ~ hospital_type + avg_patient_severity +
                     nurse_per_patient, data = hospital_type_data)

summary(model_hospital)
```

### Understanding the Output

```{r}
#| label: interpret-multi
#| echo: false

# Extract and format coefficients
hospital_coefs <- tidy(model_hospital, conf.int = TRUE) %>%
  mutate(
    term = case_when(
      term == "(Intercept)" ~ "Intercept",
      term == "hospital_typeTeaching" ~ "Hospital Type: Teaching",
      term == "hospital_typeCritical Access" ~ "Hospital Type: Critical Access",
      term == "avg_patient_severity" ~ "Average Patient Severity",
      term == "nurse_per_patient" ~ "Nurse-to-Patient Ratio"
    )
  )

hospital_coefs %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = "Regression Results: Hospital Type Effect on Readmissions") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Interpretation**:

```{r}
#| echo: false
coef_teach <- round(coef(model_hospital)["hospital_typeTeaching"], 3)
coef_crit <- round(coef(model_hospital)["hospital_typeCritical Access"], 3)
```

- **Reference Category**: Community hospitals (omitted from output)
- **Teaching Hospitals**: Readmission rate of Teaching Hospitals is `r abs(coef_teach)` **percentage points lower** than Community hospitals (β = `r coef_teach`)
- **Critical Access Hospitals**: Readmission rate of Critical Access Hospitals is `r coef_crit` **percentage points higher** than Community hospitals (β = `r coef_crit`)

**To compare Teaching vs. Critical Access hospitals**: The difference is `r coef_teach` - `r coef_crit` = `r coef_teach - coef_crit` percentage points. But is this difference statistically significant?

### Testing Comparisons Between Non-Reference Categories

The regression output provides p-values for comparisons against the **reference category** (Community hospitals), but what about comparing Teaching vs. Critical Access hospitals directly?

We need to test whether the **linear combination** of coefficients is statistically significant. We can use the `emmeans()` function with the `pairs()` function for pairwise comparisons.

```{r}
#| label: emmeans-pairwise

# Calculate estimated marginal means
emm <- emmeans(model_hospital, "hospital_type")

# Pairwise comparisons with adjustment for multiple testing
pairs(emm)
```

This provides all pairwise comparisons with confidence intervals and p-values:

- Teaching vs. Community
- Critical Access vs. Community
- Teaching vs. Critical Access (this is what we want!)

The `emmeans` approach is particularly useful because it:

1. Shows all pairwise comparisons at once
2. Adjusts p-values for multiple testing (using Tukey method by default)
3. Provides confidence intervals for the differences

Based on these tests, we can conclude whether Teaching hospitals have **statistically significantly** lower readmission rates than Critical Access hospitals, not just numerically lower rates.


### Changing the Reference Category

You can change the reference category to make different comparisons more intuitive by using the `relevel()` function in R.

```{r}
#| label: relevel

# Make Teaching the reference
hospital_type_data_reref <- hospital_type_data %>%
  mutate(hospital_type = relevel(hospital_type, ref = "Teaching"))

model_reref <- lm(readmission_rate ~ hospital_type + avg_patient_severity +
                  nurse_per_patient, data = hospital_type_data_reref)

tidy(model_reref) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = "Results with Teaching as Reference Category") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


Now the coefficients show comparisons relative to Teaching hospitals.

### Practice Checkpoint 2 {.unnumbered}

::: {.callout-important}
## Try It Yourself! (7 minutes)

**Pause here!** Open `week12_linear-regression2_practice.qmd` and complete:

- **Task 1.3**: Fit a regression model with multi-level categorical variables
- **Task 1.4**: Use `emmeans()` for pairwise comparisons

This will help you practice working with categorical variables that have more than two levels.

**After 7 minutes**, we'll continue with interaction effects.
:::

*→ We'll apply these concepts to more complex scenarios in the advanced exercises*

------------------------------------------------------------------------

# Interaction Effects {.section-time data-time="40 min"}

## What is it?

An **interaction effect** occurs when the effect of one variable on the outcome **depends on the value of another variable**. In other words, the relationship between X₁ and Y changes across different levels of X₂.

Mathematically, we add an interaction term that is the product of the two variables:

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 (X_1 \times X_2) + \epsilon$$

The interaction coefficient $\beta_3$ captures how the effect of $X_1$ changes across levels of $X_2$ (and vice versa).

## When to Use It

Use interaction terms when:

- Theory or prior research suggests an effect should vary across groups
- You observe different slopes for different groups in exploratory visualizations
- You want to test whether an intervention works differently for different populations


## Why Interactions Matter in Healthcare

Many healthcare relationships are conditional. For example:

- **Treatment effectiveness varies by patient characteristics**: A medication may work better for younger patients compared to older patients
- **Resource effects depend on hospital type**: Additional staffing may have different impacts in big vs. small hospitals
- **Policy impacts vary by region**: A new payment model may affect urban and rural hospitals differently

## Continuous × Categorical Interaction

### Continuous × Categorical Interaction Example: Does Staffing Effect Vary by Hospital Type?

#### Background and Context

**Question**: Does improving nurse staffing increase patient satisfaction equally in all hospitals?

**Why it might differ**: Hospitals have different ownership structures:

- **For-Profit hospitals**: Owned by investors, focused on financial performance
- **Non-Profit hospitals**: Community-owned, mission-driven, may reinvest more in quality

**The hypothesis**: Additional nurses might have a bigger impact on satisfaction in non-profit hospitals if they have stronger cultures of patient-centered care.

- Studies have found that nurse staffing levels significantly affect patient outcomes and satisfaction, though the magnitude of these effects may vary by hospital characteristics including ownership type.

Let's test whether the effect of nurse staffing on patient satisfaction varies by ownership type (this is a continuous (staffing) × categorical (hospital type) interaction).

#### Creating Synthetic Healthcare Data

Let's create a dataset using the code below.

```{r}
#| label: interaction-data

# Generate data for interaction example
n <- 300

satisfaction_data <- tibble(
  hospital_id = 1:n,
  ownership = sample(c("For-Profit", "Non-Profit"), n, replace = TRUE),
  nurse_ratio = round(runif(n, 2, 6), 2),
  patient_volume = round(rnorm(n, 500, 150)),
  avg_acuity = round(runif(n, 1, 3), 2)
) %>%
  mutate(
    patient_volume = pmax(100, patient_volume),
    # Interaction: staffing matters more in non-profit hospitals
    patient_satisfaction = round(
      50 +
      3 * nurse_ratio +                              # Base staffing effect
      5 * (ownership == "Non-Profit") +              # Non-profit baseline higher
      2.5 * nurse_ratio * (ownership == "Non-Profit") + # INTERACTION
      -0.01 * patient_volume +
      -3 * avg_acuity +
      rnorm(n, 0, 5),
      1
    ),
    patient_satisfaction = pmax(0, pmin(patient_satisfaction, 100)),
    ownership = factor(ownership, levels = c("For-Profit", "Non-Profit"))
  )
```

**Variables in the dataset**:

- `hospital_id`: Unique identifier for each hospital (1 to 300)
- `ownership`: Hospital ownership type (For-Profit or Non-Profit)
- `nurse_ratio`: Nurse-to-patient ratio (ranging from 2 to 6)
- `patient_volume`: Annual patient volume (number of patients)
- `avg_acuity`: Average patient acuity level (1 = low, 3 = high complexity)
- `patient_satisfaction`: Patient satisfaction score on 0-100 scale (our outcome variable)

### Visualizing the Potential Interaction

**Before running the model**, always visualize to see if an interaction might exist:

```{r}
#| label: fig-interaction-visual
#| fig-cap: "Potential interaction: different slopes suggest an interaction effect"
#| fig-height: 4
#| fig-width: 6

ggplot(satisfaction_data, aes(x = nurse_ratio, y = patient_satisfaction,
                              color = ownership)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("For-Profit" = "coral", "Non-Profit" = "steelblue")) +
  labs(
    x = "Nurse-to-Patient Ratio",
    y = "Patient Satisfaction Score",
    color = "Ownership Type",
    title = "Does Staffing Effect Differ by Hospital Ownership?",
    subtitle = "Different slopes suggest an interaction effect"
  ) +
  theme_minimal(base_size = 12)
```

The slopes appear different. Staffing seems to have a stronger positive effect in Non-Profit hospitals than in For-Profit hospitals.

### Fitting the Interaction Model

When we fit a model with an interaction, we can either use the `*` symbol between variables or the `:` symbol between variables with each variable included with the `+` sign.

```{r}
#| label: fit-interaction

# Model WITH interaction
# Alternative syntax for interaction: nurse_ratio + ownership + nurse_ratio:ownership
model_interact <- lm(patient_satisfaction ~ nurse_ratio * ownership, data = satisfaction_data)
summary(model_interact)

```


### Interpreting the Interaction

```{r}
#| label: interpret-interaction
#| echo: false

# Formatted output
tidy(model_interact, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = "Regression with Interaction Effect") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Understanding the Coefficients**:

```{r}
#| echo: false
int_coefs <- coef(model_interact)
```

1. **Intercept (`r round(int_coefs["(Intercept)"], 2)`)**: The predicted satisfaction score for a For-Profit hospital with a nurse ratio of 0.

2. **nurse_ratio (`r round(int_coefs["nurse_ratio"], 2)`)**: The effect of staffing on satisfaction **for the reference group (For-Profit hospitals)**. Each additional nurse per patient increases satisfaction by `r round(int_coefs["nurse_ratio"], 2)` points in For-Profit hospitals.

3. **ownershipNon-Profit (`r round(int_coefs["ownershipNon-Profit"], 2)`)**: The difference in satisfaction between Non-Profit and For-Profit hospitals **when nurse_ratio = 0** (not practically meaningful, since hospitals can't have zero nurses). This represents the baseline difference between ownership types at zero staffing.

4. **nurse_ratio:ownershipNon-Profit (`r round(int_coefs["nurse_ratio:ownershipNon-Profit"], 2)`)**: **THE INTERACTION TERM**. This is the additional effect of staffing in Non-Profit hospitals beyond the effect in For-Profit hospitals. The positive value means the slope is steeper for Non-Profit hospitals, and the p-value shows that this difference is statistically significant.

**Computing the Effect of Staffing for Each Group**:

- **For-Profit hospitals**: β = `r round(int_coefs["nurse_ratio"], 2)`
  - Each additional nurse increases satisfaction by `r round(int_coefs["nurse_ratio"], 2)` points

- **Non-Profit hospitals**: β = `r round(int_coefs["nurse_ratio"], 2)` + `r round(int_coefs["nurse_ratio:ownershipNon-Profit"], 2)` = `r round(int_coefs["nurse_ratio"] + int_coefs["nurse_ratio:ownershipNon-Profit"], 2)`
  - Each additional nurse increases satisfaction by `r round(int_coefs["nurse_ratio"] + int_coefs["nurse_ratio:ownershipNon-Profit"], 2)` points

The interaction reveals that the **context matters**. The same staffing increase produces different outcomes depending on hospital ownership structure. This is precisely what interaction effects help us understand when and where interventions work better.

### Visualizing the Interaction Effect

```{r}
#| label: fig-interaction-plot
#| fig-cap: "Interaction effect: Staffing effect varies by ownership type"
#| fig-height: 4
#| fig-width: 6
#| code-fold: true

# Create prediction grid
pred_grid <- expand_grid(
  nurse_ratio = seq(2, 6, by = 0.1),
  ownership = c("For-Profit", "Non-Profit")
) %>%
  mutate(
    predicted = predict(model_interact, newdata = .)
  )

# Plot
ggplot(pred_grid, aes(x = nurse_ratio, y = predicted, color = ownership)) +
  geom_line(size = 1.5) +
  scale_color_manual(values = c("For-Profit" = "coral", "Non-Profit" = "steelblue")) +
  labs(
    x = "Nurse-to-Patient Ratio",
    y = "Predicted Patient Satisfaction",
    color = "Ownership Type",
    title = "Interaction Effect: Staffing × Ownership",
    subtitle = paste0("For-Profit slope: ", round(int_coefs["nurse_ratio"], 2),
                     " | Non-Profit slope: ",
                     round(int_coefs["nurse_ratio"] + int_coefs["nurse_ratio:ownershipNon-Profit"], 2))
  ) +
  theme_minimal(base_size = 12)
```

### Practice Checkpoint 3 {.unnumbered}

::: {.callout-important}
## Try It Yourself! (10 minutes)

**Pause here!** Open `week12_linear-regression2_practice.qmd` and complete:

- **Task 1.5**: Fit a continuous × categorical interaction model
- **Task 1.6**: Calculate marginal effects for each group
- **Task 1.7**: Visualize the interaction with `probe_interaction()`

This will help you understand how to test and interpret interactions where a continuous effect varies by category.

**After 10 minutes**, we'll continue with categorical × categorical interactions.
:::

*→ We'll explore more complex interactions in the remaining sections*

## Categorical × Categorical Interaction

Interactions can also occur between two categorical variables.

### Categorical × Categorical Example: Treatment Effect Varies by Insurance Type

#### Background and Context

**The context**: Diabetes management programs help patients control blood sugar levels (measured by HbA1c). Different programs exist:

- **Standard Care**: Regular doctor visits and medication management
- **Intensive Program**: Structured education, nutrition counseling, and regular coaching

**Question**: Does program effectiveness vary by insurance type (Medicare, Medicaid, Private)?

**Why it might vary**:

- Access to additional support services differs by insurance
- Co-pays and out-of-pocket costs vary
- Transportation and time off work (easier for retirees with Medicare than workers with private insurance)
- A 2023 study on Medicare Diabetes Prevention Program found significant implementation challenges, with very limited program availability nationwide, suggesting access and effectiveness barriers vary across insurance types.

Let's test whether an intensive diabetes program works differently for patients with different insurance types. This is a categorical (insurance type) × categorical (program type) interaction.

#### Creating Synthetic Healthcare Data

Let's create a dataset using the code below:

```{r}
#| label: catcat-interaction

# Generate data
n <- 400

diabetes_data <- tibble(
  patient_id = 1:n,
  treatment = sample(c("Standard Care", "Intensive Program"), n, replace = TRUE),
  insurance = sample(c("Medicare", "Medicaid", "Private"), n, replace = TRUE,
                    prob = c(0.4, 0.3, 0.3)),
  baseline_a1c = round(rnorm(n, 8.5, 1.2), 1),
  age = round(rnorm(n, 60, 12))
) %>%
  mutate(
    baseline_a1c = pmax(6, pmin(baseline_a1c, 12)),
    age = pmax(30, pmin(age, 85)),
    # Generate A1c reduction with interaction
    a1c_reduction = round(
      0.5 +
      0.3 * (treatment == "Intensive Program") +   # Base treatment effect
      0.2 * (insurance == "Private") +              # Private insurance effect
      0.1 * (insurance == "Medicare") +
      # Interactions: treatment works better with Private insurance
      0.8 * (treatment == "Intensive Program") * (insurance == "Private") +
      0.3 * (treatment == "Intensive Program") * (insurance == "Medicare") +
      0.05 * baseline_a1c +
      rnorm(n, 0, 0.4),
      2
    ),
    a1c_reduction = pmax(0, a1c_reduction)
  )
```

**Variables in the dataset**:

- `patient_id`: Unique identifier for each patient (1 to 400)
- `treatment`: Type of diabetes program (Standard Care or Intensive Program)
- `insurance`: Insurance type (Medicare, Medicaid, or Private)
- `baseline_a1c`: Baseline HbA1c level before treatment
- `age`: Patient age in years
- `a1c_reduction`: Reduction in HbA1c after treatment (our outcome variable)

### Fitting the Interaction Model

First, let's make the variables as factors using the `factor()` function.

```{r}
diabetes_data <- diabetes_data |>
  mutate(
    treatment = factor(treatment, levels = c("Standard Care", "Intensive Program")),
    insurance = factor(insurance, levels = c("Medicaid", "Medicare", "Private"))
  )
```

When we put those factor variables into the model, R automatically creates the necessary dummy variables.

```{r}
#| label: catcat-model

# Fit model with interaction
# Alternative syntax: treatment + insurance + treatment:insurance
model_diabetes <- lm(a1c_reduction ~ treatment * insurance,
                     data = diabetes_data)
summary(model_diabetes)
```

### Understanding Categorical × Categorical Interactions

With two categorical variables, the interaction becomes more complex. Let's break down each coefficient:

```{r}
#| label: interpret-catcat
#| echo: false

# Formatted output
tidy(model_diabetes, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = "Regression Results: Treatment × Insurance Interaction") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Understanding Each Coefficient**:

```{r}
#| echo: false
diab_coefs <- coef(model_diabetes)
```

**Reference Group**: Standard Care + Medicaid (both reference categories)

1. **Intercept (`r round(diab_coefs["(Intercept)"], 3)`)**: The predicted HbA1c reduction for patients receiving **Standard Care with Medicaid insurance**. This is our baseline group.

2. **treatmentIntensive Program (`r round(diab_coefs["treatmentIntensive Program"], 3)`)**: The additional effect of the Intensive Program compared to Standard Care **for Medicaid patients only** (the reference insurance group).
   - Medicaid patients in Intensive Program: `r round(diab_coefs["(Intercept)"], 3)` + `r round(diab_coefs["treatmentIntensive Program"], 3)` = `r round(diab_coefs["(Intercept)"] + diab_coefs["treatmentIntensive Program"], 3)` reduction

3. **insuranceMedicare (`r round(diab_coefs["insuranceMedicare"], 3)`)**: The difference between Medicare and Medicaid patients **receiving Standard Care only** (the reference treatment).
   - Medicare patients in Standard Care: `r round(diab_coefs["(Intercept)"], 3)` + `r round(diab_coefs["insuranceMedicare"], 3)` = `r round(diab_coefs["(Intercept)"] + diab_coefs["insuranceMedicare"], 3)` reduction

4. **insurancePrivate (`r round(diab_coefs["insurancePrivate"], 3)`)**: The difference between Private and Medicaid patients **receiving Standard Care only**.
   - Private patients in Standard Care: `r round(diab_coefs["(Intercept)"], 3)` + `r round(diab_coefs["insurancePrivate"], 3)` = `r round(diab_coefs["(Intercept)"] + diab_coefs["insurancePrivate"], 3)` reduction

5. **treatmentIntensive Program:insuranceMedicare (`r round(diab_coefs["treatmentIntensive Program:insuranceMedicare"], 3)`)**: **INTERACTION TERM**. This is the additional treatment effect for Medicare patients beyond what Medicaid patients experience.
   - This tells us whether the Intensive Program works differently for Medicare vs. Medicaid patients
   - Medicare patients in Intensive Program: `r round(diab_coefs["(Intercept)"], 3)` + `r round(diab_coefs["treatmentIntensive Program"], 3)` + `r round(diab_coefs["insuranceMedicare"], 3)` + `r round(diab_coefs["treatmentIntensive Program:insuranceMedicare"], 3)` = `r round(diab_coefs["(Intercept)"] + diab_coefs["treatmentIntensive Program"] + diab_coefs["insuranceMedicare"] + diab_coefs["treatmentIntensive Program:insuranceMedicare"], 3)` reduction
   - A significant interaction means that the treatment effect for Medicare patients is significantly different from the treatment effect for Medicaid patients.

6. **treatmentIntensive Program:insurancePrivate (`r round(diab_coefs["treatmentIntensive Program:insurancePrivate"], 3)`)**: **INTERACTION TERM**. This is the additional treatment effect for Private insurance patients beyond what Medicaid patients experience.
   - This tells us whether the Intensive Program works differently for Private vs. Medicaid patients
   - Private patients in Intensive Program: `r round(diab_coefs["(Intercept)"], 3)` + `r round(diab_coefs["treatmentIntensive Program"], 3)` + `r round(diab_coefs["insurancePrivate"], 3)` + `r round(diab_coefs["treatmentIntensive Program:insurancePrivate"], 3)` = `r round(diab_coefs["(Intercept)"] + diab_coefs["treatmentIntensive Program"] + diab_coefs["insurancePrivate"] + diab_coefs["treatmentIntensive Program:insurancePrivate"], 3)` reduction
   - A significant interaction means that the treatment effect for Private insurance patients is significantly different from the treatment effect for Medicaid patients.

**Computing Treatment Effects for Each Insurance Group**:

- **Medicaid**: Treatment effect = `r round(diab_coefs["treatmentIntensive Program"], 3)`
- **Medicare**: Treatment effect = `r round(diab_coefs["treatmentIntensive Program"], 3)` + `r round(diab_coefs["treatmentIntensive Program:insuranceMedicare"], 3)` = `r round(diab_coefs["treatmentIntensive Program"] + diab_coefs["treatmentIntensive Program:insuranceMedicare"], 3)`
- **Private**: Treatment effect = `r round(diab_coefs["treatmentIntensive Program"], 3)` + `r round(diab_coefs["treatmentIntensive Program:insurancePrivate"], 3)` = `r round(diab_coefs["treatmentIntensive Program"] + diab_coefs["treatmentIntensive Program:insurancePrivate"], 3)`

The Intensive Program is effective across all insurance types, but significantly more effective for Private insurance patients compared to Medicaid patients. This suggests that program effectiveness depends on insurance type.


### Visualizing Categorical × Categorical Interaction

```{r}
#| label: fig-catcat-bar
#| fig-cap: "Treatment effectiveness by insurance type"
#| fig-height: 4
#| fig-width: 6
#| code-fold: true

# Get estimated marginal means
emm_df <- as.data.frame(emmeans(model_diabetes, ~ treatment * insurance))

ggplot(emm_df, aes(x = insurance, y = emmean, fill = treatment)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                position = position_dodge(width = 0.8), width = 0.2) +
  scale_fill_manual(values = c("Standard Care" = "gray60",
                               "Intensive Program" = "steelblue")) +
  labs(
    x = "Insurance Type",
    y = "Predicted HbA1c Reduction",
    fill = "Treatment",
    title = "Interaction: Treatment × Insurance",
    subtitle = "Error bars show 95% confidence intervals"
  ) +
  theme_minimal(base_size = 10)
```

## Continuous × Continuous Interaction

Interactions can also occur between two continuous variables, where the effect of one continuous variable depends on the level of another continuous variable.

### Continuous × Continuous Interaction Example: Age Moderates Physical Therapy Effect

#### Background and Context

**The context**: Physical therapy (PT) is commonly prescribed for musculoskeletal conditions like joint pain, post-surgery recovery, and mobility issues. Patients typically attend multiple sessions over several weeks to improve functional capacity.

**Question**: Does the benefit of additional PT sessions depend on patient age?

**Why it might vary**:

- **Tissue healing capacity**: Younger patients may have faster tissue repair and adaptation to exercise
- **Baseline strength and conditioning**: Younger patients often have better baseline capacity to benefit from therapy
- **Comorbidities**: Older patients may have multiple health conditions that limit their response to therapy
- **Adherence to home exercises**: Younger working adults may have different adherence patterns than older retirees

**The hypothesis**: Additional PT sessions might provide greater functional improvement for younger patients compared to older patients due to differences in physiological recovery capacity and treatment response.

- Physical therapy effectiveness can vary significantly across patient populations, with factors such as age, baseline function, and comorbidities potentially moderating treatment response.

Let's test whether the effect of PT session count on functional improvement varies by patient age. This is a continuous (PT sessions) × continuous (age) interaction.

#### Creating Synthetic Healthcare Data

Let's create a dataset using the code below:

```{r}
#| label: contcont-interaction

# Generate data
n <- 250

pt_data <- tibble(
  patient_id = 1:n,
  pt_sessions = round(runif(n, 4, 24)),
  age = round(runif(n, 25, 80)),
  baseline_function = round(runif(n, 30, 70))
) %>%
  mutate(
    # Interaction: PT sessions more effective for younger patients
    function_improvement = round(
      5 +
      1.2 * pt_sessions +
      -0.1 * age +
      -0.03 * pt_sessions * age +  # Negative interaction
      0.3 * baseline_function +
      rnorm(n, 0, 5),
      1
    ),
    function_improvement = pmax(0, function_improvement)
  )
```

**Variables in the dataset**:

- `patient_id`: Unique identifier for each patient (1 to 250)
- `pt_sessions`: Number of physical therapy sessions attended (4 to 24 sessions)
- `age`: Patient age in years
- `baseline_function`: Baseline functional capacity score before therapy (higher scores indicate better pre-treatment function)
- `function_improvement`: Improvement in functional capacity after therapy in points (our outcome variable)

### Fitting the Interaction Model

Let's fit the model using the `lm()` function.

```{r}
#| label: fit-contcont

# Fit model with interaction
# Alternative expression: pt_sessions + age + pt_sessions:age
model_pt <- lm(function_improvement ~ pt_sessions * age,
               data = pt_data)
summary(model_pt)
```

### Interpreting Continuous × Continuous Interactions

```{r}
#| label: interpret-contcont
#| echo: false

# Formatted output
tidy(model_pt, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "Regression Results: PT Sessions × Age Interaction") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Understanding the Coefficients**:

```{r}
#| echo: false
pt_coefs <- coef(model_pt)
```

1. **Intercept (`r round(pt_coefs["(Intercept)"], 2)`)**: The predicted improvement for a patient with 0 PT sessions and age 0 (not practically meaningful).

2. **pt_sessions (`r round(pt_coefs["pt_sessions"], 2)`)**: The effect of one additional PT session **when age = 0** (not practically meaningful, since we don't have newborns in PT).

3. **age (`r round(pt_coefs["age"], 2)`)**: The effect of being one year older **when pt_sessions = 0** (not practically meaningful, since patients with zero sessions wouldn't be in our study).

4. **pt_sessions:age (`r round(pt_coefs["pt_sessions:age"], 4)`)**: **THE INTERACTION TERM**. This tells us how the effect of PT sessions changes as patients get older.
   - The **negative value** means that PT sessions become less effective as age increases
   - For each additional year of age, the benefit of one PT session decreases by `r abs(round(pt_coefs["pt_sessions:age"], 4))` points

**Computing Marginal Effects at Different Ages**:

In interaction models, we're interested in the **marginal effect**, the effect of one variable (PT sessions) at specific values of another variable (age). Because of the interaction, this marginal effect changes with age.

The marginal effect of PT sessions at any given age is calculated as:

$$\frac{\partial \text{Improvement}}{\partial \text{PT Sessions}} = \beta_{\text{pt\_sessions}} + \beta_{\text{interaction}} \times \text{Age}$$

Let's calculate the marginal effect of one additional PT session at different ages:

- **Age 30**: Effect = `r round(pt_coefs["pt_sessions"], 2)` + (30 × `r round(pt_coefs["pt_sessions:age"], 4)`) = `r round(pt_coefs["pt_sessions"] + 30 * pt_coefs["pt_sessions:age"], 2)` points per session (positive = beneficial)
- **Age 50**: Effect = `r round(pt_coefs["pt_sessions"], 2)` + (50 × `r round(pt_coefs["pt_sessions:age"], 4)`) = `r round(pt_coefs["pt_sessions"] + 50 * pt_coefs["pt_sessions:age"], 2)` points per session (negative = harmful)
- **Age 70**: Effect = `r round(pt_coefs["pt_sessions"], 2)` + (70 × `r round(pt_coefs["pt_sessions:age"], 4)`) = `r round(pt_coefs["pt_sessions"] + 70 * pt_coefs["pt_sessions:age"], 2)` points per session (negative = harmful)

The effect of PT sessions **changes dramatically with age**. For younger patients (age 30), additional PT sessions improve function. However, the effect becomes **negative** for older patients (ages 50 and 70), meaning additional sessions are associated with worse functional outcomes. This crossover pattern demonstrates how interactions can reverse the direction of an effect.

### Visualizing Continuous × Continuous Interaction

For continuous × continuous interactions, we typically show the effect at different levels of the moderator. We'll use the `probe_interaction()` function from the `interactions` package, which makes it easy to visualize interactions and get marginal effects at different levels of a moderator:

```{r}
#| label: fig-contcont
#| fig-cap: "PT sessions effect varies by patient age"
#| fig-height: 4
#| fig-width: 6

# Use probe_interaction to visualize the interaction
# Specify custom age values: 30, 50, 70
probe_interaction(model_pt,
                  pred = pt_sessions,
                  modx = age,
                  modx.values = c(30, 50, 70),
                  interval = TRUE,
                  int.width = 0.95,
                  x.label = "Number of PT Sessions",
                  y.label = "Predicted Function Improvement",
                  main.title = "Interaction: PT Sessions × Age",
                  legend.main = "Patient Age")
```


**Interpretation of marginal effects**:

- **SIMPLE SLOPES ANALYSIS** (rate of change) shows the effect of PT sessions at each age level
  - The **slope** (Est.) at that age
  - **Standard error** (S.E.) and **t-statistic** (t val.)
  - **p-value** testing if the slope is significantly different from zero
- The changing slopes confirm our earlier manual calculations
- All three slopes are statistically significant (p < 0.05), but they differ in **direction**:
  - Age 30: Positive slope (`r round(pt_coefs["pt_sessions"] + 30 * pt_coefs["pt_sessions:age"], 2)`) = PT sessions are **beneficial**
  - Age 50: Negative slope (`r round(pt_coefs["pt_sessions"] + 50 * pt_coefs["pt_sessions:age"], 2)`) = PT sessions are **harmful**
  - Age 70: More negative slope (`r round(pt_coefs["pt_sessions"] + 70 * pt_coefs["pt_sessions:age"], 2)`) = PT sessions are **more harmful**

**Visual interpretation**: The graph clearly shows the interaction effect: The slopes change from positive to negative as age increases:

- **Positive slope for younger patients**: Age 30 patients (upward sloping line) gain from each additional session
- **Negative slope for middle-aged patients**: Age 50 patients (downward sloping line) are harmed by additional sessions
- **Steeper negative slope for older patients**: Age 70 patients (steeper downward slope) are harmed even more by additional sessions

In this data, the negative slopes at older ages suggest that more PT sessions harm older patients. In reality, this pattern would be unusual and would warrant investigation into potential confounding factors (e.g., sicker patients receiving more sessions). This example illustrates how interactions can produce crossover effects where treatment benefits reverse in different subgroups.

### Practice Checkpoint 4 {.unnumbered}

::: {.callout-important}
## Try It Yourself! (10 minutes)

**Pause here!** Open `week12_linear-regression2_practice.qmd` and complete:

- **Task 1.8**: Center continuous variables and fit a continuous × continuous interaction model

This will help you understand how centering improves interpretability in continuous × continuous interactions.

**After 10 minutes**, proceed to the advanced exercises in the practice file.
:::

*→ Apply all concepts in Exercises 1-3 of the practice file*

------------------------------------------------------------------------

# Practical Considerations {.section-time data-time="10 min"}

## When to Include Interactions

**Include interactions when**:

1. Theory or prior research suggests the effect should vary
2. Visual inspection shows different slopes for different groups
3. You have sufficient sample size (interactions require larger samples)

## Centering Variables

For models with continuous × continuous interactions, **centering** variables (subtracting the mean) makes the main effects much more interpretable.

**Why centering helps?**: In the uncentered PT model, the main effects were at age = 0 and pt_sessions = 0, which are meaningless. After centering, the main effects are **at the mean age and mean PT sessions**, which is much more interpretable.

Let's re-fit the PT model with centered variables and compare:

```{r}
#| label: centering

# Center variables by subtracting their means
pt_data <- pt_data %>%
  mutate(
    age_c = age - mean(age), # mean-centering the variable age
    pt_sessions_c = pt_sessions - mean(pt_sessions) # mean-centering the variable pt_sessions
  )

# Check the means of centered variables (should be approximately 0)
cat("Mean of age_c:", round(mean(pt_data$age_c), 3), "\n")
cat("Mean of pt_sessions_c:", round(mean(pt_data$pt_sessions_c), 3), "\n")

# Fit model with centered variables
model_pt_centered <- lm(function_improvement ~ pt_sessions_c * age_c,
                        data = pt_data)

# Compare uncentered vs centered models
comparison <- bind_rows(
  tidy(model_pt) %>% mutate(Model = "Uncentered"),
  tidy(model_pt_centered) %>% mutate(Model = "Centered")
) %>%
  select(Model, term, estimate, std.error, p.value) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

comparison %>%
  kable(caption = "Comparison: Uncentered vs Centered Models") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


**Key observations**:

```{r}
#| echo: false
pt_centered_coefs <- coef(model_pt_centered)
```

1. **Interaction term stays the same**: `r round(pt_coefs["pt_sessions:age"], 4)` (uncentered) = `r round(pt_centered_coefs["pt_sessions_c:age_c"], 4)` (centered)
   - Centering does NOT change the interaction effect

2. **Main effects are now interpretable**:
   - **Uncentered `pt_sessions`** (`r round(pt_coefs["pt_sessions"], 2)`): Effect when age = 0 (meaningless)
   - **Centered `pt_sessions_c`** (`r round(pt_centered_coefs["pt_sessions_c"], 2)`): Effect at **mean age** (`r round(mean(pt_data$age), 1)` years) (meaningful!)
   - **Uncentered `age`** (`r round(pt_coefs["age"], 2)`): Effect when pt_sessions = 0 (meaningless)
   - **Centered `age_c`** (`r round(pt_centered_coefs["age_c"], 2)`): Effect at **mean PT sessions** (`r round(mean(pt_data$pt_sessions), 1)` sessions) (meaningful!)

3. **Intercept is now interpretable**:
   - **Uncentered**: Predicted improvement when age = 0 AND pt_sessions = 0 (meaningless)
   - **Centered**: Predicted improvement at **mean age and mean PT sessions** = `r round(pt_centered_coefs["(Intercept)"], 2)` points

Consider centering continuous variables when you have continuous × continuous interactions. It makes the model much easier to interpret without changing the interaction effect or model fit.

------------------------------------------------------------------------

# Summary {.section-time data-time="5 min"}

## What We Covered in This Lecture

This lecture extended linear regression to handle categorical predictors and situations where effects vary across groups.

### 1. The Challenge: Including Categories in Regression

**The problem**: Regression needs numbers, but many important variables are categorical (hospital type, insurance status, treatment group).

**The solution**: Dummy variables (also called indicator variables)

- Use 0s and 1s to represent category membership
- For k categories, create k-1 dummy variables
- One category serves as the reference (baseline) for comparison

**Key concept**: Dummy variable coefficients show the difference between each category and the reference category, holding other variables constant.

### 2. How Dummy Variables Work

**Binary categories** (two groups):

- Create ONE dummy variable (0 or 1)
- Coefficient = difference in outcome between the two groups
- Example: ERAS protocol vs. Standard care

**Multiple categories** (three+ groups):

- Create k-1 dummy variables (one less than number of categories)
- Each coefficient compares that category to the reference
- Example: Teaching vs. Community vs. Critical Access hospitals
- Change reference category with `relevel()` to make different comparisons

**Why k-1 instead of k?** Including all categories creates perfect multicollinearity with the intercept.

### 3. Interaction Effects: When Effects Depend on Context

**What are they?** An interaction occurs when the effect of one variable changes depending on another variable. In plain language, "it depends!" The answer to "What's the effect of X on Y?" is "It depends on Z."

**Types of interactions**:

- **Continuous × Categorical**: Does nurse staffing effect differ by hospital ownership?
- **Categorical × Categorical**: Does treatment effectiveness vary by insurance type?
- **Continuous × Continuous**: Does PT session benefit depend on patient age?

### 4. Recognizing Interactions

**Visual clue**: Different slopes for different groups in your scatter plots

**When to test for interactions**:

- Theory suggests effect should vary across groups
- Plots show different slopes
- You want to know if an intervention works differently for different populations

### 5. Interpreting Interactions

**The key challenge**: Main effects change meaning when interactions are present!

**For Continuous × Categorical**:

- Main effect of continuous variable = effect for the reference category only
- Interaction coefficient = how much that effect changes in other categories
- To get effect for non-reference categories, add main effect + interaction

**Example**:

- Nurse staffing effect in For-Profit hospitals: β = 3.0
- Interaction (Non-Profit): β = 2.5
- Nurse staffing effect in Non-Profit hospitals: 3.0 + 2.5 = 5.5

**For Categorical × Categorical**:

- Use `emmeans()` to calculate and compare means for all combinations
- Visualize with grouped bar charts
- Each interaction term shows how treatment effects differ across categories

**For Continuous × Continuous**:

- **Main effects without centering**: Represent effects when the other variable = 0 (often meaningless)
- **Main effects with centering**: Represent effects at the mean of the other variable (meaningful!)
- **Interaction coefficient**: Shows how the slope of one variable changes as the other variable increases
- **Marginal effects (slopes)**: Calculate the effect of one variable at specific values of the moderator
  - Formula: Slope of X = β_X + (β_interaction × value of Z)
  - Use `probe_interaction()` or manual calculation to get slopes at different moderator values (and to visualize the interaction)
- **Crossover effects**: The effect can even reverse direction (positive → negative) at different moderator values

**Example**:

- PT sessions effect at age 30: β_sessions + (β_interaction × 30) = 0.30 (beneficial)
- PT sessions effect at age 70: β_sessions + (β_interaction × 70) = -0.60 (harmful)
- The interaction creates a crossover where the effect reverses from beneficial to harmful

### 6. Practical Considerations

**Sample size**: Interactions require larger samples because you're splitting data into subgroups

**Centering variables**: For models with interactions, center continuous variables (subtract the mean) to make coefficients more interpretable

## Next Steps: Practice Activities

::: {.callout-note icon=false}
## Practice Activities

Complete the hands-on practice in **`week12_linear-regression2_practice.qmd`**:

### Part 1: Guided Practice (8 tasks)

- Create and interpret binary dummy variables
- Work with multi-level categorical variables
- Use `emmeans()` for pairwise comparisons
- Test continuous × categorical interactions
- Calculate marginal effects
- Visualize interactions
- Center variables for better interpretability

### Part 2: Advanced Exercises (3 exercises)

- Apply dummy variables to real-world scenarios
- Test for interactions independently
- Make predictions using interaction models

**Expected time**: 1-2 hours during/after lecture

**Purpose**: Build confidence and fluency with dummy variables and interactions
:::

------------------------------------------------------------------------

# References and Further Reading

## Statistical Methods

- Gelman, A., & Hill, J. (2006). *Data Analysis Using Regression and Multilevel/Hierarchical Models*. Cambridge University Press.
- James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning with Applications in R* (2nd ed.). Springer.
- Fox, J. (2015). *Applied Regression Analysis and Generalized Linear Models* (3rd ed.). SAGE Publications.

## Healthcare Applications

Examples are motivated by the following:

- Katikam, S. M., Komari, L. K., & Parimi, S. B. R. (2025). Evaluation of Enhanced Recovery After Surgery (ERAS) Protocols in Abdominal Surgery. *Cureus*, 17(1), e76908. https://pmc.ncbi.nlm.nih.gov/articles/PMC12244614/

- Horwitz, L. I., Bernheim, S. M., Ross, J. S., Herrin, J., Grady, J., Krumholz, H. M., & Drye, E. E. (2017). Hospital characteristics associated with risk-standardized readmission rates. *Medical Care*, 55(5), 528-534. https://pmc.ncbi.nlm.nih.gov/articles/PMC5426655/

- Turk, M. T., Jakub, K., & Ritchie, N. D. (2023). Stakeholder Analysis: Medicare Diabetes Prevention Program Awareness and Implementation. *Preventive Medicine Reports*, 33, 102199. https://pmc.ncbi.nlm.nih.gov/articles/PMC10287029/

------------------------------------------------------------------------

# Appendix: R Code Reference

```{r}
#| eval: false

# Create dummy variables automatically
model <- lm(y ~ factor(category), data = mydata)

# Change reference category
mydata$category <- relevel(factor(mydata$category), ref = "baseline")

# Interaction syntax options
model <- lm(y ~ x1 * x2, data = mydata)        # Main effects + interaction
model <- lm(y ~ x1 + x2 + x1:x2, data = mydata) # Equivalent

# Multiple interactions
model <- lm(y ~ x1 * x2 * x3, data = mydata)   # All main effects and interactions

# Marginal means for categorical interactions
library(emmeans)
emmeans(model, ~ treatment | group)

# Simple slopes for continuous interactions
library(interactions)
probe_interaction(model, pred = x, modx = z)

# Center variables
mydata <- mydata  |>
  mutate(x_centered = x - mean(x))

```
