---
title: "Linear Regression 2: Dummy Variables and Interaction Effects"
subtitle: "MBA642: Business Analytics in Healthcare Management"
Week: "Week 12 and Week 13"
date: 03/30/2026
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: false
    theme: [lecture.scss, cosmo]
    highlight-style: github
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Load required packages
library(tidyverse)
library(car)
library(emmeans)
library(interactions)
library(patchwork)

# Set seed for reproducibility
set.seed(2026)
```

# Introduction

In Week 11, we used continuous predictors (e.g., wait time, staffing hours, patient severity) to predict outcomes. But many important variables in healthcare are **categorical** (e.g., hospital type, insurance status, treatment protocol, or region). Additionally, the effect of one variable may **depend on the value of another** (interaction effects). For instance, a treatment may work differently for different patient populations.

**Dummy variables and interactions extend linear regression** by answering more nuanced questions:

-   *Do patients in the ERAS (Enhanced Recovery After Surgery) protocol have shorter hospital stays than those in standard care, after controlling for age and severity?* (comparing group outcomes)
-   *Is the effect of nurse staffing on patient satisfaction different in for-profit vs. non-profit hospitals?* (testing conditional effects)
-   *Does treatment effectiveness vary by insurance type (if so, by how much)?* (quantifying group-specific effects)

## Why This Matters

Healthcare managers frequently encounter situations where understanding **group differences** and **conditional effects** is crucial for decision-making. For example:

-   **Resource allocation**: Do readmission rates differ across hospital types (teaching vs. community vs. critical access)?
-   **Cost management**: Does an enhanced recovery protocol reduce length of stay compared to standard care?
-   **Quality improvement**: Does the effect of staff training on patient satisfaction depend on hospital ownership type?

## What We Cover Today

This lecture introduces two extensions to the linear regression framework from Week 11:

1.  **Dummy Variable Regression**: Including categorical predictors in regression models
    -   Binary categories (e.g., ERAS vs. Standard care)
    -   Multi-level categories (e.g., Teaching vs. Community vs. Critical Access hospitals)
2.  **Interaction Effects**: Testing whether the effect of one variable depends on another
    -   Continuous × Categorical (e.g., does staffing effect vary by hospital ownership?)
    -   Categorical × Categorical (e.g., does treatment effectiveness vary by insurance type?)
    -   Continuous × Continuous (e.g., does PT session benefit depend on patient age?)

------------------------------------------------------------------------

# 1. Dummy Variables (Indicator Variables)

## What is it?

Dummy variables (also called indicator variables) allow us to include categorical predictors in regression models, which mathematically require numerical inputs. Each dummy variable uses a simple 0 vs. 1 coding to indicate whether an observation belongs to a particular category.

For a categorical variable with **k categories**, we create **k-1 dummy variables**, with one category serving as **the reference group** against which all others are compared.

For example, if we have a variable "Hospital Type" with three categories (Teaching, Community, Critical Access), we would create two dummy variables:

-   Dummy 1: Teaching (1 if Teaching, 0 otherwise)
-   Dummy 2: Critical Access (1 if Critical Access, 0 otherwise)

*Community hospitals* would be the reference category (both dummies = 0).

***Dummy 1*** captures the difference between Teaching and Community hospitals (the reference category), while ***Dummy 2*** captures the difference between Critical Access and Community hospitals.

## When to Use It

Use dummy variables when you need to:

-   Include categorical predictors in regression (hospital type, insurance status, treatment group)
-   Compare outcomes across different categories while controlling for other variables
-   Test whether group differences are statistically significant
-   Quantify the magnitude of differences between categories

## The Challenge with Categorical Variables

Regression requires numerical inputs, but categorical variables are not inherently numerical. Simply assigning numbers (1, 2, 3) to categories implies an ordering and equal spacing that usually doesn't exist.

**Solution**: Create **dummy variables** that represent category membership using 0s and 1s.

## 1.1 Binary Categorical Variables

For a variable with **two categories**, we need **one dummy variable**. For example, if we have a categorical variable called Insurance with two categories (Insured and Uninsured), we would create one dummy variable called Insured:

| Patient | Insurance Status | Dummy: Insured |
|---------|------------------|----------------|
| 1       | Insured          | 1              |
| 2       | Uninsured        | 0              |
| 3       | Insured          | 1              |
| 4       | Uninsured        | 0              |

The coefficient on the dummy variable (Insured) represents the difference in the outcome between the two groups (Insured vs. Uninsured).

### Dummy Variable Example Case: Treatment Protocol Comparison

#### Background and Context

Traditional surgical recovery involves extended hospital stays, which are costly and can lead to complications.

Enhanced Recovery After Surgery (ERAS) protocols use evidence-based practices to help patients recover faster. These include:

-   Better pain management (using less opioid medication)
-   Getting patients moving sooner after surgery
-   Starting normal eating earlier

A 2025 study found that ERAS protocols significantly reduce hospital length of stay compared to traditional care, with patients recovering and being discharged substantially earlier.

Let's compare patient outcomes (length of stay) between standard care and ERAS protocols to quantify the treatment effect using a dummy variable.

#### Data Loading

Let's load the dataset using the code below:

```{r}
#| label: binary-dummy-data

# Required packages
library(tidyverse)
library(car)
library(emmeans)
library(interactions)
set.seed(2026)

# Loading the dataset
surgical_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week12_surgical_data.csv")

# Display the dataset
head(surgical_data, 5) 
```

The dataset includes 200 surgical patients with the following variables:

-   `patient_id`: Unique identifier for each patient (1 to 200)
-   `protocol`: Treatment protocol assigned (Standard or ERAS)
-   `age`: Patient age in years
-   `bmi`: Body Mass Index (measure of body fat based on height and weight)
-   `surgery_complexity`: Complexity level of the surgery (1 = Low, 2 = Medium, 3 = High)
-   `length_of_stay`: Hospital length of stay in days (our outcome variable)

The variable `protocol` is our binary categorical variable (Standard vs. ERAS). We'll create a dummy variable for ERAS (1 = ERAS, 0 = Standard) to include in our regression model.

We can do this using the `factor()` function in R, which automatically creates the necessary dummy variable.

```{r}
#| label: make-binary-dummy

# R automatically creates dummy variables for factors
surgical_data <- surgical_data %>%
  mutate(protocol = factor(protocol, levels = c("Standard", "ERAS")))
```

Before fitting the model, let's visualize the distribution of length of stay by treatment protocol:

```{r}
#| label: fig-protocol-effect
#| code-fold: true
#| fig-cap: "Distribution of length of stay by treatment protocol"
#| fig-height: 4
#| fig-width: 6

ggplot(surgical_data, aes(x = protocol, y = length_of_stay, fill = protocol)) +
  geom_boxplot(alpha = 0.7, width = 0.4) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  scale_fill_manual(values = c("Standard" = "coral", "ERAS" = "steelblue")) +
  labs(
    x = "Treatment Protocol",
    y = "Length of Stay (days)",
    title = "Does ERAS Protocol Affect Length of Stay?"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

The boxplot suggests that ERAS patients tend to have shorter hospital stays. Let's test this formally with regression.

### Fitting the Model

We can fit a linear regression model using the `lm()` function in R. When we include the variable `protocol` in the model, which is a factor, R automatically creates the dummy variable for us.

```{r}
#| label: binary-dummy-model

# Fit model
model_protocol <- lm(length_of_stay ~ protocol + age + bmi + surgery_complexity,
                     data = surgical_data)
summary(model_protocol)
```

**Key Interpretation**:

-   The coefficient for `protocolERAS` is approximately `r round(coef(model_protocol)["protocolERAS"], 2)`.

-   This means that **patients in the ERAS protocol have a length of stay that is `r abs(round(coef(model_protocol)["protocolERAS"], 2))` days shorter than patients in the Standard protocol, holding age, BMI, and surgery complexity constant.**

-   The Standard protocol is the **reference category** (omitted from the output). The comparison is made relative to this reference group.

### Manual Calculation with Dummy Variables

```{r}
#| label: manual-calc-dummy-coeffs
#| echo: false

# Extract coefficients
coeffs <- coef(model_protocol)

# Compute predicted values for example patient (Age=65, BMI=30, Complexity=2)
pred_standard <- coeffs[1] + 0 * coeffs[2] + 65 * coeffs[3] + 30 * coeffs[4] + 2 * coeffs[5]
pred_eras <- coeffs[1] + 1 * coeffs[2] + 65 * coeffs[3] + 30 * coeffs[4] + 2 * coeffs[5]
```

Let's manually calculate predicted length of stay to understand how dummy variables work in the regression equation.

**Step 1: Recall the fitted equation**

$\widehat{\text{LOS}} = \beta_0 + \beta_1(\text{ERAS}) + \beta_2(\text{Age}) + \beta_3(\text{BMI}) + \beta_4(\text{Complexity})$

where ERAS is a dummy variable:

-   ERAS = 0 for Standard protocol
-   ERAS = 1 for ERAS protocol

**With the fitted coefficients**, this becomes:

$\widehat{\text{LOS}} = `r round(coeffs[1], 3)` + `r round(coeffs[2], 3)`(\text{ERAS}) + `r round(coeffs[3], 3)`(\text{Age}) + `r round(coeffs[4], 3)`(\text{BMI}) + `r round(coeffs[5], 3)`(\text{Complexity})$

**Step 2: Substitute and calculate**

For a patient who is 65 years old, has a BMI of 30, and underwent medium complexity surgery (complexity = 2):

**Standard protocol** (ERAS = 0):

$$\widehat{\text{LOS}} = `r round(coeffs[1], 3)` + `r round(coeffs[2], 3)` \times 0 + `r round(coeffs[3], 3)` \times 65 + `r round(coeffs[4], 3)` \times 30 + `r round(coeffs[5], 3)` \times 2 = `r round(pred_standard, 2)` \text{ days}$$

**ERAS protocol** (ERAS = 1):

$$\widehat{\text{LOS}} = `r round(coeffs[1], 3)` + `r round(coeffs[2], 3)` \times 1 + `r round(coeffs[3], 3)` \times 65 + `r round(coeffs[4], 3)` \times 30 + `r round(coeffs[5], 3)` \times 2 = `r round(pred_eras, 2)` \text{ days}$$

The difference is **`r round(pred_eras - pred_standard, 2)` days**, which is exactly the coefficient for ERAS (`r round(coeffs[2], 2)`). This confirms that the dummy variable coefficient ($\beta_1$) represents the difference in predicted outcomes between the two groups, holding all other variables constant.

## 1.2 Multi-Level Categorical Variables

When a categorical variable has **more than two levels**, we need **k-1 dummy variables** (where k is the number of categories).

### Why k-1 and Not k?

Including a dummy for every category would create **perfect multicollinearity** (the dummy variables would sum to 1 for every observation, which is perfectly correlated with the intercept).

### Multi-Level Categorical Variables Example: Hospital Type Comparison

#### Background and Context

Hospitals come in different types:

-   **Teaching hospitals**: Large academic medical centers that train medical residents and handle complex cases
-   **Community hospitals**: General hospitals serving local populations
-   **Critical Access hospitals**: Small rural hospitals in remote areas with limited resources

A 2017 study examining nearly 4,500 U.S. hospitals found that readmission rates vary by hospital type, reflecting differences in patient complexity, resources, and care patterns.

In this example, we test whether 30-day readmission rates (the percentage of patients who return to the hospital within 30 days) differ across these hospital types.

Let's compare 30-day readmission rates across three hospital types while controlling for patient characteristics and staffing levels using dummy variables.

#### Data Loading

Let's load a dataset using the code below:

```{r}
#| label: multi-dummy-data

# Generate hospital data
hospital_type_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week12_hospital_type_data.csv")

# Summary by hospital type
head(hospital_type_data, 5)
```

The dataset includes hospitals with the following variables:

-   `hospital_id`: Unique identifier for each hospital
-   `hospital_type`: Type of hospital (Community, Teaching, or Critical Access)
-   `bed_count`: Number of hospital beds
-   `avg_patient_severity`: Average patient severity score
-   `nurse_per_patient`: Nurse-to-patient ratio
-   `readmission_rate`: 30-day readmission rate (our outcome variable)

Before fitting the model, let's visualize how readmission rates differ across hospital types:

```{r}
#| label: fig-hospital-type-explore
#| fig-cap: "Distribution of readmission rates by hospital type"
#| code-fold: true
#| fig-height: 4
#| fig-width: 6

ggplot(hospital_type_data, aes(x = hospital_type, y = readmission_rate, fill = hospital_type)) +
  geom_boxplot(alpha = 0.7, width = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  labs(
    x = "Hospital Type",
    y = "30-Day Readmission Rate (%)",
    title = "Do Readmission Rates Differ by Hospital Type?"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

The boxplot suggests differences in readmission rates across hospital types. Let's use regression to test these differences formally while controlling for patient severity and staffing.

### Fitting the Model with Multiple Categories

Before fitting the model, let's make the variable `hospital_type` as a factor with "Community" as the reference category using the `factor()` function. We will put "Community" first in the `levels` argument to ensure it is the reference category.

This will create two dummy variables:

-   `hospital_typeTeaching`: 1 if the hospital is a teaching hospital, 0 otherwise
-   `hospital_typeCritical Access`: 1 if the hospital is a critical access hospital, 0 otherwise

```{r}
#| label: creating-multilevel-dummy
hospital_type_data <- hospital_type_data |>
  mutate(
    hospital_type = factor(hospital_type, levels = c("Community", "Teaching", "Critical Access"))
    )
```

We can fit a linear regression model using the `lm()` function in R. When we include the variable `hospital_type` in the model, which is a factor, R automatically creates the dummy variables for us.

```{r}
#| label: multi-dummy-model

# Fit model
model_hospital <- lm(readmission_rate ~ hospital_type + avg_patient_severity +
                     nurse_per_patient, data = hospital_type_data)

summary(model_hospital)
```

**Interpretation**:

```{r}
#| echo: false
coef_teach <- round(coef(model_hospital)["hospital_typeTeaching"], 3)
coef_crit <- round(coef(model_hospital)["hospital_typeCritical Access"], 3)
```

-   **Reference Category**: Community hospitals (omitted from output)
-   **Teaching Hospitals**: Readmission rate of Teaching Hospitals is `r abs(coef_teach)` **percentage points lower** than Community hospitals (β = `r coef_teach`)
-   **Critical Access Hospitals**: Readmission rate of Critical Access Hospitals is `r coef_crit` **percentage points higher** than Community hospitals (β = `r coef_crit`)

### Testing Comparisons Between Non-Reference Categories

The regression output provides p-values for comparisons against the **reference category** (Community hospitals), but what about comparing Teaching vs. Critical Access hospitals directly?

We need to test whether the **linear combination** of coefficients is statistically significant. We can use the `emmeans()` function with the `pairs()` function for pairwise comparisons.

```{r}
#| label: emmeans-pairwise

# Calculate estimated marginal means
emm <- emmeans(model_hospital, ~ hospital_type)

# Pairwise comparisons with Tukey adjustment for multiple testing
pairs(emm, adjust = "tukey")
```

This provides all pairwise comparisons with p-values:

-   *Teaching vs. Community*
-   *Critical Access vs. Community*
-   *Teaching vs. Critical Access (this is what we want!)*

Based on these tests, we can conclude whether Teaching hospitals have **statistically significantly** lower readmission rates than Critical Access hospitals, not just numerically lower rates.

Now that we can include categorical predictors in our regression models, let's explore what happens when the effect of one variable depends on the value of another.

------------------------------------------------------------------------

# 2. Interaction Effects

## What is it?

An **interaction effect** occurs when the effect of one variable on the outcome **depends on the value of another variable** (we talked about this in two-way ANOVA). In other words, the relationship between X₁ and Y changes across different levels of X₂.

Mathematically, we add an interaction term that is the product of the two variables:

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 (X_1 \times X_2) + \epsilon$$

The interaction coefficient $\beta_3$ captures how the effect of $X_1$ changes across levels of $X_2$ (and vice versa).

## When to Use It

Use interaction terms when:

-   Theory or prior research suggests an effect should vary across groups
-   You observe different slopes for different groups in exploratory visualizations
-   You want to test whether an intervention works differently for different populations

## Why Interactions Matter in Healthcare

Many healthcare relationships are conditional. For example:

-   **Treatment effectiveness varies by patient characteristics** (e.g., a medication may work better for younger patients compared to older patients)
-   **Resource effects depend on hospital type** (e.g., additional staffing may have different impacts in big vs. small hospitals)
-   **Policy impacts vary by region** (e.g., a new telemedicine policy may affect urban and rural hospitals differently)

Last week and today, we learned that independent variables can be continuous or categorical, so we can think about three possible combinations in the interaction terms:

-   Continuous X Categorical
-   Categorical X Categorical
-   Continuous X Continuous

Let's take a look at each one by one.

## 2.1 Continuous × Categorical Interaction

### Example: Does Staffing Effect Vary by Hospital Ownership?

#### Background and Context

Studies have found that nurse staffing levels significantly affect patient outcomes and satisfaction, though the magnitude of these effects may vary by hospital characteristics including ownership type.

Hospitals have different ownership structures, which can shape how staffing resources translate into patient experience:

-   **For-Profit hospitals**: Owned by investors, primarily focused on financial performance. Additional staff may be directed toward revenue-generating activities rather than patient interaction.
-   **Non-Profit hospitals**: Community-owned and mission-driven, often reinvesting surplus into patient care and quality improvement programs.

Thus, we might expect that additional nurses would have a bigger impact on satisfaction in non-profit hospitals if they have stronger cultures of patient-centered care.

Let's test whether the effect of nurse staffing on patient satisfaction varies by ownership type (this is a continuous (staffing) × categorical (ownership type) interaction).

#### Data Loading

Let's load the dataset using the code below.

```{r}
#| label: interaction-data

# Load the dataset
satisfaction_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week12_satisfaction_data.csv")
head(satisfaction_data, 5)
```

**Variables in the dataset**:

-   `hospital_id`: Unique identifier for each hospital (1 to 300)
-   `ownership`: Hospital ownership type (For-Profit or Non-Profit)
-   `nurse_ratio`: Nurse-to-patient ratio (ranging from 2 to 6)
-   `patient_volume`: Annual patient volume (number of patients)
-   `avg_acuity`: Average patient acuity level (1 = low, 3 = high complexity)
-   `patient_satisfaction`: Patient satisfaction score on 0-100 scale (our outcome variable)

### Visualizing the Potential Interaction

**Before running the model**, always visualize to see if an interaction might exist:

```{r}
#| label: fig-interaction-visual
#| fig-cap: "Potential interaction: different slopes suggest an interaction effect"
#| code-fold: true
#| fig-height: 4
#| fig-width: 6

ggplot(satisfaction_data, aes(x = nurse_ratio, y = patient_satisfaction,
                              color = ownership)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("For-Profit" = "coral", "Non-Profit" = "steelblue")) +
  labs(
    x = "Nurse-to-Patient Ratio",
    y = "Patient Satisfaction Score",
    color = "Ownership Type",
    title = "Does Staffing Effect Differ by Hospital Ownership?",
    subtitle = "Different slopes suggest an interaction effect"
  ) +
  theme_minimal(base_size = 12)
```

The slopes appear different. Staffing seems to have a stronger positive effect in Non-Profit hospitals than in For-Profit hospitals.

### Fitting the Interaction Model

When we fit a model with an interaction, we can use the `*` symbol, which automatically includes both main effects and the interaction. Alternatively, we can write out the terms explicitly using `+` for main effects and `:` for the interaction.

```{r}
#| label: fit-interaction

# Model WITH interaction
# Alternative syntax for interaction: nurse_ratio + ownership + nurse_ratio:ownership
model_interact <- lm(patient_satisfaction ~ nurse_ratio * ownership, data = satisfaction_data)
summary(model_interact)

```

### Interpreting the Interaction

**Understanding the Coefficients**:

```{r}
#| echo: false
int_coefs <- coef(model_interact)
```

1.  **Intercept (`r round(int_coefs["(Intercept)"], 2)`)**: The predicted satisfaction score for a For-Profit hospital with a nurse ratio of 0.

2.  **nurse_ratio (`r round(int_coefs["nurse_ratio"], 2)`)**: The effect of staffing on satisfaction **for the reference group (For-Profit hospitals)**. Each additional nurse per patient increases satisfaction by `r round(int_coefs["nurse_ratio"], 2)` points in For-Profit hospitals.

3.  **ownershipNon-Profit (`r round(int_coefs["ownershipNon-Profit"], 2)`)**: The difference in satisfaction between Non-Profit and For-Profit hospitals **when nurse_ratio = 0** (not practically meaningful, since hospitals can't have zero nurses). This represents the baseline difference between ownership types at zero staffing.

4.  **nurse_ratio:ownershipNon-Profit (`r round(int_coefs["nurse_ratio:ownershipNon-Profit"], 2)`)**: **THE INTERACTION TERM**. This is the additional effect of staffing in Non-Profit hospitals beyond the effect in For-Profit hospitals. The positive value means the slope is steeper for Non-Profit hospitals, and the p-value shows that this difference is statistically significant.

**Computing the Effect of Staffing for Each Group**:

-   **For-Profit hospitals**: β = `r round(int_coefs["nurse_ratio"], 2)`
    -   Each additional nurse increases satisfaction by `r round(int_coefs["nurse_ratio"], 2)` points
-   **Non-Profit hospitals**: β = `r round(int_coefs["nurse_ratio"], 2)` + `r round(int_coefs["nurse_ratio:ownershipNon-Profit"], 2)` = `r round(int_coefs["nurse_ratio"] + int_coefs["nurse_ratio:ownershipNon-Profit"], 2)`
    -   Each additional nurse increases satisfaction by `r round(int_coefs["nurse_ratio"] + int_coefs["nurse_ratio:ownershipNon-Profit"], 2)` points

The interaction reveals that the **context matters**. The same staffing increase produces different outcomes depending on hospital ownership structure. This is precisely what interaction effects help us understand when and where interventions work better.

## 2.2 Categorical × Categorical Interaction

Interactions can also occur between two categorical variables.

### Example: Does Treatment Effect Vary by Insurance Type?

#### Background and Context

Diabetes management programs help patients control blood sugar levels (measured by HbA1c). Different programs exist:

-   **Standard Care**: Regular doctor visits and medication management
-   **Intensive Program**: Structured education, nutrition counseling, and regular coaching

A 2023 study on Medicare Diabetes Prevention Program found significant implementation challenges, with very limited program availability nationwide, suggesting access and effectiveness barriers vary across insurance types.

Additionally, access to additional support services, costs (co-pays and out-of-pocket costs), and patients' time availability could vary by insurance type.

We might expect that the intensive program would show the greatest improvement for privately insured patients, who typically face fewer access barriers, while Medicaid patients may benefit less if co-pays, transportation, or scheduling challenges limit their ability to fully participate.

Let's test whether an intensive diabetes program works differently for patients with different insurance types. This is a categorical (insurance type) × categorical (program type) interaction.

#### Data Loading

Let's load a dataset using the code below:

```{r}
#| label: catcat-interaction

# load data
diabetes_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week12_diabetes_data.csv")
head(diabetes_data, 5)
```

**Variables in the dataset**:

-   `patient_id`: Unique identifier for each patient (1 to 400)
-   `treatment`: Type of diabetes program (Standard Care or Intensive Program)
-   `insurance`: Insurance type (Medicare, Medicaid, or Private)
-   `baseline_a1c`: Baseline HbA1c level before treatment
-   `age`: Patient age in years
-   `a1c_reduction`: Reduction in HbA1c after treatment (our outcome variable)

Before running the model, let's visualize whether the treatment effect appears to differ across insurance types:

```{r}
#| label: fig-catcat-explore
#| fig-cap: "Does treatment effectiveness vary by insurance type?"
#| code-fold: true
#| fig-height: 4
#| fig-width: 6

ggplot(diabetes_data, aes(x = insurance, y = a1c_reduction, fill = treatment)) +
  geom_boxplot(alpha = 0.7, width = 0.7) +
  labs(
    x = "Insurance Type",
    y = "HbA1c Reduction",
    fill = "Treatment"
  ) +
  theme_minimal(base_size = 12)
```

The gap between Standard Care and Intensive Program appears to vary across insurance types, suggesting a potential interaction effect. Let's test this formally.

### Fitting the Interaction Model

First, let's make the variables as factors using the `factor()` function.

-   For the variable `treatment`: "Standard Care" is the reference level
-   For the variable `insurance`: "Medicaid" is the reference level

```{r}
#| label: factor-diabetes-vars
diabetes_data <- diabetes_data |>
  mutate(
    treatment = factor(treatment, levels = c("Standard Care", "Intensive Program")),
    insurance = factor(insurance, levels = c("Medicaid", "Medicare", "Private"))
  )
```

When we put those factor variables into the model, R automatically creates the necessary dummy variables.

```{r}
#| label: catcat-model

# Fit model with interaction
# Alternative syntax: treatment + insurance + treatment:insurance
model_diabetes <- lm(a1c_reduction ~ treatment * insurance,
                     data = diabetes_data)
summary(model_diabetes)
```

### Understanding Categorical × Categorical Interactions

With two categorical variables, the interaction becomes more complex. Let's break down each coefficient:

**Understanding Each Coefficient**:

```{r}
#| echo: false
diab_coefs <- coef(model_diabetes)
```

**Reference Group**: Standard Care + Medicaid (both reference categories)

1.  **Intercept (`r round(diab_coefs["(Intercept)"], 3)`)**: The predicted HbA1c reduction for patients receiving **Standard Care with Medicaid insurance**. This is our baseline group.

2.  **treatmentIntensive Program (`r round(diab_coefs["treatmentIntensive Program"], 3)`)**: The additional effect of the Intensive Program compared to Standard Care **for Medicaid patients only** (the reference insurance group).

    -   Medicaid patients in Intensive Program: `r round(diab_coefs["(Intercept)"], 3)` + `r round(diab_coefs["treatmentIntensive Program"], 3)` = `r round(diab_coefs["(Intercept)"] + diab_coefs["treatmentIntensive Program"], 3)` reduction

3.  **insuranceMedicare (`r round(diab_coefs["insuranceMedicare"], 3)`)**: The difference between Medicare and Medicaid patients **receiving Standard Care only** (the reference treatment).

    -   Medicare patients in Standard Care: `r round(diab_coefs["(Intercept)"], 3)` + `r round(diab_coefs["insuranceMedicare"], 3)` = `r round(diab_coefs["(Intercept)"] + diab_coefs["insuranceMedicare"], 3)` reduction

4.  **insurancePrivate (`r round(diab_coefs["insurancePrivate"], 3)`)**: The difference between Private and Medicaid patients **receiving Standard Care only**.

    -   Private patients in Standard Care: `r round(diab_coefs["(Intercept)"], 3)` + `r round(diab_coefs["insurancePrivate"], 3)` = `r round(diab_coefs["(Intercept)"] + diab_coefs["insurancePrivate"], 3)` reduction

5.  **treatmentIntensive Program:insuranceMedicare (`r round(diab_coefs["treatmentIntensive Program:insuranceMedicare"], 3)`)**: **INTERACTION TERM**. This is the additional treatment effect for Medicare patients beyond what Medicaid patients experience.

    -   This tells us whether the Intensive Program works differently for Medicare vs. Medicaid patients
    -   Medicare patients in Intensive Program: `r round(diab_coefs["(Intercept)"], 3)` + `r round(diab_coefs["treatmentIntensive Program"], 3)` + `r round(diab_coefs["insuranceMedicare"], 3)` + `r round(diab_coefs["treatmentIntensive Program:insuranceMedicare"], 3)` = `r round(diab_coefs["(Intercept)"] + diab_coefs["treatmentIntensive Program"] + diab_coefs["insuranceMedicare"] + diab_coefs["treatmentIntensive Program:insuranceMedicare"], 3)` reduction
    -   A significant interaction means that the treatment effect for Medicare patients is significantly different from the treatment effect for Medicaid patients.

6.  **treatmentIntensive Program:insurancePrivate (`r round(diab_coefs["treatmentIntensive Program:insurancePrivate"], 3)`)**: **INTERACTION TERM**. This is the additional treatment effect for Private insurance patients beyond what Medicaid patients experience.

    -   This tells us whether the Intensive Program works differently for Private vs. Medicaid patients
    -   Private patients in Intensive Program: `r round(diab_coefs["(Intercept)"], 3)` + `r round(diab_coefs["treatmentIntensive Program"], 3)` + `r round(diab_coefs["insurancePrivate"], 3)` + `r round(diab_coefs["treatmentIntensive Program:insurancePrivate"], 3)` = `r round(diab_coefs["(Intercept)"] + diab_coefs["treatmentIntensive Program"] + diab_coefs["insurancePrivate"] + diab_coefs["treatmentIntensive Program:insurancePrivate"], 3)` reduction
    -   A significant interaction means that the treatment effect for Private insurance patients is significantly different from the treatment effect for Medicaid patients.

**Computing Treatment Effects for Each Insurance Group**:

-   **Medicaid**: Treatment effect = `r round(diab_coefs["treatmentIntensive Program"], 3)`
-   **Medicare**: Treatment effect = `r round(diab_coefs["treatmentIntensive Program"], 3)` + `r round(diab_coefs["treatmentIntensive Program:insuranceMedicare"], 3)` = `r round(diab_coefs["treatmentIntensive Program"] + diab_coefs["treatmentIntensive Program:insuranceMedicare"], 3)`
-   **Private**: Treatment effect = `r round(diab_coefs["treatmentIntensive Program"], 3)` + `r round(diab_coefs["treatmentIntensive Program:insurancePrivate"], 3)` = `r round(diab_coefs["treatmentIntensive Program"] + diab_coefs["treatmentIntensive Program:insurancePrivate"], 3)`

The Intensive Program is effective across all insurance types, but significantly more effective for Private insurance patients compared to Medicaid patients. This suggests that program effectiveness depends on insurance type.

## 2.3 Continuous × Continuous Interaction

Interactions can also occur between two continuous variables, where the effect of one continuous variable depends on the level of another continuous variable.

### Example: Does PT Session Benefit Depend on Patient Age?

#### Background and Context

Physical therapy (PT) is commonly prescribed for musculoskeletal conditions like joint pain, post-surgery recovery, and mobility issues. Patients typically attend multiple sessions over several weeks to improve functional capacity. However, the benefit of additional sessions may not be the same for all patients:

-   **Younger patients**: Faster tissue repair, better baseline strength, and greater capacity to respond to exercise-based therapy
-   **Older patients**: Slower recovery, more comorbidities, and potentially different adherence patterns that may limit their response to additional sessions

We might expect that additional PT sessions would provide greater functional improvement for younger patients compared to older patients, due to differences in physiological recovery capacity.

Let's test whether the effect of PT session count on functional improvement varies by patient age. This is a continuous (PT sessions) × continuous (age) interaction.

#### Data Loading

Let's load a dataset using the code below:

```{r}
#| label: contcont-interaction

# load data

pt_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week12_pt_data.csv")
head(pt_data, 5)
```

**Variables in the dataset**:

-   `patient_id`: Unique identifier for each patient (1 to 250)
-   `pt_sessions`: Number of physical therapy sessions attended (4 to 24 sessions)
-   `age`: Patient age in years
-   `baseline_function`: Baseline functional capacity score before therapy (higher scores indicate better pre-treatment function)
-   `function_improvement`: Improvement in functional capacity after therapy in points (our outcome variable)

**Before running the model**, let's visualize whether the relationship between PT sessions and improvement appears to vary by age:

```{r}
#| label: fig-contcont-explore
#| fig-cap: "Does PT session effectiveness vary by patient age?"
#| fig-height: 4
#| fig-width: 6

ggplot(pt_data, aes(x = pt_sessions, y = function_improvement, color = age)) +
  geom_point(alpha = 0.6) +
  scale_color_gradient(low = "steelblue", high = "coral") +
  labs(
    x = "Number of PT Sessions",
    y = "Function Improvement",
    color = "Age"
  ) +
  theme_minimal(base_size = 12)
```

The color gradient suggests that younger patients (blue) tend to benefit more from additional sessions, while older patients (red) may show less improvement. Let's test this formally.

### Fitting the Interaction Model

Let's fit the model using the `lm()` function.

```{r}
#| label: fit-contcont

# Fit model with interaction
# Alternative expression: pt_sessions + age + pt_sessions:age
model_pt <- lm(function_improvement ~ pt_sessions * age,
               data = pt_data)
summary(model_pt)
```

### Interpreting Continuous × Continuous Interactions

**Understanding the Coefficients**:

```{r}
#| echo: false
pt_coefs <- coef(model_pt)
```

1.  **Intercept (`r round(pt_coefs["(Intercept)"], 2)`)**: The predicted improvement for a patient with 0 PT sessions and age 0 (not practically meaningful).

2.  **pt_sessions (`r round(pt_coefs["pt_sessions"], 2)`)**: The effect of one additional PT session **when age = 0** (not practically meaningful, since we don't have newborns in PT).

3.  **age (`r round(pt_coefs["age"], 2)`)**: The effect of being one year older **when pt_sessions = 0** (not practically meaningful, since patients with zero sessions wouldn't be in our study).

4.  **pt_sessions:age (`r round(pt_coefs["pt_sessions:age"], 4)`)**: **THE INTERACTION TERM**. This tells us how the effect of PT sessions changes as patients get older.

    -   The **negative value** means that PT sessions become less effective as age increases
    -   For each additional year of age, the benefit of one PT session decreases by `r abs(round(pt_coefs["pt_sessions:age"], 4))` points

**Computing Marginal Effects at Different Ages**:

In interaction models, we're interested in the **marginal effect**, the effect of one variable (PT sessions) at specific values of another variable (age). Because of the interaction, this marginal effect changes with age.

The marginal effect of PT sessions at any given age is calculated as:

$$\frac{\partial \text{Improvement}}{\partial \text{PT Sessions}} = \beta_{\text{pt\_sessions}} + \beta_{\text{interaction}} \times \text{Age}$$

Let's calculate the marginal effect of one additional PT session at different ages:

-   **Age 30**: Effect = `r round(pt_coefs["pt_sessions"], 2)` + (30 × `r round(pt_coefs["pt_sessions:age"], 4)`) = `r round(pt_coefs["pt_sessions"] + 30 * pt_coefs["pt_sessions:age"], 2)` points per session (positive = beneficial)
-   **Age 50**: Effect = `r round(pt_coefs["pt_sessions"], 2)` + (50 × `r round(pt_coefs["pt_sessions:age"], 4)`) = `r round(pt_coefs["pt_sessions"] + 50 * pt_coefs["pt_sessions:age"], 2)` points per session (negative = associated with worse outcomes)
-   **Age 70**: Effect = `r round(pt_coefs["pt_sessions"], 2)` + (70 × `r round(pt_coefs["pt_sessions:age"], 4)`) = `r round(pt_coefs["pt_sessions"] + 70 * pt_coefs["pt_sessions:age"], 2)` points per session (negative = associated with worse outcomes)

The effect of PT sessions **changes dramatically with age**. For younger patients (age 30), additional PT sessions improve function. However, the effect becomes **negative** for older patients (ages 50 and 70), meaning additional sessions are associated with worse functional outcomes. This crossover pattern demonstrates how interactions can reverse the direction of an effect.

### Visualizing Continuous × Continuous Interaction

For continuous × continuous interactions, we typically show the effect at different levels of the moderator. We'll use the `probe_interaction()` function from the `interactions` package, which makes it easy to visualize interactions and get marginal effects at different levels of a moderator:

```{r}
#| label: fig-contcont
#| fig-cap: "PT sessions effect varies by patient age"
#| fig-height: 4
#| fig-width: 6

# Use probe_interaction to visualize the interaction
# Specify custom age values: 30, 50, 70
probe_interaction(model_pt,
                  pred = pt_sessions,
                  modx = age,
                  modx.values = c(30, 50, 70),
                  interval = TRUE,
                  int.width = 0.95,
                  x.label = "Number of PT Sessions",
                  y.label = "Predicted Function Improvement",
                  main.title = "Interaction: PT Sessions × Age",
                  legend.main = "Patient Age")
```

**Interpretation of marginal effects**:

-   **SIMPLE SLOPES ANALYSIS** (rate of change) shows the effect of PT sessions at each age level
    -   The **slope** (Est.) at that age
    -   **Standard error** (S.E.) and **t-statistic** (t val.)
    -   **p-value** testing if the slope is significantly different from zero
-   The changing slopes confirm our earlier manual calculations
-   All three slopes are statistically significant (p \< 0.05), but they differ in **direction**:
    -   Age 30: Positive slope (`r round(pt_coefs["pt_sessions"] + 30 * pt_coefs["pt_sessions:age"], 2)`) = PT sessions are **beneficial**
    -   Age 50: Negative slope (`r round(pt_coefs["pt_sessions"] + 50 * pt_coefs["pt_sessions:age"], 2)`) = PT sessions are **associated with worse outcomes**
    -   Age 70: More negative slope (`r round(pt_coefs["pt_sessions"] + 70 * pt_coefs["pt_sessions:age"], 2)`) = PT sessions are **associated with even worse outcomes**

**Visual interpretation**: The graph clearly shows the interaction effect: The slopes change from positive to negative as age increases:

-   **Positive slope for younger patients**: Age 30 patients (upward sloping line) gain from each additional session
-   **Negative slope for middle-aged patients**: Age 50 patients (downward sloping line) show worse outcomes with additional sessions
-   **Steeper negative slope for older patients**: Age 70 patients (steeper downward slope) show even worse outcomes with additional sessions

In this data, the negative slopes at older ages suggest that more PT sessions are associated with worse outcomes for older patients. In reality, this pattern would be unusual and would warrant investigation into potential confounding factors (e.g., sicker patients receiving more sessions). This example illustrates how interactions can produce crossover effects where treatment benefits reverse in different subgroups.

------------------------------------------------------------------------

## Practice: Dummy Variables and Interaction Effects {.unnumbered}

::: callout-note
## Your Turn!

Practice fitting and interpreting regression models with dummy variables and interaction effects using a new healthcare scenario.
:::

### Scenario

You are a healthcare analyst evaluating hospital quality improvement initiatives across a national hospital network. Quality scores (0-100) measure overall hospital performance based on patient outcomes, safety, and patient experience.

Your organization wants to understand:

1.  Whether quality scores differ across geographic **regions** (Midwest, Northeast, South, West), after accounting for other hospital characteristics
2.  Whether **IT investment** (e.g., electronic health records, clinical decision support systems) affects quality differently in **teaching** vs. **non-teaching** hospitals

This analysis will guide resource allocation and technology investment decisions across the hospital network.

### Data Loading

Run this code to load the dataset:

```{r}
#| label: practice-quality-data

quality_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week12_quality_data.csv")
head(quality_data, 5)
```

**Dataset Variables:**

| Variable | Description | Type |
|------------------------|------------------------------|------------------|
| `hospital_id` | Unique hospital identifier | Numeric |
| `region` | Geographic region (Midwest, Northeast, South, West) | Categorical |
| `teaching_status` | Hospital type (Non-Teaching or Teaching) | Categorical |
| `it_investment` | IT spending per bed (\$ thousands) | Numeric (continuous) |
| `avg_los` | Average length of stay (days) | Numeric (continuous) |
| `quality_score` | Hospital quality score (0-100 scale) | Numeric (continuous) |

### Task 1: Regression with Multi-Level Dummy Variables

First, convert `region` to a factor with "Midwest" as the reference category. Then fit a regression model predicting `quality_score` from `region`, controlling for `it_investment` and `avg_los`.

```{r}
#| label: practice-task-1
#| eval: false

# Step 1: Convert region to a factor with Midwest as reference
quality_data <- quality_data |>
  mutate(region = factor(___, levels = c("___", "___", "___", "___")))

# Step 2: Fit the model
model_quality <- lm(___ ~ ___ + ___ + ___, data = ___)

# Step 3: View results
summary(___)
```

**Questions:**

1.  How many dummy variables were created for `region`?
2.  Which region has the highest quality scores compared to the reference?
3.  Are all regional differences **relative to Midwest** statistically significant (p \< 0.05)?

### Task 2: Pairwise Comparisons

Use `emmeans()` to compare all pairs of regions directly.

```{r}
#| label: practice-task-2
#| eval: false

# Fill in the blanks
emm_region <- emmeans(___, ~ ___)

# Show estimated marginal means
print(___)

# Pairwise comparisons
pairs(___, adjust = "tukey")
```

**Question:** Which pair of regions has the largest difference in quality scores? Is this difference statistically significant after adjusting for multiple comparisons?

### Task 3: Interaction Model

Fit a model with an interaction between `it_investment` and `teaching_status`, controlling for `region` and `avg_los`.

```{r}
#| label: practice-task-3
#| eval: false

# Fill in the blanks
model_quality_interact <- lm(quality_score ~ ___ * ___ + ___ + ___,
                             data = ___)

# View results
summary(___)
```

**Questions:**

1.  Is the interaction between IT investment and teaching status statistically significant?
2.  How much more effective is each dollar of IT investment in teaching hospitals compared to non-teaching hospitals?

------------------------------------------------------------------------

# Summary

## Key Takeaways

**Dummy Variables**:

-   Allow categorical predictors in regression using 0/1 coding
-   For k categories, R creates k-1 dummy variables automatically
-   Coefficients represent differences from the **reference category**
-   Use `factor(x, levels = c("Reference", ...))` to set the reference
-   Use `emmeans()` with `pairs()` for all pairwise comparisons

**Interaction Effects**:

-   An interaction means the effect of one variable **depends on** another
-   Use `*` in the formula to include both main effects and the interaction: `lm(y ~ x1 * x2)`
-   Three types: Continuous × Categorical, Categorical × Categorical, Continuous × Continuous
-   With interactions present, main effects represent the effect **only for the reference group**
-   To get the effect for other groups: add main effect + interaction coefficient

------------------------------------------------------------------------

# References and Further Reading

## Statistical Methods

-   Gelman, A., & Hill, J. (2006). *Data Analysis Using Regression and Multilevel/Hierarchical Models*. Cambridge University Press.
-   James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning with Applications in R* (2nd ed.). Springer.
-   Fox, J. (2015). *Applied Regression Analysis and Generalized Linear Models* (3rd ed.). SAGE Publications.

## Healthcare Applications

Examples are motivated by the following:

-   Katikam, S. M., Komari, L. K., & Parimi, S. B. R. (2025). Evaluation of Enhanced Recovery After Surgery (ERAS) Protocols in Abdominal Surgery. *Journal of Pharmacy & Bioallied Sciences*, 17(Suppl 2), S1805-S1807.

-   Horwitz, L. I., Bernheim, S. M., Ross, J. S., Herrin, J., Grady, J., Krumholz, H. M., & Drye, E. E. (2017). Hospital characteristics associated with risk-standardized readmission rates. *Medical Care*, 55(5), 528-534. https://pmc.ncbi.nlm.nih.gov/articles/PMC5426655/

-   Turk, M. T., Jakub, K., & Ritchie, N. D. (2023). Stakeholder Analysis: Medicare Diabetes Prevention Program Awareness and Implementation. *American Journal of Managed Care*, 29(6), 308-312. https://pmc.ncbi.nlm.nih.gov/articles/PMC10287029/
