---
title: "Linear Regression: Dummy Variables and Interactions (Practice Activities)"
subtitle: "Business Analytics in Healthcare Management"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
    theme: cosmo
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
library(car)
library(emmeans)
library(interactions)
library(patchwork)

# Set seed for reproducibility
set.seed(2026)
```

# Overview

In this practice session, you will apply the regression techniques learned in the lecture for handling categorical predictors and interaction effects. This is a learning activity designed to help you gain hands-on experience with dummy variables and interactions.

**What you'll practice:**

- Creating and interpreting dummy variables for categorical predictors
- Fitting regression models with binary and multi-level categorical variables
- Testing interaction effects (continuous × categorical, categorical × categorical, continuous × continuous)
- Interpreting interaction coefficients and marginal effects
- Using `emmeans()` for pairwise comparisons
- Centering variables for better interpretation

**Format**:

- **Part 1**: Fill-in-the-blank code sections (marked with `___`) for core concepts
- **Part 2**: Advanced exercises requiring independent analysis

**Instructions**:

1. Complete the code by replacing `___` with appropriate values
2. Run each chunk to verify your results
3. Answer the interpretation questions
4. Work through the advanced exercises at the end
5. Compare your work with solutions (provided at the bottom)

**Note**: This practice is completion-based. The goal is to learn and build confidence—mistakes are part of the learning process!

---

# Recap: Key Concepts

Before starting the activities, let's review essential concepts from the lecture:

## Dummy Variables

**What they are**: Binary (0/1) variables that represent categorical group membership

**How many to create**: For k categories, create k-1 dummy variables
- One category serves as the **reference group** (omitted from model output)
- Each dummy coefficient shows the difference from the reference group

**Example**: For hospital type (Teaching, Community, Critical Access):
- Create 2 dummies if Community is reference:
  - `hospital_typeTeaching`: 1 if Teaching, 0 otherwise
  - `hospital_typeCritical Access`: 1 if Critical Access, 0 otherwise
- Community hospitals coded as 0 on both dummies

## Interaction Effects

**What they are**: Occur when the effect of one variable depends on another variable

**Model syntax in R**:
```{r}
#| eval: false
# These are equivalent:
lm(y ~ x1 * x2, data = mydata)          # Shorthand
lm(y ~ x1 + x2 + x1:x2, data = mydata)  # Explicit
```

**Types of interactions**:
- **Continuous × Categorical**: Does the effect of a continuous predictor vary across categories?
- **Categorical × Categorical**: Does the effect of one category depend on another category?
- **Continuous × Continuous**: Does the effect of one continuous variable change with another continuous variable?

## Interpreting Interaction Coefficients

**Key principle**: When interactions are present, main effects change meaning!

- **Main effect**: Effect when other variable = 0 (or reference category)
- **Interaction term**: How much the effect changes for other values/categories
- **Total effect**: Main effect + (Interaction × value of moderator)

---

# Common Errors and How to Fix Them

You **WILL** encounter errors while practicing—that's completely normal! Here are common errors specific to dummy variables and interactions:

## Error 1: "contrasts can be applied only to factors"

```r
Error in `contrasts<-`(`*tmp*`, value = contr.funs[1 + isOF[nn]]) :
  contrasts can be applied only to factors with 2 or more levels
```

**Cause**: You're trying to use a character variable as a categorical predictor

**Fix**: Convert to factor first:
```{r}
#| eval: false
mydata <- mydata %>%
  mutate(category_var = factor(category_var))
```

## Error 2: Reference category confusion

**Issue**: "The coefficient doesn't match what I expected for this category"

**Cause**: R chooses the reference category alphabetically by default

**Fix**: Explicitly set the reference category:
```{r}
#| eval: false
mydata <- mydata %>%
  mutate(category_var = factor(category_var, levels = c("Reference", "Group1", "Group2")))
# The first level becomes the reference
```

## Error 3: Interaction interpretation mistakes

**Issue**: "I thought the interaction term was the effect for the second group"

**Clarification**:
- Interaction term = **additional** effect beyond the main effect
- Total effect for non-reference group = main effect + interaction
- Always add them together!

## Error 4: "object not found" when using emmeans

```r
Error: object 'model_name' not found
```

**Cause**: The model object doesn't exist (typo or didn't run the model)

**Fix**:
1. Check spelling of model name
2. Make sure you ran the `lm()` code chunk first
3. If using `emmeans()`, ensure the model was successfully created

## Still Stuck?

If you're still getting errors:

1. **Check that categorical variables are factors**: Use `str(mydata)` to check
2. **Run all setup chunks**: Models require data and packages to be loaded
3. **Read error messages carefully**: They often tell you exactly what's wrong
4. **Compare with lecture examples**: Look at similar code in the lecture file

::: {.callout-tip}
## Pro Tip: Testing Interactions

Before fitting interaction models:
1. **Visualize first**: Create plots with different colors/facets by group
2. **Look for different slopes**: If lines are parallel, no interaction
3. **Then test formally**: Fit the model and check p-value of interaction term
:::

---

# Part 1: Guided Practice Tasks

## Task 1.1: Binary Dummy Variables - Fitting the Model

We'll use the Enhanced Recovery After Surgery (ERAS) example from the lecture. First, let's create the dataset:

```{r}
#| label: task1-data
# Generate surgical patient data
n <- 200

surgical_data <- tibble(
  patient_id = 1:n,
  protocol = sample(c("Standard", "ERAS"), n, replace = TRUE),
  age = round(rnorm(n, 62, 12)),
  bmi = round(rnorm(n, 28, 5), 1),
  surgery_complexity = sample(1:3, n, replace = TRUE)
) %>%
  mutate(
    age = pmax(30, pmin(age, 85)),
    bmi = pmax(18, pmin(bmi, 45)),
    length_of_stay = round(
      3 + 0.03 * age + 0.05 * bmi + 1.5 * surgery_complexity +
      -1.8 * (protocol == "ERAS") + rnorm(n, 0, 1), 1
    ),
    length_of_stay = pmax(1, length_of_stay),
    protocol = factor(protocol, levels = c("Standard", "ERAS"))
  )
```

**Your task**: Fit a regression model predicting `length_of_stay` from `protocol`, `age`, `bmi`, and `surgery_complexity`.

**Hint**: Use the `lm()` function. R will automatically create dummy variables when you include a factor variable.

```{r}
#| label: task-1-1
#| eval: false

# Fit the model
model_protocol <- lm(___ ~ ___ + ___ + ___ + ___,
                     data = ___)

# View the results
summary(___)
```

**Question**: What is the coefficient for `protocolERAS`? What does it mean?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.2: Interpreting Binary Dummy Variables

Using your model from Task 1.1, calculate the predicted length of stay for two patients:

- **Patient A**: Standard protocol, age 65, BMI 30, complexity = 2
- **Patient B**: ERAS protocol, age 65, BMI 30, complexity = 2

**Hint**: Use the `predict()` function with a new data frame, or calculate manually using coefficients.

```{r}
#| label: task-1-2
#| eval: false

# Create data for two patients
new_patients <- tibble(
  protocol = factor(c("___", "___"), levels = c("Standard", "ERAS")),
  age = c(___, ___),
  bmi = c(___, ___),
  surgery_complexity = c(___, ___)
)

# Get predictions
predict(___, newdata = ___)
```

**Question**: What is the difference in predicted LOS between these two patients? Does it match the dummy variable coefficient?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.3: Multi-Level Categorical Variables

Now let's work with hospital types (Teaching, Community, Critical Access). Create the dataset:

```{r}
#| label: task3-data
n_hospitals <- 250

hospital_type_data <- tibble(
  hospital_id = 1:n_hospitals,
  hospital_type = sample(c("Teaching", "Community", "Critical Access"),
                        n_hospitals, replace = TRUE,
                        prob = c(0.25, 0.50, 0.25)),
  bed_count = case_when(
    hospital_type == "Teaching" ~ round(rnorm(n_hospitals, 400, 100)),
    hospital_type == "Community" ~ round(rnorm(n_hospitals, 200, 80)),
    hospital_type == "Critical Access" ~ round(rnorm(n_hospitals, 40, 15))
  ),
  avg_patient_severity = round(runif(n_hospitals, 1, 3), 2),
  nurse_per_patient = round(runif(n_hospitals, 2, 5), 2)
) %>%
  mutate(
    bed_count = pmax(10, bed_count),
    readmission_rate = round(
      15 + 2.5 * avg_patient_severity + -0.8 * nurse_per_patient +
      case_when(
        hospital_type == "Teaching" ~ -2.5,
        hospital_type == "Community" ~ 0,
        hospital_type == "Critical Access" ~ 3.0
      ) + rnorm(n_hospitals, 0, 2), 1
    ),
    readmission_rate = pmax(5, pmin(readmission_rate, 30)),
    hospital_type = factor(hospital_type, levels = c("Community", "Teaching", "Critical Access"))
  )
```

**Your task**: Fit a model predicting `readmission_rate` from `hospital_type`, `avg_patient_severity`, and `nurse_per_patient`.

```{r}
#| label: task-1-3
#| eval: false

# Fit model with multi-level categorical variable
model_hospital <- lm(___ ~ ___ + ___ + ___,
                     data = ___)

# View results
summary(___)
```

**Question**: How many dummy variables were created? Which category is the reference group?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.4: Pairwise Comparisons with emmeans

Using your model from Task 1.3, use `emmeans()` to compare all hospital types.

**Hint**: Use `emmeans(model, "variable_name")` to get estimated means, then `pairs()` for comparisons.

```{r}
#| label: task-1-4
#| eval: false

# Get estimated marginal means
emm <- emmeans(___, "___")

# Show the means
print(___)

# Pairwise comparisons
pairs(___)
```

**Question**: Is the difference between Teaching and Critical Access hospitals statistically significant? What is the estimated difference?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.5: Continuous × Categorical Interaction

Let's test whether nurse staffing effects vary by hospital ownership. Create the data:

```{r}
#| label: task5-data
n <- 300

satisfaction_data <- tibble(
  hospital_id = 1:n,
  ownership = sample(c("For-Profit", "Non-Profit"), n, replace = TRUE),
  nurse_ratio = round(runif(n, 2, 6), 2),
  patient_volume = round(rnorm(n, 500, 150)),
  avg_acuity = round(runif(n, 1, 3), 2)
) %>%
  mutate(
    patient_volume = pmax(100, patient_volume),
    patient_satisfaction = round(
      50 + 3 * nurse_ratio + 5 * (ownership == "Non-Profit") +
      2.5 * nurse_ratio * (ownership == "Non-Profit") +
      -0.01 * patient_volume + -3 * avg_acuity + rnorm(n, 0, 5), 1
    ),
    patient_satisfaction = pmax(0, pmin(patient_satisfaction, 100)),
    ownership = factor(ownership, levels = c("For-Profit", "Non-Profit"))
  )
```

**Your task**: Fit a model with an interaction between `nurse_ratio` and `ownership`.

**Hint**: Use `*` between variables to include main effects and interaction, or use `+` for main effects and `:` for interaction.

```{r}
#| label: task-1-5
#| eval: false

# Fit model WITH interaction
model_interact <- lm(patient_satisfaction ~ ___ * ___,
                     data = ___)

# View results
summary(___)
```

**Question**: Is the interaction term statistically significant? What does this tell us?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.6: Computing Marginal Effects from Interactions

Using your model from Task 1.5, calculate the effect of nurse staffing for each ownership type.

**Hint**:
- For reference group (For-Profit): Effect = coefficient of `nurse_ratio`
- For Non-Profit: Effect = coefficient of `nurse_ratio` + interaction coefficient

```{r}
#| label: task-1-6
#| eval: false

# Extract coefficients
coefs <- coef(model_interact)

# Effect for For-Profit hospitals (reference group)
effect_forprofit <- coefs["___"]
cat("For-Profit effect:", round(effect_forprofit, 2), "\n")

# Effect for Non-Profit hospitals (reference + interaction)
effect_nonprofit <- coefs["___"] + coefs["___"]
cat("Non-Profit effect:", round(effect_nonprofit, 2), "\n")

# Difference (should equal interaction coefficient)
cat("Difference:", round(effect_nonprofit - effect_forprofit, 2))
```

**Question**: How much more effective is nurse staffing in Non-Profit vs. For-Profit hospitals?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.7: Visualizing Interactions

Using the `probe_interaction()` function from the `interactions` package, visualize the interaction from Task 1.5.

```{r}
#| label: task-1-7
#| eval: false

# Visualize the interaction
probe_interaction(___,
                  pred = ___,      # Continuous predictor
                  modx = ___,      # Categorical moderator
                  interval = TRUE,
                  x.label = "Nurse-to-Patient Ratio",
                  y.label = "Predicted Patient Satisfaction",
                  main.title = "Interaction: Staffing × Ownership",
                  legend.main = "Ownership Type")
```

**Question**: Looking at the plot, describe how the slopes differ between the two ownership types.

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.8: Centering for Continuous × Continuous Interactions

Let's work with the physical therapy example. Create the data:

```{r}
#| label: task8-data
n <- 250

pt_data <- tibble(
  patient_id = 1:n,
  pt_sessions = round(runif(n, 4, 24)),
  age = round(runif(n, 25, 80)),
  baseline_function = round(runif(n, 30, 70))
) %>%
  mutate(
    function_improvement = round(
      5 + 1.2 * pt_sessions + -0.1 * age + -0.03 * pt_sessions * age +
      0.3 * baseline_function + rnorm(n, 0, 5), 1
    ),
    function_improvement = pmax(0, function_improvement)
  )
```

**Your task**:
1. Center the `age` and `pt_sessions` variables
2. Fit a model with the interaction using centered variables
3. Compare with an uncentered model

```{r}
#| label: task-1-8
#| eval: false

# Center the variables
pt_data <- pt_data %>%
  mutate(
    age_c = ___ - mean(___),
    pt_sessions_c = ___ - mean(___)
  )

# Verify centering worked (means should be ~0)
cat("Mean of age_c:", round(mean(pt_data$age_c), 3), "\n")
cat("Mean of pt_sessions_c:", round(mean(pt_data$pt_sessions_c), 3), "\n")

# Fit model with UNCENTERED variables
model_uncentered <- lm(function_improvement ~ pt_sessions * age,
                       data = pt_data)

# Fit model with CENTERED variables
model_centered <- lm(function_improvement ~ ___ * ___,
                     data = pt_data)

# Compare results
tidy(model_uncentered)
tidy(model_centered)
```

**Question**: Does the interaction coefficient change when you center the variables? What about the main effects? Which model is easier to interpret?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

# Part 2: Advanced Exercises

These exercises require more independent analysis and critical thinking. Use the concepts you practiced above to complete them.

## Exercise 1: Dummy Variables with Multiple Categories

### Background and Context

**The context**: Healthcare quality scores (0-100 scale) measure hospitals on metrics like patient safety, outcomes, and patient experience. These scores are publicly reported and affect hospital reputation and reimbursement.

**The question**: Do quality scores vary by geographic region (Northeast, South, Midwest, West)?

**Why it might vary**:

- Different regulatory environments
- Varying patient populations
- Regional differences in healthcare infrastructure
- Economic and resource disparities

Using the `quality_data` dataset below, analyze how hospital region (`region`) affects quality scores (`quality_score`) while controlling for IT investment (`it_investment`) and average length of stay (`avg_los`).

Please create the dataset using the code below:

```{r}
#| label: practice-data

# Create practice dataset
set.seed(2026)
n <- 350

quality_data <- tibble(
  hospital_id = 1:n,
  region = sample(c("Northeast", "South", "Midwest", "West"), n, replace = TRUE,
                 prob = c(0.2, 0.3, 0.3, 0.2)),
  teaching_status = sample(c("Teaching", "Non-Teaching"), n, replace = TRUE),
  it_investment = round(runif(n, 50, 500), 0),
  avg_los = round(rnorm(n, 4.5, 1), 1)
) %>%
  mutate(
    avg_los = pmax(2, avg_los),
    quality_score = round(
      60 + 0.03 * it_investment + 5 * (teaching_status == "Teaching") +
      case_when(
        region == "Northeast" ~ 3,
        region == "West" ~ 2,
        region == "Midwest" ~ 0,
        region == "South" ~ -2
      ) +
      0.02 * it_investment * (teaching_status == "Teaching") +
      -2 * avg_los + rnorm(n, 0, 5), 1
    ),
    quality_score = pmax(0, pmin(quality_score, 100)),
    region = factor(region),
    teaching_status = factor(teaching_status, levels = c("Non-Teaching", "Teaching"))
  )
```

### Your Tasks

1. Fit a regression model with quality scores (`quality_score`) as the outcome and region (`region`) as a predictor, controlling for IT investment (`it_investment`) and average LOS (`avg_los`)

2. Interpret the dummy variable coefficients. Which region has the highest quality scores compared to the reference category?

3. Change the reference category to "South" using `relevel()`. How do the coefficients change?

```{r}
#| label: practice-exercise1
#| eval: false

# Your code here:

# 1. Fit model with region


# 2. Use emmeans to compare all regions


# 3. Change reference category to South


```

## Exercise 2: Testing for Interactions

### Background and Context

**The hypothesis**: IT investment (health information technology, electronic health records, data systems) might have different impacts on quality depending on hospital type.

**Why the effect might differ**:

- **Teaching hospitals**: Already have sophisticated IT infrastructure and trained staff, so additional investment builds on existing systems
- **Non-teaching hospitals**: May be starting from a lower baseline, facing steeper learning curves

**The question**: Does the effect of IT investment (`it_investment`) on quality scores (`quality_score`) depend on teaching status (`teaching_status`)?

Test for a continuous × categorical interaction using the same dataset.

### Your Tasks

1. Fit a model WITH the interaction between IT investment (`it_investment`) and teaching status (`teaching_status`), controlling for region (`region`) and average LOS (`avg_los`)

2. Interpret the coefficients. Does the effect of IT investment differ significantly depending on teaching status? If so, how do they differ?

3. Create a visualization showing how the effect of IT investment differs by teaching status

```{r}
#| label: practice-exercise2
#| eval: false

# Your code here:

# 1. Model with interaction


# 2. Interpret coefficients - check interaction term significance


# 3. Visualization


```

## Exercise 3: Making Predictions

### Background and Context

Now let's use your interaction model from Exercise 2 to make predictions. Imagine you're a healthcare consultant comparing two very different hospitals:

**Hospital A**: A small non-teaching hospital in the South

- `region` = "South"
- `teaching_status` = "Non-Teaching"
- `it_investment` = $100K per bed (modest investment)
- `avg_los` = 4 days

**Hospital B**: A large teaching hospital in the Northeast

- `region` = "Northeast"
- `teaching_status` = "Teaching"
- `it_investment` = $400K per bed (substantial investment)
- `avg_los` = 4 days

Predict quality scores for both hospitals and understand which factors contribute most to any differences.

### Your Tasks

1. Use your interaction model (`model_interact` from Exercise 2) to predict `quality_score` for both hospitals

2. Compare the predictions. Which hospital is expected to have higher quality scores?

```{r}
#| label: practice-exercise3
#| eval: false

# Your code here:

# Create data for the two hospitals


# Predict quality scores


```

---

# Summary and Reflection

## Key Takeaways

After completing this practice activity, you should be able to:

- ✓ Create and interpret dummy variables for categorical predictors
- ✓ Fit regression models with binary and multi-level categorical variables
- ✓ Use `emmeans()` to compare categories beyond the reference group
- ✓ Test for interaction effects between different types of variables
- ✓ Interpret interaction coefficients correctly
- ✓ Calculate marginal effects at different levels of a moderator
- ✓ Center continuous variables to improve interpretability
- ✓ Visualize interactions using `probe_interaction()`

## Common Pitfalls to Avoid

| **Mistake** | **Consequence** | **Solution** |
|-------------|-----------------|--------------|
| Forgetting to convert categorical variables to factors | R treats them as numeric, leading to wrong interpretations | Always use `factor()` for categorical variables |
| Not knowing which category is the reference | Misinterpreting coefficients | Check factor levels with `levels(mydata$category)` |
| Interpreting main effects without considering interactions | Missing the full story | When interactions are present, interpret them together |
| Not centering variables in continuous × continuous interactions | Main effects are meaningless (at zero) | Center by subtracting the mean |
| Adding interaction terms without checking | Overfitting or misleading results | Always visualize first, then test |

---

# Activity Solutions

Solutions for these activities are available below. Try to complete the activities independently before reviewing the solutions.

## Task 1.1 Solution

```{r}
#| label: solution-1-1
#| code-fold: true

# Fit the model
model_protocol <- lm(length_of_stay ~ protocol + age + bmi + surgery_complexity,
                     data = surgical_data)

# View the results
summary(model_protocol)
```

The coefficient for `protocolERAS` shows the difference in length of stay between ERAS and Standard protocols, controlling for patient characteristics. A negative coefficient indicates that ERAS reduces length of stay.

## Task 1.2 Solution

```{r}
#| label: solution-1-2
#| code-fold: true

# Create data for two patients
new_patients <- tibble(
  protocol = factor(c("Standard", "ERAS"), levels = c("Standard", "ERAS")),
  age = c(65, 65),
  bmi = c(30, 30),
  surgery_complexity = c(2, 2)
)

# Get predictions
predictions <- predict(model_protocol, newdata = new_patients)
print(predictions)

# Difference
cat("Difference:", round(predictions[2] - predictions[1], 2), "days\n")
cat("Protocol coefficient:", round(coef(model_protocol)["protocolERAS"], 2), "days")
```

The difference in predicted LOS exactly equals the dummy variable coefficient for ERAS, demonstrating that the coefficient represents the treatment effect.

## Task 1.3 Solution

```{r}
#| label: solution-1-3
#| code-fold: true

# Fit model with multi-level categorical variable
model_hospital <- lm(readmission_rate ~ hospital_type + avg_patient_severity + nurse_per_patient,
                     data = hospital_type_data)

# View results
summary(model_hospital)
```

Two dummy variables were created (`hospital_typeTeaching` and `hospital_typeCritical Access`) because we have 3 categories. Community is the reference group (first level in the factor).

## Task 1.4 Solution

```{r}
#| label: solution-1-4
#| code-fold: true

# Get estimated marginal means
emm <- emmeans(model_hospital, "hospital_type")

# Show the means
print(emm)

# Pairwise comparisons
pairs(emm)
```

The pairwise comparisons show all possible comparisons between hospital types, including Teaching vs. Critical Access, which wasn't directly available in the regression output.

## Task 1.5 Solution

```{r}
#| label: solution-1-5
#| code-fold: true

# Fit model WITH interaction
model_interact <- lm(patient_satisfaction ~ nurse_ratio * ownership,
                     data = satisfaction_data)

# View results
summary(model_interact)
```

If the interaction term (`nurse_ratio:ownershipNon-Profit`) is statistically significant, it means the effect of nurse staffing differs between ownership types.

## Task 1.6 Solution

```{r}
#| label: solution-1-6
#| code-fold: true

# Extract coefficients
coefs <- coef(model_interact)

# Effect for For-Profit hospitals (reference group)
effect_forprofit <- coefs["nurse_ratio"]
cat("For-Profit effect:", round(effect_forprofit, 2), "\n")

# Effect for Non-Profit hospitals (reference + interaction)
effect_nonprofit <- coefs["nurse_ratio"] + coefs["nurse_ratio:ownershipNon-Profit"]
cat("Non-Profit effect:", round(effect_nonprofit, 2), "\n")

# Difference (should equal interaction coefficient)
cat("Difference:", round(effect_nonprofit - effect_forprofit, 2))
```

The difference between effects equals the interaction coefficient, showing that nurse staffing is more effective in Non-Profit hospitals.

## Task 1.7 Solution

```{r}
#| label: solution-1-7
#| code-fold: true
#| fig-width: 6
#| fig-height: 4

# Visualize the interaction
probe_interaction(model_interact,
                  pred = nurse_ratio,
                  modx = ownership,
                  interval = TRUE,
                  x.label = "Nurse-to-Patient Ratio",
                  y.label = "Predicted Patient Satisfaction",
                  main.title = "Interaction: Staffing × Ownership",
                  legend.main = "Ownership Type")
```

The steeper slope for Non-Profit hospitals shows that they benefit more from additional nurse staffing compared to For-Profit hospitals.

## Task 1.8 Solution

```{r}
#| label: solution-1-8
#| code-fold: true

# Center the variables
pt_data <- pt_data %>%
  mutate(
    age_c = age - mean(age),
    pt_sessions_c = pt_sessions - mean(pt_sessions)
  )

# Verify centering worked
cat("Mean of age_c:", round(mean(pt_data$age_c), 3), "\n")
cat("Mean of pt_sessions_c:", round(mean(pt_data$pt_sessions_c), 3), "\n")

# Fit model with UNCENTERED variables
model_uncentered <- lm(function_improvement ~ pt_sessions * age,
                       data = pt_data)

# Fit model with CENTERED variables
model_centered <- lm(function_improvement ~ pt_sessions_c * age_c,
                     data = pt_data)

# Compare results
cat("\nUncentered model:\n")
tidy(model_uncentered) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable() %>% kable_styling()

cat("\nCentered model:\n")
tidy(model_centered) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable() %>% kable_styling()
```

The interaction coefficient stays the same, but the main effects are now interpretable (effect at mean values rather than at zero).

## Exercise 1 Solution

```{r}
#| label: solution-exercise1
#| code-fold: true

# 1. Fit model with region
model_region <- lm(quality_score ~ region + it_investment + avg_los,
                   data = quality_data)
summary(model_region)

# 2. Use emmeans to compare all regions
emm_region <- emmeans(model_region, "region")
print(emm_region)
pairs(emm_region)

# 3. Change reference category to South
quality_data_reref <- quality_data %>%
  mutate(region = relevel(region, ref = "South"))

model_region_south <- lm(quality_score ~ region + it_investment + avg_los,
                         data = quality_data_reref)
summary(model_region_south)
```

When you change the reference category, the coefficients change to reflect comparisons against the new reference, but the underlying relationships remain the same.

## Exercise 2 Solution

```{r}
#| label: solution-exercise2
#| code-fold: true
#| fig-width: 6
#| fig-height: 4

# 1. Model with interaction
model_interact_ex2 <- lm(quality_score ~ it_investment * teaching_status + region + avg_los,
                         data = quality_data)
summary(model_interact_ex2)

# 2. Interpret coefficients
tidy(model_interact_ex2, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = "Interaction Model Results") %>%
  kable_styling()

# 3. Visualization
probe_interaction(model_interact_ex2,
                  pred = it_investment,
                  modx = teaching_status,
                  x.label = "IT Investment ($ thousands per bed)",
                  y.label = "Predicted Quality Score",
                  main.title = "Interaction: IT Investment × Teaching Status",
                  legend.main = "Teaching Status")
```

If the interaction is significant, IT investment has different effects depending on whether the hospital is a teaching hospital or not.

## Exercise 3 Solution

```{r}
#| label: solution-exercise3
#| code-fold: true

# Create data for the two hospitals
new_hospitals <- tibble(
  it_investment = c(100, 400),
  teaching_status = factor(c("Non-Teaching", "Teaching"),
                          levels = c("Non-Teaching", "Teaching")),
  region = factor(c("South", "Northeast"),
                 levels = levels(quality_data$region)),
  avg_los = c(4, 4)
)

# Predict quality scores
predictions <- predict(model_interact_ex2, newdata = new_hospitals, interval = "confidence")
results <- bind_cols(new_hospitals, as.data.frame(predictions))

results %>%
  kable(caption = "Predicted Quality Scores with 95% Confidence Intervals") %>%
  kable_styling()

cat("\nDifference in predicted quality scores:")
cat(round(predictions[2, "fit"] - predictions[1, "fit"], 2), "points")
```

Hospital B (teaching hospital in Northeast with high IT investment) is predicted to have substantially higher quality scores due to the combination of favorable region, teaching status, high IT investment, and the positive interaction between IT investment and teaching status.

---
