---
title: "Linear Regression 2: <br> Dummy Variables and Interaction Effects"
subtitle: "Week 12"
author: ""
date: "2026-04-01"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    logo: ""
    footer: "Week 12 | Linear Regression 2"
    scrollable: true
    incremental: false
    code-line-numbers: false
    highlight-style: github
    navigation-mode: linear
execute:
  warning: false
  message: false
  echo: true
---

```{r setup}
#| include: false
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
library(emmeans)
library(interactions)
```

## Today's Learning Goals

By the end of this lecture, you will be able to:

::: incremental
1.  **Create and interpret** dummy variables for categorical predictors in regression (binary and multi-level)
2.  **Test and interpret** interaction effects in regression
3.  **Compute** marginal effects to describe how a predictor's impact varies across groups
:::

## Beyond Continuous Predictors

In Week 11, we used continuous predictors (e.g., wait time, staffing hours) to predict outcomes.

. . .

<br>

But many important variables in healthcare are **categorical** (e.g., hospital type, insurance status, treatment protocol, or region).

. . .

Moreover, the effect of a continuous variable can differ depending on those categories.

. . .

<br>

**Dummy variables and interactions** answer more nuanced questions:

::: incremental
-   *Do outcomes differ across groups, after controlling for other factors?*
-   *Does the effect of one variable change depending on the level of another?*
:::

## Why This Matters for Healthcare Management

These techniques directly support real management decisions:

::: incremental
-   **Resource allocation**: Should we prioritize staffing investments in non-profit hospitals, where the satisfaction payoff may be larger?
-   **Cost management**: Does the Enhanced Recovery After Surgery (ERAS) protocol reduce length of stay *equally* across patient severity levels, or should we target it to specific groups?
-   **Performance benchmarking**: Are we comparing hospitals fairly if we don't account for differences in patient mix and hospital type?
:::

## Today's Roadmap

::: incremental
1.  **Dummy Variables** -- Including categorical predictors (binary and multi-level) in regression
2.  **Interaction Effects** -- Testing whether the effect of one variable depends on another
    -   Continuous x Categorical
    -   Categorical x Categorical
    -   Continuous x Continuous
:::

# 1. Dummy Variables {background-color="#40666e"}

## What Are Dummy Variables?

Dummy variables (indicator variables) are **binary** variables representing categorical group membership using 0s and 1s.

. . .

<br>

For a categorical variable with **k categories**, we create **k-1 dummy variables**, with one category serving as the reference group.

. . .

**Example**: 3 Hospital Types (Teaching, Community, Critical Access) → 2 dummy variables:

::: incremental
-   Dummy 1: Teaching (1 if Teaching, 0 otherwise)
-   Dummy 2: Critical Access (1 if Critical Access, 0 otherwise)
-   Community = reference (both dummies = 0)
:::

## When to Use Dummy Variables

Regression requires numerical inputs, but categorical variables are not inherently numerical. Simply assigning numbers (e.g., Teaching=1, Community=2, Critical Access=3) would imply a false ordering.

. . .

Dummy variables solve this by representing group membership as 0s and 1s, allowing us to:

::: incremental
-   Include **categorical predictors** in regression (hospital type, insurance status, treatment group)
-   Test whether **group differences** are statistically significant while *controlling for other variables*
-   **Quantify** the magnitude of differences between categories
:::

## Binary Categorical Variables

For a variable with **two categories**, we need **one dummy variable**:

| Patient | Insurance Status | Dummy: Insured |
|:-------:|:----------------:|:--------------:|
|    1    |     Insured      |       1        |
|    2    |    Uninsured     |       0        |
|    3    |     Insured      |       1        |
|    4    |    Uninsured     |       0        |

. . .

<br>

The coefficient on the dummy variable represents the **difference in the outcome** between the two groups (Insured vs. Uninsured).

## Example Case: Treatment Protocol Comparison

**The problem**: Traditional surgical recovery involves extended hospital stays, which are costly and can lead to complications.

. . .

<br>

**A better approach**: Enhanced Recovery After Surgery (ERAS) protocols use evidence-based practices: better pain management, early mobilization, and earlier nutrition.

. . .

<br>

Let's compare patient outcomes between ***Standard care*** and ***ERAS protocols*** to quantify the treatment effect using a dummy variable.

## Loading Packages and Data

Let's load the required packages and the data using the code below:

```{r}
#| label: binary-dummy-data

library(tidyverse)    # data wrangling and visualization
library(emmeans)      # pairwise comparisons across groups
library(interactions) # visualizing interaction effects

surgical_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week12_surgical_data.csv")
head(surgical_data, 5)
```

## Setting the Reference Category

We use `factor()` to set "Standard" as the reference category:

```{r}
#| label: make-binary-dummy

surgical_data <- surgical_data %>%
  mutate(protocol = factor(protocol, levels = c("Standard", "ERAS")))
```

. . .

<br>

-   When we include `protocol` in the model, R **automatically creates** the dummy variable for us.
-   The first level specified in `factor()` (here, `"Standard"`) becomes the **reference category**.

## Visualizing the Data

Let's visually check the patterns in the data before fitting the model:

```{r}
#| label: fig-protocol-effect
#| code-fold: true
#| fig-width: 6
#| fig-height: 4

ggplot(surgical_data, aes(x = protocol, y = length_of_stay, fill = protocol)) +
  geom_boxplot(alpha = 0.7, width = 0.4) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  labs(
    x = "Treatment Protocol",
    y = "Length of Stay (days)",
    title = "Does ERAS Protocol Affect Length of Stay?"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## Fitting the Model

-   Now we can fit a linear regression model to estimate the effect of the ERAS protocol on length of stay, controlling for age, BMI, and surgery complexity
-   We can use the same `lm()` function as before, just including the `protocol` variable as a predictor

```{r}
#| label: binary-dummy-model
#| results: false

model_protocol <- lm(length_of_stay ~ protocol + age + bmi + surgery_complexity,
                     data = surgical_data)
summary(model_protocol)
```

```{r}
#| label: binary-dummy-coeffs
#| include: false

coeffs <- coef(model_protocol)
```

## Understanding the Dummy Variable Output

```{r}
#| label: interpret-binary
#| echo: false

protocol_tidy <- tidy(model_protocol, conf.int = TRUE)

protocol_tidy %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = "Regression Results: Protocol Comparison") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Key Interpretation**:

-   The coefficient for `protocolERAS` is approximately `r round(coef(model_protocol)["protocolERAS"], 2)`.
    -   **Patients in the ERAS protocol have `r abs(round(coef(model_protocol)["protocolERAS"], 2))` days shorter** length of stay than patients in the Standard protocol, holding age, BMI, and surgery complexity constant.

## Manual Calculation with Dummy Variables

**Patient profile**: A 65-year-old patient with BMI of 30 and surgery complexity of 2.

. . .

The fitted equation:

$$\widehat{\text{LOS}} = `r round(coeffs[1], 3)` + `r round(coeffs[2], 3)`(\text{ERAS}) + `r round(coeffs[3], 3)`(\text{Age}) + `r round(coeffs[4], 3)`(\text{BMI}) + `r round(coeffs[5], 3)`(\text{Complexity})$$

. . .

**Standard protocol** (ERAS = 0):

$$\widehat{\text{LOS}} = `r round(coeffs[1], 3)` + `r round(coeffs[2], 3)`(0) + `r round(coeffs[3], 3)`(65) + `r round(coeffs[4], 3)`(30) + `r round(coeffs[5], 3)`(2) = `r round(coeffs[1] + 0 * coeffs[2] + 65 * coeffs[3] + 30 * coeffs[4] + 2 * coeffs[5], 2)` \text{ days}$$

. . .

**ERAS protocol** (ERAS = 1):

$$\widehat{\text{LOS}} = `r round(coeffs[1], 3)` + `r round(coeffs[2], 3)`(1) + `r round(coeffs[3], 3)`(65) + `r round(coeffs[4], 3)`(30) + `r round(coeffs[5], 3)`(2) = `r round(coeffs[1] + 1 * coeffs[2] + 65 * coeffs[3] + 30 * coeffs[4] + 2 * coeffs[5], 2)` \text{ days}$$

## The Difference Is Exactly the Coefficient

```{r}
#| label: dummy-diff
#| echo: false

pred_standard <- coeffs[1] + 0 * coeffs[2] + 65 * coeffs[3] + 30 * coeffs[4] + 2 * coeffs[5]
pred_eras <- coeffs[1] + 1 * coeffs[2] + 65 * coeffs[3] + 30 * coeffs[4] + 2 * coeffs[5]
```

For the same patient (age 65, BMI 30, complexity 2):

-   **Standard protocol**: `r round(pred_standard, 2)` days
-   **ERAS protocol**: `r round(pred_eras, 2)` days
-   **Difference**: `r round(pred_eras - pred_standard, 2)` days = ERAS coefficient (`r round(coeffs[2], 2)`)

. . .

<br> The dummy variable coefficient represents the difference in predicted outcomes between the two groups, holding all other variables constant. This is true **regardless of the patient profile** we plug in.

# 2. Multi-Level Categorical Variables {background-color="#40666e"}

## Why k-1 Dummy Variables?

When a categorical variable has **more than two levels**, we need **k-1 dummy variables** (where k is the number of categories).

. . .

<br>

Including a dummy for every category would create **perfect multicollinearity**.

-   The dummy variables would sum to 1 for every observation, perfectly correlated with the intercept.

. . .

<br> Each of the k-1 dummy coefficients represents the **difference** between that category and the reference category.

## Example Case: Hospital Type Comparison

**Question**: Do 30-day readmission rates differ across 3 hospital types?

::: incremental
-   **Teaching hospitals**: Large academic medical centers handling complex cases
-   **Community hospitals**: General hospitals serving local populations
-   **Critical Access hospitals**: Small rural hospitals with limited resources
:::

. . .

<br> Let's compare readmission rates across these three hospital types while controlling for patient characteristics.

## Loading the Data

Let's load the data using the code below:

```{r}
#| label: multi-dummy-data

hospital_type_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week12_hospital_type_data.csv")
head(hospital_type_data, 5)
```

## Visualizing the Data

Let's visually inspect the pattern:

```{r}
#| label: fig-hospital-type-explore
#| code-fold: true
#| fig-width: 9
#| fig-height: 4.5

ggplot(hospital_type_data, aes(x = hospital_type, y = readmission_rate, fill = hospital_type)) +
  geom_boxplot(alpha = 0.7, width = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  labs(
    x = "Hospital Type",
    y = "30-Day Readmission Rate (%)",
    title = "Do Readmission Rates Differ by Hospital Type?"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## Setting Reference Category and Fitting the Model

Let's set the reference category to "Community":

```{r}
#| label: creating-multilevel-dummy

hospital_type_data <- hospital_type_data %>%
  mutate(
    hospital_type = factor(hospital_type,
                          levels = c("Community", "Teaching", "Critical Access"))
  )
```

. . .

Let's fit the model using the referenced factor variable:

```{r}
#| label: multi-dummy-model
#| results: false

model_hospital <- lm(readmission_rate ~ hospital_type + avg_patient_severity +
                     nurse_per_patient, data = hospital_type_data)

summary(model_hospital)
```

## Interpreting Multi-Level Dummy Output

```{r}
#| label: interpret-multi
#| echo: false

hospital_coefs <- tidy(model_hospital, conf.int = TRUE) %>%
  mutate(
    term = case_when(
      term == "(Intercept)" ~ "Intercept",
      term == "hospital_typeTeaching" ~ "Hospital Type: Teaching",
      term == "hospital_typeCritical Access" ~ "Hospital Type: Critical Access",
      term == "avg_patient_severity" ~ "Average Patient Severity",
      term == "nurse_per_patient" ~ "Nurse-to-Patient Ratio"
    )
  )

hospital_coefs %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = "Regression Results: Hospital Type Effect on Readmissions") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Interpreting Multi-Level Dummy Output

```{r}
#| echo: false
coef_teach <- round(coef(model_hospital)["hospital_typeTeaching"], 3)
coef_crit <- round(coef(model_hospital)["hospital_typeCritical Access"], 3)
```

Key interpretations:

-   **Reference Category**: Community hospitals (omitted from output)
-   **Teaching Hospitals**: Readmission rate `r abs(coef_teach)` percentage points **`r ifelse(coef_teach < 0, "lower", "higher")`** than Community
-   **Critical Access Hospitals**: Readmission rate `r abs(coef_crit)` percentage points **`r ifelse(coef_crit > 0, "higher", "lower")`** than Community

## Testing Non-Reference Comparisons

The regression output only compares against the reference category. To compare **Teaching vs. Critical Access** directly using the `emmeans()` function:

```{r}
#| label: emmeans-pairwise

emm <- emmeans(model_hospital, ~ hospital_type)
pairs(emm, adjust = "tukey")
```

## Changing the Reference Category

You can change the reference with `relevel()`:

```{r}
#| label: relevel

hospital_type_data_reref <- hospital_type_data %>%
  mutate(hospital_type = relevel(hospital_type, ref = "Teaching"))

model_reref <- lm(readmission_rate ~ hospital_type + avg_patient_severity +
                  nurse_per_patient, data = hospital_type_data_reref)
summary(model_reref)
```

# 3. Interaction Effects {background-color="#40666e"}

## What Is an Interaction Effect?

An **interaction effect** occurs when the effect of one variable on the outcome **depends on the value of another variable**.

. . .

<br>

Mathematically:

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 (X_1 \times X_2) + \epsilon$$

<br>

-   The interaction coefficient $\beta_3$ captures how the effect of $X_1$ **changes across levels** of $X_2$ (or vice versa)

## When to Use Interactions

Use interaction terms when:

::: incremental
-   **Theory** or prior research suggests an effect should vary across groups
-   **Visual inspection** shows different slopes for different groups
-   You want to test whether an intervention **works differently** for different populations or conditions
:::

. . .

<br>

**Examples**:

::: incremental
-   A medication may work better for younger vs. older patients
-   Additional staffing may have different impacts in big vs. small hospitals
-   A new policy may affect urban and rural hospitals differently
:::

# 3.1. Continuous x Categorical Interaction {background-color="#40666e"}

## Example: <br> Does Staffing Effect Vary by Hospital Ownership?

**Question**: Does improving nurse staffing increase patient satisfaction equally in all hospitals?

. . .

<br>

**Why it might differ**:

::: incremental
-   **For-Profit hospitals**: Investor-owned, focused on financial performance
-   **Non-Profit hospitals**: Community-owned, mission-driven, may invest more in quality
-   **Hypothesis**: Additional nurses might have a bigger impact in non-profit hospitals
:::

## Loading the Data

Let's load the data using the code below:

```{r}
#| label: interaction-data

satisfaction_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week12_satisfaction_data.csv")

# Set ownership as a factor with For-Profit as the reference category
satisfaction_data <- satisfaction_data %>%
  mutate(ownership = factor(ownership, levels = c("For-Profit", "Non-Profit")))

head(satisfaction_data, 5)
```

## Visualizing the Potential Interaction

Before running the model, always visualize to see if an interaction might exist:

```{r}
#| label: fig-interaction-visual
#| code-fold: true
#| fig-width: 9
#| fig-height: 4.5

ggplot(satisfaction_data, aes(x = nurse_ratio, y = patient_satisfaction,
                              color = ownership)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "Nurse-to-Patient Ratio",
    y = "Patient Satisfaction Score",
    color = "Ownership Type",
    title = "Does Staffing Effect Differ by Hospital Ownership?",
    subtitle = "Different slopes suggest an interaction effect"
  ) +
  theme_minimal(base_size = 12)
```

## Fitting the Interaction Model

Use `*` to include both main effects and the interaction (as we did in two-way ANOVA):

```{r}
#| label: fit-interaction
#| results: false

# nurse_ratio * ownership = nurse_ratio + ownership + nurse_ratio:ownership
model_interact <- lm(patient_satisfaction ~ nurse_ratio * ownership,
                     data = satisfaction_data)
summary(model_interact)
```

## Interpreting the Interaction

```{r}
#| label: interpret-interaction
#| echo: false

int_coefs <- coef(model_interact)

tidy(model_interact, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = "Regression with Interaction Effect") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

<br> The full model: $$\widehat{\text{Satisfaction}} = `r round(int_coefs[1], 2)` + `r round(int_coefs[2], 2)`(\text{Nurse Ratio}) + `r round(int_coefs[3], 2)`(\text{Non-Profit}) + `r round(int_coefs[4], 2)`(\text{Nurse Ratio} \times \text{Non-Profit})$$

## Understanding the Coefficients

::: incremental
1.  **nurse_ratio (`r round(int_coefs["nurse_ratio"], 2)`)**: Effect of staffing **for For-Profit hospitals** (reference group)
2.  **ownershipNon-Profit (`r round(int_coefs["ownershipNon-Profit"], 2)`)**: Difference between Non-Profit and For-Profit **when nurse_ratio = 0** (mathematical baseline, not a realistic scenario)
3.  **nurse_ratio:ownershipNon-Profit (`r round(int_coefs["nurse_ratio:ownershipNon-Profit"], 2)`)**: **THE INTERACTION** -- the additional effect of staffing in Non-Profit hospitals beyond the For-Profit effect
:::

. . .

**Computing the Effect for Each Group**:

::: incremental
-   **For-Profit**: Each additional nurse → `r round(int_coefs["nurse_ratio"], 2)` satisfaction points
-   **Non-Profit**: Each additional nurse → `r round(int_coefs["nurse_ratio"], 2)` + `r round(int_coefs["nurse_ratio:ownershipNon-Profit"], 2)` = `r round(int_coefs["nurse_ratio"] + int_coefs["nurse_ratio:ownershipNon-Profit"], 2)` satisfaction points
:::

## Visualizing the Interaction Effect

```{r}
#| label: fig-interaction-plot
#| echo: false
#| fig-width: 9
#| fig-height: 4.5

pred_grid <- expand_grid(
  nurse_ratio = seq(2, 6, by = 0.1),
  ownership = c("For-Profit", "Non-Profit")
) %>%
  mutate(predicted = predict(model_interact, newdata = .))

ggplot(pred_grid, aes(x = nurse_ratio, y = predicted, color = ownership)) +
  geom_line(linewidth = 1.5) +
  scale_color_manual(values = c("For-Profit" = "coral", "Non-Profit" = "steelblue")) +
  labs(
    x = "Nurse-to-Patient Ratio",
    y = "Predicted Patient Satisfaction",
    color = "Ownership Type",
    title = "Interaction Effect: Staffing x Ownership",
    subtitle = paste0("For-Profit slope: ", round(int_coefs["nurse_ratio"], 2),
                     " | Non-Profit slope: ",
                     round(int_coefs["nurse_ratio"] + int_coefs["nurse_ratio:ownershipNon-Profit"], 2))
  ) +
  theme_minimal(base_size = 12)
```

# 3.2. Categorical x Categorical Interaction {background-color="#40666e"}

## Example: <br> Does Treatment Effect Vary by Insurance Type?

**Question**: Does an intensive diabetes (vs. standard) program work differently for patients with different insurance types?

. . .

<br>

**Why it might vary**:

::: incremental
-   **Private**: Fewest access barriers -- broader networks, more specialist availability
-   **Medicaid**: Logistical barriers -- fewer accepting providers, transportation, time off work
-   **Medicare**: More time for appointments (often retired), but mobility/health limitations
:::

## Loading the Data

Let's load the data using the code below:

```{r}
#| label: catcat-interaction

diabetes_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week12_diabetes_data.csv")
head(diabetes_data, 5)
```

## Visualizing the Data

Let's visually inspect the patterns in the data before fitting the model:

```{r}
#| label: fig-catcat-explore
#| code-fold: true
#| fig-width: 9
#| fig-height: 4.5

ggplot(diabetes_data, aes(x = insurance, y = a1c_reduction, fill = treatment)) +
  geom_boxplot(alpha = 0.7, width = 0.7) +
  labs(
    x = "Insurance Type",
    y = "HbA1c Reduction",
    fill = "Treatment",
    title = "Does Treatment Effectiveness Vary by Insurance Type?"
  ) +
  theme_minimal(base_size = 12)
```

## Setting Factors and Fitting the Model

Let's set the reference categories for both treatment and insurance:

```{r}
#| label: factor-diabetes-vars

diabetes_data <- diabetes_data %>%
  mutate(
    treatment = factor(treatment, levels = c("Standard Care", "Intensive Program")),
    insurance = factor(insurance, levels = c("Medicaid", "Medicare", "Private"))
  )
```

. . .

Let's fit the model with the interaction between treatment and insurance:

```{r}
#| label: catcat-model
#| results: false

# treatment * insurance = treatment + insurance + treatment:insurance
model_diabetes <- lm(a1c_reduction ~ treatment * insurance,
                     data = diabetes_data)
summary(model_diabetes)
```

## Interpreting Categorical x Categorical Interactions

```{r}
#| label: interpret-catcat
#| echo: false

tidy(model_diabetes, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = "Regression Results: Treatment x Insurance Interaction") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r}
#| echo: false
diab_coefs <- coef(model_diabetes)
```

## Understanding Each Coefficient

**Reference Group**: Standard Care + Medicaid

::: incremental
1.  **Intercept (`r round(diab_coefs["(Intercept)"], 3)`)**: Predicted HbA1c reduction for Standard Care with Medicaid
2.  **treatmentIntensive Program (`r round(diab_coefs["treatmentIntensive Program"], 3)`)**: Additional effect of Intensive Program **for Medicaid patients only**
3.  **insuranceMedicare (`r round(diab_coefs["insuranceMedicare"], 3)`)**: Medicare vs. Medicaid difference **for Standard Care only**
4.  **insurancePrivate (`r round(diab_coefs["insurancePrivate"], 3)`)**: Private vs. Medicaid difference **for Standard Care only**
5.  **Interaction: Medicare (`r round(diab_coefs["treatmentIntensive Program:insuranceMedicare"], 3)`)**: How much more effective the Intensive Program is for Medicare **compared to** Medicaid
6.  **Interaction: Private (`r round(diab_coefs["treatmentIntensive Program:insurancePrivate"], 3)`)**: How much more effective the Intensive Program is for Private **compared to** Medicaid
:::

## Treatment Effects by Insurance Group

**Computing treatment effects for each group**:

::: incremental
-   **Medicaid**: Treatment effect = `r round(diab_coefs["treatmentIntensive Program"], 3)`
-   **Medicare**: Treatment effect = `r round(diab_coefs["treatmentIntensive Program"], 3)` + `r round(diab_coefs["treatmentIntensive Program:insuranceMedicare"], 3)` = `r round(diab_coefs["treatmentIntensive Program"] + diab_coefs["treatmentIntensive Program:insuranceMedicare"], 3)`
-   **Private**: Treatment effect = `r round(diab_coefs["treatmentIntensive Program"], 3)` + `r round(diab_coefs["treatmentIntensive Program:insurancePrivate"], 3)` = `r round(diab_coefs["treatmentIntensive Program"] + diab_coefs["treatmentIntensive Program:insurancePrivate"], 3)`
:::

. . .

<br>

> The Intensive Program is effective across all insurance types, but **most effective** for `r c("Medicaid", "Medicare", "Private")[which.max(c(diab_coefs["treatmentIntensive Program"], diab_coefs["treatmentIntensive Program"] + diab_coefs["treatmentIntensive Program:insuranceMedicare"], diab_coefs["treatmentIntensive Program"] + diab_coefs["treatmentIntensive Program:insurancePrivate"]))]` insurance patients.

## Visualizing Categorical x Categorical Interaction

```{r}
#| label: fig-catcat-bar
#| echo: false
#| fig-width: 9
#| fig-height: 4.5

emm_df <- as.data.frame(emmeans(model_diabetes, ~ treatment * insurance))

ggplot(emm_df, aes(x = insurance, y = emmean, fill = treatment)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                position = position_dodge(width = 0.8), width = 0.2) +
  scale_fill_manual(values = c("Standard Care" = "gray60",
                               "Intensive Program" = "steelblue")) +
  labs(
    x = "Insurance Type",
    y = "Predicted HbA1c Reduction",
    fill = "Treatment",
    title = "Interaction: Treatment x Insurance"
  ) +
  theme_minimal(base_size = 12)
```

# 3.3. Continuous x Continuous Interaction {background-color="#40666e"}

## Example: <br> Does PT Session Benefit Depend on Patient Age?

**Question**: Does the benefit of additional physical therapy (PT) sessions depend on patient age?

. . .

<br>

**Why it might vary**:

::: incremental
-   Younger patients may have faster tissue repair and adaptation
-   Older patients may have comorbidities limiting response to therapy
-   Thus, additional PT sessions might provide greater improvement for younger patients
:::

## Loading the Data

Let's load the data using the code below:

```{r}
#| label: contcont-interaction

pt_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week12_pt_data.csv")
head(pt_data, 5)
```

## Visualizing the Data

Let's visually inspect the patterns in the data before fitting the model:

```{r}
#| label: fig-contcont-explore
#| code-fold: true
#| fig-width: 9
#| fig-height: 4.5

ggplot(pt_data, aes(x = pt_sessions, y = function_improvement, color = age)) +
  geom_point(alpha = 0.6) +
  scale_color_gradient(low = "steelblue", high = "coral") +
  labs(
    x = "Number of PT Sessions",
    y = "Function Improvement",
    color = "Age",
    title = "Does PT Session Effectiveness Vary by Age?"
  ) +
  theme_minimal(base_size = 12)
```

## Fitting the Model

Let's fit the model with the interaction between PT sessions and age:

```{r}
#| label: fit-contcont
#| results: false

# pt_sessions * age = pt_sessions + age + pt_sessions:age
model_pt <- lm(function_improvement ~ pt_sessions * age,
               data = pt_data)
summary(model_pt)
```

## Interpreting the Results

```{r}
#| label: interpret-contcont
#| echo: false

tidy(model_pt, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "Regression Results: PT Sessions x Age Interaction") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r}
#| echo: false
pt_coefs <- coef(model_pt)
```

## Understanding the Coefficients

::: incremental
1.  **pt_sessions (`r round(pt_coefs["pt_sessions"], 2)`)**: Effect of one additional PT session **when age = 0** (not meaningful)
2.  **age (`r round(pt_coefs["age"], 2)`)**: Effect of one year of age **when pt_sessions = 0** (not meaningful)
3.  **pt_sessions:age (`r round(pt_coefs["pt_sessions:age"], 4)`)**: **THE INTERACTION** -- for each additional year of age, the benefit of one PT session decreases by `r abs(round(pt_coefs["pt_sessions:age"], 4))` points
:::

. . .

**Marginal effect formula**:

$$\text{Effect of PT Sessions} = \beta_{\text{pt_sessions}} + \beta_{\text{interaction}} \times \text{Age}$$

## Computing Marginal Effects at Different Ages

::: incremental
-   **Age 30**: Effect = `r round(pt_coefs["pt_sessions"], 2)` + (30 × `r round(pt_coefs["pt_sessions:age"], 4)`) = `r round(pt_coefs["pt_sessions"] + 30 * pt_coefs["pt_sessions:age"], 2)` points per session
-   **Age 50**: Effect = `r round(pt_coefs["pt_sessions"], 2)` + (50 × `r round(pt_coefs["pt_sessions:age"], 4)`) = `r round(pt_coefs["pt_sessions"] + 50 * pt_coefs["pt_sessions:age"], 2)` points per session
-   **Age 70**: Effect = `r round(pt_coefs["pt_sessions"], 2)` + (70 × `r round(pt_coefs["pt_sessions:age"], 4)`) = `r round(pt_coefs["pt_sessions"] + 70 * pt_coefs["pt_sessions:age"], 2)` points per session
:::

. . .

<br>

The effect of PT sessions **changes dramatically with age**. Positive for younger patients, potentially negative for older patients.

## Visualizing with `probe_interaction()`

We can understand the interaction better by visualizing the predicted improvement across PT sessions for different ages using `probe_interaction()`, which is a part of the `interactions` package:

```{r}
#| label: fig-contcont
#| fig-width: 9
#| fig-height: 4.5

probe_interaction(model_pt,              # fitted model
                  pred = pt_sessions,     # x-axis variable
                  modx = age,             # moderator variable
                  modx.values = c(30, 50, 70))  # values to plot separate lines
```

## Interpreting the Visualization

**Simple Slopes Analysis**:

```{r}
#| echo: false
#| results: asis
ages <- c(30, 50, 70)
for (a in ages) {
  eff <- pt_coefs["pt_sessions"] + a * pt_coefs["pt_sessions:age"]
  direction <- ifelse(eff > 0, "beneficial", "associated with worse outcomes")
  cat(sprintf("- **Age %d**: %s slope -- PT sessions are **%s**\n",
              a, ifelse(eff > 0, "Positive", "Negative"), direction))
}
```

. . .

<br> **Key insight**:

-   The interaction creates a **crossover effect** where the treatment benefit reverses direction in different subgroups.
-   In practice, this would warrant investigation into confounding factors (e.g., sicker older patients may be prescribed more PT sessions, making PT *appear* less effective).

# Summary {background-color="#40666e"}

## Key Takeaways

::: incremental
-   **Dummy variables** let us include categorical predictors in regression using 0s and 1s
-   We can not only compare two categories or groups, but also **multiple categories** by creating k-1 dummy variables
-   Use `emmeans()` for **pairwise comparisons** across multi-level categorical variables
-   **Interaction effects** test whether the effect of one variable depends on another
-   In interaction models, **main effects** have conditional interpretations (at zero or reference level)
-   Use `probe_interaction()` for **visualizing interaction effects**
:::

## Assignments

1.  **Reflection**: Please share your thoughts on the lecture
    -   3 interesting things
    -   3 confusing things
2.  **Lab Assignment 5**: Linear Regression

Both are due by 11:59pm on Sunday, April 5.

## References

**Statistical Methods**:

-   Gelman, A., & Hill, J. (2006). *Data Analysis Using Regression and Multilevel/Hierarchical Models*. Cambridge University Press.
-   James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning with Applications in R* (2nd ed.). Springer.

**Healthcare Applications**:

-   Katikam, S. M., et al. (2025). Evaluation of ERAS Protocols in Abdominal Surgery. *Journal of Pharmacy & Bioallied Sciences*, 17(Suppl 2), S1805-S1807.
-   Horwitz, L. I., et al. (2017). Hospital characteristics associated with risk-standardized readmission rates. *Medical Care*, 55(5), 528-534.
-   Turk, M. T., et al. (2023). Stakeholder Analysis: Medicare Diabetes Prevention Program. *American Journal of Managed Care*, 29(6), 308-312.
