---
title: "Assignment 1: Data Visualization Assignment (Week 04)"
subtitle: ""
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: false
    code-copy: true
    theme: 
      - cosmo
      - lecture.scss
    highlight-style: github
    mainfont: "Clear Sans, -apple-system, system-ui, sans-serif"
    fontsize: 1.1em
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(healthyR.data)
```

# Assignment Overview

- **Total Points**: 10 points  
- **Due Date**: February 8, 2026
- **Submission**: Upload your completed .qmd file to Canvas

## Learning Objectives

This assignment assesses your ability to:

- Select appropriate visualization types for different analytical questions
- Create visualizations using the `ggplot2` package
- Customize plots for professional presentation
- Interpret visualizations in a healthcare management context

## Instructions

1. Complete all 7 tasks below
2. Fill in the blanks (replace `___` with appropriate code)
3. After completing each code chunk, change `#| eval: false` to `#| eval: true` to run it
4. Answer all interpretation questions (1-2 sentences each)
5. Ensure all visualizations have proper labels, titles, and themes
6. Run all code chunks to verify they work before submitting

---

# About the Dataset

**Hospital Administrative Data**

For this assignment, we will use the `healthyR_data` dataset from the `healthyR.data` package. This is a realistic healthcare administrative dataset containing hospital visit records with patient, financial, and quality information.

## Dataset Background

The `healthyR_data` dataset simulates administrative data from a hospital system, including:

- **Patient visit information**: When patients arrived and left, how long they stayed
- **Financial data**: Charges, payments, adjustments, and amounts due
- **Service categories**: Which departments/service lines provided care
- **Quality indicators**: Readmission flags and length of stay comparisons

This type of data is commonly used by healthcare managers to analyze operational efficiency, financial performance, and quality metrics.

## Loading and Preparing the Data

First, run the code below to load required packages. If you haven't installed these packages yet, you can install them by running `install.packages("ggplot2")`, `install.packages("dplyr")`, `install.packages("tidyr")`, and `install.packages("healthyR.data")` in the R console.

```{r}
#| label: load-packages

# Load required packages
library(ggplot2)
library(dplyr)
library(tidyr)
library(healthyR.data)

set.seed(2026)  # For reproducibility
```

Now, run the following code to load a sample of the dataset for this assignment:

```{r}
#| label: load-data

# Load the healthyR_data dataset and take a sample for this assignment
hospital_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week04_assignment_data.csv")

# View first few rows
head(hospital_data, 10)
```

## Key Variables

| Variable | Description | Type |
|----------|-------------|------|
| `mrn` | Medical Record Number (patient identifier) | Character |
| `visit_id` | Unique visit identifier | Character |
| `visit_start_date_time` | Date and time of visit start | DateTime |
| `visit_end_date_time` | Date and time of visit end | DateTime |
| `total_charge_amount` | Total charges for the visit (USD) | Numeric |
| `total_amount_due` | Amount still owed | Numeric |
| `length_of_stay` | Duration of hospital stay (days) | Numeric |
| `service_line` | Department/service providing care | Categorical |
| `readmit_flag` | Whether patient was readmitted | Categorical (Y/N) |
| `payer_grouping` | Insurance payer category | Categorical |
| `visit_month` | Month of the visit (created above) | Categorical |
| `los_category` | Length of stay category (created above) | Categorical |

---

# Grading Rubric

Each task specifies point values. Points are allocated as follows:

| Component | Weight | Description |
|-----------|--------|-------------|
| Code correctness | 50% | Code runs without errors and produces correct visualization type |
| Customization | 30% | Proper labels, titles, themes, and aesthetics as specified |
| Interpretation | 20% | Thoughtful 1-2 sentence response addressing the question |

**Interpretation Guidance**: Your interpretations should be brief but specific. For example:

- ✅ Good: "The median length of stay for Cardiology (4 days) is higher than Surgery (2 days), suggesting cardiac patients typically require longer recovery."
- ❌ Too vague: "The departments are different."
- ❌ Too long: [5+ sentences of analysis]

---

# Assignment Tasks

## Task 1: Distribution of Length of Stay (1 point) {.unnumbered}

Create a **histogram** showing the distribution of patient length of stay (`length_of_stay`).

**Requirements:**

- Use appropriate binwidth (try 1 day per bin)
- Use `fill = "steelblue"` and `color = "white"`
- Include proper axis labels with units
- Apply `theme_minimal()`

```{r}
#| label: task1-histogram
#| eval: false

# Fill in the blanks
ggplot(data = ___, mapping = aes(x = ___)) +
  geom_histogram(binwidth = ___, 
                 fill = ___, 
                 color = ___) +
  labs(title = "Distribution of Patient Length of Stay",
       x = "___",
       y = "___") +
  ___()
```

**Interpretation Question**: Describe the shape of the distribution. Is it symmetric or skewed? Where do most patients fall?

::: {.callout-note icon=false title="✏️ Your Answer (1-2 sentences)"}
*[Write your answer here in the .qmd file]*

:::

---

## Task 2: Patient Volume by Service Line (1.5 points) {.unnumbered}

Create a **bar chart** showing the number of patient visits for each service line (`service_line`).

**Requirements:**

- Use `geom_bar()` to count visits automatically
- Use `fill = "darkgreen"` with `alpha = 0.7`
- Rotate x-axis labels if needed for readability (from 0 to 90 degrees)
- Include proper title and labels

```{r}
#| label: task2-barchart
#| eval: false

# Fill in the blanks
ggplot(data = ___, mapping = aes(x = ___)) +
  geom_bar(fill = ___, 
           alpha = ___) +
  labs(title = "___",
       x = "Service Line",
       y = "___") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = ___, hjust = 1))
```

**Interpretation Question**: Which service line has the highest volume? Which has the lowest?

::: {.callout-note icon=false title="✏️ Your Answer (1-2 sentences)"}
*[Write your answer here in the .qmd file]*

:::

---

## Task 3: Comparing Charges Across Service Lines (1.5 points) {.unnumbered}

Create **grouped box plots** comparing total charge amounts across different service lines.

**Requirements:**

- Map `service_line` to x-axis and `total_charge_amount` to y-axis
- Use `fill` aesthetic to color by service line
- Make the boxplot with alpha = 0.5
- Use appropriate axis labels

```{r}
#| label: task3-boxplot
#| eval: false

# Fill in the blanks
ggplot(data = ___, 
       mapping = aes(x = ___, 
                     y = ___, 
                     fill = ___)) +
  geom_boxplot(alpha = ___) +
  labs(title = "Total Charges by Service Line",
       x = "___",
       y = "___") +
  theme_minimal() +
  theme(legend.position = "",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

**Interpretation Question**: Which service line shows the highest median charges? Which shows the most variability (widest box)?

::: {.callout-note icon=false title="✏️ Your Answer (1-2 sentences)"}
*[Write your answer here in the .qmd file]*

:::

---

## Task 4: Relationship Between Length of Stay and Charges (1.5 points) {.unnumbered}

Create a **scatterplot** showing the relationship between length of stay (`length_of_stay`, x-axis) and total charges (`total_charge_amount`, y-axis). Add a linear trend line.

**Requirements:**

- Use `geom_point()` with `alpha = 0.5` for transparency
- Add `geom_smooth()` with `method = "lm"` for a linear trend line
- Color the trend line red
- Include axis labels with units

```{r}
#| label: task4-scatterplot
#| eval: false

# Fill in the blanks
ggplot(data = ___, 
       mapping = aes(x = ___, y = ___)) +
  geom_point(alpha = ___, color = "darkblue") +
  geom_smooth(method = ___, 
              se = TRUE, 
              color = ___) +
  labs(title = "Relationship Between Length of Stay and Total Charges",
       x = "___",
       y = "___") +
  theme_minimal()
```

**Interpretation Question**: Is the relationship positive, negative, or no clear pattern? What does this mean for hospital costs?

::: {.callout-note icon=false title="✏️ Your Answer (1-2 sentences)"}
*[Write your answer here in the .qmd file]*

:::

---

## Task 5: Monthly Patient Volume Trend (1.5 points) {.unnumbered}

Calculate the number of visits per month (`visit_month`) and create a **line chart** showing the trend.

**Requirements:**

- First, summarize the data to get monthly counts
- Use `geom_line()` and `geom_point()` together
- Remember to use `group = 1` in the aesthetic
- For the line, use `color = "steelblue"` and `linewidth = 1`
- For the points, use `color = "darkorange"` and `size = 3`
- Rotate x-axis labels for readability (from 0 to 90 degrees)

```{r}
#| label: task5-linechart
#| eval: false

# Step 1: Calculate monthly visit counts
monthly_visits <- hospital_data %>%
  group_by(___) %>%
  summarize(visit_count = n())

# Step 2: Create the line chart
ggplot(data = ___, 
       mapping = aes(x = ___, y = ___, group = ___)) +
  geom_line(color = "steelblue", linewidth = ___) +
  geom_point(color = "steelblue", size = ___) +
  labs(title = "Monthly Patient Visit Volume",
       x = "___",
       y = "___") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = ___, hjust = 1))
```

**Interpretation Question**: Do you see any seasonal pattern or trend? Are there months with notably high or low volumes?

::: {.callout-note icon=false title="✏️ Your Answer (1-2 sentences)"}
*[Write your answer here in the .qmd file]*

:::

---

## Task 6: Average Charges by Payer Type (1.5 points) {.unnumbered}

Create a **bar chart** showing the average total charges (`total_charge_amount`) for each payer type (`payer_grouping`).

**Requirements:**

- Use `geom_bar()` with `stat = "summary"` and `fun = "mean"` to calculate averages
- Map `payer_grouping` to the `fill` aesthetic
- Use `alpha = 0.8` for transparency
- Remove the legend (since it's redundant with x-axis labels)
- Rotate x-axis labels for readability (from 0 to 90 degrees)

```{r}
#| label: task6-barchart-means
#| eval: false

# Fill in the blanks
ggplot(data = ___, 
       mapping = aes(x = ___, y = ___, fill = ___)) +
  geom_bar(stat = ___, 
           fun = ___, 
           alpha = ___) +
  labs(title = "Average Total Charges by Payer Type",
       x = "___",
       y = "___") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = ___, hjust = 1))
```

**Interpretation Question**: Which payer type has the highest average charges? Which has the lowest?

::: {.callout-note icon=false title="✏️ Your Answer (1-2 sentences)"}
*[Write your answer here in the .qmd file]*

:::

---

## Task 7: Faceted Analysis - Charges by Payer Type (1.5 points) {.unnumbered}

Create **faceted box plots** showing total charges (`total_charge_amount`) distribution by length of stay category (`los_category`), with separate panels for each payer grouping (`payer_grouping`).

**Requirements:**

- Use `facet_wrap()` to create separate panels by `payer_grouping`
- Map `los_category` to x-axis and `total_charge_amount` to y-axis
- Use `fill` aesthetic for length of stay category
- Rotate x-axis labels for readability (from 0 to 90 degrees)

```{r}
#| label: task7-faceted
#| eval: false

# Fill in the blanks
ggplot(data = ___, 
       mapping = aes(x = ___, 
                     y = ___, 
                     fill = ___)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ ___) +
  labs(title = "Total Charges by Length of Stay Category and Payer Type",
       x = "___",
       y = "___",
       fill = "Length of Stay") +
  theme_minimal() + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = ___, hjust = 1))
```

**Interpretation Question**: Do the charge patterns differ across payer types? Do longer stays consistently result in higher charges regardless of payer?

::: {.callout-note icon=false title="✏️ Your Answer (1-2 sentences)"}
*[Write your answer here in the .qmd file]*

:::

---

# Submission Checklist

Before submitting, verify:

- [ ] All 7 code chunks run without errors
- [ ] All blanks are filled in with appropriate values
- [ ] All `#| eval: false` changed to `#| eval: true`
- [ ] All visualizations have titles and axis labels
- [ ] All 7 interpretation questions are answered (1-2 sentences each)
- [ ] Visualizations use appropriate themes and colors
- [ ] File knits to HTML successfully

---

# Points Summary

| Task | Points | Visualization Type |
|------|--------|-------------------|
| Task 1 | 1.0 | Histogram (length of stay distribution) |
| Task 2 | 1.5 | Bar chart (service line volume) |
| Task 3 | 1.5 | Grouped box plots (charges by service line) |
| Task 4 | 1.5 | Scatterplot with trend line |
| Task 5 | 1.5 | Line chart (monthly trends) |
| Task 6 | 1.5 | Bar chart (average charges by payer) |
| Task 7 | 1.5 | Faceted box plots |
| **Total** | **10.0** | |

---

**Good luck!**
