---
title: "Data Visualization (Practice Activities)"
subtitle: "Business Analytics in Healthcare Management"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
    theme: cosmo
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)

# Set Seed number for reproducibility
set.seed(2026)
```

# Overview

In this practice session, you will apply the data visualization techniques learned in the lecture.
This is a learning activity designed to help you gain a better understanding of how to visualize data using the `ggplot2` package.

**What you'll practice:**

- Histograms and density plots for distributions
- Bar charts for categorical data
- Box plots for comparing groups
- Scatterplots for relationships
- Faceting for small groups or categories
- Color grouping and customization

**Dataset**: Emergency Department (ED) Performance Metrics

**Format**: Fill-in-the-blank code sections (marked with `___`)

**About the dataset:**  

- You will be working with simulated Emergency Department (ED) Performance Metrics data. 
This dataset reflects routine measurements collected by hospitals to monitor their ED efficiency and care quality. 
- Typical variables include average patient wait times, door-to-doctor times, total ED visits, admission rates, and patient satisfaction scores, among others. 
- The data may also include categories such as hospital, week, and patient demographics (e.g., age group, gender). 
- All data are anonymized and serve as realistic practice for healthcare analytics.

**Instructions**:

1. Complete the code by replacing `___` with appropriate values
2. Run each chunk to verify your visualizations
3. Answer the interpretation questions
4. Compare your results with classmates or solutions

**Note**: This practice is completion-based. The goal is to learn and experiment—mistakes are part of the process!

---

# Recap: Key Concepts

Before starting the activities, let's review essential concepts from the lecture:

## Grammar of Graphics

`ggplot2` visualizations are built in layers:

```{r}
#| label: example-grammar
#| eval: false
# Step 1: Initialize the plot with data and aesthetic mappings
ggplot(data = ___, mapping = aes(x = ___, y = ___)) +
  # data: specify your data frame (e.g., ed_data)
  # mapping: define how variables map to visual properties
  # aes(): aesthetic function that links data columns to plot elements
  #   - x: variable for the horizontal axis
  #   - y: variable for the vertical axis (optional for some geoms)

  # Step 2: Add a geometric layer to visualize the data
  geom___() +
  # Choose the appropriate geom based on your visualization goal (see below):

  # Step 3: Add labels to make the plot informative
  labs(title = "___", x = "___", y = "___") +
  # title: descriptive title explaining what the plot shows
  # x: label for x-axis (include units, e.g., "Wait Time (minutes)")
  # y: label for y-axis (include units, e.g., "Number of Patients")

  # Step 4: Apply a theme to control overall appearance
  theme___()
  # Choose a theme for professional appearance:
  #   - theme_minimal(): clean, minimal design (recommended)
  #   - theme_classic(): traditional look with axis lines
  #   - theme_bw(): black and white theme
  #   - theme_light(): light background with subtle gridlines
```

## Common Geoms

- **geom_histogram()**: Distribution of a continuous variable
- **geom_bar()**: Counts of categorical variable
- **geom_boxplot()**: Distribution with quartiles
- **geom_point()**: Scatterplot for relationships
- **geom_smooth()**: Add trend lines
- **geom_violin()**: Distribution shape

## Data Structure

- **Wide format**: One row per subject, multiple columns for measurements
- **Long format**: One row per observation, useful for grouping and faceting

```{r}
#| label: example-pivot
#| eval: false
# Convert wide to long
long_data <- wide_data %>%
  pivot_longer(cols = c(___),        # columns to pivot from wide to long format
               names_to = "___",     # name of new column to store old column names
               values_to = "___")    # name of new column to store the values
```

---

# Common Errors and How to Fix Them

You **WILL** encounter errors while practicing—that's completely normal and part of the learning process! Here are the most common errors and their solutions:

## Error 1: "object not found"

```r
Error: object 'ed_data' not found
```

**Cause**: You didn't run the chunk that creates the dataset

**Fix**:

1. Scroll up to the "Dataset Generation" section
2. Click the green "play" button to run that chunk
3. Then try your code again

## Error 2: "could not find function"

```r
Error: could not find function "ggplot"
```

**Cause**: The ggplot2 package isn't loaded

**Fix**:

1. Scroll to the very top of the file
2. Run the `setup` chunk (with `library(ggplot2)`)
3. If that doesn't work, install the package: `install.packages("ggplot2")`

## Error 3: "mapping must be created by aes()"

**Cause**: You forgot to put variables inside `aes()`

```r
# ❌ Wrong:
ggplot(data = ed_data, x = wait_time_minutes)

# ✅ Right:
ggplot(data = ed_data, aes(x = wait_time_minutes))
```

**Fix**: Wrap your variable mappings in `aes()`

## Error 4: "unexpected symbol" or "+" at end of line

**Cause**: You added a `+` after the last line of your ggplot code

```r
# ❌ Wrong:
ggplot(ed_data, aes(x = age)) +
  geom_histogram() +
  theme_minimal() +    # ← Extra + causes error

# ✅ Right:
ggplot(ed_data, aes(x = age)) +
  geom_histogram() +
  theme_minimal()      # ← No + on the last line
```

**Fix**: Remove the `+` from the last line

## Error 5: "object of type 'closure' is not subsettable"

**Cause**: You wrote `aes` instead of `aes()`—missing the parentheses

```r
# ❌ Wrong:
ggplot(ed_data, mapping = aes)  # Missing ()

# ✅ Right:
ggplot(ed_data, mapping = aes(x = age))
```

**Fix**: Always include `()` after function names like `aes()`

## Error 6: Variable name typos

**Cause**: R is case-sensitive! Variable names must match exactly

```r
# ❌ Wrong:
aes(x = Wait_Time_Minutes)  # Capital letters

# ✅ Right:
aes(x = wait_time_minutes)  # Matches the actual column name
```

**Fix**: Use `names(ed_data)` or `str(ed_data)` to check exact variable names

## Error 7: "Computation failed" in geom_smooth()

```r
Warning: Computation failed in `stat_smooth()`
```

**Cause**: Usually too few data points or inappropriate method for your data

**Fix**:

- Make sure you have enough data points
- Try `method = "loess"` instead of `method = "lm"`
- Check that both x and y are numeric variables

## Still Stuck?

If you're still getting errors:

1. **Check spelling**: Variable names are case-sensitive
2. **Run all previous chunks**: Make sure all setup code has been executed
3. **Restart R**: Session → Restart R, then run chunks from top to bottom
4. **Read the error message carefully**: It often tells you exactly what's wrong
5. **Compare with examples**: Look at the lecture file or solutions for similar code

::: {.callout-tip}
## Pro Tip: Debugging Strategy

When you get an error:

1. **Don't panic!** Errors are learning opportunities
2. **Read the error message** - what line is it pointing to?
3. **Check the basics** - Did you run all previous chunks? Are variable names spelled correctly?
4. **Simplify** - Comment out parts of your code to find which line causes the error
5. **Search** - Copy the error message into Google (many others have had the same error!)
:::

---

# Activity: Emergency Department Performance Metrics

## Scenario

You are analyzing performance metrics from an emergency department (ED) over the past year. The hospital administrator wants to understand:

1. Distribution of patient wait times
2. Differences in wait times across shift times
3. Relationship between ED census (# of patients; crowdedness) and wait times
4. Monthly trends in patient volume

## Dataset Generation

Run this code to generate the ED performance dataset:

```{r}
#| label: generate-data
# Generate ED performance data
ed_data <- data.frame(
  visit_id = 1:500,
  month = sample(month.name, 500, replace = TRUE),
  shift = sample(c("Day", "Evening", "Night"), 500,
                 replace = TRUE, prob = c(0.45, 0.35, 0.20)),
  wait_time_minutes = round(rlnorm(500, meanlog = 3.2, sdlog = 0.8)),
  ed_census = round(rnorm(500, mean = 25, sd = 8)),
  acuity_level = sample(1:5, 500, replace = TRUE,
                        prob = c(0.05, 0.15, 0.35, 0.30, 0.15)),
  length_of_stay_hours = round(rlnorm(500, meanlog = 1.5, sdlog = 0.7), 1)
) %>%
  mutate(
    wait_time_minutes = pmax(5, pmin(wait_time_minutes, 300)),
    ed_census = pmax(5, pmin(ed_census, 50)),
    month = factor(month, levels = month.name)
  )

# Add simulated correlation: higher census = longer wait
ed_data <- ed_data %>%
  mutate(wait_time_minutes = wait_time_minutes + (ed_census - 25) * 2 +
           rnorm(500, 0, 10),
         wait_time_minutes = pmax(5, round(wait_time_minutes)))

# View first few rows
head(ed_data)
```

## Dataset Variables

This dataset contains **500 simulated emergency department visits** with the following variables:

| Variable | Type | Description | Range/Values |
|----------|------|-------------|--------------|
| **visit_id** | Numeric | Unique identifier for each ED visit | 1 to 500 |
| **month** | Factor | Month when the visit occurred | January through December |
| **shift** | Character | Time of day when patient arrived | "Day", "Evening", "Night" |
| **wait_time_minutes** | Numeric | Time from arrival to seeing a provider (minutes) | 5 to 300 minutes |
| **ed_census** | Numeric | Number of patients in ED at time of visit (crowdedness) | 5 to 50 patients |
| **acuity_level** | Numeric | Patient severity level (1=most critical, 5=least critical) | 1, 2, 3, 4, 5 |
| **length_of_stay_hours** | Numeric | Total time spent in ED from arrival to discharge (hours) | Varies (positive values) |

**Key relationships in the data:**

- Higher ED census (more crowded) tends to result in longer wait times
- The dataset simulates realistic patterns seen in emergency departments
- Shift distribution: Day (45%), Evening (35%), Night (20%)

## Examine the Dataset

```{r}
#| label: examine-data
# Check structure
str(ed_data)

# Summary statistics
summary(ed_data)
```

## Task 1.1: Distribution of Wait Times

Create a histogram showing the distribution of patient wait times.

**Hint**: Use `geom_histogram()` with an appropriate binwidth (try 15 or 20 minutes).

```{r}
#| label: task-1-1
#| eval: false

# Fill in the blanks
ggplot(data = ___, mapping = aes(x = ___)) +
  geom_histogram(binwidth = ___, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of ED Patient Wait Times",
    x = "___",
    y = "___"
  ) +
  theme_minimal()
```

**Question**: What is the approximate mean wait time? Is the distribution symmetric or skewed?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.2: Wait Time Distribution (Density Plot)

Create a density plot showing the smooth distribution of wait times. Compare it to the histogram you created in Task 1.1.

**Hint**: Use `geom_density()` instead of `geom_histogram()`. Density plots show a smooth curve rather than bars.

```{r}
#| label: task-1-2
#| eval: false
# Fill in the blanks
ggplot(data = ___, mapping = aes(x = ___)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(
    title = "___",
    x = "Wait Time (minutes)",
    y = "___"
  ) +
  theme_minimal()
```

**Question**: How does the density plot compare to the histogram from Task 1.1? Which do you find easier to interpret?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.3: Wait Times by Shift

Create side-by-side boxplots comparing wait times across different shifts (Day, Evening, Night).

**Hint**: Use `geom_boxplot()` with shift on the x-axis and wait_time_minutes on the y-axis.

```{r}
#| label: task-1-3
#| eval: false
# Fill in the blanks
ggplot(data = ed_data, mapping = aes(x = ___, y = ___)) +
  geom___() +
  labs(
    title = "___",
    x = "___",
    y = "Wait Time (minutes)"
  ) +
  theme_minimal()
```

**Question**: Which shift has the longest median wait time? Which has the most variability?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.4: Relationship Between Census and Wait Time

Create a scatterplot showing the relationship between ED census (number of patients) and wait times. Add a smoothed trend line.

**Hint**: Use `geom_point()` for the scatterplot and `geom_smooth()` for the trend line. Set `method = "lm"` for a linear trend.

```{r}
#| label: task-1-4
#| eval: false
# Fill in the blanks
ggplot(data = ___, mapping = aes(x = ___, y = ___)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  geom_smooth(method = ___, se = TRUE, color = "red") +
  labs(
    title = "___",
    x = "___",
    y = "___"
  ) +
  theme_minimal()
```

**Question**: Is there a positive or negative relationship? What does this mean for ED operations?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.5: Wait Times by Shift with Color

Recreate the scatterplot from Task 1.4, but color-code the points by shift.

**Hint**: Add `color = shift` inside the `aes()` function.

```{r}
#| label: task-1-5
#| eval: false
# Fill in the blanks
ggplot(data = ed_data, mapping = aes(x = ed_census, y = wait_time_minutes,
                                      color = ___)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "ED Wait Times by Census and Shift",
    x = "ED Census (number of patients)",
    y = "Wait Time (minutes)",
    color = "___"
  ) +
  theme_minimal()
```

**Question**: Do different shifts show different relationships between census and wait times?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.6: Monthly Patient Volume

Create a bar chart showing the number of ED visits per month.

**Hint**: Use `geom_bar()` which automatically counts observations. The months are already ordered correctly in the data.

```{r}
#| label: task-1-6
#| eval: false
# Fill in the blanks
ggplot(data = ed_data, mapping = aes(x = ___)) +
  geom___() +
  labs(
    title = "___",
    x = "___",
    y = "Number of Visits"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate month labels 45° for readability
```

**Question**: Are visits evenly distributed across months, or are there patterns?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.7: Faceted Analysis by Acuity Level

Create faceted histograms showing wait time distributions for each acuity level (1-5, where 1 is most critical).

**Hint**: Use `facet_wrap(~ acuity_level)` to create separate panels.

```{r}
#| label: task-1-7
#| eval: false
# Fill in the blanks
ggplot(data = ed_data, mapping = aes(x = wait_time_minutes)) +
  geom_histogram(binwidth = 20, fill = "coral", color = "white") +
  facet_wrap(~ ___) +
  labs(
    title = "___",
    x = "Wait Time (minutes)",
    y = "Count"
  ) +
  theme_minimal()
```

**Question**: Do you notice differences in wait time distributions across acuity levels?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---


# Summary and Reflection

## Key Takeaways

After completing this practice activity, you should be able to:

- ✓ Create histograms and density plots to visualize distributions
- ✓ Build bar charts to compare categorical data
- ✓ Use box plots to compare groups
- ✓ Create scatterplots to examine relationships between variables
- ✓ Add color grouping to reveal patterns across categories
- ✓ Use faceting to create small multiples for comparison
- ✓ Customize plots with labels, themes, and colors

## Common Visualization Types Reference

| **Question** | **Visualization Type** | **ggplot2 Geom** |
|--------------|------------------------|------------------|
| What is the distribution of a continuous variable? | Histogram or density plot | `geom_histogram()`, `geom_density()` |
| How do groups compare on a continuous variable? | Boxplot or violin plot | `geom_boxplot()` |
| What is the frequency of categories? | Bar chart | `geom_bar()` |
| What is the relationship between two continuous variables? | Scatterplot | `geom_point()` |
| How does a trend change over time or categories? | Line chart or bar chart | `geom_line()` |
| How do proportions differ across groups? | Stacked or filled bar chart | `geom_bar(position = "fill")` |


---

# Activity Solutions

Solutions for these activities are available below. Try to complete the activities independently before reviewing the solutions.

## Task 1.1 Solution

```{r}
#| label: solution-1-1
#| code-fold: true

ggplot(data = ed_data, mapping = aes(x = wait_time_minutes)) +
  geom_histogram(binwidth = 20, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of ED Patient Wait Times",
    x = "Wait Time (minutes)",
    y = "Number of Patients"
  ) +
  theme_minimal()
```

The distribution shows wait times are right-skewed with most patients waiting 30-80 minutes, but some experiencing much longer waits.

### Task 1.2 Solution

```{r}
#| label: solution-1-2
#| code-fold: true

ggplot(data = ed_data, mapping = aes(x = wait_time_minutes)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(
    title = "Density Plot of ED Patient Wait Times",
    x = "Wait Time (minutes)",
    y = "Density"
  ) +
  theme_minimal()
```

The density plot shows a smooth curve of the wait time distribution, making it easier to see the overall shape compared to the histogram. The right-skewed pattern is clearly visible.

### Task 1.3 Solution

```{r}
#| label: solution-1-3
#| code-fold: true

ggplot(data = ed_data, mapping = aes(x = shift, y = wait_time_minutes)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "ED Wait Times by Shift",
    x = "Shift",
    y = "Wait Time (minutes)"
  ) +
  theme_minimal()
```

Wait times are relatively similar across shifts, though there may be slightly longer waits during certain shifts (night > evening > day) depending on staffing patterns.

### Task 1.4 Solution

```{r}
#| label: solution-1-4
#| code-fold: true

ggplot(data = ed_data, mapping = aes(x = ed_census, y = wait_time_minutes)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Relationship Between ED Census and Wait Times",
    x = "ED Census (number of patients)",
    y = "Wait Time (minutes)"
  ) +
  theme_minimal()
```

There is a positive relationship: as ED census increases, wait times increase. This suggests capacity constraints during busy periods.

### Task 1.5 Solution

```{r}
#| label: solution-1-5
#| code-fold: true

ggplot(data = ed_data, mapping = aes(x = ed_census, y = wait_time_minutes,
                                      color = shift)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "ED Wait Times by Census and Shift",
    x = "ED Census (number of patients)",
    y = "Wait Time (minutes)",
    color = "Shift"
  ) +
  theme_minimal()
```

The positive relationship between census and wait times appears consistent across shifts, though the slopes may differ slightly.

### Task 1.6 Solution

```{r}
#| label: solution-1-6
#| code-fold: true

ggplot(data = ed_data, mapping = aes(x = month)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Monthly ED Patient Volume",
    x = "Month",
    y = "Number of Visits"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Since the data is simulated with random sampling, visits are relatively evenly distributed. In real data, you might see seasonal patterns (e.g., higher volumes in winter due to flu season).

### Task 1.7 Solution

```{r}
#| label: solution-1-7
#| code-fold: true

ggplot(data = ed_data, mapping = aes(x = wait_time_minutes)) +
  geom_histogram(binwidth = 20, fill = "coral", color = "white") +
  facet_wrap(~ acuity_level) +
  labs(
    title = "Wait Time Distribution by Acuity Level",
    x = "Wait Time (minutes)",
    y = "Count"
  ) +
  theme_minimal()
```

Ideally, higher acuity patients (level 1-2) should have shorter wait times, but this would need to be specifically coded into the data generation process.

---
