---
title: "Logistic Regression: Predicting Binary Outcomes in Healthcare"
subtitle: "MBA642: Business Analytics in Healthcare Management"
Week: "Week 14"
date: 04/13/2026
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: false
    theme: [lecture.scss, cosmo]
    highlight-style: github
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Load required packages
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
library(scales)


# Set seed for reproducibility
set.seed(2026)
```

# Introduction to Logistic Regression {.section-time data-time="10 min"}

## Why Logistic Regression Matters in Healthcare

In healthcare management, many of the most critical decisions involve predicting binary outcomes. For example,

-   Will an event happen or not?
-   Will a patient be readmitted within 30 days or not?
-   Will a patient adhere to their medication regimen or not?
-   Will a treatment be successful or not?

These yes/no questions are fundamentally different from predicting continuous outcomes like blood pressure or length of stay.

Let's think about the following scenario:

A hospital administrator wants to identify patients at high risk of readmission (returning to the hospital after being released) so that care coordinators can provide additional support before discharge. The outcome variable, *readmission*, can only take two values: YES (the patient was readmitted) or NO (the patient was not readmitted). We cannot use regular linear regression for this problem because linear regression is designed to predict continuous outcomes that can take any value.

**Logistic regression** is an appropriate statistical technique when your outcome variable is binary. It allows us to:

-   Predict the **probability** that an event will occur
-   Identify which factors **increase or decrease** the likelihood of the outcome
-   **Quantify the strength** of each predictor's effect
-   Make **data-driven decisions** about resource allocation and interventions

## The Problem with Linear Regression for Binary Outcomes

You might wonder why we cannot just use linear regression to predict a binary outcome coded as 0 and 1. Let's understand why this approach fails.

Imagine we want to predict whether a patient will be readmitted (coded as 1) or not (coded as 0) based on their length of hospital stay (in days). If we use linear regression, we might get an equation like:

$$P(\text{Readmission}) = -0.2 + 0.05 \times \text{Length of Stay}$$

This seems reasonable at first, but there's a fundamental problem. What if a patient has a very long length of stay of 30 days?

$$P(\text{Readmission}) = -0.2 + 0.05 \times 30 = 1.3$$

We get a probability greater than 1! But probabilities must be between 0 and 1. Similarly, for patients with very short stays, we could get negative probabilities. This is nonsensical. You cannot have a 130% chance or a -5% chance of being readmitted.

Linear regression assumes a straight-line relationship that extends infinitely in both directions. But probabilities are bounded between 0 and 1. We need a different approach that respects these boundaries.

## The Logistic Solution: An S-Shaped Curve

Logistic regression solves this problem by using a special S-shaped curve (called the **logistic** function) that naturally stays between 0 and 1. Instead of predicting the probability directly with a straight line, logistic regression:

1.  Transforms the probability into something called **log odds**
2.  Models the log odds as a linear function of the predictors
3.  Transforms the result back to a probability

This transformation ensures that no matter what values your predictors take, the predicted probability will always be between 0 and 1. The relationship "flattens out" at the extremes, creating the characteristic S-shape.

```{r}
#| label: sigmoid-curve
#| echo: false
#| fig-width: 6
#| fig-height: 4

# Create data for comparison
x <- seq(-6, 6, length.out = 200)
linear <- 0.5 + 0.15 * x
logistic <- 1 / (1 + exp(-x))

plot_data <- data.frame(
  x = rep(x, 2),
  y = c(linear, logistic),
  Model = rep(c("Linear Regression", "Logistic Regression"), each = 200)
)

ggplot(plot_data, aes(x = x, y = y, color = Model)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = c(0, 1), linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("Linear Regression" = "red",
                                 "Logistic Regression" = "steelblue")) +
  labs(title = "Linear Regression vs. Logistic Regression",
       subtitle = "Logistic regression keeps predictions between 0 and 1",
       x = "Predictor Value",
       y = "Predicted Probability") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom") +
  annotate("text", x = 6, y = 1.15, label = "Impossible!\n(p > 1)",
           color = "darkred", size = 3) +
  annotate("text", x = -5.5, y = -0.15, label = "Impossible!\n(p < 0)",
           color = "darkred", size = 3)
```

As you can see in the graph above, the linear regression line (red) extends beyond the valid probability range, while the logistic regression curve (blue) smoothly transitions between 0 and 1, approaching but never crossing these boundaries.

Now that we understand **why** we need logistic regression and **how** it solves the boundary problem, let's explore the mathematical foundations that make it work.

------------------------------------------------------------------------

# Understanding the Core Concepts

Before we jump into running logistic regression analyses, it's essential to understand three related concepts: **probability**, **odds**, and **log odds**. These transformations are at the heart of how logistic regression works, and understanding them will help you interpret your results correctly.

## From Probability to Odds

You may already be familiar with **probability**. It's the chance that something will happen, expressed as a number between 0 and 1 (or as a percentage between 0% and 100%). For example, if 30 out of 100 patients are readmitted, the probability of readmission is 0.30 or 30%.

**Odds** express the same information in a different way. Instead of comparing the number of events to the total, odds compare the number of events happening to the number of events not happening:

$$\text{Odds} = \frac{p}{1-p} = \frac{\text{Probability of event happening}}{\text{Probability of event not happening}}$$

Using our readmission example:

-   Probability of readmission = 0.30
-   Probability of NOT being readmitted = 0.70
-   Odds of readmission = 0.30 / 0.70 = 0.429

This means that for every patient who is readmitted, there are about 2.3 patients who are not readmitted (since 1/0.429 ≈ 2.3). Alternatively, we can say the odds are "0.429 to 1" or roughly "7 to 3" in favor of not being readmitted.

### Understanding Odds Values

Here's a helpful table showing how probability and odds relate to each other:

```{r}
#| label: odds-table
#| echo: false

prob_list <- c(0.10, 0.25, 0.33, 0.50, 0.67, 0.75, 0.90)
complement_list <- 1 - prob_list
odds_list <- round(prob_list / complement_list, 2)

odds_table <- data.frame(
  Probability = prob_list,
  `Complement (1-p)` = complement_list,
  Odds = odds_list,
  Interpretation = c(
    "Very unlikely",
    "",
    "",
    "Equal chance (50/50)",
    "",
    "",
    "Very likely"
  )
)

kable(odds_table, col.names = c("Probability (p)", "1 - p", "Odds", "Interpretation"),
      align = "c", digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Key insight about odds:**

-   When odds = 1, the event is equally likely to happen or not happen (50/50 chance)
-   When odds \< 1, the event is less likely than 50/50
-   When odds \> 1, the event is more likely than 50/50
-   Odds can only be **positive** values (they range **from 0 to infinity**)

## From Odds to Log Odds (Logit)

While odds are more useful than probabilities for our purposes (they range from 0 to infinity instead of 0 to 1), they still have a limitation: they cannot be negative. This asymmetry makes them difficult to model with linear equations.

The solution is to take the **natural logarithm** of the odds, which gives us the **log odds** (also called the **logit**):

$$\text{Log Odds} = \ln(\text{Odds}) = \ln\left(\frac{p}{1-p}\right)$$

This transformation has a beautiful property: log odds can range from negative infinity to positive infinity, making them perfect for linear modeling.

```{r}
#| label: logit-table
#| echo: false

logodds_list <- log(odds_list)

logit_table <- data.frame(
  Probability = prob_list,
  `Complement (1-p)` = complement_list,
  Odds = odds_list,
  `Log Odds` = logodds_list
)

kable(logit_table, col.names = c("Probability (p)", "1 - p", "Odds", "Log Odds (Logit)"),
      align = "c", digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Key insight about log odds:**

-   When log odds = 0, the event has a 50/50 chance (probability = 0.5)
-   When log odds \< 0, the event is less likely than 50/50
-   When log odds \> 0, the event is more likely than 50/50
-   Log odds can be any real number (positive or negative)

The **log odds transformation** is what makes logistic regression work. It converts the bounded probability (0 to 1) into an unbounded scale that can be modeled with a linear equation.

## The Logistic Regression Equation

Now we can write the logistic regression equation. Instead of predicting probability directly (which would cause the boundary problems we discussed), we predict the log odds:

$$\ln\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_k X_k$$

where **p** is the probability of the outcome (e.g., probability of readmission).

This equation indicates that "the log odds of the outcome is a linear combination of the predictors."

The coefficients ($\beta$ values) in this equation are **log odds ratios**, which tell us how much the log odds change when the predictor increases by one unit.

By using this equation, we can get **predicted log odds**.

### Converting Back to Probability

Once we have predicted log odds, we can convert back to probability using these relationships:

1.  **From log odds to odds:** $\text{Odds} = e^{\text{log odds}}$ (exponentiated log odds)
2.  **From odds to probability:** $p = \frac{\text{Odds}}{1 + \text{Odds}}$

Or in one step:

$$p = \frac{e^{\beta_0 + \beta_1 X_1 + \cdots}}{1 + e^{\beta_0 + \beta_1 X_1 + \cdots}}$$

Don't worry if these formulas seem complex. R will do all these calculations for us (that's why we use statistical software!). The important thing is to understand the conceptual flow: (1) we model log odds linearly, then (2) transform the results back to probabilities.

## Estimation Method: Maximum Likelihood Estimation

Unlike linear regression, we cannot use the OLS method, because:

1.  **Probability must stay between 0 and 1**: OLS could predict probabilities outside this range (e.g., -0.3 or 1.4), which doesn't make sense.
2.  **Binary outcomes violate OLS assumptions**: The residuals in binary outcomes can't be normally distributed (as they can only take on two values).
3.  **Non-constant variance**: When outcomes are binary, the variance changes systematically with the predicted probability, violating the OLS assumption of constant variance (homoscedasticity).

Logistic regression uses **Maximum Likelihood Estimation (MLE)**, a different approach that:

-   Finds the coefficient values (parameters) that make the observed data **most likely** to have occurred
-   Naturally respects the probability boundaries (predictions always stay between 0 and 1)
-   Works well with the binomial distribution of binary outcomes

You don't need to understand the mathematics of MLE for this course. The key point is that **the estimation method is different from linear regression**, but R handles all the computation automatically. When you use the `glm()` function with `family = binomial`, R is using MLE behind the scenes.

------------------------------------------------------------------------

# 1. Simple Logistic Regression

Simple logistic regression involves predicting a binary outcome using a single predictor variable. This is the foundation for understanding more complex models, so we'll work through this carefully with a detailed healthcare example.

## 1.1 What is it?

Simple logistic regression models the relationship between **one predictor variable** (continuous or categorical) and a **binary outcome** (0 or 1). It estimates the probability of the outcome occurring as a function of the predictor, using the logistic (S-shaped) curve to keep predictions between 0 and 1.

## 1.2 When to Use It

Use simple logistic regression when:

-   Your **outcome variable is binary** (two possible values: 0/1, yes/no)
-   You have **one predictor variable** you want to examine
-   You want to understand **how the predictor affects the probability** of the outcome
-   You want to **quantify the effect** using log odds ratios and odds ratios

## 1.3 Example Case: Predicting Hospital Readmission

You are a quality improvement analyst at a regional hospital. Hospital readmissions within 30 days are a major concern because they indicate potential gaps in care quality and result in financial penalties under Medicare's Hospital Readmissions Reduction Program.

Your hospital has been assessing patient discharge readiness (measured on a 1-10 scale at discharge) and wants to understand: **Does discharge readiness predict the likelihood of 30-day readmission?**

The hypothesis is that patients who feel more ready for discharge have a better understanding of their follow-up care instructions, are more confident in managing their health at home, and therefore are less likely to be readmitted.

The Hospital Readmissions Reduction Program (HRRP), established by the Affordable Care Act, penalizes hospitals with excess readmissions for specific conditions. Research has shown that perceived readiness for hospital discharge is associated with post-discharge outcomes including readmission rates, though the relationship is influenced by multiple factors including patient demographics, clinical complexity, and post-discharge support systems.

## 1.4 Loading and Exploring the Data

Let's first load our dataset and explore it to understand what we're working with.

```{r}
#| label: simple-data

library(tidyverse)
set.seed(2026)

# load data
readmit_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week13_readmit_data.csv")

readmit_data <- readmit_data |>
  mutate(insurance = factor(insurance, levels = c("Medicaid", "Medicare", "Private")))

head(readmit_data, 10) 
```

**Variables in the dataset**:

-   `patient_id`: Unique identifier for each patient (1 to `r nrow(readmit_data)`)
-   `discharge_readiness`: Discharge readiness score (1-10 scale, where higher scores indicate greater readiness for discharge)
-   `age`: Patient age in years
-   `comorbidities`: Number of chronic health conditions (e.g., diabetes, hypertension, heart disease)
-   `insurance`: Insurance type (Private, Medicare, or Medicaid)
-   `readmitted`: Whether the patient was readmitted within 30 days (1 = yes, 0 = no)

Let's visually look at the relationship between discharge readiness and readmission:

```{r}
#| label: simple-viz
#| fig-width: 6
#| fig-height: 4
#| code-fold: true

# Create visualization
n_not <- sum(readmit_data$readmitted == 0)
n_yes <- sum(readmit_data$readmitted == 1)

p1 <- ggplot(readmit_data, aes(x = factor(readmitted), y = discharge_readiness,
                                fill = factor(readmitted))) +
  geom_boxplot(alpha = 0.7, width = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  scale_fill_manual(values = c("0" = "steelblue", "1" = "darkred"),
                    labels = c("Not Readmitted", "Readmitted")) +
  scale_x_discrete(labels = c(
    paste0("Not Readmitted\n(n = ", n_not, ")"),
    paste0("Readmitted\n(n = ", n_yes, ")")
  )) +
  labs(title = "Discharge Readiness Scores by Readmission Status",
       x = "Readmission Status",
       y = "Discharge Readiness Score (1-10)") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))

p1
```

Looking at this boxplot, we can see that patients who were readmitted tend to have lower discharge readiness scores than those who were not readmitted. This suggests there may be a relationship worth investigating with logistic regression.

## 1.5 Running the Logistic Regression

In R, we use the `glm()` function (Generalized Linear Model) to run logistic regression. The key arguments are:

-   `formula`: Specifies the outcome and predictors (e.g., `readmitted ~ discharge_readiness`)
-   `data`: The dataset to use
-   `family = binomial`: Tells R to use logistic regression for binary outcomes

The syntax is similar to linear regression with `lm()`, but we add `family = binomial(link = "logit")` or just `family = binomial` (the default link function for binomial is logit):

```{r}
#| label: simple-model

# Fit logistic regression model
model_simple <- glm(readmitted ~ discharge_readiness,
                    data = readmit_data,
                    family = binomial(link = "logit"))

# View the results
summary(model_simple)
```

## 1.6 Interpreting the Results Step by Step

The output contains a lot of information. Let's work through it piece by piece to understand what it all means.

### Step 1: Understanding the Coefficients Table

The most important part of the output is the "Coefficients" table. Let's take a look at it:

```{r}
#| label: simple-coef
#| echo: false

# Extract coefficients
coef_table <- summary(model_simple)$coefficients
kable(coef_table, digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Each row in this table represents one term in our model:

-   **`(Intercept)`**: This is $\beta_0$, the predicted log odds when discharge readiness = 0
-   **`discharge_readiness`**: This is $\beta_1$, the change in log odds for each one-unit increase in discharge readiness

The columns tell us:

-   **Estimate**: The coefficient value (log odds ratio)
-   **Std. Error**: The uncertainty in our estimate
-   **z value**: The test statistic (Estimate / Std. Error)
-   **Pr(\>\|z\|)**: The p-value for testing whether the coefficient significantly differs from zero

### Step 2: Writing the Model Equation

From our output, we can write the logistic regression equation:

```{r}
#| label: simple-interpret-log-odds-2
#| echo: false

b0 <- round(coef(model_simple)[1], 4)
b1 <- round(coef(model_simple)[2], 4)
b1_p <- format(summary(model_simple)$coefficients[2, 4], digits = 3)

```

**Log Odds of Readmission = `r b0` + (`r b1`) × Discharge Readiness**

This equation tells us that, for each one-unit increase in discharge readiness score, the log odds of readmission changes by the coefficient value.

### Step 3: Interpreting the Log Odds Ratio

The coefficient for discharge readiness is our **log odds ratio**. Let's interpret it:

**Log Odds Ratio for Discharge Readiness:** `r b1`

**Interpretation:**

1.  **DIRECTION**: The coefficient is `r if(b1 < 0) "NEGATIVE" else "POSITIVE"`, which means higher discharge readiness is associated with `r if(b1 < 0) "LOWER" else "HIGHER"` log odds of readmission.

2.  **STATISTICAL SIGNIFICANCE**: The p-value is `r b1_p`. Since p `r if(b1_p < 0.05) "< 0.05" else ">= 0.05"`, this relationship is `r if(b1_p < 0.05) "statistically significant" else "NOT statistically significant"`.

3.  **MAGNITUDE**: For each one-unit increase in discharge readiness score, the log odds of readmission change by `r b1`.

### Step 4: Converting to Odds Ratio

While log odds ratios are mathematically convenient, **odds ratios** are much easier to interpret. We convert by exponentiating:

$$\text{Odds Ratio} = e^{\text{Log Odds Ratio}}$$

We can get Odds Ratio by using the following **R functions**:

-   `coef()`: Extracts coefficients from the model
-   `exp()`: Exponentiates values (raises e to the power of the value) to convert log odds ratios to odds ratios

```{r}
#| label: simple-odds-ratio1
#| echo: true

# Calculate odds ratios
odds_ratios <- exp(coef(model_simple))
```

Then, we get **Odds Ratios:**

-   Intercept: `r round(odds_ratios[1], 4)`
-   Discharge Readiness: `r round(odds_ratios[2], 4)`

Let's interpret the odds ratio for discharge readiness:

**Interpretation:**

```{r}
#| label: simple-odds-ratio2
#| echo: false

or_dr <- odds_ratios[2]
pct_change <- if(or_dr < 1) round((1 - or_dr) * 100, 1) else round((or_dr - 1) * 100, 1)
```

-   Since the odds ratio is `r if(or_dr < 1) "LESS than 1" else "GREATER than 1"`, higher discharge readiness is associated with `r if(or_dr < 1) "LOWER" else "HIGHER"` odds of readmission.

-   For each one-unit increase in discharge readiness score, the odds of readmission `r if(or_dr < 1) "DECREASE" else "INCREASE"` by `r pct_change`%. (or alternatively, the odds of readmission are multiplied by `r round(or_dr, 3)` for each one-unit increase in discharge readiness).

## 1.7 Predicting Probabilities

One of the most useful applications of logistic regression is predicting the probability of the outcome for specific values of the predictor. Let's calculate predicted probabilities step by step.

### 1.7.1 Manual Calculation

Let's predict the probability of readmission for ***a new patient with a discharge readiness score of 5***:

```{r}
#| label: simple-predict-manual
#| echo: false

# Get coefficients
b0_pred <- coef(model_simple)[1]
b1_pred <- coef(model_simple)[2]

# For discharge_readiness = 5
dr_value <- 5

# Step 1: Calculate log odds
log_odds_pred <- b0_pred + b1_pred * dr_value

# Step 2: Convert to odds
odds_pred <- exp(log_odds_pred)

# Step 3: Convert to probability
prob_pred <- odds_pred / (1 + odds_pred)
```

**Step 1: Calculate Predicted Log Odds**

First, we use the logistic regression equation to calculate the log odds. We plug the discharge readiness score into our model equation with the estimated coefficients:

Log odds = `r round(b0_pred, 3)` + (`r round(b1_pred, 3)`) × `r dr_value` = `r round(log_odds_pred, 4)`

This log odds value represents the predicted log odds of readmission on the logit scale. To interpret this meaningfully, we need to convert it to odds and then to probability.

**Step 2: Convert Log Odds to Odds**

Next, we convert the log odds to odds by exponentiating (taking e to the power of the log odds):

$$\text{Odds} = e^{`r round(log_odds_pred, 4)`} = `r round(odds_pred, 4)`$$

This odds value tells us the ratio of the probability of readmission to the probability of not being readmitted. An odds value of `r round(odds_pred, 4)` means that readmission is `r if(odds_pred < 1) "less likely than not being readmitted" else "more likely than not being readmitted"`.

**Step 3: Convert Odds to Probability**

Finally, we convert odds to probability using the formula: Probability = Odds / (1 + Odds)

$$\text{Probability} = \frac{`r round(odds_pred, 4)`}{1 + `r round(odds_pred, 4)`} = `r round(prob_pred, 4)`$$

A patient with a discharge readiness score of `r dr_value` has a `r round(prob_pred * 100, 1)`% predicted probability of being readmitted within 30 days. This means that if we had 100 patients with this discharge readiness level, we would expect approximately `r round(prob_pred * 100)` of them to be readmitted.

### 1.7.2 Using R's predict() Function

While the manual calculation helps us understand the underlying process, in practice we use R's `predict()` function to efficiently calculate predicted probabilities for many observations at once. This function automatically performs all three steps (log odds → odds → probability) for us.

The `predict()` function in R generates predictions from a fitted model. Key arguments are:

-   First argument: The fitted model object
-   `newdata`: A data frame containing the predictor values for which you want predictions
-   `type = "response"`: Returns predicted probabilities (for logistic regression). Without this, it returns log odds.

Let's predict readmission probabilities for patients with different discharge readiness levels:

**Predicted Probabilities of Readmission:**

```{r}
#| label: simple-predict-r
#| code-fold: true

# Create data frame with values to predict (5 new patients)
new_patients <- data.frame(
  discharge_readiness = c(4, 5, 6, 7, 8)
)

new_patients$predicted_prob <- predict(model_simple,
                                        newdata = new_patients,
                                        type = "response")

kable(round(new_patients, 3),
      col.names = c("Discharge Readiness Score", "Predicted Probability"),
      align = c("c", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

As discharge readiness scores increase from 4 to 8, the predicted probability of readmission systematically decreases. This quantifies the relationship we observed in our earlier boxplot: higher discharge readiness is associated with lower readmission risk. We can use these probabilities to identify which patients may need additional discharge support.

### 1.7.3 Visualizing the Predicted Probabilities

Visualizing the relationship between our predictor and the predicted probability helps us understand the model and communicate findings to stakeholders. Let's create a graph showing how the predicted probability of readmission changes across the full range of discharge readiness scores:

```{r}
#| label: simple-predict-viz
#| fig-width: 6
#| fig-height: 4
#| code-fold: true

# Create prediction data across range of discharge readiness
pred_data <- data.frame(
  discharge_readiness = seq(3, 10, by = 0.1)
)
pred_data$probability <- predict(model_simple, newdata = pred_data, type = "response")

# Plot
ggplot() +
  # Add prediction curve
  geom_line(data = pred_data, aes(x = discharge_readiness, y = probability),
            color = "steelblue", linewidth = 1.2) +
  # Add actual data points
  geom_point(data = readmit_data, aes(x = discharge_readiness, y = readmitted),
             alpha = 0.3, position = position_jitter(width = 0.1, height = 0)) +
  # Add reference line at 0.5
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  labs(title = "Predicted Probability of Readmission by Discharge Readiness Score",
       subtitle = "Points show actual patient outcomes (0 = not readmitted, 1 = readmitted)",
       x = "Discharge Readiness Score",
       y = "Predicted Probability of Readmission") +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

**Understanding this visualization:**

The **blue curve** represents our model's predicted probabilities across all discharge readiness levels. The **black points** show the actual patient outcomes in our dataset.

Key features of the logistic curve:

-   **At low discharge readiness scores** (left side): The probability of readmission is relatively high.
-   **At high discharge readiness scores** (right side): The probability of readmission is relatively low.

The **dashed line at 0.5** (50%) represents the traditional classification threshold used in many statistical applications. However, **in practice, we would likely use a much lower threshold for readmission prediction** (e.g., 10-20%). Why? Because:

-   Actual readmission rates are typically around 15%, not 50%
-   The cost of missing a high-risk patient (false negative) is high
-   The cost of providing extra discharge support to a lower-risk patient (false positive) is relatively low

For example, if we used a 15% threshold, any patient with predicted probability ≥ 15 % would receive enhanced discharge planning. This more conservative approach helps ensure we don't miss high-risk patients who could benefit from intervention.

## 1.8 Evaluating Model Performance

After fitting a logistic regression model, we want to assess its practical usefulness. In other words: **How well does this model actually predict outcomes?** While our model provides probabilities, in many healthcare applications we need to make binary decisions (e.g., "provide extra discharge support" vs. "standard discharge"). This requires evaluating how accurately our model classifies patients.

### 1.8.1 Creating a Classification Table

To evaluate classification performance, we first convert predicted probabilities into predicted classes using a **decision threshold**.

**For this example, we'll use the traditional 0.5 (50%) threshold** to demonstrate how classification tables work:

-   If predicted probability ≥ 0.5 → predict "readmitted" (class = 1)
-   If predicted probability \< 0.5 → predict "not readmitted" (class = 0)

While 0.5 is the standard default, in practice a lower threshold (e.g., 0.15-0.20) may be more appropriate for readmission prediction, since readmission rates are typically well below 50%.

We then create a **classification table** (also called a **confusion matrix**) that compares our predictions to actual outcomes:

```{r}
#| label: simple-classification
#| echo: false

# Add predictions to our data
readmit_data$predicted_prob <- predict(model_simple, type = "response")
readmit_data$predicted_class <- ifelse(readmit_data$predicted_prob >= 0.5, 1, 0)

# Create classification table with explicit factor levels to ensure 2x2 table
class_table <- table(
  Predicted = factor(readmit_data$predicted_class, levels = c(0, 1)),
  Actual = factor(readmit_data$readmitted, levels = c(0, 1))
)
```

**Classification Table (Confusion Matrix):**

```{r}
#| echo: false
cm_df <- as.data.frame.matrix(class_table)
cm_df <- cbind(`Predicted Outcome` = rownames(cm_df), cm_df)
rownames(cm_df) <- NULL

kable(cm_df, align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE,
                position = "left") %>%
  add_header_above(c(" " = 1, "Actual Outcome" = 2))
```

### 1.8.2 Interpreting the Classification Table

This 2×2 table shows four possible outcomes:

-   **True Negatives (TN, top-left)**: Patients we correctly predicted would NOT be readmitted
-   **False Negatives (FN, top-right)**: Patients we incorrectly predicted would NOT be readmitted (missed case)
-   **False Positives (FP, bottom-left)**: Patients we incorrectly predicted would be readmitted (false alarm)
-   **True Positives (TP, bottom-right)**: Patients we correctly predicted would be readmitted

### 1.8.3 Calculating Model Accuracy

**Accuracy** measures the overall proportion of correct predictions (both true positives and true negatives) out of all predictions:

$$\text{Accuracy} = \frac{\text{True Positives} + \text{True Negatives}}{\text{Total Number of Observations}} = \frac{TP + TN}{TP + TN + FP + FN}$$

Let's calculate our model's accuracy:

```{r}
#| label: simple-accuracy
#| echo: false

# Calculate metrics
correct <- sum(diag(class_table))
total <- sum(class_table)
accuracy <- correct / total

# Break down by actual class
tn <- class_table[1, 1]  # True negatives
fp <- class_table[2, 1]  # False positives
fn <- class_table[1, 2]  # False negatives
tp <- class_table[2, 2]  # True positives
```

**Model Performance:**

-   Correctly classified: `r correct` out of `r total` patients
-   Overall accuracy: `r round(accuracy * 100, 1)`%

**Breakdown:**

-   Correctly predicted NOT readmitted (True Negatives): `r tn`
-   Correctly predicted readmitted (True Positives): `r tp`
-   Incorrectly predicted readmitted (False Positives - false alarm): `r fp`
-   Incorrectly predicted NOT readmitted (False Negatives - missed): `r fn`

**What does this mean?**

Our model correctly classified `r round(accuracy * 100, 1)`% of patients. While accuracy is a useful starting point, in healthcare applications we often care more about specific types of errors. For example:

-   **False negatives** (missed readmissions) mean high-risk patients don't receive intervention
-   **False positives** (false alarms) mean resources are spent on patients who don't need extra support

The relative importance of these errors depends on the costs and benefits of intervention in your specific context.

## 1.9 Practice: Simple Logistic Regression {.unnumbered}

::: callout-note
Complete the activity below. This activity will help you practice:

-   Running simple logistic regression models
-   Interpreting log odds ratios and odds ratios
-   Making predictions from the model
:::

**Practice Activity: Medication Adherence**

**Scenario**: A pharmacy wants to understand whether the number of medications a patient takes predicts their adherence (whether they take medications as prescribed) to their medication regimen. Adherence is measured as a binary variable: 1 = adherent (taking medications correctly), 0 = non-adherent (not taking medications as prescribed).

Medication non-adherence is a major challenge in managing chronic diseases (long-term health conditions). Studies have shown that patients taking multiple medications often have lower adherence rates due to increased complexity, higher costs, greater risk of side effects, and difficulties managing multiple medication schedules.

**Dataset Loading**

Run this code to load the medication adherence dataset:

```{r}
#| label: activity1-data

# Medication adherence data
adherence_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week13_adherence.data.csv")

# View the data
head(adherence_data, 5)
```

**Task 1: Explore the Data**

Before modeling, let's explore the relationship between number of medications and adherence. **Fill in the blanks** to create a summary table:

```{r}
#| label: activity1-explore
#| eval: false

# Calculate adherence rate by number of medications (grouped)
adherence_summary <- adherence_data %>%
  mutate(med_group = cut(num_medications,
                         breaks = c(0, 4, 8, 12),
                         labels = c("1-4 meds", "5-8 meds", "9-12 meds"))) %>%
  group_by(___) %>%  # Group by medication category
  summarize(
    n_patients = n(),
    adherence_rate = round(mean(___) * 100, 1)  # Calculate % adherent
  )

adherence_summary
```

**Question**: What pattern do you see? Does adherence seem related to number of medications?

**Task 2: Run Simple Logistic Regression**

**Fill in the blanks** to run a logistic regression predicting adherence from number of medications:

```{r}
#| label: activity1-model
#| eval: false

# Fit logistic regression model
model_adhere <- glm(___ ~ ___,  # outcome ~ predictor
                    data = ___,
                    family = ___)  # Don't forget this!

# View the results
summary(model_adhere)
```

**Task 3: Interpret the Log Odds Ratio**

From your `summary()` output, find the row for `num_medications` and answer the following:

1.  What is the **coefficient** (log odds ratio) for `num_medications`?
2.  What is the **p-value**? Is it statistically significant (p \< 0.05)?
3.  What is the **direction** of the effect? (Does taking more medications increase or decrease the odds of adherence?)

**Task 4: Calculate and Interpret Odds Ratio**

**Fill in the blanks** to calculate the odds ratio:

```{r}
#| label: activity1-or
#| eval: false

# Calculate odds ratio
or_medications <- ___(coef(model_adhere)["num_medications"])
round(or_medications, 2)

```

**Questions**:

1.  For each additional medication, how much do the odds of adherence change (in %)?
2.  What does this mean for patient management?

**Task 5: Make Predictions**

Suppose three new patients arrive at your clinic. Patient A takes 2 medications, Patient B takes 6, and Patient C takes 10. Using your model, predict each patient's probability of medication adherence.

**Fill in the blanks** to generate these predictions:

```{r}
#| label: activity1-predict
#| eval: false

# Create data for prediction
new_patients <- data.frame(
  num_medications = c(2, 6, 10)  # Low, medium, high medication counts
)

# Predict probabilities
new_patients$predicted_prob <- predict(___,
                                       newdata = ___,
                                       type = ___)  # Don't forget this!

# Display results
new_patients
```

**Question**: Suppose your clinic wants to provide timely medication reminders (e.g., text messages or phone calls) to patients at risk of non-adherence. If you offer this intervention to patients whose predicted adherence probability falls below 0.5, which of the three patients (A, B, C) would receive the intervention?

------------------------------------------------------------------------

# 2. Multiple Logistic Regression

In practice, outcomes like hospital readmission are influenced by multiple factors simultaneously. **Multiple logistic regression** allows us to include several predictors in a single model.

## 2.1 What is it?

Multiple logistic regression extends simple logistic regression by including **two or more predictor variables** to predict a binary outcome. The model estimates the independent effect of each predictor while holding all other predictors constant, just like multiple linear regression, but for binary outcomes.

## 2.2 When to Use It

Use multiple logistic regression when:

-   Your **outcome variable is binary** (0/1)
-   You have **two or more predictor variables** (continuous, categorical, or both)
-   You want to **control for confounding variables**
-   You want to understand the **independent effect** of each predictor

Multiple logistic regression enables us to:

1.  **Control for confounding variables** (factors that might distort the relationship we're studying)
2.  **Make more accurate predictions** by incorporating more information

## 2.3 Example Case: Comprehensive Readmission Risk Model

In the simple regression section, we used only discharge readiness to predict readmission. But what if that effect was partly due to **confounding**? For example, younger, healthier patients may feel more ready for discharge AND are naturally less likely to be readmitted. Without controlling for age and comorbidities, we might be overstating the true effect of discharge readiness.

Our `readmit_data` dataset (n = `r nrow(readmit_data)`) already includes additional variables we can use to build a more comprehensive model:

-   `discharge_readiness`: Discharge readiness score (1-10 scale)
-   `age`: Patient age in years
-   `comorbidities`: Number of chronic health conditions (e.g., diabetes, hypertension, heart disease)
-   `insurance`: Insurance type (Private, Medicare, or Medicaid) — a **categorical predictor**
-   `readmitted`: Whether the patient was readmitted within 30 days (1 = yes, 0 = no)

Let's test whether discharge readiness still predicts readmission when we account for age, health complexity, and insurance type.

## 2.4 Running Multiple Logistic Regression

We understand our data, so we're ready to fit the multiple logistic regression model. The good news is that the syntax is very similar to simple logistic regression. We just add more predictors to the formula using the `+` sign.

Remember from simple regression, we had: `readmitted ~ discharge_readiness`

Now with multiple predictors, we expand this to: `readmitted ~ discharge_readiness + age + comorbidities + insurance`

Each predictor is added to the right side of the formula, separated by plus signs. Let's fit the model:

```{r}
#| label: multiple-model

# Fit multiple logistic regression model
model_multiple <- glm(readmitted ~ discharge_readiness + age + comorbidities + insurance,
                      data = readmit_data,
                      family = binomial(link = "logit"))

# View the results
summary(model_multiple)
```

## 2.5 Interpreting Multiple Predictors

The output looks similar to simple logistic regression, but now we have coefficients for each predictor. The key difference in interpretation is that **each coefficient represents the effect of that variable while holding all other variables constant**. This is crucial for proper interpretation.

For example, the coefficient for discharge readiness tells us, "How does readmission risk change when discharge readiness increases by 1 point, **assuming age, comorbidities, and insurance type stay the same**?" This allows us to isolate the independent effect of each predictor.

Let's create a summary table to see all coefficients, odds ratios, and p-values together:

### 2.5.1 Interpreting Coefficients

```{r}
#| label: multiple-coef
#| echo: false

# Create a nice summary table
coef_summary <- data.frame(
  Variable = names(coef(model_multiple)),
  `Log Odds Ratio` = round(coef(model_multiple), 4),
  `Odds Ratio` = round(exp(coef(model_multiple)), 4),
  `p-value` = round(summary(model_multiple)$coefficients[, 4], 4)
)

kable(coef_summary, row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Let's interpret each predictor:

```{r}
#| label: multiple-interpret
#| echo: false

coefs <- coef(model_multiple)
ors <- exp(coefs)
pvals <- summary(model_multiple)$coefficients[, 4]

# Calculate percentage changes
pct_readiness <- if(ors["discharge_readiness"] < 1) round((1 - ors["discharge_readiness"]) * 100, 1) else round((ors["discharge_readiness"] - 1) * 100, 1)
pct_age <- round((ors["age"] - 1) * 100, 1)
pct_comorbid <- round((ors["comorbidities"] - 1) * 100, 1)
```

**Interpretation of Each Predictor:**

**1. Discharge Readiness**

-   **Log Odds Ratio:** `r round(coefs["discharge_readiness"], 4)` (p = `r format(round(pvals["discharge_readiness"], 4), scientific = FALSE)`)
    -   This means that for each one-unit increase in discharge readiness, the log odds of readmission change by `r round(coefs["discharge_readiness"], 4)`, holding other variables constant.
-   **Odds Ratio:** `r round(ors["discharge_readiness"], 3)`
    -   This means that for each one-unit increase in discharge readiness, the odds of readmission are multiplied by `r round(ors["discharge_readiness"], 3)`, holding other variables constant. (i.e., the odds `r if(ors["discharge_readiness"] < 1) "DECREASE" else "INCREASE"` by `r pct_readiness`%)

**2. Age**

-   **Log Odds Ratio:** `r round(coefs["age"], 4)` (p = `r format(round(pvals["age"], 4), scientific = FALSE)`)
    -   For each one-year increase in age, the log odds of readmission change by `r round(coefs["age"], 4)`, holding other variables constant.
-   **Odds Ratio:** `r round(ors["age"], 4)`
    -   For each one-year increase in age, the odds of readmission are multiplied by `r round(ors["age"], 4)`, holding other variables constant. (i.e., the odds `r if(ors["age"] < 1) "decrease" else "increase"` by `r pct_age`%)

**3. Comorbidities**

-   **Log Odds Ratio:** `r round(coefs["comorbidities"], 4)` (p = `r format(round(pvals["comorbidities"], 4), scientific = FALSE)`)
    -   For each additional comorbidity, the log odds of readmission change by `r round(coefs["comorbidities"], 4)`, holding other variables constant.
-   **Odds Ratio:** `r round(ors["comorbidities"], 3)`
    -   For each additional comorbidity, the odds of readmission are multiplied by `r round(ors["comorbidities"], 3)`, holding other variables constant. (i.e., the odds `r if(ors["comorbidities"] < 1) "decrease" else "increase"` by `r pct_comorbid`%)

**4. Insurance Type** (Reference category: Medicaid)

-   **Medicare vs. Medicaid:**
    -   **Log Odds Ratio:** `r round(coefs["insuranceMedicare"], 4)` (p = `r format(round(pvals["insuranceMedicare"], 4), scientific = FALSE)`)
    -   **Odds Ratio:** `r round(ors["insuranceMedicare"], 3)`
    -   **Interpretation:** Holding discharge readiness, age, and comorbidities constant, patients with Medicare have odds of readmission that are `r round(ors["insuranceMedicare"], 3)` times the odds for patients with Medicaid. `r if(ors["insuranceMedicare"] < 1) paste0("In other words, Medicare patients have ", round((1 - ors["insuranceMedicare"]) * 100, 1), "% lower odds compared to Medicaid patients.") else paste0("In other words, Medicare patients have ", round((ors["insuranceMedicare"] - 1) * 100, 1), "% higher odds compared to Medicaid patients.")`
-   **Private vs. Medicaid:**
    -   **Log Odds Ratio:** `r round(coefs["insurancePrivate"], 4)` (p = `r format(round(pvals["insurancePrivate"], 4), scientific = FALSE)`)
    -   **Odds Ratio:** `r round(ors["insurancePrivate"], 3)`
    -   **Interpretation:** Holding discharge readiness, age, and comorbidities constant, patients with Private insurance have odds of readmission that are `r round(ors["insurancePrivate"], 3)` times the odds for patients with Medicaid. `r if(ors["insurancePrivate"] < 1) paste0("In other words, Private insurance patients have ", round((1 - ors["insurancePrivate"]) * 100, 1), "% lower odds compared to Medicaid patients.") else paste0("In other words, Private insurance patients have ", round((ors["insurancePrivate"] - 1) * 100, 1), "% higher odds compared to Medicaid patients.")`

## 2.6 Making Predictions with Multiple Predictors

With multiple logistic regression, we can make more nuanced predictions by considering a patient's complete risk profile, not just one. Instead of just knowing their discharge readiness score, we now incorporate their age, health complexity, and insurance status to get a more accurate prediction.

### 2.6.1 Manual Calculation for Multiple Predictors

To understand how multiple predictors combine, let's walk through a manual calculation for a specific patient. This shows the mathematical process, though in practice R will do this for us automatically.

The multiple logistic regression equation extends the simple version by adding more terms:

$$\text{Log Odds} = \beta_0 + \beta_1 \times \text{Discharge Readiness} + \beta_2 \times \text{Age} + \beta_3 \times \text{Comorbidities}$$ $$+ \beta_4 \times \text{Medicare} + \beta_5 \times \text{Private}$$

**Example Patient:**

Let's assume that we have a new patient whose data is as follows, and we would like to predict this patient's predicted probability of readmission using our multiple logistic regression model.

-   Discharge Readiness = 6
-   Age = 70
-   Comorbidities = 3
-   Insurance = Medicare

```{r}
#| label: multiple-manual-predict
#| echo: false

# Get coefficients
b0_m <- coef(model_multiple)[1]
b1_dr <- coef(model_multiple)["discharge_readiness"]
b2_age <- coef(model_multiple)["age"]
b3_comor <- coef(model_multiple)["comorbidities"]
b4_medicare <- coef(model_multiple)["insuranceMedicare"]
b5_private <- coef(model_multiple)["insurancePrivate"]

# Patient values
dr_val <- 6
age_val <- 70
comor_val <- 3
# Medicare dummy = 1, Otherwise = 0
medicare_val <- 1
# Private dummy = 1, Otherwise = 0
private_val <- 0

# Calculate log odds
log_odds_m <- b0_m + b1_dr * dr_val + b2_age * age_val + b3_comor * comor_val + b4_medicare * medicare_val + b5_private * private_val

# Convert to odds
odds_m <- exp(log_odds_m)

# Convert to probability
prob_m <- odds_m / (1 + odds_m)
```

**Step 1: Calculate Log Odds**

Using the multiple logistic regression equation, we plug in all predictor values:

Log odds = `r round(b0_m, 3)` + (`r round(b1_dr, 3)`) × 6 + (`r round(b2_age, 4)`) × 70 + (`r round(b3_comor, 3)`) × 3 + (`r round(b4_medicare, 3)`) × 1 + (`r round(b5_private, 3)`) × 0

Log odds = `r round(log_odds_m, 4)`

Note: For Medicare insurance, we use 1 for the Medicare dummy variable and 0 for the Private dummy variable (since the patient doesn't have private insurance).

**Step 2: Convert to Odds**

We can get the predicted odds by taking the exponent of the log odds.

$$\text{Odds} = e^{`r round(log_odds_m, 4)`} = `r round(odds_m, 4)`$$

**Step 3: Convert to Probability**

We can get the predicted probability by using the equation we learned above.

$$\text{Probability} = \frac{`r round(odds_m, 4)`}{1 + `r round(odds_m, 4)`} = `r round(prob_m, 4)`$$

This patient has a `r round(prob_m * 100, 1)`% predicted probability of readmission within 30 days.

### 2.6.2 Predictions for Multiple Patient Profiles

While manual calculations help us understand the mechanics, in practice we use R to generate predictions efficiently. One powerful use of multiple regression is creating **risk profiles**: comparing predicted probabilities for patients with different combinations of risk factors.

For example, we can contrast a "Low Risk" patient (young, few comorbidities, high discharge readiness, private insurance) with a "High Risk" patient (older, many comorbidities, low discharge readiness, Medicaid). Let's create three representative profiles and see how their predicted probabilities differ:

```{r}
#| label: multiple-predict

# Create patient profiles
patient_profiles <- data.frame(
  discharge_readiness = c(8, 6, 4),
  age = c(50, 65, 80),
  comorbidities = c(1, 2, 4),
  insurance = factor(c("Private", "Medicare", "Medicaid"))
)

# Predict probabilities
patient_profiles$predicted_prob <- predict(model_multiple,
                                           newdata = patient_profiles,
                                           type = "response")
patient_profiles
```

Notice how combining multiple risk factors leads to dramatically different predicted probabilities. The third patient (low discharge readiness, age 80, 4 comorbidities, Medicaid) has a much higher predicted probability than the first patient (high readiness, age 50, 1 comorbidity, Private insurance). This demonstrates the power of multiple regression — it captures the cumulative effect of multiple risk factors.

## 2.7 Evaluating the Multiple Regression Model

Finally, let's assess how well our multiple regression model performs at classifying patients. Just as we did with simple regression, we'll create a confusion matrix to see how many patients are correctly classified as readmitted or not readmitted.

As before, we'll use the 0.5 (50%) threshold:

```{r}
#| label: multiple-evaluation
#| echo: true

# Add predictions
readmit_data$pred_prob_multiple <- predict(model_multiple, type = "response")
readmit_data$pred_class_multiple <- ifelse(readmit_data$pred_prob_multiple >= 0.5, 1, 0)
```

**Classification Table (Confusion Matrix):**

```{r}
#| echo: false
# Classification table with explicit factor levels
class_table_multiple <- table(
  Predicted = factor(readmit_data$pred_class_multiple, levels = c(0, 1)),
  Actual = factor(readmit_data$readmitted, levels = c(0, 1))
)

cm_df_m <- as.data.frame.matrix(class_table_multiple)
cm_df_m <- cbind(`Predicted Outcome` = rownames(cm_df_m), cm_df_m)
rownames(cm_df_m) <- NULL

kable(cm_df_m, align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE,
                position = "left") %>%
  add_header_above(c(" " = 1, "Actual Outcome" = 2))
```

```{r}
#| label: multiple-accuracy
#| echo: false

# Calculate metrics
correct_m <- sum(diag(class_table_multiple))
total_m <- sum(class_table_multiple)
accuracy_m <- correct_m / total_m

# Break down by actual class
tn_m <- class_table_multiple[1, 1]  # True negatives
fp_m <- class_table_multiple[2, 1]  # False positives
fn_m <- class_table_multiple[1, 2]  # False negatives
tp_m <- class_table_multiple[2, 2]  # True positives
```

**Model Performance:**

-   Correctly classified: `r correct_m` out of `r total_m` patients
-   Overall accuracy: `r round(accuracy_m * 100, 1)`%

**Breakdown:**

-   Correctly predicted NOT readmitted (True Negatives): `r tn_m`
-   Correctly predicted readmitted (True Positives): `r tp_m`
-   Incorrectly predicted readmitted (False Positives): `r fp_m`
-   Incorrectly predicted NOT readmitted (False Negatives): `r fn_m`

The multiple logistic regression model achieves `r round(accuracy_m * 100, 1)`% accuracy. This demonstrates the value of incorporating multiple risk factors in predicting readmission.

## 2.8 Practice: Multiple Logistic Regression {.unnumbered}

::: callout-note
Complete the activity below. This activity will help you practice:

-   Running multiple logistic regression with several predictors
-   Interpreting odds ratios while controlling for confounders
-   Comparing effects of different predictors
-   Making predictions with multiple risk factors
:::

**Practice Activity: Surgical Complication Risk**

**Scenario**: A surgical department wants to predict the risk of post-operative complications (problems that occur after surgery). They have collected data on patient age (`age`), BMI (`bmi` - body mass index, a measure of body weight relative to height), surgery duration in minutes (`surgery_duration`), and whether the surgery was elective (planned) or emergency (`surgery_type`).

Post-operative complications represent a significant concern in surgical care, with factors such as patient characteristics, procedure complexity, and surgical urgency all potentially contributing to complication risk. Understanding which factors predict complications can help with risk stratification and targeted prevention strategies.

**Dataset Loading**

Load the dataset using the code below:

```{r}
#| label: activity3-data

# Surgical complication data
surgery_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week13_surgery_data.csv")
surgery_data$surgery_type <- as.factor(surgery_data$surgery_type)

# View the data
head(surgery_data, 5)
```

**Task 1: Explore the Data**

**Fill in the blanks** to summarize complication rates by surgery type:

```{r}
#| label: activity3-explore
#| eval: false

# Summary by surgery type
surgery_summary <- surgery_data %>%
  group_by(___) %>%
  summarize(
    n = n(),
    complication_rate = round(mean(___) * 100, 1),
    avg_age = round(mean(___), 1),
    avg_bmi = round(mean(___), 1)
  )

surgery_summary
```

**Question**: What initial patterns do you notice? Do emergency surgeries appear to have different characteristics?

**Task 2: Run Multiple Logistic Regression**

**Fill in the blanks** to fit a model with all four predictors:

```{r}
#| label: activity3-model
#| eval: false

# Fit multiple logistic regression
model_surg <- glm(complication ~ ___ + ___ + ___ + ___,  # All 4 predictors
                  data = ___,
                  family = ___)

# View results
summary(___)
```

**Task 3: Interpret the Results**

From your `summary()` output, answer the following:

1.  Which predictors are statistically significant (p \< 0.05)?
2.  For each significant predictor, what is the direction of the effect (increases or decreases complication risk)?

**Task 4: Calculate and Interpret Odds Ratios**

**Fill in the blank** to calculate the odds ratios for all predictors:

```{r}
#| label: activity3-or
#| eval: false

# Calculate odds ratios from coefficients
___(coef(model_surg))
```

Using the output, answer the following:

1.  What is the odds ratio for `age`? For each additional year of age, how do the odds of complication change (in %)?
2.  What does the odds ratio for `surgery_typeEmergency` tell you about emergency vs. elective surgeries?

**Task 5: Make Predictions for Patient Profiles**

Assume you have three new patients scheduled for surgery:

-   **Patient A**: 45 years old, BMI of 24, scheduled for a 90-minute **elective** procedure
-   **Patient B**: 65 years old, BMI of 30, scheduled for a 150-minute **elective** procedure
-   **Patient C**: 80 years old, BMI of 38, undergoing a 210-minute **emergency** procedure

**Fill in the blanks** to predict each patient's complication risk:

```{r}
#| label: activity3-predict
#| eval: false

# Create patient profiles
patient_profiles <- data.frame(
  age = c(45, 65, 80),
  bmi = c(24, 30, 38),
  surgery_duration = c(90, 150, 210),
  surgery_type = c("Elective", "Elective", "Emergency")
)

# Predict probabilities
patient_profiles$predicted_prob <- predict(___,
                                          newdata = ___,
                                          type = ___)

# Display
patient_profiles
```

**Question**: Which patient(s) have the highest predicted complication risk? What combination of factors seems to drive the difference?

------------------------------------------------------------------------

# Summary

## Key Takeaways

### When to Use Logistic Regression

-   Use logistic regression when your **outcome variable is binary** (two possible values)
-   Use `glm(outcome ~ predictors, data = data, family = binomial)` in R
-   The model predicts the **probability** of the event occurring

### Interpreting Coefficients

Ask three questions for each coefficient:

1.  **Direction**: Is the log odds ratio positive or negative? (Positive = increases odds; Negative = decreases odds)
2.  **Magnitude**: What is the odds ratio? How much do the odds change per unit increase in the predictor?
3.  **Significance**: Is p \< 0.05? (But practical significance matters too!)

### Log Odds Ratios vs. Odds Ratios

| Measure        | No Effect | Decreases Risk | Increases Risk |
|----------------|-----------|----------------|----------------|
| Log Odds Ratio | = 0       | \< 0           | \> 0           |
| Odds Ratio     | = 1       | \< 1           | \> 1           |

### Converting Between Measures

**Probability → Odds → Log Odds**

$$\text{Odds} = \frac{p}{1-p}$$

$$\text{Log Odds} = \ln(\text{Odds})$$

**Log Odds → Odds → Probability**

$$\text{Odds} = e^{\text{Log Odds}}$$

$$\text{Probability} = \frac{\text{Odds}}{1 + \text{Odds}}$$

### Simple vs. Multiple Logistic Regression

| Aspect         | Simple              | Multiple                               |
|----------------|---------------------|----------------------------------------|
| Predictors     | One                 | Two or more                            |
| Interpretation | Direct effect       | Effect controlling for other variables |
| Use case       | Initial exploration | Comprehensive modeling                 |
| Confounding    | Not addressed       | Can be controlled                      |

### Model Evaluation

-   **Accuracy** measures overall correct predictions but can be misleading with imbalanced outcomes
-   Use the **confusion matrix** (classification table) to see where the model makes errors (false positives vs. false negatives)
-   In healthcare, consider which type of error is more costly for your specific application

------------------------------------------------------------------------

# References and Further Reading

## Statistical Methods

-   Hosmer, D. W., Lemeshow, S., & Sturdivant, R. X. (2013). *Applied Logistic Regression* (3rd ed.). Wiley.
-   Field, A. (2018). *Discovering Statistics Using IBM SPSS Statistics* (5th ed.). SAGE Publications.
-   James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning* (2nd ed.). Springer.

## Healthcare Applications

Examples are motivated by the following:

-   Weiss, M. E., Piacentine, L. B., Lokken, L., Ancona, J., Archer, J., Ber, S., ... & Vega-Stromberg, T. (2007). Perceived readiness for hospital discharge in adult medical-surgical patients. *Clinical Nurse Specialist*, 21(1), 31-42. https://doi.org/10.1097/00002800-200701000-00006

-   Wang, X., Juntasopeepun, P., Panuthai, S., & Suwankruhasn, N. (2025). Factors influencing medication adherence among Chinese older adults with physical multimorbidity and polypharmacy: A cross-sectional study. *Nursing & Health Sciences*, 27(3), e70192. https://doi.org/10.1111/nhs.70192

-   Sheikh, S., Booth-Norse, A., Smotherman, C., Kalynych, C., Lukens-Bull, K., Guerrido, E., Henson, M., Gautam, S., & Hendry, P. (2020). Predicting pain-related 30-day emergency department return visits in middle-aged and older adults. *Pain Medicine*, 21(11), 2748-2756. https://pmc.ncbi.nlm.nih.gov/articles/PMC8557807/

-   Dharap, S. B., Barbaniya, P., & Navgale, S. (2022). Incidence and risk factors of postoperative complications in general surgery patients. *Cureus*, 14(11), e30975. https://doi.org/10.7759/cureus.30975
