---
title: "Logistic Regression (Practice Activities)"
subtitle: "Business Analytics in Healthcare Management"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
    theme: cosmo
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)

# Set Seed number for reproducibility
set.seed(2026)
```

# Overview

In this practice session, you will apply the logistic regression techniques learned in the lecture.
This is a learning activity designed to help you gain a better understanding of how to predict binary outcomes using the `glm()` function in R.

**What you'll practice:**

- Running simple logistic regression models
- Interpreting log odds ratios and odds ratios
- Making predictions from logistic regression
- Running multiple logistic regression with several predictors
- Evaluating classification accuracy
- Understanding the effect of controlling for confounders

**Format**: Fill-in-the-blank code sections (marked with `___`)

**Instructions**:

1. Complete the code by replacing `___` with appropriate values
2. Run each chunk to verify your results
3. Answer the interpretation questions
4. Compare your results with solutions at the end

**Note**: This practice is completion-based. The goal is to learn and experiment—mistakes are part of the process!

---

# Recap: Key Concepts

Before starting the activities, let's review essential concepts from the lecture:

## The Three Scales

Logistic regression works with three related measures:

- **Probability (p)**: Ranges from 0 to 1
- **Odds**: Ranges from 0 to ∞, calculated as $\frac{p}{1-p}$
- **Log Odds (Logit)**: Ranges from −∞ to +∞, calculated as $\ln(\text{Odds})$

## Converting Between Scales

```{r}
#| label: example-conversion
#| eval: false

# Probability → Odds → Log Odds
probability <- 0.3
odds <- probability / (1 - probability)
log_odds <- log(odds)

# Log Odds → Odds → Probability
log_odds <- -0.5
odds <- exp(log_odds)
probability <- odds / (1 + odds)
```

## Running Logistic Regression in R

**Simple logistic regression** (one predictor):

```{r}
#| label: example-simple
#| eval: false

model <- glm(outcome ~ predictor,
             data = your_data,
             family = binomial(link = "logit"))

summary(model)  # View results
```

**Multiple logistic regression** (multiple predictors):

```{r}
#| label: example-multiple
#| eval: false

model <- glm(outcome ~ predictor1 + predictor2 + predictor3,
             data = your_data,
             family = binomial(link = "logit"))

summary(model)  # View results
```

## Interpreting Results

**Coefficients** (log odds ratios):
- **Negative** coefficient → predictor decreases log odds of outcome
- **Positive** coefficient → predictor increases log odds of outcome

**Odds Ratios** (exponentiated coefficients):

```{r}
#| label: example-or
#| eval: false

# Calculate odds ratios
odds_ratios <- exp(coef(model))
```

- **Odds ratio < 1** → predictor decreases odds of outcome
- **Odds ratio = 1** → no effect
- **Odds ratio > 1** → predictor increases odds of outcome

**Percent change in odds**:

If odds ratio = 0.85, then odds decrease by (1 - 0.85) × 100 = 15%
If odds ratio = 1.20, then odds increase by (1.20 - 1) × 100 = 20%

## Making Predictions

```{r}
#| label: example-predict
#| eval: false

# Create new data
new_data <- data.frame(
  predictor1 = c(5, 7),
  predictor2 = c(60, 80)
)

# Predict probabilities
predicted_prob <- predict(model,
                         newdata = new_data,
                         type = "response")
```

**Important**: `type = "response"` returns probabilities. Without it, you get log odds.

## Evaluating Classification

```{r}
#| label: example-classification
#| eval: false

# Add predicted probabilities to data
data$predicted_prob <- predict(model, type = "response")

# Classify using 0.5 threshold
data$predicted_class <- ifelse(data$predicted_prob >= 0.5, 1, 0)

# Create confusion matrix
table(Predicted = data$predicted_class, Actual = data$outcome)

# Calculate accuracy
accuracy <- mean(data$predicted_class == data$outcome)
```

---

# Common Errors and How to Fix Them

You **WILL** encounter errors while practicing—that's completely normal and part of the learning process! Here are the most common errors specific to logistic regression and their solutions:

## Error 1: "non-numeric argument to binary operator"

```r
Error in predictor1 + predictor2: non-numeric argument to binary operator
```

**Cause**: You're trying to use a character/factor variable directly in arithmetic operations

**Fix**: Make sure categorical variables are properly coded as factors:

```{r}
#| eval: false
# Check variable types
str(your_data)

# Convert to factor if needed
your_data$category <- factor(your_data$category)
```

## Error 2: Forgetting `family = binomial`

```r
# ❌ Wrong - this runs linear regression:
model <- glm(outcome ~ predictor, data = data)

# ✅ Right - this runs logistic regression:
model <- glm(outcome ~ predictor, data = data, family = binomial)
```

**Fix**: Always include `family = binomial(link = "logit")` or just `family = binomial` for logistic regression.

## Error 3: Misinterpreting coefficients as odds ratios

**Cause**: The coefficients from `summary()` are **log odds ratios**, not odds ratios

**Fix**: Exponentiate to get odds ratios:

```{r}
#| eval: false
# ❌ Wrong - these are log odds ratios:
coef(model)

# ✅ Right - these are odds ratios:
exp(coef(model))
```

## Error 4: Forgetting `type = "response"` in predictions

```{r}
#| eval: false
# ❌ Wrong - returns log odds:
predict(model, newdata = new_data)

# ✅ Right - returns probabilities:
predict(model, newdata = new_data, type = "response")
```

**Fix**: Always use `type = "response"` to get predicted probabilities.

## Error 5: "contrasts can be applied only to factors with 2 or more levels"

**Cause**: Your categorical variable has only one level in your data

**Fix**: Check your data and make sure you have variation:

```{r}
#| eval: false
# Check levels
table(your_data$category)

# Make sure you have at least 2 levels with observations
```

## Error 6: Incorrect calculation of percent change

```{r}
#| eval: false
# Example: odds ratio = 0.75
or <- 0.75

# ❌ Wrong:
pct_change <- or * 100  # This gives 75, which is wrong

# ✅ Right (for OR < 1):
pct_decrease <- (1 - or) * 100  # This gives 25% decrease

# Example: odds ratio = 1.30
or <- 1.30

# ✅ Right (for OR > 1):
pct_increase <- (or - 1) * 100  # This gives 30% increase
```

## Still Stuck?

If you're still getting errors:

1. **Check spelling**: Variable names are case-sensitive
2. **Run setup chunk**: Make sure you've loaded all libraries
3. **Check data structure**: Use `str(your_data)` to verify variable types
4. **Read error messages carefully**: They often tell you exactly what's wrong
5. **Compare with lecture examples**: Look for similar code patterns

::: {.callout-tip}
## Pro Tip: Debugging Strategy

When you get an error:

1. **Don't panic!** Errors are learning opportunities
2. **Read the error message** - what line is it pointing to?
3. **Check the basics** - Did you include `family = binomial`? Did you use `type = "response"`?
4. **Simplify** - Try running just the `glm()` first before making predictions
5. **Check the lecture** - Find a similar example and compare your code
:::

---

# Activity 1: Medication Adherence (Simple Logistic Regression)

## Scenario

A pharmacy wants to understand whether the number of medications a patient takes predicts their adherence (whether they take medications as prescribed) to their medication regimen. Adherence is measured as a binary variable: 1 = adherent (taking medications correctly), 0 = non-adherent (not taking medications as prescribed).

Medication non-adherence is a major challenge in managing chronic diseases (long-term health conditions). Studies have shown that patients taking multiple medications often have lower adherence rates due to increased complexity, higher costs, greater risk of side effects, and difficulties managing multiple medication schedules.

## Dataset Generation

Run this code to generate the medication adherence dataset:

```{r}
#| label: activity1-data

# Medication adherence data
n_patients <- 100
num_medications <- round(runif(n_patients, 1, 12))
log_odds_adhere <- 1.5 - 0.2 * num_medications + rnorm(n_patients, 0, 0.5)
prob_adhere <- exp(log_odds_adhere) / (1 + exp(log_odds_adhere))
adherent <- rbinom(n_patients, 1, prob_adhere)

adherence_data <- data.frame(
  patient_id = 1:n_patients,
  num_medications = num_medications,
  adherent = adherent
)

# View the data
head(adherence_data, 10)
```

## Task 1.1: Explore the Data

Before modeling, let's explore the relationship between number of medications and adherence.

**Fill in the blanks** to create a summary table:

```{r}
#| label: activity1-explore
#| eval: false

# Calculate adherence rate by number of medications (grouped)
adherence_summary <- adherence_data %>%
  mutate(med_group = cut(num_medications,
                         breaks = c(0, 4, 8, 12),
                         labels = c("1-4 meds", "5-8 meds", "9-12 meds"))) %>%
  group_by(___) %>%  # Group by medication category
  summarize(
    n_patients = n(),
    adherence_rate = round(mean(___) * 100, 1)  # Calculate % adherent
  )

adherence_summary
```

**Question**: What pattern do you see? Does adherence seem related to number of medications?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.2: Run Simple Logistic Regression

**Fill in the blanks** to run a logistic regression predicting adherence from number of medications:

```{r}
#| label: activity1-model
#| eval: false

# Fit logistic regression model
model_adhere <- glm(___ ~ ___,  # outcome ~ predictor
                    data = ___,
                    family = ___)  # Don't forget this!

# View the results
summary(model_adhere)
```

## Task 1.3: Interpret the Log Odds Ratio

Look at the coefficient for `num_medications` in your output.

**Fill in the interpretation**:

```{r}
#| label: activity1-interpret
#| eval: false

# Extract the coefficient
log_or <- coef(model_adhere)["num_medications"]
p_value <- summary(model_adhere)$coefficients["num_medications", "Pr(>|z|)"]

# Print interpretation
cat("Log Odds Ratio:", round(log_or, 4), "\n")
cat("P-value:", round(p_value, 4), "\n")
```

**Question**: What is the direction of the effect? Is it statistically significant (p < 0.05)?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.4: Calculate and Interpret Odds Ratio

**Fill in the blanks** to calculate the odds ratio:

```{r}
#| label: activity1-or
#| eval: false

# Calculate odds ratio
or_medications <- exp(___(model_adhere)["num_medications"])

cat("Odds Ratio:", round(or_medications, 4), "\n")

# Calculate percent change
if (or_medications < 1) {
  pct_change <- (1 - ___) * 100  # Percent decrease
  cat("For each additional medication, odds of adherence DECREASE by",
      round(pct_change, 1), "%\n")
} else {
  pct_change <- (___ - 1) * 100  # Percent increase
  cat("For each additional medication, odds of adherence INCREASE by",
      round(pct_change, 1), "%\n")
}
```

**Question**: Interpret the odds ratio in plain language. What does this mean for patient management?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 1.5: Make Predictions

**Fill in the blanks** to predict adherence probability for patients taking different numbers of medications:

```{r}
#| label: activity1-predict
#| eval: false

# Create data for prediction
new_patients <- data.frame(
  num_medications = c(2, 6, 10)  # Low, medium, high medication counts
)

# Predict probabilities
new_patients$predicted_prob <- predict(___,
                                       newdata = ___,
                                       type = ___)  # Don't forget this!

# Display results
kable(new_patients, digits = 3,
      col.names = c("Number of Medications", "Predicted Probability of Adherence"))
```

**Question**: How does predicted adherence change as medication count increases? What threshold would you use to identify high-risk patients?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

# Activity 2: Emergency Department Return Visits (Simple Logistic Regression)

## Scenario

An emergency department wants to predict whether patients will return within 72 hours based on their initial pain score (0-10 scale).

Early return visits to the emergency department (within 72 hours) are often considered an indicator of incomplete treatment or inadequate discharge planning (preparing patients for safe transition home). Pain management is a critical component of emergency department care, and uncontrolled pain is a common reason for return visits.

## Dataset Generation

```{r}
#| label: activity2-data

# ED return visit data
n_ed <- 120
pain_score <- round(runif(n_ed, 1, 10))
log_odds_return <- -2.5 + 0.35 * pain_score + rnorm(n_ed, 0, 0.6)
prob_return <- exp(log_odds_return) / (1 + exp(log_odds_return))
returned <- rbinom(n_ed, 1, prob_return)

ed_data <- data.frame(
  visit_id = 1:n_ed,
  pain_score = pain_score,
  returned_72hr = returned
)

# View the data
head(ed_data, 10)
```

## Task 2.1: Run Logistic Regression

**Fill in the blanks** to predict 72-hour return from pain score:

```{r}
#| label: activity2-model
#| eval: false

# Fit the model
model_ed <- glm(___ ~ ___,
                data = ___,
                family = ___)

# View results
summary(___)
```

## Task 2.2: Interpret Results

**Fill in the blanks** to extract and interpret the odds ratio:

```{r}
#| label: activity2-or
#| eval: false

# Calculate odds ratio
or_pain <- ___(coef(model_ed)["pain_score"])

cat("Odds Ratio for pain_score:", round(or_pain, 3), "\n")

# Interpret
if (or_pain > 1) {
  pct_increase <- (or_pain - ___) * 100
  cat("For each 1-point increase in pain score,")
  cat("the odds of returning within 72 hours INCREASE by",
      round(pct_increase, 1), "%\n")
}
```

**Question**: Is pain score a significant predictor? What does this suggest about pain management at discharge?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 2.3: Evaluate Classification Accuracy

**Fill in the blanks** to assess model performance:

```{r}
#| label: activity2-classification
#| eval: false

# Add predictions to data
ed_data$predicted_prob <- predict(___, type = ___)
ed_data$predicted_class <- ifelse(ed_data$predicted_prob >= ___, 1, 0)  # Use 0.5 threshold

# Create confusion matrix
confusion_matrix <- table(
  Predicted = ed_data$___,
  Actual = ed_data$___
)

# Display
confusion_matrix

# Calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(___)
cat("\nOverall Accuracy:", round(accuracy * 100, 1), "%\n")
```

**Question**: What is the model's accuracy? Looking at the confusion matrix, does the model make more false positives or false negatives? Which type of error is more concerning in this context?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

# Activity 3: Surgical Complication Risk (Multiple Logistic Regression)

## Scenario

A surgical department wants to predict the risk of post-operative complications (problems that occur after surgery). They have collected data on patient age (`age`), BMI (`bmi` - body mass index, a measure of body weight relative to height), surgery duration in minutes (`surgery_duration`), and whether the surgery was elective (planned) or emergency (`surgery_type`).

Post-operative complications represent a significant concern in surgical care, with factors such as patient characteristics, procedure complexity, and surgical urgency all potentially contributing to complication risk. Understanding which factors predict complications can help with risk stratification and targeted prevention strategies.

## Dataset Generation

```{r}
#| label: activity3-data

# Surgical complication data
n_surg <- 180

surgery_data <- data.frame(
  patient_id = 1:n_surg,
  age = round(rnorm(n_surg, 60, 15)),
  bmi = round(rnorm(n_surg, 28, 5), 1),
  surgery_duration = round(rnorm(n_surg, 120, 45)),
  surgery_type = sample(c("Elective", "Emergency"), n_surg,
                        replace = TRUE, prob = c(0.7, 0.3))
)

# Generate outcomes
surgery_data$age <- pmax(pmin(surgery_data$age, 90), 25)
surgery_data$bmi <- pmax(pmin(surgery_data$bmi, 45), 18)
surgery_data$surgery_duration <- pmax(surgery_data$surgery_duration, 30)

type_effect <- ifelse(surgery_data$surgery_type == "Emergency", 0.8, 0)
log_odds_comp <- -4 + 0.03 * surgery_data$age + 0.05 * surgery_data$bmi +
                 0.01 * surgery_data$surgery_duration + type_effect +
                 rnorm(n_surg, 0, 0.5)
prob_comp <- exp(log_odds_comp) / (1 + exp(log_odds_comp))
surgery_data$complication <- rbinom(n_surg, 1, prob_comp)

# View the data
head(surgery_data, 10)
```

## Task 3.1: Explore the Data

**Fill in the blanks** to summarize complication rates by surgery type:

```{r}
#| label: activity3-explore
#| eval: false

# Summary by surgery type
surgery_summary <- surgery_data %>%
  group_by(___) %>%
  summarize(
    n = n(),
    complication_rate = round(mean(___) * 100, 1),
    avg_age = round(mean(___), 1),
    avg_bmi = round(mean(___), 1)
  )

surgery_summary
```

**Question**: What initial patterns do you notice? Do emergency surgeries appear to have different characteristics?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 3.2: Run Multiple Logistic Regression

**Fill in the blanks** to fit a model with all four predictors:

```{r}
#| label: activity3-model
#| eval: false

# Fit multiple logistic regression
model_surg <- glm(complication ~ ___ + ___ + ___ + ___,  # All 4 predictors
                  data = ___,
                  family = ___)

# View results
summary(___)
```

## Task 3.3: Interpret Odds Ratios

**Fill in the blanks** to create a summary table of all odds ratios:

```{r}
#| label: activity3-or
#| eval: false

# Create summary table
or_summary <- data.frame(
  Variable = names(coef(model_surg)),
  Log_Odds_Ratio = round(coef(___), 4),
  Odds_Ratio = round(exp(___(___))), 4),
  P_value = round(summary(___)$coefficients[, 4], 4)
)

kable(or_summary, row.names = FALSE)
```

**Question**: Which predictors are statistically significant (p < 0.05)? Which factor has the strongest effect on complication risk?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 3.4: Interpret Specific Coefficients

**Fill in the blanks** to interpret the effect of age:

```{r}
#| label: activity3-interpret-age
#| eval: false

# Extract OR for age
or_age <- exp(coef(model_surg)["age"])
pct_change_age <- (or_age - 1) * ___  # Calculate percent change

cat("For each 1-year increase in age,")
cat("holding BMI, surgery duration, and surgery type constant,")
cat("the odds of complication change by", round(pct_change_age, 2), "%\n")
```

**Question**: Interpret the odds ratio for `surgery_typeEmergency`. What does it tell you about emergency vs. elective surgeries?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 3.5: Make Predictions for Patient Profiles

**Fill in the blanks** to predict complication risk for different patient types:

```{r}
#| label: activity3-predict
#| eval: false

# Create patient profiles
patient_profiles <- data.frame(
  Profile = c("Low Risk", "Medium Risk", "High Risk"),
  age = c(45, 65, 80),
  bmi = c(24, 30, 38),
  surgery_duration = c(90, 150, 210),
  surgery_type = c("Elective", "Elective", "Emergency")
)

# Predict probabilities
patient_profiles$predicted_prob <- predict(___,
                                          newdata = ___,
                                          type = ___)

# Display
kable(patient_profiles, digits = 3)
```

**Question**: How much does the predicted probability change from low to high risk? What does this tell you about the combined effect of risk factors?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

# Activity 4: Diabetes Management Success (Multiple Logistic Regression)

## Scenario

A diabetes clinic wants to predict which patients will achieve good blood sugar control. This is measured using HbA1c (a blood test that shows average blood sugar levels over the past 2-3 months). A value below 7% is considered good control.

The clinic has data on patient age (`age`), years since diagnosis (`years_since_dx`), number of clinic visits in the past year (`clinic_visits`), and whether they use insulin injections (`uses_insulin`).

Diabetes management success depends on multiple factors including patient engagement (reflected in clinic visit frequency), disease duration, treatment complexity, and age-related considerations. Understanding which factors predict successful glycemic control can help clinics target support resources effectively.

## Dataset Generation

```{r}
#| label: activity4-data

# Diabetes management data
n_dm <- 150

diabetes_data <- data.frame(
  patient_id = 1:n_dm,
  age = round(rnorm(n_dm, 58, 12)),
  years_since_dx = round(runif(n_dm, 1, 20)),
  clinic_visits = rpois(n_dm, lambda = 3),
  uses_insulin = sample(c("Yes", "No"), n_dm, replace = TRUE, prob = c(0.4, 0.6))
)

# Generate outcomes
diabetes_data$age <- pmax(pmin(diabetes_data$age, 85), 30)
insulin_effect <- ifelse(diabetes_data$uses_insulin == "Yes", -0.5, 0)

log_odds_control <- -1 + 0.02 * diabetes_data$age - 0.08 * diabetes_data$years_since_dx +
                    0.3 * diabetes_data$clinic_visits + insulin_effect +
                    rnorm(n_dm, 0, 0.6)
prob_control <- exp(log_odds_control) / (1 + exp(log_odds_control))
diabetes_data$good_control <- rbinom(n_dm, 1, prob_control)

# View the data
head(diabetes_data, 10)
```

## Task 4.1: Run Multiple Logistic Regression

**Fill in the blanks** to model good control from all predictors:

```{r}
#| label: activity4-model
#| eval: false

# Fit the model
model_dm <- glm(___ ~ age + ___ + ___ + ___,
                data = ___,
                family = ___)

# View results
summary(___)
```

## Task 4.2: Compare Predictor Effects

**Fill in the blanks** to create a comparison table:

```{r}
#| label: activity4-compare
#| eval: false

# Extract key statistics
coef_table <- data.frame(
  Predictor = names(coef(model_dm)),
  Odds_Ratio = round(___(coef(___)), 3),
  P_value = round(summary(___)$coefficients[, "Pr(>|z|)"], 4),
  Significant = ifelse(summary(___)$coefficients[, "Pr(>|z|)"] < 0.05, "Yes", "No")
)

kable(coef_table, row.names = FALSE)
```

**Question**: Which predictors are significant? Which factor has the largest effect on achieving good control?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 4.3: Interpret Clinic Visits Effect

**Fill in the blanks**:

```{r}
#| label: activity4-interpret-visits
#| eval: false

# Calculate effect of clinic visits
or_visits <- exp(coef(model_dm)["clinic_visits"])
pct_change <- (___ - 1) * 100

cat("For each additional clinic visit per year,")
cat("holding age, years since diagnosis, and insulin use constant,")
cat("the odds of achieving good control",
    ifelse(or_visits > 1, "INCREASE", "DECREASE"),
    "by", round(abs(pct_change), 1), "%\n")
```

**Question**: What does this suggest about the importance of regular clinic follow-up?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 4.4: Make Prediction for Specific Patient

**Fill in the blanks** to predict for a 55-year-old patient diagnosed 8 years ago, with 4 clinic visits, using insulin:

```{r}
#| label: activity4-predict
#| eval: false

# Create patient data
new_patient <- data.frame(
  age = ___,
  years_since_dx = ___,
  clinic_visits = ___,
  uses_insulin = ___  # Use "Yes" or "No"
)

# Predict probability
predicted_prob <- predict(___,
                         newdata = ___,
                         type = ___)

cat("Predicted probability of good control:", round(predicted_prob * 100, 1), "%\n")
```

**Question**: Based on this probability, would you classify this patient as likely to achieve good control? What interventions might you recommend?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

## Task 4.5: Compare Simple vs. Multiple Regression

**Fill in the blanks** to see how controlling for confounders affects the clinic visits coefficient:

```{r}
#| label: activity4-compare-models
#| eval: false

# Simple regression (clinic visits only)
model_simple <- glm(___ ~ clinic_visits,
                   data = ___,
                   family = ___)

# Compare coefficients
comparison <- data.frame(
  Model = c("Simple (visits only)", "Multiple (all predictors)"),
  Odds_Ratio_Visits = c(
    round(exp(coef(model_simple)["clinic_visits"]), 3),
    round(exp(coef(___)["clinic_visits"]), 3)
  )
)

kable(comparison)
```

**Question**: How does the odds ratio for clinic visits change when we control for other factors? What does this tell you about confounding?

<div style="background-color: #f0f0f0; padding: 10px; margin: 10px 0;">
**Your answer**:

</div>

---

# Summary and Reflection

## Key Takeaways

After completing this practice activity, you should be able to:

- ✓ Run simple and multiple logistic regression using `glm()` with `family = binomial`
- ✓ Interpret log odds ratios and convert them to odds ratios using `exp()`
- ✓ Calculate percent changes in odds from odds ratios
- ✓ Make predictions using `predict()` with `type = "response"`
- ✓ Evaluate classification accuracy using confusion matrices
- ✓ Understand how controlling for confounders affects coefficient estimates

## Logistic Regression Quick Reference

| **Task** | **R Code** |
|----------|-----------|
| **Run simple logistic regression** | `glm(outcome ~ predictor, data, family = binomial)` |
| **Run multiple logistic regression** | `glm(outcome ~ pred1 + pred2 + pred3, data, family = binomial)` |
| **Get odds ratios** | `exp(coef(model))` |
| **Make predictions** | `predict(model, newdata, type = "response")` |
| **Classification** | `ifelse(predicted_prob >= 0.5, 1, 0)` |
| **Confusion matrix** | `table(Predicted = pred_class, Actual = actual)` |
| **Accuracy** | `mean(pred_class == actual)` |

## Interpretation Guide

**Odds Ratio Interpretation:**

- **OR < 1**: Protective factor (decreases odds)
  - Example: OR = 0.75 → 25% decrease in odds
  - Formula: `(1 - OR) × 100`

- **OR = 1**: No effect

- **OR > 1**: Risk factor (increases odds)
  - Example: OR = 1.30 → 30% increase in odds
  - Formula: `(OR - 1) × 100`

**In multiple regression**, remember:
- Each coefficient represents the effect **holding all other variables constant**
- This helps isolate the **independent effect** of each predictor
- Comparing simple vs. multiple regression reveals **confounding**

---

# Activity Solutions

Solutions for these activities are available below. Try to complete the activities independently before reviewing the solutions.

## Activity 1 Solutions

### Task 1.1 Solution

```{r}
#| label: solution-1-1
#| code-fold: true

# Calculate adherence rate by number of medications (grouped)
adherence_summary <- adherence_data %>%
  mutate(med_group = cut(num_medications,
                         breaks = c(0, 4, 8, 12),
                         labels = c("1-4 meds", "5-8 meds", "9-12 meds"))) %>%
  group_by(med_group) %>%
  summarize(
    n_patients = n(),
    adherence_rate = round(mean(adherent) * 100, 1)
  )

adherence_summary
```

**Pattern**: Adherence rates tend to decrease as the number of medications increases, suggesting polypharmacy (taking many medications) is associated with lower adherence.

### Task 1.2 Solution

```{r}
#| label: solution-1-2
#| code-fold: true

# Fit logistic regression model
model_adhere <- glm(adherent ~ num_medications,
                    data = adherence_data,
                    family = binomial)

summary(model_adhere)
```

### Task 1.3 Solution

```{r}
#| label: solution-1-3
#| code-fold: true

# Extract the coefficient
log_or <- coef(model_adhere)["num_medications"]
p_value <- summary(model_adhere)$coefficients["num_medications", "Pr(>|z|)"]

cat("Log Odds Ratio:", round(log_or, 4), "\n")
cat("P-value:", round(p_value, 4), "\n")
```

The coefficient is negative, indicating that more medications are associated with lower adherence. The p-value indicates whether this effect is statistically significant.

### Task 1.4 Solution

```{r}
#| label: solution-1-4
#| code-fold: true

# Calculate odds ratio
or_medications <- exp(coef(model_adhere)["num_medications"])

cat("Odds Ratio:", round(or_medications, 4), "\n")

# Calculate percent change
if (or_medications < 1) {
  pct_change <- (1 - or_medications) * 100
  cat("For each additional medication, odds of adherence DECREASE by",
      round(pct_change, 1), "%\n")
} else {
  pct_change <- (or_medications - 1) * 100
  cat("For each additional medication, odds of adherence INCREASE by",
      round(pct_change, 1), "%\n")
}
```

**Interpretation**: Each additional medication is associated with decreased odds of adherence. This suggests that medication regimen simplification may improve adherence rates.

### Task 1.5 Solution

```{r}
#| label: solution-1-5
#| code-fold: true

# Create data for prediction
new_patients <- data.frame(
  num_medications = c(2, 6, 10)
)

# Predict probabilities
new_patients$predicted_prob <- predict(model_adhere,
                                       newdata = new_patients,
                                       type = "response")

kable(new_patients, digits = 3,
      col.names = c("Number of Medications", "Predicted Probability of Adherence"))
```

**Insight**: Predicted adherence probability decreases substantially as medication count increases. Patients taking 10+ medications may need additional support interventions.

## Activity 2 Solutions

### Task 2.1 Solution

```{r}
#| label: solution-2-1
#| code-fold: true

# Fit the model
model_ed <- glm(returned_72hr ~ pain_score,
                data = ed_data,
                family = binomial)

summary(model_ed)
```

### Task 2.2 Solution

```{r}
#| label: solution-2-2
#| code-fold: true

# Calculate odds ratio
or_pain <- exp(coef(model_ed)["pain_score"])

cat("Odds Ratio for pain_score:", round(or_pain, 3), "\n")

# Interpret
if (or_pain > 1) {
  pct_increase <- (or_pain - 1) * 100
  cat("For each 1-point increase in pain score,\n")
  cat("the odds of returning within 72 hours INCREASE by",
      round(pct_increase, 1), "%\n")
}
```

**Implication**: Higher pain scores at discharge significantly predict return visits, suggesting the importance of adequate pain management before discharge.

### Task 2.3 Solution

```{r}
#| label: solution-2-3
#| code-fold: true

# Add predictions to data
ed_data$predicted_prob <- predict(model_ed, type = "response")
ed_data$predicted_class <- ifelse(ed_data$predicted_prob >= 0.5, 1, 0)

# Create confusion matrix
confusion_matrix <- table(
  Predicted = ed_data$predicted_class,
  Actual = ed_data$returned_72hr
)

confusion_matrix

# Calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
cat("\nOverall Accuracy:", round(accuracy * 100, 1), "%\n")
```

**Analysis**: The confusion matrix shows the balance of errors. False negatives (missed returns) may be more concerning as these patients didn't receive appropriate discharge planning.

## Activity 3 Solutions

### Task 3.1 Solution

```{r}
#| label: solution-3-1
#| code-fold: true

# Summary by surgery type
surgery_summary <- surgery_data %>%
  group_by(surgery_type) %>%
  summarize(
    n = n(),
    complication_rate = round(mean(complication) * 100, 1),
    avg_age = round(mean(age), 1),
    avg_bmi = round(mean(bmi), 1)
  )

surgery_summary
```

### Task 3.2 Solution

```{r}
#| label: solution-3-2
#| code-fold: true

# Fit multiple logistic regression
model_surg <- glm(complication ~ age + bmi + surgery_duration + surgery_type,
                  data = surgery_data,
                  family = binomial)

summary(model_surg)
```

### Task 3.3 Solution

```{r}
#| label: solution-3-3
#| code-fold: true

# Create summary table
or_summary <- data.frame(
  Variable = names(coef(model_surg)),
  Log_Odds_Ratio = round(coef(model_surg), 4),
  Odds_Ratio = round(exp(coef(model_surg)), 4),
  P_value = round(summary(model_surg)$coefficients[, 4], 4)
)

kable(or_summary, row.names = FALSE)
```

### Task 3.4 Solution

```{r}
#| label: solution-3-4
#| code-fold: true

# Extract OR for age
or_age <- exp(coef(model_surg)["age"])
pct_change_age <- (or_age - 1) * 100

cat("For each 1-year increase in age,\n")
cat("holding BMI, surgery duration, and surgery type constant,\n")
cat("the odds of complication increase by", round(pct_change_age, 2), "%\n")
```

**Interpretation of surgery_typeEmergency**: This coefficient compares emergency surgeries to elective surgeries (the reference category), holding all other factors constant. An OR > 1 indicates emergency surgeries have higher complication odds.

### Task 3.5 Solution

```{r}
#| label: solution-3-5
#| code-fold: true

# Create patient profiles
patient_profiles <- data.frame(
  Profile = c("Low Risk", "Medium Risk", "High Risk"),
  age = c(45, 65, 80),
  bmi = c(24, 30, 38),
  surgery_duration = c(90, 150, 210),
  surgery_type = c("Elective", "Elective", "Emergency")
)

# Predict probabilities
patient_profiles$predicted_prob <- predict(model_surg,
                                          newdata = patient_profiles,
                                          type = "response")

kable(patient_profiles, digits = 3)
```

**Insight**: The cumulative effect of multiple risk factors leads to dramatically different predicted probabilities, demonstrating the power of multiple regression.

## Activity 4 Solutions

### Task 4.1 Solution

```{r}
#| label: solution-4-1
#| code-fold: true

# Fit the model
model_dm <- glm(good_control ~ age + years_since_dx + clinic_visits + uses_insulin,
                data = diabetes_data,
                family = binomial)

summary(model_dm)
```

### Task 4.2 Solution

```{r}
#| label: solution-4-2
#| code-fold: true

# Extract key statistics
coef_table <- data.frame(
  Predictor = names(coef(model_dm)),
  Odds_Ratio = round(exp(coef(model_dm)), 3),
  P_value = round(summary(model_dm)$coefficients[, "Pr(>|z|)"], 4),
  Significant = ifelse(summary(model_dm)$coefficients[, "Pr(>|z|)"] < 0.05, "Yes", "No")
)

kable(coef_table, row.names = FALSE)
```

### Task 4.3 Solution

```{r}
#| label: solution-4-3
#| code-fold: true

# Calculate effect of clinic visits
or_visits <- exp(coef(model_dm)["clinic_visits"])
pct_change <- (or_visits - 1) * 100

cat("For each additional clinic visit per year,\n")
cat("holding age, years since diagnosis, and insulin use constant,\n")
cat("the odds of achieving good control",
    ifelse(or_visits > 1, "INCREASE", "DECREASE"),
    "by", round(abs(pct_change), 1), "%\n")
```

**Clinical implication**: Regular follow-up visits are strongly associated with better diabetes control, supporting the value of continuous care management.

### Task 4.4 Solution

```{r}
#| label: solution-4-4
#| code-fold: true

# Create patient data
new_patient <- data.frame(
  age = 55,
  years_since_dx = 8,
  clinic_visits = 4,
  uses_insulin = "Yes"
)

# Predict probability
predicted_prob <- predict(model_dm,
                         newdata = new_patient,
                         type = "response")

cat("Predicted probability of good control:", round(predicted_prob * 100, 1), "%\n")
```

### Task 4.5 Solution

```{r}
#| label: solution-4-5
#| code-fold: true

# Simple regression (clinic visits only)
model_simple <- glm(good_control ~ clinic_visits,
                   data = diabetes_data,
                   family = binomial)

# Compare coefficients
comparison <- data.frame(
  Model = c("Simple (visits only)", "Multiple (all predictors)"),
  Odds_Ratio_Visits = c(
    round(exp(coef(model_simple)["clinic_visits"]), 3),
    round(exp(coef(model_dm)["clinic_visits"]), 3)
  )
)

kable(comparison)
```

**Understanding confounding**: The change in the clinic visits coefficient between simple and multiple regression reveals confounding. Other factors (age, disease duration, insulin use) were related to both clinic visits and outcomes, potentially inflating or deflating the simple relationship.

---
