---
title: "Logistic Regression"
subtitle: "Week 13"
author: ""
date: "2026-04-8"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    preview-links: auto
    logo: ""
    footer: "Week 13 | Logistic Regression"
    scrollable: true
    incremental: false
    code-line-numbers: false
    highlight-style: github
    navigation-mode: linear
execute:
  warning: false
  message: false
  echo: true
---

```{r setup}
#| include: false
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
library(scales)
library(patchwork)

set.seed(2026)
```

## Today's Learning Goals

By the end of this lecture, you will be able to:

::: incremental
1.  **Explain** why logistic regression is necessary for **binary outcomes** and how it differs from linear regression
2.  **Interpret** logistic regression coefficients including log odds ratios and odds ratios
3.  **Run and interpret** both simple and multiple logistic regression models in R
4.  **Evaluate** model performance using classification tables
:::

## Why Logistic Regression?

In healthcare management, many critical decisions involve **binary** outcomes:

::: incremental
-   Will a patient be **readmitted** within 30 days?
-   Will a patient **adhere** to their medication?
-   Will a treatment be **successful** or not?
-   Will a patient develop a **complication** after surgery?
:::

. . .

<br>

These yes/no questions are fundamentally different from predicting continuous outcomes (e.g., blood pressure, weight, LOS). We cannot use linear regression because it is designed for continuous outcomes.

. . .

We need a different approach: **logistic regression**.

## The Problem with Linear Regression for Binary Outcomes

Imagine predicting readmission (1 = yes, 0 = no) from length of hospital stay (in days):

$$P(\text{Readmission}) = -0.2 + 0.05 \times \text{Length of Stay}$$

. . .

What if a patient has a very long length of stay of 30 days?

$$P(\text{Readmission}) = -0.2 + 0.05 \times 30 = 1.3$$

. . .

<br>

Probabilities must be between 0 and 1, but the predicted probability of readmission exceeds 1. Linear regression can produce impossible predictions for binary outcomes.

## The Logistic Solution: An S-Shaped Curve

Logistic regression uses a special **S-shaped curve** (logistic function) that naturally stays between 0 and 1:

```{r}
#| label: fig-sigmoid-curve
#| echo: false
#| code-fold: true
#| fig-width: 9
#| fig-height: 4.5

x <- seq(-6, 6, length.out = 200)
linear <- 0.5 + 0.15 * x
logistic <- 1 / (1 + exp(-x))

plot_data <- data.frame(
  x = rep(x, 2),
  y = c(linear, logistic),
  Model = rep(c("Linear Regression", "Logistic Regression"), each = 200)
)

ggplot(plot_data, aes(x = x, y = y, color = Model)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = c(0, 1), linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("Linear Regression" = "red",
                                 "Logistic Regression" = "steelblue")) +
  labs(title = "Linear Regression vs. Logistic Regression",
       subtitle = "Logistic regression keeps predictions between 0 and 1",
       x = "Predictor Value",
       y = "Predicted Probability") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom") +
  annotate("text", x = 6, y = 1.15, label = "Impossible!\n(p > 1)",
           color = "darkred", size = 3) +
  annotate("text", x = -5.5, y = -0.15, label = "Impossible!\n(p < 0)",
           color = "darkred", size = 3)
```

# 1. Core Concepts of Logistic Regression {background-color="#40666e"}

## From Probability to Odds

**Probability**: Chance that something happens (0 to 1)

. . .

**Odds**: Ratio of event happening to event not happening

$$\text{Odds} = \frac{p}{1-p} \qquad \Longleftrightarrow \qquad p = \frac{\text{Odds}}{1 + \text{Odds}}$$

. . .

<br> **Example** (readmission probability = 0.30):

-   Probability of readmission = 0.30
-   Probability of NOT readmitted = 0.70
-   Odds = 0.30 / 0.70 = 0.429

## Understanding Odds Values

```{r}
#| label: odds-table
#| echo: false

prob_list <- c(0.10, 0.25, 0.33, 0.50, 0.67, 0.75, 0.90)
complement_list <- 1 - prob_list
odds_list <- round(prob_list / complement_list, 2)

odds_table <- data.frame(
  Probability = prob_list,
  `Complement (1-p)` = complement_list,
  Odds = odds_list,
  Interpretation = c(
    "Very unlikely", "", "",
    "Equal chance (50/50)",
    "", "", "Very likely"
  )
)

kable(odds_table, col.names = c("Probability (p)", "1 - p", "Odds", "Interpretation"),
      align = "c", digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 22)
```

. . .

**Key**:

-   Odds = 1 means 50/50 chance
-   Odds \< 1 = less likely; Odds \> 1 = more likely
-   Odds can be from 0 to positive infinity <br> (unbounded above, but cannot be negative)

## From Odds to Log Odds (Logit)

Odds cannot be negative (range: 0 to positive infinity). We need an **unbounded** scale (-∞ to +∞) for linear modeling.

. . .

**Log Odds** = natural logarithm of odds:

$$\text{Log Odds} = \ln\left(\frac{p}{1-p}\right)$$

. . .

:::::: columns
::: {.column width="60%"}
```{r}
#| label: logit-table
#| echo: false

logodds_list <- log(odds_list)

logit_table <- data.frame(
  Probability = prob_list,
  `Complement (1-p)` = complement_list,
  Odds = odds_list,
  `Log Odds` = logodds_list
)

kable(logit_table, col.names = c("Probability (p)", "1 - p", "Odds", "Log Odds (Logit)"),
      align = "c", digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 22)
```
:::

:::: {.column width="40%"}
**Key points**

::: {style="font-size: 0.8em;"}
-   When **p = 0.5**, **log odds = 0**
-   When **p \> 0.5**, log odds are **positive**
-   When **p \< 0.5**, log odds are **negative**
-   Log odds map probabilities (0, 1) onto the full real line (−∞, +∞)
:::
::::
::::::

## Visualizing the Three Scales

```{r}
#| label: fig-three-scales
#| echo: false
#| code-fold: true
#| fig-width: 9
#| fig-height: 4.5

prob_seq <- seq(0.01, 0.99, by = 0.01)
odds_seq <- prob_seq / (1 - prob_seq)
logodds_seq <- log(odds_seq)

transform_data <- data.frame(
  probability = prob_seq,
  odds = odds_seq,
  log_odds = logodds_seq
)

p1 <- ggplot(transform_data, aes(x = prob_seq, y = probability)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  labs(x = "Probability", y = "Probability", title = "Probability Scale") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p2 <- ggplot(transform_data, aes(x = prob_seq, y = odds)) +
  geom_line(color = "darkgreen", linewidth = 1.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  annotate("text", x = 0.15, y = 1.5, label = "Odds = 1\n(50/50 chance)",
           size = 3, color = "gray30") +
  labs(x = "Probability", y = "Odds", title = "Odds Scale") +
  scale_x_continuous(limits = c(0, 1)) +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p3 <- ggplot(transform_data, aes(x = prob_seq, y = log_odds)) +
  geom_line(color = "darkred", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  annotate("text", x = 0.15, y = 0.5, label = "Log Odds = 0\n(50/50 chance)",
           size = 3, color = "gray30") +
  labs(x = "Probability", y = "Log Odds (Logit)", title = "Log Odds Scale") +
  scale_x_continuous(limits = c(0, 1)) +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p1 + p2 + p3 +
  plot_annotation(
    title = "Three Ways to Express the Same Information",
    theme = theme(plot.title = element_text(face = "bold", size = 12))
  )
```

## The Logistic Regression Equation

Instead of predicting probability directly, we predict the **log odds**:

$$\ln\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_k X_k$$

. . .

<br> The coefficients ($\beta$ values) represent the **change in log odds** per one-unit increase in the predictor.

. . .

<br> **Converting back to probability**:

$$p = \frac{e^{\beta_0 + \beta_1 X_1 + \cdots}}{1 + e^{\beta_0 + \beta_1 X_1 + \cdots}} = \frac{odds}{1 + odds}$$

## Maximum Likelihood Estimation (MLE)

Unlike linear regression (OLS), logistic regression uses **MLE** because:

::: incremental
-   OLS could predict probabilities **outside the plausible range (0-1)**
-   Binary outcomes **violate OLS assumptions** (non-normal residuals, non-constant variance)
-   MLE finds coefficient values that make the observed data **most likely**
:::

. . .

<br>

You don't need to understand the math of MLE as R does it for you (`glm()` with `family = binomial`, R uses MLE automatically).

# 2. Simple Logistic Regression {background-color="#40666e"}

## What Is It? / When to Use It

**Simple logistic regression** models the relationship between **one predictor** and a **binary outcome** (0/1).

. . .

<br> **Use it when:**

::: incremental
-   Your **outcome variable is binary** (two possible values)
-   You have **one predictor variable** to examine
-   You want to understand **how the predictor affects the probability** of the outcome, quantified through **log odds ratios** and **odds ratios**
:::

## Example Case: Predicting Hospital Readmission

**Scenario**: A quality improvement analyst wants to understand whether *discharge readiness* predicts *30-day readmission*.

<br> The dataset contains records for **200 patients** discharged from a single hospital, with each row representing one patient and including their discharge readiness score, demographics, clinical characteristics, and whether they were readmitted within 30 days.

. . .

<br> **Hypothesis**: Patients who feel more ready for discharge have a better understanding of follow-up care and are less likely to be readmitted.

. . .

<br> **Hospital Readmissions Reduction Program (HRRP)**: Hospitals with excess readmissions face financial penalties under Medicare. Understanding predictors of readmission helps target interventions.

## Loading and Exploring the Data

Let's load the data and required packages.

```{r}
#| label: simple-data

# load required packages
library(tidyverse)

# load data
readmit_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week13_readmit_data.csv")

# Set the Medicaid as the reference category for insurance
readmit_data <- readmit_data |>
  mutate(insurance = factor(insurance, levels = c("Medicaid", "Medicare", "Private")))

head(readmit_data, 3) 
```

## Key Variables

| Variable | Description | Type |
|----|----|----|
| `patient_id` | Unique identifier for each patient | ID |
| `discharge_readiness` | Discharge readiness score (1-10, higher = more ready) | Continuous |
| `age` | Patient age in years | Continuous |
| `comorbidities` | Number of chronic conditions (e.g., diabetes, hypertension) | Continuous |
| `insurance` | Insurance type (Medicaid, Medicare, or Private) | Categorical |
| `readmitted` | Whether readmitted within 30 days (1 = yes, 0 = no) — **outcome** | Binary |

## Exploring the Data

Let's start by exploring the distribution of our outcome variable (readmission) and how it relates to our key predictor (discharge readiness).

::::: columns
::: {.column width="50%"}
```{r}
#| label: simple-explore-table
#| echo: false

readmission_summary <- data.frame(
  Outcome = c("Not Readmitted", "Readmitted", "Total"),
  Count = c(
    sum(readmit_data$readmitted == 0),
    sum(readmit_data$readmitted == 1),
    nrow(readmit_data)
  ),
  Percentage = c(
    round(100 * mean(readmit_data$readmitted == 0), 1),
    round(100 * mean(readmit_data$readmitted == 1), 1),
    100
  )
)

kable(readmission_summary,
      col.names = c("Outcome", "N", "%"),
      align = c("l", "c", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```
:::

::: {.column width="50%"}
```{r}
#| label: fig-simple-explore
#| code-fold: true
#| fig-width: 5
#| fig-height: 4

ggplot(readmit_data, aes(x = factor(readmitted), y = discharge_readiness,
                          fill = factor(readmitted))) +
  geom_boxplot(alpha = 0.7, width = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(title = "Discharge Readiness by Readmission Status",
       x = "Readmission Status",
       y = "Discharge Readiness Score (1-10)") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))
```
:::
:::::

## Running Simple Logistic Regression

Let's run a simple logistic regression model to see if discharge readiness predicts readmission. We can do this by using `glm()` with `family = binomial`:

```{r}
#| label: simple-model
#| output-location: slide

model_simple <- glm(readmitted ~ discharge_readiness,
                    data = readmit_data,
                    family = binomial(link = "logit")) # link = "logit" is the default, so you can omit this

summary(model_simple)
```

## Step 1: Understanding the Coefficients

```{r}
#| label: simple-coef-table
#| echo: false

coef_table <- summary(model_simple)$coefficients
kable(coef_table, digits = 4,
      col.names = c("Estimate", "Std. Error", "z value", "p-value")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r}
#| echo: false
b0 <- round(coef(model_simple)[1], 4)
b1 <- round(coef(model_simple)[2], 4)
b1_p <- format(summary(model_simple)$coefficients[2, 4], digits = 3)
```

. . .

<br> **Model equation**:

-   Log Odds of Readmission = `r b0` + (`r b1`) x Discharge Readiness

. . .

1.  For each one-unit increase in discharge readiness (IV), the **log odds** of readmission change by `r b1`.
2.  The p-value for discharge readiness is `r b1_p`, which indicates the effect of discharge readiness on readmission is statistically significant.

## Step 2: Converting to Odds Ratio

Log odds ratios are hard to interpret. **Exponentiate** to get odds ratios:

::: {style="font-size: 0.75em;"}
-   `coef()`: to get regression coefficients (log odds ratios)
-   `exp()`: to exponentiate log odds ratios to get odds ratios
:::

```{r}
#| label: simple-odds-ratio

odds_ratios <- exp(coef(model_simple))
odds_ratios
```

```{r}
#| echo: false
or_dr <- odds_ratios[2]
pct_change <- if(or_dr < 1) round((1 - or_dr) * 100, 1) else round((or_dr - 1) * 100, 1)
```

. . .

**Interpretation**:

::: {.incremental style="font-size: 0.75em;"}
-   **Direction**: Odds ratio is `r if(or_dr < 1) "LESS than 1" else "GREATER than 1"` -\> higher discharge readiness = `r if(or_dr < 1) "LOWER" else "HIGHER"` odds of readmission
-   **Magnitude**: For each 1-point increase in discharge readiness, odds of readmission `r if(or_dr < 1) "DECREASE" else "INCREASE"` by `r pct_change`%
:::

## Step 3: Predicting Probabilities

We can use the logistic regression model to get a predicted probability. For a patient with discharge readiness = 5:

```{r}
#| label: simple-predict-manual
#| echo: false

b0_pred <- coef(model_simple)[1]
b1_pred <- coef(model_simple)[2]
dr_value <- 5

log_odds_pred <- b0_pred + b1_pred * dr_value
odds_pred <- exp(log_odds_pred)
prob_pred <- odds_pred / (1 + odds_pred)
```

::: incremental
1.  **Log odds** = `r round(b0_pred, 3)` + (`r round(b1_pred, 3)`) x 5 = `r round(log_odds_pred, 4)`
2.  **Odds** = $e^{`r round(log_odds_pred, 4)`}$ = `r round(odds_pred, 4)`
3.  **Probability** = `r round(odds_pred, 4)` / (1 + `r round(odds_pred, 4)`) = `r round(prob_pred, 4)` (**`r round(prob_pred * 100, 1)`%**)
:::

## Step 4: Using R's predict() Function

Let's get the predicted probabilities of readmission for a range of discharge readiness scores using R's `predict()` function with `type = "response"`:

```{r}
#| label: simple-predict-r

new_patients <- data.frame(
  discharge_readiness = c(4, 5, 6, 7, 8)
)

new_patients$predicted_prob <- predict(model_simple,
                                        newdata = new_patients,
                                        type = "response") # to get predicted probabilities, type = "response" is required. Without it, you would get log odds instead.
round(head(new_patients, 3), 3)
```

## Step 5: Visualizing Predicted Probabilities

```{r}
#| label: fig-simple-predict-viz
#| code-fold: true
#| fig-width: 9
#| fig-height: 4.5

library(scales)
pred_data <- data.frame(discharge_readiness = seq(3, 10, by = 0.1))
pred_data$probability <- predict(model_simple, newdata = pred_data, type = "response")

ggplot() +
  geom_line(data = pred_data, aes(x = discharge_readiness, y = probability),
            color = "steelblue", linewidth = 1.2) +
  geom_point(data = readmit_data, aes(x = discharge_readiness, y = readmitted),
             alpha = 0.3, position = position_jitter(width = 0.1, height = 0)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  labs(title = "Predicted Probability of Readmission by Discharge Readiness Score",
       subtitle = "Points show actual outcomes (0 = not readmitted, 1 = readmitted)",
       x = "Discharge Readiness Score",
       y = "Predicted Probability of Readmission") +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Step 6: From Probabilities to Decisions

Our model gives us **probabilities**, but in practice we often need to make **binary decisions** (e.g., "provide extra discharge support" vs. "standard discharge").

. . .

<br> To classify patients, we use a **decision threshold** (default = 0.5):

-   If predicted probability ≥ 0.5 → predict **readmitted** (1)
-   If predicted probability \< 0.5 → predict **not readmitted** (0)

. . .

<br> We then compare our predictions to actual outcomes using a **classification table** (also called a **confusion matrix**).

## Understanding the Classification Table

The confusion matrix has four possible outcomes:

|   | **Actual: Not Readmitted** | **Actual: Readmitted** |
|:---|:--:|:--:|
| **Predicted: Not Readmitted** | True Negative (TN) | False Negative (FN) |
| **Predicted: Readmitted** | False Positive (FP) | True Positive (TP) |

<br>

-   **True Negative (TN)**: Correctly predicted NOT readmitted
-   **True Positive (TP)**: Correctly predicted readmitted
-   **False Positive (FP)**: Predicted readmitted, but was NOT (false alarm)
-   **False Negative (FN)**: Predicted NOT readmitted, but WAS (miss)

## Our Model's Classification Table

Let's make a classification table for our simple logistic regression model using a 0.5 threshold:

```{r}
#| label: simple-classification
#| code-fold: false

readmit_data$predicted_prob <- predict(model_simple, type = "response")
readmit_data$predicted_class <- ifelse(readmit_data$predicted_prob >= 0.5, 1, 0)

class_table <- table(
  Predicted = factor(readmit_data$predicted_class, levels = c(0, 1)),
  Actual = factor(readmit_data$readmitted, levels = c(0, 1))
)
class_table
```

```{r}
#| echo: false
correct <- sum(diag(class_table))
total <- sum(class_table)
accuracy <- correct / total
tn <- class_table[1, 1]
fp <- class_table[2, 1]
fn <- class_table[1, 2]
tp <- class_table[2, 2]
```

. . .

**Accuracy** = (TP + TN) / Total = `r correct` / `r total` = **`r round(accuracy * 100, 1)`%**

# 3. Practice 1: Simple Logistic Regression {background-color="#40666e"}

## Practice 1: Medication Adherence

**Scenario**: A pharmacy wants to understand whether the number of medications a patient takes predicts whether they actually take their medications as prescribed.

. . .

<br> **Sample & data**: The dataset contains records for **100 patients** from a single community pharmacy. For each patient, staff recorded the **number of medications** currently prescribed and whether the patient was classified as **adherent** (1 = took medications as prescribed, 0 = did not) based on refill history and self-report during a follow-up visit.

. . .

<br> **Hypothesis**: When patients have more medications to take, it becomes harder to keep track of everything; more pills to remember, different times of day, and more possible side effects. So we expect patients with more medications to be **less likely to take their medications as prescribed**.

## Practice 1: Load Data

Let's load the data using the code below.

```{r}
#| label: activity1-data

# Medication adherence data
adherence_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week13_adherence.data.csv")

# View the data
head(adherence_data, 5)
```

## Practice 1: Key Variables

| Variable | Description | Type |
|----|----|----|
| `patient_id` | Unique identifier for each patient | ID |
| `num_medications` | Number of medications the patient is taking | Continuous |
| `adherent` | Whether the patient is adherent to medication (1 = yes, 0 = no) — **outcome** | Binary |

## Practice 1: Tasks (1/3)

**Task 1**: Explore the data

Let's first explore the relationship between the number of medications and adherence rates by creating medication groups.

```{r}
#| label: task-1-1
#| eval: false

adherence_summary <- adherence_data %>%
  mutate(med_group = cut(num_medications,
                         breaks = c(0, 4, 8, 12),
                         labels = c("1-4 meds", "5-8 meds", "9-12 meds"))) %>% # create medication groups
  group_by(___) %>%
  summarize(
    n_patients = n(),
    adherence_rate = round(mean(___) * 100, 1)
  )
adherence_summary
```

## Practice 1: Tasks (2/3)

**Task 2**: Run simple logistic regression

```{r}
#| label: task-1-2
#| eval: false

model_adhere <- glm(___ ~ ___,
                    data = ___,
                    family = ___)
summary(model_adhere)
```

. . .

<br> **Task 3**: Calculate and interpret odds ratio

```{r}
#| label: task-1-3
#| eval: false

or_medications <- exp(___(model_adhere)["num_medications"])
round(or_medications, 3)
```

## Practice 1: Tasks (3/3)

**Task 4**: Make predictions for three new patients

Use your fitted model to predict the probability of adherence for the following patients:

-   **Patient A** — takes **2 medications**
-   **Patient B** — takes **6 medications**
-   **Patient C** — takes **10 medications**

```{r}
#| label: task-1-4
#| eval: false

new_patients <- data.frame(num_medications = c(___, ___, ___))
new_patients$predicted_prob <- predict(___,
                                       newdata = ___,
                                       type = ___)
round(new_patients, 3)
```

# 4. Multiple Logistic Regression {background-color="#40666e"}

## What Is It? / When to Use It

**Multiple logistic regression** includes **two or more predictors** to predict a binary outcome.

. . .

<br> **Use it when:**

::: incremental
-   Your **outcome is binary** (0/1)
-   You have **multiple predictor variables** (continuous, categorical, or both)
-   You want the **independent effect** of each predictor while **controlling for** other variables
:::

## Example Case: Comprehensive Readmission Model

Our `readmit_data` already contains additional predictors:

::: incremental
-   **Discharge Readiness**: Discharge readiness score (1-10)
-   **Age**: Patient age in years
-   **Comorbidities**: Number of chronic conditions
-   **Insurance**: Private, Medicare, or Medicaid (reference)
:::

. . .

<br>

**The questions we have are**:

-   Does discharge readiness still predict readmission when we account for age, health complexity, and insurance type?
-   And, what are the effects of **age**, **health complexity**, and **insurance type** on the odds of readmission?

## Running Multiple Logistic Regression

We can run a multiple logistic regression model by including all predictors in the formula with the same `glm()` function and `family = binomial`:

```{r}
#| label: multiple-model
#| output-location: slide

model_multiple <- glm(readmitted ~ discharge_readiness + age + comorbidities + insurance,
                      data = readmit_data,
                      family = binomial(link = "logit"))

summary(model_multiple)
```

## Interpreting Multiple Predictors: Odds Ratios

```{r}
#| label: multiple-coef-table
#| echo: false

coef_summary <- data.frame(
  Variable = names(coef(model_multiple)),
  `Log Odds Ratio` = round(coef(model_multiple), 4),
  `Odds Ratio` = round(exp(coef(model_multiple)), 4),
  `p-value` = round(summary(model_multiple)$coefficients[, 4], 4)
)

kable(coef_summary, row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r}
#| echo: false
coefs <- coef(model_multiple)
ors <- exp(coefs)
pvals <- summary(model_multiple)$coefficients[, 4]
pct_readiness <- if(ors["discharge_readiness"] < 1) round((1 - ors["discharge_readiness"]) * 100, 1) else round((ors["discharge_readiness"] - 1) * 100, 1)
pct_comorbid <- round((ors["comorbidities"] - 1) * 100, 1)
```

. . .

<br> **Each coefficient represents the effect while holding all other variables constant.**

## Interpreting Each Predictor

::: incremental
-   **Discharge Readiness** (OR = `r round(ors["discharge_readiness"], 3)`, p = `r round(pvals["discharge_readiness"], 4)`): Each 1-point increase -\> odds of readmission `r if(ors["discharge_readiness"] < 1) "DECREASE" else "INCREASE"` by `r pct_readiness`%, holding other variables constant
-   **Age** (OR = `r round(ors["age"], 4)`, p = `r round(pvals["age"], 4)`): Each additional year -\> odds `r if(ors["age"] > 1) "increase" else "decrease"` by `r round(abs(ors["age"] - 1) * 100, 1)`%
-   **Comorbidities** (OR = `r round(ors["comorbidities"], 3)`, p = `r round(pvals["comorbidities"], 4)`): Each additional condition -\> odds `r if(ors["comorbidities"] > 1) "increase" else "decrease"` by `r pct_comorbid`%
-   **Insurance (Medicare)** (OR = `r round(ors["insuranceMedicare"], 3)`, p = `r round(pvals["insuranceMedicare"], 4)`): Compared to Medicaid, Medicare patients have `r if(ors["insuranceMedicare"] < 1) paste0(round((1 - ors["insuranceMedicare"]) * 100, 1), "% lower") else paste0(round((ors["insuranceMedicare"] - 1) * 100, 1), "% higher")` odds of readmission
-   **Insurance (Private)** (OR = `r round(ors["insurancePrivate"], 3)`, p = `r format(pvals["insurancePrivate"], digits = 3, scientific = FALSE)`): Compared to Medicaid, Private insurance patients have `r if(ors["insurancePrivate"] < 1) paste0(round((1 - ors["insurancePrivate"]) * 100, 1), "% lower") else paste0(round((ors["insurancePrivate"] - 1) * 100, 1), "% higher")` odds of readmission
:::

## Making Predictions: Manual Calculation

Let's predict readmission probability for a patient manually:

-   Discharge readiness = 6, Age = 65, Comorbidities = 2, Insurance = Medicare

```{r}
#| label: manual-predict
#| echo: false

cm <- coef(model_multiple)
log_odds_m <- cm["(Intercept)"] +
              cm["discharge_readiness"] * 6 +
              cm["age"] * 65 +
              cm["comorbidities"] * 2 +
              cm["insuranceMedicare"] * 1
odds_m <- exp(log_odds_m)
prob_m <- odds_m / (1 + odds_m)
```

::: incremental
1.  **Log odds** = `r round(cm["(Intercept)"], 3)` + (`r round(cm["discharge_readiness"], 3)`) x 6 + (`r round(cm["age"], 3)`) x 65 + (`r round(cm["comorbidities"], 3)`) x 2 + (`r round(cm["insuranceMedicare"], 3)`) x 1 = `r round(log_odds_m, 4)`
2.  **Odds** = $e^{`r round(log_odds_m, 4)`}$ = `r round(odds_m, 4)`
3.  **Probability** = `r round(odds_m, 4)` / (1 + `r round(odds_m, 4)`) = `r round(prob_m, 4)` (**`r round(prob_m * 100, 1)`%**)
:::

## Making Predictions: Using `predict()`

We can more easily use `predict()` with `type = "response"` to obtain predicted probabilities for multiple patients at once. Consider three hypothetical patients:

::: {style="font-size: 0.75em;"}
-   **Patient A**: discharge readiness = 8, age = 50, 1 comorbidity, Private insurance
-   **Patient B**: discharge readiness = 6, age = 65, 2 comorbidities, Medicare
-   **Patient C**: discharge readiness = 4, age = 80, 4 comorbidities, Medicaid
:::

```{r}
#| label: multiple-predict
#| code-fold: true

patient_profiles <- data.frame(
  discharge_readiness = c(8, 6, 4),
  age = c(50, 65, 80),
  comorbidities = c(1, 2, 4),
  insurance = c("Private", "Medicare", "Medicaid")
)

patient_profiles$predicted_prob <- predict(model_multiple,
                                           newdata = patient_profiles,
                                           type = "response")
patient_profiles
```

## Evaluating the Multiple Regression Model

Let's check the accuracy of our multiple logistic regression model by creating a classification table using the same 0.5 threshold:

```{r}
#| label: multiple-evaluation

readmit_data$pred_prob_multiple <- predict(model_multiple, type = "response") # getting predicted probabilities of the data
readmit_data$pred_class_multiple <- ifelse(readmit_data$pred_prob_multiple >= 0.5, 1, 0) # classification

class_table_multiple <- table(
  Predicted = readmit_data$pred_class_multiple,
  Actual = readmit_data$readmitted
)
class_table_multiple
```

```{r}
#| echo: false
correct_m <- sum(diag(class_table_multiple))
total_m <- sum(class_table_multiple)
accuracy_m <- correct_m / total_m
tn_m <- class_table_multiple[1, 1]
fp_m <- class_table_multiple[2, 1]
fn_m <- class_table_multiple[1, 2]
tp_m <- class_table_multiple[2, 2]
```

. . .

**Multiple regression accuracy**: `r round(accuracy_m * 100, 1)`% (`r correct_m` / `r total_m` patients correctly classified)

# 5. Practice 2: Multiple Logistic Regression {background-color="#40666e"}

## Practice 2: Surgical Complication Risk

**Scenario**: A surgical department wants to predict the risk of post-operative complications from patient age, BMI, surgery duration, and surgery type.

. . .

<br> **Sample & data**: The dataset contains records for **180 patients** who underwent surgery at a single hospital. For each patient, the team recorded **age**, **BMI**, **surgery duration (minutes)**, whether the surgery was **elective or emergency**, and whether the patient experienced a **post-operative complication** (1 = yes, 0 = no) within 30 days of surgery.

. . .

<br> **Hypothesis**: Older patients, patients with higher BMI, longer surgeries, and emergency (vs. elective) procedures are each expected to be associated with **higher odds of complications**.

## Practice 2: Load Data

Let's start by loading the surgical complications dataset and converting `surgery_type` into a factor so R treats it as a categorical variable:

```{r}
#| label: activity3-data

# Surgical complication data
surgery_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week13_surgery_data.csv")
surgery_data$surgery_type <- as.factor(surgery_data$surgery_type)

# View the data
head(surgery_data, 5)
```

## Practice 2: Key Variables

Before we build the model, let's review the variables we'll be working with:

| Variable | Description | Type |
|----|----|----|
| `patient_id` | Unique identifier for each patient | ID |
| `age` | Patient age in years | Continuous |
| `bmi` | Body mass index (weight relative to height) | Continuous |
| `surgery_duration` | Duration of surgery in minutes | Continuous |
| `surgery_type` | Type of surgery (Elective or Emergency) | Categorical |
| `complication` | Whether a post-operative complication occurred (1 = yes, 0 = no) — **outcome** | Binary |

## Practice 2: Tasks (1/2)

Now it's your turn. Fit a multiple logistic regression model using all four predictors, then interpret what each coefficient tells you about complication risk.

**Task 1**: Run multiple logistic regression

```{r}
#| label: task-3-2
#| eval: false

model_surg <- glm(complication ~ ___ + ___ + ___ + ___,
                  data = ___,
                  family = ___)
summary(___)
```

. . .

<br> **Task 2**: Interpret odds ratios

```{r}
#| label: task-3-3
#| eval: false

round(___(___(model_surg)), 3)
```

## Practice 2: Tasks (2/2)

**Task 3**: Predict for three patient profiles

Suppose we have three new patient profiles. Using the multiple logistic regression model you just fitted, predict each patient's probability of experiencing a post-surgery complication.

-   **Patient A** — age 45, BMI 24, 90-min elective surgery
-   **Patient B** — age 65, BMI 30, 150-min elective surgery
-   **Patient C** — age 80, BMI 38, 210-min emergency surgery

```{r}
#| label: task-3-4
#| eval: false

patient_profiles <- data.frame(
  age = c(___, ___, ___),
  bmi = c(___, ___, ___),
  surgery_duration = c(___, ___, ___),
  surgery_type = c(___, ___, ___)
)
patient_profiles$predicted_prob <- predict(___, newdata = ___, type = ___)
patient_profiles
```

# Summary {background-color="#40666e"}

## Key Takeaways

::: incremental
-   **Logistic regression** is for explaining or predicting **binary outcomes**
    -   Use `glm()` with `family = binomial` in R
-   Coefficients are **log odds ratios**. Exponentiate with `exp()` to get **odds ratios**
    -   **Odds ratio \< 1** = decreases risk; **OR = 1** = no effect; **OR \> 1** = increases risk
-   In **multiple regression**, each coefficient shows the **individual effect** holding other variables constant
-   **Accuracy** measures the overall percentage of correct predictions
    -   The **confusion matrix** is helpful to understand where errors occur
    -   In healthcare, consider which type of error (false positive vs. false negative) is **more costly** for your specific application
:::

## Assignments

1.  **Week 13 Reflection**: Please share your thoughts on the lecture

-   3 interesting things
-   3 confusing or unclear things

Due by 11:59 PM on Sunday, April 12.

. . .

<br> 2. **Lab Assignment 6**: Logistic Regression

We will postpone this assignment (will inform you next week)

## References

::: {style="font-size: 0.85em;"}
-   Hosmer, D. W., Lemeshow, S., & Sturdivant, R. X. (2013). *Applied Logistic Regression* (3rd ed.). Wiley.
-   Weiss, M. E., Piacentine, L. B., Lokken, L., Ancona, J., Archer, J., Ber, S., ... & Vega-Stromberg, T. (2007). Perceived readiness for hospital discharge in adult medical-surgical patients. *Clinical Nurse Specialist*, 21(1), 31-42.
-   Wang, X., Juntasopeepun, P., Panuthai, S., & Suwankruhasn, N. (2025). Factors influencing medication adherence among Chinese older adults with physical multimorbidity and polypharmacy: A cross-sectional study. *Nursing & Health Sciences*, 27(3), e70192.
-   Dharap, S. B., Barbaniya, P., & Navgale, S. (2022). Incidence and risk factors of postoperative complications in general surgery patients. *Cureus*, 14(11), e30975.
:::
