---
title: "Week 2: Practice Activities"
subtitle: ""
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: false
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
set.seed(2026)
```

# Overview

Welcome to the hands-on practice session! This document contains guided exercises to reinforce what you learned in the lecture.

**What you'll practice:**

- Creating variables and working with data types
- Building and manipulating vectors
- Working with data frames
- Using dplyr functions for data wrangling
- Creating your first analysis in Quarto

**Format**: Fill-in-the-blank code sections (marked with `___`)

**Instructions**:

1. Complete the code by replacing `___` with appropriate values
2. Run each chunk to verify your code works
3. Answer the interpretation questions
4. Don't worry about mistakes - they're part of learning!

---

# Common Errors and How to Fix Them

You **WILL** encounter errors - that's completely normal! Here are the most common ones:

## Error 1: "object not found"

```r
Error: object 'patient_data' not found
```

**Cause**: You didn't run the chunk that creates the dataset

**Fix**: Scroll up, run the data generation chunk, then try again

## Error 2: "could not find function"

```r
Error: could not find function "filter"
```

**Cause**: The tidyverse package isn't loaded

**Fix**: Run the `setup` chunk at the top of this document that has `library(tidyverse)`

## Error 3: Spelling and capitalization

R is case-sensitive! `Age` ≠ `age`

**Fix**: Use `names(data_frame)` to check exact column names

## Error 4: Missing comma or parenthesis

```r
# ❌ Wrong:
c(1, 2, 3 4, 5)

# ✅ Right:
c(1, 2, 3, 4, 5)
```

**Fix**: Count your parentheses and commas carefully

::: {.callout-tip}
## Tip: When You Get Stuck

1. Read the error message: It often tells you exactly what's wrong
2. Check your spelling and capitalization
3. Make sure you ran all previous chunks
4. Try running your code line by line to find the problem
5. Ask your classmates or instructor!
:::

---

# Activity 1: R Basics {.activity-time data-time="15 min"}

## Scenario

You're a new data analyst at City General Hospital. Before diving into complex analyses, you need to practice basic R operations using simple patient data.

## Task 1.1: Creating Variables

Create variables for a single patient's information:

```{r}
#| label: task1_1
#| eval: false

# Fill in the blanks
patient_name <- "___"  # Enter any name
patient_age <- ___  # Enter an age (number)
is_admitted <- ___  # TRUE or FALSE
department <- "___"  # Choose: "ER", "ICU", or "Surgery"

# View the values (run this to check your work)
patient_name
patient_age
is_admitted
department
```

**Question**: What data type is each variable? Use `class()` to check:

```{r}
#| label: task1_1_var_class
#| eval: false

# Fill in the blanks to check types
class(___)  # patient_name
class(___)  # patient_age
class(___)  # is_admitted
class(___)  # department
```

::: {.callout-note icon=false}
## Your Answer

- patient_name is type: ___
- patient_age is type: ___
- is_admitted is type: ___
- department is type: ___
:::

## Task 1.2: Working with Vectors

Create vectors containing information for 5 patients:

```{r}
#| label: task1_2
#| eval: false

# Fill in the blanks - create a vector with 5 ages
ages <- c(___, ___, ___, ___, ___)

# Create a vector with 3 departments
departments <- c("___", "___", "___")

# View the vectors
ages
departments

# Calculate average age
mean(___)  # Fill in the vector name

# Find the youngest patient's age
min(___)  # Fill in the vector name

# Find the oldest patient's age
max(___)  # Fill in the vector name
```

**Question**: What is the average age? What are the minimum and maximum ages?

::: {.callout-note icon=false}
## Your Answer

- the average age of the patients is: ___
- the minimum age of the patients is: ___
- the maximum age of the patients is: ___
:::


## Task 1.3: Accessing Vector Elements

Practice extracting specific elements from vectors:

```{r}
#| label: task1_3
#| eval: false

# Use the ages vector from above
ages

# Get the first patient's age
ages[___]  # Fill in the position number

# Get the third patient's age
ages[___]

# Get the first three patients' ages
ages[___:___]  # Fill in the range

# Get ages for patients 2 and 4
ages[c(___, ___)]  # Fill in the positions
```

---

# Activity 2: Working with Data Frames {.activity-time data-time="10 min"}

## Generate Patient Dataset

Run this code to load a patient dataset:

```{r}
# Patient data
patients <- read_csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week02_patients_data.csv")

# View the data
patients
```

## Task 2.1: Exploring the Data Frame

::: {.callout-tip}
## New Helper Functions

These functions help you explore data frames:

- `nrow(df)`: counts the number of rows (observations)
- `ncol(df)`: counts the number of columns (variables)
- `names(df)`: shows the column names
- `str(df)`: shows the structure (data types of each column)
- `summary(df)`: shows summary statistics for each column
:::

```{r}
#| label: task2_1
#| eval: false

# How many rows (patients)?
nrow(___)  # Fill in the data frame name

# How many columns (variables)?
ncol(___)

# What are the column names?
names(___)

# View structure
str(___)

# View summary statistics
summary(___)
```

**Question**: How many patients are in the dataset? What is the average age?

::: {.callout-note icon=false}
## Your Answer

- The number of patients in the dataset is: ______
- The average age of the patients is: ______
:::

## Task 2.2: Accessing Data Frame Elements

```{r}
#| label: task2_2
#| eval: false

# Get the age column (two ways)
patients$___  # method 1: Dollar sign notation
patients[, "___"]  # method 2: Bracket notation

# Get the first patient (all columns)
patients[___, ]  # Fill in the row number

# Get patient 3's total charges (total_charges)
patients[___, "___"]  # Fill in row number and column name

# Get the first 3 patients
patients[___:___, ]  # Fill in the range
```

---

# Activity 3: Data Wrangling with dplyr {.activity-time data-time="20 min"}

## Generate Healthcare Dataset

Run this code to load a larger dataset for analysis:

```{r}
# Generate hospital performance data
practice_data <- read_csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week02_practice_data.csv")
# View first few rows
head(practice_data)
```

## Task 3.1: Select Columns

Choose specific columns from the dataset:

```{r}
#| label: task3_1_1
#| eval: false

# Select patient_id, age, and department
practice_data |>
  select(___, ___, ___)
```

```{r}
#| label: task3_1_2
#| eval: false

# Select all columns from patient_id to department
practice_data |>
  select(___:___)
```

```{r}
#| label: task3_1_3
#| eval: false

# Select all columns that start with "total"
practice_data |>
  select(starts_with("___"))
```

## Task 3.2: Filter Rows

Select specific patients based on criteria:

```{r}
#| label: task3_2_1
#| eval: false

# Find all patients age 65 or older
practice_data |>
  filter(___ >= ___)
```

```{r}
#| label: task3_2_2
#| eval: false

# Find all Emergency department patients
practice_data |>
  filter(___ == "___")
```

```{r}
#| label: task3_2_3
#| eval: false

# Find senior patients (age >= 65) in the Emergency department
practice_data |>
  filter(___ >= ___, ___ == "___")
```

```{r}
#| label: task3_2_4
#| eval: false

# Find patients in either Cardiology or Surgery
practice_data |>
  filter(department %in% c("___", "___"))
```

**Question**: How many patients are 65 or older? (Hint: count the rows in your filtered result)

::: {.callout-note icon=false}
## Your Answer

- The number of patients who are 65 or older is: ______
:::

## Task 3.3: Create New Columns with Mutate

Create new variables based on existing ones:

```{r}
#| label: task3_3_1
#| eval: false

# Create a cost_per_day column
practice_data |>
  mutate(cost_per_day = ___ / ___) |>
  select(patient_id, total_charges, los, cost_per_day)
```

```{r}
#| label: task3_3_2
#| eval: false

# Create age groups
practice_data |>
  mutate(
    age_group = case_when(
      age < ___ ~ "Under 40",
      age < ___ ~ "40-64",
      TRUE ~ "65+"
    )
  ) |>
  select(patient_id, age, age_group)
```

```{r}
#| label: task3_3_3
#| eval: false

# Flag high-cost patients (total charges > $25,000)
practice_data |>
  mutate(high_cost = ___ > ___) |>
  select(patient_id, total_charges, high_cost)
```

## Task 3.4: Sort with Arrange

Order your data:

```{r}
#| label: task3_4_1
#| eval: false

# Sort by age (youngest first)
practice_data |>
  arrange(___) |>
  select(patient_id, age, department)
```

```{r}
#| label: task3_4_2
#| eval: false

# Sort by total charges (highest first)
# Hint: use desc() for descending order
practice_data |>
  arrange(desc(___)) |>
  select(patient_id, total_charges)
```

```{r}
#| label: task3_4_3
#| eval: false

# Sort by department, then by age within each department
practice_data |>
  arrange(___, ___) |>
  select(patient_id, department, age)
```

## Task 3.5: Summarize Data

Calculate summary statistics:

```{r}
#| label: task3_5_1
#| eval: false

# Calculate overall statistics
practice_data |>
  summarize(
    total_patients = n(),
    mean_age = mean(___),
    mean_los = mean(___),
    mean_charges = mean(___),
    readmit_rate = mean(___)  # TRUE=1, FALSE=0, so mean gives proportion
  )
```

**Question**: What is the average age and average length of stay?

::: {.callout-note icon=false}
## Your Answer

- The average age is: ______
- The average length of stay is: ______
:::

## Task 3.6: Group By and Summarize

This is where dplyr becomes powerful! Calculate statistics for each group:

```{r}
#| label: task3_6_1
#| eval: false

# Statistics by department
practice_data |>
  group_by(___) |>
  summarize(
    patient_count = n(),
    mean_age = mean(___),
    mean_los = mean(___),
    mean_charges = mean(___),
    readmit_rate = mean(___)
  )
```

**Question**: Which department has the highest average charges? Which has the highest readmission rate?

::: {.callout-note icon=false}
## Your Answer

- The department that has the highest average total charges is: ______
- The department that has the highest 30-day readmission rate is: ______
:::

```{r}
#| label: task3_6_2
#| eval: false

# Statistics by insurance type
practice_data |>
  group_by(___) |>
  summarize(
    patient_count = n(),
    mean_age = round(mean(___), 1),
    mean_charges = round(mean(___), 2)
  ) |>
  arrange(desc(mean_charges))
```

## Task 3.7: Combining Multiple Operations

Put it all together! Create a complex analysis pipeline:

```{r}
#| label: task3_7_1
#| eval: false

# Analyze high-cost senior patients by department
practice_data |>
  filter(___ >= 65) |>  # Seniors only
  mutate(high_cost = total_charges > 25000) |>  # Flag expensive cases
  group_by(___, ___) |>  # Group by department and high_cost
  summarize(
    count = n(),
    avg_los = round(mean(___), 1),
    .groups = "drop"
  ) |>
  arrange(___, desc(___))  # Sort by department, then by high_cost
```

**Question**: What insight does this analysis provide about senior patients?

::: {.callout-note icon=false}
## Your Answer

:::

---

# Summary and Reflection

## What You Practiced Today

✓ Creating variables with different data types
✓ Working with vectors and data frames
✓ Using dplyr verbs: `select()`, `filter()`, `mutate()`, `arrange()`, `summarize()`
✓ Grouping data with `group_by()`
✓ Combining operations with the pipe `|>`
✓ Creating a Quarto report

## Key Concepts to Remember

::: {.callout-important}
## Essential Takeaways

1. **The pipe `|>` makes code readable** - chain operations together
2. **`group_by()` + `summarize()` is powerful** - analyze by groups
3. **Comments explain your code** - use `#` liberally
4. **Quarto makes work reproducible** - code + narrative + output
5. **Practice makes perfect** - the more you code, the easier it gets!
:::

---

# Solutions

Solutions are provided below. Try to complete the exercises independently before checking!

## Activity 1 Solutions

### Task 1.1

```{r}
patient_name <- "John Doe"
patient_age <- 45
is_admitted <- TRUE
department <- "ER"

class(patient_name)    # "character"
class(patient_age)     # "numeric"
class(is_admitted)     # "logical"
class(department)      # "character"
```

### Task 1.2

```{r}
ages <- c(45, 62, 33, 58, 71)
departments <- c("ER", "ICU", "Surgery", "ER", "Cardiology")

mean(ages)  # 53.8
min(ages)   # 33
max(ages)   # 71
```

### Task 1.3

```{r}
ages[1]      # 45
ages[3]      # 33
ages[1:3]    # 45 62 33
ages[c(2, 4)]  # 62 58
```

## Activity 2 Solutions

### Task 2.1

```{r}
nrow(patients)    # 10
ncol(patients)    # 7
names(patients)
str(patients)
summary(patients)
```

### Task 2.2

```{r}
patients$age
patients[, "age"]
patients[1, ]
patients[3, "total_charges"]
patients[1:3, ]
```

## Activity 3 Solutions

### Task 3.1

```{r}
practice_data |>
  select(patient_id, age, department)

practice_data |>
  select(patient_id:department)

practice_data |>
  select(starts_with("total"))
```

### Task 3.2

```{r}
practice_data |>
  filter(age >= 65)

practice_data |>
  filter(department == "Emergency")

practice_data |>
  filter(age >= 65, department == "Emergency")

practice_data |>
  filter(department %in% c("Cardiology", "Surgery"))
```

### Task 3.3

```{r}
practice_data |>
  mutate(cost_per_day = total_charges / los) |>
  select(patient_id, total_charges, los, cost_per_day)

practice_data |>
  mutate(
    age_group = case_when(
      age < 40 ~ "Under 40",
      age < 65 ~ "40-64",
      TRUE ~ "65+"
    )
  ) |>
  select(patient_id, age, age_group)

practice_data |>
  mutate(high_cost = total_charges > 25000) |>
  select(patient_id, total_charges, high_cost)
```

### Task 3.4

```{r}
practice_data |>
  arrange(age) |>
  select(patient_id, age, department)

practice_data |>
  arrange(desc(total_charges)) |>
  select(patient_id, total_charges)

practice_data |>
  arrange(department, age) |>
  select(patient_id, department, age)
```

### Task 3.5

```{r}
practice_data |>
  summarize(
    total_patients = n(),
    mean_age = mean(age),
    mean_los = mean(los),
    mean_charges = mean(total_charges),
    readmit_rate = mean(readmitted_30day)
  )
```

### Task 3.6

```{r}
practice_data |>
  group_by(department) |>
  summarize(
    patient_count = n(),
    mean_age = mean(age),
    mean_los = mean(los),
    mean_charges = mean(total_charges),
    readmit_rate = mean(readmitted_30day)
  )

practice_data |>
  group_by(insurance) |>
  summarize(
    patient_count = n(),
    mean_age = round(mean(age), 1),
    mean_charges = round(mean(total_charges), 2)
  ) |>
  arrange(desc(mean_charges))
```

### Task 3.7

```{r}
practice_data |>
  filter(age >= 65) |>
  mutate(high_cost = total_charges > 25000) |>
  group_by(department, high_cost) |>
  summarize(
    count = n(),
    avg_los = round(mean(los), 1),
    .groups = "drop"
  ) |>
  arrange(department, desc(high_cost))
```

