---
title: "Penalized Regression in Healthcare Analytics (Lecture Notes)"
subtitle: " "
author: ""
Week: "Week 14"
date: 01/01/2026
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: false
    code-copy: true
    theme: custom-theme.scss
    highlight-style: github
    mainfont: "Clear Sans, -apple-system, system-ui, sans-serif"
    fontsize: 1.1em
    grid:
      sidebar-width: 200px
      body-width: 900px
      margin-width: 300px
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Install required packages if needed
required_packages <- c("glmnet", "dplyr", "ggplot2", "tidyr", "caret")
new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

# Load required packages
library(glmnet)    # Penalized regression
library(dplyr)     # Data manipulation
library(ggplot2)   # Visualization
library(tidyr)     # Data reshaping
library(caret)     # Model training utilities
library(gridExtra) # Multiple plots

# Set seed for reproducibility
set.seed(2026)
```

# Introduction and Learning Objectives

## Prediction vs. Explanation: A Critical Distinction

In previous courses, we learned to use regression to **explain** why things happen (e.g., understand which factors cause higher costs or worse outcomes). This module takes a different approach: we focus on **prediction**, how we can more accurately predict outcomes, rather than explanation.

**What's the difference?**

- **Explanation**: Understanding *why* certain patients cost more (e.g., "Does smoking cause higher costs?")
- **Prediction**: Forecasting *how much* a patient will cost (regardless of why)

When explaining, we care about individual variables and their effects. When predicting, we only care about **accuracy** (e.g., how well we can forecast outcomes for new patients). The combination of variables matters more than understanding each one.

## Why Prediction Matters in Healthcare

Healthcare managers routinely need accurate forecasts:

- **Budget planning**: How much will our patient panel cost next year?
- **Resource allocation**: Which patients need care management outreach?
- **Readmission prevention**: Who is at highest risk of coming back to the hospital?

Traditional regression often fails at prediction when we have many potential factors to consider, like age, diagnoses, medications, prior utilization, and more. With dozens or hundreds of variables, standard methods either break down mathematically or produce unreliable forecasts.

**Penalized regression** addresses this by penalizing overly complex models: it automatically reduces the influence of less important variables (sometimes to zero), keeping only what matters for prediction. 

## Learning Objectives

By the end of this module, you will be able to:

1. **Distinguish** between prediction and explanation, and recognize when each approach is appropriate for healthcare management decisions

2. **Explain** why standard regression struggles with prediction when you have many potential factors, and how penalized regression addresses this challenge

3. **Compare** the three main types of penalized regression (Ridge, LASSO, and Elastic Net) and choose the right method for different situations

4. **Build** penalized regression models in R to predict healthcare outcomes like costs and readmissions

5. **Interpret** prediction results appropriately, understanding what the model tells you (future outcomes) and what it doesn't tell you (causal relationships)

---

# Conceptual Foundations

This section provides the theoretical foundation you need to understand penalized regression. We'll start with a review of ordinary least squares regression, identify its limitations, and then see how adding a penalty term addresses these limitations.

## Review: Ordinary Least Squares (OLS) Regression

Before diving into penalized methods, let's recall how standard linear regression works. This review will help us understand what we're modifying and why.

### The Basic Setup

In linear regression, we want to predict a continuous outcome variable $Y$ (e.g., healthcare costs) using a set of predictor variables $X_1, X_2, \ldots, X_p$ (e.g., age, diagnoses, prior utilization). We assume the relationship takes the form:

$$\hat{Y} = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_p X_p$$

Where:

- $\hat{Y}$ is our predicted value
- $\beta_0$ is the intercept (predicted value when all X's are zero)
- $\beta_1, \beta_2, \ldots, \beta_p$ are coefficients that tell us how much $Y$ changes for a one-unit change in each predictor

**Note on terminology**: Throughout this document, we use the terms "predictor," "variable," and "feature" interchangeably to refer to the X variables used to predict the outcome.

### How OLS Finds the Best Coefficients

OLS finds the coefficient values that minimize the **Sum of Squared Errors (SSE)**, the total squared difference between what we predicted and what actually happened:

$$\text{Minimize: } SSE = \sum_{i=1}^{n} (Y_i - \hat{Y}_i)^2$$

We square the prediction errors (so positive and negative errors don't cancel out), add them up across all observations, and find the intercept and coefficients that make this sum as small as possible.

This approach works excellently when:

- We have many more observations than predictors ($n >> p$)
- Predictors are not highly correlated with each other
- Our primary goal is to understand relationships (statistical inference)

However, modern healthcare analytics often violates these conditions, leading to several major problems.

##  Problems with Standard Regression in Healthcare

1. Related variables create confusion

- Healthcare factors naturally cluster together. For example, diabetes patients often have high blood pressure and obesity, patients with many doctor visits also have many prescriptions, etc.

- When variables are related, standard regression can't determine which one truly drives the outcome. This leads to unstable results: coefficients can become wildly inflated, have wrong signs, or change dramatically with small data changes. 


2. Too many variables, too few observations

- Healthcare data often has hundreds of potential predictors (diagnosis codes, medications, lab tests) but limited patients. For example, a hospital might have 300 patients but 250 potential predictor variables (or even more predictors than patients)

- The result: the model "memorizes" your specific patients rather than learning general patterns (overfitting). It predicts your current patients perfectly but fails miserably on new patients.


3. Hard to identify what factors are really important

- Standard regression uses *every* variable you give it, even irrelevant ones. You end up with all hundreds of variables in your model.

- Healthcare organizations need interpretable, actionable models (clinicians can't use hundreds of factors to make decisions).



## The Solution: Penalized Regression

### The Key Insight

Those problems above stem from too much flexibility (meaning OLS can set coefficients to any value to minimize training error). A potential solution is to **constrain** the coefficients, preventing them from becoming too large.

We do this by modifying the OLS objective function:

$$\text{Minimize: } SSE + \lambda \times \text{Penalty}$$

*In plain language, find coefficients that balance [prediction accuracy] + [penalty for model complexity]*

Where:

- **SSE** = Sum of Squared Errors (how well the model fits the training data)
- **Penalty** = A function of the coefficient sizes (how complex the model is)
- **λ (lambda)** = A tuning parameter that controls how much we penalize complexity

### How the Penalty Helps

**Why does the penalty improve predictions?** Without constraints, standard regression overfits—it "memorizes" the quirks of your specific patients rather than learning general patterns. The penalty prevents this by keeping the model simple enough to work well on new patients. Think of it like studying concepts versus memorizing answers: concepts work on new problems, memorization doesn't.

The penalty term creates a tradeoff:

- **Making SSE smaller** (better fit) typically requires larger, more complex coefficients
- **Making the penalty smaller** requires smaller coefficients
- **The optimal solution balances** these competing goals: slightly worse fit on your current data in exchange for better predictions on new patients

**How this addresses each problem**:

- **Handles correlated variables**: When variables are related (like diabetes, obesity, and blood pressure), the penalty prevents coefficients from becoming wildly inflated. 

- **Prevents overfitting with many variables**: The penalty stops the model from "memorizing" your specific patients. By shrinking coefficients, it forces the model to focus only on the strongest, most reliable patterns that will generalize to new patients.

- **Automatic variable selection (LASSO)**: The L1 penalty pushes weak predictors to exactly zero, eliminating them from the model entirely. You end up with a small set of important variables instead of hundreds.


### The Role of Lambda (λ)

The tuning parameter λ controls how much we care about the penalty:

- **λ = 0**: No penalty, model equals OLS
- **Small λ**: Light penalty, close to OLS
- **Large λ**: Heavy penalty, coefficients shrunk toward zero
- **λ → ∞**: All coefficients forced to zero (except intercept)

We need to choose λ carefully (too small and we don't solve our problems; too large and we throw away useful information). We'll discuss how to select λ using cross-validation later.


# Penalized Regression Models 
## Ridge Regression: The L2 Penalty

Now let's examine the first specific type of penalized regression: Ridge regression. This method addresses multicollinearity and overfitting by shrinking all coefficients toward zero.

### Mathematical Formulation

Ridge regression adds a penalty based on the **squared values** of coefficients:

$$\text{Minimize: } \sum_{i=1}^{n} (Y_i - \hat{Y}_i)^2 + \lambda \sum_{j=1}^{p} \beta_j^2$$

The penalty term $\sum_{j=1}^{p} \beta_j^2$ is called the **L2 penalty** or **squared penalty** because we're summing squared coefficients.


**What this formula says:**

"Find the coefficients that make predictions accurate (small SSE) while also keeping coefficients small (small sum of squared betas). The parameter λ controls how much we care about keeping coefficients small."


### How Ridge Regression Works

Ridge regression works by gently discouraging large coefficients. Instead of allowing the model to rely too heavily on any single predictor, it nudges all coefficients toward smaller values. This has several important consequences.

1. Ridge regression shrinks all coefficients toward zero, but does not remove variables.

As the penalty becomes stronger, coefficient values get smaller and smaller. However, they do not drop all the way to zero. This means Ridge regression keeps all predictors in the model, but limits how strongly each one can influence the outcome. 

2. Larger coefficients are shrunk more than smaller ones.

Ridge regression is especially uncomfortable with very large coefficients. If one predictor has a much larger effect than others, the penalty pushes it down more strongly. Smaller coefficients are affected less. This helps prevent the model from overreacting to noise or peculiarities in the data.

3. Ridge regression handles correlated predictors smoothly.

When predictors are highly correlated, ordinary regression often struggles to decide which one deserves credit, leading to unstable or counterintuitive estimates. Ridge regression avoids this problem by sharing the effect across correlated variables. Instead of picking one variable and inflating its coefficient, it assigns moderate coefficients to both, producing more sensible and reliable results.


### Visual Understanding: The Constraint Region

One of the best ways to understand Ridge regression is geometrically. Imagine we have just two predictors, so we're looking for β₁ and β₂.

![[LASSO and Ridge Penalty](https://towardsdatascience.com/courage-to-learn-ml-unraveling-l1-l2-regularization-part-2-1bb171e43b35/)](https://towardsdatascience.com/wp-content/uploads/2023/11/1VdcJFf5CEOsXojxHG9b9Yw.jpeg) 


**Understanding the figure:**

- **Red ellipses**: Contour lines of SSE. The center is where SSE is minimized (the OLS solution). Moving away from the center increases SSE.

- **Blue shaded region**: The constraint region - all combinations of β₁ and β₂ that satisfy the penalty constraint. For Ridge, this is a circle because we're constraining $\beta_1^2 + \beta_2^2 \leq c$ (equation of a circle).

- **The solution**: We want to be as close to the OLS center as possible (small SSE) while staying within the constraint region (small penalty). The Ridge solution is where the (smallest) ellipse touches the circle.

**Key insight**: Because the constraint region is a circle, the solution rarely lands exactly on an axis. This means Ridge keeps all variables with non-zero coefficients.

### Advantages of Ridge Regression

**1. Handles correlated predictors well**

When predictors overlap heavily, Ridge regression avoids extreme or unstable estimates. Instead of forcing the model to “choose sides,” it spreads influence across related variables in a balanced way.

**2. Reduces coefficient instability from multicollinearity**

When predictors are correlated, ordinary regression produces unstable coefficient estimates that can change dramatically with small data changes. Ridge regression stabilizes these estimates by limiting how large they can grow.

**3. Improves prediction accuracy on new data**

By preventing overfitting to training data, Ridge regression often achieves better prediction performance on new, unseen patients compared to ordinary regression.

**4. Keeps all relevant information**

If most predictors are believed to matter at least a little, Ridge regression retains them all while preventing any single variable from dominating.

### Disadvantages of Ridge Regression

**1. Does not simplify the model**

Ridge regression keeps every predictor in the model. When there are many variables, this can make the model hard to summarize or deploy.

**2. Limited interpretability**

Because effects are spread across many small coefficients, it is difficult to identify which predictors are truly important. Ridge is better suited for prediction than explanation.

**3. Retains noisy predictors**

Variables with no real signal are not removed, only weakened. When many such variables are present, their combined influence can reduce performance.

### When to Use Ridge Regression

Ridge regression is best when:

- Predictors are strongly correlated
- Most variables are expected to carry some information
- Prediction is the main goal (interpretation is less important)
- The number of predictors is moderate rather than very large


## LASSO Regression: The L1 Penalty

LASSO (Least Absolute Shrinkage and Selection Operator) is the other major penalized regression method. Its key advantage over Ridge is **automatic feature selection**: LASSO can eliminate variables entirely by setting their coefficients exactly to zero.

### Mathematical Formulation

LASSO adds a penalty based on the **absolute values** of coefficients:

$$\text{Minimize: } \sum_{i=1}^{n} (Y_i - \hat{Y}_i)^2 + \lambda \sum_{j=1}^{p} |\beta_j|$$

The penalty term $\sum_{j=1}^{p} |\beta_j|$ is called the **L1 penalty** or **absolute value penalty**.

The critical difference from Ridge is that LASSO uses absolute values instead of squared values, which changes the geometry in a way that allows coefficients to become exactly zero.

### How LASSO Achieves Feature Selection

LASSO’s key feature selection property comes from its constraint shape (see the figure above).
The LASSO constraint has sharp corners along the axes. When the model searches for the best fit, it often lands on one of these corners—where one or more coefficients are exactly zero.

In contrast, Ridge regression uses a smooth, round constraint. This makes it unlikely for coefficients to be driven all the way to zero, so all predictors remain in the model.


### Advantages of LASSO Regression

**1. Automatic feature selection**

LASSO identifies a small set of important predictors and removes the rest, making it especially useful when many variables are available but only some are informative.

**2. Interpretability**

With fewer active predictors, the model’s logic is easier to understand and explain in practical terms.

**3. Effective in high-dimensional settings**

LASSO works even when the number of predictors exceeds the number of observations, which is common with diagnostic codes, procedures, or lab measures.

**4. Controls overfitting**

By excluding noisy predictors, LASSO reduces the risk of capturing patterns that do not generalize.

### Disadvantages of LASSO Regression

**1. Unstable with correlated predictors**

When predictors are highly correlated, LASSO tends to select one and drop the others. Which variable is chosen can vary across samples, making interpretation fragile.

**2. May discard useful information**

By setting coefficients exactly to zero, LASSO can remove predictors that still carry some signal. In purely predictive tasks, Ridge may perform slightly better.

**3. Selection is limited in very small samples**

When the sample size is small, LASSO can only select a limited number of predictors, potentially missing relevant factors.


### When to Use LASSO Regression

LASSO is best when:

- Identifying key predictors is a primary goal
- Simplicity and interpretability matters
- You have many potential predictors and need feature selection
- Predictors are not highly correlated 


## Elastic Net

Elastic Net combines the strengths of both Ridge and LASSO by using both penalties together. It's often the best choice in practice because it handles correlated variables better than LASSO while still performing feature selection.

### Mathematical Formulation

Elastic Net minimizes:

$$\text{Minimize: } \sum_{i=1}^{n} (Y_i - \hat{Y}_i)^2 + \lambda \left[ \alpha \sum_{j=1}^{p} |\beta_j| + (1-\alpha) \sum_{j=1}^{p} \beta_j^2 \right]$$

Where:

- **α (alpha)** controls the mix of L1 and L2 penalties
- **α = 1**: Pure LASSO
- **α = 0**: Pure Ridge
- **0 < α < 1**: Elastic Net blend

### How Elastic Net Handles Correlated Variables

When predictors are strongly correlated, LASSO tends to select one and discard the rest, even if they all carry meaningful information. Ridge takes the opposite approach, keeping everything and shrinking coefficients together. Elastic Net sits between these two.

The key idea is the grouping effect. Elastic Net tends to keep or remove groups of related variables together, rather than making arbitrary choices among them. The Ridge component encourages correlated predictors to stay together, while the LASSO component removes predictors (or groups) that contribute little to prediction.

For example, imagine you include several variables that capture a patient's overall health status in different ways (e.g., past diagnoses, prior hospital use, overall medical complexity, etc.) along with many unrelated predictors. 

- LASSO might keep one of these health measures and drop others
- Ridge would keep all measures, including many irrelevant ones
- Elastic Net would likely keep all of the health measures together, while removing unrelated noise variables

### Tuning Elastic Net

Elastic Net balances two competing goals: simplicity and stability. 
It uses two parameters two control this balance. 

1. **λ (lambda)**: How strongly coefficients are penalized overall
2. **α (alpha)**: How much the model behaves like LASSO vs. Ridge

In practice, the model tries different balances and selects the one that performs best on new data. Rather than committing to LASSO or Ridge upfront, Elastic Net lets the data decide how much feature selection and how much stabilization are needed.


### Advantages of Elastic Net

1. Combines feature selection and stability 

Elastic Net can remove unimportant predictors while keeping correlated, meaningful ones together.

2. More stable than LASSO

When predictors overlap, Elastic Net produces more consistent selections across samples, reducing the risk of arbitrary choices.

3. Effective in very high dimensions

Unlike LASSO, Elastic Net can retain more predictors when the number of variables far exceeds the number of observations.

4. Highly adaptable

By adjusting the balance between LASSO and Ridge, Elastic Net can be tailored to different data structures and research goals.


### Disadvantages of Elastic Net

1. More tuning required

Elastic Net introduces an extra decision (parameter) compared to Ridge or LASSO, which significantly increases complexity. 

2. Same modeling assumptions as linear regression

Elastic Net improves estimation and selection, but it does not address nonlinearity or misspecified relationships.

3. Not always necessary

If predictors are largely independent, LASSO alone may suffice. If feature selection is unnecessary, Ridge alone may be simpler.


### When to Use Elastic Net

Elastic Net is often a strong default choice, especially when: 

- Predictors are correlated and feature selection is desired
- You are not sure whether Ridge or LASSO is more appropriate
- The data are high-dimensional
- Stability in selected predictors matters

Elastic Net is often the best starting choice because it combines Ridge's stability with LASSO's feature selection. If you're unsure which method to use, start with Elastic Net and let cross-validation determine the optimal balance between Ridge and LASSO.


## Comparing the Three Methods

Let's summarize the key differences to help you choose the right method for your healthcare analytics problem.

### Side-by-Side Comparison

| Criterion | Ridge | LASSO | Elastic Net |
|-----------|-------|-------|-------------|
| **Penalty** | L2 (squared) | L1 (absolute) | Both L1 and L2 |
| **Feature selection** | No - keeps all variables | Yes - zeros out variables | Yes - zeros out variables |
| **Correlated predictors** | Excellent - shrinks together | Poor - arbitrary selection | Good - groups together |
| **Works when p > n** | Yes | Yes (selects ≤ n) | Yes (can select > n) |
| **Number of tuning parameters** | 1 (λ) | 1 (λ) | 2 (λ and α) |
| **Interpretability** | Low (many variables) | High (sparse model) | Moderate |
| **Best for** | Correlated predictors, all relevant | Feature selection, uncorrelated | General use, correlated with selection |



## Choosing Lambda: Cross-Validation

We need to choose the right λ value, but how?

**The challenge:** We can't just pick the λ that works best on our current patients. That would be like a student memorizing practice test answers. We need to know which λ will work best on *new* patients we haven't seen yet.

**The solution: Cross-validation**

Think of cross-validation as creating multiple "practice tests":

1. **Divide your data into 10 groups** (called "folds"; the number of groups can vary)
2. **For each λ value you want to try:**
   - Train the model on 9 groups, test on the 10th group
   - Repeat 10 times, each time holding out a different group
   - Average the prediction errors across all 10 tests
3. **Pick the λ with the lowest average error**

**Why this works:** Every patient gets used for both training AND testing (just not at the same time). This efficiently uses all your data to estimate how well each λ will perform on new patients.

**In practice:** The software (like R's `glmnet` package) does this automatically. You'll see which λ performs best, and we'll discuss how to interpret the results in the hands-on section.

---

# Hands-On Implementation with R

Now that you understand the theory behind penalized regression, let's apply these methods to a realistic healthcare analytics problem. In this section, you'll learn how to implement Ridge, LASSO, and Elastic Net in R and interpret the results.

**What You'll Learn in This Section:**

- How to fit Ridge, LASSO, and Elastic Net models in R
- Use cross-validation to select optimal penalty parameters
- Interpret coefficient paths and model comparisons
- Evaluate prediction performance on test data

## The Case: Predicting Patient Healthcare Costs

**Problem:**

Imagine you work for a health insurance company. Every year, your company needs to predict how much each patient will cost in healthcare spending so you can:

- Set appropriate insurance premiums
- Budget for the coming year
- Identify high-cost patients who might benefit from care management programs (like helping diabetics control their blood sugar to avoid expensive complications)

**Your Task:**

Build a model that predicts next year's total healthcare spending for each patient based on information you currently have about them.

**Available Data:**

You have data on 500 patients, including:

- **Demographics**: Age, sex
- **Socioeconomic factors**: Income level, insurance status
- **Health behaviors**: Smoking, BMI, exercise frequency
- **Chronic conditions**: Diabetes, hypertension, heart disease, COPD (chronic obstructive pulmonary disease; a chronic lung disease), depression, arthritis
- **Prior healthcare use**: Emergency room visits, hospitalizations, doctor visits, number of medications

- Outcome variable is annual healthcare cost (the total cost of care; the insurance company pays out to hospitals, doctors, and other healthcare providers for a patient's care over a year).

**The Challenge:**

With so many potential predictors (some of which are related to each other), standard regression may struggle. We'll use penalized regression to build a model that predicts well on new patients.

---

## Creating a Healthcare Cost Prediction Dataset

For this tutorial, we'll create a realistic simulated dataset representing this scenario.

```{r}
#| label: create-data
#| code-fold: true

# Number of patients in our dataset
n <- 500

# Generate patient characteristics
healthcare_data <- data.frame(
  patient_id = 1:n,

  # Demographics
  age = round(rnorm(n, mean = 55, sd = 15)),
  female = rbinom(n, 1, 0.52),

  # Socioeconomic factors
  income_level = sample(1:5, n, replace = TRUE,
                        prob = c(0.15, 0.20, 0.30, 0.20, 0.15)),
  has_insurance = rbinom(n, 1, 0.88),

  # Health behaviors
  smoker = rbinom(n, 1, 0.18),
  bmi = round(rnorm(n, mean = 28, sd = 6), 1),
  exercise_days = rpois(n, lambda = 2),

  # Chronic conditions - these will be major cost drivers
  diabetes = rbinom(n, 1, 0.12),
  hypertension = rbinom(n, 1, 0.30),
  heart_disease = rbinom(n, 1, 0.08),
  copd = rbinom(n, 1, 0.06),
  depression = rbinom(n, 1, 0.15),
  arthritis = rbinom(n, 1, 0.22),

  # Healthcare utilization from prior year
  ed_visits = rpois(n, lambda = 0.5),
  hospitalizations = rpois(n, lambda = 0.15),
  physician_visits = rpois(n, lambda = 4),
  num_medications = rpois(n, lambda = 3)
)

# Make chronic conditions more realistic by adding correlations
healthcare_data <- healthcare_data %>%
  mutate(
    diabetes = ifelse(bmi > 30 & age > 50,
                      rbinom(n(), 1, 0.25), diabetes),
    hypertension = ifelse(diabetes == 1,
                          rbinom(n(), 1, 0.65), hypertension),
    heart_disease = ifelse(diabetes == 1 & hypertension == 1 & age > 60,
                           rbinom(n(), 1, 0.30), heart_disease)
  )

# Generate the outcome variable: Annual healthcare costs
healthcare_data <- healthcare_data %>%
  mutate(
    annual_cost =
      # Base cost
      2000 +
      # Demographics
      age * 50 +
      (1 - has_insurance) * 1500 +
      # Health behaviors
      smoker * 2000 +
      pmax(bmi - 25, 0) * 100 +
      # Chronic conditions
      diabetes * 5000 +
      hypertension * 2000 +
      heart_disease * 8000 +
      copd * 6000 +
      depression * 1500 +
      arthritis * 1000 +
      # Prior utilization
      ed_visits * 1500 +
      hospitalizations * 15000 +
      physician_visits * 200 +
      num_medications * 500 +
      # Random noise
      rnorm(n(), 0, 3000)
  ) %>%
  mutate(annual_cost = pmax(annual_cost, 500))

# Display first few rows
head(healthcare_data)
```

Let's examine the distribution of annual healthcare costs in our dataset:

```{r}
#| label: data-summary

# Summary statistics
summary(healthcare_data$annual_cost)

# Visualize distribution
ggplot(healthcare_data, aes(x = annual_cost)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.8) +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Distribution of Annual Healthcare Costs",
    subtitle = paste0("n = ", n, " patients | Mean = $",
                      format(round(mean(healthcare_data$annual_cost)), big.mark = ",")),
    x = "Annual Cost ($)",
    y = "Number of Patients"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

The distribution shows that most patients have moderate costs (around $8,000-$18,000), but some have very high costs. This right-skewed pattern is typical of healthcare data: most people use healthcare moderately, but a small number have very expensive care (major surgeries, long hospitalizations, etc.).

## Data Preparation for glmnet

Before we can use penalized regression, we need to prepare our data in a specific format and split it into training and test sets.

Let's use 70% of the data to build the model (training set) and 30% to evaluate how well the model predicts on new patients (test set; data not used for training the model). Within the training set, we perform 10-fold cross-validation to choose λ.


```{r}
#| label: data-prep

# Split into training (70%) and test (30%) sets
train_index <- sample(1:n, size = 0.7 * n)
train_data <- healthcare_data[train_index, ] # Training Dataset
test_data <- healthcare_data[-train_index, ] # Test Dataset

# Define predictor variables
predictor_vars <- setdiff(names(healthcare_data), c("patient_id", "annual_cost"))

# Create matrix format required by glmnet
x_train <- as.matrix(train_data[, predictor_vars])
y_train <- train_data$annual_cost

x_test <- as.matrix(test_data[, predictor_vars])
y_test <- test_data$annual_cost
```

Our dataset now contains 350 patients (70%) in the training set and 150 (30%) in the test set. We have 17 predictor variables (age, sex, income, insurance status, health behaviors, chronic conditions, and prior healthcare use).

**Note**: The `glmnet` package requires predictors in matrix format (not a data frame) and automatically standardizes them before fitting.


## Baseline: Ordinary Least Squares (OLS) Regression

Before fitting penalized models, let's establish a baseline using OLS regression:

```{r}
#| label: ols-model

# Fit OLS model
ols_model <- lm(annual_cost ~ ., data = train_data[, c(predictor_vars, "annual_cost")])

# View summary
summary(ols_model)
```

Evaluate on test set:

```{r}
#| label: ols-performance

# Make predictions
ols_pred <- predict(ols_model, newdata = test_data)

# Calculate performance metrics
ols_mse <- mean((y_test - ols_pred)^2)
ols_rmse <- sqrt(ols_mse)
```

The OLS model's test set performance shows a Root Mean Squared Error (RMSE) that represents the average prediction error in dollars (OLS RMSE = `r round(ols_rmse, 2)`). This baseline performance will help us evaluate whether penalized regression methods offer improvements.

## Ridge Regression

Now let's fit a Ridge regression model using the L2 penalty.

### Fitting the Ridge Model

```{r}
#| label: ridge-fit

# Fit Ridge regression (alpha = 0)
ridge_model <- glmnet(
  x = x_train,
  y = y_train,
  alpha = 0
)
```

The `glmnet` package automatically tests a sequence of lambda values to explore different levels of penalty strength.


### Visualizing the Coefficient Paths

```{r}
#| label: ridge-paths
#| fig-width: 10
#| fig-height: 6

# Plot coefficient paths
plot(ridge_model, xvar = "lambda", label = TRUE)
title("Ridge Regression: Coefficient Paths", line = 3)
```

**Interpretation of the plot:**

- **X-axis**: -Log(λ) (negative log lambda). Left = large penalty (heavy shrinkage), Right = small penalty (close to OLS)
- **Y-axis**: Coefficient values
- **Each line**: One predictor variable
- **Key observation**: As we move left on the plot (larger λ, stronger penalty), all coefficients shrink toward zero, but none reach exactly zero

### Cross-Validation to Find Optimal Lambda

Let's find an optimal Lambda value by using a cross-validation approach. We will use a 10-fold cross-validation to evaluate different Lambda values and select the one that minimizes the prediction error.

```{r}
#| label: ridge-cv

# Perform 10-fold cross-validation
ridge_cv <- cv.glmnet(
  x = x_train,
  y = y_train,
  alpha = 0,
  nfolds = 10 # how many folds will be used
)

# Plot CV results
plot(ridge_cv)
title("Ridge Regression: Cross-Validation Results", line = 3)
```

The cross-validation plot shows prediction error (vertical axis) for different lambda values (horizontal axis). The two vertical dashed lines mark the two lambda options that cross-validation identifies.

**Understanding the two lambda choices:**

Cross-validation provides two options:

- **lambda.min**: The λ with the absolute lowest prediction error (best accuracy)
- **lambda.1se**: Uses the "1 standard error rule": selects the largest λ (simplest model) whose prediction error is within 1 standard error of the minimum. This provides a simpler, more robust model with negligible loss in accuracy.

Let's use **lambda.1se** as it is often preferred in practice. It produces simpler models that are less likely to overfit and more likely to generalize well to new patients.


### Examining Ridge Coefficients

```{r}
#| label: ridge-coef

# Extract coefficients at lambda.1se
ridge_coef <- coef(ridge_cv, s = "lambda.1se")
ridge_coef

# All coefficients are non-zero
n_nonzero_ridge <- sum(ridge_coef[-1] != 0)
n_nonzero_ridge
```

Notice that Ridge regression keeps all variables in the model (i.e., every coefficient is non-zero). Ridge shrinks the coefficients toward zero but never eliminates variables entirely.


### Ridge Test Set Performance

Let's get the RMSE of the Ridge model using the test dataset. 

```{r}
#| label: ridge-performance

# Make predictions
ridge_pred <- predict(ridge_cv, s = "lambda.1se", newx = x_test)

# Calculate metrics
ridge_mse <- mean((y_test - ridge_pred)^2)
ridge_rmse <- sqrt(ridge_mse)

# Compare to OLS
improvement <- (ols_rmse - ridge_rmse) / ols_rmse * 100
```

Ridge regression's test set RMSE (`r round(ridge_rmse, 2)`) shows whether the penalty improves prediction accuracy compared to OLS. The improvement percentage (`r round(improvement, 2)`%) indicates how much better Ridge performs on new, unseen patients (untrained data).

## LASSO Regression

Now let's fit a LASSO model using the L1 penalty for automatic feature selection.

### Fitting the LASSO Model

```{r}
#| label: lasso-fit

# Fit LASSO regression (alpha = 1)
lasso_model <- glmnet(
  x = x_train,
  y = y_train,
  alpha = 1
)
```

### Visualizing LASSO Coefficient Paths

```{r}
#| label: lasso-paths
#| fig-width: 10
#| fig-height: 6

# Plot coefficient paths
plot(lasso_model, xvar = "lambda", label = TRUE)
title("LASSO Regression: Coefficient Paths", line = 3)
```

**Key observation**: Unlike Ridge, the LASSO paths hit zero! As we move left on the plot (larger λ, stronger penalty), variables are progressively eliminated. This is feature selection.


### Cross-Validation for LASSO

Same as our Ridge demonstration, let's find an optimal lambda value for our LASSO model using a 10-fold cross-validation approach.

```{r}
#| label: lasso-cv

# Perform cross-validation
lasso_cv <- cv.glmnet(
  x = x_train,
  y = y_train,
  alpha = 1,
  nfolds = 10
)

# Plot CV results
plot(lasso_cv)
title("LASSO Regression: Cross-Validation Results", line = 3)
```

The cross-validation plot for LASSO is similar to Ridge, but notice the numbers at the top—they show how many variables remain in the model at each lambda value. As we move left on the plot (larger λ, stronger penalty), LASSO eliminates more variables.


### LASSO Feature Selection

```{r}
#| label: lasso-coef

# At lambda.1se (simpler model)
lasso_coef_1se <- coef(lasso_cv, s = "lambda.1se")
lasso_coef_1se

# Count non-zero coefficients
n_nonzero_1se <- sum(lasso_coef_1se[-1] != 0)
n_nonzero_1se
```

LASSO's key feature is automatic variable selection. At lambda.1se (the more conservative choice), even fewer variables are selected, resulting in a simpler model.

### Visualizing Important Predictors

```{r}
#| label: lasso-important
#| fig-width: 8
#| fig-height: 6

# Extract non-zero coefficients
lasso_coef_df <- data.frame(
  variable = rownames(lasso_coef_1se),
  coefficient = as.vector(lasso_coef_1se)
) %>%
  filter(variable != "(Intercept)", coefficient != 0) %>%
  arrange(desc(abs(coefficient)))

# Display
lasso_coef_df

# Visualize
ggplot(lasso_coef_df, aes(x = reorder(variable, abs(coefficient)),
                          y = coefficient,
                          fill = coefficient > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(
    values = c("firebrick", "steelblue"),
    labels = c("Decreases Cost", "Increases Cost"),
    name = "Effect on Cost"
  ) +
  labs(
    title = "LASSO: Key Predictors of Healthcare Costs",
    subtitle = paste0("Variables with non-zero coefficients (", length(rownames(lasso_coef_df)),
                      " of ", length(predictor_vars), " selected)"),
    x = "",
    y = "Coefficient (Effect on Annual Cost)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

LASSO has identified the most important cost drivers. Hospitalizations has the largest coefficient, followed by chronic conditions like heart disease and COPD.

**Interpreting coefficient magnitudes**: These coefficients represent the change in annual cost (in dollars) for a one-unit change in each predictor, holding others constant. For example, if hospitalizations has a coefficient of approximately 15,000, this means each additional hospitalization is associated with roughly $15,000 higher annual costs.



### LASSO Test Set Performance

Let's test the performance of our LASSO model using the test dataset. 

```{r}
#| label: lasso-performance

# Make predictions
lasso_pred <- predict(lasso_cv, s = "lambda.1se", newx = x_test)

# Calculate metrics
lasso_mse <- mean((y_test - lasso_pred)^2)
lasso_rmse <- sqrt(lasso_mse)

# Compare to OLS
improvement_lasso <- (ols_rmse - lasso_rmse) / ols_rmse * 100
```

LASSO achieves competitive prediction accuracy (RMSE = `r round(lasso_rmse, 2)`, reduction of `r round(improvement_lasso, 2)`% compared to the OLS model) while using far fewer variables than OLS. This demonstrates the power of automatic feature selection (better predictions with simpler models).



## Elastic Net Regression

Finally, let's fit an Elastic Net model that combines Ridge and LASSO penalties.

### Tuning Alpha and Lambda

For the alpha parameter (weights of L1 and L2 regularization), we will use a grid search method (starting from 0, we will increase alpha by 0.1 every time). 

```{r}
#| label: elastic-net-tuning

# Define grid of alpha values
alpha_grid <- seq(0, 1, by = 0.1)

# Store results
elastic_results <- data.frame(
  alpha = alpha_grid,
  lambda_1se = NA,
  mse_1se = NA,
  n_features = NA
)

# Use consistent CV folds (10)
fold_id <- sample(1:10, size = length(y_train), replace = TRUE) 

# Test each alpha value
for (i in seq_along(alpha_grid)) {
  cv_fit <- cv.glmnet(
    x = x_train,
    y = y_train,
    alpha = alpha_grid[i],
    foldid = fold_id # instead of nfolds, we use foldid approach to have the same data in each fold 
  )

  elastic_results$lambda_1se[i] <- cv_fit$lambda.1se
  # Get MSE at lambda.1se
  idx_1se <- which(cv_fit$lambda == cv_fit$lambda.1se)
  elastic_results$mse_1se[i] <- cv_fit$cvm[idx_1se]

  coef_temp <- coef(cv_fit, s = "lambda.1se")
  elastic_results$n_features[i] <- sum(coef_temp != 0) - 1
}

# Display results
elastic_results

# Find best alpha
best_alpha <- elastic_results$alpha[which.min(elastic_results$mse_1se)]
best_alpha
```

The tuning results show the cross-validated MSE for each alpha value (ranging from 0=pure Ridge to 1=pure LASSO). The alpha with the lowest MSE represents the optimal balance between Ridge and LASSO penalties for this dataset (here, when alpha is `r best_alpha`).

### Visualizing Alpha Tuning

```{r}
#| label: elastic-net-viz
#| fig-width: 8
#| fig-height: 5
#| echo: false

# Plot MSE vs Alpha
ggplot(elastic_results, aes(x = alpha, y = mse_1se)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_vline(xintercept = best_alpha, linetype = "dashed", color = "firebrick") +
  annotate("text", x = best_alpha + 0.08, y = max(elastic_results$mse_1se),
           label = paste("Best α =", best_alpha), hjust = 0, color = "firebrick") +
  labs(
    title = "Elastic Net: Cross-Validated MSE vs Alpha",
    subtitle = "Alpha = 0 (Pure Ridge) to Alpha = 1 (Pure LASSO)",
    x = "Alpha (Mixing Parameter)",
    y = "Cross-Validated Mean Squared Error"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

### Fit Final Elastic Net Model

```{r}
#| label: elastic-net-final

# Fit with best alpha
elastic_cv <- cv.glmnet(
  x = x_train,
  y = y_train,
  alpha = best_alpha,
  nfolds = 10
)

# Extract coefficients
elastic_coef <- coef(elastic_cv, s = "lambda.1se")
n_features_elastic <- sum(elastic_coef != 0) - 1

# Make predictions
elastic_pred <- predict(elastic_cv, s = "lambda.1se", newx = x_test)

# Calculate metrics
elastic_mse <- mean((y_test - elastic_pred)^2)
elastic_rmse <- sqrt(elastic_mse)
elastic_rmse

# Compare to OLS
improvement_elastic <- (ols_rmse - elastic_rmse) / ols_rmse * 100
improvement_elastic
```

Elastic Net combines the strengths of Ridge and LASSO, achieving strong prediction accuracy while performing feature selection. The model uses the optimal alpha value identified through cross-validation.

## Comprehensive Model Comparison

Now let's compare all four models:

### Performance Summary Table

```{r}
#| label: model-comparison

# Create comparison table
comparison <- data.frame(
  Model = c("OLS", "Ridge", "LASSO", "Elastic Net"),
  RMSE = round(c(ols_rmse, ridge_rmse, lasso_rmse, elastic_rmse), 2),
  MSE = round(c(ols_mse, ridge_mse, lasso_mse, elastic_mse), 0),
  Features = c(length(predictor_vars),
               length(predictor_vars),
               length(rownames(lasso_coef_df)),
               n_features_elastic)
)

# Calculate improvement over OLS
comparison$Improvement = paste0(
  round((1 - comparison$RMSE / ols_rmse) * 100, 2), "%"
)

# Display
comparison
```

The comparison table shows test set performance metrics for all four models. Notice how penalized regression methods (Ridge, LASSO, and Elastic Net) typically outperform OLS, and LASSO/Elastic Net achieve this with fewer variables.

```{r}
#| label: model-comparison-plot
#| fig-width: 10
#| fig-height: 5
#| echo: false

# Visualize performance
comparison %>%
  ggplot(aes(x = reorder(Model, -RMSE), y = RMSE, fill = Model)) +
  geom_col() +
  geom_text(aes(label = paste0("$", format(round(RMSE), big.mark = ","))),
            vjust = -0.5, fontface = "bold") +
  geom_text(aes(label = paste0("(", Features, " vars)")),
            vjust = 1.5, color = "white", size = 3.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Model Comparison: Test Set RMSE",
    subtitle = "Lower RMSE indicates better prediction accuracy",
    x = "",
    y = "Root Mean Squared Error ($)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  ) +
  coord_cartesian(ylim = c(0, max(comparison$RMSE) * 1.15))
```

### Coefficient Comparison

```{r}
#| label: coef-comparison
#| fig-width: 12
#| fig-height: 8

# Extract all coefficients
coef_comparison <- data.frame(
  variable = rownames(ridge_coef),
  OLS = c(coef(ols_model)),
  Ridge = as.vector(ridge_coef),
  LASSO = as.vector(lasso_coef_1se),
  ElasticNet = as.vector(elastic_coef)
) %>%
  filter(variable != "(Intercept)") %>%
  pivot_longer(
    cols = -variable,
    names_to = "Model",
    values_to = "Coefficient"
  )

# Plot
ggplot(coef_comparison, aes(x = variable, y = Coefficient, fill = Model)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Coefficient Comparison Across Models",
    subtitle = "Note how LASSO zeros out less important variables while others keep all",
    x = "",
    y = "Coefficient Value"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
```

**Key observations:**

1. **OLS** has the largest coefficients (no shrinkage)
2. **Ridge** shrinks all coefficients but keeps them non-zero
3. **LASSO** eliminates several variables entirely (zero coefficients)
4. **Elastic Net** balances between Ridge and LASSO

---

# Summary and Key Takeaways

## When to Use Each Method

**Ridge Regression**:

- Use when you have correlated predictors and want to keep all variables
- Best for prediction when you believe most variables are relevant
- Handles multicollinearity well
- Not suitable when you need feature selection

**LASSO Regression**:

- Use when you need automatic feature selection
- Best for interpretability and sparse models
- Works well with many potential predictors
- Caution with highly correlated predictors (arbitrary selection)

**Elastic Net**:

- Often the best default choice
- Combines strengths of Ridge and LASSO
- Handles correlated predictors better than LASSO
- Performs feature selection like LASSO
- Requires tuning two parameters (α and λ)

## Cross-Validation Best Practices

1. **Always use cross-validation** to select λ (and α for Elastic Net)
2. **Use consistent folds** when comparing different α values
3. **Consider lambda.1se** for simpler, more interpretable models 
4. **Use lambda.min** when prediction accuracy is paramount
5. **Check the CV curve** - flat regions indicate room for simplification


## Important Caveats

1. **Prediction vs. Causation**: Penalized regression is for prediction, not causal inference. Selected variables are associated with outcomes, not necessarily causal drivers.

2. **Feature Selection Stability**: LASSO selections can be unstable, especially with correlated predictors. Run multiple samples or use Elastic Net.

3. **Clinical Validation**: Always validate that selected features make clinical sense. Statistical selection doesn't guarantee clinical relevance.

4. **Generalizability**: Models should be validated on external data from different time periods or institutions.

---

# References and Additional Resources

## Primary Methodological Papers

- **Tibshirani, R. (1996)**. Regression shrinkage and selection via the lasso. *Journal of the Royal Statistical Society Series B*, 58(1), 267-288.

- **Zou, H., & Hastie, T. (2005)**. Regularization and variable selection via the elastic net. *Journal of the Royal Statistical Society Series B*, 67(2), 301-320.

## Recommended Textbooks

- **James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021)**. *An Introduction to Statistical Learning with Applications in R* (2nd ed.). Springer.

- **Etzioni, R., et. al. (2020)**. *Statistics for Health Data Science*. Springer.

## R Package Documentation

- **glmnet package**: https://glmnet.stanford.edu/articles/glmnet.html

- **UC Business Analytics R Programming Guide**: [Regularized Regression](https://uc-r.github.io/regularized_regression)

