---
title: "Lab Assignment 2: t-tests (Week 05)"
subtitle: ""
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: false
    code-copy: true
    theme: 
      - cosmo
      - lecture.scss
    highlight-style: github
    mainfont: "Clear Sans, -apple-system, system-ui, sans-serif"
    fontsize: 1.1em
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(parameters)
```

# Assignment Overview

- **Total Points**: 10 points  
- **Due Date**: February 15, 2026
- **Submission**: Upload your completed .qmd and rendered html files to Canvas

## Learning Objectives

This assignment assesses your ability to:

- Identify the appropriate t-test for different research scenarios
- Conduct one-sample, independent-samples, and paired-samples t-tests using R
- Interpret t-test output including t-statistics and p-values
- Draw data-driven conclusions in a healthcare management context

## Instructions

1. Complete all 6 tasks below
2. Fill in the blanks (replace `___` with appropriate code or values)
3. After completing each code chunk, change `#| eval: false` to `#| eval: true` to run it
4. Answer all interpretation questions (1-2 sentences each)
5. Run all code chunks to verify they work before submitting

---

# About the Dataset

**Hospital Quality Improvement Study**

For this assignment, we will use data from a hospital quality improvement initiative. The dataset contains records from 40 patients **who were readmitted** across two departments, tracking clinical outcomes and the impact of a **Mindfulness-Based Stress Reduction (MBSR)** program on patient anxiety.

## Dataset Background

This dataset simulates a quality improvement study at a multi-department hospital system. It captures:

- **Readmission outcomes**: How quickly readmitted patients returned after discharge (compared to the 30-day national benchmark)
- **Departmental costs**: Treatment costs across Cardiology and Orthopedics
- **Intervention effectiveness**: Patient anxiety scores measured before and after an MBSR program integrating guided meditation, breathing exercises, and progressive relaxation

This type of data is commonly used by healthcare managers to evaluate clinical benchmarks, compare departmental performance, and assess the effectiveness of patient wellness programs.

## Loading and Preparing the Data

First, run the code below to load required packages. If you haven't installed these packages yet, run `install.packages("ggplot2")`, `install.packages("dplyr")`, and `install.packages("parameters")` in the R console.

```{r}
#| label: load-packages

# Load required packages
library(ggplot2)
library(dplyr)
library(parameters)

set.seed(2026)  # For reproducibility
```

Now, run the following code to load the dataset for this assignment:

```{r}
#| label: load-data

# Load the assignment dataset
qi_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/week05_assignment_data.csv")

# View first few rows
head(qi_data, 10)
```

## Key Variables

| Variable | Description | Type |
|----------|-------------|------|
| `patient_id` | Unique patient identifier | Numeric |
| `department` | Hospital department (Cardiology or Orthopedics) | Categorical |
| `cost_of_care` | Total treatment cost (USD) | Numeric |
| `readmission_days` | Days from discharge to readmission | Numeric |
| `anxiety_before` | Anxiety score before MBSR program (0–10 scale) | Numeric |
| `anxiety_after` | Anxiety score after MBSR program (0–10 scale) | Numeric |
| `satisfaction` | Patient satisfaction score (1–10 scale) | Numeric |

---

# Grading Rubric

Each task specifies point values. Points are allocated as follows:

| Component | Weight | Description |
|-----------|--------|-------------|
| Code correctness | 50% | Code runs without errors and produces correct output |
| Completeness | 30% | All blanks filled in, correct t-test type and arguments used |
| Interpretation | 20% | Thoughtful 1-2 sentence response addressing the question |

**Interpretation Guidance**: Your interpretations should be brief but specific. For example:

- ✅ Good: "The p-value of 0.0032 is less than 0.05, so we reject the null hypothesis. The hospital's average readmission time (26.6 days) is significantly lower than the 30-day national benchmark."
- ❌ Too vague: "The result is significant."
- ❌ Too long: [3+ sentences of analysis]

---

# Assignment Tasks

## Task 1: Identifying the Right t-Test (2 points) {.unnumbered}

For each scenario below, identify which type of t-test is most appropriate. Write **"one-sample"**, **"independent-samples"**, or **"paired-samples"** in the answer box.

**(a)** A hospital administrator wants to know whether the average patient satisfaction score at their facility differs from the national average of 7.5 out of 10. *(0.5 points)*

::: {.callout-note icon=false title="✏️ Your Answer"}
*[Write your answer here]*

:::

**(b)** A researcher compares recovery times between patients who received robotic-assisted surgery and those who had traditional open surgery. The patients are different individuals in each group. *(0.5 points)*

::: {.callout-note icon=false title="✏️ Your Answer"}
*[Write your answer here]*

:::

**(c)** A clinic measures patient blood pressure before and after a 6-month exercise intervention to see if the program was effective. *(0.5 points)*

::: {.callout-note icon=false title="✏️ Your Answer"}
*[Write your answer here]*

:::

**(d)** A quality improvement team wants to determine whether their hospital's average length of stay is different from the CMS (Centers for Medicare & Medicaid Services) benchmark of 4.5 days. *(0.5 points)*

::: {.callout-note icon=false title="✏️ Your Answer"}
*[Write your answer here]*

:::

---

## Task 2: One-Sample t-Test — Readmission Benchmark (2 points) {.unnumbered}

The national benchmark for hospital readmission is **30 days**. Your hospital's quality improvement team wants to determine whether your patients' average readmission time differs from this benchmark.

**Step 1**: Extract the readmission data and compute descriptive statistics.

```{r}
#| label: task2-descriptives
#| eval: false

# Extract readmission days
readmission <- qi_data$___

# Descriptive statistics
cat("Sample size:", length(readmission), "\n")
cat("Sample mean:", round(mean(readmission), 2), "days\n")
cat("Sample SD:", round(sd(readmission), 2), "\n")
cat("Benchmark:", ___, "days\n")
```

**Step 2**: Perform the one-sample t-test and view the results.

```{r}
#| label: task2-ttest
#| eval: false

# Perform one-sample t-test comparing readmission days to the benchmark
result_readmission <- t.test(___, mu = ___)
parameters(result_readmission)
```

**Interpretation Question**: Based on the p-value, is the hospital's average readmission time significantly different from the 30-day benchmark? What does this suggest for the quality improvement team?

::: {.callout-note icon=false title="✏️ Your Answer (1-2 sentences)"}
*[Write your answer here]*

:::

---

## Task 3: Independent-Samples t-Test — Departmental Costs (2 points) {.unnumbered}

Hospital leadership wants to compare treatment costs between the **Cardiology** and **Orthopedics** departments to inform budget allocation decisions.

**Step 1**: Separate the cost data by department and compute descriptive statistics.

```{r}
#| label: task3-descriptives
#| eval: false

# Extract treatment costs for each department
cardiology_cost <- qi_data$cost_of_care[qi_data$department == "___"]
orthopedics_cost <- qi_data$cost_of_care[qi_data$department == "___"]

# Descriptive statistics
cat("Cardiology:\n")
cat("  Mean: $", round(mean(cardiology_cost), 2), "\n")
cat("  SD: $", round(sd(cardiology_cost), 2), "\n")
cat("  n:", length(cardiology_cost), "\n\n")

cat("Orthopedics:\n")
cat("  Mean: $", round(mean(orthopedics_cost), 2), "\n")
cat("  SD: $", round(sd(orthopedics_cost), 2), "\n")
cat("  n:", length(orthopedics_cost), "\n\n")

cat("Difference: $", round(mean(cardiology_cost) - mean(orthopedics_cost), 2))
```

**Step 2**: Perform the independent-samples t-test.

```{r}
#| label: task3-ttest
#| eval: false

# Perform independent-samples t-test
result_cost <- t.test(___, ___, var.equal = ___)
parameters(result_cost)
```

**Interpretation Question**: Is there a statistically significant difference in treatment costs between the two departments? What recommendation would you make to hospital leadership regarding budget allocation?

::: {.callout-note icon=false title="✏️ Your Answer (1-2 sentences)"}
*[Write your answer here]*

:::

---

## Task 4: Visualizing Group Differences (1 point) {.unnumbered}

Create a **box plot** comparing treatment costs between Cardiology and Orthopedics. This visualization supports your analysis from Task 3.

```{r}
#| label: task4-boxplot
#| eval: false

# Fill in the blanks
ggplot(data = qi_data, 
       mapping = aes(x = ___, 
                     y = ___, 
                     fill = ___)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Treatment Costs by Department",
       x = "___",
       y = "___") +
  theme_minimal() +
  theme(legend.position = "none")
```

**Interpretation Question**: Based on the box plot, do the two departments appear to have different cost distributions? How does the visualization support (or contradict) your t-test result from Task 3?

::: {.callout-note icon=false title="✏️ Your Answer (1-2 sentences)"}
*[Write your answer here]*

:::

---

## Task 5: Paired-Samples t-Test — MBSR Program Effectiveness (2 points) {.unnumbered}

The hospital implemented a **Mindfulness-Based Stress Reduction (MBSR)** program to reduce patient anxiety. Anxiety was measured on a 0–10 scale before and after the program for the same 40 patients. The quality team wants to know if the program significantly reduced anxiety.

**Step 1**: Extract the paired data and compute descriptive statistics.

```{r}
#| label: task5-descriptives
#| eval: false

# Extract anxiety scores
anxiety_pre <- qi_data$___
anxiety_post <- qi_data$___

# Descriptive statistics
cat("Before MBSR:\n")
cat("  Mean:", round(mean(anxiety_pre), 2), "\n")
cat("  SD:", round(sd(anxiety_pre), 2), "\n\n")

cat("After MBSR:\n")
cat("  Mean:", round(mean(anxiety_post), 2), "\n")
cat("  SD:", round(sd(anxiety_post), 2), "\n\n")

cat("Mean difference:", round(mean(anxiety_pre - anxiety_post), 2), "points\n")
```

**Step 2**: Perform the paired-samples t-test.

```{r}
#| label: task5-ttest
#| eval: false

# Perform paired-samples t-test
result_anxiety <- t.test(___, ___, paired = ___)
parameters(result_anxiety)
```

**Interpretation Question**: Based on the p-value, did the MBSR program significantly reduce patient anxiety? Would you recommend continuing the program?

::: {.callout-note icon=false title="✏️ Your Answer (1-2 sentences)"}
*[Write your answer here]*

:::

---

## Task 6: Visualizing Paired Data (1 point) {.unnumbered}

Create a **spaghetti plot** (paired line plot) showing individual patient anxiety changes before and after the MBSR program.

```{r}
#| label: task6-spaghetti
#| eval: false

# Create data frame for visualization
anxiety_viz <- data.frame(
  patient = rep(1:___, 2),
  time = factor(rep(c("Before", "After"), each = ___),
                levels = c("Before", "After")),
  score = c(___, ___)
)

# Create spaghetti plot
ggplot(anxiety_viz, aes(x = ___, y = ___, group = ___)) +
  geom_line(alpha = 0.3, color = "gray30") +
  geom_point(aes(color = ___), size = 2, alpha = 0.7) +
  stat_summary(aes(group = 1), fun = mean, geom = "line",
               linewidth = 2, color = "red") +
  stat_summary(aes(group = 1), fun = mean, geom = "point",
               size = 4, color = "red") +
  labs(title = "Patient Anxiety Scores Before and After MBSR Program",
       subtitle = "Red line shows mean change",
       x = "",
       y = "___") +
  theme_minimal() +
  theme(legend.position = "none")
```

**Interpretation Question**: What does the red line (mean) tell you about the overall effect? Do all patients show the same pattern, or is there individual variability?

::: {.callout-note icon=false title="✏️ Your Answer (1-2 sentences)"}
*[Write your answer here]*

:::

---

# Submission Checklist

Before submitting, verify:

- [ ] All 6 tasks are completed
- [ ] All blanks (`___`) are filled in with appropriate values
- [ ] All `#| eval: false` changed to `#| eval: true`
- [ ] All code chunks run without errors
- [ ] All interpretation questions are answered (1-2 sentences each)
- [ ] Visualizations have appropriate titles and axis labels
- [ ] File renders to HTML successfully

---

# Points Summary

| Task | Points | Description |
|------|--------|-------------|
| Task 1 | 2.0 | Identifying the correct t-test type (4 scenarios × 0.5) |
| Task 2 | 2.0 | One-sample t-test (readmission vs. benchmark) |
| Task 3 | 2.0 | Independent-samples t-test (departmental costs) |
| Task 4 | 1.0 | Box plot visualization (group comparison) |
| Task 5 | 2.0 | Paired-samples t-test (MBSR program effectiveness) |
| Task 6 | 1.0 | Spaghetti plot visualization (paired data) |
| **Total** | **10.0** | |

---

**Good luck!**
