---
title: "Final Exam: Healthcare Data Analytics"
subtitle: "MBA642: Business Analytics in Healthcare Management"
author: "Your Name Here"
date: 5/6/2026
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: false
    code-copy: true
    theme: [lecture.scss, cosmo]
    highlight-style: github
    mainfont: "Clear Sans, -apple-system, system-ui, sans-serif"
    fontsize: 1.1em
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Exam Overview

- **Total Points**: 50 points
- **Duration**: 24 hours (12:00 AM - 11:59 PM, 5/6 Wednesday)
- **Policy**: Open-book, open-note (all course materials allowed)
- **Submission**: Upload your completed `.qmd` file to Canvas (no need to submit the rendered HTML). I'll render your `.qmd` to HTML and review the outputs and interpretations there. Before submitting, render it yourself to confirm every code chunk produces output. Missing output (whether from rendering errors OR from leaving `#| eval: false` unchanged) will be treated as incomplete work.

## Instructions

1.  Replace **"Your Name Here"** in the YAML header with your name
2.  Complete all 5 sections below
3.  Fill in the blanks (replace `___` with appropriate code or values)
4.  After completing each code chunk, change `#| eval: false` to `#| eval: true` to run it
5.  Answer all interpretation questions (**1-2 sentences each**)
6.  Render the document to HTML and verify all outputs display correctly before submitting

## Grading Rubric

Each task uses the following weighting:

| Component | Weight | Description |
|----|----|----|
| Code correctness | 50% | Code runs without errors and produces correct output |
| Completeness | 30% | All blanks filled in; correct variables, arguments, and aesthetics used |
| Interpretation | 20% | Brief, specific 1-2 sentence response that references statistics and connects to a healthcare decision |

**Interpretation Guidance:**

- **Good**: "The slope for `physical_therapy_hours` is -0.53 (p \< 0.001), so each additional hour of PT is associated with about half a day shorter recovery, holding other variables constant. This supports investing in expanded post-op PT programs."
- **Too vague**: "PT helps recovery."
- **Too long**: \[3+ sentences restating the obvious\]

------------------------------------------------------------------------

# Loading Packages

Run the code below to load all required R packages. Each dataset is loaded at the start of the section that uses it.

```{r}
#| label: load-packages

library(tidyverse)
library(emmeans)
library(glmnet)
```

------------------------------------------------------------------------

# Section 1: Multiple Linear Regression with Dummy Variables (12 points)

*Flyer's chief of orthopedics wants to know which patient characteristics drive post-operative recovery time. Build a multiple regression model that controls for age, BMI, diabetes status, surgery type, and physical therapy hours, and use it to compare recovery times across surgery types.*

## About the Data: Flyer Orthopedic Surgical Recovery {.unnumbered}

Flyer Orthopedic Hospital tracked **250 surgical patients** undergoing knee, hip, or shoulder surgery. The hospital's quality team wants to (1) understand what drives **post-operative recovery time**, (2) test whether **physical therapy** has different effects across surgery types, and (3) build a model to flag patients at risk of **post-operative complications**.

Load the data using the code below:

```{r}
#| label: load-surgery-data

surgery_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/exam2_surgery_data.csv")

surgery_data <- surgery_data |>
  mutate(
    surgery_type = factor(surgery_type, levels = c("Knee", "Hip", "Shoulder")),
    diabetes     = factor(diabetes,     levels = c("No", "Yes"))
  )

head(surgery_data, 3)
```

| Variable | Type | Description |
|----|----|----|
| `patient_id` | Integer | Unique patient identifier |
| `age` | Continuous | Patient age in years |
| `sex` | Categorical | Patient sex (Male, Female) |
| `bmi` | Continuous | Body Mass Index |
| `diabetes` | Categorical | Diabetes status (No, Yes) |
| `surgery_type` | Categorical | Type of surgery (Knee, Hip, Shoulder) |
| `num_comorbidities` | Count | Number of comorbid conditions |
| `surgery_minutes` | Continuous | Surgery duration (minutes) |
| `physical_therapy_hours` | Continuous | Weekly post-op PT hours |
| `recovery_days` | Continuous | Days to full recovery — **outcome (Sections 1-2)** |
| `complication` | Binary | Post-op complication (1 = Yes, 0 = No) — **outcome (Section 3)** |

## Task 1a: Visualization (1.5 points) {.unnumbered}

Create a boxplot comparing `recovery_days` across the three surgery types. Use different fill colors for each surgery type.

```{r}
#| label: task1a-boxplot
#| eval: false

ggplot(surgery_data, aes(x = ___, y = ___, fill = ___)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Recovery Days by Surgery Type",
    x     = "Surgery Type",
    y     = "Recovery Days"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Task 1b: Multiple Regression (5 points) {.unnumbered}

Fit a multiple linear regression model predicting `recovery_days` from `age`, `bmi`, `diabetes`, `surgery_type`, and `physical_therapy_hours`.

```{r}
#| label: task1b-model
#| eval: false

model_recovery <- lm(___ ~ ___ + ___ + ___ + ___ + ___,
                     data = ___)
summary(___)
```

**Interpretation (1-2 sentences each):**

For each question below, interpret the coefficient using the **Direction–Significance–Magnitude (DSM)** framework: report (1) the **direction** (positive or negative), (2) the **statistical significance** (cite the p-value), and (3) the **magnitude** of the effect in plain language.

1.  Interpret the coefficient on `physical_therapy_hours` (DSM). For magnitude, state how many days recovery changes for each additional weekly PT hour, holding other variables constant.
2.  Interpret the coefficient on `surgery_typeHip` (DSM). For magnitude, state how many days Hip patients differ from the **reference category** (Knee patients) on average, holding other variables constant.
3.  Interpret the coefficient on `diabetesYes` (DSM). For magnitude, state how many days diabetic patients differ from non-diabetic patients on average, holding other variables constant.

::: {.callout-note icon="false" title="Your Answer (1-2 sentences each)"}
1.  Answer 1:

2.  Answer 2:

3.  Answer 3:
:::

## Task 1c: Pairwise Comparisons with `emmeans` (2.5 points) {.unnumbered}

Use `emmeans()` and `pairs()` to perform Tukey-adjusted pairwise comparisons of `surgery_type`, holding the other predictors constant.

```{r}
#| label: task1c-emmeans
#| eval: false

emm_surgery <- emmeans(___, ~ ___)
pairs(emm_surgery, adjust = "___")
```

**Interpretation:**

4.  Using the `emmeans` pairwise comparison, are **Hip** and **Shoulder** patients significantly different in recovery time? If they are significantly different, which group takes significantly longer to recover; if not, state that they are not significantly different.

::: {.callout-note icon="false" title="Your Answer (1-2 sentences)"}
4.  Answer 4:
:::

## Task 1d: Predicting Recovery for Multiple Patients (3 points) {.unnumbered}

The orthopedic team wants to forecast recovery times for the next four upcoming surgical patients in order to plan post-op resources. The team will flag patients for **extended-stay support** (additional nursing visits, social-work check-ins, and bed reservations) if their predicted recovery time exceeds **30 days**.

The four upcoming patient profiles are listed below. Translate the table into the `patients` data frame in the code chunk, then use your model from Task 1b to predict each patient's recovery time.

| Profile   | Age | BMI | Diabetes | Surgery Type | Weekly PT hours |
|-----------|-----|-----|----------|--------------|-----------------|
| Patient A | 45  | 24  | No       | Knee         | 20              |
| Patient B | 50  | 25  | No       | Shoulder     | 22              |
| Patient C | 65  | 30  | Yes      | Hip          | 18              |
| Patient D | 78  | 33  | Yes      | Hip          | 10              |

```{r}
#| label: task1d-predict
#| eval: false

patients <- data.frame(
  Profile                = c("Patient A", "Patient B", "Patient C", "Patient D"),
  age                    = c(___, ___, ___, ___),
  bmi                    = c(___, ___, ___, ___),
  diabetes               = c("___", "___", "___", "___"),
  surgery_type           = c("___", "___", "___", "___"),
  physical_therapy_hours = c(___, ___, ___, ___)
)

patients$predicted_days <- round(predict(___, newdata = ___), 2)
patients
```

**Interpretation (1-2 sentences each):**

5.  What is the predicted recovery time for **Patient C**?
6.  Using the **30-day threshold**, which patient(s) would the orthopedic team flag for extended-stay support? List all that apply.

::: {.callout-note icon="false" title="Your Answer (1-2 sentences each)"}
5.  Answer 5:

6.  Answer 6:
:::

------------------------------------------------------------------------

# Section 2: Linear Regression with Interaction Effects (8 points)

*Flyer's PT director hypothesizes that PT does not help all surgery types equally; some patients may benefit much more from extra PT than others. Test this hypothesis using an interaction model.*

## Task 2a: Visualization with Group-Specific Trend Lines (1.5 points) {.unnumbered}

Create a scatterplot that visualizes how **recovery days** change with **weekly physical therapy hours** across the three surgery types. Include a linear trend line for each surgery type.

```{r}
#| label: task2a-plot
#| eval: false

ggplot(surgery_data, aes(x = ___, y = ___, color = ___)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "___", se = FALSE) +
  labs(
    title = "Recovery Days vs. Weekly PT Hours by Surgery Type",
    x     = "Weekly Physical Therapy Hours",
    y     = "Recovery Days",
    color = "Surgery Type"
  ) +
  theme_minimal()
```

## Task 2b: Interaction Model (2 points) {.unnumbered}

Fit a regression model to test whether the effect of weekly PT hours on recovery time **depends on** surgery type. Use only these two predictors (no age, BMI, or diabetes).

```{r}
#| label: task2b-model
#| eval: false

# Hint: use * to include both main effects AND the interaction in one term
model_interaction <- lm(___ ~ ___ * ___, data = ___)
summary(___)
```

## Task 2c: Interpretation (4.5 points) {.unnumbered}

**Interpretation (1-2 sentences each):**

7.  Interpret the coefficient on `physical_therapy_hours:surgery_typeHip`. State its **direction** (positive or negative) and what it tells you about how the PT effect for **Hip** patients differs from the PT effect for **Knee** patients (the reference category) — is PT more effective, less effective, or equally effective for Hip patients?
8.  Interpret the coefficient on `physical_therapy_hours:surgery_typeShoulder`. State its **direction** and what it tells you about how the PT effect for **Shoulder** patients differs from the PT effect for **Knee** patients.
9.  For **Hip** patients, by how many days does recovery time change for each additional weekly PT hour? What about for **Knee** patients? Show the calculation for each.

::: {.callout-note icon="false" title="Your Answer (1-2 sentences each)"}
7.  Answer 7:

8.  Answer 8:

9.  Answer 9:
:::

------------------------------------------------------------------------

# Section 3: Logistic Regression (12 points)

*The patient safety committee wants a model to flag surgical patients at risk for post-operative complications. They will eventually use the model to triage which patients receive enhanced monitoring.*

## Task 3a: Simple Logistic Regression: Age and Complications (3 points) {.unnumbered}

The committee suspects that **age** is a key driver of post-operative complications; older patients tend to be frailer and recover less reliably. Before building a more comprehensive model, you want to quantify how strongly age alone predicts complication risk so the team has a baseline benchmark.

**Step 1**: Fit a simple logistic regression predicting `complication` from `age`.

```{r}
#| label: task3a-model
#| eval: false

model_simple <- glm(___ ~ ___,
                    data   = ___,
                    family = ___)
summary(___)
```

**Step 2**: Calculate and report the odds ratio for `age`.

```{r}
#| label: task3a-or
#| eval: false

or_age <- exp(coef(___)["age"])
round(or_age, 4)
```

**Interpretation (1-2 sentences each):**

10. Interpret the `age` predictor using the **Direction–Significance–Magnitude (DSM)** framework: state (1) the **direction** (does age increase or decrease the odds of complication?), (2) the **statistical significance** (cite the p-value), and (3) the **magnitude** (as age increases by 1 year, how do the odds of having a complication change)?
11. The odds of complication for a **70-year-old** patient are approximately how many times the odds for a **50-year-old** patient (a 20-year age gap)?

::: {.callout-note icon="false" title="Your Answer (1-2 sentences each)"}
10. Answer 10:

11. Answer 11:
:::

## Task 3b: Multiple Logistic Regression (3.5 points) {.unnumbered}

The committee suspects that complication risk is shaped by more than age alone; comorbidities, surgery duration, and diabetes status all plausibly matter. Build a multivariable model so the team can see which factors affect complications after controlling for the others.

**Step 1**: Fit a multiple logistic regression predicting `complication` from `age`, `num_comorbidities`, `surgery_minutes`, and `diabetes`.

```{r}
#| label: task3b-model
#| eval: false

model_complication <- glm(___ ~ ___ + ___ + ___ + ___,
                          data   = ___,
                          family = ___)
summary(___)
```

**Step 2**: Compute odds ratios for all predictors.

```{r}
#| label: task3b-or-table
#| eval: false

round(exp(coef(___)), 4)
```

**Interpretation (1-2 sentences each):**

12. Which predictors are statistically significant at α = 0.05? For each significant predictor, indicate whether it **increases** or **decreases** the odds of complication.
13. Interpret the odds ratio for `diabetesYes`. By what percentage do the odds of complication change for a diabetic patient compared to a non-diabetic patient with the same age, comorbidities, and surgery duration?

::: {.callout-note icon="false" title="Your Answer (1-2 sentences each)"}
12. Answer 12:

13. Answer 13:
:::

## Task 3c: Predicted Probabilities for Patient Profiles (2.5 points) {.unnumbered}

The safety committee wants to see predicted complication probabilities for three upcoming surgical cases that span the typical risk spectrum (low, moderate, high). These predictions will inform whether each patient should be routed to **enhanced monitoring**.

Use `model_complication` from Task 3b to predict the probability of complication for each of the three patients below.

- **Patient A**: 50 years old, 1 comorbidity, 90-minute surgery, no diabetes
- **Patient B**: 65 years old, 3 comorbidities, 130-minute surgery, no diabetes
- **Patient C**: 78 years old, 5 comorbidities, 180-minute surgery, has diabetes

```{r}
#| label: task3c-profiles
#| eval: false

patient_profiles <- data.frame(
  Profile           = c("Patient A", "Patient B", "Patient C"),
  age               = c(50, ___, ___),
  num_comorbidities = c(___, 3, ___),
  surgery_minutes   = c(___, ___, 180),
  diabetes          = c("No", "___", "___")
)

patient_profiles$predicted_prob <- round(
  predict(___, newdata = patient_profiles, type = "___"),
  4
)
patient_profiles
```

**Interpretation:**

14. List the predicted complication probability for each patient (Patient A, Patient B, and Patient C). Then, using **0.50** as the threshold for "enhanced monitoring," identify which patient(s) would be flagged.

::: {.callout-note icon="false" title="Your Answer (1-2 sentences)"}
14. Answer 14:
:::

## Task 3d: Confusion Matrix and Accuracy (3 points) {.unnumbered}

Before the safety committee deploys `model_complication` for triage, they need to know how often it is right and where it goes wrong. Apply the model back to the original 250 patients, classify each one as "predicted complication" vs. "no complication" using the 0.50 probability threshold, and build a confusion matrix to evaluate overall accuracy and error patterns.

```{r}
#| label: task3d-cm
#| eval: false

# Add predictions to the data
surgery_data$pred_prob  <- predict(___, type = "___")
surgery_data$pred_class <- ifelse(surgery_data$pred_prob > ___, 1, 0)

# Confusion matrix
confusion_matrix <- table(
  Predicted = factor(surgery_data$___, levels = c(0, 1)),
  Actual    = factor(surgery_data$___, levels = c(0, 1))
)
confusion_matrix

# Accuracy
tn <- confusion_matrix[1, 1]
fp <- confusion_matrix[2, 1]
fn <- confusion_matrix[1, 2]
tp <- confusion_matrix[2, 2]

accuracy <- (___ + ___) / (tp + tn + fp + fn)
round(accuracy, 4)
```

**Interpretation (1-2 sentences each):**

15. Report the accuracy. Out of all 250 patients, how many were correctly classified?
16. Report how many patients who **actually had a complication** the model **failed to flag** (i.e., actual = 1 but predicted = 0; the false negatives). Between **false positives** and **false negatives**, which is more concerning for the patient safety committee, and why in one sentence?

::: {.callout-note icon="false" title="Your Answer (1-2 sentences each)"}
15. Answer 15:

16. Answer 16:
:::

------------------------------------------------------------------------

# Section 4: Conjoint Analysis: Telehealth Platform Design (10 points)

*TeleHealth Plus is choosing between several platform designs that vary on four attributes: wait time, video quality, provider type, and visit cost. Use the rating-based conjoint data to estimate part-worths, attribute importance, willingness-to-pay, and predicted market shares.*

## About the Data: TeleHealth Plus Conjoint Survey {.unnumbered}

A **national health insurance company** is considering adding a new telehealth platform as a member benefit. They conducted a **rating-based conjoint survey** of 100 enrollees to estimate preferences across four design attributes. Each respondent rated 16 hypothetical platforms.

Load the data using the code below:

```{r}
#| label: load-telehealth-data

telehealth_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/exam2_telehealth_ratings.csv")

head(telehealth_data, 3)
```

| Variable | Type | Description |
|----|----|----|
| `respondent_id` | Integer | Unique respondent identifier |
| `wait_5min` | Binary | Short wait indicator (1 = 5 min wait, 0 = 30 min wait) |
| `video_hd` | Binary | Video quality (1 = HD, 0 = SD) |
| `specialist` | Binary | Provider type (1 = Specialist, 0 = General Practitioner) |
| `visit_cost` | Continuous | Out-of-pocket cost per visit (\$25 or \$75) |
| `rating` | Continuous | Respondent's rating (1-10) — **outcome variable** |

## Task 4a: Rating-Based Conjoint Model (2 points) {.unnumbered}

The first step is to translate the raw 1–10 ratings into **part-worth utilities** (numerical values the insurer can use to quantify how much each attribute drives respondent preference).

Fit a linear regression model predicting `rating` from the four attributes.

```{r}
#| label: task4a-model
#| eval: false

model_telehealth <- lm(___ ~ ___ + ___ + ___ + ___,
                       data = ___)
summary(___)
```

**Interpretation:**

17. Report the **sign** (positive or negative) of each coefficient. Do all four signs make intuitive sense given what each attribute represents? Explain briefly.

::: {.callout-note icon="false" title="Your Answer (1-2 sentences)"}
17. Answer 17:
:::

## Task 4b: Attribute Importance (2.5 points) {.unnumbered}

The product team has a fixed budget and cannot perfect every attribute at once. They want to know **which attributes matter most to respondents** so they can focus engineering, contracting, and marketing investment effectively. Translate the part-worths from Task 4a into relative importance percentages.

Compute utility ranges and importance percentages.

```{r}
#| label: task4b-importance
#| eval: false

coef_uc <- coef(model_telehealth)

range_wait       <- abs(coef_uc["___"] * 1)            # Binary: 5 min vs 30 min
range_video      <- abs(coef_uc["___"] * 1)            # Binary: HD vs SD
range_specialist <- abs(coef_uc["___"] * 1)            # Binary: Specialist vs GP
range_cost       <- abs(coef_uc["___"] * (75 - 25))    # $50 cost difference

total_range <- range_wait + range_video + range_specialist + range_cost

importance_tbl <- data.frame(
  attribute      = c("Wait Time", "Video Quality", "Provider Type", "Visit Cost"),
  utility_range  = c(range_wait, range_video, range_specialist, range_cost),
  importance_pct = c(range_wait, range_video, range_specialist, range_cost) /
                     total_range * 100
)
importance_tbl
```

**Interpretation:**

18. Identify the **most important** and **least important** attribute, and report each importance percentage (rounded to one decimal place).

::: {.callout-note icon="false" title="Your Answer"}
18. Answer 18:
:::

## Task 4c: Willingness to Pay (2.5 points) {.unnumbered}

The insurer's product committee needs to know what each attribute upgrade is **worth in dollars**, so they can compare it directly to the cost of providing it (e.g., contracting specialists, upgrading video infrastructure, expanding hours). Convert each non-price part-worth into a willingness-to-pay (WTP) figure.

Compute the willingness-to-pay (WTP) for each non-price attribute. Recall the formula: $$\text{WTP} = -\frac{\beta_{\text{attribute}}}{\beta_{\text{price}}}$$

```{r}
#| label: task4c-wtp
#| eval: false

wtp_wait       <- -coef_uc["___"] / coef_uc["___"]
wtp_video      <- -coef_uc["___"] / coef_uc["___"]
wtp_specialist <- -coef_uc["___"] / coef_uc["___"]

wtp_table <- data.frame(
  Attribute = c("Short Wait (5 min vs 30 min)",
                "HD Video (vs SD)",
                "Specialist (vs GP)"),
  WTP       = paste0("$", round(c(wtp_wait, wtp_video, wtp_specialist), 2))
)
wtp_table
```

**Interpretation:**

19. Report the WTP for **specialist provider**. If hiring specialists adds \$50 per visit to TeleHealth's costs, is offering specialist visits financially justified based on respondent preferences? Explain by comparing WTP to the \$50 cost.

::: {.callout-note icon="false" title="Your Answer (1-2 sentences)"}
19. Answer 19:
:::

## Task 4d: Market Share Prediction (3 points) {.unnumbered}

A **Premium** telehealth product from a competitor is already on the market. The insurer is considering entering the market with a lower-cost **Basic** alternative for its enrollees, but only if the Basic product would capture a meaningful share of enrollee preference. Use the conjoint model to simulate the share each product would capture if enrollees had to choose between the two:

- **Premium** (competitor, already on market): 5-min wait, HD video, Specialist, \$75 per visit
- **Basic** (insurer's proposed entry): 30-min wait, SD video, GP, \$25 per visit

**Step 1**: Predict ratings for both platforms.

```{r}
#| label: task4d-predict
#| eval: false

premium <- data.frame(wait_5min = ___, video_hd = ___,
                      specialist = ___, visit_cost = ___)
basic   <- data.frame(wait_5min = ___, video_hd = ___,
                      specialist = ___, visit_cost = ___)

rating_premium <- predict(___, newdata = premium)
rating_basic   <- predict(___, newdata = basic)

cat("Premium predicted rating:", round(rating_premium, 2), "\n")
cat("Basic predicted rating:  ", round(rating_basic,   2), "\n")
```

**Step 2**: Calculate proportional market shares.

```{r}
#| label: task4d-share
#| eval: false

total_rating <- rating_premium + rating_basic
share_premium <- ___ / total_rating
share_basic   <- ___ / total_rating

market_share <- data.frame(
  Platform     = c("Premium", "Basic"),
  Rating       = round(c(rating_premium, rating_basic), 2),
  Market_Share = paste0(round(c(share_premium, share_basic) * 100, 1), "%")
)
market_share
```

**Interpretation (1-2 sentences each):**

20. Report each product's predicted market share. Which product captures the larger share?
21. The insurer's market-entry rule is to launch the Basic alternative **only if its predicted share exceeds 20%**. Based on your computed market share for Basic, does it clear the 20% threshold? State **Yes** or **No**, and recommend whether the insurer should launch.

::: {.callout-note icon="false" title="Your Answer (1-2 sentences each)"}
20. Answer 20:

21. Answer 21:
:::

------------------------------------------------------------------------

# Section 5: Penalized Regression: Ridge & LASSO (8 points)

*Dayton General has 18 candidate predictors of 30-day post-discharge cost. Some are clinical (likely meaningful); others are administrative fields recorded for every patient (likely noise). You will use Ridge and LASSO regression (both with 10-fold cross-validation) to compare a model that keeps all predictors against one that performs automatic feature selection.*

## About the Data: Dayton General 30-Day Cost {.unnumbered}

Dayton General Hospital wants to predict **30-day post-discharge cost** for 400 recently discharged patients. The dataset includes **18 candidate predictors** (a mix of clinical fields the team believes matter and administrative fields recorded for every patient).

Load the data using the code below:

```{r}
#| label: load-cost-data

cost_data <- read.csv("https://raw.githubusercontent.com/mba642-spring-26/mba642-spring-26.github.io/refs/heads/main/datasets/exam2_cost_data.csv")

head(cost_data, 3)
```

| Variable | Type | Description |
|----|----|----|
| `patient_id` | Integer | Unique patient identifier |
| `age` | Continuous | Patient age in years |
| `num_comorbidities` | Count | Number of comorbid conditions |
| `asa_score` | Ordinal | ASA physical status (1-4) |
| `emergency_admit` | Binary | Emergency admission (1 = Yes) |
| `los_days` | Continuous | Length of stay (days) |
| `num_procedures` | Count | Procedures during admission |
| `preop_albumin` | Continuous | Pre-op albumin (g/dL) |
| `has_ckd` | Binary | Chronic kidney disease (1 = Yes) |
| `admit_day_of_week` | Integer | Day of admission (1-7) |
| `assigned_nurse_id` | Integer | ID of primary nurse (1-15) |
| `room_floor` | Integer | Hospital floor (1-5) |
| `bed_number` | Integer | Bed number (1-30) |
| `visitor_count` | Count | Visitors on admission day |
| `chart_template_version` | Integer | EHR chart template version (1-3) |
| `distance_to_nurse_m` | Continuous | Meters from room to nurse station |
| `has_window` | Binary | Room has a window (1 = Yes) |
| `discharge_planner_id` | Integer | Assigned discharge planner (1-10) |
| `intake_form_pages` | Integer | Pages of intake paperwork (3-8) |
| `cost_30day` | Continuous | 30-day post-discharge cost (USD) — **outcome variable** |

## Task 5a: Data Preparation (1 point) {.unnumbered}

To accurately evaluate how well a model predicts **new** patients, you cannot use the same data both to fit the model and to evaluate it. Hold out a portion of the data as a **test set** that the model never uses during fitting, then evaluate it there. `glmnet` also requires its predictors in **matrix** form (not a data frame), so do that conversion here too.

Split the cost data into 70% training and 30% test sets, then build matrix inputs for `glmnet`.

```{r}
#| label: task5a-prep
#| eval: false

# Use this seed for reproducibility; do not change it.
set.seed(2026)

predictor_vars <- cost_data |>
  select(-patient_id, -cost_30day) |>
  names()

# Train/test split
train_idx  <- sample(1:nrow(cost_data), size = ___ * nrow(cost_data))
train_data <- cost_data[train_idx, ]
test_data  <- cost_data[-train_idx, ]

x_train <- as.matrix(train_data[, predictor_vars])
y_train <- train_data$___

x_test  <- as.matrix(test_data[, predictor_vars])
y_test  <- test_data$___

cat("Training set:", nrow(x_train), "patients,", ncol(x_train), "predictors\n")
cat("Test set:    ", nrow(x_test), "patients\n")
```

## Task 5b: Ridge Regression with Cross-Validation (2 points) {.unnumbered}

Fitting an OLS regression with 18 predictors (many of which are likely correlated administrative noise) risks overfitting and unstable coefficients. **Ridge regression** addresses this by shrinking coefficients toward zero, trading a bit of bias for substantially lower variance. The strength of shrinkage is controlled by **lambda**, which you choose via 10-fold cross-validation.

Fit a Ridge model with 10-fold cross-validation, plot the CV curve, and compute the test RMSE at `lambda.1se`.

```{r}
#| label: task5b-ridge
#| eval: false

ridge_cv <- cv.glmnet(
  x      = ___,
  y      = ___,
  alpha  = ___,
  nfolds = 10
)

plot(___)
title("Ridge Regression: Cross-Validation Results", line = 3)

ridge_coef <- coef(ridge_cv, s = "___")
ridge_pred <- predict(ridge_cv, s = "___", newx = ___)
ridge_rmse <- sqrt(mean((y_test - ridge_pred)^2))

cat("Non-zero Ridge coefficients:",
    sum(ridge_coef[-1] != 0), "out of", length(predictor_vars), "\n")
cat("Ridge test RMSE: $", round(ridge_rmse, 0), "\n", sep = "")
```

## Task 5c: LASSO Regression with Cross-Validation (2 points) {.unnumbered}

Ridge keeps every predictor in the model, even ones that contribute almost nothing, making the model harder for clinicians to interpret and operationalize. **LASSO regression** uses a different penalty (L1) that drives weak coefficients to **exactly zero**, performing automatic variable selection. The result is a sparser, simpler model that tends to keep only the predictors that genuinely matter.

Fit a LASSO model on the same data with 10-fold cross-validation, identify the selected variables, and compute test RMSE at `lambda.1se`.

```{r}
#| label: task5c-lasso
#| eval: false

lasso_cv <- cv.glmnet(
  x      = ___,
  y      = ___,
  alpha  = ___,
  nfolds = 10
)

plot(___)
title("LASSO Regression: Cross-Validation Results", line = 3)

lasso_coef <- coef(lasso_cv, s = "___")

# Selected variables (non-zero, excluding intercept)
selected_vars <- rownames(lasso_coef)[which(lasso_coef != 0)][-1]

# Test RMSE
lasso_pred <- predict(lasso_cv, s = "___", newx = ___)
lasso_rmse <- sqrt(mean((y_test - lasso_pred)^2))

cat("LASSO selected", length(selected_vars), "of", length(predictor_vars),
    "predictors:\n", paste(selected_vars, collapse = ", "), "\n\n")
cat("LASSO test RMSE: $", round(lasso_rmse, 0), "\n", sep = "")
```

## Task 5d: Comparison and Recommendation (3 points) {.unnumbered}

Dayton General has to choose **one** model to deploy in their planning workflow. The choice involves a trade-off: Ridge tends to keep every predictor (often slightly more accurate but harder for staff to operate), while LASSO keeps only a handful (simpler but sometimes less accurate). Compare both on the same test set and decide which model best fits the hospital's operational needs.

Build a side-by-side comparison of Ridge and LASSO and answer the recommendation questions.

```{r}
#| label: task5d-compare
#| eval: false

comparison <- data.frame(
  Model               = c("Ridge", "LASSO"),
  Test_RMSE_USD       = round(c(ridge_rmse, lasso_rmse), 0),
  Predictors_Retained = c(sum(ridge_coef[-1] != 0), length(selected_vars)),
  Total_Candidates    = length(predictor_vars)
)
comparison
```

**Interpretation (1-2 sentences each):**

22. How many of the 18 candidate predictors does each model retain? Why does Ridge keep all of them, while LASSO sets several coefficients to **exactly zero** (what does each penalty actually do to the coefficients)?

23. Which model has a lower RMSE? Is the difference practically meaningful for hospital cost forecasting (i.e., relative to the mean 30-day cost a patient generates, which is roughly **\$22,000**)?

24. Dayton General will adopt the LASSO model **only if both** of the following are true:

    - 

      (a) LASSO's test RMSE is **within \$200** of Ridge's test RMSE

    - 

      (b) LASSO retains **fewer predictors** than Ridge

    For LASSO, state **Yes** or **No** for each of (a) and (b) using your numbers, then state your **final recommendation** (Ridge or LASSO) in one sentence.

::: {.callout-note icon="false" title="Your Answer (1-2 sentences each)"}
22. Answer 22:

23. Answer 23:

24. Answer 24:
:::

------------------------------------------------------------------------

# Submission Checklist

Before submitting, verify:

- [ ] Your name is in the YAML header
- [ ] All 5 sections are completed
- [ ] All blanks (`___`) are filled in with appropriate values
- [ ] All `#| eval: false` changed to `#| eval: true`
- [ ] All code chunks run without errors
- [ ] All interpretation questions are answered (1-2 sentences each)
- [ ] Visualizations have appropriate titles and axis labels
- [ ] Document renders to HTML successfully (a rendering error indicates a code error — fix it before submitting)

------------------------------------------------------------------------

# Points Summary

| Section   | Topic                                                  | Points |
|-----------|--------------------------------------------------------|--------|
| Section 1 | Multiple Linear Regression with Dummy Variables        | 12     |
| Section 2 | Linear Regression with Interaction Effects             | 8      |
| Section 3 | Logistic Regression (simple, multiple, classification) | 12     |
| Section 4 | Conjoint Analysis (importance, WTP, market share)      | 10     |
| Section 5 | Penalized Regression (Ridge & LASSO)                   | 8      |
| **Total** |                                                        | **50** |

------------------------------------------------------------------------

**Good luck!**
